### 2010-05-06 doc/typing_rules.tex
Ne pas regarder ce fichier!
C'est pas fini d'expliquer et en plus ça gâcherait tout l'effet de surprise :)

git-svn-id: http://gasp.dev.regis-gianas.com/svn/trunk@40 403e655e-980a-47bd-8c60-1ee5348bb6b7

\documentclass[12pt]{article}
\usepackage{mathpartir}
\usepackage{ucs}
\usepackage[utf8x]{inputenc}
\usepackage{mathpartir}
\usepackage{color}
%\usepackage[pdftex,backref=page,colorlinks=true]{hyperref}
\usepackage{amsmath, amstext, amsthm, amsfonts}
\usepackage{stmaryrd}
\usepackage[protrusion=true,expansion=true]{microtype}

\title{Typing rules for the data-oriented patch system}

\newcommand{\sort}{\textsf{s}}
\newcommand{\gor}{\ |\ }
\newcommand{\gdecl}[2]{{#1}\ &::=\ {#2}}
\newcommand{\subst}[2]{\{#1/#2\}}
\begin{document}

\maketitle

\paragraph{Introduction}

Representing syntax and logics is nicely done in a \emph{logical
  framework} like LF: both the syntactic elements and the typing
derivations sit in the same tree structure, and both can be rechecked
at the same time. The PTSes are a concise presentation of a family of
logical frameworks.

For the purpose of incremental type checking though, our needs are a
little bit different: first, we need to record, that is to name all
intermediate values of our developments, so as to be able to address
and reuse them multiple times. Secondly, we need to make sure that
intermediate values are not recorded twice, which would imply that we
will verify the same subterm twice (precisely what we are trying to
avoid). These two conditions together entail that we are looking to
represent syntactic and typing objects as a \emph{DAG} rather than a
tree, and moreover a DAG with \emph{maximal sharing} among its
sub-DAGs.

Let's take the PTS approach, and tailor it to our needs. We make
several changes to the usual presentation:
\begin{itemize}
\item In a first iteration of our project, we do not need computations
  to take place within our DAGs. No $\lambda$'s, no conversion
  rule. We stratify the syntax of the system into terms and types.
\item Compound terms are what we want to avoid. Every term should be a
  flat application of variables. 
\item Finally we need a way to record these intermediate definitions:
  we introduce a new kind of product, singleton products, which
  contain the definition of the new object along with its name and
  type. Environments are extended for these definitions.
\end{itemize}

\paragraph{Parameters}

As for a regular PTS, our system depends on three variables:

\begin{itemize}
  \item$\mathcal S$ a set of sorts \sort,
  \item$\mathcal A \in \mathcal S^2$ the axioms of the system,
  \item$\mathcal R \in \mathcal S^3$ the allowed products.
\end{itemize}

\paragraph{Syntax} There are three syntactic categories: applicative
terms $a$, types $t$ and environments $\Gamma$. Environments look like
list, but in fact they aren't (see below).

\begin{align*}
 \gdecl{a}{x \gor a\ x } \\
 \gdecl{t}{a \gor \sort \gor (x:t)\to t \gor (x=a:t)\to t} \\
 \gdecl{\Gamma}{\cdot \gor \Gamma[x:t] \gor \Gamma[x=a:t]}
\end{align*}

Following Agda's syntax, we denote by $(x:t)\to u$ the dependent
product (elsewhere written with a $\Pi$). The new construct
$(x=a:t)\to u$ is a singleton type in argument position, whose value
is $a$. Such a singleton product should be read as ``the function
taking as argument any values syntactically equal to $a$ of type $t$
(\emph{i.e.} a pedantic way to say $a$). We write as usual
$(x_1:t_1)(x_2:t_2)\ldots(x_n:t_n)\to t$ for
$(x_1:t_1)\to(x_2:t_2)\to\ldots\to(x_n:t_n)\to t$. The lookup in the
environment is written either $\Gamma(x):t$, denoting ``$\Gamma$
contains a declaration $[x:t]$ \emph{or} a definition $[x=a:t]$'', or
$\Gamma(x)=a:t$ denoting ``$\Gamma$ contains a definition
$[x=a:t]$''. Capture-avoiding renaming of variables in a term $t$ is
written $t\subst{x}{y}$ for ``replace all occurences of $x$ by $y$''
and is defined the usual way.

We maintain the strong invariant that the environment does not contain
two variables of the same name, nor two syntactically equal
declarations (the $a$ part is unique). More on that in the
implementation part.

\paragraph{Typing} The typing of our system is a restriction of the
PTS's typing rules with the additional two rules for typing a singleton
product and the application of a singleton argument (featuring the
syntactic verification in the environment). It relies on two distincts
jugements:
\begin{itemize}
\item $\Gamma\vdash a : t$ ($a$ has type $t$ in $\Gamma$),
\item $\Gamma\vdash t : \sort$ ($t$ has sort $s$ in $\Gamma$).
\end{itemize}
The rules are:
\begin{mathpar}
  \infer[Axiom]{
  }{\Gamma\vdash\sort_1:\sort_2}
  \quad(\sort_1,\sort_2)\in\mathcal A
  \and
  \infer[Init]{
    \Gamma(x):t
  }{\Gamma\vdash x:t}
  \and
  \infer[Prod]{\Gamma\vdash t:\sort_1 \and 
    \Gamma[x:t]\vdash u:\sort_2
  }{\Gamma\vdash(x:t)\to u : \sort_3}
  \quad(\sort_1,\sort_2,\sort_3)\in\mathcal R
  \and
  \infer[Sing]{\Gamma\vdash a:t \and
    \Gamma\vdash t:\sort_1 \and
    \Gamma[x=a:t]\vdash u:\sort_2
  }{\Gamma\vdash(x=a:t)\to u : \sort_3} 
  \quad(\sort_1,\sort_2,\sort_3)\in\mathcal R
  \and
  \infer[AppP]{
    \Gamma\vdash a:(y:t)\to u \and \Gamma(x):t
    }{\Gamma\vdash a\ x : u\subst{x}{y}}
  \and
  \infer[AppS]{
    \Gamma\vdash a:(y=b:t)\to u \and \Gamma(x)=b:t
    }{\Gamma\vdash a\ x : u\subst{x}{y}}
\end{mathpar}

A term $t$ represents a state of the repository at a given moment. It
contains the declaration of the object language's syntax, its typing
rules, the syntactic objects and derivations recorded in it along with
the patches that let the repository evolve. If $\cdot\vdash t:\sort$,
then the repository is well-typed.

\paragraph{An example}

\paragraph{Implementation}

The main problem when it comes with implementation is to guarantee the
unicity of declarations and definitions in the environment. One
approach would be to use the locally-nameless method, along with a
special type of free names: A term is stored in deBruijn syntax, but
as soon as a variable is introduced, it takes a real name.

How to choose the names in order for each of them to be uniquely bound
to their content (for definitions)? Let's follow \textsf{git} on this:
each definition is given for name the hash of its content
(definiens). In turn, the content of a definition is itself an
application of variables, which have been given unique names earlier,
so taking the hash of the content guarantees that the whole sub-DAG is
contained in the name.

The receipt to choose a name at introduction-time (\textsc{Sing} or
\textsc{Prod}) is:
\begin{itemize}
\item If we introduce a \emph{declaration}, it has no content. Let's
  assign it a big random hash for a name: it will ensure its
  uniqueness.
\item If we introduce a \emph{definition}, then we compute the hash of
  its content. If this hash is already present in the environment, we
  don't introduce anything and take this hash as a name. Otherwise we
  introduce the definition and assign it the hash as a name.
\end{itemize}

\end{document}

### 2010-05-11 doc/typing_rules.tex
Definition de la deuxième version des règles de typages.
On a simplifié un certain nombre de caractéristiques inutiles du langage
précédent: fusion atome/transformer, plus d'arité mais un système
curryfié, plus de conclusion multiple possible.

git-svn-id: http://gasp.dev.regis-gianas.com/svn/trunk@41 403e655e-980a-47bd-8c60-1ee5348bb6b7

@@
 \usepackage[protrusion=true,expansion=true]{microtype}
+\usepackage{fullpage}
 
-\title{Typing rules for the data-oriented patch system}
+\title{A kernel for incremental type-checking}
 
@@
  \gdecl{a}{x \gor a\ x } \\
- \gdecl{t}{a \gor \sort \gor (x:t)\to t \gor (x=a:t)\to t} \\
+ \gdecl{t}{a \gor \sort \gor (x:t)\cdot t \gor (x=a:t)\cdot t} \\
  \gdecl{\Gamma}{\cdot \gor \Gamma[x:t] \gor \Gamma[x=a:t]}
@@
 
-Following Agda's syntax, we denote by $(x:t)\to u$ the dependent
-product (elsewhere written with a $\Pi$). The new construct
-$(x=a:t)\to u$ is a singleton type in argument position, whose value
-is $a$. Such a singleton product should be read as ``the function
-taking as argument any values syntactically equal to $a$ of type $t$
-(\emph{i.e.} a pedantic way to say $a$). We write as usual
-$(x_1:t_1)(x_2:t_2)\ldots(x_n:t_n)\to t$ for
-$(x_1:t_1)\to(x_2:t_2)\to\ldots\to(x_n:t_n)\to t$. The lookup in the
-environment is written either $\Gamma(x):t$, denoting ``$\Gamma$
-contains a declaration $[x:t]$ \emph{or} a definition $[x=a:t]$'', or
+Following the standard syntax, we denote by $(x:t)\cdot u$ the
+dependent product (elsewhere written with a $\Pi$). The new construct
+$(x=a:t)\cdot u$ is a kind of singleton type, whose value is $a$. Such
+a singleton product should be read as ``the (implicit) function taking
+as argument any values syntactically equal to $a$ of type $t$
+(\emph{i.e.} a pedantic way to say $a$).
+
+We write as usual $(x_1:t_1)(x_2:t_2)\ldots(x_n:t_n)\cdot t$ for
+$(x_1:t_1)\cdot(x_2:t_2)\cdot\ldots\cdot(x_n:t_n)\cdot t$, $t \to
+u$ for $(x:t)\cdot u$ where $x\notin u$, and $(x\ y : t)\cdot u$ for
+$(x:t)(y:t)\cdot u$. The lookup in the environment
+is written either $\Gamma(x):t$, denoting ``$\Gamma$ contains a
+declaration $[x:t]$ \emph{or} a definition $[x=a:t]$'', or
 $\Gamma(x)=a:t$ denoting ``$\Gamma$ contains a definition
@@
 two variables of the same name, nor two syntactically equal
-declarations (the $a$ part is unique). More on that in the
+definitions (the $a$ part is unique). More on that in the
 implementation part.
@@
     \Gamma[x:t]\vdash u:\sort_2
-  }{\Gamma\vdash(x:t)\to u : \sort_3}
+  }{\Gamma\vdash(x:t)\cdot u : \sort_3}
   \quad(\sort_1,\sort_2,\sort_3)\in\mathcal R
@@
     \Gamma[x=a:t]\vdash u:\sort_2
-  }{\Gamma\vdash(x=a:t)\to u : \sort_3} 
+  }{\Gamma\vdash(x=a:t)\cdot u : \sort_3} 
   \quad(\sort_1,\sort_2,\sort_3)\in\mathcal R
   \and
-  \infer[AppP]{
-    \Gamma\vdash a:(y:t)\to u \and \Gamma(x):t
+  \infer[App]{
+    \Gamma\vdash a:(y:t)\cdot u \and \Gamma(x):t
     }{\Gamma\vdash a\ x : u\subst{x}{y}}
   \and
-  \infer[AppS]{
-    \Gamma\vdash a:(y=b:t)\to u \and \Gamma(x)=b:t
-    }{\Gamma\vdash a\ x : u\subst{x}{y}}
+  \infer[Skip]{
+    \Gamma\vdash a:(y=b:t)\cdot u \and \Gamma(x)=b:t
+    }{\Gamma\vdash a : u\subst{x}{y}}
 \end{mathpar}
@@
 
-\paragraph{An example}
-
 \paragraph{Implementation}
@@
 special type of free names: A term is stored in deBruijn syntax, but
-as soon as a variable is introduced, it takes a real name.
+as soon as a variable is introduced, it takes a real name. So in fact,
+environments are not syntactic list objects, instead the operations
+$\Gamma[x:t]$ and $\Gamma[x=a:t]$ are the partial meta-operations of
+adding new entries into the environment, which returns a new, unique
+name for them. The last meta step is to substitute $x$ for this new
+name.
 
@@
 
+\paragraph{An example}
+
+Let's put aside the problem of configuring the PTS for the moment,
+since it's not clear yet: we work in system U \emph{i.e.}
+$\langle\{*\},\{*:*\},\{(*,*,*)\}\rangle$.
+
+We are going to construct a repository gradually as a big term, but
+stopping at some steps to explain the construct. The beginning of the
+term will consist in the declaration of the language (a small
+arithmetic language). Keep in mind that this simply typed language is
+just an example: more evolved languages can and will be declared, even
+on-the-fly as part of a commit!
+\newcommand\nat{\mathbb N}
+\newcommand\prop{prop}
+\begin{align}&
+  (\prop : *)(\nat : *)(0:\nat)(S:\nat) \\&
+  (\land\ \lor\Rightarrow\ : \prop \to \prop \to \prop) \\&
+  (+\ -\ : \nat\to\nat\to\nat)
+  (=\ \neq\ \leq:\nat\to\nat\to \prop )
+  \intertext{Now let's begin the definition of the expression
+    $2+2=4$. The syntax restricts us to define it step-by-step:} &
+  (1 = S\ 0 : \nat)(2 = S\ 1 : \nat)(3 = S\ 2 :
+  \nat)(4=S\ 3 : \nat) \\&
+  (t = (+)\ 2\ 2 : \nat) (p : (=)\ t\ 4) \\&
+  \intertext{Forever, the name $p$ will be associated to that
+    expression. Let's mark it with a name. First we introduce -- on
+    the fly -- the marking feature to our repository (how cool is that?):} &
+  (commit : *)
+  (mark : \prop\to string\to commit)
+  \intertext{The type $string$ is predefined in some way, and the
+    choice of $prop$ as the argument to $mark$ fixes that we want to
+    record propositions as commits, not naturals: it is in some sense
+    the ``tip'' of our syntax. Now we can mark, or commit that $p$: } &
+  (x = mark\ p\ \mathtt{``Cool\ equation}" : commit)
+  \intertext{Now let's make it evolve to $2+2=1+3$, and commit the result} &
+  (u = (+)\ 1\ 3 : \nat)(q = (=)\ t\ u : \prop) \\&
+  (y = mark\ q\ \mathtt{``Also\ true}")
+  \intertext{We could stop here: we have a nice object for commits,
+    and a way to verify by typing that our repository is valid. Let's
+    try to exploit the type system a little bit more: We  now
+    introduce a higher-order object -- a
+    patch maybe? -- taking a piece of syntax matching a certain form
+    and constructing another object. In that case, it takes an
+    equation and transforms it into an inequation:} &
+  (P : (t\ u:\nat)(p=(=)\ t\ u:\prop)(q=(\leq)\ t\ u)\cdot mark\ q\
+  \mathtt{``A\ fortiori}")
+  \intertext{Now we can show that this patch applies to our
+    repository by filling in the missing items:} &
+  (z = P\ t\ u : (p=(=)\ t\ u:\prop)(q=(\leq)\ t\ u)\cdot mark\ q\
+  \mathtt{``A\ fortiori}")
+\end{align}
+When the type of $P\ t\ u$ will be typed, the variable $p$ will be
+given the same name as the previously-introduced $q$, because their
+content match; whereas the new $q$ will be introduced into the
+environment, and given a new unique name.
+\\
+
+The status of commits, patches etc. is still not very clear and the
+previous example should be taken with a grain of salt, but this
+hopefully gives a hint of what is possible with a dependently-typed,
+memoizing language.
+
 \end{document}

### 2010-05-20 doc/synthese.tex
* Un brouillon brouillon d'un squelette de synthese pour s'y retrouver un peu.
git-svn-id: http://gasp.dev.regis-gianas.com/svn/trunk@58 403e655e-980a-47bd-8c60-1ee5348bb6b7

\documentclass{article}
\usepackage[francais]{babel}
\usepackage[utf8]{inputenc}

\begin{document}

\title{Synthèse des idées sur \textsc{Gasp}}

Ce document synthétise les idées sur le développement du
\textit{framework} de patchs sémantiques \textsc{Gasp}, dans le but de
décider efficacement des directions de recherche à explorer.

\section{Le noyau}

\section{Les outils}

\subsection{Édition avec historique}

\subsubsection{Édition \textit{via} des vues}

La syntaxe concrète au choix. Un choix intéressant~: les constructeurs des termes du langage,
des noms pour des transformations locales (de refactoring), des annotations d'aide à la 
reconstruction de l'objet interne (généralisation des annotations de type dans le cas
de l'inférence de type).

\subsubsection{Algorithme de \textit{diff}}

Un outil pour l'incrémentalité. Une application du cadre de la programmation bidirectionnelle (?).

\subsection{Développement collaboratif}

Qu'est-ce qu'une branche? Qu'un merge? Toutes les opérations de \textsc{Git}, de \textsc{Darcs}? 
Une hypothèse~: la métathéorie du langage de patchs
fournit des outils méta semblant capturer ces notions. Une auto-application du système
fait apparaître un langage de description de l'évolution d'un ensemble d'historique.

\subsection{Unification du système de modules (de seconde classe) et du système de construction}

\section{Certification en \textsc{Coq}}

Coq pour donner une interprétation au langage et pour construire un noyau de primitives
certifiées sur lesquelles on peut s'appuyer pour construire des transformations plus 
complexes.

\section{Plan d'attaque}

\end{document}

### 2010-06-17 experiments/alagit/tests/gasp-interface/test1.sh
* Checkout for raw syntax seems operational.
git-svn-id: http://gasp.dev.regis-gianas.com/svn/trunk@90 403e655e-980a-47bd-8c60-1ee5348bb6b7

@@ -7,3 +7,5 @@ $GASP --do init
 $GASP --do commit foo foo.module save 
 # Again, just save the "foo2.module" in the repository
 $GASP --do commit foo2 foo2.module save
+# Now, we check out this saved module, again, just as raw syntax.
+$GASP --do checkout foo2 foo3.raw-fragment

### 2010-06-17 experiments/alagit/tests/gasp-interface/config.sh
* Allow reference to meta-vars in input syntax.
git-svn-id: http://gasp.dev.regis-gianas.com/svn/trunk@94 403e655e-980a-47bd-8c60-1ee5348bb6b7

@@ -1,1 +1,3 @@
+set -e # If a test fails in a sequence of tests, 
+       # we abort the entire sequence.
 GASP=../../gasp.native

### 2010-06-17 experiments/alagit/tests/gasp-interface/test1.sh

@@ -1,11 +1,11 @@
 #!/usr/bin/env bash
 
 source config.sh
-# Create a repository
-$GASP --do init                    
-# Require an "just-save" integration of "foo.module" in the repository
+echo '* Create a repository.'
+$GASP --do init
+echo '* Require a "just-save" integration of "foo.module" in the repository.'
 $GASP --do commit foo foo.module save 
-# Again, just save the "foo2.module" in the repository
+echo '* Again, just save the "foo2.module" in the repository.'
 $GASP --do commit foo2 foo2.module save
-# Now, we check out this saved module, again, just as raw syntax.
+echo '* Now, we check out this saved module, again, just as raw syntax.'
 $GASP --do checkout foo2 foo3.raw-fragment

### 2010-06-18 doc/abstract.tex
Abstract MIPS2010, version 0.01
git-svn-id: http://gasp.dev.regis-gianas.com/svn/trunk@95 403e655e-980a-47bd-8c60-1ee5348bb6b7

\documentclass{article}

\usepackage{ucs}
\usepackage[utf8x]{inputenc}
\usepackage{mathpartir}
\usepackage{color}
\usepackage[pdftex,backref=page,colorlinks=true]{hyperref}
\usepackage{amsmath, amstext, amsthm, amsfonts}
\usepackage{stmaryrd}
\usepackage{geometry}
\usepackage{natbib}
\usepackage{fancyhdr}
\input xy
\xyoption{all}
\usepackage[protrusion=true,expansion=true]{microtype}

\definecolor{bwgreen}{rgb}{0.183,1,0.5}
\definecolor{bwmagenta}{rgb}{1,0.169,0.909}
\definecolor{bwblue}{rgb}{0.317,0.161,1}
\hypersetup{
  linkcolor=blue,
  citecolor=blue
}

\newcommand{\sort}{\textsf{s}}
\newcommand{\gor}{\ |\ }
\newcommand{\gdecl}[2]{{#1}\ &::=\ {#2}}
\newcommand{\subst}[2]{\{#1/#2\}}

\title{Semantic Patches
  % pas sûr que le nom soit encore adapté
}
\author{Matthias Puech $\quad\Leftrightarrow\quad$ Yann Régis-Gianas}
\date{}

\begin{document}
\maketitle

\section{Motivations}

It is interesting to investigate the possible relationship between the
very \emph{methodologies} or even \emph{daily workflow} employed in
both fields of mathematics and computer science: one trying to prove a
theorem, the other constructing a program to fulfill a task. How does
one or the other elaborates his object of study? What kind of \emph{a
  posteriori} modification is he prone to doing? What do these
modifications imply on the validity of the whole edifice or the other
way around, how do one rely on existing work to build up new results?
Finally, how to relate how both scientists collaborate in a team?

Recent efforts in the formalization of mathematical results have
naturally led to these questions and many of them remain largely
unanswered, but the tendency seems to be to adapt existing methods
coming from software development, witness for example the recent
introduction of modules, file-based scripts and separate compilation
in proof assistants like \textsf{Coq} or \textsf{Matita}, the use of
dependency management tools (\textsf{make}) or version control system
(\textsf{git}, \textsf{svn}) to build and manage versions of a
project. Both for the development of proofs or programs, these tools
attempt to cope with the fact that most of a mathematician or
programmer's time is actually spent \emph{editing}, not
\emph{writing}.

As intelligent and widely adopted as this tool chain has gotten, we
believe that they are not adapted for the new, demanding requirements
of proof developments. Indeed, whereas compilation of a program is
usually fast enough for the programmer to rely on the usual
interaction loop ((edit; compile)*; commit)*, the operation of proof
checking is usually too expensive computationally to mimic this
workflow. But even besides the time factor, this ``traditional'' way
of formalizing mathematics hinders the process of mathematical
discovery process: once a concept contained in a file is compiled, it
is considered frozen and any changes to it require the recompilation
of the whole file; the linearity of the development also gives no room
for alternate, inequivalent definitions. This fact has nonetheless
been shown to be of crucial to the mathematical discovery process by
\cite{lakatos1964proofs}, and we believe that they should be taken
into account in the realization of mathematical assistants like
\textsf{Coq} or \textsf{Matita}.

We propose to discuss a small part of these questions, namely the
enhancement and adaptation of version control paradigms to the
management of mathematical repositories, to witness with more
precision the \emph{impact of changes}. By studying the dependencies
between objects in a proof, we can get a much finely grained vision of
what a future evolution would impact. In proof assistants based on the
propositions-as-types paradigm, Type Theory offers a powerful tool to
witness the effect of changes finely, \emph{i.e.} types. Eventually,
we hope to be capable of expressing evolutions on formal proofs, check
them efficiently and incrementally, and manage distributed
repositories of mathematics with it, guaranteeing \emph{by typing} the
stepwise global consistency of the repository.

In the light of the propositions-as-types paradigms where
proof-checking boils down to type-checking, we abstract from whether
we are talking about proofs or programs, so that our development could
eventually also be used in traditional software development.

\section{Methodologies}

We propose here to devise a system for \emph{semantic patches}. It
% TODO changer le nom?
substitutes the idea of textual transformation by:
\begin{itemize}
\item First preferring an abstract syntax tree representation instead
  of plain text;
\item Secondly embedding semantic, or typing data into the
  transformations, in order to be able to reason on them.
\item Finally, perform type-checking in an incremental way, that is
  type only the syntactic difference and reuse derivation for the
  already known subprograms.
\end{itemize}

We want to design a \emph{language of patches}, able to represent
% TODO changer le nom?
these transformations and their semantic properties, thus capturing
local changes in programs, as well as their global effect on the whole
project. We can see this goal as a refinement of the former idea of
``dependency''.

\subsection{Data \emph{vs.} patch-oriented}

\subsection{Related Works}

\section{A kernel for incremental type-checking}

Representing syntax and logics is nicely done in a \emph{logical
  framework} like LF (\cite{harper1993framework}): both the syntactic
elements and the typing derivations can sit in the same tree
structure, and both can be rechecked at the same time, thanks to
dependent types.

For the purpose of incremental type checking though, our needs are a
little bit different: first, we need to record, that is to name all
intermediate values of our developments, so as to be able to address
and reuse them multiple times. Secondly, we need to make sure that
those intermediate values (sub-terms) are not recorded twice, so as to
not type-check them twice. Thus we are looking to represent syntactic
and typing objects as a \emph{DAG} rather than a tree, and ensure
\emph{maximal sharing} among its sub-DAGs.

Our system will have these properties \emph{wrt.} LF:
\begin{itemize}
\item In this first iteration of the project, we do not need
  computations to take place within our DAGs. Our syntax will then be
  restricted to product types and applications;
\item Every term should be a flat application of variables, so that we
  don't introduce compound terms without naming them;
\item Finally we need a way to record intermediate definitions: we
  introduce a new kind of binder, the equality binder. We maintain the
  invariant that not two same definition can sit in the context while
  typing.
\end{itemize}

\paragraph{Syntax} There are three syntactic categories: applicative
terms $a$, types $t$ and environments $\Gamma$.

\begin{align*}
 \gdecl{a}{x \gor a\ x } \\
 \gdecl{t}{a \gor \sort \gor (x:t)\cdot t \gor (x=a:t)\cdot t} \\
 \gdecl{\Gamma}{\cdot \gor \Gamma[x:t] \gor \Gamma[x=a:t]}
\end{align*}

Following the standard syntax, we denote by $(x:t)\cdot u$ the
dependent product (elsewhere written with a $\Pi$). The new construct
$(x=a:t)\cdot u$ is a definition, whose value is $a$. We write as
usual $(x_1:t_1)(x_2:t_2)\ldots(x_n:t_n)\cdot t$ for
$(x_1:t_1)\cdot(x_2:t_2)\cdot\ldots\cdot(x_n:t_n)\cdot t$, $t \to u$
for $(x:t)\cdot u$ where $x\notin u$, and $(x\ y : t)\cdot u$ for
$(x:t)(y:t)\cdot u$. The lookup in the environment is written either
$\Gamma(x):t$, denoting ``$\Gamma$ contains a declaration $[x:t]$
\emph{or} a definition $[x=a:t]$'', or $\Gamma(x)=a:t$ denoting
``$\Gamma$ contains a definition $[x=a:t]$''. Capture-avoiding
renaming of variables in a term $t$ is written $t\subst{x}{y}$ for
``replace all occurences of $x$ by $y$'' and is defined the usual way.

\paragraph{Typing} The typing of our system is a restriction of the
usual typing rules for dependently-typed systems with the additional
two rules for typing a singleton product and the application of a
singleton argument (featuring the syntactic verification in the
environment). It depends on: 
\begin{itemize}
  \item$\mathcal S$ a set of sorts \sort,
  \item$\mathcal A \in \mathcal S^2$ the axioms of the system,
  \item$\mathcal R \in \mathcal S^3$ the allowed products.
\end{itemize}
It relies on two distincts judgments:
\begin{itemize}
\item $\Gamma\vdash a : t$ ($a$ has type $t$ in $\Gamma$),
\item $\Gamma\vdash t : \sort$ ($t$ has sort $s$ in $\Gamma$).
\end{itemize}
The rules are:
\begin{mathpar}
  \infer[Axiom]{
  }{\Gamma\vdash\sort_1:\sort_2}
  \quad(\sort_1,\sort_2)\in\mathcal A
  \and
  \infer[Init]{
    \Gamma(x):t
  }{\Gamma\vdash x:t}
  \and
  \infer[Prod]{\Gamma\vdash t:\sort_1 \and 
    \Gamma[x:t]\vdash u:\sort_2
  }{\Gamma\vdash(x:t)\cdot u : \sort_3}
  \quad(\sort_1,\sort_2,\sort_3)\in\mathcal R
  \and
  \infer[Sing]{\Gamma\vdash a:t \and
    \Gamma\vdash t:\sort_1 \and
    \Gamma[x=a:t]\vdash u:\sort_2
  }{\Gamma\vdash(x=a:t)\cdot u : \sort_3} 
  \quad(\sort_1,\sort_2,\sort_3)\in\mathcal R
  \and
  \infer[App]{
    \Gamma\vdash a:(y:t)\cdot u \and \Gamma(x):t
    }{\Gamma\vdash a\ x : u\subst{x}{y}}
  \and
  \infer[Skip]{
    \Gamma\vdash a:(y=b:t)\cdot u \and \Gamma(x)=b:t
    }{\Gamma\vdash a : u\subst{x}{y}}
\end{mathpar}

A term $t$ represents a state of the repository at a given moment. It
contains the declaration of the object language's syntax, its typing
rules, the syntactic objects and derivations recorded in it along with
the patches that let the repository evolve. If $\cdot\vdash t:\sort$,
then the repository is well-typed.

\bibliographystyle{plainnat}

\bibliography{english}

\end{document}

### 2010-06-18 doc/proposition-these-matthias.tex
ma proposition de thèse
git-svn-id: http://gasp.dev.regis-gianas.com/svn/trunk@98 403e655e-980a-47bd-8c60-1ee5348bb6b7

\documentclass[twoside,a4paper,12pt]{article}

\usepackage{ucs}
\usepackage[utf8x]{inputenc}
\usepackage{mathpartir}
\usepackage{color}
\usepackage[pdftex,backref=page,colorlinks=true]{hyperref}
\usepackage{amsmath, amstext, amsthm, amsfonts}
\usepackage{stmaryrd}
\usepackage{geometry}
\usepackage{natbib}
\usepackage{fancyhdr}
\input xy
\xyoption{all}
\usepackage[protrusion=true,expansion=true]{microtype}

\newcommand{\Coq}{\textsf{Coq} }
\newcommand{\Matita}{\textsf{Matita} }
\newcommand{\letin}[3]{\textsf{let}\ #1=#2\ \textsf{in}\ #3}
\newcommand{\To}{\longrightarrow}

\newtheorem{example}{Example}
\newtheorem{theorem}{Theorem}
\newtheorem{conjecture}{Conjecture}
\newtheorem{definition}{Definition}
\newtheorem{property}{Property}

\definecolor{bwgreen}{rgb}{0.183,1,0.5}
\definecolor{bwmagenta}{rgb}{1,0.169,0.909}
\definecolor{bwblue}{rgb}{0.317,0.161,1}
\hypersetup{
  linkcolor=blue,
  citecolor=blue
}

\pagestyle{fancy}
\renewcommand{\headrulewidth}{0.3pt}
\renewcommand{\footrulewidth}{0pt}

\fancyhf{}
\fancyhead[LE,RO]{\bfseries\thepage}
\fancyhead[LO]{\small\slshape\rightmark}
\fancyhead[RE]{\small\slshape\leftmark}
\renewcommand{\sectionmark}[1]
  {\markboth{\MakeUppercase{\thesection.\ \ #1}}
  {\MakeUppercase{\thesection.\ \ #1}}}

\date{\normalsize February 28, 2010}

\title{{\small Thesis Proposal --
    Dottorato di Ricerca in Informatica} \\[1em]
  Towards formalized mathematics repositories based on type theory}

\author{Matthias \textsc{Puech}\thanks{\texttt{\href{mailto:puech@cs.unibo.it}{puech@cs.unibo.it}}}\\[0.5em]
  \footnotesize \it Dipartimento di Scienze dell'Informazione ---
  Universit\`a di Bologna \\[-0.4em]
  \footnotesize \it Laboratoire PPS --- Université Paris 7--Denis Diderot \\[1em]
  \small \emph{dir.} Andrea \textsc{Asperti}\ \ {\it\&}\ \ Hugo
  \textsc{Herbelin}
  \\[1.3em]
}

\begin{document}

\maketitle
\thispagestyle{empty}
\tableofcontents

\section*{Introduction
  \markboth{INTRODUCTION}{INTRODUCTION}}
\pagenumbering{roman}

The fields of formal proofs, automated reasoning and verified
software, have made great advances in the last forty years, since the
seminal works of on the checkable proof language \textsf{Automath} by
\cite{de1970mathematical} and the automated theorem prover
\textsf{Nqthm} from \cite{boyer1988theorem}.  They appear today as a
mature technology, used not only academically but also in the
industry; the diversity of the approaches also witnesses the vivacity
of the field: there are plethora of systems implemented for various,
more-or-less specific uses, from the all-automated theorem prover for
a domain specific application to the general-purpose, interactive
mathematical assistant embedding a rich and expressive logic.

In particular, \emph{proof assistants}, also known as interactive
theorem provers, are tools intended to help the human to develop
formal proofs: the proof itself is constructed and checked by the
machine, but guided by the human in an interactive collaboration.
Among these, we count \textsf{Isabelle/HOL}, \textsf{PVS},
\textsf{Coq}, \textsf{Mizar}, \textsf{Matita} and much more. These
tools have recently enjoyed a great success, leading to rich formal
mathematics libraries and complex developments, see
\cite{asperti2009social} for a critical and historical account on
formal verification.

The factors of this success are in our opinion threefold: first,
advances on computational logic have lead to the development of
powerful and solid foundational languages of proofs, not only made
intuitive for the mathematician by borrowing their methods, but also
suitable to the exploitation of the computational power of modern
computers. Secondly, the wealth of implementation efforts has lead to
a well-understood architecture (\cite{asperti2007crafting}), and
efficient methods for the multitude of algorithmic tasks performed in
this context. Finally, the tight interaction between users and
developers of these systems (it is actually far from exceptional to be
both) has helped creating rich and usable interfaces, witness for
example the reinterpretation of the proof language of \textsf{Coq} by
\cite{gonthier2008small} originally intended as domain-specific but
inspiring new generations of systems.

However, a lot of the paradigms now taken for granted in the
development of formal mathematics are inherited from programming. For
example, most proof assistant are based on an LCF-like, procedural
tactic language for devising proofs, making the act of proving alike
to the one of programming and strayed from the common vernacular used
in informal mathematics. Moreover, the validation process of a formal
proof, stepwise and linear because modelled after the automatic
process of compilation, do not reflect the way a user would like to
interact with the system: not only would he like to make progress and
advance toward the solution (the only choice today), but also to
retract from previous declaration, refine and modify his idea\ldots\
Also, this rigidity does not leave room at all for team-work, where
several people interact on the same (large) document, possibly
modifying concurrently the main development in an incompatible way,
between each other or with respect to previous developments.

If, following \cite{dowek2007metamorphoses}, we believe that
machine-checkable proofs and the mechanization of proof-search made
mathematics enter its \emph{``industrial era''} after a long history
of handicraft, an era where unverified, one-man work is not acceptable
anymore due to the increased complexity of the manipulated concepts,
then we are in urge of finding techniques to adapt these new tools to
the practice of mathematics.

We believe that the simultaneous work on two aspects of this problem
would bring up new potentialities and ways of formalizing mathematics.
Increasing the inference power of the proof language is one of them:
most of today's proof languages are tightly bound to the underlying
logic, whereas informal mathematics use extensively and implicitly
notations, isomorphisms and shortcuts. Most proof assistants provide
the strict minimum to deal with them, usually in a more-or-less ad-hoc
way; increasing the capability of automating ``trivial'' parts of the
reasoning would be a first step to the design of a high-level, highly
ambiguous but still fully checkable proof language. A second step,
concerning the interactivity of the formal development process, is the
study of the impact of changes in large developments. A high-level
proof language comes with possibly high computational power required,
and one cannot afford then to recheck the whole development after each
change, as small and insignificant as they may be. Moreover, we want
to develop the interactive aspect of these tools, to be able to use
them even during the discovery phase, where change happen often and
deeply into the structure of the development. By analyzing finely the
dependency between concepts and even intra-concept, we hope to render
this interaction between the computer and the user, but also even
between several users.

All these tools are, in a nutshell, implementations of a particular
logic, along with a rich language to conduct the construction of
proofs, mechanisms of inference to relieve the user from the most
tedious and repetitive tasks, and input facilities. While the ideas
underlying the input of proofs are shared by most of these tools, a
particularly interesting subclass of them are those based on
\emph{type theory}. Type theory is a calculus having the notable
feature of being able to be seen dually as an intuitionistic logic and
as a programming language allowing to give and check rich
specifications to programs. This duality, also known as the
Curry-Howard correspondence, opened a whole new field of research,
both practical and theoretical.
\\[-0.5em]

We propose here to tackle the two issues developed above in the
context of proof assistants relying on type theory, and especially the
two developed in both our universities: \textsf{Matita} for the
University of Bologna, \textsf{Coq} for the University Paris 7. 
This proposal is therefore organized in
% three parts. The first is an
% overview of the history and main ideas underlying type theory and its
% impact on both programming languages and proof assistants. The second
two parts.
The first
is an account of the state-of-the art in terms of automation of
proofs, both on a proof-theoretic point of view and on existing
solutions implemented.  It finishes with some directions considered
for further work and works in progress in this area. 
%The last part
The second part
presents an ongoing work on \emph{semantic patches}, a preliminary
theoretical study for the management of changes in type theory that
could eventually serve as a basis for an integration into an existing
system. It finishes also with some possible directions for further
work.
We begin with a short overview of the architecture of the proof
assistants we plan to work on.

\newpage
\pagenumbering{arabic}

% \section{From construction to programming}

% When Brouwer proposed at the turn of the twenties its vision of an
% intuitionistic logic, what is in question is to reject all forms of
% formalism, proofs referring to thought experiments and nothing more:
% it is what he call the \emph{constructions}. The mathematician do not
% verify his proofs, but he must convince himself of their legitimacy,
% as he says:
% \begin{quote}
%   {\it [\ldots] in his most profound hearth}
%   (\cite{brouwer1927intuitionistic}, in \cite{van1967frege}).
% \end{quote}

% This profoundly subjectivist view of logics contrasts interestingly
% with the contemporary work on formalizing set theory by Zermelo, and
% Hilbert's dream of making mathematics so formal that an automatic
% process could generate all valid and interesting theorems of
% mathematics. He actually leaves almost entire the question, yet
% crucial, of the nature of these constructions. The interpretation by
% Heyting in the thirties of his work and the partial formalization of
% the notion gave birth to what we now call ``intuitionistic logic'', or
% constructive mathematics.

% Fifty years later, computer systems emerge that can verify the
% validity of proofs, totally automatically, and even efficient
% processes to discover proofs. A large amount of these tools, as we
% will see, are profoundly linked with the ideas that have emerged from
% the notion of constructive mathematics: in fact, many of them
% implement at their very core an ``intuitionistic logic'': the theory
% of types. Moreover, underlying the proofs that one can carry in these
% logics, there are real programs, that can be executed by the machine.
% How has the notion of construction shifted so much as to change nature
% almost entirely? It is the question we will try to explore here.

% For this, we will draw the material for this analysis in two, strongly
% dependent discoveries: one is the Curry-Howard correspondence which
% appears today as the first tangible incarnation of the idea of
% construction, the second is one of its sequel, that gave birth to a
% whole field of research at the boundaries between Mathematics, Logics
% and Computer Sciences: type theory.

% Finally, we will expose the formal definitions, the variants and some
% of the issues regarding the incarnation of type theory that will be
% the framework of this proposal: the Calculus of Constructions. It is
% this type theory that is implemented in the proof assistants
% \textsf{Coq} and \textsf{Matita}.

% \subsection{Constructive mathematics}

% We shall try in this section to give some historical references and
% clues that will help us comprehend this paradoxical development of
% history: how, from the subjectivism of Brouwer, could emerge these
% tools, that witness such an externalist vision of Logic?  How this
% intimate, almost mystical experience -- the construction -- could be
% subject to such a change of nature as to become a process, an
% algorithm for the totally external verification of proofs?

% \subsubsection{The constructivist program}

% At the turn of the 20\textsuperscript{th} century, the dominant
% ideology in mathematical logic was that of \emph{formalism}. This
% theory, as advocated among others by the mathematicians Hilbert,
% Bernays and von Neumann, holds that mathematics, and in particular
% logic, are nothing but a game on formulas and statements, and that
% theorems and their proofs can be assimilated to a bare manipulation of
% strings. For example, Euclidean geometry can be thought as a game,
% whose goal is to transform a series of axioms by a number of
% transformations into a given statement, and once this goal is
% attained, the statement has been proven. It is then a theory about,
% and only about axiomatic systems: for these mathematicians, the system
% of axioms and rules considered is of practically no importance except
% for its intrinsic properties, and the meaning of the actual choice of
% a system is (almost) totally irrelevant.

% In spite of the many deceptions that arose at the beginning of the
% century when attempting to devise a general enough formalism -- among
% others, Frege's \emph{Begriffschrift} attempt to build a universal
% formal language was doomed by Russell's paradox, see
% \cite{van1967frege} -- this view was so much anchored in its time that
% David Hilbert, maybe the earliest proponent of formalism, proposed his
% \emph{program}. From the ``foundational crisis'' mathematics were
% suffering, that is the search for a proper foundation for mathematics,
% he proposed to ground all existing theories to a finite, complete set
% of axioms, and provide a proof that these axioms were consistent, in
% order to provide mathematics with a universal, solid foundation: a
% complete and consistent axiomatization of \emph{all of mathematics}.
% However, a few years later, this program was mathematically shown to
% be unattainable by Gödel's incompleteness theorem.
% \\[-.5em]

% In this context, \emph{constructivism} emerged as a strong opponent to
% the formalist school. Lead by Brouwer, Markov, Bishop and late Hilbert
% and Bernays with their ``finitism'', constructivism, as a
% philosophical view on the existence of the mathematical objects,
% asserts that it is necessary to find (or "construct") a mathematical
% object in order to prove that it exists.  Therefore, the ``game''
% of applying inference rules from axioms is not necessarily a proof of
% existence of an object: only the \emph{construction} -- the exhibit of
% such an object -- is able to convince the mathematician of its
% validity. Thus, the formalist notion of inference disappears almost
% entirely here: the choice of the system is not only the only valuable
% question, since it would make no sense to conduct a proof in a system
% that doesn't allow to perform the construction of a witness of
% existence of the object, but it also looses all reason to be: only one
% system is capable of producing such a witness to perform constructive
% mathematics.

% In particular, one form of reasoning is therefore strongly rejected:
% when one assumes that an object does not exist and derives a
% contradiction fromc that assumption, one still has not found the object
% and therefore not proved its existence, according to constructivists:
% the \emph{law of excluded middle} is therefore rejected. Often,
% constructivism is assimilated with this rejection; it is in fact only
% a consequence of the foundation of the notion of proofs on the
% construction.

% Brouwer's \emph{intuitionism} is one of the form taken by the
% constructivists' thesis statements. It was strongly influenced by
% Kant's notion of intuition and redefined the notion of truth as it was
% considered by traditional mathematics. In classical logic like in set
% theory, truth is a static property of a proposition (or as Brouwer
% prefers to name it after Kant, a \emph{judgement}): it is either true
% or false, depending on its \emph{real nature}. This is expressed by
% the interpretation of the logical statements in a model of truth
% values: traditionally, the meaning of a proposition was interpreted in
% an algebra of two values, true and false. On the contrary, the major
% methodological turning point of Brouwer is to abandon the semantical
% interpretation of propositions. He suggests that this usual denotation
% of propositions as truth values is simply inadequate, this for four
% reasons:
% \begin{itemize}
% \item It is first a philosophical platitude: Tarski for example
%   interprets the conjunction as: $A\wedge B$ is true if $A$ is true
%   and $B$ is true.  The \emph{meta} operator ``and'' doesn't help us
%   to understand $\wedge$ in any way.
% \item Justify the quantification in this way is undecidable in most
%   cases, since we have to verify it on the whole ``domain of the
%   discourse''.
% \item Fermat's theorem denotes in this case the same object as
%   $2+2=4$. It is really reasonable to let so little room to the notion
%   of proof?
% \item Finally, this interpretation simply doesn't scale to
%   higher-order logic.
% \end{itemize}

% He takes a much more subjectivist approach to the notion of proof. For
% him, it is even close to the juridical notion of \emph{evidence}: a
% proof of $A$ is the set of processes that make one be convinced of
% $A$. It is in some way an incriminating evidence, an object the
% observer -- the judge or the mathematician -- can convince himself,
% either directly (the immediate proof), or through a series of steps, a
% mechanism (the mediate proof, like the mediate jurisdiction) that can
% be ``replayed'' at will, as many times as one needs to convince
% himself. The intuitionistic slogan of Brouwer's ``there are no
% unexperienced truth'' takes then all its meaning. The experience, that
% Brouwer call construction, is the only evidence that constitutes
% truth. In this sense, the act of ``proving'' that $12$ is not a prime
% number is to exhibit the computation of $4*3$ as an evidence.

% Heyting, and later Kolmogorov will formulate this idea more precisely
% (in some of the articles in \cite{benacerraf1983philosophy}): the
% proposition (statement, judgement) must be understood as a problem, an
% expectation or an intention. Better, a task. To prove a statement (a
% problem, a task) is to know what constitutes a solution to its
% problem, a fulfilment to its task. Heyting and Kolmogorov
% (\cite{heyting1971intuitionism}) even propose a full interpretation of
% the logical statements in terms of these analogy: the famous BHK
% interpretation: \newcommand{\subst}[2]{[{#1}/{#2}]}
% \begin{align*}
%   \llbracket A\Rightarrow B \rrbracket &= f \in \llbracket
%   A\rrbracket\to \llbracket B\rrbracket\\
%   \llbracket\forall x\in A,\ B\rrbracket &= f \in
%   \{\llbracket B_x\rrbracket\}_{x\in \llbracket A\rrbracket} \\
%   \llbracket \exists x\in A,\ B\rrbracket &= \langle a,b\rangle \in
%   \llbracket A\rrbracket\times \llbracket B\subst{x}{a}\rrbracket \\
%   \llbracket A\wedge B\rrbracket &= \langle a, b \rangle \in
%   \llbracket A\rrbracket\times \llbracket B\rrbracket
% \end{align*}

% This interpretation gave birth to the first formalization by Heyting
% of the intuitionistic logic. It is nowadays presented as a natural
% deduction, minus the rule \emph{Reduction ad absurdum} which justifies
% the excluded middle.

% This interpretation put one idea forward that takes the form of two
% properties. A logic that would satisfy this interpretation will see
% its connectives $\exists$ and $\vee$ act more precisely, with more
% control than in usual, classical logics: they are the properties of
% the witness and of the disjunction.
% \begin{property}[Witness]
%   From a proof of a statement of the form $\exists x, P(x)$, one can
%   \emph{extract} a \emph{witness} $a$ such that $P(a)$ is provable.
% \end{property}
% \begin{property}[Disjunction]
%   From a proof of a statement of the form $A\vee B$, one can decide
%   which of $A$ or $B$ is provable.
% \end{property}

% As we can see, the excluded middle $A\vee¬A$ is obviously not provable
% in such a logic: if it were, one of $A$ or $¬A$ would be provable
% (without any other knowledge about $A$), which is obviously false.
% \\[-.5em]

% The first real incarnation of what we could call a guideline for the
% elaboration of logics, was discovered in the sixties, almost by
% accident, by linking two notions that had previously no commonalities,
% the $\lambda$-calculus and the intuitionistic logic. This link
% actually turned out to be an exact and astonishing correspondence.

% \subsubsection{The Curry-Howard correspondence}

% In the 1930s, Church introduced, as part of an investigation into the
% foundation of mathematics, a very simple yet powerful formal system to
% reason on function definition and their application: the
% $\lambda$-calculus.

% After the failure of Frege's (variant of) na\"ive set theory,
% logicians were left with the difficult problem of finding a natural
% yet consistent formulation of set theory. Two direction were taken
% from this point. Zermelo and Fr\"ankel on one side modified Frege's
% formulation to obtain ZF, an axiomatic set theory. Church on the other
% side, took the approach of expressing higher-order logic (a set
% theory) with the help of a minimal language for functions. In his
% seminal article, \cite{church1940formulation} introduced, with the
% help of the $\lambda$-calculus, a formulation of classical,
% higher-order logic based on first-order logic with the addition of
% only three symbols: $\lambda$ which represents an anonymous function,
% the application of a function to its argument, and the arrow $\to$ to
% express what he call the \emph{type} of these functions. This
% formulation is called the Simple Theory of Types. It has the
% peculiarity of introducing, for the first time explicitly, a notion of
% computation or \emph{reduction}: the $\beta$-reduction.

% This reduction is at the heart of the higher-order character of his
% logics: in fact, it mimics the behavior of an axiom of set theory, the
% \emph{comprehension scheme}: \[ \forall x_1\ldots x_n, \exists y,
% \forall z (z\in y \Leftrightarrow P)\] This scheme states that it is
% possible to define sets \emph{in comprehension}, i.e. by a property on
% its elements. At each property corresponds a set. In fact, by
% skolemizing this scheme and introducing the notation $\{ x\ |\ P \}$
% for the function associated to $y$, we obtain: \[ \forall x_1\ldots
% x_n, z, (z \in \{z\ |\ P\} \Leftrightarrow P)\] This is read: if I
% constructed the set $z$ from the property $P$, and if in fact, $z$ is
% in this set, then $P$ is true.

% Some years later, Church, extracting the computational content of his
% logic, was left with a very simple formal system that had nothing to
% do with logic anymore, but with computational complexity:
% \[ t\ ::=\ x\ |\ \lambda x. t\ |\ t\ t \qquad\qquad (\lambda x. t)\ u
% \to t[x/u]\]. This systems supposedly allowed to represent \emph{any}
% kind of computation, and still today is a touchstone for the notion of
% computation. One problem with it is that, viewed as a set theory, it
% allows to define the set of all sets that do not belong to themselves,
% Russel's paradox: \[ \{x\ |\ ¬x\in x\}\in\{x\ |\ ¬x\in x\}\]. Thus,
% the above system is not only logically inconsistent but Russel's
% paradox is also a counter-example to its termination: the
% corresponding term $(\lambda x. x\ x)(\lambda x. x\ x)$ reduces on
% itself \emph{ad infinitum}. Obviously, the problem with the above
% system is that it allows the variable $x$ is applied to itself, which
% lead Church to put each variable in a distinct \emph{type}. A variable
% denoting a function would have a type $A\to B$, and could be applied
% only to an argument of type $A$. This lead to the Simple Type Theory,
% an isolation of a subset of well-behaved terms, the typed terms. They
% are defined inductively as follows:
% \begin{mathpar}
%   \infer[Init]{(x:A)\in\Gamma}{\Gamma\vdash x:A} \and
%   \infer[Abs]{\Gamma, x:A\vdash t:B}
%   {\Gamma\vdash\lambda x. t : A\to B} \and 
%   \infer[App]{\Gamma\vdash t:A\to B \and \Gamma\vdash
%   u:A}{\Gamma\vdash t\ u: B}
% \end{mathpar}
% These three rules formed so to say the core of the functional aspect
% of the logic discovered by Church, but also a universal programming
% language, in which all possible programs are expressible as terms. For
% a full history of $\lambda$-calculus and their applications, see
% \cite{cardone2006history}.
% \\[-.5em]

% This idea of a well-behaved fragment of the $\lambda$-terms stayed
% contained to the study of computable functions and higher-order
% logics, for example by \cite{russell1908mathematical} in his Ramified
% Theory of Types. In the 1960's however, H. Curry and W. A. Howard
% noted a striking similarity between these three rules and those of the
% minimal fragment of intuitionistic logic: in fact, if one removes all
% information concerning terms in these rules, he obtains exactly the
% rules of minimal intuitionistic logic:
% \begin{mathpar}
%   \infer[Init]{A\in\Gamma}{\Gamma\vdash A} \and
%   \infer[Intro]{\Gamma,A\vdash B}{\Gamma\vdash A\Rightarrow B} \and
%   \infer[Elim]{\Gamma\vdash A\Rightarrow B \and \Gamma\vdash A}{\Gamma\vdash B} \and
% \end{mathpar}

% This remark, yet so simple, circulated on an unpublished note that was
% published only twenty years after by \cite{howard1980formulae}, and
% was called the \emph{proposition-as-types} paradigm. It is easily
% expanded to full first-order intuitionistic propositional logic by
% adding new connectives in the types and new constructions in the term:
% pairs and conjunctions, pattern-matching and disjunction, empty type
% and $\bot$, unit type and $\top$\ldots\ and allows then to build a
% very precise correspondence between notions of proof-theory and their
% now-equivalent in computation theory and programming language
% theory:
% \\

% \begin{center}
%   \begin{tabular}{|l|l|}
%     \hline 
%     \bf proof-theory & \bf computation \\ \hline
%     proposition & type \\ \hline
%     proof & program \\ \hline
%     deduction rules & typing rules \\ \hline
%     cut-elimination & evaluation \\ \hline
%     cut-free proof & value \\ \hline
%   \end{tabular}
% \end{center}
% \vspace{1em}
% However, the problem of extending this correspondence to full
% first-order logic, and besides to higher-order logic stayed
% entire. One of the (many) answer to this problem appeared at the end
% of the seventies under the name of \emph{Intuitionistic Type Theory}.

% \subsection{Type theory as a foundational language}

% Among the formalisms proposed to extend the Curry-Howard
% correspondence, two of them were particularly successful. One is the
% system F, discovered independently by \cite{girard1972interpr} and
% \cite{reynolds1983types}. It is a functional programming language with
% polymorphism that can be viewed logically through the Curry-Howard
% correspondence as the minimal second-order fragment of intuitionistic
% logic. The second, which will be our interest in this section is
% Martin-L\"of's Intuitionistic Type Theory
% (\cite{martin1984intuitionistic}). It differs from the previous
% formalisms in an important way, as it is \emph{dependent}, that is, it
% makes no distinction anymore between types and terms and allows
% computation to be performed not only on terms but also on types.

% We will expose this formalism and try to hint its correspondence to
% the intuitionistic program of Brouwer, but before that we will have to
% expose one of its major underlying philosophical idea, taking its
% ground directly in the notion of judgement of Kant's. We will then see
% that it allows a very general presentation, unifying most of the
% variants proposed, including system F: pure type systems.

% \subsubsection{The judgemental method}
% \label{sec:jmeth}

% To justify the notion of dependency as introduced in his type theory,
% Martin-L\"of proposes a new method for constructing a logic, that we
% will call the \emph{judgemental method} following Pfenning. It is a
% shift of focus from the notion of formula to the one of judgement,
% allowing a new flexibility in the definition and justification of the
% meaning of the rules. This method is exposed in the course notes
% \cite{martin1996meanings}.

% What is the object of a rule of inference? What are $A$ and $A\vee B$
% when I write the rule: $$\infer{A}{A\vee B}$$ The traditional answer
% to that question is that they are \emph{formulas}, or
% \emph{propositions}. They constitute in the traditional course on
% logic, e.g. when natural deduction is explained, the central notion,
% the one on which all other are depending on: inference, axioms,
% models\ldots\ A formula is an inductively defined expression,
% that is fixed before starting to define the rules of inference, and is
% generally represented as a BNF grammar:
% $$ A,B\ ::=\ P(t_1,\ldots,t_n)\ |\ A\wedge B\ |\ A\vee B\ |\ \forall x, A\ |\
% \exists x, A\ |\ \top\ |\ \bot$$ \emph{Expressions} on the contrary
% form so to say an open concept: they are tree-like objects that one
% can represent, \emph{``can recognize in its many occurrences and can
%   reproduce in many copies''} (\cite{martin1996meanings}). From this
% traditional view, we can draw an implicit order of precedence of
% concepts:
% \begin{quotation}
%   expression $<$ formula $<$ logical rule
% \end{quotation}

% This vision has a severe conceptual flaw however. When I define
% natural deduction, the \emph{signature}, i.e. the set of predicates,
% function symbols and axioms, are to be defined before the formulas
% since formulas depend on them, and \emph{a fortiori} before logical
% rules. Then, if I define the reasoning on arithmetics (natural
% deduction on Peano's arithmetic), and then set theory (natural
% deduction on ZF axioms), I define to totally disjoints theories that
% don't even share their logical connectives: we will have
% $\vee_{\text{Peano}}$ and $\vee_{\rm ZF}$ etc. And yet, these
% connectives behave exactly the same way on the formulas: they are so
% to say two unnecessarily specific versions of the same logical
% concept.

% This remark suggests a more liberal, a looser handling of the notion
% of formula. In the above conception, an important point is missing:
% when a rule is stated, it is implicitly stated that if the premises
% are true, then the conclusion must be true as well. If we make this
% character explicit, the rule above must be read as:
% \[ \infer{A\vee B\text{ is true}}{A\text{ is true}}\] The previously
% implicit modality applied to the formulas -- the fact that they are
% true, make up what we will call a \emph{judgement}. As it was done by
% Bolzano and Frege, we can even declare different forms of judgement,
% like the fact that a statement is true or false, the fact that an
% expression is a natural number\ldots\ provided that we provide a way
% to verify these forms of judgments: provided that we know what counts
% as a proof of these judgements. Introducing this intermediate concept
% of judgment allows us to liberate ourself completely from the too
% restrictive notion of formula. Let's even replace it completely with
% the open notion of expression, and constrain \emph{by the rules} their
% character of formula or proposition. We introduce two different forms
% of judgements:
% \begin{itemize}
% \item $A$ is a proposition ($A\text{ prop}$)
% \item $A$ is true ($A\text{ true}$)
% \end{itemize}
% The previous rule becomes:
% \[ \infer{A\text{ prop}\and B\text{ prop}\and A\text{ true}}{A\vee
%   B\text{ true}}\]
% So to say, the charge of work let previously to the definition of
% formula -- discriminate against well- or ill-formed propositions -- is
% deferred on the inference rules, and a derivation in the obtained
% system not only constitutes the establishment of the truthfulness of
% a formula, but also of this well-forming. For each symbol, we now have
% to give the adequate rules:
% \begin{mathpar}
%   \infer{A\text{ prop}\and B\text{ prop}}{A\vee B\text{ prop}} \and
%   \infer{A\text{ prop}}{¬A\text{ prop}} \and \ldots
% \end{mathpar}

% The syntax of formulas, deferred this way to the logical rules, allows
% a tight control of the system, because judgement types can be mixed
% together. This remark, as innocent as it seems, is in fact a central
% point of the theory of types as we will see.

% \subsubsection{Martin-L\"of type theory}

% The original presentation of Martin-L\"of's type theory
% (\cite{martin1984intuitionistic}) is an extensible theory composed of
% a core \emph{calculus of judgements}, on top of which one declares the
% logical connectives and their rules.

% It relies on two primitive notions: the first is called indifferently
% \emph{set} or \emph{type} and corresponds intuitively to the --
% intuitionistic -- idea that a type is nothing but the set of its
% values. Beware though: ``set'' is not to be understood in the broad
% sense of set theory only as ``what we can define inductively''. We
% will write following the original presentation $\in$ the membership of
% such a set. The second notion is the one of \emph{judgmental
%   equality}. It is a form of judgement that defines the set of purely
% computation substitution operations allowed on respectively the sets
% and their inhabitants. Think for the moment of this equality as a
% quotient on the other forms of judgment.

% \paragraph{Forms of judgment}

% \newcommand\set{\text{ set}}

% We introduce the following forms of judgment:
% \begin{enumerate}
% \item $A$ is a set ($A\set$)
% \item $A$ and $B$ are equal judgments ($A=B$)
% \item $a$ is an element of $A$ ($a\in A$)
% \item $a$ and $b$ are two equal elements of $A$ ($a=b\in A$)
% \end{enumerate}

% They come with constraints on the rules they will apply to: if we want
% to define a particular set $S$, $S\set$ can be introduced as a rule
% only if we introduce at the same time all rules that describe all
% elements of this set, and how to compare two elements of this set. For
% example, this is a valid introduction of a set:
% \begin{mathpar}
%   \infer{ }{\textsf{0}\in\mathbb N} \and
%   \infer{n\in\mathbb N}{\textsf{S}\ n\in\mathbb N} \and
%   \infer{ }{\textsf{0}=\textsf{0}\in \mathbb N} \and
%   \infer{n=m\in\mathbb N}{\textsf{S}\ n = \textsf{S}\ m \in\mathbb N}
% \end{mathpar}

% All of these judgements form in fact families of judgement by what is
% introduced under the name of \emph{hypothetical judgement}. An
% hypothetical judgement is a family of judgements indexed by another
% judgement, and will be written $$ (J_1)\ldots(J_n)\ J $$

% Hypothetical judgements are judgements of their conclusion $J$ whose
% proof can be made complete by the substitution of their ``holes'' by
% judgements of their hypothesis $J_1\ldots J_n$. An important novelty
% is that it is an inductive construct: an hypothesis $J_i$ is also
% supposed to be an hypothetical judgement, parameterized by all $J_j$
% with $j<i$, and this is reflected even in its statement. This
% mechanism allows a very natural expression of mathematical statements.
% \begin{example} The mathematical(ly void) statement:
%   \begin{quote}
%     Let $R$ be a relation on a set $A$, and $R*$ its reflexive
%     closure. Then for all $x\in A$, $R*(x,x)$.
%   \end{quote}
%   is translated into the hypothetical judgement:
%   \begin{quote}
%     ($A$ is a set) \\
%     ($R$ is a relation on $A$) \\
%     ($R*$ the reflexive closure of $R$) \\
%     ($x$ an element of $A$) \\
%     $R*(x,x)$
%   \end{quote}

% \end{example}

% \paragraph{Interpretation}

% You won't be surprised to learn that each of these forms of judgement
% accept several reading, as based on the BHK interpretation:

% \begin{center}
%   \begin{tabular}{l|l}
%     $\bf A\ \rm\bf set$ & $\bf a\in A$ \\ \hline
%     $A$ is a set & $a$ is an element of $A$ 
%     \emph{or} $A$ is non-empty \\\hline
%     $A$ is a proposition & $a$ is a proof of $A$ 
%     \emph{or} $A$ is true \\ \hline
%     $A$ is a problem & $a$ is a method to solve $A$
%     \emph{or} $A$ is solvable \\ \hline
%     $A$ is a specification & $a$ is a program realizing $A$
%     \emph{or} $A$ is inhabited
%   \end{tabular}
% \end{center}

% In other words, $A\set$ can be viewed (interpreted) as the judgment
% $A\text{ prop}$ introduced in \ref{sec:jmeth}. The same way, by
% ``degrading'' the judgement $a\in A$ and forgetting the $a$, it is
% read as $A\text{ true}$. These remarks have for Martin-L\"of only an
% indicative value though: they only allow us to give intuitions on
% their meaning, and do not constitute an formal interpretation: they
% will allow, once the logic fully defined, to verify that the logical
% constants introduced have the meaning we intended to give them.

% \paragraph{Typing rules}

% We can now begin to construct the system of rules for the core
% theory. All rules can be justified by the underlying construction, but
% we detail only some of them.
% \\[-.5em]

% \underline{\bf Equality rules}
% \begin{mathpar}
%   \infer[Refl/EqS]{A\set}{A=A} \and
%   \infer[Refl/EqE]{a\in A}{a=a\in A} \\
%   \infer[Sym/EqS]{A=B}{B=A} \and
%   \infer[Sym/EqE]{a=b\in A}{b=a\in A} \\
%   \infer[Trans/EqS]{A=B \\ B=C}{A=C} \and
%   \infer[Trans/EqE]{a=b\in A \\ b=c\in A}{a=c\in A} \and
% \end{mathpar}

% These rules reflect the fact that the equality is an equivalence
% relation, on elements and on sets. They are justified by the fact that
% they define a congruence on elements and sets. Remark the use of other
% kind of judgments as premises of the reflexivity rules. They ensure
% that equality is always done in a ``typed setting'': one cannot prove
% $2=2$, but only $2=2\in\mathbb N$.
% \\[-.5em]

% \label{sec:mltt}
% \underline{\bf Set conversion}
% \begin{mathpar}
%   \infer[Conv/E]{a\in A \\ A=B}{a\in B} \and
%   \infer[Conv/EqE]{a=b\in A\\ A=B}{a=b\in B}
% \end{mathpar}

% These rules state that equal elements of different sets can be
% assimilated provided that the sets are equal. They will be of crucial
% importance and are the key to the comprehension of type theory and its
% sequel, because they break the usual discrepancy made between types as
% static values and terms as dynamic entities that can be evaluated: in
% type theory, sets are also subject to computation. For example, the
% set of vectors of length $x+y$ is the same as the set of vectors of
% length $y+x$ up to conversion.
% \\

% \newcommand{\hypo}[2]{(#1)\ #2}

% \underline{\bf Substitution}
% \begin{mathpar}
%   \infer[Subst/E/S]{a\in A \\ \hypo{x\in A}{B\set}}{B\subst{x}{a}\set} \and
%   \infer[Subst/E/EqS]{a\in A \\ \hypo{x\in
%       A}{B=D}}{B\subst{x}{a}=D\subst{x}{a}} \and
%   \infer[Subst/E/E]{a\in A \\ \hypo{x\in A}{b\in B}}{b\subst{x}{a}\in
%     B\subst{x}{a}} \and
%   \infer[Subst/E/EqE]{a\in A \\ \hypo{x\in A}{b=d\in
%       B}}{b\subst{x}{a}=d\subst{x}{a}\in B\subst{x}{a}} \and
%   \infer[Subst/EqE/S]{a=c\in A \\ \hypo{x\in
%       A}{}B\set}{B\subst{x}{a}=B\subst{x}{c}} \and
% \end{mathpar}

% Substitution is the basic operation on hypothetical judgements. They
% state that an hypothetical judgement can be completed by the knowledge
% of its hypothesis; there are theoretically one rule by pair of
% judgement, but we keep only the useful ones here. Here we have the
% witness of the \emph{operationality} of this language: this mechanism
% of completion of partial knowledge forms already an algorithm for the
% reduction of a proof, or the execution of the solution to a problem in
% Martin-L\"of's words. This basic system, forms the core of
% Martin-L\"of's type theory, in which one can define the usual logical
% as well as general connectives such as the dependent products and
% sum.

% This language, by its operationality and its ability to represent
% logical concepts in an intuitionistic way, is the basis of many modern
% proof assistants. In fact, Martin-L\"of goes further and declares:

% \begin{quote} \textit{ ``If programming is understood not as the
%     writing of instructions for this or that computing machine but as
%     the design of methods of computation that it is the computer's
%     duty to execute (a difference that Dijkstra has referred to as the
%     difference between computer science and computing science), then
%     it no longer seems possible to distinguish the discipline of
%     programming from constructive mathematics. This explains why the
%     intuitionistic theory of types [\ldots], which was originally
%     developed as a symbolism for the precise codification of
%     constructive mathematics, may equally well be viewed as a
%     programming language.''} --- \cite{martin1984constructive}
% \end{quote}

% \subsection{Architecture of a proof assistant}
\subsection*{Architecture of a proof assistant}

Both our proof assistants of interest, \textsf{Coq} and
\textsf{Matita} are based on the \emph{Calculus of Inductive
  Constructions (CIC)} of \cite{paulin1996definitions}. It is a very
expressive logic, built on a variant of Martin-L\"of's type theory
with polymorphism \`a la System F and a hierarchy of predicative
universes. It supports, as in Martin-L\"of presentation of his type
theory, the inductive definitions of objects, and supports the
proofs-as-types paradigm. Among other features, it provides recursive
function definitions and an impredicative universe for dealing with
logical propositions.

Numerous systems based on a variant of this framework have been
implemented: \textsf{Agda}, \textsf{NuPRL}, \textsf{Epigram} and much
more. As time went by, the architecture of these tools has become
better and better understood. One particularly important criterion in
their conception is the so-called \emph{De Bruijn criterion}. They all
share a common layered software architecture, as sketched for
\textsf{Matita} in \cite{asperti2007crafting} for example, and are
often qualified of ``skeptical proof assistants''. We give a brief
overview of each layer from the innermost to the outermost one, as it
is implemented in \textsf{Matita} (this may vary for other systems):

\begin{description}
\item[Kernel] Their principle is to isolate a little portion of
  human-understandable code, the \emph{kernel}, only dedicated to the
  task of proof-checking, i.e. the verification of the well-typing.
  This component is responsible for the validation of all generated
  proofs, and is this way the only piece of code that must be trusted
  in order to trust the whole system.

\item[Refiner] Writing directly proofs in the language of the CIC is
  an almost impossible operation: not only would it be very hard to
  follow, but also very boring, because the language is very
  redundant. On top of the kernel, the \emph{refiner} is in charge of
  completing all logical information omitted by the user, with the
  help of dedicated mechanisms: \emph{type inference} allows you to
  omit some type constraints if they can be inferred by the system
  from the context. \emph{Coercions} simulate a common practice in the
  mathematical discourse, i.e. the use of a super-structure in place
  of one of its components (example, a group instead of its carrier
  set). The more recent \emph{canonical structures} and \emph{type
    classes} allow to avoid mentioning the context by choosing, on one
  side a canonical representative of the context (a particular
  mathematical structure) or relying on inheritance between
  structures.

\item[Library management] All proofs and definitions in these proof
  assistants are devised thanks to a mechanism of definitional
  equality (in the sense of definition expansion). All the definitions
  are stored in a library, with the ability to import and reuse
  previous development. This layer is in charge of the management of
  such a library: loading, saving, managing namespaces as well as
  modifying previous definitions.

\item[Proof management] Some of the systems mentioned allow the
  stepwise construction of proofs thanks to a library of tactics,
  which are tools to build proofs incrementally by refining the
  previous state of the proof. This layer manages the call to tactics
  in this process, and the verification of their result by the refiner
  (which in turns calls the kernel).

\item[Tactics] This layer makes up the set of specific tactics of
  proof refinement. It is usually built in a hierarchical manner, some
  tactics calling some others for specific subtasks.

\item[Disambiguation] Very often in the context of informal
  mathematical vernacular, ambiguity of the notations used happens.
  This layer is responsible for mapping potentially ambiguous
  notations and names to their logical, non-ambiguous counterpart by
  inferring it from the context.

\item[User interface] Then, some logic-independent layers are in
  charge of the parsing of formulas, their display, and graphical user
  interface management etc. The traditional model of GUI for a big
  family of these tools (implemented in the emacs editor as
  \textsf{ProofGeneral}) is a text editor with three panes: one
  displaying the actual script, the second the state of the current
  proof, and the last the possible messages to the user. One compiles
  a part of the script by moving a cursor forward and backward.
\end{description}

\newpage

\section{Automation in proof assistants}
\label{sec:auto}

The kernel of interactive theorem provers based on type theory, namely
the logic they implement and their typing algorithm, is now a
well-understood part of such systems. But the very fine granularity of
the proofs one can write directly for the kernel to check makes this
act almost impossible for any relatively involved proofs. This led to
the direction of improving the capacity of these systems to receive
partial proofs, possibly with omitted parts or information, and infer
them back fully, to the point where the kernel can check them. Many
different mechanisms fulfill this task on different levels:
\emph{coercions} permit to assimilate notions corresponding to
different underlying objects, \emph{implicit arguments} allow to omit
redundant arguments of functions and theorems, the recent \emph{type
  classes} are an incarnation of the ad-hoc polymorphism present in
some programming languages. To a certain extent, most of the tactics
of \textsf{Matita} or \textsf{Coq} belong to this category as well:
their application triggers the construction of complex proof objects.

In this section, we will focus on one of the most involved of such
mechanism, commonly called \emph{automation} tactics or mechanism. As
opposed to the others, the goal of these is to provide a high-level
method for (more-or-less blindly) searching a proof of a given
statement. The utility of these helpers are needless to be proved
anymore, but we want to advocate here a particular use of them. Two
problems arise from the current, low-level view on formalization.
First, devising proofs directly in type theory or with the help of
low-level tactics is a tedious exercise and reading an already proof
is often difficult. Secondly, the different variations and
peculiarities of each proof assistants make their proofs not
interoperable, resulting in isolated, balkanized formalizations and
duplicated efforts. The development of the automation techniques of
all kind are a direction towards the compensation of both these
problems: they allow to alleviate the proof burden and are a required
step towards the design of very light proof languages that can be
embedded in different foundational dialects and proof assistants, and
is therefore necessary for the design of large, formalized mathematics
repositories.

\subsection{State of the art}

We begin by reviewing some of the common techniques of automated
theorem proving, and discuss their adaptation and limitations in the
setting of type theory, along with implemented generic methods of
proof search in proof assistants.

\subsubsection{First-order theorem proving}

\paragraph{Automated theorem proving}

Automated theorem proving, or automated deduction, is the field
dedicated to the search of proofs of mathematical theorems by computer
software. Theorem provers are tools taking a mathematical statement as
input, and outputting a proof of it (or just succeed) or fail or loop
forever if the statement isn't provable. Most general purpose theorem
provers rely on classical first-order logic (FOL).  These include for
example \textsf{Vampire}, \textsf{Ergo}, \textsf{Waldmeister},
\textsf{Mace4/Prover9}, \textsf{SPASS} or more recently
\textsf{Imogen}. In this part we will review some of the popular
techniques and framework for classical and intuitionistic first-order
theorem proving and discuss their potential use in type theory.

\emph{Resolution} of \cite{robinson1965machine} is the one, if not the
most popular complete method for semi-deciding the refutation proof
problem: given a statement, we can apply this method; if the statement
is unsatisfiable in FOL, the method will eventually terminate with a
refutation of it. The design and implementation of such a method is
discussed at length in \cite{riazanov2002design}.  Many recent provers
rely on resolution, which has proved empirically to be the most
efficient. Since Robinson's discovery, many complete refinements have
been proposed since its introduction: ordered resolution,
selection etc. (see \cite{bachmair2001resolution} for an overview).
It consists of two phases: the transformation of the problem into
clausal form, and the resolution itself, which is the iterated
application of a (single, complete) rule mimicking the \emph{cut} rule
of natural deduction.

Equality is an ubiquitous notion in mathematics. Yet, FOL doesn't
include any special treatment for this symbol, which makes it very
inefficient to use in practice. A better way to treat the equality is
by means of rewriting. It is the goal of \emph{paramodulation}
(\cite{wos1968paramodulation}), \emph{superposition} and their
derivatives (see \cite{nieuwenhuis2001paramodulation} for a complete
survey).  They all are dedicated inference rules for treating equality
externally to the logic, as a mean to replace equals by equals in
formulas. For this purpose, \emph{completion}
(\cite{bachmair2001resolution})is a useful tool to orient equations
and avoid uncontrolled rewriting as much as possible.

Besides these popular techniques, the \emph{tableaux} method and their
variants (see \cite{hahnle2001tableaux}) is a proof search method for
first-order sequent calculus. It has been implemented in various
theorem provers, e.g. 3TAP (\cite{beckert1996tableau}) and the one
from \cite{paulson99}. FOL extended with inductive definitions has
been also the object of studies in ATP systems, resulting in methods
like \emph{rippling} (\cite{bundy1993rippling},
\cite{bundy2001automation}), or \emph{inductionless induction}
(\cite{comon2001inductionless}). The first is a method to cope with
the proofs of inductive cases by syntactic methods, the second is an
efficient embedding of induction into a extension of FOL.
Intuitionistic theorem proving is still an active field of research, a
recent success being \textsf{Imogen} from
\cite{mclaughlin2009efficient}, relying on the focusing discipline
(\cite{andreoli1992logic}).

\subsubsection{Proof search in ITP systems}

In order to give more inference power to existing proof assistants,
recent efforts have been put on adapting the efficient proof-search
methods designed by the automated deduction community to proof
assistants, by adapting them in order to generate proofs and translate
them to the tool's format, for it to recheck them afterward. We will
first do a quick review of the efforts in this domain.

\paragraph{Embedding first-order calculi}

First-order theorem automated theorem provers and the paradigms they
rely on are rich and very efficient. It would seem quite natural to
make use of these advanced tools for interactive theorem proving. One
can achieve this in two ways, either by interfacing an existing tool
through e.g. a dedicated tactic, or by designing a purpose-built
theorem prover that fits the particular needs of the interactive
system. \textsf{PVS} has a rich collection of decision procedures and
generic proof search tools. \textsf{HOL} was interfaced with various
first-order theorem provers such as resolution-based \textsf{Gandalf}
(\cite{hurd1999integrating}) and \textsf{Metis}
(\cite{hurd2003first}). \textsf{Isabelle} was first interfaced with
its dedicated prover based on the tableaux method by \cite{paulson99}
and more recently with \textsf{Vampire} (\cite{meng2006automation}). A
generic communication protocol between \textsf{Coq} and external
provers is being developed (the tactic \textsf{extern}). Since
recently, \textsf{Matita} uses its own paramodulation-based tool
(\cite{asperti2010paramodulation}).

\paragraph{Generic proof search in type theory}

Besides the embedding of first-order automation techniques, proof
assistants generally come with some generic proof search facilities,
devised this time entirely in the type theory. These tactics, known as
\textsf{auto} in \textsf{Coq} and \textsf{//} in \textsf{Matita}, have
at their core a mechanism analogous to the resolution process
introduced earlier. There are important differences with the
integration of external tools as introduced above: their integration
directly into the type theory allows them to exploit its full
higher-order nature, along with their implementation details like
definitions unfolding or inference mechanisms.

\subsection{Issues in ITP automation}

\subsubsection{Embedding first-order calculi}

The deep conceptual differences between both approaches (interactive
and automated theorem proving, higher-order vs. first-order) raises
questions that still need generic answers. We plan on developing the
line of work initiated by \cite{asperti2010paramodulation} in
\textsf{Matita}.

\paragraph{Proof translation}

Some of the interfaces from interactive systems to automatic ones
share the same architecture, at least in systems based on type theory.
We will expose here the approach taken in
\cite{asperti2010paramodulation} With the exception of \textsf{PVS}
which doesn't require a proof and trusts its decision procedures, they
all need to get back a \emph{trace} from the external tool to
reconstruct a proof in their own language. The architecture is then
composed of:
\begin{enumerate}
\item A forgetful translation from the type theory to the
  (first-order) language of the tool. It takes the current goal along
  with a set of hypothesis to a (axioms, problem) pair. This pair is
  run by the external tool and possibly returns a trace of the proof;
\item An interpretation function, which interprets the trace into a
  (possibly incomplete) proof object in the type theory;
\item A retyping mechanism, which infers the missing part of the
  returned proof object and checks it in the type theory.
\end{enumerate}

From this methodology, we see that the integration of external tools
relying on weaker or at least different logics is necessarily an
incomplete method based on heuristics, and it is often difficult to
predict their behavior.

The forgetful translation is not only the operation of removing all
type information from a statement to get a first-order term: type
theory possibly performs computations in statements, has defined
objects that can be unfolded. For example, how do we translate the
higher-order goal $\forall n\ m\ p, (n+1) * m \leq (n+1) * p \to m <
p$? Possibilities are numerous: $S\ n * m \leq S\ n * p \vdash S\
m\leq p$ or $m + n * m \leq p + n * p \to m+1 \leq p$ or any
intermediate steps of unfolding the constants $+$, $*$ or $<$ and
reducing their content. Depending on the set of axioms provided to the
external tool, some will be provable, some not.

The interpretation of trace object is also a delicate operation,
heavily depending on the calculus used by the external tool. Traces
are representations of the success of the proof process: they could be
the list of rewriting steps from the axioms to the problem statement
for paramodulation, or the list of clauses and literals used by the
resolution. The interpretation of a paramodulation step is pretty easy
since its logical meaning (rewriting) is admissible in type theory (by
means of the induction scheme of equality). It becomes quite involved
when considering a \emph{blind} method like resolution: the initial
preprocessing transforming the problem into a set of clause, which is
not admissible in intuitionistic logic, has to be traced carefully and
checked for validity. The same goes for another blind method, SMT
solving.

Finally the last part of the process -- retyping the result of the
interpretation -- is not guaranteed to be feasible since the type
information was lost in the first place. Sometimes the external prover
would success and return a valid trace of the first-order translated
goal, which has no meaning in the type theory and couldn't be
retyped. The inference of the missing type information can also be
quite involved and may itself rely on proof search (see
\cite{asperti2010paramodulation} for details).

\paragraph{Automated vs. interactive paradigms}

But the conceptual differences between automated and interactive
theorem proving go beyond the only untyped vs. typed or first-order
vs. higher-order problems.

One lies in the very purpose of automation in these two paradigms: ATP
systems are usually provided with a small set of axioms for a given
theory, often minimal, and deduce new facts from these onwards, until
the goal is reached. Thus, they generate \emph{cut-free} proofs, in
the sense that they don't infer useful intermediate lemmas that would
need to be proved afterward. On the other hand, automation in ITP
system usually relies on a huge set of objects, both axioms defined by
the user, logical rules in the form of constructors and destructors
for logical connectives, and intermediate proved lemmas, and this set
grows each time a new object is introduced.

From this emerges two difficulties of integrating ITP techniques into
ATP. The first is that provers are not necessarily optimized for
treating big sets of initial axioms, nor for dynamically extending
those sets as is required for an efficient integration. The
pre-treatment of generating and translating these sets might be
prohibitive if not correctly designed in the ITP software from the
beginning. The second and most important difficulty is the one of
choosing among those sets the needed objects. Usually, the whole
library of proof assistants contain very redundant objects: proofs
trivially subsuming other proofs (e.g. by computation, or by
superposition of two results), different proofs of the same statement,
different implementations of the same logical connective\ldots\ To be
treated efficiently and avoiding duplicated work, one would have to
choose in the whole set of objects of the library the relevant subset
for the given problem, minimal but with enough information for the
proof to succeed. This issue has not been throughly tackled and would
need a careful study.

\subsubsection{Generic proof search in type theory}
\label{sec:psitp}

The tactics and methods of proof search as implemented in \textsf{Coq}
and \textsf{Matita} can fulfill different purposes and usages, that we
can classify according to two criteria. First, we discriminate on the
\emph{purpose} intended by the user: it might be used to automatically
fill in some of the small ``trivial'' gaps leaved by the user in a
proof (what we call small-scale automation), for example leaving to
the machine the inference of one trivial case among others; it might
also be intended as a help to devise the proof when the user totally
lacks intuition of it (large-scale automation). This second purpose is
obviously more difficult to achieve, and it is probably unrealistic
considering the results obtained in the dedicated field of automated
theorem proving. Secondly, we separate two different \emph{modus
  operandi} of the automation mechanism: some of them allow to
automatically finish a proof and fill in the proof tree until its
leaves, may it be a trivial step of reasoning (small-scale) or a more
involved one requiring to explore deeply the search space
(large-scale). These take the form of classic tactics like
\textsf{auto} or \textsf{//}. Some of them on the other hand are
\emph{interleaved} in the proof process and allow to incarnate a
certain \emph{quotient} on applicable proof steps or statements. This
is the case of \textsf{Coq}'s type classes, or the recent \emph{smart
  apply} tactic of \textsf{Matita}.

\paragraph{Intuitionistic resolution}

As explained earlier, the resolution mechanism as usually presented is
classical by nature. However, it is well known that this appears as
nothing more than a notational facility, and its core reasoning scheme
is in fact intuitionistic. It proceeds simply by maintaining a list of
hypothesis and lemmas along with the goal; at each step, it applies in
turn all hypothesis and lemmas whose head unifies with the current
goal:
\begin{mathpar}
  \infer[Intro]{\Gamma,x:A\vdash B}{\Gamma\vdash \Pi x:A. B} \and
  \infer[Apply]{\Gamma\vdash A_1\ \ldots\ \Gamma\vdash A_n \and
    \Gamma\vdash B\equiv C \and f:\Pi (x_1:A_1)\ldots(x_n:A_n). B \in
    \Gamma}{\Gamma\vdash C}
\end{mathpar}

This \emph{backward-chaining} method (from the goal to the
hypothesis), although not complete, helps the completion of easy
goals. Because of the dependent nature of the statements in the CIC
however, this process is made more complex by the fact that the
application of a \textsc{Apply} rule can introduce
\emph{metavariables} in the goal, that is holes in a term of a known
type. For example, the application of a transitivity lemma on the
relation $R$ to a goal of the form $R(a,b)$ produces two goals sharing
a metavariable: $R(a,\mathsf{X})$ and $R(\mathsf{X},b)$. This
complicates greatly the control of the proof-search process: one could
na\"ively think of it as a backtracking process on each \textsc{Apply}
rules in turn, but this is erroneous: a goal, even closed by the
application of a 0-ary application (an instance of the \textsc{Init}
rule of natural deduction), could result in a backtracking if a
metavariable was instantiated. Imagine in our example that the goal
$R(a,\mathsf{X})$ is closed by a proof of $R(a,c)$, but that the
second instantiated branch $R(c,b)$ fails to be closed. One has then
to backtrack on the first goal rather than on the previous one, to be
able to find a second, successful instantiation e.g. $R(a,d)$ and
$R(d,b)$.

Due to some limitation of its implemented tactic language, this
subtlety is not taken into account in \textsf{Coq}; whereas it is
solved correctly by \textsf{Matita}. Both implement a depth-first
algorithm of the search tree.

\paragraph{Choice of hypothesis}

Contexts of proof assistants are not simple objects as are their
theoretical counterparts above: they are composed of many segments and
many kind of different objects, like definitions, inductive objects,
constructors, axioms. One question relative to these sort of
automation is: should we consider all of them as equally probable
candidates? Both proof assistants solve the question quite
differently. They both implement a mechanism of priority among these
objects, taking some heuristics to determine them like the number of
goals opened by the application or the number of metavariables
introduced. Besides that, \textsf{Matita} takes however the approach
of considering all objects as candidates except the most general ones
(induction principles for example that can be applied to any goal),
whereas \textsf{Coq} chose the approach of user-defined \emph{hints}
databases, associating to each lemma a priority and a strategy such as
\textsf{Resolve} (general strategy) or \textsf{Immediate} (the goal
has to be closed in one step after its application). Possible
directions like the refinement of these heuristics as well as the fine
control of the strategies are discussed in \ref{sec:strat}.

\paragraph{Definitions and computations}

Due to its higher-order nature and its definitional mechanism, the
logics underlying these proof assistants are subject to some notions
of reduction. Actually, the conversion rule of the CIC ensures that
the statements of the logic are all quotiented w.r.t computation.
These computations can be split into two different forms, that serve
different uses: first, some real computation can occur, as in $P(2+2)
= P(4)$ or $Q(S\ n * m) = Q(n * m + n)$, and a lemma applicable to one
of its part should be equally applicable to the other. The second is
the use of the \emph{definitional mechanism} to abbreviate some
often-used statement forms: for example, one might want to define
$\mathrm{transitive}(R)$ to expand to $\forall xyz, R(x,y) \to
R(y,z)\to R(x,z)$.

This is the responsibility of the \emph{unification} algorithm to
generally determine when these situation happen, when applying a lemma
for example. It is a complex part of a proof system, and often
involves expansive computations. Thus, iterating this process again
and again when trying to \textsc{Apply} all possible lemmas during
proof search can lead to substantial delays. To achieve good
performance, indexing techniques are generally used to rule out as
fast as possible lemmas that don't need to be treated for a given goal
because they won't ever unify. However, these data structures are
purely first-order, and it is non-trivial to use them enough so as to
improve the efficiency but not to restrict too much the candidates and
miss some possibly successful ones. These techniques have been
experimented, and are discussed in more details in
\ref{sec:implem}. In particular, the adaptation of these
data structures to take abbreviation-unfolding is a non-trivial, open
problem.

\subsection{Further directions}

Besides the integration of the first-order prover in \textsf{Matita},
We would like to improve the existing general-purpose proof search
techniques available in these proof assistants. The general
methodology adopted is to design a procedure directly for the Calculus
of Inductive Constructions, by focusing as much as we can on the
user's need. A preliminary analysis has been done by identifying clear
examples of situations where automation is wanted but that the tools
available (in \textsf{Coq} for the moment) are unable to cope with.

\subsubsection{Sequent calculus proof search}

The calculus of constructions is usually implemented in a natural
deduction fashion, at least in \textsf{Coq} and \textsf{Matita}: the
context is used only to retrieve previously defined object, and there
is no way to modify it after its introduction. This proof style is
supported by the \textsf{apply} family of tactics that applies a lemma
or an hypothesis to the goal, progressing in a
\emph{backward-reasoning} fashion: one progresses from the goal to the
assumptions. This approach is however not always intuitive, and a
second kind of reasoning is supported in both tools. The
\textsf{lapply} family of tactics (or \textsf{apply with}) apply a
lemma or an hypothesis to another hypothesis in the dual
\emph{forward-reasoning} fashion. In this case, one progresses by
transforming the assumptions until they eventually match the goal.
These two complementary approaches correspond to two presentation of
the theory of types, respectively the natural deduction and sequent
calculus style. Pure Type Sequent Calculus (PTSC) was introduced in
\cite{lengrand2006normalisation}.

\paragraph{Forward-reasoning}

The automation methods described in \ref{sec:psitp} are, to the best
of our knowledge, relying on a \emph{backward-reasoning} mechanism:
the system applies iteratively all possible applicable lemma and
hypothesis. This method has severe disadvantages: consider the sequent
$A\wedge B\vdash A$. The only way of proving it automatically is to
introduce a metavariable and apply the projection $proj2 : \forall X
Y, X\wedge Y \to X$ to produce the sequent $A\wedge B\vdash A\wedge
\mathsf{Y}$ and to conclude by instantiating \textsf{Y} to
$B$. However, allowing the application of these projections, and most
lemmas that produce metavariables leads to an enormous search space
since they can be applied to any goals. They should be avoided as much
as possible, preferring simpler proofs: here an obvious solution is to
apply the rule of sequent calculus:
$$ \infer{\Gamma, A, B\vdash C}{\Gamma, A\wedge B\vdash C} $$
to obtain the sequent $A,B\vdash A$ and conclude immediately.

In the CIC, the conjunction is a defined object. In fact, $\wedge$ is
an inductively defined non-recursive predicate with one constructor
and two independent hypothesis: $A\wedge B := conj : A \to B \to
A\wedge B$. By $\alpha$-conversion, the above rule of the sequent
calculus is valid if we replace $\wedge$ by any connective of the same
shape. It is also valid for connectives of the same ``kind'': if we call
conjunctions all inductive definitions having an unique constructor,
we can devise rules of forward-reasoning for all of them.
\begin{example} All of $\exists$, $\wedge$ and record types are
  conjunctions. Therefore we can devise the rules:
  \begin{mathpar}
    \infer{\Gamma, A, B\vdash C}{\Gamma, A\wedge B\vdash C} \and
    \infer{\Gamma, x:A, B(x)\vdash C}{\Gamma, \exists x:A. B\vdash C}
    \and \infer{\Gamma, x_1:A_1, x_n:A_n\vdash C}{\Gamma,
      \{x_1:A_1;\dots;x_n:A_n\}\vdash C}
  \end{mathpar}
\end{example}

\paragraph{The case of disjunction}

Can we treat inductive definitions with several constructors the same
way? The na\"ive application of the previous idea leads to a very
redundant proof search. Consider the goal $A\vee B\vdash C$. If we
apply the rule of left disjunction to it, we are left with two goals
$A\vdash C$ and $B\vdash C$. If we can prove the sequent $\vdash \pi :
C$ without using either $A$ or $B$, then the previous application
doubles the work: we have to copy the proof $\pi$ in both
sequent. Therefore, the left rules are to be applied more carefully
for disjunctions.

A second (but wrong) idea is to apply them lazily: we keep them
intact, until the moment where we meet one of their components in the
goal. Then we can destruct them and continue with the second branch:
$$ \infer*{\infer*{\infer*{\infer*{ }{\Gamma, A\vdash A} \and
      \infer*{\vdots}{\Gamma,B\vdash A}}{\Gamma, A\vee B\vdash
      A}}{\vdots}}{\Gamma, A\vee B\vdash C}
$$ Unfortunately this method is not complete, as it may be possible to
prove $\Gamma, B\vdash C$ but not the degraded judgement $\Gamma,
B\vdash A$.

We propose the following improvement: each disjunction sees its
components recorded, and proof search continues leaving the
disjunction untouched. If an instance of one of the component is met
later once a proof $\pi$ has been generated, we backtrack to the point
where the disjunction was introduced, break the disjunction, apply
$\pi$ on its corresponding component, and we are left with the second
branch of the disjunction in the hypothesis.
\begin{mathpar}
  \infer*{\infer*{\infer*{\vdots}{\Gamma,A\vee B\vdash A}}{\pi\left\{\vdots\right.}}{\Gamma,A\vee B\vdash C}
  \and \Longrightarrow \and
  \infer*{\infer*{\infer*{\infer*{ }{\Gamma,A\vdash
          A}}{\left.\vdots\right\}\pi}}{\Gamma,A\vdash C} \and
    \infer*{\vdots}{\Gamma,B\vdash C}}{\Gamma,A\vee B\vdash C}
\end{mathpar}

To sum up, this technique allows to rewrite part of a proof
\emph{during} proof search to optimize its number of inferences. This
way, the depth of a proof can be reduced and since most automated
proof search have their depth bound, we increase the inference power
of these techniques.

\newcommand{\nat}{\ensuremath{\mathbb N}}

Our goal is to generalize these techniques to all ``kinds'' of
inductive. The strategy described above could be generalized to
recursive inductive by taking into account the induction hypothesis,
then performing induction by itself:
\begin{mathpar}
  \infer*{\infer*{\infer*{\vdots}{\Gamma,n:\nat\vdash
        C(0)}}{\pi\left\{\vdots\right.}}{\Gamma,n:\nat\vdash C(n)}
  \and \Longrightarrow \and \infer*{\infer*{\infer*{\infer*{
        }{\Gamma,C(0)\vdash
          C(0)}}{\left.\vdots\right\}\pi}}{\Gamma\vdash C(0)} \and
    \infer*{\vdots}{\Gamma,m:\nat,C(m)\vdash
      C(\textsf{S}(m))}}{\Gamma,n:\nat\vdash C}
\end{mathpar}

By combining forward- and backward-reasoning step in an efficient way
and respecting the underlying logic, we believe that we can build an
automatic proof search method, not only more efficient than the
currently implemented methods, but also more expressive and able to
cope with the problem of deferring ``trivial'' proof steps to the
theorem prover, leaving to the user the load of only sketching the
proof.

\subsubsection{Strategic proof search}
\label{sec:strat}

Proof search however is not only about data structures and inference
rules. It is a dynamic process involving complex backtracking among
goals, and even more complex in the presence of metavariables. The
goal of fast automation requires a fine control on backtracking on the
search tree, as was already hinted in the previous section: attaining
a given goal made us backtrack to a special position and remember some
pieces of information (the proof $\pi$) to replay it without possible
backtracking on the next goal.

\paragraph{Focusing}

We can go further in this idea by putting into practice ideas that
emerged in the \emph{focusing} discipline of \cite{andreoli1992logic}.
Focusing was introduced as a winning strategy for proof search in the
Linear Logic of \cite{girard1995linear}, exploiting the redundancies
present in na\"ive proof search. These redundancies are actually not
on space, but on time, i.e. it is not a problem of data structure for
the proofs (as was solved by \emph{proof nets}), but a problem of
backtracking in the proof search. Indeed, it was identified that some
rules of Linear Logic are \emph{invertible}, that is, they can be read
in both directions, from top to bottom or from bottom to top. Thus the
application of such a rule does not change the provability of a
sequent and can be applied \emph{eagerly}. The eager application of a
rule is its application without the need to backtrack before them:
these rules just simplify the goal without modifying their
provability. Examples of invertible rules are the rules on conjunction
and disjunction, both left and right. This remark led to the
specification of \emph{polarities} defined on sub-formulas: the
\emph{negative} parts of a formula comprises all chain of invertible
connectives, whereas the \emph{positive} parts make up for the
non-invertible one. A \emph{phase change} occurs during proof search
when a connective is met that is not of the same polarity as the
previous one. Focusing is the strategy defined by:
\begin{itemize}
\item Applying eagerly all the connectives during the negative phase,
\item Focusing on an arbitrary hypothesis during the positive phase
  and apply eagerly all possible rules,
\item Backtracking occurs only at phase change.
\end{itemize}

These ideas, although having been defined for linear logic, can apply
to intuitionistic logic as well, by embedding one logic into another
by a (non-unique) polarized translation. The automated theorem prover
\textsf{Imogen} by \cite{mclaughlin2009efficient} is based on these
ideas, and proved to be a success.

A focusing approach applied to dependent type theory is still to
devise though, and would constitute a theoretical advance with clear
applications to automated proof search in type theory.

\paragraph{Backtracking strategies}

More generally, an efficient proof search requires a control of
backtracking that goes beyond what is currently implemented in
\textsf{Coq} and \textsf{Matita}: both proof assistants consider the
all leaf of the search tree as potential backtrack nodes, and don't
provide a way to specify strategies in a general way. Furthermore, no
distinction is made in the search between logical objects like the
defined connectives or the equality, and full lemmas with non-trivial
proofs. \textsf{Coq}'s $\mathcal L_{tac}$
(\cite{delahaye2001conception}) is an attempt to provide a
full-featured programming language for automatic tactic application.
It provides a recursion operator as well as a matching construct on
the current goal. However, it fails to define general and extensible
way of defining general-purpose automation tactics, specifically to
deal with non-determinism, and is confined to domain-specific
application, like the propositional fragment of logic
(\textsf{tauto}), or the decision of some order-related statements
(\textsf{order}).

We propose, as a first step towards the design of strategies, to have
a clear developer-side language of specification of these strategies,
expressive enough to cope with previous ideas: backtracking by keeping
a (partial) proof, focusing on a particular formula, eagerly advance
on the application of a particular rule. Also, we would like to
abstract from the \emph{names} of the inductive objects to select them
according to their \emph{kind}: (dependent, indexed, recursive)
disjunctions, conjunctions\ldots\ The design of such a general method
is of course interleaved with, and will be nurtured by the discovery
of particular wanted strategies. A good starting point could be the
backtracking monad of \cite{kiselyov2005backtracking}.

\subsubsection{Implementation issues}
\label{sec:implem}

When it comes to implementation, all the techniques mentioned above
require the manipulation of large sets of terms, and a recurring
problem is to retrieve terms satisfying some structural conditions,
like unification with a given pattern. One-to-one unification has a
well-known, simple and efficient algorithm, but it becomes
under-optimal when applied iteratively on large sets of terms, because
a lot of sharing is lost. Term indexes are data structure used to
store these sets, making the retrieval of unifiers to a given pattern
more efficient.

One data structure used for this purpose has been \emph{discrimination
  trees} or \emph{nets} (\cite{mccune1992experiments}). It operates on
the same principle as \emph{tries}, viewing the terms as flattened
strings and retrieving following a given prefix (the pattern), like a
dictionary. A critical inefficiency of this structure is due to the
loss of information resulting from the flattening of
terms. \emph{Substitution trees} (\cite{graf1995substitution}) are a
popular alternative. Their main idea is to index the terms according
to their \emph{mgu}'s, and form a tree of substitutions. All these
techniques are devised for first-order term algebras. With the
emergence of automated provers for higher-order logics like
\textsf{Twelf}, some attention has been drawn on the problem of
\emph{higher order indexing}, i.e. indexing of $\lambda$-terms modulo
$\beta$-reduction (\cite{pientka2003higher}).

This problem is in general undecidable, but such a behavior is
definitely a wanted feature for our use. We will focus at first on a
subproblem, already interesting and not yet tackled: the goal of
adapting these structure to be used modulo definitional equality to
abstract from certain abbreviations used commonly.

Moreover, if we decide to tackle the problem of the implementation of
sequent calculus proof search in a proof assistant, where the search
is not only goal-directed anymore but to some extend also directed by
the hypotheses, we then want to index hypothesis and be able to
retrieve efficiently a term having e.g. a given form of hypothesis.
However, due to the set nature of the context, the pair (hypothesis,
goal) cannot be presented as a term since it is to a certain extend
possible to permute all hypothesis. We then need to devise an
efficient data structure of term indexing taking into account to these
permutations. It seems not trivial to adapt existing term indexing
methods for this purpose. Moreover, hypothesis in a context are in
type theory possibly dependent: the instantiation of one hypothesis
triggers the modification of all subsequent dependent hypothesis. Our
data structure should be able to instantiate the variables
accordingly. This is however a known problem, and most of the time a
simple solution exists: the carry of a closure substitution during the
instantiation.

\subsubsection{Isomorphisms in the CIC}

The calculus of inductive constructions has the nice property of being
a very minimal foundational language, in which even very low-level
mathematical concepts can be defined. For example, the usual logical
connectors $\vee, \exists, \wedge\ldots$, even the equality are not
elementary objects but defined ones, thanks to the dependent and
inductive types features. However, this freedom implies that a given
usual \emph{concept} can have a lot of different
\emph{implementations}. Some of them will be convenient or efficient
for a given task, other for other tasks.

A simple example is the definition of the logical equivalence in
\textsf{Coq}'s standard library. It is historically defined as
$A\Leftrightarrow B := A\to B \land B\to A$, but a recent idea was to
change it to an inductive definition $A\Leftrightarrow B :=
\mathsf{build\_iff} : (A\to B)\to(B\to A)\to A\Leftrightarrow
B$. Although both definitions are strictly equivalent and represent
the same ``usual'' notion, such a change implied to rewrite almost all
proofs relying on the first definition and was then aborted.

Another less-trivial example is the representation of (natural)
numbers. They accept a wealth of different structural definitions ---
Peano, binary, machine, axiomatic integers\ldots --- each supporting
its own reasoning style, having different efficiency properties
etc. but, in the end, representing the same usual notion. When one
wants to begin a new development using natural numbers, she has to
choose from the beginning her implementation and stick to it
throughout the development. A change of representation would require
to rewrite a large part of the development and is therefore not
affordable; moreover, accessing the results, or using properties of
another representation is simply forbidden.

These facts suggest that a certain notion of \emph{isomorphism}, or
simply of \emph{morphisms} between the \emph{concepts} to be
formalized and their (possibly numerous) \emph{implementation} would
be a valuable study for both the modularity of the proofs (more
results to share among developments) and the abstraction of the proof
language, which would rely not on the actual implementation details of
structures but on higher-level concepts. The notion of type
isomorphisms has been already well-studied in the setting of
functional programming languages like ML, notably by
\cite{di1995isomorphisms} and proposals of integration in type theory
by \cite{barthe2001type} have been made.

There seem to be many different approaches to this problem, some
already considered, some not. A first approach is the definition of
axiomatic \emph{interfaces} for data structures or concepts, by means
of records or module system in the fashion of modular programming,
like it is done in the formalizations of algebra undertaken in the
CIC. A natural number is a ``\emph{complete}'' set of its
``\emph{atomic}'' properties (terms to be defined): successor lemmas,
recursion\ldots\ Individual representations of natural numbers are
then implementations of the interface. But what are these properties,
and how can one be sure that they are ``minimal'' or ``complete''
wrt. the common concepts, or appropriate for her particular needs?
Another pragmatic approach is to ``elect'' a canonical representation,
and to have translation functions/morphisms from each of the other
representations to it. This way, we lose their interesting structural
properties (e.g. efficiency), but we inherit from all properties
proved on them. A more \emph{bottom-up} but automatizable proposition
is to work directly at the syntactic level and assimilate isomorphic
terms in the calculus. Contrary to the other suggestions, this is a
meta-investigation of the isomorphisms in the logical framework that
could lead to the (partial) automation of reasoning modulo some
implementational details (like e.g. the \emph{iff} example
above). Inductive types seem to provide a rich framework for studying
these isomorphisms, since their purpose is precisely to provide
one-to-one correspondence between a ``concept'' (the inductive type)
and its possible alternative ``implementations'' (the
constructors). For example, we could assimilate function types which
are equal modulo permutation of arguments, or inductive types sharing
the same structure, types defined on an inductive type and the
corresponding substituted inductive type\ldots\ Building up on this
principle, we can hope to capture a large class of equivalences. A
third proposition is to rely on existing automation techniques to deal
with isomorphisms. Most instances of these isomorphisms can be
expressed as equalities, for example curryfying propositions: $\forall
A B C, A\to(B\to C) = (A\wedge B)\to C$. By providing a way to apply
logical rules up-to these equalities, we can somehow abstract from a
part of the logical specificities. This is the approach taken in the
new \texttt{sapply} facility of \textsf{Matita} described in
\cite{asperti2010paramodulation}.

The applications of such a study are numerous: abstract proof and
logical language mapped ``by construction'' to the foundational
language, modularity of the developments, increased automation\ldots

% \subsubsection{Proof mining}

% Over the years, large developments of formally verified proofs have
% been made in various fields of science, from mathematics to hardware
% verification, with the help of various interactive proof
% assistants. These large proofs along with many others are stored in
% repositories and can be rechecked at will. Yet, the information
% provided by these developments remains largely unexploited. Most proof
% assistants provide a very minimal set of interaction with previous
% developments: loading, reusing a result, searching by statement, and
% usually some statement-based automation like described above. In these
% interactions, the proofs themselves are totally passive, kept only to
% testify of the validity of the reasoning. Unlike their equivalents in
% programming languages---libraries---proof assistants repositories do
% not only consist of names of objects (functions, classes etc.)  but
% they also form a very structured and valuable source of information,
% both on the proof process and on the use one makes of the logical
% framework.

% By analyzing these large databases of already human-formalized proofs
% with the help of \emph{empirical methods} (statistics, data-mining,
% information theory, machine-learning), one could extract some useful
% information. It is expected to be of interest for a number of
% tasks. By analyzing them ``by hand'', some common reasoning schemes
% could be isolated and serve as a basis for improving or designing
% high-level proof languages and procedures, validating their design
% \emph{by the usage}.  It could also, as a side-effect, be treated
% automatically or semi-automatically, to provide a valuable help to the
% user in a given situation about what has been previously done in
% similar situations -- taking the form of suggestions or hints, or even
% to fully automatize some easy parts of proofs in a machine-learning
% fashion. The analysis of proof patterns could help also to
% automatically factorize large proofs, to infer useful intermediate
% lemmas or to compress the representation of a proof. This methodology
% has already been advocated and experimented in \cite{duncan2004udm}
% with the help of standard data-mining concepts, and \cite{urban08} in
% a machine-learning fashion, both leading to some interesting
% results. Their approaches however, concentrates on some features of
% the proofs. For example, they analyze the uses of lemmas and axioms in
% a given proof, preferring in a future situation a more productive
% lemma over one that hasn't been used at all. Proofs are complex
% objects, which structure itself is much richer than these features,
% and it might be even more productive to reflect the structure itself
% in the analysis.

% Among all information contained in the databases of formal proofs
% consisting proof assistants' standard libraries, the order of chaining
% of logical rules in the proofs, i.e. their very \emph{structure}, is
% already of crucial interest: it could contain enough information to
% both spot previously unknown practices of mathematicians, and even
% help the resolution of analogous cases, as an automation procedure. By
% extracting from these big databases of proof trees the common,
% recurring or generally interesting chaining of rules, that is some
% sub-trees of our trees, we can expect to isolate some information
% about the proof process, even some useful methods for future proofs.

% We propose to data-mine proofs by means of \emph{frequent subtree
%   mining}. Frequent subtree mining is a relatively new research field,
% dedicated to the extraction of recurring patterns inside the structure
% of trees (see an overview in \cite{chi2005frequent}). These methods
% will have to be adapted for our purpose: proof-terms have a certain
% structure that follows from the logical framework they were written
% in, whereas most algorithms in the literature have been devised for
% general trees. The direct reuse of a known algorithm has been proved
% in a preliminary investigation to be unadapted: the large majority of
% results are be non-relevant and the interesting ones drowned in this
% mass. Also, some of the notions used in the algorithms will have to be
% redefined: the notion of subtree differs in presence of (proof-)terms,
% and we will have to refine the definition of frequency to capture what
% are ``interesting'' patterns. Some preliminary results have been
% exposed by the author in \cite{puech2009formal}.

\newpage

\section{Semantic patches}

The second, complementary axis of this proposal is the attempt to
tackle the problem of modularity and evolution in formally checked
mathematical developments.

Even beside the notion of constructivism and the assimilation of
proving and programming in type theory, it is interesting to
investigate the possible relationship between the very
\emph{methodologies} employed in both fields, mathematics and computer
science. And by that we even intend the comparison of the \emph{daily
  workflow} of both scientist: one trying to prove a theorem, the
other constructing a program to fulfill a task. How does one or the
other elaborates his object of study? What kind of \emph{a posteriori}
modification is he prone to doing? What do these modifications imply
on the validity of the whole edifice? How to relate how both
scientists collaborate in a team? How do they rely on existing work to
build up new results?

Recent efforts in the formalization of mathematical results have
naturally led to these questions and many of them remain largely
unanswered, but the tendency seems to be to adapt existing methods
coming from software development, witness for example the recent
introduction of modules or separate compilation in proof assistants
like \Coq or \Matita, the use of dependency management tools
(\textsf{make}) or version control system (\textsf{git}, \textsf{svn})
to build and manage versions of a project. The use of software
engineering techniques for the management of a large formalization
project is even the thesis advocated in the manifesto of the
Mathematical Components team (\cite{gonthiermanifesto}).
\\[-0.5em]

We propose to address a small part of these questions, namely the
enhancement and adaptation of version control systems to the
management of mathematical repositories, by means of what we
preliminarily call a \emph{semantic patch system}\footnote{Joint work
  with \href{http://www.pps.jussieu.fr/~yrg/}{Yann R\'egis-Gianas}}.
It is a generic system allowing one to express \emph{transformations}
on formal developments, and check that some semantic properties are
preserved by the transformations, for example well-typing or
operational equivalence. Eventually, we hope to be capable of
expressing complex transformations on formal proofs, check them
efficiently, and manage distributed repositories of mathematics with
it, guaranteeing \emph{by typing} the stepwise global consistency of
the repository.

In the light of the Curry-Howard isomorphism though, we abstract
totally from whether we are talking about proofs or programs. That is
why, in the following, we start from today's situation, where existing
version control system are used both in software and proof
development, and we will not make any difference between both
use. Depending on his interest, the reader can thus choose his own
side of the isomorphism by translating program into proof, type into
statement\ldots\ or vice versa.

\subsection{Motivations}

The programmer and the mathematician know very well that developing a
new theory or program is not a linear task: one do not open her text
editor and write down her ideas in a unique pass. In the case of
program development, most of the time is actually spent editing
previously written code, correcting bugs or implementing new features
which are interleaved with previous ones. This idea is not present at
all in the text-editor/compiler paradigm, but to reflect it,
\emph{version control systems} usually rely on a textual
differentiation mechanism (\emph{diff}s): the comparison between two
versions of a text file generates a \emph{patch} which indicates which
lines of text are to be added or deleted to transform one into the
other.

Developments, may it be proofs or programs, are usually split into
files, each of these containing a self-contained module of the whole
development, an ``atom'' in some sense. To compile -- or check -- the
whole developments, we use dependency management tools like
\texttt{make}. These tools generate dependencies among the files, and
launch the compilation only on files that have been changed since last
compilation, and their dependencies. This process, known as
\emph{separate compilation} has files as atomic objects, and
dependency generation is performed uniquely on them, ignoring their
internal structure.

As intelligent and widely adopted as these tools have gotten, we
believe that they are not adapted for proof developments. Indeed,
whereas compilation of a program is usually fast enough for the
programmer to rely on the usual interaction loop ((edit; compile)*;
commit)*, the operation of proof checking is usually too expensive
computationally to mimic this workflow. But even besides the time
factor, this ``traditional'' way of formalizing mathematics hinders
the process of mathematical discovery process: once a concept
contained in a file is compiled, it is considered frozen and any
changes to it require the recompilation of all files depending on it;
the linearity of the development also gives no room for alternate,
inequivalent definitions. This fact has nonetheless been shown to be
of crucial to the mathematical discovery process by
\cite{lakatos1964proofs}, and we believe that they should be taken
into account in the realization of mathematical assistants like
\textsf{Coq} or \textsf{Matita}.

Actually, we even dare to state that these tools and the workflow they
suggest were not even adapted for program development in the first
place, but were just a convenient approximation of the user's intent.
Here are some hints to justify this argument: first of all, \emph{any}
change in a source file requires full recompilation of all files
depending on it, whereas a finer management of dependencies is
possible (function-wise for example). The basic operations of a
version control system such as \texttt{svn} or \texttt{git} are often
very sensitive, error-prone and based on heuristics (example, a
\emph{merge}). This is among other due to the fact that the
signal-to-noise ratio of a textual \emph{diff} is often high, because
semantically void changes like altering indentation have the same
value as changing the argument of a function. Finally, as they are not
aware of the meaning of the text they are managing, these tools don't
provide any consistency guarantee on the resulting files (except
textual ones).

All these facts, if they are not a major drawbacks when applied to
weakly structured languages (e.g. assembler, or a \LaTeX\ document),
gain a crucial importance when considering a typed language (a
functional programming language for example) : we would like to make
sure that a patch does not break the well-typing (or any other
property) of a source code.  One more step forward, more
structure-awareness becomes \emph{essential} when they apply to a
proof language, where the structure (the typing) \emph{is} the only
valuable information.

\subsection{Methodology}

We propose here to devise a system for \emph{semantic patches}. It
substitutes the idea of textual transformation by:
\begin{itemize}
\item First preferring an abstract syntax tree representation instead
  of plain text;
\item Secondly embedding semantic data into the transformations, in
  order to be able to reason on them.
\end{itemize}

We want to design a \emph{language of patches}, able to represent
these transformations and their semantic properties, thus capturing
local changes in programs, as well as their global effect on the whole
project. We can see this goal as a refinement of the former idea of
``dependency''.

A semantic patch is a program intended to transform a program written
in an \emph{object language} $\mathcal L$. We focus, on the first
iteration of the project, on the semantic assertion of well-typing:
from the object language's syntax and typing rules $\mathcal L$, we
derive typing rules for the patch languages $\mathcal P(\mathcal L)$,
and we can then type the patch. Our motto is:

\begin{quote} \label{motto}
  A well-typed patch is the guarantee that it transforms a well-typed
  program into a well-typed program.
\end{quote}

In this sense, it is closely related to \emph{incremental typing}: if
a modification has been made on a program, it is expressible by a
patch. To ensure that it will result in a valid program, it suffices
to type the patch, and not the whole resulting program.

Moreover, our approach is meant to be totally formal, so that we can
prove not only validity of our transformations, but also the
completeness of our patch language and other properties. For this, we
place ourselves in a constructive, higher-order metalanguage, namely
the Calculus of Inductive Constructions (CIC), in which we will
formalize in turn:
\begin{enumerate}
\item The object language $\mathcal L$ of our choice
\item Its metatheory
\item The patch language $\mathcal P(\mathcal L)$, parameterized by
  $\mathcal L$.
\item Its metatheory
\end{enumerate}

\subsection{Related work}

Both the study of metatheorical properties as proof transformations
and of change impact in structured documents (e.g. proofs, programs)
is not a new subject.

The \textsf{Twelf} project (\cite{pfenning1999system}) is an
implementation of the Logical Framework (LF,
\cite{harper1993framework}). It is a logic programming language able
to represent logics or programming languages, and embeds an inductive
metatheorem prover able to construct transformations of the
$\Pi\Sigma$ form. It was used in \cite{anderson1993program} to devise
transformations of proofs in order to extract efficient programs.

The problems of managing a formal mathematical library have been dealt
with in various proof assistant and mathematical repositories. The
HELM project (\cite{asperti2000content}, \cite{asperti2006content})
was an attempt to create a large library of mathematics, importing
\textsf{Coq}'s developments into a searchable and browsable database.
Most ideas from this project were imported into the \textsf{Matita}
proof assistants (\cite{asperti2007hop}), especially a mechanism of
\emph{invalidation and regeneration} to ensure the global consistency
of its library w.r.t changes, with granularity the whole definitions
or proofs and their dependencies. The MBase project
(\cite{kohlhase2001mbase}) attempts at creating a web-based,
distributed mathematical knowledge database putting forward the idea
of \emph{development graph} (\cite{hutter2000management},
\cite{autexier2000towards}) to manage changes in the database,
allowing semantic-based retrieval and object-level dependency
management.

This idea, generalized over structured, semi-formal documents gave
birth to \texttt{\it locutor} (\cite{muller2008fine}), a fine-grained
extension of the \texttt{svn} version control system for XML
documents, embedding ontology-driven, user-defined semantic knowledge
which allows to go across the filesystem border. It embeds a
\emph{diff} algorithm, operating on the source text modulo some
equality theory to quotient the syntax. On the same line of work, we
should mention the \emph{Coccinelle} tool
(\cite{padioleau2008documenting}). It is an evolution over textual
patches, specialized on the C language, allowing more flexibility in
the matching process, and was developed to deal with the problem of
\emph{collateral evolutions} in the Linux kernel. It embeds a
declarative language for matching and transforming C source code,
operating on text modulo defined isomorphisms.

Our approach to the ``impact of changes'' problem seems novel on
several aspects: first, it applies uniformly on proofs and programming
languages by virtue of the Curry-Howard isomorphism, and because we
operate at the AST level. Secondly, by taking \emph{types} as a
witnesses for the evolution of a development, we refine the usual,
dependency-based approach for a finer granularity. Thirdly, the
formality of our approach ensures not only a maximal trust in the
system, but also the ability to ``meta-reason'' on the defined
language.
\\

In the following, we will give preliminary hints on the construction
of such a system, and how it could be useful to supersede not only
textual \emph{diff}s and version control systems, but also other
paradigms for modular programming and go towards an eased mathematical
discovery process with the help of proof assistants.

\subsection{Patches as metatheorem composition}

The approach we propose is the following: to transform a program into
another and preserve well-typing, we need to transform accordingly the
typing derivation of the source program into a valid derivation for
the target program. On one hand, these transformations, if performed
by hand, are very error-prone: the theory of the object language may
be tricky and side conditions are frequently omitted in informal
presentations. On the other hand, there is a wealth of formalization
techniques of programming languages in the literature, and numerous
implementations in proof assistants. Theoretical results on the object
language are called \emph{metatheorems}; if these metatheorems are
proved in a constructive formalism, like the one of the proof
assistants \textsf{Coq} or \textsf{Matita}, they constitute algorithms
of program transformation. Then we just have to choose an adequate,
``complete'' set among those. Take the simply typed $\lambda$-calculus
(STLC) for instance. Examples of metatheorems are:
\begin{itemize}
\item \textsf{app} : from two terms $t$ and $u$, we can always form
  the application $tu$. This is a purely syntactical constructor of
  the object language, promoted as a metatheorem;
\item \textsf{typ\_app} : if $\Gamma \vdash t : A \to B$ and $\Gamma
  \vdash u : A$ then $\Gamma\vdash tu : B$. It is part of the typing
  definition of the language, and its proof constitutes a
  transformation building an application node at the root of a
  program;
\item \textsf{inv\_app} : if $\Gamma \vdash tu : B$ then there is a
  type $A$ such that $\Gamma\vdash t:A\to B$ and $\Gamma\vdash
  u:A$. Its proof is a transformation decomposing an application into
  its two subprograms: the function and its argument;
\item \textsf{weaken} : if $\Gamma\vdash t : A$ and $x$ is not free in
  $t$ then $\Gamma, x:B\vdash t:A$. Its proof has no effect on the
  program itself, but enlarges the context in which it is typed.
\item \textsf{strengthen} : if $\Gamma,x:B\vdash t:A$ and $x$ is not
  free in $t$ then $\Gamma\vdash t:A$.  \\\ldots
\end{itemize}
Considering a formalization working directly on the abstract syntax
tree of the programming language (named, possibly implicitly
typed\ldots), our patch language is formed on the composition of a
given, carefully chosen set of metatheorems.

\subsubsection{Definitions}

\paragraph{Transformers}

We consider only a subset of the possible metatheorems, those of the
following form. A \emph{patch variable} is an element $x,y,\ldots$
taken in a countable, infinite set $\mathcal X$. An \emph{atom} is
either a syntactic category of the object language, or a meta property
we want to reason on, i.e. for STLC:
\begin{align*}
  A,B\ ::=\ &\mathrm{var}\ |\ \mathrm{term}\ |\ \mathrm{env}\ |\
  \mathrm{type}\\
  |\ &{}\mathrm{deriv}\ {of}\ (\mathcal X \times \mathcal
  X\times\mathcal X)\\
  |\ &{}\mathrm{variable}\ of\ (\mathcal X\times
  \mathcal X\times\mathcal X) \\
  |\ &{}\ldots
\end{align*}

A \emph{judgement} $j$ can be of two kinds: $x$ is an atom $A$, or $x$
has form $\mathsf{T}(x_1,\ldots,x_n)$, where \textsf{T} is a transformer.
\[ j,k\ ::=\ x:A\ |\ x=\mathsf{T}(x_1,\ldots,x_n)\]

Now a \emph{transformer} $\emph{T}$ corresponds to a metatheorem in
$\Pi\Sigma$ form, with added definitional constraints. The statement
of the metatheorem is ``reified'' into a syntactical \emph{signature}
of the form:
\[ \mathsf{T}\ ::\ j_1, \ldots, j_n \longrightarrow k_1,\ldots, k_m \]

The semantics of these signature is: all tuples on the left of the
arrow are arguments, all tuples on the right are results. Arguments of
kind $x:A$ are interpreted as bare arguments to the transformer,
arguments of kind $x=\mathsf{T}(\vec x)$ as \emph{constraining
  arguments}: the actual argument supplied must be the result of
applying $\mathsf{T}$ to $\vec x$. Dually, results of kind $x:A$ are
bare results of unknown form, whereas results of kind
$x=\mathsf{T}(\vec x)$ are results ensured to be constructed from the
application of $\mathsf{T}$.

\begin{example}
Here are the signatures of some of the metatheorems given above:
\begin{align*}
  \mathsf{app} :: {} & (t\ u : \mathrm{term}) \longrightarrow
  (v : \mathrm{term}) \\
  \mathsf{typ\_app} ::{}& (\Gamma:\mathrm{env}) (A\ B:\mathrm{type})
  (C=\mathsf{arrow}(A,B)) (t\ u:\mathrm{term}) \\
  {}& (d_1 : \mathrm{deriv}(\Gamma,t,C)) (d_2 :
  \mathrm{deriv}(\Gamma,u,A)) \longrightarrow (v=\mathsf{app}(t,u))
  (d:\mathrm{deriv}(\Gamma,v,B)) \\
  \mathsf{inv\_app} ::{}& (\Gamma:\mathrm{env}) (t\ u:\mathrm{term})
  (v=\mathsf{app}(t,u))(d:\mathrm{deriv}(\Gamma,v,B)) \longrightarrow
  \\ {}&
  (A:\mathrm{type})(C=\mathrm{arrow}(A,B))(d_1:\mathrm{deriv}(\Gamma,t,C))
  (d_2:\mathrm{deriv}(\Gamma,u,A))
\end{align*}
but we can also imagine higher-level transformers:
\begin{align*}
  \mathsf{cps} ::{}& (t:\mathrm{term}) \longrightarrow (u:\mathrm{term}) \\
  \mathsf{aa\_trans} ::{}& (A:\mathrm{type}) \longrightarrow
  (B:\mathrm{type}) \\
  \mathsf{typ\_cps} ::{}& (\Gamma:\mathrm{env}) (t:\mathrm{term}) 
  (A:\mathrm{type}) (d:\mathrm{deriv}(\Gamma,t,A)) 
  \longrightarrow \\ {}&
  (t'=\mathsf{cps}(t))
  (A'=\mathsf{aa\_trans}(A)) (d':\mathrm{deriv}(\Gamma,t',A'))
\end{align*}
\end{example}

\paragraph{Patches}

Given a set of transformers $\Sigma$ and their signatures, a
\emph{patch} $\Delta$ is a sequence of variable assignment constructed
from the application of a transformer $\mathsf{T}_i\in\Sigma$ to bound
variables, much like a \textsf{let} construct:
\begin{align*}
  \Delta\ {}::=& \vec x_1 = \mathsf{T_1}(\vec y_1);\ \ldots;
  \\ {}& \vec x_n = \mathsf{T_n}(\vec y_n)
\end{align*}

Here, all $\vec x_i$ are bound in subsequent assignments.  A
\emph{repository} is a patch with no free variable. A patch therefore
represents the evolution of a set of atoms: the free variables
represent what the patch expects as input, all bound variables the
newly constructed objects. A repository represents the history of
modification of a program, starting from ground up. Each variable has
an implicit type, taken into the set of atoms.

We now describe informally a typing discipline for the patches: a
transformer should be applied to exactly its number and type of
arguments, and bind exactly its number of results, which take their
type from the signature. Results of kind $x=\mathsf{T}(\vec y)$ are to
be interpreted as additional assignments specifying the structure of
the variable bound. A transformer with an argument of kind
$x=\mathsf{T}(\vec y)$ should be applied to a variable having exactly
been created by the definiens $\mathsf{T}(\vec y)$: same transformer
name \textsf{T}, same variable names $\vec y$.

As we said earlier, a well-typed repository (in the way we just
hinted) describes well-typed changes of its objects, for examples
programs and derivations. An important property of this typing
discipline is its independence w.r.t the metatheorems associated to
the transformers in $\Sigma$. All the typing process is done
\emph{entirely syntactically}, without having to peak into the
semantics of transformers. This allows the patch language to be only
parametric on $\Sigma$.

\newcommand{\interp}[1]{\llbracket{#1}\rrbracket}

\paragraph{Interpretation}

The same way we reified our metatheorem's types into a syntax (the
transformers) to obtain the patch language, we can do the opposite
transformation, i.e. interpret a patch or a repository $\Delta$ into
an object of the language. This is done the obvious way, by following
the definitions in $\Delta$ and interpreting transformers by their
associated metatheorem:
\begin{align*}
  \interp x_\Delta &= \Delta(x) \\
  \interp{\mathsf{T}(\vec y)} &= T(\interp{\vec y})
\end{align*}

The following result ensures that our motto (\autoref{motto}) is correct:
\begin{theorem}[Soundness]
    If $\Delta$ is well-typed and $d : \mathrm{deriv}(\Gamma,t,A) \in
    \Delta$, then $\interp d_\Delta$ is a derivation of
    $\interp\Gamma_\Delta\vdash\interp t_\Delta : \interp{A}_\Delta $
\end{theorem}

\subsubsection{Examples}

We make a pause here and carry on a simple example of repository
construction, assuming the object language to be the simply typed
$\lambda$-calculus. The transformers involved are the following:

\begin{align*}
  \mathsf{int}::{}&\mathrm{type} \\
  \mathsf{string}::{}&\mathrm{string}\To\mathrm{var}\\
  \mathsf{var}::{}&\mathrm{var} \longrightarrow \mathrm{term} \\
  \mathsf{nil}::{}&\mathrm{env} \\
  \mathsf{cons}::{}&(x:\mathrm{var})(A:\mathrm{type})(\Gamma:\mathrm{env})
  \To
  \mathrm{env} \\
  \mathsf{init}::{}&(\Gamma:\mathrm{env}) (x:\mathrm{var})
  (t=\mathsf{var}(x)) (A:\mathrm{type}) \\ {}&
  (v:\mathrm{variable}(\Gamma,x,A)) \To
  (d:\mathrm{deriv}(\Gamma,t,A))
  \\
  \mathsf{zero}::{}& (\Gamma:\mathrm{env})(x:\mathrm{var})
  (A:\mathrm{type})(\Gamma'=\mathsf{cons}(\Gamma,x,A)) \To \\ {}&
  (v:\mathrm{variable}(\Gamma',x,A)) \\
  \mathsf{succ}::{}& (\Gamma:\mathrm{env}) (x\ y:\mathrm{var}) (A\
  B:\mathrm{type}) (v:\mathrm{variable}(\Gamma,x,A)) \\ {}&
  (\Gamma'=\mathsf{cons}(\Gamma,y,B)) \To
  (v':\mathrm{variable}(\Gamma',x,A)) \\
  \mathsf{typ\_lam}::{}& (\Gamma:\mathrm{env})(x:\mathrm{var})(A\
  B:\mathrm{type}) (t:\mathrm{term}) \\ {}&
  (\Gamma'=\mathsf{cons}(x,A,\Gamma))(d:\mathrm{deriv}(\Gamma',t,B))\To
  \\ {}&
  (t'=\mathsf{lam}(t,x,A))(C=\mathsf{arrow}(A,B))(d':\mathrm{deriv}
  (\Gamma,t',C)) \\
  \mathsf{typ\_app}::{}&(\Gamma:\mathrm{env})(t\ u:\mathrm{term})(A\
  B:\mathrm{type}) (C=\mathsf{arrow}(A,B)) \\ {}&
  (d_1:\mathrm{deriv}(\Gamma,t,C))(d_2:\mathrm{deriv}(\Gamma,u,A))
  \To (d_3:\mathrm{deriv}(\Gamma,v,B)) \\
  \mathsf{weak}::{}&(\Gamma:\mathrm{env})(x:\mathrm{var})(A\
  B:\mathrm{type})(t:\mathrm{term})(\Gamma'=\mathsf{cons}(\Gamma,x,A))
  \\{}& (d_1:\mathrm{deriv}(\Gamma,t,B)) \To
  (d_2:\mathrm{deriv}(\Gamma',t,B))
\end{align*}

\begin{example}
  From the empty repository, we build a derivation for $\lambda^{int} x. x$
  by constructing the repository:
  \begin{align}
    \Gamma &= \mathsf{nil} \\
    A &= \mathsf{int} \\
    x &= \mathsf{string}(\mathtt{"x"}) \\
    t_0 &= \mathsf{var}(x) \\
    \Gamma' &= \mathsf{cons}(\Gamma,x,A) \\
    V_0 &= \mathsf{zero}(\Gamma,x,A,\Gamma') \\
    D_0 &= \mathsf{init}(\Gamma,x,t_0,A,V_0) \\
    (t_1,C,D_1) &= \mathsf{typ\_lam}(\Gamma,x,A,A,t_0,\Gamma',D_0)
  \end{align}
  The last variable, $D_1$ is of type $\mathrm{deriv}(\Gamma,t_1,C)$.
  By unfolding the definitions and interpreting back the transformers,
  this expression becomes $\vdash\lambda^{int}x.x : int\to int$ which
  is what is expected.

\end{example}

In this example, we didn't use anything more than the transformers
already provided by the language construction, so this was just a
simple derivation. By now it should be clear that we can already build
whatever derivation we need: we just reified the object language into
the patch language by means of its constructors.

\begin{example}
  \label{ex:transf}
  Suppose now that we want to transform our program $D_1$ into a
  program $\lambda x^{int} f^{int\to int}.\ f x$. We need to transform
  the derivations as well. We add the following to our previous
  repository:
  \begin{align}
    f &= \mathsf{string}(\mathtt{"f"}) \\
    t_2 &= \mathsf{var}(f) \\
    \Gamma'' &= \mathsf{cons}(\Gamma',f,C) \\
    V_1 &= \mathsf{zero}(\Gamma',f,C,\Gamma'') \\
    D_2 &= \mathsf{init}(\Gamma'',f,t_2,C,V_1) \\
    D_3 &= \mathsf{weak}(\Gamma',f,C,A,t_0,\Gamma'',D_1) \\
    (t_3,D_4) &= \mathsf{typ\_app}(\Gamma'',t_2,t_0,A,A,C,D_2,D_3) \\
    (t_4,D_5) &= \mathsf{typ\_lam}(\Gamma',f,C,A,D_4) \\
    (t_5,D_6) &= \mathsf{typ\_lam}(\Gamma,x,A,D,D_5)
  \end{align}

  The resulting objects $t_5$ and $D_6$ are respectively interpreted
  back as the term $\lambda x^{int} f^{int\to int}.fx$ and its
  derivation $\vdash \lambda x^{int} f^{int\to int}.fx : int\to
  (int\to int)\to int$.
\end{example}

\subsection{Directions}

This formalization of proof/program transformations is a simple, yet
powerful idea, which opens numerous perspectives. The remaining
points of this sections will be directions for further investigations,
not definitive results \emph{per se}.

\paragraph{A basis for transformers}

In order to be able to represent \emph{any} transformation on the
object language $\mathcal L$, we need to specify the set of
metatheorems that we choose to promote as transformers. Let's try to
identify a minimal set of transformers able to represent all
transformations, a \emph{basis} so to speak for the construction of
the patch language.

One approach would be to identify, with the help of justified
examples, a set of common transformations on $\mathcal L$, as they are
used in practice when programming in $\mathcal L$. We would then need
to prove that they are \emph{complete} w.r.t all possible
transformations (more on completeness later). For a common, functional
programming language, we can imagine the following transformations:
\begin{itemize}
\item Rename a function or an argument, and all its subsequent
  occurrences in their scope
\item Add an argument to a function, and add a default parameter to
  all its calls
\item Replace a name by another name of the same type
\item add a constructor to a type declaration, adding a branch to all
  its \texttt{match\ldots with} constructs
\end{itemize}

We prefer a more low-level, syntactical approach based on the tree
representation of all objects, both syntactical and of typing: we
promote as transformers all operations of \emph{construction} and
\emph{destruction} on these trees, namely the constructors and
destructors of all syntactical objects, and the typing rules and their
inversion. Both these categories of operations manipulate directly the
tree structures (insert/delete a node) and provide a unifying way to
describe transformations. This way, we are not only guaranteed to be
complete, but also to help the design of an important piece of the
architecture of our system that what we will introduce later: the diff
algorithm.

\paragraph{Maximal sharing and completeness}

Variables assignments by transformer application is a simple way to
serve two different purposes. The first is to implement a dependent,
multi-hypothesis, multi-conclusion logic if we see transformers as
logical connectives. The second is to name all intermediate
construction steps of all objects, to allow free non-linear reuse of
previously constructed object so as to maximize sharing between
objects in a repository. This is the \emph{memoizing} function of
variable assignments, and it is the way we do incremental typing.

Maximal sharing of constructed object in a repository is a wanted
property. Not only does it ensures that we save the most work, which
is simply an efficiency concern, but it also ensures that we can
address objects by their variables: It states

\begin{definition}[Maximal sharing]
  A patch $\Delta$ has \emph{maximal sharing} if every object is
  assigned to a patch variable only once, namely that for all
  $x,y\in\Delta$, $\interp x_\Delta\neq\interp y_\Delta$.
\end{definition}

Knowing this, what is completeness in our system? A first, trivially
true and not very interesting statement is the following:
\begin{theorem}[Dummy completeness]
  All valid transformation from $\Delta$ to a program $p$ and its
  derivation $\vdash p : A$ are expressible as a program in $\mathcal
  P(\mathcal L)$.
\end{theorem}
\begin{proof}
  Since our patch language contains all typing rules, we just have to
  reconstruct the derivation of $\vdash p:A$ entirely (without using
  $\Delta$ in any ways).
\end{proof}

Obviously, this notion of completeness is not the one intended, as we
want to maximally reuse previous results from $\Delta$. It suggests
another notion of completeness, that we still to formalize and prove:

\begin{conjecture}[Completeness]
  All valid transformation from $\Delta$ to a program $p$ and its
  derivation $\vdash p : A$ are expressible as a program in $\mathcal
  P(\mathcal L)$ with maximal sharing.
\end{conjecture}

\paragraph{Bootstrapping the system}

We thus have a method, which can be automated, to generate from an
object language $\mathcal L$ and its typing rules, a patch language
$\mathcal P(\mathcal L)$ that, we saw, is also typed. Then nothing
prevents us from reapplying the functor $\mathcal P$ to the resulting
object. What do we obtain? Following what we just said, $\mathcal
P(\mathcal P(\mathcal L))$ is a language describing repository
transformations. For example, it should be able to describe the
application of patches at the end of a repository, the way we
implicitly did in our example \ref{ex:transf}.

Can this transformation help us describe what would be the core of a
version control system? To answer this, we would need to describe all
primitive operation of these systems (application, merge, commutation
of patches\ldots), and encode them as programs into the language. In
particular, we should try carefully to define a certain notion of
concurrency: what are patches that can be applied concurrently, what
are those that need a special sequentiality? Many interesting, open
questions concerning this bootstrapping arise.  Here are some of them:

\begin{itemize}
\item What metatheorem can we prove about the language $\mathcal
  P(\mathcal L)$
\item Which of them shall we choose as transformers for $\mathcal
  P(\mathcal P(\mathcal L))$?
\item Do they depend on the object language $\mathcal L$?
\item What can we say about the ``free'' language $\forall \mathcal X,
  \mathcal P(\mathcal X)$
\end{itemize}

\paragraph{A \emph{diff} algorithm}

The user of such a system surely does not want to write by hand what
we wrote in our previous example. It is both very verbose, as it shows
and names all intermediate constructs, and not useful as the
construction of a typing derivation is usually not his responsibility
but the work of the type inference algorithm. Moreover, one does not
want to change as drastically their habits as not to write
programs/proofs anymore, but only patches applied to previous versions
of its project.

Therefore, an interesting line of work would be to devise an algorithm
for computing the \emph{diff} between a repository and a file. It
would take a program and return the patch from the current repository
to that program. This patch could be typed afterwards, to guarantee
that the program given was well-typed with minimum effort.

If we see the patch language as a dependent, multi-hypothesis,
multi-conclusion logic, the work of searching for a resembles
proof-search as in \autoref{sec:auto}. Indeed, starting from a repository
$\Delta$ and having written the target program $t : A$, we are looking
for a composition of transformers $\Theta$ (viewed as logical
constants) realizing: \[ \Theta\quad:\quad\Delta \To (x = t) (d =
\mathrm{deriv(\Gamma_0,t,A)}) \] (the actual logic in which this proof
search is conducted still need to be specified formally).

But we have to be careful in this process, not to redo previous
searches, i.e. reusing maximally the objects of $\Delta$. This is
usually absent from traditional proof search, in which the proof
accounts only for its existence. Recall that we chose as transformers
the exact set of insert/delete operation on the derivation tree built.
Our process in this sense resembles much more a \emph{tree edit
  distance} algorithm: the best ``proof'' we can find is the one
minimizing the distance between (one of) the trees already built in
$\Delta$, and the new program we are willing to commit to $\Delta$. Of
course, these techniques will have to be adapted to the case of trees
with binders.

This algorithmic search is a challenging direction as it needs to be
very efficient, to be transparent to the user and not make us loose
the time we gained by typing incrementally. Moreover, the dual view of
this process as algorithmic content of proofs in automated proof
search seems a novel approach.

\paragraph{Functors as first-class patches}

Describing transformation of proofs/programs seems not only
interesting to implement a type-safe version control system as we
hinted above, but can be useful to encode other mechanisms commonly
used in proof development and programming, like modules.

Modules were designed as a tool for generic programming, and was
integrated in many functional programming languages, its archetype
being \textsf{SML}. One implements a given \emph{module} (a set of
functions and values), and give it a \emph{signature}, i.e. the bare
specification of its exported objects. Then, \textsf{functors} are
second-class functions, mapping a module of a certain signature to
another module. These functors can be applied, resulting in a new
module.

Given a language including modules but not functors, we conjecture
that we can represent functors as patches in the patch language.
Applying a functor would be the operation of applying the patch in a
given place in the repository.

Consider the following, toy language for libraries:
\begin{mathpar}
  A ::= \alpha\ |\ A\to A \and
  t ::= x\ |\ \lambda x. t\ |\ t\ t \\
  D ::= \{x:A; D\}\ |\ \cdot \and
  d :: \textsf{let}\ x=t; d\ |\ \cdot \\

  \infer[Nil]{ }{\Gamma\vdash\cdot\Rightarrow\cdot} \and
  \infer[Cons]{\Gamma\vdash t:A \\ \Gamma, x:A\vdash d\Rightarrow D}
  {\Gamma\vdash \textsf{let}\ x=t; d \Rightarrow\{x:A; D\}} \\
  \infer[Init]{(x:A)\in\Gamma}{\Gamma\vdash x:A} \and
  \infer[Lambda]{\Gamma;x,A\vdash t:B}{\Gamma\vdash\lambda x.t:A\to B}
  \and \infer[App]{\Gamma\vdash t:A\to B\\\Gamma\vdash
    u:A}{\Gamma\vdash t u:B} \and
\end{mathpar}

As opposed to the STLC, this language does not represents programs but
\emph{libraries} of programs (simple modules), having an
implementation ($\mathsf{let}\ x=t;\ \mathsf{let}\ y=u;\ \ldots$) and
a signature ($\{x:A;\ y:B;\ \ldots\ \}$).

A functor in this language can be encoded as a patch, taking as input
(i.e. having as free variable) a declaration $d$ of type the signature
$D$ expected, from which it constructs a typing environment $\Gamma$
and a new library, well-typed in $\Gamma$. This way, we encoded
functors as transformation operations on modules and we provided a
unified approach to them. Although this approach would require to be
formaized more carefully, we can already raise an interesting question
besides it: in the second-order language ($\mathcal P(\mathcal L)$) we
encoded functors, but we know that we can iterate this process
\emph{ad infinitum}. Can we then represent higher-order functors, like
functors taking functors as arguments, in $\mathcal P(\ldots (\mathcal
P(\mathcal L))\ldots)$?

\paragraph{User interaction}

Along with being a theoretical study on the impact of transformations
in proofs and programs, we believe that some practical application
could already come out, namely a simple, linear version control
system. We can already imagine the kind of user interaction this tool
would allow for the editing of developments: first of all, liberated
from the separate compilation paradigm, we don't edit whole files
anymore, but functions and values. For example, a user could ask:
\begin{verbatim}
> edit Set.add
\end{verbatim}
which would spawn a text editor with this function displayed (the
actual text being generated by pretty-printing the associated
\emph{term} sub-object of the current program). He then edit the
function and save the file. At this point, a patch is generated by the
\emph{diff} algorithm, possibly not well-typed. Then the patch is
temporarily committed (added to the repository at the current point),
and its type-checking begins. If it type-checks correctly, the patch
is definitively added and the system waits for a new user interaction.
If it doesn't type-checks, i.e. if there is a type error either in the
edited function or in the subsequent portion, then the errors are
displayed (possibly with additional context), and the user is asked
how to solve them. This is done either by hand, or by calling an
appropriate automatic strategy, defined by a high-level
transformer. Examples of high-level automatic resolution strategies
are:
\begin{description}
\item[$\alpha$-conversion] If we changed the name of an element of the
  context, we can decide to automatically rename all its occurrences;
\item[default argument] If we added an argument to a function, we can
  add a default parameter to all its occurrences;
\item[weakening] If the environment has been enlarged at some point,
  weakening allows us to modify any subsequent type derivation with
  the new environment;
\item[strengthening] If a name is not used in any part of the program,
  we can safely delete it from the environment;
\end{description}

\vspace{1em}

On the long term, we see other directions on this project. They
include:
\begin{itemize}
\item The use of special representations of the terms of the language
  using forward-pointers from binders to their occurrences,
\item The definition \emph{in the language} of patches of the
  higher-level transformers we described above,
\item Investigate the relationship with the version control tool
  \texttt{git}, as the structure of our repository resembles the
  \emph{object database} and its content-addressable namespace;
  possibly import ideas from it,
\item Approximation of type systems: when complex type systems are at
  stake (like dependent type theory), we could possibly devise a
  conservative approximation and use the real typing algorithm as a
  black box for retyping parts of the constructions, and still remain
  more efficient than typing the whole program,
\item Finally, we focused here on the semantic property of typing, but
  we could experiment other properties, like the preservation of
  operational semantics. The applications are numerous for programming
  (certified refactoring for example) but it seems more difficult
  because the base property is in the general case undecidable.
\end{itemize}

\bibliographystyle{plainnat}

\bibliography{../biblio}

% \appendix

% \section{ATP techniques}

% \subsubsection{Resolution}

% Resolution is the one, if not the most popular complete method for
% semi-deciding the refutation proof problem: given a statement, we can
% apply this method; if the statement is unsatisfiable in FOL, the
% method will eventually terminate with a refutation of it. The design
% and implementation of such a method is discussed at length in
% \cite{riazanov2002design}.  Many recent provers rely on resolution,
% which has proved empirically to be the most efficient.

% It was discovered in 1965 by \cite{robinson1965machine}, but since
% then many complete refinements have been proposed since its
% introduction: ordered resolution, selection\ldots (see
% \cite{bachmair2001resolution} for an overview). It consists of two
% phases: the transformation of the problem into clausal form, and the
% resolution itself, which is the iterated application of a (single,
% complete) rule mimicking the \emph{cut} rule of natural deduction.
% This is done with the help of a \emph{unification} algorithm to select
% relevant clause.

% Consider a set of formulas $H_i$ (the hypotheses) and a particular
% formula $G$ (the goal) of classical FOL. The satisfiability problem is
% the problem of finding assignations to the variables in $H_i, G$ such
% that the interpretation, in a boolean model satisfying all $H_i$, of
% the formula $G$ has truth value ``true''. Given a model of such type,
% it is easy to verify the satisfiability of the formula $G$ but the
% search for such a model is a very impractical exercise for a computer.
% We thus have to rely on \emph{syntactical methods} rather than such a
% semantic one, and resolution is one of them. It is easy to prove (by
% the De Morgan laws) that for all formula $B$, there is an
% equisatisfiable formula $B'$ which is in clausal normal form, i.e. has
% the form: \[ B = \bigwedge_{i=0}^{m}\left(\bigvee_{j=0}^n \pm
%   A_{ij}\right) \] where $A_{ij}$ are all \emph{literals}, i.e. terms
% of the signature and all variables in the literals are implicitly
% universally quantified.  Moreover, the satisfiability of the goal $G$
% provided that all $H_i$ are satisfiable is equivalent to the
% satisfiability of: $$\bigwedge_i(-H_i) \vee G$$ This allows us to work
% directly on sets of \emph{clauses}, or disjunctive set of literals and
% treat the literal as plain terms, without having to consider their
% semantics.

% Now let's take the negation of our starting formulas, and put it into
% clausal normal form. The resolution procedure works by iteratively
% enlarging this set of clauses with new clauses (the resolvants) until
% it generates an empty clause. The empty clause, writted $\bot$,
% denotes the unsatisifability of the clause, thus the unsatisifability
% of the whole set of clause, and if the method is sound, deriving
% $\bot$ means that the starting set of clause was unsatisfiable, i.e.
% that our formula is satisfiable, since the negation of a satisfiable
% formula is unsatisfiable. Here are the two rules of resolution:

% \begin{mathpar}
%   \infer[Resolution]{C,+A \and C',-A'}{(C, C')\sigma}\ 
%   \sigma=mgu(A,A') \and \infer[Factoring]{C,+A,+A'}{(C,+A)\sigma}
%   \ \sigma=mgu(A,A')
% \end{mathpar}

% The function $mgu$ computes the \emph{most general unifier} of two
% literals, and the resolution rule states that two unifiable literals
% of inverse signs in two sets of clauses can help generate a new clause
% formed by the concatenation of the previous sets, provided that we
% instantiate their variable by the result of the unification. The
% factoring rule, which is unnecessary in the degraded case of Horn
% clauses (clause with \emph{at most} one positive literal) but
% necessary for general clauses, states that two positive (resp.
% negative) unifiable literals can be factorized into one, provided that
% we instantiate the rest of the clause. We see that only a resolution
% step can lead to the final result: in the case where both premises
% are singleton clauses containing two unifiable literals, resolution
% leads to the empty clause.

% This method enjoys soundness and completeness.

% \begin{theorem}[Soundness]
%   Let $S$ be a set of clauses and $S'$ a set obtained by resolution from
%   $S$. All models of $S$ are models of $S'$.

%   In particular, if $\bot$ is deducible from $S$, then $S$ is
%   unsatisfiable.
% \end{theorem}

% Soundness is trivial to prove: both rules are interpreted valid in any
% model of first-order logic. From a proof-theoretic point-of-view, the
% resolution rule is much alike to the \emph{cut} rule of sequent
% calculus, and the factoring the contraction rule:
% \begin{mathpar}
%   \infer[Cut]{\Gamma,A \vdash \Delta \and \Gamma\vdash A,\Delta}
%   {\Gamma\vdash\Delta} \and
%   \infer[CR]{\Gamma\vdash\Delta, A, A}{\Gamma\vdash\Delta, A}
% \end{mathpar}

% \begin{theorem}[Completeness]
%   Let $S$ be a set of clause. It is unsatisfiable \emph{iff} we can
%   derive $\bot$ by the only application of \textsc{Resolution} and
%   \textsc{Factoring}.
% \end{theorem}

% Completeness is harder and requires complex proof techniques. A
% popular one is the \emph{semantic tree method}: we construct a
% \emph{Herbrand tree} of the set of clause, and consider the
% \emph{failure} nodes where the set becomes unsatisfiable. We see that
% resolution let these nodes climb up the tree, until it arrives to the
% root. Thus we prove that the iterated application of resolution is
% terminating, and that for an unsatisfiable $S$, $\bot$ is eventually
% derived. Note that this does not guarantee the termination of the
% process in all cases: if the given clauses are satisfiable, it
% \emph{will} loop forever. See \cite{goubault2005blossom} for an
% overview of this proof method.
% \\[-0.5em]

% The high non-determinism of this rule makes the bare resolution rather
% inefficient. Indeed, the number of clause in the system during
% resolution keeps growing exponentially: each application of resolution
% of factoring leads to a new clause which is added in the global pool
% and ready to be in turn combined with other at the next iteration. The
% search space is huge (unless a trivial case is met), and refinements of
% the resolution principle have been devised to reduce it when possible,
% i.e. by still remaining complete.

% \paragraph{Ordered resolution} One of the inefficiencies of this
% method is the possibility to combine every literal of a clause with
% every literal from another (or possibly the same) clause. It is
% possible to restrict the application of resolution to \emph{maximal}
% literal in every clause, according to an ordering on terms respecting
% some criteria. The rule becomes:
% \[ \infer {C,+A_1,\ldots,+A_n \and C', -A'}
% {(C,C')\sigma}\quad\sigma=mgu(A_1,\ldots,A_n,A') \] where $n\geq 1$,
% and $A_1\ldots A_n$ are all maximal in the first premise, $A$ in the
% second with respect to a strict, stable order $\succ$. Stable means
% that $A\succ B$ implies $A\sigma\succ B\sigma$ for all $A,B$ literals
% and $\sigma$ substitution. We emit the same restriction for the
% factoring rule.

% In practice, stable orders are chosen among reduction orderings, i.e.
% ordering expressing the fact that a term is ``smaller'' than another,
% so that unification has to do the minimum job possible.

% \paragraph{Ordered resolution with selection} An even bigger
% restriction on the application of the resolution principle is to
% parameterize it with respect to a special \emph{selection function}
% $sel : C \to \wp(C)$ that selects negative literals, whose role is to
% select a default choice of literals to resolve in a clause. The rule
% becomes:
% \[ \infer
% {\{C_i,A_{i1},\ldots,A_{in_i}\}_{1\leq i\leq l} \and
%   C',-A'_1,\ldots,-A'_l}
% {(C_1,\ldots,C_l,C')\sigma}
% \]
% where:
% \begin{enumerate}
% \item $n_i\geq 1$ for all $i$, $1\leq i\leq l$,
% \item $\sigma = mgu\{ A_{ij} = A'_i\}_{1\leq i\leq l,\ 1\leq j\leq n_i}$,
% \item $sel(C_i, A_{i1},\ldots,A_{in_i}) = \emptyset$ and
%   $A_{i1},\ldots,A_{in_i}$ maximals in their clause,
% \item either $sel(C',-A'_1,\ldots,-A'_l) = -A'_1,\ldots,-A'_l$ with
%   $l\geq 1$, or $sel(C',-A'_1,\ldots,-A'_l) = \emptyset$ and $-A'1$
%   maximal in its clause.
% \end{enumerate}

% As counter-intuitive as it may sound, even with an arbitrary selection
% function, this method remains complete.
% \\[-0.5em]

% As we have seen, these method of proof search are based on the
% decomposition of a first-order formula into a set of clauses by
% applying the De Morgan laws. Most of these laws are well-known to be
% false in intuitionistic logic, in fact most of them are true only in
% one direction. One could easily drop to the conclusion that these
% methods are inapplicable to an intuitionistic framework. In fact, it
% turns out that the process of resolution in itself is well-applicable
% in this case, and the preprocessing into clausal normal form is only a
% commodity. One could translate this process into a fully
% intuitionistic one by basing the system, not on clausal forms anymore,
% but on implicative form. What we get by doing this is a system that
% applies its hypothesis to the goal creating new goals, until all of
% them vanish. We will discuss this approach in more depth in
% \ref{sec:psitp}.

% \subsubsection{Equality treatment}

% Equality is an ubiquitous notion in mathematics. Yet, FOL doesn't
% include any special treatment for this symbol: it is defined
% axiomatically inside the logic (as the smallest congruent equivalence
% relation), which makes it very inefficient to use in practice: proving
% an equality triggers the blind application of the transitivity and
% symmetry lemmas in all possible ways:
% \begin{align*}
%   x=x & \qquad (reflexivity)\\
%   x=y \to y=x & \qquad (symmetry) \\
%   x=y \to y=z \to x=z & \qquad (transitivity)\\
%   x=y \to f(x)=f(y) & \qquad (congruence,\ \forall f)
% \end{align*}

% A better way to treat the equality is by means of rewriting. It is the
% goal of paramodulation, superposition and their derivatives. They all
% are inference rules for treating equality externally to the logic, as
% a mean to replace equals by equals in formulas. Again, a wealth of
% variants have been devised from the main, relatively inefficient idea:
% E-unification (paramodulation treated inside the unification
% algorithm), oriented paramodulation with selection. Completion is an
% important ingredient to the treatment of equality: Given a set of
% equalities, it is an algorithm for transforming it, when possible,
% into a set of confluent rewrite rules.  This accelerates the procedure
% of paramodulation without loss of completeness: we know that for some
% rules rewriting isn't necessary in both directions. Finally,
% congruence closure is a decision method for equalities in ground
% equational theories, used in practice in the SMT provers, a new
% generation of provers that has proved to be very efficient.

% \paragraph{Paramodulation}

% Paramodulation (See \cite{nieuwenhuis2001paramodulation} for a
% complete survey) originated as a development of resolution, introduced
% soon after it by \cite{wos1968paramodulation}: for improving
% resolution-based methods, the study of the equality predicate has been
% particularly important because its axiomatic treatment was unusable.
% This lead to the design of an additional, dedicated inference rule:
% \[ \infer{C, s=t \and D}{(C, D[t]_p)\sigma}\quad\sigma=mgu(s,D|_p)\]
% where $D|_p$ is the subterm at position $p$, and $D[t]_p$ denotes the
% result of replacing in $D$ the subterm at position $p$ with $t$. It is
% based on Leibnitz's law of replacing equal by equal: every instance of
% a member of an equation inside a clause can be rewritten to the other
% member of the equation. This applies as well to clauses containing
% equations. The addition of this rule to general resolution was proven
% to be a complete way to semi-decide FOL with equality, but, the rule
% being so general, it is almost unusable as-is.

% Later on, it was proved that paramodulation was unnecessary to perform
% into variables, that is, paramodulations where $D|_p$ is a variable.
% However, even under these restrictions, paramodulation is difficult to
% control: unless additional refinements are considered, it quickly
% produces a large amount of unnecessary clauses, expanding the search
% space excessively.

% \paragraph{Superposition}

% Superposition is a way to control the rule of paramodulation by
% restricting the application of the rule with respect to an
% \emph{order}, similarly to ordered resolution. The intuition behind
% superposition is to only rewrite \emph{big} terms into \emph{smaller}
% ones, these predicates being defined by a special \emph{rewriting
%   order} (a monotone, stable, well-founded order). Superposition
% consist in these two rules, parameterized on the order $\succ$:
% \begin{align}
%   \infer{C, l=r\and C',\pm (s=t)} {C,C', \pm (s[r]_p=t)}\qquad
%   &\text{if}\quad
%   \begin{array}{l}
%     s|_p\cong l, l\succ r, s\succ t \\
%     l\succ u\text{ for all } u\in C \\
%     s\succ v\text{ for all } v\in C' 
%   \end{array} \\
%   \infer{C, - (s = s)}{C}\qquad
%   &\text{if}\quad\ \ 
%   s\succ v \text{ for all } v\in C
% \end{align}

% Numerous term orderings have been proposed in the literature that
% tackle the problem of finding an ordering compatible with the
% requirements of superposition, general enough to capture the intuitive
% notion of the size of a term, and efficiently computable. Among the
% \emph{syntactic} term orderings, the Recursive Path Ordering (RPO) and
% the Knuth-Bendix ordering are the most successful. They both contain
% the subterm order, and satisfy the required properties: they are
% congruences (or monotone), stable under substitution and
% well-founded. By their syntactical nature, they are easily computable.
% \emph{Semantic} term ordering are rewriting orders based on semantical
% properties rather than their syntactic form. It usually requires to
% find a model of the pair of term which is often undecidable but can be
% approximated. Examples of such models are graph models (dependency
% pairs) or polynomial models (which requires to solve diophantine
% equations).

% \paragraph{Completion}

% The first use of the superposition rule has been by Knuth and Bendix
% for \emph{completion}. Completion is a procedure that attempts to
% transform a given rewrite system into an equivalent confluent one. It
% finds critical pairs in the rewrite system by superposing its
% equations, orienting them according to the given order. During the
% completion procedure, equations are simplified by rewriting and
% tautologies are removed. For example, group theory can be defined as
% the set of equations:
% \[ E\ :=\ \{ x\cdot e=x; I(x)=e; x\cdot(y\cdot z)=(x\cdot y)\cdot z
% \} \] and is transformed by the completion procedure into the
% confluent rewrite system:
% \[ R\ :=\ \{ x\cdot e\to x; I(x)\to e; x\cdot(y\cdot z)\to (x\cdot
% y)\cdot z \} \]

\end{document}

### 2010-06-18 papers/MIPS2010/abstract.tex
* Just move the abstract in a clean space to clarify my ideas. :)
git-svn-id: http://gasp.dev.regis-gianas.com/svn/trunk@99 403e655e-980a-47bd-8c60-1ee5348bb6b7

--- 2010-06-18 doc/abstract.tex
@@
 these transformations and their semantic properties, thus capturing
-local changes in programs, as well as their global effect on the whole
-project. We can see this goal as a refinement of the former idea of
-``dependency''.
+local syntactic changes in programs (the abstract syntax tree), as
+well as their global effect on the whole project (the typing
+derivations). We can see this goal as a refinement of the former idea
+of ``dependency''.
 

### 2010-06-18 tools/latexpp/whizzy.sh
* Preprocesseur.
git-svn-id: http://gasp.dev.regis-gianas.com/svn/trunk@100 403e655e-980a-47bd-8c60-1ee5348bb6b7

@@ -1,0 +1,1 @@
+export MKSLICE="./latexpp.opt -w -e ocaml ocaml-sf"

### 2010-06-18 papers/MIPS2010/abstract.tex
* Preprocesseur.
git-svn-id: http://gasp.dev.regis-gianas.com/svn/trunk@100 403e655e-980a-47bd-8c60-1ee5348bb6b7
* Quelques commentaires.
git-svn-id: http://gasp.dev.regis-gianas.com/svn/trunk@101 403e655e-980a-47bd-8c60-1ee5348bb6b7

--- 2010-06-18 doc/abstract.tex
@@
 \usepackage{color}
-\usepackage[pdftex,backref=page,colorlinks=true]{hyperref}
+\usepackage[backref=page,colorlinks=true]{hyperref}
 \usepackage{amsmath, amstext, amsthm, amsfonts}
@@
 \usepackage{fancyhdr}
-\input xy
-\xyoption{all}
-\usepackage[protrusion=true,expansion=true]{microtype}
+\usepackage[all,cmtip]{xy}
+%\usepackage[protrusion=true,expansion=true]{microtype}
 
@@
 
-\title{Semantic Patches
+% Commandes à désactiver avant la soumission.
+% Remarques sur la structure.
+\newcommand{\remplan}[1]{\noindent\textcolor{bwblue}{$\triangleright$ \textbf{#1}}}
+% Remarques sur le contenu.
+\newcommand{\remtext}[1]{\textcolor{bwgreen}{$\triangleright$ \textsl{#1}}}
+
+\title{Towards Semantic Patches}
   % pas sûr que le nom soit encore adapté
-}
+
 \author{Matthias Puech $\quad\Leftrightarrow\quad$ Yann Régis-Gianas}
@@
 \begin{document}
+
 \maketitle
@@
 
-It is interesting to investigate the possible relationship between the
+\remplan{Motivations générales}
+
+It may be fruitful to investigate the possible relationship between the
 very \emph{methodologies} or even \emph{daily workflow} employed in
@@
 way around, how do one rely on existing work to build up new results?
+% Y: Je ne comprends pas tout à fait la dernière question: 
 Finally, how to relate how both scientists collaborate in a team?
@@
 unanswered, but the tendency seems to be to adapt existing methods
-coming from software development, witness for example the recent
-introduction of modules, file-based scripts and separate compilation
-in proof assistants like \textsf{Coq} or \textsf{Matita}, the use of
-dependency management tools (\textsf{make}) or version control system
-(\textsf{git}, \textsf{svn}) to build and manage versions of a
+coming from software development, as illustrated for example by the
+recent introduction of modules, file-based scripts and separate
+compilation in proof assistants like \textsf{Coq}~{\cite{CoqDocWeb}}
+or \textsf{Matita}~{\cite{Asperti06userinteraction}}, the use of
+dependency management tools (\textsf{Make}) or version control system
+(\textsf{GIT}, \textsf{Subversion}) to build and manage versions of a
 project. Both for the development of proofs or programs, these tools
@@
 for alternate, inequivalent definitions. This fact has nonetheless
-been shown to be of crucial to the mathematical discovery process by
-\cite{lakatos1964proofs}, and we believe that they should be taken
-into account in the realization of mathematical assistants like
-\textsf{Coq} or \textsf{Matita}.
+been shown to be of crucial to the mathematical discovery
+process~\cite{lakatos1964proofs}, and we believe that they should be
+taken into account in the realization of mathematical assistants.
+% \textsf{Coq} or \textsf{Matita}.
+
+\remtext{Fri Jun 18, 2010 9:55. AM Ici, on pourrait rajouter un
+  paragraphe pour dire qu'il existe des travaux sur la formalisation
+  de metatheories dans les assistants de preuve, que ce soit en TWELF
+  ou Coq. Puis, on explicite notre but général qui est de construire
+  un système/framework qui remplace la chaîne d'outils traditionnelle
+  en important/adaptant la technologie issue de la formalisation de
+  metatheories.}
+
+\remplan{Sujet du papier : on focalise sur un problème bien précis}
+
+\remtext{Fri Jun 18, 2010 10:00 AM.
+  Un peu de liant. Pour arriver à notre but, il y a une
+  question importante à résoudre: si on te donne une méta-théorie
+  formalisée, es-tu capable de décrire l'historique d'un développement
+  dans cette théorie? Cela induit deux questions : dans quel langage
+  et est-ce que ce langage est utilisable (parce qu'a priori, ce n'est
+  pas gagné car le développeur travaille en mode ``instantané'' et ne
+  veut pas se trimballer ses casseroles tout le temps\ldots).}
 
@@
 
-In the light of the propositions-as-types paradigms where
-proof-checking boils down to type-checking, we abstract from whether
-we are talking about proofs or programs, so that our development could
-eventually also be used in traditional software development.
+\remtext{Fri Jun 18, 2010 10:06 AM.  Le paragraphe précédent n'est pas
+  assez précis je pense et en même temps en dit trop. À la place, je
+  propose d'énoncer succintement les contributions de ce papier puis 
+  le plan. Ce serait quelque chose comme:
+  1. A description language for semantic repositories.
+  2. Theory development through semantic patches. 
+  3. Related work
+  4. Further work
+}
 
-\section{Methodologies}
+\remplan{Fri Jun 18, 2010 10:08 AM. Du coup, je restructure un peu la suite.}
 
-We propose here to devise a system for \emph{semantic patches}. It
-% TODO changer le nom?
-substitutes the idea of textual transformation by:
-\begin{itemize}
-\item First preferring an abstract syntax tree representation instead
-  of plain text;
-\item Secondly embedding semantic, or typing data into the
-  transformations, in order to be able to reason on them.
-\item Finally, perform type-checking in an incremental way, that is
-  type only the syntactic difference and reuse derivation for the
-  already known subprograms.
-\end{itemize}
+\section{A kernel language to describe semantic repositories}
 
-We want to design a \emph{language of patches}, able to represent
-% TODO changer le nom?
-these transformations and their semantic properties, thus capturing
-local changes in programs, as well as their global effect on the whole
-project. We can see this goal as a refinement of the former idea of
-``dependency''.
+\subsection{How to describe a development?}
 
-\subsection{Data \emph{vs.} patch-oriented}
+\remplan{Fri Jun 18, 2010 10:09 AM. Pour commencer, on peut essayer 
+  de répondre à la question: ``À quoi sert ce langage?''. }
 
-\subsection{Related Works}
+\remtext{Fri Jun 18, 2010 10:10 AM. Il sert à décrire des objets d'une
+  théorie et, surtout, des transformations de ces objets (valides) en
+  de nouveaux objets (valides). Il sert donc à décrire des applications
+  de méta-théorèmes.}
 
-\section{A kernel for incremental type-checking}
+\remplan{Fri Jun 18, 2010 10:13 AM. Ensuite, est-ce qu'il n'existe pas déjà
+  des langages adaptés à ça? }
+
+\remtext{Fri Jun 18, 2010 10:15 AM. Pas tout à fait. Certes, LF
+  fournit un cadre qui autorise la représentation des metathéories et
+  des objets, construits par application de règles
+  admissibles/métathéorèmes. Cependant, la syntaxe de ces objets 
+  est statique/extensionnel~: on décrit l'objet comme le résultat
+  de sa construction en oubliant son histoire. Pour représenter 
+  l'historique d'un développement, notre langage
+  doit au contraire maintenir l'ordre dans lequel les objets ont
+  été construits ainsi que la façon dont ils l'ont été.}
+
+\remtext{Fri Jun 18, 2010 10:26 AM. Ce n'est pas peut être pas tout à fait vrai que personne n'a
+  fait de proposition de langages qui décrivent un développement ...
+  On peut peut-être citer Pfenning, Klint and co et faire une référence
+  en avant vers le related work.}
+
+\remtext{Fri Jun 18, 2010 10:20 AM. Remarque annexe: je pense qu'à ce
+  stade de l'explication, il n'est pas nécessaire de parler
+  d'incrémentalité et de partage maximal. Je crois que c'est une
+  conséquence pratique heureuse de notre formalisation mais qui ne
+  fait pas réellement partie des nécessités/besoins initiaux.
+  Ce que je veux dire, c'est que l'incrémentalité est une propriété
+  orthogonale à la capacité de dénoter une historique.}
 
@@
 
+\remtext{Fri Jun 18, 2010 10:26 AM. 
+  Du coup, il faudrait reformuler un peu les paragraphes précédents.}
+
+\subsection{Formalization}
+
+\remplan{On rentre dans les détails techniques. Je crois qu'il faut
+  prendre un peu de temps pour définir précisément les termes utilisés
+  (repository, patch, etc) et expliquer/paraphraser les règles.}
+
 \paragraph{Syntax} There are three syntactic categories: applicative
 terms $a$, types $t$ and environments $\Gamma$.
+
+% Y: Remarque en passant: pour maintenir des notations cohérentes, il
+% vaut mieux se définir des macros pour absolument tout
+% (metavariables, jugements, ...). Ca passe mieux à l'échelle.  On
+% verra ca pour le prochain papier :-).
 
@@
 
+\section{Development through semantic patches}
+
+\remplan{Fri Jun 18, 2010 10:50 AM. L'objectif ici, c'est de décrire
+  une (possible) architecture pour rendre utilisable le langage 
+  décrit dans la section précédente. En gros, on énumère les 
+  opérations qui interviennent dans l'utilisation et on les décrit
+  informellement à l'aide d'un exemple.}
+
+\remtext{Fri Jun 18, 2010 10:52 AM. Par contre, je ne sais pas quel
+  exemple utilisé. Le STLC? Est-ce que ça fait ``assez'' système
+  logique pour ce workshop?}
+
+\subsection{A layered architecture}
+
+% À mettre dans une figure. 
+\begin{figure}
+\begin{center}
+% FIXME: il y a un problème avec la flèche double phi/psi.
+\xymatrix{
+\textrm{History of metatheory objects}
+&R 
+\ar@{->}[d]_{\phi}
+& \Delta \ar@{->}[r]^{\theta(R)} & R' \\
+\textrm{Metatheory objects} 
+&t \ar@{->}[d]^{\pi}
+\ar@{->}[u]_{\psi}
+& t' \ar@{->}[u]_{\nabla(t)}\\
+\textrm{Concrete views on metatheory objects} 
+&w \ar@{.>}[r] & w' \ar@{->}[u]_{\sigma(t, w)}
+}
+\end{center}
+\label{fig:architecture}
+\caption{A layered architecture for a semantic patch framework.}
+\end{figure}
+
+\remtext{Fri Jun 18, 2010  1:32 PM. Ici, on décrit la figure en suivant
+  le modèle de flot du système. On décrit le rôle des fonctions, 
+les problèmes posés.}
+
+\paragraph{From an abstract proof tree to a concrete view, and back}
+\-
+
+\remtext{Fri Jun 18, 2010  1:36 PM. Un arbre de preuves, ce n'est
+  pas très pratique. Pourtant, c'est moralement sur cet objet 
+  que s'opère le développement et non sur le code source, qui
+  n'en est qu'une représentation. On justifie la nécessité de
+  projection et on fait référence aux lentilles de Pierce pour
+  expliquer $\sigma$ et $\pi$.}
+
+\remtext{Fri Jun 18, 2010 1:46 PM. Autre point important sur les vues
+  concrètes~: elles peuvent servir à parler de l'historique.  Voir le
+  module MetaAST de l'implémentation. Ces vues sont utiles par exemple
+  lorsque l'utilisateur doit construire un patch d'adaptation. Il a
+  alors toute la puissance des métathéorèmes ainsi que tous les objets
+  déjà construits pour décrire son patch.}
+
+\paragraph{Capturing the evolution of abstract proof trees}
+\-
+
+\remtext{Fri Jun 18, 2010 1:38 PM. Il faut différencier deux arbres de
+  syntaxe. Une méthode : la distance d'édition syntaxique permet de
+  déduire des applications des règles d'élimination et d'introduction
+  que l'on peut utiliser pour construire $\Delta$. Référence sur 
+  la littérature sur les distances d'édition entre arbres. Discussion
+  sur l'élaboration de patchs de haut niveau (comme le renommage etc). }
+
+\paragraph{Propagating changes through dependencies}
+\-
+
+\remtext{Fri Jun 18, 2010 1:42 PM. Si on se contente de concaténer
+  $\Delta$ à $R$, on n'a intégré qu'une nouvelle version de l'objet
+  initial~$t$ dans $R$ mais on n'a pas propagé ces modifications
+  auprès des clients de $t$. Cette propagation n'est \textit{a priori}
+  pas triviale~: rien ne dit que $t'$ à un sens équivalent à $t$. Il
+  faut donc adapter les clients en demandant un patch d'adaptation à
+  l'utilisateur. On présente ces demandes sous la forme de patch à
+  compléter (on fournit un peu de contexte et on a un trou pour y
+  injecter une nouvelle sous-dérivation mais on peut aussi modifier le
+  contexte pour l'adapter). La construction d'un patch d'adaptation
+  est un processus interactif~: on ne sait pas a priori borner le
+  contexte nécessaire à la construction du patchs d'adaptation donc on
+  peut imaginer un sorte de jeu où le système fait augmenter le
+  contexte (en suivant les dépendances) si l'utilisateur ne sait pas
+  construire le patch d'adaptation. En passant, quand on parle d'utilisateur
+  ici, ce peut être aussi bien un programme, comme un type-checker
+  pour le langage objet, ou bien un programmeur.}
+
+\subsection{Application on simply typed $\lambda$-calculus}
+
+\section{Related work}
+
+\remtext{Dolphin, Twelf, etc}
+
+\remtext{Semantic patches à la coccinnelle}
+
+\section{Further work}
+
+\remtext{Extension vers une version constructiviste?}
+
+\remtext{Algorithme d'élaboration de patchs?}
+
+\remtext{Parler du bootstraping?}
+
+\section{Conclusion}
+
+%% In the light of the propositions-as-types paradigms where
+%% proof-checking boils down to type-checking, we abstract from whether
+%% we are talking about proofs or programs, so that our development could
+%% eventually also be used in traditional software development.
+
+%% \section{Methodologies}
+
+%% We propose here to devise a system for \emph{semantic patches}. It
+%% % TODO changer le nom?
+%% substitutes the idea of textual transformation by:
+%% \begin{itemize}
+%% \item First preferring an abstract syntax tree representation instead
+%%   of plain text;
+%% \item Secondly embedding semantic, or typing data into the
+%%   transformations, in order to be able to reason on them.
+%% \item Finally, perform type-checking in an incremental way, that is
+%%   type only the syntactic difference and reuse derivation for the
+%%   already known subprograms.
+%% \end{itemize}
+
+%% We want to design a \emph{language of patches}, able to represent
+%% % TODO changer le nom?
+%% these transformations and their semantic properties, thus capturing
+%% local syntactic changes in programs (the abstract syntax tree), as
+%% well as their global effect on the whole project (the typing
+%% derivations). We can see this goal as a refinement of the former idea
+%% of ``dependency''.
+
 \bibliographystyle{plainnat}

### 2010-06-18 papers/MIPS2010/whizzy.sh

@@ -1,0 +1,2 @@
+export MKSLICE="../../tools/latexpp/latexpp.opt -w -g color yes"
+export TEXINPUTS="include":$TEXINPUTS

### 2010-06-18 papers/MIPS2010/abstract.tex
* Some quickly written paragraphs.
git-svn-id: http://gasp.dev.regis-gianas.com/svn/trunk@102 403e655e-980a-47bd-8c60-1ee5348bb6b7

@@
 
+The kernel language described in the previous section is clearly not
+adapted to human interaction, for several reasons~: 
+\begin{itemize}
+
+\item Every single object is flattened, which makes unobvious the
+  structure of a program or a proof.
+
+\item A term of this language denotes the entire history of a
+  development. Most of the time, the user is more interested in a
+  instantaneous snapshot of his object of study.
+
+\item Even though we were able to provide a snapshot of a particular
+  state of the development, the repository represents this state 
+  using proof terms, which contain too many syntactic details from the
+  user point of view and are not suitable for edition.
+
+\end{itemize}
+
 \remtext{Fri Jun 18, 2010 10:52 AM. Par contre, je ne sais pas quel
@@
 \subsection{A layered architecture}
+
+\remtext{Fri Jun 18, 2010  1:32 PM. Ici, on décrit la figure en suivant
+  le modèle de flot du système. On décrit le rôle des fonctions, 
+les problèmes posés.}
+
+With the previous remarks in mind, we designed a generic framework of semantic
+patches that can be described using three layers which are summarized in 
+Figure~\ref{fig:architecture}.
+
+The most internal layer focused on the history of metatheory
+objects. It is composed of a type-checker for the kernel language. The
+purpose of this layer is to maintain a repository~$R$ and the
+invariant that it is well-typed. Through the layer interface, a client
+can ask for the \textit{integration} $\theta(R)$ of a patch $\Delta$
+to obtain an extended version $R'$ of the repository $R$.
+
+The middle layer is a mechanized formalization of a metatheory. This
+module provides a syntax for the objects of the metatheory (formulae,
+judgments, \ldots) as well as the admissible rules (that we also call
+metatheorems).  An \textit{internalization} function $\phi$ translates
+the constructors of these objects and the specification of the
+admissible rules as terms of the kernel language. As a consequence,
+every object~$t$ of the theory can be injected into the
+repository~$R$. An \textit{externalization} function $\psi$ is
+implementing the inverse operation. Given a modified version $t'$ of
+an object~$t$, we are able to compute the patch $\Delta$ using
+the \textit{differienciation} function $\nabla(t)$.
+
+The external layer offers views on the objects of the theory. A
+view~$w$ is a human readable, and editable, partial representation of
+an object~$t$. A view is obtained from an object through a
+projection~$\pi$.  The user transforms the view $w$ into a view $w'$
+which must be $\textit{put back}$ into the metatheory as a modified
+object $t'$ using a $\textit{putback}$ function $\sigma(t, w)$.
 
@@
 \end{figure}
-
-\remtext{Fri Jun 18, 2010  1:32 PM. Ici, on décrit la figure en suivant
-  le modèle de flot du système. On décrit le rôle des fonctions, 
-les problèmes posés.}
 

### 2010-06-18 papers/MIPS2010/abstract.tex
* Un peu de texte en plus.
git-svn-id: http://gasp.dev.regis-gianas.com/svn/trunk@103 403e655e-980a-47bd-8c60-1ee5348bb6b7

@@
 % Remarques sur la structure.
-\newcommand{\remplan}[1]{\noindent\textcolor{bwblue}{$\triangleright$ \textbf{#1}}}
+%\newcommand{\remplan}[1]{\noindent\textcolor{bwblue}{$\triangleright$ \textbf{#1}}}
 % Remarques sur le contenu.
-\newcommand{\remtext}[1]{\textcolor{bwgreen}{$\triangleright$ \textsl{#1}}}
+%\newcommand{\remtext}[1]{\textcolor{bwgreen}{$\triangleright$ \textsl{#1}}}
+
+\newcommand{\remplan}[1]{}
+\newcommand{\remtext}[1]{}
 
@@
 
+% À mettre dans une figure. 
+\begin{figure}
+\begin{center}
+% FIXME: il y a un problème avec la flèche double phi/psi.
+\xymatrix{
+\textrm{History of metatheory objects}
+&R 
+\ar@{->}[d]_{\phi}
+& \Delta \ar@{->}[r]^{\theta(R)} & R' \\
+\textrm{Metatheory objects} 
+&t \ar@{->}[d]^{\pi}
+\ar@{->}[u]_{\psi}
+& t' \ar@{->}[u]_{\nabla(t)}\\
+\textrm{Concrete views on metatheory objects} 
+&w \ar@{.>}[r] & w' \ar@{->}[u]_{\sigma(t, w)}
+}
+\end{center}
+\label{fig:architecture}
+\caption{A layered architecture for a semantic patch framework.}
+\end{figure}
+
 With the previous remarks in mind, we designed a generic framework of semantic
 patches that can be described using three layers which are summarized in 
-Figure~\ref{fig:architecture}.
+Figure~\ref{fig:architecture}. We now describe the role of each layer.
 
@@
 which must be $\textit{put back}$ into the metatheory as a modified
-object $t'$ using a $\textit{putback}$ function $\sigma(t, w)$.
+object $t'$ using a $\textit{putback}$ function $\sigma(t, w)$.  In
+concrete syntax, an object~$t$ may be denoted simply using its
+syntactic definition but also, more interestingly, by a reference into
+the repository or even as the application of a metatheorem.
 
-% À mettre dans une figure. 
-\begin{figure}
-\begin{center}
-% FIXME: il y a un problème avec la flèche double phi/psi.
-\xymatrix{
-\textrm{History of metatheory objects}
-&R 
-\ar@{->}[d]_{\phi}
-& \Delta \ar@{->}[r]^{\theta(R)} & R' \\
-\textrm{Metatheory objects} 
-&t \ar@{->}[d]^{\pi}
-\ar@{->}[u]_{\psi}
-& t' \ar@{->}[u]_{\nabla(t)}\\
-\textrm{Concrete views on metatheory objects} 
-&w \ar@{.>}[r] & w' \ar@{->}[u]_{\sigma(t, w)}
-}
-\end{center}
-\label{fig:architecture}
-\caption{A layered architecture for a semantic patch framework.}
-\end{figure}
+The implementation of this architecture raises some interesting
+technical difficulties that can be handled, but only partially, 
+by existing technology. Indeed, even if internalization and 
+externalization are easily implemented thanks to the expressivity
+of the kernel language, the projection, putback, differenciation
+and integration operations are more intricate to realize.
 
@@
 \-
-
 \remtext{Fri Jun 18, 2010  1:36 PM. Un arbre de preuves, ce n'est
@@
   expliquer $\sigma$ et $\pi$.}
+
+The problem of updating some data according to a modification that was
+done on a partial view of it is a well-known issue in database
+litterature called the \textit{view-update} problem. Pierce \textit{et
+  al} have recently attacked it from a language design approach. Their
+idea is to develop projection and putback functions simultaneously by
+writing a so-called lens in a bidirectional programming
+language~\cite{pierce-bidirectional}. Roughly speaking, a lense is a
+software component that can be used to produce output from input but
+also input from output (hence the bidirectionality). A lense is able
+to keep track of the information that was erased by a projection to
+produce a projected output from the initial input. When we supply a
+modified output to the lense, it can elaborate, from the initial input
+and under certain conditions, the missing part of the modified input.
+
+We plan to design a bidirectional language to define the
+projection~function~$\pi$ and the putback~function~$\sigma$ as a
+lense. Yet, we cannot just import existing lenses combinators from the
+Harmony system that already works on trees. Indeed, most of the time,
+our trees have extra structural constraints induced by the presence of
+binders.
 
@@
 
+Computing the minimal edition script between two abstract syntax trees
+is a NP-hard problem that has been heavily studied, in particular in
+bioinformatics.  We guess that, in practice, we can restrict the
+search to local transformations that preserve substructures (that
+roughly consist in inserting, removing or moving around entire
+subtrees). In that case, an existing polynomial algorithm might be
+used~\cite{REF!}. Besides, abstract trees are relativily small in
+practice.
+
+We are more concerned about our ability to elaborate high-level patchs
+from these low-level edition scripts. Actually, it is easy to deduce
+from a low-level edition script, a low-level patch, based on object
+syntax constructor elimination and introduction. However, such a
+low-level patch may forbid a more efficient reuse of existing
+objects. For instance, let us assume a transformation of an
+object-level term~$x + x$ into~ he term~$2 + 2$. A low-level patch
+would represents the transformation as the replacement of the first
+subterm $x$ by the subterm~$2$ followed by the replacament of the
+second subterm $x$ by the subterm~$2$. If we had an object-level type
+derivation for $x + x$, the integration function would have to build
+two subderivations, one for each replaced subterms.  Instead, a
+high-level patch would represent that transformation as a substitution
+of the free occurrences of~$x$ by the term $2$ in the term~$x + x$. In
+that case, the integration function has the opportunity to apply a
+metatheorem about the stability of typing derivations under
+substitution and finally reuse the initial typing derivation
+directly. Inferring automatically such a high-level patch seems an
+unreachable objective~: there is clearly no most general high-level
+patch. Nonetheless, an interactive process can be designed to help the
+user building his high-level patch from a set of pre-computed
+interpretations of the low-level patch.
+
 \remtext{Fri Jun 18, 2010 1:38 PM. Il faut différencier deux arbres de
@@
   déduire des applications des règles d'élimination et d'introduction
-  que l'on peut utiliser pour construire $\Delta$. Référence sur 
-  la littérature sur les distances d'édition entre arbres. Discussion
-  sur l'élaboration de patchs de haut niveau (comme le renommage etc). }
+  que l'on peut utiliser pour construire $\Delta$. Référence sur la
+  littérature sur les distances d'édition entre arbres. Discussion sur
+  l'élaboration de patchs de haut niveau (comme le renommage etc). }
 

### 2010-06-18 papers/MIPS2010/abstract.tex
* Un peu de texte en plus.
git-svn-id: http://gasp.dev.regis-gianas.com/svn/trunk@104 403e655e-980a-47bd-8c60-1ee5348bb6b7

@@
 \end{itemize}
-It relies on two distincts judgments:
+It relies on two distinct judgments:
 \begin{itemize}
@@
 an object~$t$, we are able to compute the patch $\Delta$ using
-the \textit{differienciation} function $\nabla(t)$.
+the \textit{differentiation} function $\nabla(t)$.
 
@@
 externalization are easily implemented thanks to the expressivity
-of the kernel language, the projection, putback, differenciation
+of the kernel language, the projection, putback, differentiation
 and integration operations are more intricate to realize.
@@
 done on a partial view of it is a well-known issue in database
-litterature called the \textit{view-update} problem. Pierce \textit{et
+literature called the \textit{view-update} problem. Pierce \textit{et
   al} have recently attacked it from a language design approach. Their
@@
 writing a so-called lens in a bidirectional programming
-language~\cite{pierce-bidirectional}. Roughly speaking, a lense is a
+language~\cite{pierce-bidirectional}. Roughly speaking, a lens is a
 software component that can be used to produce output from input but
-also input from output (hence the bidirectionality). A lense is able
+also input from output (hence the bidirectionality). A lens is able
 to keep track of the information that was erased by a projection to
 produce a projected output from the initial input. When we supply a
-modified output to the lense, it can elaborate, from the initial input
+modified output to the lens, it can elaborate, from the initial input
 and under certain conditions, the missing part of the modified input.
@@
 projection~function~$\pi$ and the putback~function~$\sigma$ as a
-lense. Yet, we cannot just import existing lenses combinators from the
+lens. Yet, we cannot just import existing lenses combinators from the
 Harmony system that already works on trees. Indeed, most of the time,
@@
 is a NP-hard problem that has been heavily studied, in particular in
-bioinformatics.  We guess that, in practice, we can restrict the
+bio-informatics.  We guess that, in practice, we can restrict the
 search to local transformations that preserve substructures (that
 roughly consist in inserting, removing or moving around entire
-subtrees). In that case, an existing polynomial algorithm might be
-used~\cite{REF!}. Besides, abstract trees are relativily small in
+sub-trees). In that case, an existing polynomial algorithm might be
+used~\cite{REF!}. Besides, abstract trees are relatively small in
 practice.
 
-We are more concerned about our ability to elaborate high-level patchs
+We are more concerned about our ability to elaborate high-level patches
 from these low-level edition scripts. Actually, it is easy to deduce
@@
 would represents the transformation as the replacement of the first
-subterm $x$ by the subterm~$2$ followed by the replacament of the
-second subterm $x$ by the subterm~$2$. If we had an object-level type
+sub-term $x$ by the sub-term~$2$ followed by the replacement of the
+second sub-term $x$ by the sub-term~$2$. If we had an object-level type
 derivation for $x + x$, the integration function would have to build
-two subderivations, one for each replaced subterms.  Instead, a
+two sub-derivations, one for each replaced sub-terms.  Instead, a
 high-level patch would represent that transformation as a substitution
@@
 
+There is several ways of integrating a patch $\Delta$ inside a
+repository~$R$.  The naive attempt would be a simple concatenation
+of~$\Delta$ at the end of~$R$.  This kind of integration, though
+possible and correct, is of little interest because it does not take
+into account the fact that $\Delta$ probably defines new versions of
+objects already present in~$R$.
+
+A more useful integration propagates changes to the context that uses
+the modified object in order to trigger the updating this context
+too. Again, in full generality, it is not possible to update the
+context automatically. Instead, we choose to generate
+\textit{challenges}, which are adaptation patches with holes. These
+challenges can be addressed by the author that wrote the patch to be
+integrated, by the author of the context or automatically when there
+is no ambiguity (think of a context that uses a function whose
+body has changed but not its type).
+
 \remtext{Fri Jun 18, 2010 1:42 PM. Si on se contente de concaténer
@@
   contexte (en suivant les dépendances) si l'utilisateur ne sait pas
-  construire le patch d'adaptation. En passant, quand on parle d'utilisateur
-  ici, ce peut être aussi bien un programme, comme un type-checker
-  pour le langage objet, ou bien un programmeur.}
+  construire le patch d'adaptation. En passant, quand on parle
+  d'utilisateur ici, ce peut être aussi bien un programme, comme un
+  type-checker pour le langage objet, ou bien un programmeur.}
 
@@
 \remtext{Extension vers une version constructiviste?}
-
-\remtext{Algorithme d'élaboration de patchs?}
 

### 2010-06-18 papers/MIPS2010/abstract.tex
Reactive les commandes de commentaire.
git-svn-id: http://gasp.dev.regis-gianas.com/svn/trunk@105 403e655e-980a-47bd-8c60-1ee5348bb6b7
fix intro, add related works (a tailler)
git-svn-id: http://gasp.dev.regis-gianas.com/svn/trunk@106 403e655e-980a-47bd-8c60-1ee5348bb6b7

@@
 \usepackage[all,cmtip]{xy}
-%\usepackage[protrusion=true,expansion=true]{microtype}
+\usepackage[protrusion=true,expansion=true]{microtype}
 
@@
 % Remarques sur la structure.
-%\newcommand{\remplan}[1]{\noindent\textcolor{bwblue}{$\triangleright$ \textbf{#1}}}
+\newcommand{\remplan}[1]{\noindent\textcolor{bwblue}{$\triangleright$ \textbf{#1}}}
 % Remarques sur le contenu.
-%\newcommand{\remtext}[1]{\textcolor{bwgreen}{$\triangleright$ \textsl{#1}}}
+\newcommand{\remtext}[1]{\textcolor{bwgreen}{$\triangleright$ \textsl{#1}}}
 
-\newcommand{\remplan}[1]{}
-\newcommand{\remtext}[1]{}
+%\newcommand{\remplan}[1]{}
+%\newcommand{\remtext}[1]{}
 
@@
 
-\author{Matthias Puech $\quad\Leftrightarrow\quad$ Yann Régis-Gianas}
+\author{Matthias Puech, Yann Régis-Gianas}
 \date{}
@@
 taken into account in the realization of mathematical assistants.
-% \textsf{Coq} or \textsf{Matita}.
 
-\remtext{Fri Jun 18, 2010 9:55. AM Ici, on pourrait rajouter un
-  paragraphe pour dire qu'il existe des travaux sur la formalisation
-  de metatheories dans les assistants de preuve, que ce soit en TWELF
-  ou Coq. Puis, on explicite notre but général qui est de construire
-  un système/framework qui remplace la chaîne d'outils traditionnelle
-  en important/adaptant la technologie issue de la formalisation de
-  metatheories.}
+In fact, although dedicated tools exist to formalize the description
+of languages and their metatheory (e.g. \textsf{Twelf},
+{\cite{pfenning1999system}}), and substantial formalizations have been
+undertaken \remtext{TODO citer, mais quoi?}, we still use legacy tools
+based on text representation to manage our developments. The general
+goal exposed here is to replace this tool chain and make it
+language-aware, both on a syntactic side (AST) and on a semantic side
+(typing).
 
@@
 management of mathematical repositories, to witness with more
-precision the \emph{impact of changes}. By studying the dependencies
-between objects in a proof, we can get a much finely grained vision of
-what a future evolution would impact. In proof assistants based on the
+precision the \emph{impact of changes}. In proof assistants based on the
 propositions-as-types paradigm, Type Theory offers a powerful tool to
-witness the effect of changes finely, \emph{i.e.} types. Eventually,
-we hope to be capable of expressing evolutions on formal proofs, check
-them efficiently and incrementally, and manage distributed
-repositories of mathematics with it, guaranteeing \emph{by typing} the
-stepwise global consistency of the repository.
+witness the effect of changes finely: types.
 
-\remtext{Fri Jun 18, 2010 10:06 AM.  Le paragraphe précédent n'est pas
-  assez précis je pense et en même temps en dit trop. À la place, je
-  propose d'énoncer succintement les contributions de ce papier puis 
-  le plan. Ce serait quelque chose comme:
-  1. A description language for semantic repositories.
-  2. Theory development through semantic patches. 
-  3. Related work
-  4. Further work
-}
-
-\remplan{Fri Jun 18, 2010 10:08 AM. Du coup, je restructure un peu la suite.}
+We will describe here some of the possible directions to develop a
+tool to analyze the impact of change through types. It involves at its
+core a typed description language for repositories -- described in
+\S\ref{kernel} -- driven by a user interface that we sketch in
+\S\ref{devel}. In the first iteration of this project, we focus on a
+static, or data-driven model for repositories; in \S\ref{rw}, we
+quickly review related approaches, and hint a possible generalization
+in \S\ref{fw}.
 
 \section{A kernel language to describe semantic repositories}
+\label{kernel}
 
@@
 \section{Development through semantic patches}
+\label{devel}
 
@@
 \section{Related work}
+\label{rw}
 
-\remtext{Dolphin, Twelf, etc}
+The \textsf{Twelf} project (\cite{pfenning1999system}) is an
+implementation of the Logical Framework (LF,
+\cite{harper1993framework}). It was used in \cite{anderson1993program}
+to devise transformations of proofs in order to extract efficient
+programs.
 
-\remtext{Semantic patches à la coccinnelle}
+The problems of managing a formal mathematical library have been dealt
+with in various proof assistant and mathematical repositories. The
+HELM project (\cite{asperti2000content}, \cite{asperti2006content})
+was an attempt to create a large library of mathematics, importing
+\textsf{Coq}'s developments into a searchable and browsable database.
+Most ideas from this project were imported into the \textsf{Matita}
+proof assistants (\cite{asperti2007hop}), especially a mechanism of
+\emph{invalidation and regeneration} to ensure the global consistency
+of its library w.r.t changes, with granularity the whole definitions
+or proofs and their dependencies. The MBase project
+(\cite{kohlhase2001mbase}) attempts at creating a web-based,
+distributed mathematical knowledge database putting forward the idea
+of \emph{development graph} (\cite{hutter2000management},
+\cite{autexier2000towards}) to manage changes in the database,
+allowing semantic-based retrieval and object-level dependency
+management.
+
+This idea, generalized over structured, semi-formal documents gave
+birth to \texttt{\it locutor} (\cite{muller2008fine}), a fine-grained
+extension of the \texttt{svn} version control system for XML
+documents, embedding ontology-driven, user-defined semantic knowledge
+which allows to go across the filesystem border. It embeds a
+\emph{diff} algorithm, operating on the source text modulo some
+equality theory to quotient the syntax. On the same line of work, we
+should mention the \emph{Coccinelle} tool
+(\cite{padioleau2008documenting}). It is an evolution over textual
+patches, specialized on the C language, allowing more flexibility in
+the matching process, and was developed to deal with the problem of
+\emph{collateral evolutions} in the Linux kernel. It embeds a
+declarative language for matching and transforming C source code,
+operating on text modulo defined isomorphisms.
+
+Our approach to the ``impact of changes'' problem seems novel on
+several aspects: first, it applies uniformly on proofs and programming
+languages by virtue of the Curry-Howard isomorphism, and because we
+operate at the AST level. Secondly, by taking \emph{types} as a
+witnesses for the evolution of a development, we refine the usual,
+dependency-based approach for a finer granularity.
 
 \section{Further work}
+\label{fw}
 

### 2010-06-18 papers/MIPS2010/abstract.tex
factorisation de l'intro, introduction du langage kernel
git-svn-id: http://gasp.dev.regis-gianas.com/svn/trunk@107 403e655e-980a-47bd-8c60-1ee5348bb6b7

@@
 
-It may be fruitful to investigate the possible relationship between the
-very \emph{methodologies} or even \emph{daily workflow} employed in
-both fields of mathematics and computer science: one trying to prove a
-theorem, the other constructing a program to fulfill a task. How does
-one or the other elaborates his object of study? What kind of \emph{a
+Practical efforts in the formalization of mathematical results have
+naturally led to the questions of how to manage large repositories of
+proofs, i.e. what is the \emph{daily workflow} of users of a proof
+assistant. How does one elaborate a formal proof? What kind of \emph{a
   posteriori} modification is he prone to doing? What do these
-modifications imply on the validity of the whole edifice or the other
+modifications imply on the validity of the whole edifice or, the other
 way around, how do one rely on existing work to build up new results?
-% Y: Je ne comprends pas tout à fait la dernière question: 
-Finally, how to relate how both scientists collaborate in a team?
-
-Recent efforts in the formalization of mathematical results have
-naturally led to these questions and many of them remain largely
-unanswered, but the tendency seems to be to adapt existing methods
-coming from software development, as illustrated for example by the
-recent introduction of modules, file-based scripts and separate
-compilation in proof assistants like \textsf{Coq}~{\cite{CoqDocWeb}}
-or \textsf{Matita}~{\cite{Asperti06userinteraction}}, the use of
+Many of these questions remain largely unanswered, but the tendency
+seems to be to adapt existing methods coming from software
+development, as illustrated for example by the introduction of
+modules, file-based scripts and separate compilation in proof
+assistants like \textsf{Coq}~{\cite{CoqDocWeb}} or
+\textsf{Matita}~{\cite{Asperti06userinteraction}}, the use of
 dependency management tools (\textsf{Make}) or version control system
@@
 
-As intelligent and widely adopted as this tool chain has gotten, we
-believe that they are not adapted for the new, demanding requirements
-of proof developments. Indeed, whereas compilation of a program is
-usually fast enough for the programmer to rely on the usual
+We believe that these tools are not adapted for the new, demanding
+requirements of proof developments. Indeed, whereas compilation of a
+program is usually fast enough for the programmer to rely on the usual
 interaction loop ((edit; compile)*; commit)*, the operation of proof
@@
 management of mathematical repositories, to witness with more
-precision the \emph{impact of changes}. In proof assistants based on the
-propositions-as-types paradigm, Type Theory offers a powerful tool to
-witness the effect of changes finely: types.
+precision the \emph{impact of changes}. In proof assistants based on
+the propositions-as-types paradigm, Type Theory offers a powerful tool
+to witness the effect of changes finely: types, and we base our work
+on these.
 
@@
 
-\section{A kernel language to describe semantic repositories}
+\section{A core language to describe typed repositories}
 \label{kernel}
 
+The kernel of our system is a type-checker algorithm for a typed
+meta-language. In this language, we will declare both the syntax of
+the object (proof-)language and its typing rules, and define pieces of
+syntax (our proofs) and their derivations. Describing transformations
+among syntax objects is done by sharing common subterms or
+subderivations.
+
 \subsection{How to describe a development?}
-
-\remplan{Fri Jun 18, 2010 10:09 AM. Pour commencer, on peut essayer 
-  de répondre à la question: ``À quoi sert ce langage?''. }
-
-\remtext{Fri Jun 18, 2010 10:10 AM. Il sert à décrire des objets d'une
-  théorie et, surtout, des transformations de ces objets (valides) en
-  de nouveaux objets (valides). Il sert donc à décrire des applications
-  de méta-théorèmes.}
-
-\remplan{Fri Jun 18, 2010 10:13 AM. Ensuite, est-ce qu'il n'existe pas déjà
-  des langages adaptés à ça? }
-
-\remtext{Fri Jun 18, 2010 10:15 AM. Pas tout à fait. Certes, LF
-  fournit un cadre qui autorise la représentation des metathéories et
-  des objets, construits par application de règles
-  admissibles/métathéorèmes. Cependant, la syntaxe de ces objets 
-  est statique/extensionnel~: on décrit l'objet comme le résultat
-  de sa construction en oubliant son histoire. Pour représenter 
-  l'historique d'un développement, notre langage
-  doit au contraire maintenir l'ordre dans lequel les objets ont
-  été construits ainsi que la façon dont ils l'ont été.}
-
-\remtext{Fri Jun 18, 2010 10:26 AM. Ce n'est pas peut être pas tout à fait vrai que personne n'a
-  fait de proposition de langages qui décrivent un développement ...
-  On peut peut-être citer Pfenning, Klint and co et faire une référence
-  en avant vers le related work.}
-
-\remtext{Fri Jun 18, 2010 10:20 AM. Remarque annexe: je pense qu'à ce
-  stade de l'explication, il n'est pas nécessaire de parler
-  d'incrémentalité et de partage maximal. Je crois que c'est une
-  conséquence pratique heureuse de notre formalisation mais qui ne
-  fait pas réellement partie des nécessités/besoins initiaux.
-  Ce que je veux dire, c'est que l'incrémentalité est une propriété
-  orthogonale à la capacité de dénoter une historique.}
 
@@
 structure, and both can be rechecked at the same time, thanks to
-dependent types.
-
-For the purpose of incremental type checking though, our needs are a
-little bit different: first, we need to record, that is to name all
-intermediate values of our developments, so as to be able to address
-and reuse them multiple times. Secondly, we need to make sure that
-those intermediate values (sub-terms) are not recorded twice, so as to
-not type-check them twice. Thus we are looking to represent syntactic
-and typing objects as a \emph{DAG} rather than a tree, and ensure
-\emph{maximal sharing} among its sub-DAGs.
+dependent types. For the purpose of incremental type checking though,
+our needs are a little bit different: first, we need to record, that
+is to name all intermediate values of our developments, so as to be
+able to address and reuse them multiple times. Secondly, we need to
+make sure that those intermediate values (sub-terms) are not recorded
+twice, so as to not type-check them twice: we are looking to represent
+syntactic and typing objects as a DAG rather than a tree.
 
@@
 \end{itemize}
-
-\remtext{Fri Jun 18, 2010 10:26 AM. 
-  Du coup, il faudrait reformuler un peu les paragraphes précédents.}
 

### 2010-06-18 papers/MIPS2010/abstract.tex
* Remove reference to metatheorems.
git-svn-id: http://gasp.dev.regis-gianas.com/svn/trunk@108 403e655e-980a-47bd-8c60-1ee5348bb6b7

@@
 concrete syntax, an object~$t$ may be denoted simply using its
-syntactic definition but also, more interestingly, by a reference into
-the repository or even as the application of a metatheorem.
+syntactic definition but also, more interestingly, by a reference to this 
+object into the repository.
 
@@
 roughly consist in inserting, removing or moving around entire
-sub-trees). In that case, an existing polynomial algorithm might be
-used~\cite{REF!}. Besides, abstract trees are relatively small in
+sub-trees). In that case, an existing polynomial algorithm may be
+used~\cite{Touzet05alinear}. Besides, abstract trees are relatively small in
 practice.
 
-We are more concerned about our ability to elaborate high-level patches
-from these low-level edition scripts. Actually, it is easy to deduce
-from a low-level edition script, a low-level patch, based on object
-syntax constructor elimination and introduction. However, such a
-low-level patch may forbid a more efficient reuse of existing
-objects. For instance, let us assume a transformation of an
-object-level term~$x + x$ into~ he term~$2 + 2$. A low-level patch
-would represents the transformation as the replacement of the first
-sub-term $x$ by the sub-term~$2$ followed by the replacement of the
-second sub-term $x$ by the sub-term~$2$. If we had an object-level type
-derivation for $x + x$, the integration function would have to build
-two sub-derivations, one for each replaced sub-terms.  Instead, a
-high-level patch would represent that transformation as a substitution
-of the free occurrences of~$x$ by the term $2$ in the term~$x + x$. In
-that case, the integration function has the opportunity to apply a
-metatheorem about the stability of typing derivations under
-substitution and finally reuse the initial typing derivation
-directly. Inferring automatically such a high-level patch seems an
-unreachable objective~: there is clearly no most general high-level
-patch. Nonetheless, an interactive process can be designed to help the
-user building his high-level patch from a set of pre-computed
-interpretations of the low-level patch.
+We are more concerned about the fact that existing algorithms are not
+aware of binders that occur in proof-terms. Indeed, some standard
+edition scripts must be forbidden because they could break
+well-scopedness. As a consequence, we will have to adapt standard
+algorithms to only look for scope-preserving edition scripts.
+
+%% We are more concerned about our ability to elaborate high-level patches
+%% from these low-level edition scripts. Actually, it is easy to deduce
+%% from a low-level edition script, a low-level patch, based on object
+%% syntax constructor elimination and introduction. However, such a
+%% low-level patch may forbid a more efficient reuse of existing
+%% objects. For instance, let us assume a transformation of an
+%% object-level term~$x + x$ into~ he term~$2 + 2$. A low-level patch
+%% would represents the transformation as the replacement of the first
+%% sub-term $x$ by the sub-term~$2$ followed by the replacement of the
+%% second sub-term $x$ by the sub-term~$2$. If we had an object-level type
+%% derivation for $x + x$, the integration function would have to build
+%% two sub-derivations, one for each replaced sub-terms.  Instead, a
+%% high-level patch would represent that transformation as a substitution
+%% of the free occurrences of~$x$ by the term $2$ in the term~$x + x$. In
+%% that case, the integration function has the opportunity to apply a
+%% metatheorem about the stability of typing derivations under
+%% substitution and finally reuse the initial typing derivation
+%% directly. Inferring automatically such a high-level patch seems an
+%% unreachable objective~: there is clearly no most general high-level
+%% patch. Nonetheless, an interactive process can be designed to help the
+%% user building his high-level patch from a set of pre-computed
+%% interpretations of the low-level patch.
 
@@
 A more useful integration propagates changes to the context that uses
-the modified object in order to trigger the updating this context
-too. Again, in full generality, it is not possible to update the
-context automatically. Instead, we choose to generate
-\textit{challenges}, which are adaptation patches with holes. These
-challenges can be addressed by the author that wrote the patch to be
-integrated, by the author of the context or automatically when there
-is no ambiguity (think of a context that uses a function whose
-body has changed but not its type).
+the modified object in order to trigger the updating of this context
+(which may trigger the updating of other tree, and so forth). Again,
+in full generality, it is not possible to update the context
+automatically. Instead, we choose to generate \textit{challenges},
+which are adaptation patches with holes. These challenges can be
+addressed by the author that wrote the patch to be integrated, by the
+author of the context or automatically when there is no ambiguity
+(think of a context that uses a function whose body has changed but
+not its type).
 

### 2010-06-18 papers/MIPS2010/abstract.tex
Peaufinage de la description du kernel
git-svn-id: http://gasp.dev.regis-gianas.com/svn/trunk@109 403e655e-980a-47bd-8c60-1ee5348bb6b7

@@
 
-\remplan{On rentre dans les détails techniques. Je crois qu'il faut
-  prendre un peu de temps pour définir précisément les termes utilisés
-  (repository, patch, etc) et expliquer/paraphraser les règles.}
+We describe the state of a repository at a given moment by a type in
+the following syntax. Well-typed types in the repository meta-language
+guarantee that it contains only well-typed proofs in the object
+language.
 
@@
 terms $a$, types $t$ and environments $\Gamma$.
-
-% Y: Remarque en passant: pour maintenir des notations cohérentes, il
-% vaut mieux se définir des macros pour absolument tout
-% (metavariables, jugements, ...). Ca passe mieux à l'échelle.  On
-% verra ca pour le prochain papier :-).
 
@@
 dependent product (elsewhere written with a $\Pi$). The new construct
-$(x=a:t)\cdot u$ is a definition, whose value is $a$. We write as
-usual $(x_1:t_1)(x_2:t_2)\ldots(x_n:t_n)\cdot t$ for
+$(x=a:t)\cdot u$ is a definition binder, whose value is $a$. We write
+as usual $(x_1:t_1)(x_2:t_2)\ldots(x_n:t_n)\cdot t$ for
 $(x_1:t_1)\cdot(x_2:t_2)\cdot\ldots\cdot(x_n:t_n)\cdot t$, $t \to u$
@@
 
-\paragraph{Typing} The typing of our system is a restriction of the
-usual typing rules for dependently-typed systems with the additional
-two rules for typing a singleton product and the application of a
-singleton argument (featuring the syntactic verification in the
-environment). It depends on: 
+\paragraph{Typing} Typing is parameterized by:
 \begin{itemize}
@@
 \item $\Gamma\vdash a : t$ ($a$ has type $t$ in $\Gamma$),
-\item $\Gamma\vdash t : \sort$ ($t$ has sort $s$ in $\Gamma$).
+\item $\Gamma\vdash t : \sort$ (the repository $t$ is well-typed of
+  sort $s$ in $\Gamma$).
 \end{itemize}
@@
   \and
-  \infer[Sing]{\Gamma\vdash a:t \and
+  \infer[Let1]{\Gamma\vdash a:t \and
     \Gamma\vdash t:\sort_1 \and
@@
   }{\Gamma\vdash(x=a:t)\cdot u : \sort_3} 
-  \quad(\sort_1,\sort_2,\sort_3)\in\mathcal R
+  \quad (x=a:t)\notin\Gamma
+  \and
+  \infer[Let2]{\Gamma\vdash t:\sort_1 \and
+    \Gamma\vdash u:\sort_2
+  }{\Gamma\vdash(x=a:t)\cdot u : \sort_3} 
+  \quad(x=a:t)\in\Gamma
   \and
@@
   \and
-  \infer[Skip]{
+  \infer[Skip1]{
     \Gamma\vdash a:(y=b:t)\cdot u \and \Gamma(x)=b:t
+    }{\Gamma\vdash a : u\subst{x}{y}} \and
+  \infer[Skip2]{
+    \Gamma[x=b:t]\vdash a:(y=b:t)\cdot u
     }{\Gamma\vdash a : u\subst{x}{y}}
@@
 
-A term $t$ represents a state of the repository at a given moment. It
-contains the declaration of the object language's syntax, its typing
-rules, the syntactic objects and derivations recorded in it along with
-the patches that let the repository evolve. If $\cdot\vdash t:\sort$,
-then the repository is well-typed.
+The first four rules are standard. As application is only performed on
+variables, the \textsc{App} rule looks up directly in the
+environment. Rules \textsc{Let1} and \textsc{Skip1} show the
+verification and the use of the equality binder in case the definition
+is not already known. Rule \textsc{Let2} and \textsc{Skip2} allow
+incremental typing in our system: if the definition was already known
+in the environment, we do not need to recheck it. If $\cdot\vdash
+t:\sort$, then the repository $t$ is well-typed.
 

### 2010-06-18 papers/MIPS2010/abstract.tex
Exemple, le fichier stlc2.ga beautified
git-svn-id: http://gasp.dev.regis-gianas.com/svn/trunk@110 403e655e-980a-47bd-8c60-1ee5348bb6b7

@@
 \newcommand{\subst}[2]{\{#1/#2\}}
+\newcommand{\nat}{\mathbb N}
 
@@
 
-%\newcommand{\remplan}[1]{}
-%\newcommand{\remtext}[1]{}
+%\renewcommand{\remplan}[1]{}
+%\renewcommand{\remtext}[1]{}
 
@@
 t:\sort$, then the repository $t$ is well-typed.
+
+\paragraph{Example}
+
+Here is a well-typed repository describing the simply-typed
+$\lambda$-calculus and constructing two $\lambda$-terms sharing some
+sub-derivations: 
+{\small
+\begin{align}&
+(\nat : *).
+(o : \nat).
+(s : \nat \to \nat).
+(type : *).
+(base : type).
+(arr : type \to type \to type). \\&
+(env : *).
+(nil : env).
+(cons : type \to env \to env). \\&
+(term : *).
+(var : \nat \to term).
+(lam : term \to term).
+(app : term \to term \to term). \\&
+(lookup : env \to \nat \to type \to *).
+(lookup_o : (e : env).(a : type).(f = cons\ a\ e : env). lookup\ e\ o\ a). \\&
+(lookup_s : (e : env).(a b : type).(n : \nat).(sn = s\ n : \nat).(f =
+cons\ a\ e : env). lookup\ e\ n\ b \to lookup\ f\ sn\ b). \\&
+(of : env \to term \to type \to *). \\&
+(of_v : (e : env).(a : type).(n : \nat).(v = var\ n : term).
+          lookup\ e\ n\ a \to of\ e\ v\ a). \\&
+(of_l : (e : env).(t : term).(a\ b: type).(c = arr\ a\ b : type).
+          (u = lam\ t : term).(f = cons\ a\ e : env).
+          of\ f\ t\ b \to of\ e\ u\ c). \\&
+(of_a : (e : env).(t\ u: term).(a\ b: type).
+          (c = arr\ a\ b : type).(v = app\ t\ u : term).
+          of\ e\ t\ c \to of\ e\ u\ a \to of\ e\ v\ b). \\&
+(1 = s\ o : \nat).(v_0 = var\ o : term).(v_1 = var\ 1 : term).
+(t_3 = lam\ v_0 : term).(ch_0 = lam\ t_3 : term).\\&
+(t_1 = app\ v_0\ v_1 : term).(t_2 = lam\ t_1 : term).
+(t_4 = lam\ t_2 : term).
+(bb = arr\ base\ base : type). \\&
+(e_0 = cons\ bb\ nil : env).
+(e_1 = cons\ base\ e0 : env).
+(d_0 = lookup_o\ e_1\ base : lookup\ e_1\ o\ base). \\& %        (* a→a, a ⊢ 0 : a *)
+(d_1 = of_v\ e_1\ base\ o\ d0 : of\ e_1\ v_0\ base). %	   (* a→a, a ⊢ 0 : a *)
+(d_2 = of_l\ e_0\ v_0\ base\ base d_1 : of\ e_0\ t_3\ bb). \\& %   (* a→a ⊢ λ0 : a → a *)
+(bbb = arr\ bb\ bb : type).
+(d_3 = of_l\ nil\ t_3\ bb\ bb\ d_2 : of\ nil\ ch_0\ bbb). \\& %  (* ⊢ λλ0 : (a→a)→(a→a) *)
+(g_1 = lookup_o\ e_0\ bb : lookup\ e_0\ o\ bb).
+(g_2 = lookup_s\ e_0\ base\ bb\ o\ g_1 : lookup\ e_1\ 1\ bb). \\&
+(g_3 = of_v\ e_1\ bb\ 1\ g_2 : of\ e_1\ v_1\ bb).	          % (* a→a, a ⊢ 1 : a→a *)
+(t_5 = app\ v_1\ v0 : term). \\&
+(g_4 = of_a\ e_1\ v_1\ v_0\ base\ base\ g_3\ d_1 : of\ e1\ t5\ base). % (* ... ⊢ 1 0 : a *)
+(t_6 = lam\ t_5 : term). \\&
+(g_5 = of_l\ e_0\ t_5\ base\ base\ g_4 : of\ e_0\ t_6\ bb).  %  (* a→a ⊢ λ(1 0) : a→a *)
+(t_7 = lam\ t_6 : term).
+(g_6 = of_l\ nil\ t_6\ bb\ bb\ g_5 : of\ nil\ t_7\ bbb). *	 %  (* ⊢ λλ(1 0) : (a→a)→a→a  *)
+\end{align}
+}
 

### 2010-06-18 papers/MIPS2010/abstract.tex
* SML bib.
git-svn-id: http://gasp.dev.regis-gianas.com/svn/trunk@111 403e655e-980a-47bd-8c60-1ee5348bb6b7

@@
 {\cite{pfenning1999system}}), and substantial formalizations have been
-undertaken \remtext{TODO citer, mais quoi?}, we still use legacy tools
+undertaken~[]\cite{lee07}, we still use legacy tools
 based on text representation to manage our developments. The general

### 2010-06-18 papers/MIPS2010/abstract.tex
déplacement de la partie sur l'ui.
git-svn-id: http://gasp.dev.regis-gianas.com/svn/trunk@112 403e655e-980a-47bd-8c60-1ee5348bb6b7

@@
 
-\title{Towards Semantic Patches}
-  % pas sûr que le nom soit encore adapté
+\title{Towards typed repositories of proofs}
 
-\author{Matthias Puech, Yann Régis-Gianas}
+\author{Matthias Puech \and Yann Régis-Gianas}
 \date{}
@@
 tool to analyze the impact of change through types. It involves at its
-core a typed description language for repositories -- described in
-\S\ref{kernel} -- driven by a user interface that we sketch in
-\S\ref{devel}. In the first iteration of this project, we focus on a
+core a typed description language for repositories, described in
+\S\ref{kernel}. In the first iteration of this project, we focus on a
 static, or data-driven model for repositories; in \S\ref{rw}, we
@@
 
-Our system will have these properties \emph{wrt.} LF:
+Our system will have these properties w.r.t. LF:
 \begin{itemize}
@@
 (g_5 = of_l\ e_0\ t_5\ base\ base\ g_4 : of\ e_0\ t_6\ bb).  %  (* a→a ⊢ λ(1 0) : a→a *)
-(t_7 = lam\ t_6 : term).
+(t_7 = lam\ t_6 : term). \\&
 (g_6 = of_l\ nil\ t_6\ bb\ bb\ g_5 : of\ nil\ t_7\ bbb). *	 %  (* ⊢ λλ(1 0) : (a→a)→a→a  *)
@@
 }
-
-\section{Development through semantic patches}
-\label{devel}
-
-\remplan{Fri Jun 18, 2010 10:50 AM. L'objectif ici, c'est de décrire
-  une (possible) architecture pour rendre utilisable le langage 
-  décrit dans la section précédente. En gros, on énumère les 
-  opérations qui interviennent dans l'utilisation et on les décrit
-  informellement à l'aide d'un exemple.}
-
-The kernel language described in the previous section is clearly not
-adapted to human interaction, for several reasons~: 
-\begin{itemize}
-
-\item Every single object is flattened, which makes unobvious the
-  structure of a program or a proof.
-
-\item A term of this language denotes the entire history of a
-  development. Most of the time, the user is more interested in a
-  instantaneous snapshot of his object of study.
-
-\item Even though we were able to provide a snapshot of a particular
-  state of the development, the repository represents this state 
-  using proof terms, which contain too many syntactic details from the
-  user point of view and are not suitable for edition.
-
-\end{itemize}
-
-\remtext{Fri Jun 18, 2010 10:52 AM. Par contre, je ne sais pas quel
-  exemple utilisé. Le STLC? Est-ce que ça fait ``assez'' système
-  logique pour ce workshop?}
-
-\subsection{A layered architecture}
-
-\remtext{Fri Jun 18, 2010  1:32 PM. Ici, on décrit la figure en suivant
-  le modèle de flot du système. On décrit le rôle des fonctions, 
-les problèmes posés.}
-
-% À mettre dans une figure. 
-\begin{figure}
-\begin{center}
-% FIXME: il y a un problème avec la flèche double phi/psi.
-\xymatrix{
-\textrm{History of metatheory objects}
-&R 
-\ar@{->}[d]_{\phi}
-& \Delta \ar@{->}[r]^{\theta(R)} & R' \\
-\textrm{Metatheory objects} 
-&t \ar@{->}[d]^{\pi}
-\ar@{->}[u]_{\psi}
-& t' \ar@{->}[u]_{\nabla(t)}\\
-\textrm{Concrete views on metatheory objects} 
-&w \ar@{.>}[r] & w' \ar@{->}[u]_{\sigma(t, w)}
-}
-\end{center}
-\label{fig:architecture}
-\caption{A layered architecture for a semantic patch framework.}
-\end{figure}
-
-With the previous remarks in mind, we designed a generic framework of semantic
-patches that can be described using three layers which are summarized in 
-Figure~\ref{fig:architecture}. We now describe the role of each layer.
-
-The most internal layer focused on the history of metatheory
-objects. It is composed of a type-checker for the kernel language. The
-purpose of this layer is to maintain a repository~$R$ and the
-invariant that it is well-typed. Through the layer interface, a client
-can ask for the \textit{integration} $\theta(R)$ of a patch $\Delta$
-to obtain an extended version $R'$ of the repository $R$.
-
-The middle layer is a mechanized formalization of a metatheory. This
-module provides a syntax for the objects of the metatheory (formulae,
-judgments, \ldots) as well as the admissible rules (that we also call
-metatheorems).  An \textit{internalization} function $\phi$ translates
-the constructors of these objects and the specification of the
-admissible rules as terms of the kernel language. As a consequence,
-every object~$t$ of the theory can be injected into the
-repository~$R$. An \textit{externalization} function $\psi$ is
-implementing the inverse operation. Given a modified version $t'$ of
-an object~$t$, we are able to compute the patch $\Delta$ using
-the \textit{differentiation} function $\nabla(t)$.
-
-The external layer offers views on the objects of the theory. A
-view~$w$ is a human readable, and editable, partial representation of
-an object~$t$. A view is obtained from an object through a
-projection~$\pi$.  The user transforms the view $w$ into a view $w'$
-which must be $\textit{put back}$ into the metatheory as a modified
-object $t'$ using a $\textit{putback}$ function $\sigma(t, w)$.  In
-concrete syntax, an object~$t$ may be denoted simply using its
-syntactic definition but also, more interestingly, by a reference to this 
-object into the repository.
-
-The implementation of this architecture raises some interesting
-technical difficulties that can be handled, but only partially, 
-by existing technology. Indeed, even if internalization and 
-externalization are easily implemented thanks to the expressivity
-of the kernel language, the projection, putback, differentiation
-and integration operations are more intricate to realize.
-
-\paragraph{From an abstract proof tree to a concrete view, and back}
-\-
-\remtext{Fri Jun 18, 2010  1:36 PM. Un arbre de preuves, ce n'est
-  pas très pratique. Pourtant, c'est moralement sur cet objet 
-  que s'opère le développement et non sur le code source, qui
-  n'en est qu'une représentation. On justifie la nécessité de
-  projection et on fait référence aux lentilles de Pierce pour
-  expliquer $\sigma$ et $\pi$.}
-
-The problem of updating some data according to a modification that was
-done on a partial view of it is a well-known issue in database
-literature called the \textit{view-update} problem. Pierce \textit{et
-  al} have recently attacked it from a language design approach. Their
-idea is to develop projection and putback functions simultaneously by
-writing a so-called lens in a bidirectional programming
-language~\cite{pierce-bidirectional}. Roughly speaking, a lens is a
-software component that can be used to produce output from input but
-also input from output (hence the bidirectionality). A lens is able
-to keep track of the information that was erased by a projection to
-produce a projected output from the initial input. When we supply a
-modified output to the lens, it can elaborate, from the initial input
-and under certain conditions, the missing part of the modified input.
-
-We plan to design a bidirectional language to define the
-projection~function~$\pi$ and the putback~function~$\sigma$ as a
-lens. Yet, we cannot just import existing lenses combinators from the
-Harmony system that already works on trees. Indeed, most of the time,
-our trees have extra structural constraints induced by the presence of
-binders.
-
-\remtext{Fri Jun 18, 2010 1:46 PM. Autre point important sur les vues
-  concrètes~: elles peuvent servir à parler de l'historique.  Voir le
-  module MetaAST de l'implémentation. Ces vues sont utiles par exemple
-  lorsque l'utilisateur doit construire un patch d'adaptation. Il a
-  alors toute la puissance des métathéorèmes ainsi que tous les objets
-  déjà construits pour décrire son patch.}
-
-\paragraph{Capturing the evolution of abstract proof trees}
-\-
-
-Computing the minimal edition script between two abstract syntax trees
-is a NP-hard problem that has been heavily studied, in particular in
-bio-informatics.  We guess that, in practice, we can restrict the
-search to local transformations that preserve substructures (that
-roughly consist in inserting, removing or moving around entire
-sub-trees). In that case, an existing polynomial algorithm may be
-used~\cite{Touzet05alinear}. Besides, abstract trees are relatively small in
-practice.
-
-We are more concerned about the fact that existing algorithms are not
-aware of binders that occur in proof-terms. Indeed, some standard
-edition scripts must be forbidden because they could break
-well-scopedness. As a consequence, we will have to adapt standard
-algorithms to only look for scope-preserving edition scripts.
-
-%% We are more concerned about our ability to elaborate high-level patches
-%% from these low-level edition scripts. Actually, it is easy to deduce
-%% from a low-level edition script, a low-level patch, based on object
-%% syntax constructor elimination and introduction. However, such a
-%% low-level patch may forbid a more efficient reuse of existing
-%% objects. For instance, let us assume a transformation of an
-%% object-level term~$x + x$ into~ he term~$2 + 2$. A low-level patch
-%% would represents the transformation as the replacement of the first
-%% sub-term $x$ by the sub-term~$2$ followed by the replacement of the
-%% second sub-term $x$ by the sub-term~$2$. If we had an object-level type
-%% derivation for $x + x$, the integration function would have to build
-%% two sub-derivations, one for each replaced sub-terms.  Instead, a
-%% high-level patch would represent that transformation as a substitution
-%% of the free occurrences of~$x$ by the term $2$ in the term~$x + x$. In
-%% that case, the integration function has the opportunity to apply a
-%% metatheorem about the stability of typing derivations under
-%% substitution and finally reuse the initial typing derivation
-%% directly. Inferring automatically such a high-level patch seems an
-%% unreachable objective~: there is clearly no most general high-level
-%% patch. Nonetheless, an interactive process can be designed to help the
-%% user building his high-level patch from a set of pre-computed
-%% interpretations of the low-level patch.
-
-\remtext{Fri Jun 18, 2010 1:38 PM. Il faut différencier deux arbres de
-  syntaxe. Une méthode : la distance d'édition syntaxique permet de
-  déduire des applications des règles d'élimination et d'introduction
-  que l'on peut utiliser pour construire $\Delta$. Référence sur la
-  littérature sur les distances d'édition entre arbres. Discussion sur
-  l'élaboration de patchs de haut niveau (comme le renommage etc). }
-
-\paragraph{Propagating changes through dependencies}
-\-
-
-There is several ways of integrating a patch $\Delta$ inside a
-repository~$R$.  The naive attempt would be a simple concatenation
-of~$\Delta$ at the end of~$R$.  This kind of integration, though
-possible and correct, is of little interest because it does not take
-into account the fact that $\Delta$ probably defines new versions of
-objects already present in~$R$.
-
-A more useful integration propagates changes to the context that uses
-the modified object in order to trigger the updating of this context
-(which may trigger the updating of other tree, and so forth). Again,
-in full generality, it is not possible to update the context
-automatically. Instead, we choose to generate \textit{challenges},
-which are adaptation patches with holes. These challenges can be
-addressed by the author that wrote the patch to be integrated, by the
-author of the context or automatically when there is no ambiguity
-(think of a context that uses a function whose body has changed but
-not its type).
-
-\remtext{Fri Jun 18, 2010 1:42 PM. Si on se contente de concaténer
-  $\Delta$ à $R$, on n'a intégré qu'une nouvelle version de l'objet
-  initial~$t$ dans $R$ mais on n'a pas propagé ces modifications
-  auprès des clients de $t$. Cette propagation n'est \textit{a priori}
-  pas triviale~: rien ne dit que $t'$ à un sens équivalent à $t$. Il
-  faut donc adapter les clients en demandant un patch d'adaptation à
-  l'utilisateur. On présente ces demandes sous la forme de patch à
-  compléter (on fournit un peu de contexte et on a un trou pour y
-  injecter une nouvelle sous-dérivation mais on peut aussi modifier le
-  contexte pour l'adapter). La construction d'un patch d'adaptation
-  est un processus interactif~: on ne sait pas a priori borner le
-  contexte nécessaire à la construction du patchs d'adaptation donc on
-  peut imaginer un sorte de jeu où le système fait augmenter le
-  contexte (en suivant les dépendances) si l'utilisateur ne sait pas
-  construire le patch d'adaptation. En passant, quand on parle
-  d'utilisateur ici, ce peut être aussi bien un programme, comme un
-  type-checker pour le langage objet, ou bien un programmeur.}
-
-\subsection{Application on simply typed $\lambda$-calculus}
 
@@
 
-%% We propose here to devise a system for \emph{semantic patches}. It
+%% We propose here to devise a system for \emph{semantic repositories}. It
 %% % TODO changer le nom?
@@
 
-%% We want to design a \emph{language of patches}, able to represent
+%% We want to design a \emph{language of repositories}, able to represent
 %% % TODO changer le nom?

### 2010-06-18 papers/MIPS2010/abstract.tex
On allège encore: abstract court sans technique
git-svn-id: http://gasp.dev.regis-gianas.com/svn/trunk@113 403e655e-980a-47bd-8c60-1ee5348bb6b7

@@
 
-%\renewcommand{\remplan}[1]{}
-%\renewcommand{\remtext}[1]{}
+\renewcommand{\remplan}[1]{}
+\renewcommand{\remtext}[1]{}
 
@@
 
-\section{Motivations}
+\paragraph{Motivations}
 
@@
 modules, file-based scripts and separate compilation in proof
-assistants like \textsf{Coq}~{\cite{CoqDocWeb}} or
+assistants like \textsf{Coq} ({\cite{CoqDocWeb}}) or
 \textsf{Matita}~{\cite{Asperti06userinteraction}}, the use of
@@
 
-We will describe here some of the possible directions to develop a
-tool to analyze the impact of change through types. It involves at its
-core a typed description language for repositories, described in
-\S\ref{kernel}. In the first iteration of this project, we focus on a
-static, or data-driven model for repositories; in \S\ref{rw}, we
-quickly review related approaches, and hint a possible generalization
-in \S\ref{fw}.
+We will describe some of the possible directions to develop a tool to
+analyze the impact of change through types. It involves at its core a
+typed description language for repositories, and is strongly related
+to incremental type-checking: only differences between versions are
+type-checked and not the entire development. In the first iteration of
+this project, we focus on a static, or data-driven model for
+repositories inspired by the repository model of \textsf{GIT}.
 
-\section{A core language to describe typed repositories}
-\label{kernel}
+\paragraph{A core language to describe typed repositories}
 
@@
 subderivations.
-
-\subsection{How to describe a development?}
 
@@
 
-\subsection{Formalization}
-
 We describe the state of a repository at a given moment by a type in
@@
 language.
-
-\paragraph{Syntax} There are three syntactic categories: applicative
-terms $a$, types $t$ and environments $\Gamma$.
-
 \begin{align*}
@@
 
-Following the standard syntax, we denote by $(x:t)\cdot u$ the
-dependent product (elsewhere written with a $\Pi$). The new construct
-$(x=a:t)\cdot u$ is a definition binder, whose value is $a$. We write
-as usual $(x_1:t_1)(x_2:t_2)\ldots(x_n:t_n)\cdot t$ for
-$(x_1:t_1)\cdot(x_2:t_2)\cdot\ldots\cdot(x_n:t_n)\cdot t$, $t \to u$
-for $(x:t)\cdot u$ where $x\notin u$, and $(x\ y : t)\cdot u$ for
-$(x:t)(y:t)\cdot u$. The lookup in the environment is written either
-$\Gamma(x):t$, denoting ``$\Gamma$ contains a declaration $[x:t]$
-\emph{or} a definition $[x=a:t]$'', or $\Gamma(x)=a:t$ denoting
-``$\Gamma$ contains a definition $[x=a:t]$''. Capture-avoiding
-renaming of variables in a term $t$ is written $t\subst{x}{y}$ for
-``replace all occurences of $x$ by $y$'' and is defined the usual way.
+% \label{kernel}
 
-\paragraph{Typing} Typing is parameterized by:
-\begin{itemize}
-  \item$\mathcal S$ a set of sorts \sort,
-  \item$\mathcal A \in \mathcal S^2$ the axioms of the system,
-  \item$\mathcal R \in \mathcal S^3$ the allowed products.
-\end{itemize}
-It relies on two distinct judgments:
-\begin{itemize}
-\item $\Gamma\vdash a : t$ ($a$ has type $t$ in $\Gamma$),
-\item $\Gamma\vdash t : \sort$ (the repository $t$ is well-typed of
-  sort $s$ in $\Gamma$).
-\end{itemize}
-The rules are:
-\begin{mathpar}
-  \infer[Axiom]{
-  }{\Gamma\vdash\sort_1:\sort_2}
-  \quad(\sort_1,\sort_2)\in\mathcal A
-  \and
-  \infer[Init]{
-    \Gamma(x):t
-  }{\Gamma\vdash x:t}
-  \and
-  \infer[Prod]{\Gamma\vdash t:\sort_1 \and 
-    \Gamma[x:t]\vdash u:\sort_2
-  }{\Gamma\vdash(x:t)\cdot u : \sort_3}
-  \quad(\sort_1,\sort_2,\sort_3)\in\mathcal R
-  \and
-  \infer[Let1]{\Gamma\vdash a:t \and
-    \Gamma\vdash t:\sort_1 \and
-    \Gamma[x=a:t]\vdash u:\sort_2
-  }{\Gamma\vdash(x=a:t)\cdot u : \sort_3} 
-  \quad (x=a:t)\notin\Gamma
-  \and
-  \infer[Let2]{\Gamma\vdash t:\sort_1 \and
-    \Gamma\vdash u:\sort_2
-  }{\Gamma\vdash(x=a:t)\cdot u : \sort_3} 
-  \quad(x=a:t)\in\Gamma
-  \and
-  \infer[App]{
-    \Gamma\vdash a:(y:t)\cdot u \and \Gamma(x):t
-    }{\Gamma\vdash a\ x : u\subst{x}{y}}
-  \and
-  \infer[Skip1]{
-    \Gamma\vdash a:(y=b:t)\cdot u \and \Gamma(x)=b:t
-    }{\Gamma\vdash a : u\subst{x}{y}} \and
-  \infer[Skip2]{
-    \Gamma[x=b:t]\vdash a:(y=b:t)\cdot u
-    }{\Gamma\vdash a : u\subst{x}{y}}
-\end{mathpar}
+% The kernel of our system is a type-checker algorithm for a typed
+% meta-language. In this language, we will declare both the syntax of
+% the object (proof-)language and its typing rules, and define pieces of
+% syntax (our proofs) and their derivations. Describing transformations
+% among syntax objects is done by sharing common subterms or
+% subderivations.
 
-The first four rules are standard. As application is only performed on
-variables, the \textsc{App} rule looks up directly in the
-environment. Rules \textsc{Let1} and \textsc{Skip1} show the
-verification and the use of the equality binder in case the definition
-is not already known. Rule \textsc{Let2} and \textsc{Skip2} allow
-incremental typing in our system: if the definition was already known
-in the environment, we do not need to recheck it. If $\cdot\vdash
-t:\sort$, then the repository $t$ is well-typed.
+% \subsection{How to describe a development?}
 
-\paragraph{Example}
+% Representing syntax and logics is nicely done in a \emph{logical
+%   framework} like LF (\cite{harper1993framework}): both the syntactic
+% elements and the typing derivations can sit in the same tree
+% structure, and both can be rechecked at the same time, thanks to
+% dependent types. For the purpose of incremental type checking though,
+% our needs are a little bit different: first, we need to record, that
+% is to name all intermediate values of our developments, so as to be
+% able to address and reuse them multiple times. Secondly, we need to
+% make sure that those intermediate values (sub-terms) are not recorded
+% twice, so as to not type-check them twice: we are looking to represent
+% syntactic and typing objects as a DAG rather than a tree.
 
-Here is a well-typed repository describing the simply-typed
-$\lambda$-calculus and constructing two $\lambda$-terms sharing some
-sub-derivations: 
-{\small
-\begin{align}&
-(\nat : *).
-(o : \nat).
-(s : \nat \to \nat).
-(type : *).
-(base : type).
-(arr : type \to type \to type). \\&
-(env : *).
-(nil : env).
-(cons : type \to env \to env). \\&
-(term : *).
-(var : \nat \to term).
-(lam : term \to term).
-(app : term \to term \to term). \\&
-(lookup : env \to \nat \to type \to *).
-(lookup_o : (e : env).(a : type).(f = cons\ a\ e : env). lookup\ e\ o\ a). \\&
-(lookup_s : (e : env).(a b : type).(n : \nat).(sn = s\ n : \nat).(f =
-cons\ a\ e : env). lookup\ e\ n\ b \to lookup\ f\ sn\ b). \\&
-(of : env \to term \to type \to *). \\&
-(of_v : (e : env).(a : type).(n : \nat).(v = var\ n : term).
-          lookup\ e\ n\ a \to of\ e\ v\ a). \\&
-(of_l : (e : env).(t : term).(a\ b: type).(c = arr\ a\ b : type).
-          (u = lam\ t : term).(f = cons\ a\ e : env).
-          of\ f\ t\ b \to of\ e\ u\ c). \\&
-(of_a : (e : env).(t\ u: term).(a\ b: type).
-          (c = arr\ a\ b : type).(v = app\ t\ u : term).
-          of\ e\ t\ c \to of\ e\ u\ a \to of\ e\ v\ b). \\&
-(1 = s\ o : \nat).(v_0 = var\ o : term).(v_1 = var\ 1 : term).
-(t_3 = lam\ v_0 : term).(ch_0 = lam\ t_3 : term).\\&
-(t_1 = app\ v_0\ v_1 : term).(t_2 = lam\ t_1 : term).
-(t_4 = lam\ t_2 : term).
-(bb = arr\ base\ base : type). \\&
-(e_0 = cons\ bb\ nil : env).
-(e_1 = cons\ base\ e0 : env).
-(d_0 = lookup_o\ e_1\ base : lookup\ e_1\ o\ base). \\& %        (* a→a, a ⊢ 0 : a *)
-(d_1 = of_v\ e_1\ base\ o\ d0 : of\ e_1\ v_0\ base). %	   (* a→a, a ⊢ 0 : a *)
-(d_2 = of_l\ e_0\ v_0\ base\ base d_1 : of\ e_0\ t_3\ bb). \\& %   (* a→a ⊢ λ0 : a → a *)
-(bbb = arr\ bb\ bb : type).
-(d_3 = of_l\ nil\ t_3\ bb\ bb\ d_2 : of\ nil\ ch_0\ bbb). \\& %  (* ⊢ λλ0 : (a→a)→(a→a) *)
-(g_1 = lookup_o\ e_0\ bb : lookup\ e_0\ o\ bb).
-(g_2 = lookup_s\ e_0\ base\ bb\ o\ g_1 : lookup\ e_1\ 1\ bb). \\&
-(g_3 = of_v\ e_1\ bb\ 1\ g_2 : of\ e_1\ v_1\ bb).	          % (* a→a, a ⊢ 1 : a→a *)
-(t_5 = app\ v_1\ v0 : term). \\&
-(g_4 = of_a\ e_1\ v_1\ v_0\ base\ base\ g_3\ d_1 : of\ e1\ t5\ base). % (* ... ⊢ 1 0 : a *)
-(t_6 = lam\ t_5 : term). \\&
-(g_5 = of_l\ e_0\ t_5\ base\ base\ g_4 : of\ e_0\ t_6\ bb).  %  (* a→a ⊢ λ(1 0) : a→a *)
-(t_7 = lam\ t_6 : term). \\&
-(g_6 = of_l\ nil\ t_6\ bb\ bb\ g_5 : of\ nil\ t_7\ bbb). *	 %  (* ⊢ λλ(1 0) : (a→a)→a→a  *)
-\end{align}
-}
+% Our system will have these properties w.r.t. LF:
+% \begin{itemize}
+% \item In this first iteration of the project, we do not need
+%   computations to take place within our DAGs. Our syntax will then be
+%   restricted to product types and applications;
+% \item Every term should be a flat application of variables, so that we
+%   don't introduce compound terms without naming them;
+% \item Finally we need a way to record intermediate definitions: we
+%   introduce a new kind of binder, the equality binder. We maintain the
+%   invariant that not two same definition can sit in the context while
+%   typing.
+% \end{itemize}
 
-\section{Related work}
+% \subsection{Formalization}
+
+% We describe the state of a repository at a given moment by a type in
+% the following syntax. Well-typed types in the repository meta-language
+% guarantee that it contains only well-typed proofs in the object
+% language.
+
+% \paragraph{Syntax} There are three syntactic categories: applicative
+% terms $a$, types $t$ and environments $\Gamma$.
+
+% \begin{align*}
+%  \gdecl{a}{x \gor a\ x } \\
+%  \gdecl{t}{a \gor \sort \gor (x:t)\cdot t \gor (x=a:t)\cdot t} \\
+%  \gdecl{\Gamma}{\cdot \gor \Gamma[x:t] \gor \Gamma[x=a:t]}
+% \end{align*}
+
+% Following the standard syntax, we denote by $(x:t)\cdot u$ the
+% dependent product (elsewhere written with a $\Pi$). The new construct
+% $(x=a:t)\cdot u$ is a definition binder, whose value is $a$. We write
+% as usual $(x_1:t_1)(x_2:t_2)\ldots(x_n:t_n)\cdot t$ for
+% $(x_1:t_1)\cdot(x_2:t_2)\cdot\ldots\cdot(x_n:t_n)\cdot t$, $t \to u$
+% for $(x:t)\cdot u$ where $x\notin u$, and $(x\ y : t)\cdot u$ for
+% $(x:t)(y:t)\cdot u$. The lookup in the environment is written either
+% $\Gamma(x):t$, denoting ``$\Gamma$ contains a declaration $[x:t]$
+% \emph{or} a definition $[x=a:t]$'', or $\Gamma(x)=a:t$ denoting
+% ``$\Gamma$ contains a definition $[x=a:t]$''. Capture-avoiding
+% renaming of variables in a term $t$ is written $t\subst{x}{y}$ for
+% ``replace all occurences of $x$ by $y$'' and is defined the usual way.
+
+% \paragraph{Typing} Typing is parameterized by:
+% \begin{itemize}
+%   \item$\mathcal S$ a set of sorts \sort,
+%   \item$\mathcal A \in \mathcal S^2$ the axioms of the system,
+%   \item$\mathcal R \in \mathcal S^3$ the allowed products.
+% \end{itemize}
+% It relies on two distinct judgments:
+% \begin{itemize}
+% \item $\Gamma\vdash a : t$ ($a$ has type $t$ in $\Gamma$),
+% \item $\Gamma\vdash t : \sort$ (the repository $t$ is well-typed of
+%   sort $s$ in $\Gamma$).
+% \end{itemize}
+% The rules are:
+% \begin{mathpar}
+%   \infer[Axiom]{
+%   }{\Gamma\vdash\sort_1:\sort_2}
+%   \quad(\sort_1,\sort_2)\in\mathcal A
+%   \and
+%   \infer[Init]{
+%     \Gamma(x):t
+%   }{\Gamma\vdash x:t}
+%   \and
+%   \infer[Prod]{\Gamma\vdash t:\sort_1 \and 
+%     \Gamma[x:t]\vdash u:\sort_2
+%   }{\Gamma\vdash(x:t)\cdot u : \sort_3}
+%   \quad(\sort_1,\sort_2,\sort_3)\in\mathcal R
+%   \and
+%   \infer[Let1]{\Gamma\vdash a:t \and
+%     \Gamma\vdash t:\sort_1 \and
+%     \Gamma[x=a:t]\vdash u:\sort_2
+%   }{\Gamma\vdash(x=a:t)\cdot u : \sort_3} 
+%   \quad (x=a:t)\notin\Gamma
+%   \and
+%   \infer[Let2]{\Gamma\vdash t:\sort_1 \and
+%     \Gamma\vdash u:\sort_2
+%   }{\Gamma\vdash(x=a:t)\cdot u : \sort_3} 
+%   \quad(x=a:t)\in\Gamma
+%   \and
+%   \infer[App]{
+%     \Gamma\vdash a:(y:t)\cdot u \and \Gamma(x):t
+%     }{\Gamma\vdash a\ x : u\subst{x}{y}}
+%   \and
+%   \infer[Skip1]{
+%     \Gamma\vdash a:(y=b:t)\cdot u \and \Gamma(x)=b:t
+%     }{\Gamma\vdash a : u\subst{x}{y}} \and
+%   \infer[Skip2]{
+%     \Gamma[x=b:t]\vdash a:(y=b:t)\cdot u
+%     }{\Gamma\vdash a : u\subst{x}{y}}
+% \end{mathpar}
+
+% The first four rules are standard. As application is only performed on
+% variables, the \textsc{App} rule looks up directly in the
+% environment. Rules \textsc{Let1} and \textsc{Skip1} show the
+% verification and the use of the equality binder in case the definition
+% is not already known. Rule \textsc{Let2} and \textsc{Skip2} allow
+% incremental typing in our system: if the definition was already known
+% in the environment, we do not need to recheck it. If $\cdot\vdash
+% t:\sort$, then the repository $t$ is well-typed.
+
+% \paragraph{Example}
+
+% Here is a well-typed repository describing the simply-typed
+% $\lambda$-calculus and constructing two $\lambda$-terms sharing some
+% sub-derivations: 
+% {\small
+% \begin{align}&
+% (\nat : *).
+% (o : \nat).
+% (s : \nat \to \nat).
+% (type : *).
+% (base : type).
+% (arr : type \to type \to type). \\&
+% (env : *).
+% (nil : env).
+% (cons : type \to env \to env). \\&
+% (term : *).
+% (var : \nat \to term).
+% (lam : term \to term).
+% (app : term \to term \to term). \\&
+% (lookup : env \to \nat \to type \to *).
+% (lookup_o : (e : env).(a : type).(f = cons\ a\ e : env). lookup\ e\ o\ a). \\&
+% (lookup_s : (e : env).(a b : type).(n : \nat).(sn = s\ n : \nat).(f =
+% cons\ a\ e : env). lookup\ e\ n\ b \to lookup\ f\ sn\ b). \\&
+% (of : env \to term \to type \to *). \\&
+% (of_v : (e : env).(a : type).(n : \nat).(v = var\ n : term).
+%           lookup\ e\ n\ a \to of\ e\ v\ a). \\&
+% (of_l : (e : env).(t : term).(a\ b: type).(c = arr\ a\ b : type).
+%           (u = lam\ t : term).(f = cons\ a\ e : env).
+%           of\ f\ t\ b \to of\ e\ u\ c). \\&
+% (of_a : (e : env).(t\ u: term).(a\ b: type).
+%           (c = arr\ a\ b : type).(v = app\ t\ u : term).
+%           of\ e\ t\ c \to of\ e\ u\ a \to of\ e\ v\ b). \\&
+% (1 = s\ o : \nat).(v_0 = var\ o : term).(v_1 = var\ 1 : term).
+% (t_3 = lam\ v_0 : term).(ch_0 = lam\ t_3 : term).\\&
+% (t_1 = app\ v_0\ v_1 : term).(t_2 = lam\ t_1 : term).
+% (t_4 = lam\ t_2 : term).
+% (bb = arr\ base\ base : type). \\&
+% (e_0 = cons\ bb\ nil : env).
+% (e_1 = cons\ base\ e0 : env).
+% (d_0 = lookup_o\ e_1\ base : lookup\ e_1\ o\ base). \\& %        (* a→a, a ⊢ 0 : a *)
+% (d_1 = of_v\ e_1\ base\ o\ d0 : of\ e_1\ v_0\ base). %	   (* a→a, a ⊢ 0 : a *)
+% (d_2 = of_l\ e_0\ v_0\ base\ base d_1 : of\ e_0\ t_3\ bb). \\& %   (* a→a ⊢ λ0 : a → a *)
+% (bbb = arr\ bb\ bb : type).
+% (d_3 = of_l\ nil\ t_3\ bb\ bb\ d_2 : of\ nil\ ch_0\ bbb). \\& %  (* ⊢ λλ0 : (a→a)→(a→a) *)
+% (g_1 = lookup_o\ e_0\ bb : lookup\ e_0\ o\ bb).
+% (g_2 = lookup_s\ e_0\ base\ bb\ o\ g_1 : lookup\ e_1\ 1\ bb). \\&
+% (g_3 = of_v\ e_1\ bb\ 1\ g_2 : of\ e_1\ v_1\ bb).	          % (* a→a, a ⊢ 1 : a→a *)
+% (t_5 = app\ v_1\ v0 : term). \\&
+% (g_4 = of_a\ e_1\ v_1\ v_0\ base\ base\ g_3\ d_1 : of\ e1\ t5\ base). % (* ... ⊢ 1 0 : a *)
+% (t_6 = lam\ t_5 : term). \\&
+% (g_5 = of_l\ e_0\ t_5\ base\ base\ g_4 : of\ e_0\ t_6\ bb).  %  (* a→a ⊢ λ(1 0) : a→a *)
+% (t_7 = lam\ t_6 : term). \\&
+% (g_6 = of_l\ nil\ t_6\ bb\ bb\ g_5 : of\ nil\ t_7\ bbb). *	 %  (* ⊢ λλ(1 0) : (a→a)→a→a  *)
+% \end{align}
+% }
+
+\paragraph{Related work}
 \label{rw}
@@
 
-\section{Further work}
-\label{fw}
+% \section{Further work}
+% \label{fw}
 
-\remtext{Extension vers une version constructiviste?}
+% \remtext{Extension vers une version constructiviste?}
 
-\remtext{Parler du bootstraping?}
+% \remtext{Parler du bootstraping?}
 
-\section{Conclusion}
+% \section{Conclusion}
 

### 2010-06-18 papers/MIPS2010/abstract.tex
...
git-svn-id: http://gasp.dev.regis-gianas.com/svn/trunk@114 403e655e-980a-47bd-8c60-1ee5348bb6b7

@@
 Representing syntax and logics is nicely done in a \emph{logical
-  framework} like LF (\cite{harper1993framework}): both the syntactic
+  framework} like LF: both the syntactic
 elements and the typing derivations can sit in the same tree

### 2010-06-18 papers/MIPS2010/abstract.tex
citations ok
git-svn-id: http://gasp.dev.regis-gianas.com/svn/trunk@115 403e655e-980a-47bd-8c60-1ee5348bb6b7

@@
 assistants like \textsf{Coq} ({\cite{CoqDocWeb}}) or
-\textsf{Matita}~{\cite{Asperti06userinteraction}}, the use of
+\textsf{Matita}~({\cite{AspertiCTZ07}}), the use of
 dependency management tools (\textsf{Make}) or version control system
@@
 been shown to be of crucial to the mathematical discovery
-process~\cite{lakatos1964proofs}, and we believe that they should be
+process~(\cite{lakatos1964proofs}), and we believe that they should be
 taken into account in the realization of mathematical assistants.
@@
 {\cite{pfenning1999system}}), and substantial formalizations have been
-undertaken~[]\cite{lee07}, we still use legacy tools
+undertaken~(\cite{lee07}), we still use legacy tools
 based on text representation to manage our developments. The general
@@
 with in various proof assistant and mathematical repositories. The
-HELM project (\cite{asperti2000content}, \cite{asperti2006content})
+HELM project (\cite{asperti2006content})
 was an attempt to create a large library of mathematics, importing
@@
 Most ideas from this project were imported into the \textsf{Matita}
-proof assistants (\cite{asperti2007hop}), especially a mechanism of
+proof assistants (\cite{AspertiCTZ07}), especially a mechanism of
 \emph{invalidation and regeneration} to ensure the global consistency

### 2010-06-18 papers/MIPS2010/abstract.tex
vire les commentaires, on les retrouvera dans l'histo
git-svn-id: http://gasp.dev.regis-gianas.com/svn/trunk@116 403e655e-980a-47bd-8c60-1ee5348bb6b7
* Modif mineure dans intro.
git-svn-id: http://gasp.dev.regis-gianas.com/svn/trunk@117 403e655e-980a-47bd-8c60-1ee5348bb6b7

@@
 modifications imply on the validity of the whole edifice or, the other
-way around, how do one rely on existing work to build up new results?
+way around, how does one rely on existing work to build up new results?
 Many of these questions remain largely unanswered, but the tendency
@@
 modules, file-based scripts and separate compilation in proof
-assistants like \textsf{Coq} ({\cite{CoqDocWeb}}) or
-\textsf{Matita}~({\cite{AspertiCTZ07}}), the use of
+assistants like \textsf{Coq}~\cite{CoqDocWeb} or
+\textsf{Matita}~\cite{AspertiCTZ07}, the use of
 dependency management tools (\textsf{Make}) or version control system
@@
 {\cite{pfenning1999system}}), and substantial formalizations have been
-undertaken~(\cite{lee07}), we still use legacy tools
-based on text representation to manage our developments. The general
-goal exposed here is to replace this tool chain and make it
-language-aware, both on a syntactic side (AST) and on a semantic side
-(typing).
+undertaken\cite{lee07}, we still use legacy tools based on text
+representation to manage our developments. The general goal exposed
+here is to replace this tool chain and make it language-aware, both on
+a syntactic side through the use of abstract syntax trees instead of
+concrete syntax and on a semantic side by using typing to ensure 
+repository consistency.
 
-\remplan{Sujet du papier : on focalise sur un problème bien précis}
+% \remplan{Sujet du papier : on focalise sur un problème bien précis}
 
-\remtext{Fri Jun 18, 2010 10:00 AM.
-  Un peu de liant. Pour arriver à notre but, il y a une
-  question importante à résoudre: si on te donne une méta-théorie
-  formalisée, es-tu capable de décrire l'historique d'un développement
-  dans cette théorie? Cela induit deux questions : dans quel langage
-  et est-ce que ce langage est utilisable (parce qu'a priori, ce n'est
-  pas gagné car le développeur travaille en mode ``instantané'' et ne
-  veut pas se trimballer ses casseroles tout le temps\ldots).}
+%% \remtext{Fri Jun 18, 2010 10:00 AM.
+%%   Un peu de liant. Pour arriver à notre but, il y a une
+%%   question importante à résoudre: si on te donne une méta-théorie
+%%   formalisée, es-tu capable de décrire l'historique d'un développement
+%%   dans cette théorie? Cela induit deux questions : dans quel langage
+%%   et est-ce que ce langage est utilisable (parce qu'a priori, ce n'est
+%%   pas gagné car le développeur travaille en mode ``instantané'' et ne
+%%   veut pas se trimballer ses casseroles tout le temps\ldots).}
 
@@
 \end{align*}
-
-% \label{kernel}
-
-% The kernel of our system is a type-checker algorithm for a typed
-% meta-language. In this language, we will declare both the syntax of
-% the object (proof-)language and its typing rules, and define pieces of
-% syntax (our proofs) and their derivations. Describing transformations
-% among syntax objects is done by sharing common subterms or
-% subderivations.
-
-% \subsection{How to describe a development?}
-
-% Representing syntax and logics is nicely done in a \emph{logical
-%   framework} like LF (\cite{harper1993framework}): both the syntactic
-% elements and the typing derivations can sit in the same tree
-% structure, and both can be rechecked at the same time, thanks to
-% dependent types. For the purpose of incremental type checking though,
-% our needs are a little bit different: first, we need to record, that
-% is to name all intermediate values of our developments, so as to be
-% able to address and reuse them multiple times. Secondly, we need to
-% make sure that those intermediate values (sub-terms) are not recorded
-% twice, so as to not type-check them twice: we are looking to represent
-% syntactic and typing objects as a DAG rather than a tree.
-
-% Our system will have these properties w.r.t. LF:
-% \begin{itemize}
-% \item In this first iteration of the project, we do not need
-%   computations to take place within our DAGs. Our syntax will then be
-%   restricted to product types and applications;
-% \item Every term should be a flat application of variables, so that we
-%   don't introduce compound terms without naming them;
-% \item Finally we need a way to record intermediate definitions: we
-%   introduce a new kind of binder, the equality binder. We maintain the
-%   invariant that not two same definition can sit in the context while
-%   typing.
-% \end{itemize}
-
-% \subsection{Formalization}
-
-% We describe the state of a repository at a given moment by a type in
-% the following syntax. Well-typed types in the repository meta-language
-% guarantee that it contains only well-typed proofs in the object
-% language.
-
-% \paragraph{Syntax} There are three syntactic categories: applicative
-% terms $a$, types $t$ and environments $\Gamma$.
-
-% \begin{align*}
-%  \gdecl{a}{x \gor a\ x } \\
-%  \gdecl{t}{a \gor \sort \gor (x:t)\cdot t \gor (x=a:t)\cdot t} \\
-%  \gdecl{\Gamma}{\cdot \gor \Gamma[x:t] \gor \Gamma[x=a:t]}
-% \end{align*}
-
-% Following the standard syntax, we denote by $(x:t)\cdot u$ the
-% dependent product (elsewhere written with a $\Pi$). The new construct
-% $(x=a:t)\cdot u$ is a definition binder, whose value is $a$. We write
-% as usual $(x_1:t_1)(x_2:t_2)\ldots(x_n:t_n)\cdot t$ for
-% $(x_1:t_1)\cdot(x_2:t_2)\cdot\ldots\cdot(x_n:t_n)\cdot t$, $t \to u$
-% for $(x:t)\cdot u$ where $x\notin u$, and $(x\ y : t)\cdot u$ for
-% $(x:t)(y:t)\cdot u$. The lookup in the environment is written either
-% $\Gamma(x):t$, denoting ``$\Gamma$ contains a declaration $[x:t]$
-% \emph{or} a definition $[x=a:t]$'', or $\Gamma(x)=a:t$ denoting
-% ``$\Gamma$ contains a definition $[x=a:t]$''. Capture-avoiding
-% renaming of variables in a term $t$ is written $t\subst{x}{y}$ for
-% ``replace all occurences of $x$ by $y$'' and is defined the usual way.
-
-% \paragraph{Typing} Typing is parameterized by:
-% \begin{itemize}
-%   \item$\mathcal S$ a set of sorts \sort,
-%   \item$\mathcal A \in \mathcal S^2$ the axioms of the system,
-%   \item$\mathcal R \in \mathcal S^3$ the allowed products.
-% \end{itemize}
-% It relies on two distinct judgments:
-% \begin{itemize}
-% \item $\Gamma\vdash a : t$ ($a$ has type $t$ in $\Gamma$),
-% \item $\Gamma\vdash t : \sort$ (the repository $t$ is well-typed of
-%   sort $s$ in $\Gamma$).
-% \end{itemize}
-% The rules are:
-% \begin{mathpar}
-%   \infer[Axiom]{
-%   }{\Gamma\vdash\sort_1:\sort_2}
-%   \quad(\sort_1,\sort_2)\in\mathcal A
-%   \and
-%   \infer[Init]{
-%     \Gamma(x):t
-%   }{\Gamma\vdash x:t}
-%   \and
-%   \infer[Prod]{\Gamma\vdash t:\sort_1 \and 
-%     \Gamma[x:t]\vdash u:\sort_2
-%   }{\Gamma\vdash(x:t)\cdot u : \sort_3}
-%   \quad(\sort_1,\sort_2,\sort_3)\in\mathcal R
-%   \and
-%   \infer[Let1]{\Gamma\vdash a:t \and
-%     \Gamma\vdash t:\sort_1 \and
-%     \Gamma[x=a:t]\vdash u:\sort_2
-%   }{\Gamma\vdash(x=a:t)\cdot u : \sort_3} 
-%   \quad (x=a:t)\notin\Gamma
-%   \and
-%   \infer[Let2]{\Gamma\vdash t:\sort_1 \and
-%     \Gamma\vdash u:\sort_2
-%   }{\Gamma\vdash(x=a:t)\cdot u : \sort_3} 
-%   \quad(x=a:t)\in\Gamma
-%   \and
-%   \infer[App]{
-%     \Gamma\vdash a:(y:t)\cdot u \and \Gamma(x):t
-%     }{\Gamma\vdash a\ x : u\subst{x}{y}}
-%   \and
-%   \infer[Skip1]{
-%     \Gamma\vdash a:(y=b:t)\cdot u \and \Gamma(x)=b:t
-%     }{\Gamma\vdash a : u\subst{x}{y}} \and
-%   \infer[Skip2]{
-%     \Gamma[x=b:t]\vdash a:(y=b:t)\cdot u
-%     }{\Gamma\vdash a : u\subst{x}{y}}
-% \end{mathpar}
-
-% The first four rules are standard. As application is only performed on
-% variables, the \textsc{App} rule looks up directly in the
-% environment. Rules \textsc{Let1} and \textsc{Skip1} show the
-% verification and the use of the equality binder in case the definition
-% is not already known. Rule \textsc{Let2} and \textsc{Skip2} allow
-% incremental typing in our system: if the definition was already known
-% in the environment, we do not need to recheck it. If $\cdot\vdash
-% t:\sort$, then the repository $t$ is well-typed.
-
-% \paragraph{Example}
-
-% Here is a well-typed repository describing the simply-typed
-% $\lambda$-calculus and constructing two $\lambda$-terms sharing some
-% sub-derivations: 
-% {\small
-% \begin{align}&
-% (\nat : *).
-% (o : \nat).
-% (s : \nat \to \nat).
-% (type : *).
-% (base : type).
-% (arr : type \to type \to type). \\&
-% (env : *).
-% (nil : env).
-% (cons : type \to env \to env). \\&
-% (term : *).
-% (var : \nat \to term).
-% (lam : term \to term).
-% (app : term \to term \to term). \\&
-% (lookup : env \to \nat \to type \to *).
-% (lookup_o : (e : env).(a : type).(f = cons\ a\ e : env). lookup\ e\ o\ a). \\&
-% (lookup_s : (e : env).(a b : type).(n : \nat).(sn = s\ n : \nat).(f =
-% cons\ a\ e : env). lookup\ e\ n\ b \to lookup\ f\ sn\ b). \\&
-% (of : env \to term \to type \to *). \\&
-% (of_v : (e : env).(a : type).(n : \nat).(v = var\ n : term).
-%           lookup\ e\ n\ a \to of\ e\ v\ a). \\&
-% (of_l : (e : env).(t : term).(a\ b: type).(c = arr\ a\ b : type).
-%           (u = lam\ t : term).(f = cons\ a\ e : env).
-%           of\ f\ t\ b \to of\ e\ u\ c). \\&
-% (of_a : (e : env).(t\ u: term).(a\ b: type).
-%           (c = arr\ a\ b : type).(v = app\ t\ u : term).
-%           of\ e\ t\ c \to of\ e\ u\ a \to of\ e\ v\ b). \\&
-% (1 = s\ o : \nat).(v_0 = var\ o : term).(v_1 = var\ 1 : term).
-% (t_3 = lam\ v_0 : term).(ch_0 = lam\ t_3 : term).\\&
-% (t_1 = app\ v_0\ v_1 : term).(t_2 = lam\ t_1 : term).
-% (t_4 = lam\ t_2 : term).
-% (bb = arr\ base\ base : type). \\&
-% (e_0 = cons\ bb\ nil : env).
-% (e_1 = cons\ base\ e0 : env).
-% (d_0 = lookup_o\ e_1\ base : lookup\ e_1\ o\ base). \\& %        (* a→a, a ⊢ 0 : a *)
-% (d_1 = of_v\ e_1\ base\ o\ d0 : of\ e_1\ v_0\ base). %	   (* a→a, a ⊢ 0 : a *)
-% (d_2 = of_l\ e_0\ v_0\ base\ base d_1 : of\ e_0\ t_3\ bb). \\& %   (* a→a ⊢ λ0 : a → a *)
-% (bbb = arr\ bb\ bb : type).
-% (d_3 = of_l\ nil\ t_3\ bb\ bb\ d_2 : of\ nil\ ch_0\ bbb). \\& %  (* ⊢ λλ0 : (a→a)→(a→a) *)
-% (g_1 = lookup_o\ e_0\ bb : lookup\ e_0\ o\ bb).
-% (g_2 = lookup_s\ e_0\ base\ bb\ o\ g_1 : lookup\ e_1\ 1\ bb). \\&
-% (g_3 = of_v\ e_1\ bb\ 1\ g_2 : of\ e_1\ v_1\ bb).	          % (* a→a, a ⊢ 1 : a→a *)
-% (t_5 = app\ v_1\ v0 : term). \\&
-% (g_4 = of_a\ e_1\ v_1\ v_0\ base\ base\ g_3\ d_1 : of\ e1\ t5\ base). % (* ... ⊢ 1 0 : a *)
-% (t_6 = lam\ t_5 : term). \\&
-% (g_5 = of_l\ e_0\ t_5\ base\ base\ g_4 : of\ e_0\ t_6\ bb).  %  (* a→a ⊢ λ(1 0) : a→a *)
-% (t_7 = lam\ t_6 : term). \\&
-% (g_6 = of_l\ nil\ t_6\ bb\ bb\ g_5 : of\ nil\ t_7\ bbb). *	 %  (* ⊢ λλ(1 0) : (a→a)→a→a  *)
-% \end{align}
-% }
 

### 2010-06-18 papers/MIPS2010/abstract.tex
* Tentative de reformulation de l approche a base de theorie des types
git-svn-id: http://gasp.dev.regis-gianas.com/svn/trunk@118 403e655e-980a-47bd-8c60-1ee5348bb6b7

@@
 management of mathematical repositories, to witness with more
-precision the \emph{impact of changes}. In proof assistants based on
-the propositions-as-types paradigm, Type Theory offers a powerful tool
-to witness the effect of changes finely: types, and we base our work
-on these.
+precision the \emph{impact of changes}. Following the Type Theory
+approach, our work is based on an algebra of expressive types that are
+meant to assign precise specifications to object-level term
+constructors and, in the meantime, to capture fine-grained
+dependencies between these objects.
+
+%% In proof assistants based on
+%% the propositions-as-types paradigm, Type Theory offers a powerful tool
+%% to witness the effect of changes finely: types, and we base our work
+%% on these.
 
 We will describe some of the possible directions to develop a tool to
-analyze the impact of change through types. It involves at its core a
+analyze the impact of changes through types. It involves at its core a
 typed description language for repositories, and is strongly related

### 2010-06-18 papers/MIPS2010/abstract.tex
* Quelque chose ne va pas avec cette phrase.
git-svn-id: http://gasp.dev.regis-gianas.com/svn/trunk@119 403e655e-980a-47bd-8c60-1ee5348bb6b7

@@
 meant to assign precise specifications to object-level term
-constructors and, in the meantime, to capture fine-grained
+constructors and, in the meantime, capture fine-grained
 dependencies between these objects.

### 2010-06-18 papers/MIPS2010/abstract.tex
passe sur le kernel
git-svn-id: http://gasp.dev.regis-gianas.com/svn/trunk@120 403e655e-980a-47bd-8c60-1ee5348bb6b7

@@
 Representing syntax and logics is nicely done in a \emph{logical
-  framework} like LF: both the syntactic
-elements and the typing derivations can sit in the same tree
-structure, and both can be rechecked at the same time, thanks to
-dependent types. For the purpose of incremental type checking though,
-our needs are a little bit different: first, we need to record, that
-is to name all intermediate values of our developments, so as to be
-able to address and reuse them multiple times. Secondly, we need to
-make sure that those intermediate values (sub-terms) are not recorded
-twice, so as to not type-check them twice: we are looking to represent
-syntactic and typing objects as a DAG rather than a tree.
-
-Our system will have these properties w.r.t. LF:
+  framework} like LF: both the syntactic elements and the typing
+derivations can sit in the same tree structure, and both can be
+rechecked at the same time, thanks to dependent types. For the purpose
+of incremental type checking though, our needs are a little bit
+different: first, we need to record, that is to name all intermediate
+values of our developments, so as to be able to address and reuse them
+multiple times. Secondly, we need to make sure that those intermediate
+values (sub-terms) are not recorded twice, so as to not type-check
+them twice: we are looking to represent syntactic and typing objects
+as a DAG rather than a tree. Moreover, we enforce by typing a property
+of \emph{maximal sharing}: every different subterm can be constructed
+exactly once. Our system will have these properties w.r.t. LF:
 \begin{itemize}
@@
 \item Every term should be a flat application of variables, so that we
-  don't introduce compound terms without naming them;
+  don't introduce compound terms without naming them and recording
+  their types;
 \item Finally we need a way to record intermediate definitions: we
@@
   invariant that not two same definition can sit in the context while
-  typing.
+  typing, thus guaranteeing \emph{maximal sharing} among terms and
+  derivations.
 \end{itemize}
 
-We describe the state of a repository at a given moment by a type in
-the following syntax. Well-typed types in the repository meta-language
-guarantee that it contains only well-typed proofs in the object
-language.
+We describe the state of a repository at a given moment by a type $t$
+in the following syntax. Well-typed types in the repository
+meta-language guarantee that it contains only well-typed proofs in the
+object language.
 \begin{align*}
  \gdecl{a}{x \gor a\ x } \\
- \gdecl{t}{a \gor \sort \gor (x:t)\cdot t \gor (x=a:t)\cdot t} \\
- \gdecl{\Gamma}{\cdot \gor \Gamma[x:t] \gor \Gamma[x=a:t]}
+ \gdecl{t}{a \gor \sort \gor (x:t)\cdot t \gor (x=a:t)\cdot t}
 \end{align*}
 
-\paragraph{Related work}
+From an implementation point of view, following \textsf{GIT}, we store
+these terms in a database of objects indexed by \emph{keys}, which are
+hash values of their contents, so that retreiving the type of a whole
+term boils down to compute its hash, and look for it in the database.
+
+Note that in this system, computation has no existence: we only
+provide ways to verify that syntax and typing rules of an object
+language are correct. Therefore, patches -- functions from
+repositories to repositories -- have no existence \emph{per se}. A
+necessary enhancement of our theory would be to consider computation
+as a way to apply patches as if they were constructive
+metatheorems. We conjecture that it would be done by re-extending the
+system towards LF (adding abstraction and reduction).
+
+\paragraph{Related works}
 \label{rw}

### 2010-06-18 papers/MIPS2010/abstract.tex
* Related work, encore des choses a ameliorer.
git-svn-id: http://gasp.dev.regis-gianas.com/svn/trunk@121 403e655e-980a-47bd-8c60-1ee5348bb6b7

@@
 here is to replace this tool chain and make it language-aware, both on
-a syntactic side through the use of abstract syntax trees instead of
+a syntactic side through the use of abstract syntax trees (AST) instead of
 concrete syntax and on a semantic side by using typing to ensure 
@@
 
-The \textsf{Twelf} project (\cite{pfenning1999system}) is an
+The \textsf{Twelf} project~\cite{pfenning1999system} is an
 implementation of the Logical Framework (LF,
@@
 to devise transformations of proofs in order to extract efficient
-programs.
+programs. Our kernel language reformulates LF by a syntactic
+explicitation of dependencies.
 
@@
 with in various proof assistant and mathematical repositories. The
-HELM project (\cite{asperti2006content})
-was an attempt to create a large library of mathematics, importing
-\textsf{Coq}'s developments into a searchable and browsable database.
-Most ideas from this project were imported into the \textsf{Matita}
-proof assistants (\cite{AspertiCTZ07}), especially a mechanism of
+HELM project~\cite{asperti2006content} was an attempt to create a
+large library of mathematics, importing \textsf{Coq}'s developments
+into a searchable and browsable database.  Most ideas from this
+project were imported into the \textsf{Matita} proof
+assistants~\cite{AspertiCTZ07}, especially a mechanism of
 \emph{invalidation and regeneration} to ensure the global consistency
 of its library w.r.t changes, with granularity the whole definitions
-or proofs and their dependencies. The MBase project
-(\cite{kohlhase2001mbase}) attempts at creating a web-based,
+or proofs and their dependencies. The MBase
+project~\cite{kohlhase2001mbase} attempts at creating a web-based,
 distributed mathematical knowledge database putting forward the idea
-of \emph{development graph} (\cite{hutter2000management},
-\cite{autexier2000towards}) to manage changes in the database,
-allowing semantic-based retrieval and object-level dependency
-management.
+of \emph{development
+  graph}~\cite{hutter2000management,autexier2000towards} to manage
+changes in the database, allowing semantic-based retrieval and
+object-level dependency management. 
+% Y: Un mot pour se comparer a ces travaux?
 
 This idea, generalized over structured, semi-formal documents gave
-birth to \texttt{\it locutor} (\cite{muller2008fine}), a fine-grained
-extension of the \texttt{svn} version control system for XML
+birth to \texttt{\it locutor}~\cite{muller2008fine}, a fine-grained
+extension of the \textsc{svn} version control system for XML
 documents, embedding ontology-driven, user-defined semantic knowledge
@@
 operating on text modulo defined isomorphisms.
+% Y: meme chose, comment on se compare sur ce point?
 
@@
 languages by virtue of the Curry-Howard isomorphism, and because we
-operate at the AST level. Secondly, by taking \emph{types} as a
+operate at the AST level. Secondly, by taking \emph{types} as 
 witnesses for the evolution of a development, we refine the usual,

### 2010-06-18 papers/MIPS2010/abstract.tex
* typo.
git-svn-id: http://gasp.dev.regis-gianas.com/svn/trunk@122 403e655e-980a-47bd-8c60-1ee5348bb6b7
* correct display of citation.
git-svn-id: http://gasp.dev.regis-gianas.com/svn/trunk@123 403e655e-980a-47bd-8c60-1ee5348bb6b7

@@
 modules, file-based scripts and separate compilation in proof
-assistants like \textsf{Coq}~\cite{CoqDocWeb} or
-\textsf{Matita}~\cite{AspertiCTZ07}, the use of
+assistants like \textsf{Coq}~\citep{CoqDocWeb} or
+\textsf{Matita}~\citep{AspertiCTZ07}, the use of
 dependency management tools (\textsf{Make}) or version control system
@@
 been shown to be of crucial to the mathematical discovery
-process~(\cite{lakatos1964proofs}), and we believe that they should be
+process~\citep{lakatos1964proofs}, and we believe that they should be
 taken into account in the realization of mathematical assistants.
@@
 of languages and their metatheory (e.g. \textsf{Twelf},
-{\cite{pfenning1999system}}), and substantial formalizations have been
-undertaken\cite{lee07}, we still use legacy tools based on text
+{\citep{pfenning1999system}}), and substantial formalizations have been
+undertaken~\citep{lee07}, we still use legacy tools based on text
 representation to manage our developments. The general goal exposed
@@
 syntax (our proofs) and their derivations. Describing transformations
-among syntax objects is done by sharing common subterms or
+among syntactic objects is done by sharing common subterms or
 subderivations.
@@
 
-The \textsf{Twelf} project~\cite{pfenning1999system} is an
-implementation of the Logical Framework (LF,
-\cite{harper1993framework}). It was used in \cite{anderson1993program}
-to devise transformations of proofs in order to extract efficient
-programs. Our kernel language reformulates LF by a syntactic
-explicitation of dependencies.
+The \textsf{Twelf} project~\citep{pfenning1999system} is an
+implementation of the Logical
+Framework~\citep{harper1993framework}. It was used in
+\citep{anderson1993program} to devise transformations of proofs in
+order to extract efficient programs. Our kernel language reformulates
+LF by a syntactic explicitation of dependencies.
 
@@
 with in various proof assistant and mathematical repositories. The
-HELM project~\cite{asperti2006content} was an attempt to create a
+HELM project~\citep{asperti2006content} was an attempt to create a
 large library of mathematics, importing \textsf{Coq}'s developments
@@
 project were imported into the \textsf{Matita} proof
-assistants~\cite{AspertiCTZ07}, especially a mechanism of
+assistants~\citep{AspertiCTZ07}, especially a mechanism of
 \emph{invalidation and regeneration} to ensure the global consistency
@@
 or proofs and their dependencies. The MBase
-project~\cite{kohlhase2001mbase} attempts at creating a web-based,
+project~\citep{kohlhase2001mbase} attempts at creating a web-based,
 distributed mathematical knowledge database putting forward the idea
 of \emph{development
-  graph}~\cite{hutter2000management,autexier2000towards} to manage
+  graph}~\citep{hutter2000management,autexier2000towards} to manage
 changes in the database, allowing semantic-based retrieval and
@@
 This idea, generalized over structured, semi-formal documents gave
-birth to \texttt{\it locutor}~\cite{muller2008fine}, a fine-grained
+birth to \texttt{\it locutor}~\citep{muller2008fine}, a fine-grained
 extension of the \textsc{svn} version control system for XML
@@
 equality theory to quotient the syntax. On the same line of work, we
-should mention the \emph{Coccinelle} tool
-(\cite{padioleau2008documenting}). It is an evolution over textual
+should mention the \emph{Coccinelle}
+tool~\citep{padioleau2008documenting}. It is an evolution over textual
 patches, specialized on the C language, allowing more flexibility in

### 2010-06-18 papers/MIPS2010/abstract.tex
* Work est indenombrable.
git-svn-id: http://gasp.dev.regis-gianas.com/svn/trunk@124 403e655e-980a-47bd-8c60-1ee5348bb6b7
...
git-svn-id: http://gasp.dev.regis-gianas.com/svn/trunk@125 403e655e-980a-47bd-8c60-1ee5348bb6b7

@@
 the object (proof-)language and its typing rules, and define pieces of
-syntax (our proofs) and their derivations. Describing transformations
-among syntactic objects is done by sharing common subterms or
-subderivations.
+syntax (our proofs, potentially with omitted informations) and their
+derivations (fully explicit application of typing rules). Describing
+transformations among syntax objects is done by sharing common
+subterms or subderivations.
 
@@
 them twice: we are looking to represent syntactic and typing objects
-as a DAG rather than a tree. Moreover, we enforce by typing a property
-of \emph{maximal sharing}: every different subterm can be constructed
-exactly once. Our system will have these properties w.r.t. LF:
+as a directed acyclic graph (DAG) rather than a tree. Moreover, we
+enforce by typing a property of \emph{maximal sharing}: every
+different subterm can be constructed exactly once. Our system will
+have these properties w.r.t. LF:
 \begin{itemize}
@@
   computations to take place within our DAGs. Our syntax will then be
-  restricted to product types and applications;
-\item Every term should be a flat application of variables, so that we
-  don't introduce compound terms without naming them and recording
-  their types;
+  restricted to product types $(x:t)\cdot t$ and applications;
+\item Every term should be a flat application of variables $a ::= \vec x$, so
+  that we don't introduce compound terms without naming them and
+  recording their types;
 \item Finally we need a way to record intermediate definitions: we
-  introduce a new kind of binder, the equality binder. We maintain the
-  invariant that not two same definition can sit in the context while
-  typing, thus guaranteeing \emph{maximal sharing} among terms and
-  derivations.
+  introduce a new kind of binder, the equality binder $(x=a)\cdot
+  t$. We maintain the invariant that no two syntactically equal
+  definition can sit in the context while typing, thus guaranteeing
+  \emph{maximal sharing} among terms and derivations.
 \end{itemize}
@@
 
-\paragraph{Related works}
+\paragraph{Related work}
 \label{rw}

### 2010-06-18 papers/MIPS2010/abstract.tex
* Reformulation.
git-svn-id: http://gasp.dev.regis-gianas.com/svn/trunk@126 403e655e-980a-47bd-8c60-1ee5348bb6b7
* Institution.
git-svn-id: http://gasp.dev.regis-gianas.com/svn/trunk@127 403e655e-980a-47bd-8c60-1ee5348bb6b7

@@
 
-\author{Matthias Puech \and Yann Régis-Gianas}
+\author{
+\begin{tabular}{cc}
+Matthias Puech &
+Yann Régis-Gianas \\ 
+&
+{\small \url{yrg@pps.jussieu.fr}}\\
+& 
+{\small Laboratoire PPS, équipe ${\pi}r^2$} \\
+&
+{\small University Paris 7, CNRS, and INRIA}
+\end{tabular}
+}
 \date{}
@@
 Practical efforts in the formalization of mathematical results have
-naturally led to the questions of how to manage large repositories of
+naturally led to the question of how to manage large repositories of
 proofs, i.e. what is the \emph{daily workflow} of users of a proof
@@
 order to extract efficient programs. Our kernel language reformulates
-LF by a syntactic explicitation of dependencies.
+a fragment of LF to make dependencies syntactically explicit.
 

### 2010-06-18 papers/MIPS2010/abstract.tex
* Derniere passe sur intro.
git-svn-id: http://gasp.dev.regis-gianas.com/svn/trunk@128 403e655e-980a-47bd-8c60-1ee5348bb6b7

@@
 checking is usually too expensive computationally to mimic this
-workflow. But even besides the time factor, this ``traditional'' way
+workflow. But even beyond the time factor, this ``traditional'' way
 of formalizing mathematics hinders the process of mathematical
@@
 for alternate, inequivalent definitions. This fact has nonetheless
-been shown to be of crucial to the mathematical discovery
+been shown to be crucial to the mathematical discovery
 process~\citep{lakatos1964proofs}, and we believe that they should be
@@
 a syntactic side through the use of abstract syntax trees (AST) instead of
-concrete syntax and on a semantic side by using typing to ensure 
+concrete syntax, and on a semantic side by using typing to ensure 
 repository consistency.

### 2010-06-18 papers/MIPS2010/abstract.tex
* Adresse.
git-svn-id: http://gasp.dev.regis-gianas.com/svn/trunk@129 403e655e-980a-47bd-8c60-1ee5348bb6b7

@@
 &
-{\small University Paris 7, CNRS, and INRIA}
+{\small University Paris 7, CNRS, and INRIA} \\
+&
+{\small 23, avenue d'Italie. 75013 Paris}
 \end{tabular}

### 2010-06-18 papers/MIPS2010/abstract.tex
adresses correctes
git-svn-id: http://gasp.dev.regis-gianas.com/svn/trunk@130 403e655e-980a-47bd-8c60-1ee5348bb6b7

--- 2010-06-18 papers/MIPS2010/abstract.tex
@@
 \begin{tabular}{cc}
-Matthias Puech &
-Yann Régis-Gianas \\ 
-&
-{\small \url{yrg@pps.jussieu.fr}}\\
-& 
-{\small Laboratoire PPS, équipe ${\pi}r^2$} \\
-&
-{\small University Paris 7, CNRS, and INRIA}
+Matthias Puech & Yann Régis-Gianas \\
+{\small \url{puech@cs.unibo.it}} & {\small \url{yrg@pps.jussieu.fr}}\\
+{\small Dept. of Computer Science} & {\small Laboratoire PPS, équipe ${\pi}r^2$} \\
+{\small University of Bologna}& {\small University Paris 7, CNRS, and INRIA} \\
+{\small Mura Anteo Zamboni 7, 40127 Bologna} & {\small 23, avenue d'Italie, 75013 Paris}
 \end{tabular}

### 2010-07-05 talks/MIPS2010/slides.tex
mise en place des slides (beaucoup plus manuel, mais c'est la seule chose que je comprends)
git-svn-id: http://gasp.dev.regis-gianas.com/svn/trunk@134 403e655e-980a-47bd-8c60-1ee5348bb6b7

\documentclass[red]{beamer}
\usepackage{ucs}
\usepackage{mathpartir}
\usepackage{amsfonts,amsmath,amscd}
\usepackage{stmaryrd}
\usepackage[utf8]{inputenc}
\usepackage[protrusion=true,expansion=true]{microtype}

\title{Towards typed repositories of proofs}

\date{July 10, 2010}

\author[Matthias Puech \& Yann Régis-Gianas] {
Matthias Puech\inst{1} \and Yann Régis-Gianas\inst{2} \\
{\small \url{puech@cs.unibo.it}} \and {\small \url{yrg@pps.jussieu.fr}}
}
\institute {
  \inst 1 {\small Dept. of Computer Science, University of Bologna} \and
  \inst 2 {\small University Paris 7, CNRS, and INRIA, PPS, team ${\pi}r^2$}
}

\setbeamertemplate{footline}
{ \sf
  \hbox{%
  \begin{beamercolorbox}[wd=.5\paperwidth,ht=2.25ex,dp=1ex,left]{button}%
    \usebeamerfont{title in head/foot} MIPS 2010
  \end{beamercolorbox}%
  \begin{beamercolorbox}[wd=.5\paperwidth,ht=2.25ex,dp=1ex,right]{frame
    number}%
    \usebeamerfont{date in head/foot}
    \insertframenumber{} / \inserttotalframenumber\hspace*{2ex} 
  \end{beamercolorbox}}%
  \vskip0pt%
}

%\setbeamertemplate{footline}[frame number]
\usenavigationsymbolstemplate{}

\AtBeginSection[]
{\begin{frame}<beamer>{Outline}
    \tableofcontents[currentsection]
  \end{frame}
}

\usefonttheme{serif}

\begin{document}

\frame{\titlepage}

\section*{Introduction}

\frame{un \pause deux}

\begin{frame}{Outline}
  \tableofcontents
\end{frame}
\section{Section 1}
\frame{bla}
\section{Section 2}
\frame{blo}

\end{document}

### 2010-07-05 talks/MIPS2010/whizzy.sh

@@ -1,2 +1,0 @@
-export MKSLICE="latexpp -w -g color yes"
-export TEXINPUTS="common":$TEXINPUTS

### 2010-07-05 talks/MIPS2010/slides.tex
structure prête, un peu de contenu
git-svn-id: http://gasp.dev.regis-gianas.com/svn/trunk@135 403e655e-980a-47bd-8c60-1ee5348bb6b7

@@
-\documentclass[red]{beamer}
+% \documentclass[ignorenonframetext,red]{beamer}
+
+\documentclass{article} \usepackage{beamerarticle}
+
 \usepackage{ucs}
@@
 \author[Matthias Puech \& Yann Régis-Gianas] {
-Matthias Puech\inst{1} \and Yann Régis-Gianas\inst{2} \\
+Matthias Puech\inst{1,2} \and Yann Régis-Gianas\inst{2} \\
 {\small \url{puech@cs.unibo.it}} \and {\small \url{yrg@pps.jussieu.fr}}
@@
 
-\setbeamertemplate{footline}
-{ \sf
-  \hbox{%
-  \begin{beamercolorbox}[wd=.5\paperwidth,ht=2.25ex,dp=1ex,left]{button}%
-    \usebeamerfont{title in head/foot} MIPS 2010
-  \end{beamercolorbox}%
-  \begin{beamercolorbox}[wd=.5\paperwidth,ht=2.25ex,dp=1ex,right]{frame
-    number}%
-    \usebeamerfont{date in head/foot}
-    \insertframenumber{} / \inserttotalframenumber\hspace*{2ex} 
-  \end{beamercolorbox}}%
-  \vskip0pt%
-}
-
-%\setbeamertemplate{footline}[frame number]
-\usenavigationsymbolstemplate{}
+\setbeamertemplate{footline}[frame number]
+\setbeamertemplate{navigation symbols}{}
 
@@
 
-\frame{\titlepage}
-
+\begin{frame}
+  \titlepage
+  \mode<article>{\newcommand\url\texttt\maketitle}
+\end{frame}
 \section*{Introduction}
 
-\frame{un \pause deux}
+This talk will be about some remarks and directions on the way the
+edition of formal proof is done in proof assistants, especially those
+based on type theory like \textsf{Coq} or \textsf{Matita}.
+
+\begin{frame}{A paradoxical situation}
+  \begin{itemize}
+  \item Most time spent \emph{editing}, not \emph{writing}
+  \item Workflow of formal mathematics largely inspired by software development
+  \end{itemize}  
+\end{frame}
+
+Let me start with two simple observations relating the activity of
+both the working mathematician and the programmer. A first, obvious
+fact is that both spend more time \emph{editing} than \emph{writing}:
+their workflow is highly non-linear, involving experiments, fixes,
+backtrack on previous modifications\ldots, eventually validated or
+invalidated by some criteria like the absence of bugs, the conformance
+to a specifications on one side, or the validity of a proof, the
+adequacy of a definition on the other.
+
+The second observation is that since the advent of modern proof
+assistants, large bodies of mathematics have been formalized and the
+way people managed these developments is very similar to the one used
+in software development: we write proofs in text buffers, split the
+development into separate files
 
@@
 \end{frame}
-\section{Section 1}
-\frame{bla}
-\section{Section 2}
-\frame{blo}
+
+\section{}
+
+\subsection{Practical applications}
+
+\section{A core meta-language for proof repositories}
 

### 2010-07-05 talks/MIPS2010/slides.tex
bilan de la journée: niet!
git-svn-id: http://gasp.dev.regis-gianas.com/svn/trunk@136 403e655e-980a-47bd-8c60-1ee5348bb6b7

--- 2010-07-05 talks/MIPS2010/slides.tex
@@
-\documentclass[red]{beamer}
+\documentclass[ignorenonframetext,red]{beamer}
+
+% \documentclass{article} \usepackage{beamerarticle}
+
 \usepackage{ucs}
@@
 \usepackage{stmaryrd}
-\usepackage[utf8]{inputenc}
+\usepackage[utf8x]{inputenc}
 \usepackage[protrusion=true,expansion=true]{microtype}
+\usepackage{dot2texi}
+\usepackage{tikz}
+\usetikzlibrary{shapes,arrows}
 
@@
 
-\frame{\titlepage}
-
+\begin{frame}
+  \titlepage
+  \mode<article>{
+    \newcommand\url\texttt
+    \maketitle}
+\end{frame}
 \section*{Introduction}
 
-\frame{un \pause deux}
+This talk will be about some remarks and directions on the way the
+edition of formal proof is done in proof assistants.
 
-\begin{frame}{Outline}
+\begin{frame}{A paradoxical situation}
+  Most time is spent \emph{editing}, not \emph{writing} proofs.
+
+  Workflow of formal mathematics is largely inspired by software
+  development:
+  \begin{itemize}
+  \item Text-based
+  \item Separate compilation (\textsf{make})
+  \item Versioning (\textsf{cvs}, \textsf{svn}, \textsf{diff}s\ldots)
+  \end{itemize}
+\end{frame}
+
+Let me start with two simple observations relating the activity of
+both the working mathematician and the programmer. A first, obvious
+fact is that both spend more time \emph{editing} than \emph{writing}:
+their workflow is highly non-linear, involving experiments, fixes,
+backtrack on previous modifications\ldots, eventually validated or
+invalidated by some criteria like the absence of bugs or the validity
+of a proof.
+
+The second observation is that since the advent of modern proof
+assistants, large bodies of mathematics have been formalized and the
+way people managed these developments is very similar to the one used
+in software development: we write proofs in text buffers, which are
+linear, and launch proof-checking when we are done. To avoid
+re-checking the whole development, we split it into mutually dependent
+files and re-check only those who have changed. To keep track of old
+versions, we use line-based version control systems that have no
+awareness of the actual content of the files. The separation into
+multiple files then serves two different purpose, and the fact that
+they coincide is pure contingency: the purpose of splitting a
+development into logical units, and the purpose of easing the
+non-linear process of proof editing. Also, recording the actual text
+of the proofs for versioning is often non-satisfactory: on one side it
+contains mostly non-relevant informations (line breaks, spaces\ldots),
+on the other the relevant ones, the validity of the proof, the
+dependency between objects, is absent.
+
+The paradox in this situation is that these very tools can be used to
+reason about formal languages and especially proof languages. Couldn't
+we adapt the methods developped to reason about a formal language to
+the proof language of proof assistants itself? Could we replace this
+legacy toolchain and make it language-aware?
+
+\begin{frame}{Expectations}
+
+\end{frame}
+
+\begin{frame}{\mode<beamer>{Outline}}
   \tableofcontents
 \end{frame}
-\section{Section 1}
-\frame{bla}
-\section{Section 2}
-\frame{blo}
+
+\section{}
+
+\subsection{Practical applications}
+
+\section{A core meta-language for proof repositories}
+
+\begin{frame}{Inspiration: the \textsf{git} storage model}
+  \begin{itemize}
+  \item ``Content-adressable''
+  \item Name reflects content
+  \item Maximal sharing (or hashconsing)
+  \end{itemize}
+\end{frame}
+
+\begin{frame}{Incremental type-checking}
+  A small language for representing proofs:
+  \only<1>{\[ t\ ::=\ (x:t)\cdot t\ |\ x\ |\ t\ t\ |\ *\] \[\]}%
+  \only<2>{\[ t\ ::=\ (x:t)\cdot t\ |\ \alert x\ |\ \alert{t\ t}\ |\ *\] \[\]}%
+  \only<3>{\[ t\ ::=\ (x:t)\cdot t\ |\ \alert a\ |\ *\]%
+    \[ a\ ::=\ x\ |\ a\ x \]} 
+  \only<4>{\[ t\ ::=\ (x:t)\cdot t\ |\ a\
+    |\ *\ |\ \alert{(x = a)\cdot t}\]%
+    \[ a\ ::=\ x\ |\ a\ x \]}
+  \begin{example}
+    \[ \infer[Lam]{\Gamma,A\vdash B}{\Gamma\vdash A\to B} \]
+    \[ \textsc{Lam}\ :\ (\Gamma : env)\cdot(A B : type)\cdot
+    (d : \Gamma,A\vdash B)\cdot \Gamma\vdash A\to B \]
+  \end{example}
+\end{frame}
 

### 2010-07-05 talks/MIPS2010/slides.tex
slides mips2010: Partie motivations bien avancées
git-svn-id: http://gasp.dev.regis-gianas.com/svn/trunk@137 403e655e-980a-47bd-8c60-1ee5348bb6b7

@@
 \usepackage[protrusion=true,expansion=true]{microtype}
-\usepackage{dot2texi}
-\usepackage{tikz}
-\usetikzlibrary{shapes,arrows}
+\usepackage{setspace}
 
@@
 
-\AtBeginSection[]
-{\begin{frame}<beamer>{Outline}
-    \tableofcontents[currentsection]
-  \end{frame}
-}
+% \AtBeginSection[]
+% {\begin{frame}<beamer>{Outline}
+%     \tableofcontents[currentsection]
+%   \end{frame}
+% }
 
@@
 \end{frame}
-\section*{Introduction}
+
+\section{Motivation}
 
@@
 \begin{frame}{A paradoxical situation}
-  Most time is spent \emph{editing}, not \emph{writing} proofs.
-
-  Workflow of formal mathematics is largely inspired by software
+  \begin{block}{Observation 1}
+    Most of a programmer/mathematician's time is spent \emph{editing}, not \emph{writing}
+  \end{block}
+  \begin{block}{Observation 2}
+    Workflow of formal mathematics was largely inspired by software
     development:
     \begin{itemize}
-  \item Text-based
+    \item File-based scripts
     \item Separate compilation (\textsf{make})
-  \item Versioning (\textsf{cvs}, \textsf{svn}, \textsf{diff}s\ldots)
+    \item Text-based versioning (\textsf{cvs}, \textsf{svn},
+      \textsf{diff}s\ldots)
     \end{itemize}
+  \end{block}
+  \begin{block}{Postulate}
+    \it Current administrative toolchain should be adapted to proof languages
+  \end{block}
+  {\Large }
 \end{frame}
 
-Let me start with two simple observations relating the activity of
+Let me start with some simple observations relating the activity of
 both the working mathematician and the programmer. A first, obvious
@@
 linear, and launch proof-checking when we are done. To avoid
-re-checking the whole development, we split it into mutually dependent
+re-checking the whole development, we split it into inter-dependent
 files and re-check only those who have changed. To keep track of old
@@
 awareness of the actual content of the files. The separation into
-multiple files then serves two different purpose, and the fact that
-they coincide is pure contingency: the purpose of splitting a
-development into logical units, and the purpose of easing the
-non-linear process of proof editing. Also, recording the actual text
-of the proofs for versioning is often non-satisfactory: on one side it
-contains mostly non-relevant informations (line breaks, spaces\ldots),
-on the other the relevant ones, the validity of the proof, the
-dependency between objects, is absent.
+multiple files serves two different purpose, and the fact that they
+coincide is pure contingency: the purpose of splitting a development
+into logical units, and the purpose of easing the non-linear process
+of proof editing. Also, recording the actual text of the proofs for
+versioning is often non-satisfactory: on one side it contains mostly
+non-relevant informations (line breaks, spaces\ldots), on the other
+the relevant ones, the validity of the proof, the dependency between
+objects, is absent.
 
@@
 reason about formal languages and especially proof languages. Couldn't
-we adapt the methods developped to reason about a formal language to
-the proof language of proof assistants itself? Could we replace this
-legacy toolchain and make it language-aware?
+we adapt the methods developped to reason about formal languages to
+the language of proof assistants itself? Could we replace this legacy
+toolchain and make it language-aware?
 
-\begin{frame}{Expectations}
+\begin{frame}{Motivations}
+  Current toolset is not adapted to proof edition:
+  \begin{itemize}
+  \item ((edit; compile)*; commit)* loop does not scale to proofs
+  \item Concept freeze inhibits the discovery process
+  \item Any change in a file requires full recompilation of
+    dependencies
+  \item No room for alternate definitions (linearity of scripts)
+  \item Textual scripts \textsf{diff}s do not reflect the semantics
+  \item Not even the syntax
+  \end{itemize}
+  \vspace{2em}
+  {\tiny \ldots\ Maybe it wasn't adapted to software development}
+\end{frame}
+
+More precisely, here are some problems that arise with the traditional
+tools when editing proofs. Whereas the usual edit, compile, commit
+loop may be a satisfactory approximation when developing software, the
+additional demands of proof-checking makes it unsatisfactory as the
+development grows: often the time of recompilation is prohibitive.
+
+But it is not only a matter of time: once a concept is proof-checked,
+it is frozen and any future modification of it will require to
+re-check a large part of the development, often much more than what is
+actually required. We believe that this facts inhibits the use of
+proof assistants as tools to help the mathematician in the discovery
+process, and not only to formally transcribe paper proofs.
+
+The same way, the linearity of proof scripts does not allow any
+flexibility in the alternate definition of concepts.
+
+Finally of course, the way we store mathematical developments, that is
+as textual files or textual diffs does not reflect neither their syntax
+nor their semantics, and as I will try to show, there is a lot to gain
+in changing this base representation.
+
+In fact, some will say that this model of interaction wasn't even
+adapted to programming in the first place.
+
+\begin{frame}{The impact of changes}
+  These problems were addressed by a number of (ad-hoc) fixes:
+  \begin{itemize}
+  \item File-based separate compilation
+  \item Interface/implementation
+  \item Interaction loop with global undo
+  \end{itemize}
+  \vspace{2em}
+  \begin{center} \Large
+    \begin{spacing}{1.3}
+      \textit{Need for a refined notion of dependency to analyze the
+        impact of changes}
+    \end{spacing}
+  \end{center}
+\end{frame}
+
+We can then see a number of extra-logical features of modern proof
+assistants as ways to cope with these problems. Starting from a
+hypothetical ``dark age'' of proof assistants where one would write
+its complete development in one unique file and let the program check
+it when done, were added:
+\begin{itemize}
+\item The ability to split it into files organized in a dependency
+  graph as discussed earlier, which is a coarse approximation of
+  actual dependencies amongst objects
+\item The ability to separate interfaces (statements) from
+  implementations (proofs), and recheck dependencies only if
+  implementations changed.
+\item Most proof assistants base their interactivity on a mechanism
+  for backtracking their entire states globally, as a way to make
+  changes to already checked parts.
+\end{itemize}
+
+\begin{frame}{Example: \textsf{coqide}}
 
@@
 
-\begin{frame}{\mode<beamer>{Outline}}
-  \tableofcontents
+\begin{frame}{Methodology}
+  \begin{center}
+    \only<1-2>{\fbox{\parbox{20em}{\only<2>\alert{version control}\\[1em]%
+        \fbox{\parbox{18em}{script files\\[1em]%
+            \fbox{\parbox{16em}{parsing\\[1em]%
+                \fbox{\parbox{14em}{proof-checking\\} }}}}}}}}%
+    \only<3>{\fbox{\parbox{20em}{script files\\[1em]%
+        \fbox{\parbox{18em}{\alert{version control}\\[1em]%
+            \fbox{\parbox{16em}{parsing\\[1em]%
+                \fbox{\parbox{14em}{proof-checking\\} }}}}}}}}%
+    \only<4->{\fbox{\parbox{20em}{script files\\[1em]%
+        \fbox{\parbox{18em}{parsing\\[1em]%
+            \fbox{\parbox{16em}{\alert{version control \only<5->{+ proof-checking}}\\[1em]%
+                \uncover<4>{\fbox{\parbox{14em}{proof-checking\\}}}}}}}}}}%
+  \end{center}
+  \uncover<6>{
+    \begin{itemize}
+    \item AST representation
+    \item Typing annotations
+    \item Incremental type-checking
+    \end{itemize}
+  }
 \end{frame}
-
-\section{}
-
-\subsection{Practical applications}
 

### 2010-07-06 talks/MIPS2010/slides.tex
Commentaires et small fixes
git-svn-id: http://gasp.dev.regis-gianas.com/svn/trunk@138 403e655e-980a-47bd-8c60-1ee5348bb6b7
slides mips 2010: première partie finie.
git-svn-id: http://gasp.dev.regis-gianas.com/svn/trunk@140 403e655e-980a-47bd-8c60-1ee5348bb6b7

--- 2010-07-05 talks/MIPS2010/slides.tex
@@
+% Mode d'emploi: Switcher les deux lignes suivantes pour passer du
+% mode présentation au handout
+% ===================================================
 \documentclass[ignorenonframetext,red]{beamer}
-
-% \documentclass{article} \usepackage{beamerarticle}
+% \documentclass[12pt]{article} \usepackage{beamerarticle} \usepackage{fullpage}
+% ===================================================
 
@@
 \usepackage[protrusion=true,expansion=true]{microtype}
-\usepackage{dot2texi}
-\usepackage{tikz}
-\usetikzlibrary{shapes,arrows}
+\usepackage{setspace}
+\usepackage{graphicx}
 
-\title{Towards typed repositories of proofs}
+\title{Towards typed repositories of proofs \\[0.6em] 
+  \small \textsf{MIPS 2010}}
 
@@
 
+\hrule
+\begin{frame}{How are \emph{constructed} formal mathematics?}
+  \large
+  \begin{tabular}{ll}
+    {\Huge Q :} & \parbox{0.8\textwidth}{What is the common point between the working
+      mathematician and the working programmer?} \\[2em]
+    \pause
+    {\Huge A :} & They both spend more time \emph{editing} than \emph{writing}
+  \end{tabular}
+\end{frame}
+\hrule
+
+Let me start with a simple observation relating the activity of both
+the working mathematician and the computer scientist. Whereas both
+activities are percieved, from the outside, as apodictic because the
+final object of their study, the proof or the program, seems to be
+unquestionably valid (or not).
+
+If we observe the day-to-day work of these scientists though, their
+workflow resembles a lot more that of experimental sciences: it is
+highly non-linear, involving multiple parallel experiments, fixes,
+backtrack on previous modifications\ldots\ eventually validated or
+invalidated by some \emph{a posteriori} criterion, the absence of
+bugs, the validity of a proof.
+
+In other words, the mathematician and the programmer spend more time
+\emph{editing} their developments non-linearly than monotonously
+\emph{writing} new material.
+
+It is the case for programs, for classical, paper mathematics so it
+should be the case for formal mathematics, edited in a proof
+assistant.
+
+\hrule
 \begin{frame}{A paradoxical situation}  
-  Most time is spent \emph{editing}, not \emph{writing} proofs.
-
-  Workflow of formal mathematics is largely inspired by software
+  \begin{block}{Observation}
+    Workflow of formal mathematics was largely inspired by software
     development:
     \begin{itemize}
-  \item Text-based
+    \item File-based scripts (\textsf{emacs})
     \item Separate compilation (\textsf{make})
-  \item Versioning (\textsf{cvs}, \textsf{svn}, \textsf{diff}s\ldots)
+    \item Text-based versioning (\textsf{svn}, \textsf{diff}s\ldots)
+    \end{itemize}
+  \end{block}
+  \pause
+  \begin{block}{Mission}
+    \large \it Isn't it time to adapt these tools to proofs?
+  \end{block}
+  {\Large }
+\end{frame}
+\hrule
+
+Since the advent of modern proof assistants, large bodies of
+mathematics have been formalized and the way people managed these
+developments is very similar to the one used in software development
+for years: we write proofs in text files, which are linear, and launch
+proof-checking when we are done. To avoid re-checking the whole
+development, we split it into inter-dependent files and re-check only
+those who have changed. To keep track of old versions, we use
+line-based version control systems that have no awareness of the
+actual content of the files. In a sense, the non-linearity of proof
+development is managed outside the scope of the tool, by manual,
+textual transformations that are often very unsatisfactory.
+
+The paradox about this situation is that these very tools can be used
+to reason about formal languages and especially proof
+languages: some of these tools are even dedicated to the formalization
+of proof systems and their properties. But we still use the legacy
+generic tools to manage our developments.
+
+Couldn't we adapt the methods developed to reason about formal
+languages to the language of the proof assistant itself? Could we
+replace this legacy toolchain and make it language-aware, even
+semantic-aware?
+
+\hrule
+\begin{frame}{Motivations}
+  \begin{block}{Rigidity of linear edition}
+    \begin{itemize}
+    \item ((edit; compile)*; commit)* loop does not scale to proofs
+    \item Concept freeze inhibits the discovery process
+    \item No room for alternate definitions
+    \end{itemize}
+  \end{block}
+  \pause
+  \begin{block}{Laxity of textual representation}
+    \begin{itemize}
+    \item Textual scripts \textsf{diff}s do not reflect the semantics
+    \item Not even the syntax
+    \end{itemize}
+  \end{block}
+  \vspace{2em}
+  \pause
+  {\tiny \ldots\ Maybe it wasn't adapted to software development}
+\end{frame}
+\hrule
+
+More precisely, here are some problems that arise with the traditional
+tools when editing proofs. Whereas the usual edit, compile, commit
+loop may be a satisfactory approximation when developing software, the
+additional demands of proof-checking makes it unsatisfactory as the
+development grows: often the time of recompilation is prohibitive.
+
+But it is not only a matter of time: once a concept is proof-checked,
+it is frozen and any future modification of it will require to
+re-check a large part of the development, often much more than what is
+actually required. We believe that this fact inhibits the use of proof
+assistants as tools to help the mathematician in the discovery
+process, as an experimentation tool, and not only to formally
+transcribe paper proofs.
+
+The same way, the linearity of proof scripts does not allow any
+flexibility in the alternate definition of concepts.
+
+On the other side, the way we store mathematical developments, that is
+as textual files or textual diffs does not reflect neither their
+syntax nor their semantics, and as I will try to show, there is a lot
+to gain in changing this base representation.
+
+In fact, some will say that this model of interaction wasn't even
+adapted to programming in the first place.
+
+\begin{frame}{The impact of changes}
+  \vspace{0.5em}
+  \begin{center}
+    \only<1-3>{\includegraphics[width=0.75\textwidth]{images/coqide.png}}%
+    \only<4>{\includegraphics[width=0.75\textwidth]{images/coqide2.png}}%
+    \only<5>{\includegraphics[width=0.75\textwidth]{images/coqide4.png}}%
+    \only<6>{\includegraphics[width=0.75\textwidth]{images/coqide5.png}}%
+    \only<7>{\includegraphics[width=0.75\textwidth]{images/coqide6.png}}%
+    \only<8>{\includegraphics[width=0.75\textwidth]{images/coqide7.png}}%
+  \end{center}
+  \pause
+  \uncover<2->{
+    \begin{itemize}
+    \item File-based separate compilation
+      \pause
+    \item Interaction loop with global undo
+    \end{itemize}
+  }
+\end{frame}
+
+We can then see a number of extra-logical features of modern proof
+assistants as ways to cope with these problems.
+
+Instead of writing our complete development in one unique file and let
+the program check it when done, we can split it in several files or
+modules and establish dependencies among them: only those who changed
+(and their dependencies) have to be rechecked.
+
+But this separation serves two different purpose, and the fact that
+they coincide is pure contingency: the purpose of splitting a
+development into logical units, and the purpose of providing a coarse
+notion of dependency to ease the non-linear process of proof editing
+by accelerating recompilation upon changes.
+
+Secondly, the traditional interaction loop makes unnecessary the
+recompilation upon local changes by providing a global undo feature of
+the whole state of the proof assistant. But if I want to change this
+proof, I have to backtrack all the way here (because the green region
+is locked), do my changes and then replay the whole proofs in between,
+even if my change had no impact on this one, even if I just added a
+comment.
+
+The actual text of the proofs for versioning is often
+non-satisfactory: on one side it contains mostly non-relevant
+informations (line breaks, spaces\ldots), on the other the relevant
+ones, the validity of the proof, the dependency between objects, is
+absent.
+
+So both these facilities, among others, seem to provide ad-hoc ways to
+deal with the non-linearity of the script and its edition. They
+provide a coarse approximation of the dependency amongst objects. But
+this notion of dependency \emph{is} a metatheoretical property of a
+logic, and it could be used to provide an \emph{exact} way of dealing
+with the impact of changes, in particular to guide proof-checking to
+recheck only minimal parts of the developments.
+
+This is what we will try to do in the following.
+
+\begin{frame}{Methodology}
+  \begin{center}
+    \only<1-2>{\fbox{\parbox{20em}{\only<2>\alert{version
+            management}\\[1em]%
+          \fbox{\parbox{18em}{script files\\[1em]%
+              \fbox{\parbox{17em}{parsing\\[1em]%
+                  \fbox{\parbox{14em}{proof-checking\\} }}}}}}}}%
+    \only<3>{\fbox{\parbox{20em}{script files\\[1em]%
+          \fbox{\parbox{18em}{\alert{version management}\\[1em]%
+              \fbox{\parbox{17em}{parsing\\[1em]%
+                  \fbox{\parbox{14em}{proof-checking\\} }}}}}}}}%
+    \only<4->{\fbox{\parbox{20em}{\alt<-5>{script files}{user
+            interaction}\\[1em]%
+          \fbox{\parbox{18em}{parsing\\[1em]%
+              \fbox{\parbox{17em}{\alert{version management
+                    \only<7->{+ proof-checking}}\\[1em]%
+                  \uncover<4-6>{\fbox{\parbox{14.5em}{proof-checking\\}}}}}}}}}}%
+  \end{center}
+  \begin{itemize}
+  \item<4-> AST representation
+  \item<5-> Explicit dependency DAG
+  \item<7-> Typing annotations
+  \item<8-> Incremental type-checking
   \end{itemize}
@@
 
-Let me start with two simple observations relating the activity of
-both the working mathematician and the programmer. A first, obvious
-fact is that both spend more time \emph{editing} than \emph{writing}:
-their workflow is highly non-linear, involving experiments, fixes,
-backtrack on previous modifications\ldots, eventually validated or
-invalidated by some criteria like the absence of bugs or the validity
-of a proof.
+Here is a schematic view of the traditional way to manage formal
+developments. The user writes scripts which are version controlled,
+and these scripts are parsed and checked on demand. By pushing version
+management inwards, we abstract from syntax, and more informations
+become available. For instance we can compute the graph of
+dependencies. At this level already, the situation is greatly
+enhanced: syntax, and file-based development, get the place they
+deserve: they are relegated to a tool for user interaction, the actual
+development being stored and manipulated in an abstract form. We could
+even imagine syntax being dynamically switchable, part of user
+preferences!
 
-The second observation is that since the advent of modern proof
-assistants, large bodies of mathematics have been formalized and the
-way people managed these developments is very similar to the one used
-in software development: we write proofs in text buffers, which are
-linear, and launch proof-checking when we are done. To avoid
-re-checking the whole development, we split it into mutually dependent
-files and re-check only those who have changed. To keep track of old
-versions, we use line-based version control systems that have no
-awareness of the actual content of the files. The separation into
-multiple files then serves two different purpose, and the fact that
-they coincide is pure contingency: the purpose of splitting a
-development into logical units, and the purpose of easing the
-non-linear process of proof editing. Also, recording the actual text
-of the proofs for versioning is often non-satisfactory: on one side it
-contains mostly non-relevant informations (line breaks, spaces\ldots),
-on the other the relevant ones, the validity of the proof, the
-dependency between objects, is absent.
+But let's go one step further and embed version management into the
+type-checker. We gain access to type informations at the level of
+changes, that is, we can track changes, dependencies and types at the
+same time, and decide to type-check only local changes of the
+development and their global effect, in a minimal way. We can perform
+\emph{incremental type-checking}.
 
-The paradox in this situation is that these very tools can be used to
-reason about formal languages and especially proof languages. Couldn't
-we adapt the methods developped to reason about a formal language to
-the proof language of proof assistants itself? Could we replace this
-legacy toolchain and make it language-aware?
-
-\begin{frame}{Expectations}
-
+\begin{frame}{A core meta-language for incremental type-checking}
+  \begin{columns}[t]
+    \begin{column}{0.5\textwidth}
+      \begin{block}{Expresses}
+        \begin{itemize}
+        \item (abstract) Syntax
+        \item (object-) Logics
+        \item Proofs (-terms)
+        \end{itemize}
+      \end{block}
+    \end{column}
+    \begin{column}{0.5\textwidth}
+      \begin{block}{Features}
+        \begin{itemize}
+        \item Typing
+        \item Incrementality
+        \end{itemize}
+      \end{block}
+    \end{column}
+  \end{columns}
+  \vspace{2em}
+  \begin{center}
+    \emph{A kernel for a typed version control system?}
+  \end{center}
 \end{frame}
 
-\begin{frame}{\mode<beamer>{Outline}}
-  \tableofcontents
-\end{frame}
-
-\section{}
-
-\subsection{Practical applications}
-
-\section{A core meta-language for proof repositories}
+In the following, I will present a preliminary implementation of these
+ideas. It is a meta-language for incremental proof-checking, that is a
+language in which one can define abstract syntaxes, logical rules and
+proofs, and allows to check validity, and express the incrementality
+of the constructions. It can be viewed as a kernel for a typed version
+control system.
 
@@
   \end{itemize}
+
+  Let's design a \emph{typed} \textsf{git} to do incremental type-checking
 \end{frame}

### 2010-07-06 talks/MIPS2010/slides.tex
2e slide corrigé
git-svn-id: http://gasp.dev.regis-gianas.com/svn/trunk@141 403e655e-980a-47bd-8c60-1ee5348bb6b7

@@
   \pause
-  \begin{block}{Mission}
-    \large \it Isn't it time to adapt these tools to proofs?
+  \begin{block}{\ldots\ And yet,}
+    Powerful tools to mechanize the metatheory of (proof) languages
   \end{block}
-  {\Large }
+  \vspace{0.6em}
+  \pause
+  \begin{center}
+    {\large \it Isn't it time to make these tools metatheory-aware?}
+  \end{center}
 \end{frame}
@@
 The paradox about this situation is that these very tools can be used
-to reason about formal languages and especially proof
-languages: some of these tools are even dedicated to the formalization
-of proof systems and their properties. But we still use the legacy
-generic tools to manage our developments.
+to reason about formal languages and especially proof languages: some
+of these tools are even dedicated to the mechanization of the
+metatheory of proof languages. But we still use the legacy generic
+tools to manage our developments.
 
@@
 proofs, and allows to check validity, and express the incrementality
-of the constructions. It can be viewed as a kernel for a typed version
-control system.
+of the constructions. It can be viewed as a kernel for a generic,
+typed version control system.
 

### 2010-07-06 talks/MIPS2010/slides.tex
Git inspiration...
git-svn-id: http://gasp.dev.regis-gianas.com/svn/trunk@142 403e655e-980a-47bd-8c60-1ee5348bb6b7

@@
 loop may be a satisfactory approximation when developing software, the
-additional demands of proof-checking makes it unsatisfactory as the
-development grows: often the time of recompilation is prohibitive.
+extra computational cost of proof-checking makes it unsatisfactory as
+the development grows: often the time of recompilation is prohibitive.
 
@@
   \begin{center}
-    \only<1-3>{\includegraphics[width=0.75\textwidth]{images/coqide.png}}%
+    \only<1>{\includegraphics[width=0.75\textwidth]{images/coqide00.png}}%
+    \only<2>{\includegraphics[width=0.75\textwidth]{images/coqide0.png}}%
+    \only<3>{\includegraphics[width=0.75\textwidth]{images/coqide.png}}%
     \only<4>{\includegraphics[width=0.75\textwidth]{images/coqide2.png}}%
@@
 \begin{frame}{Inspiration: the \textsf{git} storage model}
-  \begin{itemize}
-  \item ``Content-adressable''
-  \item Name reflects content
-  \item Maximal sharing (or hashconsing)
-  \end{itemize}
+  \begin{center}
+    \begin{overlayarea}{\textwidth}{12em}
+      \only<2>{\includegraphics[width=0.85\textwidth]{images/git1.pdf}}%
+      \only<3>{\includegraphics[width=0.85\textwidth]{images/git2.pdf}}%
+      \only<4>{\includegraphics[width=0.85\textwidth]{images/git3.pdf}}%
+      \only<5>{\includegraphics[width=0.85\textwidth]{images/git4.pdf}}%
+      \only<6-9>{\includegraphics[width=0.85\textwidth]{images/git5.pdf}}%
+      \only<10>{
+        \begin{center}
+          \vspace{5em}
+          {\Large Let's design a \emph{typed} \textsf{git}}
+        \end{center}
+      }%
+    \end{overlayarea}
+  \end{center}%
+  \begin{itemize}\small %
+  \item<7-> ``Content-adressable''%
+  \item<8-> Name reflects content%
+  \item<9-> Maximal sharing (or hash-consing)%
+  \end{itemize}%
+\end{frame}
 
-  Let's design a \emph{typed} \textsf{git} to do incremental type-checking
-\end{frame}
+\textsf{Git} is one of the most successful version control system, and
+part of its success is due to the very simple, yet very efficient
+storage model. It doesn't keep track of changes \emph{per se}, but of
+similarities between version. Given a base directory structure,
+already committed in its database, and a new version of it with, let's
+say, one file changed -- \texttt{f3.txt}, it stores the new version by
+linking all common subdirectory structure to the previous version.
+This is possible because every node in the tree, directories and
+files, have a unique name. So actually, the database of \textsf{git}
+looks like this.
+
+Moreover, similarities are easily tracked because every name is
+carefully chosen to reflect the \emph{content} of a file or directory.
+In practice, \texttt{git} uses a hash function to produce all names in
+the database: files are given the hash of their content, and
+directories are given the hash of the hash of all the files it
+contains, recursively.
+
+This way, we ensure that no two subdirectories with different content
+appear in the database, that is, we enforce maximal sharing among
+subdirectories. In functional programming, this is called
+hash-consing.
 

### 2010-07-07 talks/MIPS2010/slides.tex
Première passe sur l'explication de la syntaxe et des règles de typage
git-svn-id: http://gasp.dev.regis-gianas.com/svn/trunk@143 403e655e-980a-47bd-8c60-1ee5348bb6b7

@@
 Let me start with a simple observation relating the activity of both
-the working mathematician and the computer scientist. Whereas both
-activities are percieved, from the outside, as apodictic because the
-final object of their study, the proof or the program, seems to be
+the working mathematician and the computer scientist. Both activities
+could be percieved, from the outside, as apodictic because the final
+object of their study, the proof or the program, seems to be
 unquestionably valid (or not).
@@
 management inwards, we abstract from syntax, and more informations
-become available. For instance we can compute the graph of
+become available. For instance we can compute the acyclic graph of
 dependencies. At this level already, the situation is greatly
@@
         \item Incrementality
+        \item Dependency
         \end{itemize}
@@
 language in which one can define abstract syntaxes, logical rules and
-proofs, and allows to check validity, and express the incrementality
-of the constructions. It can be viewed as a kernel for a generic,
-typed version control system.
+proofs, and allows to check validity, express the incrementality of
+the constructions, and their dependency. It can be viewed as a kernel
+for a generic, typed version control system.
 
@@
 \textsf{Git} is one of the most successful version control system, and
-part of its success is due to the very simple, yet very efficient
+part of its success is due to the very simple, yet very powerful
 storage model. It doesn't keep track of changes \emph{per se}, but of
@@
 files, have a unique name. So actually, the database of \textsf{git}
-looks like this.
+looks like this: it associates to every name, a content (for files) or
+a list of names (for directories). The invariant is that this database
+forms a DAG. This naming policy makes all content adressable.
 
@@
 carefully chosen to reflect the \emph{content} of a file or directory.
-In practice, \texttt{git} uses a hash function to produce all names in
+In practice, \textsf{git} uses a hash function to produce all names in
 the database: files are given the hash of their content, and
@@
 
-\begin{frame}{Incremental type-checking}
-  A small language for representing proofs:
-  \only<1>{\[ t\ ::=\ (x:t)\cdot t\ |\ x\ |\ t\ t\ |\ *\] \[\]}%
-  \only<2>{\[ t\ ::=\ (x:t)\cdot t\ |\ \alert x\ |\ \alert{t\ t}\ |\ *\] \[\]}%
-  \only<3>{\[ t\ ::=\ (x:t)\cdot t\ |\ \alert a\ |\ *\]%
-    \[ a\ ::=\ x\ |\ a\ x \]} 
-  \only<4>{\[ t\ ::=\ (x:t)\cdot t\ |\ a\
-    |\ *\ |\ \alert{(x = a)\cdot t}\]%
-    \[ a\ ::=\ x\ |\ a\ x \]}
-  \begin{example}
-    \[ \infer[Lam]{\Gamma,A\vdash B}{\Gamma\vdash A\to B} \]
-    \[ \textsc{Lam}\ :\ (\Gamma : env)\cdot(A B : type)\cdot
-    (d : \Gamma,A\vdash B)\cdot \Gamma\vdash A\to B \]
-  \end{example}
+\begin{frame}{A typed repository}
+  % exemple
+\end{frame}
+
+on va plus loin grace a l'approche LF où tous les objets ont un type,
+et font partie de l'arbre de syntaxe. Donc tout est hash-consé
+(règles logiques, syntaxe, preuves).
+
+\begin{frame}{Incremental type-checking, incrementally}%
+  \begin{block}{Syntax}%
+    \only<1>{\[ t\ ::=\ [x:t]\cdot t\ |\ (x:t)\cdot t\ |\ x\ |\ t\ t\ %
+      |\ *\] \[\]}%
+    \only<2>{\[ t\ ::=\ [x:t]\cdot t\ |\ (x:t)\cdot t\ |\ \alert x\ |\ %
+      \alert{t\ t}\ |\ *\] \[\]}%
+    \only<3>{\[ t\ ::=\ [x:t]\cdot t\ |\ (x:t)\cdot t\ |\ \alert a\ |\ %
+      *\]%
+      \[ a\ ::=\ x\ |\ a\ x \]}%
+    \only<4>{\[ t\ ::=\ [x:t]\cdot t\ |\ (x:t)\cdot t\ |\ a\ |\ *\ |\ %
+      \alert{(x = a)\cdot t}\]%
+      \[ a\ ::=\ x\ |\ a\ x \]}%
+    \only<5>{\[ t\ ::=\ \alert{[x:t]\cdot t}\ |\ (x:t)\cdot t\ |\ a\ %
+      |\ *\ |\ (x = a)\cdot t\]%
+      \[ a\ ::=\ x\ |\ a\ x \]}%
+    \only<6->{\[ t\ ::=\ (x:t)\cdot t\ |\ a\ |\ *\ |\ (x = a)\cdot t\]%
+      \[ a\ ::=\ x\ |\ a\ x \]}%
+  \end{block}
+  \begin{block}{Semantics}
+      \[ \Gamma\ ::=\ \cdot\ |\ \Gamma [x:t]\only<4>{\ |\
+        \alert{\Gamma[x=a:t]}}
+      \only<5->{\ |\ \Gamma[x=a:t]}\] 
+      \vspace{1em}
+      \begin{center}
+        \begin{overlayarea}{0.8\textwidth}{5em}
+          \only<-6>{ $ \Gamma \vdash t : u $ \hfill `` In environment
+            $\Gamma$, term $t$ has type $u$ '' } \only<7>{ $ \Gamma
+            \vdash t : u \alert{\Rightarrow \Delta} $ \hfill ``
+            From \alert{repository} $\Gamma$, term $t$ of type $u$ \alert{leads to
+            the new repository $\Delta$}'' }
+        \end{overlayarea}
+      \end{center}
+  \end{block}
+\end{frame}
+
+\begin{frame}{Typing rules}
+  \begin{block}{Product - Init}
+  \begin{mathpar}
+    \large
+    \infer                      % prod
+    {\Gamma\vdash t : *\uncover<2->{ \Rightarrow \_} \and \Gamma[x:t]\vdash u :
+      *\uncover<2->{\Rightarrow\Delta}}
+    {\Gamma\vdash(x:t)\cdot u : *\uncover<2->{\Rightarrow\Delta}}
+    \and
+    \uncover<3->{                               %init
+      \infer{ }
+      {\Gamma\vdash x : t \uncover<4->{\Rightarrow\Gamma}}
+      \quad {\small [x:t]\in\Gamma}
+    }
+  \end{mathpar}
+  \end{block}
+\end{frame}
+
+\begin{frame}{Typing rules}
+  \begin{block}{Equality binder}
+    \begin{mathpar}
+      \large
+      \infer
+      {\Gamma\vdash a : u \Rightarrow \_ \and \Gamma[x=a:u]\vdash t :
+        *\Rightarrow\Delta }
+      {\Gamma\vdash (x=a)\cdot t : * \Rightarrow\Delta}
+      \quad {\small [y=a:u] \notin\Gamma}
+      \and \pause
+      \infer
+      {\Gamma\vdash t \{y/x\} : * \Rightarrow\Delta}
+      {\Gamma\vdash(x=a)\cdot t : * \Rightarrow\Delta}
+      \quad {\small [y=a:u] \in\Gamma}
+    \end{mathpar}
+  \end{block}
+  \pause
+  \begin{block}{Application}
+    \[ \infer
+      {\Gamma\vdash a : (y:u)\cdot t \Rightarrow\Delta}
+      {\Gamma\vdash a\ x : t \{x/y\} \Rightarrow\Delta}
+      \quad [x : u] \in \Gamma
+      \]
+  \end{block}
+\end{frame}
+
+\begin{frame}{Naming}
+
+\end{frame}
+
+environnement : nom -> contenu * type
+
+\begin{frame}{Conclusion}
+
 \end{frame}

### 2010-07-07 talks/MIPS2010/slides.tex
Texte handout pour le slide de syntaxe
git-svn-id: http://gasp.dev.regis-gianas.com/svn/trunk@144 403e655e-980a-47bd-8c60-1ee5348bb6b7

@@
 
+\hrule
 \begin{frame}{The impact of changes}
@@
 \end{frame}
+\hrule
 
@@
 
+\hrule
 \begin{frame}{Methodology}
@@
 
+\hrule
 \begin{frame}{A core meta-language for incremental type-checking}
@@
 
+\hrule
 \begin{frame}{Inspiration: the \textsf{git} storage model}
@@
 
+\hrule
 \begin{frame}{A typed repository}
@@
 
+\hrule
 \begin{frame}{Incremental type-checking, incrementally}%
@@
   \end{block}
-  \begin{block}{Semantics}
+  \begin{block}{Typing}
       \[ \Gamma\ ::=\ \cdot\ |\ \Gamma [x:t]\only<4>{\ |\
@@
       \only<5->{\ |\ \Gamma[x=a:t]}\] 
-      \vspace{1em}
+      \vspace{0em}
       \begin{center}
@@
 \end{frame}
+\hrule
 
+Let's construct a typed language for this. We start from the usual
+Pure Type Systems here, and we will modify it gradually to fit our
+needs.
+
+First, we want to name all intermediate results, that is we don't want
+to allow compound, deep applications. We remove general application
+and make it a list of variables. Then, we need a new binder to
+remember these flat applications. It will behave like a \textsf{let}
+binding, and we add a corresponding entry to the environment. Note
+that in terms, the type is not present whereas it is in the
+environment.
+
+We will use this syntax in quite an unusual way as we will write and
+check only \emph{types}. Terms will be sequences of binders declaring
+and defining inter-depentent pieces of syntax, and what finishes them
+(* or $a$) does not matter in most cases. For the moment, we remove
+the abstraction, and with it all notion of computation.
+
+These simple modifications allow us to see this system as a typed
+repository of proofs with a last modification: the usual judgement
+becomes ``From repository Γ, term t of type u leads to repository
+∆''. The analogy here is: Γ is the current database of versions, $t$
+is a working directory with possible redundancies (it's not maximally
+shared), and integration of $t$ into $\Gamma$ leads to $\Delta$, which
+is maximally shared.
+
+We thread the environment along the typing process and return it as
+the result.
+
+\hrule
 \begin{frame}{Typing rules}
@@
 \end{frame}
+\hrule
 
+\hrule
 \begin{frame}{Typing rules}
@@
 \end{frame}
+\hrule
 
+\hrule
 \begin{frame}{Naming}
@@
 
-\begin{frame}{Conclusion}
+\hrule
+\begin{frame}{Further Work}
 
 \end{frame}
+\hrule
 

### 2010-07-08 talks/MIPS2010/slides.tex
* Iteration.
git-svn-id: http://gasp.dev.regis-gianas.com/svn/trunk@145 403e655e-980a-47bd-8c60-1ee5348bb6b7
unicode -> math
git-svn-id: http://gasp.dev.regis-gianas.com/svn/trunk@146 403e655e-980a-47bd-8c60-1ee5348bb6b7

@@
 % ===================================================
-\documentclass[ignorenonframetext,red]{beamer}
-% \documentclass[12pt]{article} \usepackage{beamerarticle} \usepackage{fullpage}
+%\documentclass[ignorenonframetext,red]{beamer}
+\documentclass[12pt]{article} \usepackage{beamerarticle} \usepackage{fullpage}
 % ===================================================
@@
 and make it a list of variables. Then, we need a new binder to
-remember these flat applications. It will behave like a \textsf{let}
-binding, and we add a corresponding entry to the environment. Note
-that in terms, the type is not present whereas it is in the
-environment.
+remember these flat applications and give a name to them. It will
+behave like a \textsf{let} binding, and we add a corresponding entry
+to the environment. Note that in terms, the type is not present
+whereas it is in the environment.
 
@@
 repository of proofs with a last modification: the usual judgement
-becomes ``From repository Γ, term t of type u leads to repository
-∆''. The analogy here is: Γ is the current database of versions, $t$
-is a working directory with possible redundancies (it's not maximally
-shared), and integration of $t$ into $\Gamma$ leads to $\Delta$, which
-is maximally shared.
+becomes ``From repository $\Gamma$, term $t$ of type $u$ leads to
+repository $\Delta$''. The analogy here is: $\Gamma$ is the current
+database of versions, $t$ is a working directory with possible
+redundancies (it's not maximally shared), and integration of $t$ into
+$\Gamma$ leads to $\Delta$, which is maximally shared.
 
@@
       *\uncover<2->{\Rightarrow\Delta}}
-    {\Gamma\vdash(x:t)\cdot u : *\uncover<2->{\Rightarrow\Delta}}
+    {\Gamma\vdash(x:t)\cdot u : *\uncover<2->{\textcolor{gray}{\Rightarrow\Delta}}}
     \and
@@
 \hrule
-\begin{frame}{Naming}
+\begin{frame}{Content-aware names (implementation details)}
+
+\begin{itemize}
+\item Given ``$a$'', how to decide efficiently ``$[y = a : u] \in \Gamma$''?
+
+\item Inspiration from \textsf{Git}~:
+\begin{center}
+\begin{math}
+\begin{array}{lllr}
+| \dots | &:& \kappa^\star \rightarrow \kappa & \text{Hash function}\\
+\equiv    &:& \kappa \rightarrow \kappa \rightarrow \mathbb{B} & \text{Efficient comparison}\\
+\nu       &:& \textsf{unit} \rightarrow \kappa & \text{Fresh key generator}
+\end{array}
+\end{math}
+\end{center}
+\end{itemize}
+
+\begin{mathpar}
+  \large
+  \infer
+      {\Gamma\vdash a : u \Rightarrow \_ \and \Gamma[|a|=a:u]\vdash t\{x/|a|\} :
+        *\Rightarrow\Delta }
+      {\Gamma\vdash (x=a)\cdot t : * \Rightarrow\Delta}
+      \quad {\small [y=a:u] \notin\Gamma}
+   \and
+   \large
+   \infer                      % prod
+       {\Gamma\vdash t : *\uncover<2->{ \Rightarrow \_} \and \Gamma[k:t]\vdash u\{x/k\} :
+         *\uncover<2->{\Rightarrow\Delta}}
+       {\Gamma\vdash(x:t)\cdot u : * \textcolor{gray}{\Rightarrow\Delta}}
+       \quad{\small k=\nu()}
+
+\end{mathpar}
+
+environnement : nom -> contenu * type
 
@@
 
-environnement : nom -> contenu * type
-
 \hrule
@@
 
+\begin{itemize}
+\item What if we reintroduce the lambda?
+\begin{enumerate}
+\item Constructive metatheory.
+\item A language to express patches?
+\end{enumerate}
+\end{itemize}
 \end{frame}

### 2010-07-09 talks/MIPS2010/slides.tex
exemple gasp, qqes détails
git-svn-id: http://gasp.dev.regis-gianas.com/svn/trunk@147 403e655e-980a-47bd-8c60-1ee5348bb6b7
Version finale des slides (?)
git-svn-id: http://gasp.dev.regis-gianas.com/svn/trunk@148 403e655e-980a-47bd-8c60-1ee5348bb6b7

--- 2010-07-07 talks/MIPS2010/slides.tex
@@
 \documentclass[ignorenonframetext,red]{beamer}
-% \documentclass[12pt]{article} \usepackage{beamerarticle} \usepackage{fullpage}
+%\documentclass[12pt]{article} \usepackage{beamerarticle} \usepackage{fullpage}
 % ===================================================
@@
 
-This talk will be about some remarks and directions on the way the
-edition of formal proof is done in proof assistants.
+I am going to present a work-in-progress, realized together with Yann
+Régis-Gianas. It is far from a finished product, so feel free to
+interrupt, comment or criticize. So this talk will be about some
+observations and directions to improve the way we edit and store
+formal proofs in proof assistants.
 
@@
 
-The paradox about this situation is that these very tools can be used
-to reason about formal languages and especially proof languages: some
-of these tools are even dedicated to the mechanization of the
-metatheory of proof languages. But we still use the legacy generic
-tools to manage our developments.
+The paradox about this situation is that these very tools (Coq,
+Matita, Agda) can be used to reason about formal languages and
+especially proof languages, some of these tools are even dedicated to
+the mechanization of the metatheory of proof languages. But we still
+use this legacy generic tools to manage our developments.
 
@@
 the constructions, and their dependency. It can be viewed as a kernel
-for a generic, typed version control system.
+for a generic, typed version control system, and for the first
+iteration of this project, we were inspired by the way \textsf{git}
+stores the history of directories.
 
+\hrule
 \begin{frame}{Inspiration: the \textsf{git} storage model}
@@
 say, one file changed -- \texttt{f3.txt}, it stores the new version by
-linking all common subdirectory structure to the previous version.
-This is possible because every node in the tree, directories and
-files, have a unique name. So actually, the database of \textsf{git}
-looks like this: it associates to every name, a content (for files) or
-a list of names (for directories). The invariant is that this database
-forms a DAG. This naming policy makes all content adressable.
+linking all common subdirectory structure to the previous
+version. This way, similarities are physically relalized by this kind
+of ``compression''. This is possible because every node in the tree,
+directories and files, have a unique name. So actually, the database
+of \textsf{git} looks like this: it associates to every name, a
+content (for files) or a list of names (for directories). The
+invariant is that this database forms a DAG. This naming policy makes
+all content adressable.
 
@@
 
+Let's apply the same discipline to proofs, and validate them by
+typing. What would we get?
+
+\hrule
 \begin{frame}{A typed repository}
-  % exemple
+  \only<1>{\includegraphics[width=\textwidth]{images/gasp1.pdf}}%
+  \only<2>{\includegraphics[width=\textwidth]{images/gasp2.pdf}}%
+  \only<3>{\includegraphics[width=\textwidth]{images/gasp3.pdf}}%
+  \only<4>{\includegraphics[width=\textwidth]{images/gasp4.pdf}}%
+  \only<5>{\includegraphics[width=\textwidth]{images/gasp5.pdf}}%
 \end{frame}
+\hrule
 
-on va plus loin grace a l'approche LF où tous les objets ont un type,
-et font partie de l'arbre de syntaxe. Donc tout est hash-consé
-(règles logiques, syntaxe, preuves).
+On the same model, let's start with a proof of $C$ already committed,
+and starting with this structure. $\pi_1$, $\pi_2$ and $\pi_3$ are the
+rest of the proof, upwards. Each node is labelled with the logical
+rule (or constructor) used, but also annotated by the judgement it
+proves. Now let's imagine that we come up with a new, better proof for
+$B$. Integrating it in the repository means, like before, to share all
+common sub-derivations. But now, we don't need to retype the whole new
+proof, just the new parts (in red). Like before, this is possible if
+we give a name to all sub-derivations, and the database looks like
+this: it associates with every name, how it is formed.
 
+\hrule
 \begin{frame}{Incremental type-checking, incrementally}%
@@
   \end{block}
-  \begin{block}{Semantics}
+  \begin{block}{Environments}
       \[ \Gamma\ ::=\ \cdot\ |\ \Gamma [x:t]\only<4>{\ |\
         \alert{\Gamma[x=a:t]}}
-      \only<5->{\ |\ \Gamma[x=a:t]}\] 
+      \only<5->{\ |\ \Gamma[x=a:t]} \] 
+    \end{block}
+    \begin{block}{Judgement}
       \vspace{1em}
-      \begin{center}
-        \begin{overlayarea}{0.8\textwidth}{5em}
-          \only<-6>{ $ \Gamma \vdash t : u $ \hfill `` In environment
-            $\Gamma$, term $t$ has type $u$ '' } \only<7>{ $ \Gamma
-            \vdash t : u \alert{\Rightarrow \Delta} $ \hfill ``
-            From \alert{repository} $\Gamma$, term $t$ of type $u$ \alert{leads to
-            the new repository $\Delta$}'' }
-        \end{overlayarea}
-      \end{center}
+        \begin{tabular}{ll} \Large
+          $\Gamma\vdash t : u \uncover<7>{\alert{\Rightarrow\Delta}}$
+          &
+          \only<-6>{`` In environment $\Gamma$, term $t$ has type $u$
+            ''}
+          \uncover<7>{`` From
+            \alert{repository} $\Gamma$, term $t$ of type $u$ \\ &
+            \alert{\hspace{1.2em}leads to the new repository $\Delta$}''
+          }
+        \end{tabular}
     \end{block}
 \end{frame}
+\hrule
 
+Let's construct a calculus for this. We start from the usual Pure Type
+Systems, with abstraction, products, variables, applications and just
+one sort for the sake of simplicity. Typing is done in the usual
+name-type environment, by the usual judgement: ``In $\Gamma$, $t$ has
+type $u$. We modify it gradually, keeping its expressivity but
+tailoring the syntax to name all subterms.
+
+First we don't want to allow compound, deep applications. We remove
+general application and make it a list of variables. Then, we need a
+new binder to remember these flat applications and give a name to
+them. It will behave like a \textsf{let} binding, and we add a
+corresponding entry to the environment. Note that in terms, the type
+is not present whereas it is in the environment.
+
+It happens that the system is already interesting enough for our
+purpose without abstraction: the notions of dependency and
+incrementality do not need the abstraction. For the moment, we remove
+it, and with it all notion of computation.
+
+These simple modifications allow us to see this system as a typed
+repository of proofs with a last modification: the usual judgement
+becomes ``From repository $\Gamma$, term $t$ of type $u$ leads to
+repository $\Delta$''. The analogy here is: $\Gamma$ is the current
+repository, $t$ is a piece of proof that has to be checked and
+committed, it can have redundancies (it's not maximally shared), and
+integration of $t$ into $\Gamma$ leads to $\Delta$, which is maximally
+shared.
+
+We thread the environment along the typing process and return it as
+the result.
+
+\newcommand\g[1]{\textcolor{gray}{#1}}
+
+\hrule
 \begin{frame}{Typing rules}
-  \begin{block}{Product - Init}
-  \begin{mathpar}
-    \large
+  \begin{block}{Product}
+    \[ \large
     \infer                      % prod
-    {\Gamma\vdash t : *\uncover<2->{ \Rightarrow \_} \and \Gamma[x:t]\vdash u :
-      *\uncover<2->{\Rightarrow\Delta}}
-    {\Gamma\vdash(x:t)\cdot u : *\uncover<2->{\Rightarrow\Delta}}
-    \and
-    \uncover<3->{                               %init
-      \infer{ }
-      {\Gamma\vdash x : t \uncover<4->{\Rightarrow\Gamma}}
+    {\Gamma\vdash t : *\uncover<2->{ \textcolor{gray}{\ \Rightarrow \_}} \and \Gamma[x:t]\vdash u :
+      *\uncover<2->{\textcolor{gray}{\ \Rightarrow\Delta}}}
+    {\Gamma\vdash(x:t)\cdot u : * \uncover<2->{\textcolor{gray}{\ \Rightarrow\Delta}}}
+    \]
+  \end{block}
+  \pause\pause
+  \begin{block}{Init}
+    \[\infer{ }
+      {\Gamma\vdash x : t \uncover<4->{\textcolor{gray}{\ \Rightarrow\Gamma}}}
       \quad {\small [x:t]\in\Gamma}
-    }
-  \end{mathpar}
+      \]
   \end{block}
 \end{frame}
+\hrule
 
+I will present briefly some of the typing rules of this calculus. Here
+is for example the usual product: we thread the return environment in
+its main branch. Here is the usual init rule, we return the enlarged
+environment.
+
+\hrule
 \begin{frame}{Typing rules}
@@
       \infer
-      {\Gamma\vdash a : u \Rightarrow \_ \and \Gamma[x=a:u]\vdash t :
-        *\Rightarrow\Delta }
-      {\Gamma\vdash (x=a)\cdot t : * \Rightarrow\Delta}
+      {\Gamma\vdash a : u \textcolor{gray}{\ \Rightarrow \_} \and \Gamma[x=a:u]\vdash t :
+        *\textcolor{gray}{\ \Rightarrow\Delta} }
+      {\Gamma\vdash (x=a)\cdot t : * \textcolor{gray}{\ \Rightarrow\Delta}}
       \quad {\small [y=a:u] \notin\Gamma}
@@
       \infer
-      {\Gamma\vdash t \{y/x\} : * \Rightarrow\Delta}
-      {\Gamma\vdash(x=a)\cdot t : * \Rightarrow\Delta}
+      {\Gamma\vdash t \{y/x\} : * \textcolor{gray}{\ \Rightarrow\Delta}}
+      {\Gamma\vdash(x=a)\cdot t : * \textcolor{gray}{\ \Rightarrow\Delta}}
       \quad {\small [y=a:u] \in\Gamma}
@@
     \[ \infer
-      {\Gamma\vdash a : (y:u)\cdot t \Rightarrow\Delta}
-      {\Gamma\vdash a\ x : t \{x/y\} \Rightarrow\Delta}
+      {\Gamma\vdash a : (y:u)\cdot t \textcolor{gray}{\ \Rightarrow\Delta}}
+      {\Gamma\vdash a\ x : t \{x/y\} \textcolor{gray}{\ \Rightarrow\Delta}}
       \quad [x : u] \in \Gamma
@@
 \end{frame}
+\hrule
 
-\begin{frame}{Naming}
+The rules for the equality binder are two. The first resembles a lot
+that of the \textsf{let} binding: we check the body, then the rest by
+putting a new definition in the environment. But it applies only when
+there is \emph{no} other equal definitions with body $a$ in the
+environment. If there is, then the second rule apply.
 
+In this case, we just ignore the first premise, as $a$ was already
+type-checked. This rule is the key to incrementality.
+
+Last rule, the application, is not very different from the usual one,
+except that since there can be only a variable in the argument
+position, we inline the Init rule and get rid of the second premise.
+
+\hrule
+\begin{frame}{Content-aware names (implementation)}
+  Given ``$a$'', how to decide efficiently ``$[y = a : u] \in
+  \Gamma$''?
+  \pause
+  \[ \begin{array}{lllr}
+    | \cdot | &:& \vec\kappa \rightarrow \kappa & \text{Hash function}\\
+    \equiv    &:& \kappa \rightarrow \kappa \rightarrow \mathbb{B} & \text{Efficient comparison}\\
+    \nu       &:& \textsf{unit} \rightarrow \kappa & \text{Fresh key generator}
+  \end{array} \]
+  \pause
+  \begin{mathpar}
+    \infer
+    {\Gamma\vdash a : u \textcolor{gray}{\ \Rightarrow \_} 
+      \and \Gamma[|a|=a:u]\vdash t\{x/|a|\} :
+      *\textcolor{gray}{\ \Rightarrow\Delta} }
+    {\Gamma\vdash (x=a)\cdot t : * \textcolor{gray}{\ \Rightarrow\Delta}}
+    \quad {\small [y=a:u] \notin\Gamma}
+    \and
+    \pause
+    \infer
+    {\Gamma\vdash t : *\uncover<2->{ \Rightarrow \_} \and \Gamma[k:t]\vdash u\{x/k\} :
+      *\uncover<2->{\textcolor{gray}{\ \Rightarrow\Delta}}}
+    {\Gamma\vdash(x:t)\cdot u : * \textcolor{gray}{\textcolor{gray}{\ \Rightarrow\Delta}}}
+    \quad{\small k=\nu()}
+  \end{mathpar}
+  \pause
+  \Large
+  \[ \Gamma\ :\ \kappa \to \vec\kappa * \tau \]
 \end{frame}
+\hrule
 
-environnement : nom -> contenu * type
+Let's delve a moment into a particular implementation detail: how to
+choose names. We saw a quite unusual query to the environment before:
+look for all variables with body exactly $a$. How to decide this query
+efficiently?
 
-\begin{frame}{Conclusion}
+We inspire once more from \textsf{git}, and give \emph{internal}
+names, or keys, to all objects, that reflect their contents. We
+introduce a morally injective hash function, from lists of keys to
+keys, an efficient comparison function between two hashes, and a fresh
+key generator.
 
+Here are the introduction rules of the system: When we introduce a
+definition $a$, we choose the hash of $a$ as a name for it, and do the
+substitution in the rest of the term. When we introduce a declaration
+(a product), we choose a new key for its name, and do the
+substitution.
+
+Therefore, the environment is then a function from a key to the
+content of its definition (if it's a definition) and its type.
+
+But let me finish with a remark: all of this is very static, content
+is determined exactly by its syntactic shape. There seems to lack a
+notion of computation into the calculus.
+
+\hrule
+\begin{frame}{Further Work}
+  \Large
+What if we reintroduce {\Huge $[x:t]\cdot t$} ?
+\begin{enumerate} \normalsize
+\pause
+\item Constructive metatheory
+\pause
+\item A language to express patches?
+\end{enumerate}
 \end{frame}
+\hrule
+
+What would happen if we reintroduce computation, and in particular the
+abstraction? We could not only declare logics but also develop its
+metatheory, as we do in Pure Type Systems. For example, the weakening
+property would have an existence, and we could even compute it since
+it's constructively defined. In a sense, there is hope to relax the
+very strict notion of equality implicit in the calculus, from
+syntactic equality to ``semantic'' equality.
+
+Moreover, we would be able to express functions that transform the
+syntax, that is \emph{patches}. This opens a set of research
+directions : are they valid with respect to a logic? what can we say
+about dependencies between patches? Their commutation etc.
 

### 2010-07-09 talks/MIPS2010/slides.tex
bug corrigé ds les illustrations, qqes fixes du handout
git-svn-id: http://gasp.dev.regis-gianas.com/svn/trunk@149 403e655e-980a-47bd-8c60-1ee5348bb6b7

@@
 
-The actual text of the proofs for versioning is often
-non-satisfactory: on one side it contains mostly non-relevant
-informations (line breaks, spaces\ldots), on the other the relevant
-ones, the validity of the proof, the dependency between objects, is
-absent.
-
-So both these facilities, among others, seem to provide ad-hoc ways to
-deal with the non-linearity of the script and its edition. They
-provide a coarse approximation of the dependency amongst objects. But
-this notion of dependency \emph{is} a metatheoretical property of a
-logic, and it could be used to provide an \emph{exact} way of dealing
-with the impact of changes, in particular to guide proof-checking to
-recheck only minimal parts of the developments.
+So both these facilities provide a coarse approximation of the
+dependency between objects. But this notion of dependency \emph{is} a
+metatheoretical property of a logic, and it could be used to provide
+an \emph{exact} way of dealing with the impact of changes, in
+particular to guide proof-checking to recheck only minimal parts of
+the developments.
 
@@
 
-This way, we ensure that no two subdirectories with different content
+This way, we ensure that no two subdirectories with the same content
 appear in the database, that is, we enforce maximal sharing among
@@
   \begin{block}{Syntax}%
-    \only<1>{\[ t\ ::=\ [x:t]\cdot t\ |\ (x:t)\cdot t\ |\ x\ |\ t\ t\ %
+    \begin{overlayarea}{\textwidth}{3.5em}
+      \only<1>{\[ t\ ::=\ [x:t]\cdot t\ |\ (x:t)\cdot t\ |\ x\ |\ t\
+        t\ %
         |\ *\] \[\]}%
-    \only<2>{\[ t\ ::=\ [x:t]\cdot t\ |\ (x:t)\cdot t\ |\ \alert x\ |\ %
+      \only<2>{\[ t\ ::=\ [x:t]\cdot t\ |\ (x:t)\cdot t\ |\ \alert x\
+        |\ %
         \alert{t\ t}\ |\ *\] \[\]}%
-    \only<3>{\[ t\ ::=\ [x:t]\cdot t\ |\ (x:t)\cdot t\ |\ \alert a\ |\ %
+      \only<3>{\[ t\ ::=\ [x:t]\cdot t\ |\ (x:t)\cdot t\ |\ \alert a\
+        |\ %
         *\]%
         \[ a\ ::=\ x\ |\ a\ x \]}%
-    \only<4>{\[ t\ ::=\ [x:t]\cdot t\ |\ (x:t)\cdot t\ |\ a\ |\ *\ |\ %
+      \only<4>{\[ t\ ::=\ [x:t]\cdot t\ |\ (x:t)\cdot t\ |\ a\ |\ *\
+        |\ %
         \alert{(x = a)\cdot t}\]%
         \[ a\ ::=\ x\ |\ a\ x \]}%
-    \only<5>{\[ t\ ::=\ \alert{[x:t]\cdot t}\ |\ (x:t)\cdot t\ |\ a\ %
+      \only<5>{\[ t\ ::=\ \alert{[x:t]\cdot t}\ |\ (x:t)\cdot t\ |\
+        a\ %
         |\ *\ |\ (x = a)\cdot t\]%
         \[ a\ ::=\ x\ |\ a\ x \]}%
-    \only<6->{\[ t\ ::=\ (x:t)\cdot t\ |\ a\ |\ *\ |\ (x = a)\cdot t\]%
+      \only<6->{\[ t\ ::=\ (x:t)\cdot t\ |\ a\ |\ *\ |\ (x = a)\cdot
+        t\]%
         \[ a\ ::=\ x\ |\ a\ x \]}%
+    \end{overlayarea}
   \end{block}
@@
 name-type environment, by the usual judgement: ``In $\Gamma$, $t$ has
-type $u$. We modify it gradually, keeping its expressivity but
+type $u$''. We modify it gradually, keeping its expressivity but
 tailoring the syntax to name all subterms.
@@
 
-It happens that the system is already interesting enough for our
-purpose without abstraction: the notions of dependency and
-incrementality do not need the abstraction. For the moment, we remove
-it, and with it all notion of computation.
+To understand how incrementality is handled, we focus for the moment
+on a minimal system. It happens that it is already interesting enough
+without abstraction. For the moment, we remove it, and with it all
+notion of computation.
 

### 2010-07-09 talks/MIPS2010/handout.tex
Ajout du handout ds un fichier séparé
git-svn-id: http://gasp.dev.regis-gianas.com/svn/trunk@150 403e655e-980a-47bd-8c60-1ee5348bb6b7

--- 2010-07-09 talks/MIPS2010/slides.tex
@@
-% Mode d'emploi: Switcher les deux lignes suivantes pour passer du
-% mode présentation au handout
-% ===================================================
-\documentclass[ignorenonframetext,red]{beamer}
-%\documentclass[12pt]{article} \usepackage{beamerarticle} \usepackage{fullpage}
-% ===================================================
-
+\documentclass[12pt]{article}
+\usepackage{fullpage}
 \usepackage{ucs}
@@
 \usepackage[protrusion=true,expansion=true]{microtype}
-\usepackage{setspace}
-\usepackage{graphicx}
 
@@
 
-\author[Matthias Puech \& Yann Régis-Gianas] {
-Matthias Puech\inst{1} \and Yann Régis-Gianas\inst{2} \\
-{\small \url{puech@cs.unibo.it}} \and {\small \url{yrg@pps.jussieu.fr}}
-}
-\institute {
-  \inst 1 {\small Dept. of Computer Science, University of Bologna} \and
-  \inst 2 {\small University Paris 7, CNRS, and INRIA, PPS, team ${\pi}r^2$}
-}
+\author{Matthias Puech \and Yann Régis-Gianas}
 
-\setbeamertemplate{footline}[frame number]
-\setbeamertemplate{navigation symbols}{}
-
-% \AtBeginSection[]
-% {\begin{frame}<beamer>{Outline}
-%     \tableofcontents[currentsection]
-%   \end{frame}
-% }
-
-\usefonttheme{serif}
+\newcommand{\slide}{\vspace{1em} \hrule \vspace{1em}}
 
@@
 
-\begin{frame}
-  \titlepage
-  \mode<article>{
-    \newcommand\url\texttt
-    \maketitle}
-\end{frame}
-
-\section{Motivation}
+\maketitle
 
@@
 
-\hrule
-\begin{frame}{How are \emph{constructed} formal mathematics?}
-  \large
-  \begin{tabular}{ll}
-    {\Huge Q :} & \parbox{0.8\textwidth}{What is the common point between the working
-      mathematician and the working programmer?} \\[2em]
-    \pause
-    {\Huge A :} & They both spend more time \emph{editing} than \emph{writing}
-  \end{tabular}
-\end{frame}
-\hrule
+\slide
 
@@
 
-\hrule
-\begin{frame}{A paradoxical situation}  
-  \begin{block}{Observation}
-    Workflow of formal mathematics was largely inspired by software
-    development:
-    \begin{itemize}
-    \item File-based scripts (\textsf{emacs})
-    \item Separate compilation (\textsf{make})
-    \item Text-based versioning (\textsf{svn}, \textsf{diff}s\ldots)
-    \end{itemize}
-  \end{block}
-  \pause
-  \begin{block}{\ldots\ And yet,}
-    Powerful tools to mechanize the metatheory of (proof) languages
-  \end{block}
-  \vspace{0.6em}
-  \pause
-  \begin{center}
-    {\large \it Isn't it time to make these tools metatheory-aware?}
-  \end{center}
-\end{frame}
-\hrule
+\slide
 
@@
 
-\hrule
-\begin{frame}{Motivations}
-  \begin{block}{Rigidity of linear edition}
-    \begin{itemize}
-    \item ((edit; compile)*; commit)* loop does not scale to proofs
-    \item Concept freeze inhibits the discovery process
-    \item No room for alternate definitions
-    \end{itemize}
-  \end{block}
-  \pause
-  \begin{block}{Laxity of textual representation}
-    \begin{itemize}
-    \item Textual scripts \textsf{diff}s do not reflect the semantics
-    \item Not even the syntax
-    \end{itemize}
-  \end{block}
-  \vspace{2em}
-  \pause
-  {\tiny \ldots\ Maybe it wasn't adapted to software development}
-\end{frame}
-\hrule
+\slide
 
@@
 
-\hrule
-\begin{frame}{The impact of changes}
-  \vspace{0.5em}
-  \begin{center}
-    \only<1>{\includegraphics[width=0.75\textwidth]{images/coqide00.png}}%
-    \only<2>{\includegraphics[width=0.75\textwidth]{images/coqide0.png}}%
-    \only<3>{\includegraphics[width=0.75\textwidth]{images/coqide.png}}%
-    \only<4>{\includegraphics[width=0.75\textwidth]{images/coqide2.png}}%
-    \only<5>{\includegraphics[width=0.75\textwidth]{images/coqide4.png}}%
-    \only<6>{\includegraphics[width=0.75\textwidth]{images/coqide5.png}}%
-    \only<7>{\includegraphics[width=0.75\textwidth]{images/coqide6.png}}%
-    \only<8>{\includegraphics[width=0.75\textwidth]{images/coqide7.png}}%
-  \end{center}
-  \pause
-  \uncover<2->{
-    \begin{itemize}
-    \item File-based separate compilation
-      \pause
-    \item Interaction loop with global undo
-    \end{itemize}
-  }
-\end{frame}
-\hrule
+\slide
 
@@
 
-\hrule
-\begin{frame}{Methodology}
-  \begin{center}
-    \only<1-2>{\fbox{\parbox{20em}{\only<2>\alert{version
-            management}\\[1em]%
-          \fbox{\parbox{18em}{script files\\[1em]%
-              \fbox{\parbox{17em}{parsing\\[1em]%
-                  \fbox{\parbox{14em}{proof-checking\\} }}}}}}}}%
-    \only<3>{\fbox{\parbox{20em}{script files\\[1em]%
-          \fbox{\parbox{18em}{\alert{version management}\\[1em]%
-              \fbox{\parbox{17em}{parsing\\[1em]%
-                  \fbox{\parbox{14em}{proof-checking\\} }}}}}}}}%
-    \only<4->{\fbox{\parbox{20em}{\alt<-5>{script files}{user
-            interaction}\\[1em]%
-          \fbox{\parbox{18em}{parsing\\[1em]%
-              \fbox{\parbox{17em}{\alert{version management
-                    \only<7->{+ proof-checking}}\\[1em]%
-                  \uncover<4-6>{\fbox{\parbox{14.5em}{proof-checking\\}}}}}}}}}}%
-  \end{center}
-  \begin{itemize}
-  \item<4-> AST representation
-  \item<5-> Explicit dependency DAG
-  \item<7-> Typing annotations
-  \item<8-> Incremental type-checking
-  \end{itemize}
-\end{frame}
-\hrule
+\slide
 
@@
 
-\hrule
-\begin{frame}{A core meta-language for incremental type-checking}
-  \begin{columns}[t]
-    \begin{column}{0.5\textwidth}
-      \begin{block}{Expresses}
-        \begin{itemize}
-        \item (abstract) Syntax
-        \item (object-) Logics
-        \item Proofs (-terms)
-        \end{itemize}
-      \end{block}
-    \end{column}
-    \begin{column}{0.5\textwidth}
-      \begin{block}{Features}
-        \begin{itemize}
-        \item Typing
-        \item Incrementality
-        \item Dependency
-        \end{itemize}
-      \end{block}
-    \end{column}
-  \end{columns}
-  \vspace{2em}
-  \begin{center}
-    \emph{A kernel for a typed version control system?}
-  \end{center}
-\end{frame}
-\hrule
+\slide
 
@@
 
-\hrule
-\begin{frame}{Inspiration: the \textsf{git} storage model}
-  \begin{center}
-    \begin{overlayarea}{\textwidth}{12em}
-      \only<2>{\includegraphics[width=0.85\textwidth]{images/git1.pdf}}%
-      \only<3>{\includegraphics[width=0.85\textwidth]{images/git2.pdf}}%
-      \only<4>{\includegraphics[width=0.85\textwidth]{images/git3.pdf}}%
-      \only<5>{\includegraphics[width=0.85\textwidth]{images/git4.pdf}}%
-      \only<6-9>{\includegraphics[width=0.85\textwidth]{images/git5.pdf}}%
-      \only<10>{
-        \begin{center}
-          \vspace{5em}
-          {\Large Let's design a \emph{typed} \textsf{git}}
-        \end{center}
-      }%
-    \end{overlayarea}
-  \end{center}%
-  \begin{itemize}\small %
-  \item<7-> ``Content-adressable''%
-  \item<8-> Name reflects content%
-  \item<9-> Maximal sharing (or hash-consing)%
-  \end{itemize}%
-\end{frame}
-\hrule
+\slide
 
@@
 
-\hrule
-\begin{frame}{A typed repository}
-  \only<1>{\includegraphics[width=\textwidth]{images/gasp1.pdf}}%
-  \only<2>{\includegraphics[width=\textwidth]{images/gasp2.pdf}}%
-  \only<3>{\includegraphics[width=\textwidth]{images/gasp3.pdf}}%
-  \only<4>{\includegraphics[width=\textwidth]{images/gasp4.pdf}}%
-  \only<5>{\includegraphics[width=\textwidth]{images/gasp5.pdf}}%
-\end{frame}
-\hrule
+\slide
 
@@
 
-\hrule
-\begin{frame}{Incremental type-checking, incrementally}%
-  \begin{block}{Syntax}%
-    \begin{overlayarea}{\textwidth}{3.5em}
-      \only<1>{\[ t\ ::=\ [x:t]\cdot t\ |\ (x:t)\cdot t\ |\ x\ |\ t\
-        t\ %
-        |\ *\] \[\]}%
-      \only<2>{\[ t\ ::=\ [x:t]\cdot t\ |\ (x:t)\cdot t\ |\ \alert x\
-        |\ %
-        \alert{t\ t}\ |\ *\] \[\]}%
-      \only<3>{\[ t\ ::=\ [x:t]\cdot t\ |\ (x:t)\cdot t\ |\ \alert a\
-        |\ %
-        *\]%
-        \[ a\ ::=\ x\ |\ a\ x \]}%
-      \only<4>{\[ t\ ::=\ [x:t]\cdot t\ |\ (x:t)\cdot t\ |\ a\ |\ *\
-        |\ %
-        \alert{(x = a)\cdot t}\]%
-        \[ a\ ::=\ x\ |\ a\ x \]}%
-      \only<5>{\[ t\ ::=\ \alert{[x:t]\cdot t}\ |\ (x:t)\cdot t\ |\
-        a\ %
-        |\ *\ |\ (x = a)\cdot t\]%
-        \[ a\ ::=\ x\ |\ a\ x \]}%
-      \only<6->{\[ t\ ::=\ (x:t)\cdot t\ |\ a\ |\ *\ |\ (x = a)\cdot
-        t\]%
-        \[ a\ ::=\ x\ |\ a\ x \]}%
-    \end{overlayarea}
-  \end{block}
-  \begin{block}{Environments}
-      \[ \Gamma\ ::=\ \cdot\ |\ \Gamma [x:t]\only<4>{\ |\
-        \alert{\Gamma[x=a:t]}}
-      \only<5->{\ |\ \Gamma[x=a:t]} \] 
-    \end{block}
-    \begin{block}{Judgement}
-      \vspace{1em}
-        \begin{tabular}{ll} \Large
-          $\Gamma\vdash t : u \uncover<7>{\alert{\Rightarrow\Delta}}$
-          &
-          \only<-6>{`` In environment $\Gamma$, term $t$ has type $u$
-            ''}
-          \uncover<7>{`` From
-            \alert{repository} $\Gamma$, term $t$ of type $u$ \\ &
-            \alert{\hspace{1.2em}leads to the new repository $\Delta$}''
-          }
-        \end{tabular}
-    \end{block}
-\end{frame}
-\hrule
+\slide
 
@@
 
-\newcommand\g[1]{\textcolor{gray}{#1}}
-
-\hrule
-\begin{frame}{Typing rules}
-  \begin{block}{Product}
-    \[ \large
-    \infer                      % prod
-    {\Gamma\vdash t : *\uncover<2->{ \textcolor{gray}{\ \Rightarrow \_}} \and \Gamma[x:t]\vdash u :
-      *\uncover<2->{\textcolor{gray}{\ \Rightarrow\Delta}}}
-    {\Gamma\vdash(x:t)\cdot u : * \uncover<2->{\textcolor{gray}{\ \Rightarrow\Delta}}}
-    \]
-  \end{block}
-  \pause\pause
-  \begin{block}{Init}
-    \[\infer{ }
-      {\Gamma\vdash x : t \uncover<4->{\textcolor{gray}{\ \Rightarrow\Gamma}}}
-      \quad {\small [x:t]\in\Gamma}
-      \]
-  \end{block}
-\end{frame}
-\hrule
+\slide
 
@@
 
-\hrule
-\begin{frame}{Typing rules}
-  \begin{block}{Equality binder}
-    \begin{mathpar}
-      \large
-      \infer
-      {\Gamma\vdash a : u \textcolor{gray}{\ \Rightarrow \_} \and \Gamma[x=a:u]\vdash t :
-        *\textcolor{gray}{\ \Rightarrow\Delta} }
-      {\Gamma\vdash (x=a)\cdot t : * \textcolor{gray}{\ \Rightarrow\Delta}}
-      \quad {\small [y=a:u] \notin\Gamma}
-      \and \pause
-      \infer
-      {\Gamma\vdash t \{y/x\} : * \textcolor{gray}{\ \Rightarrow\Delta}}
-      {\Gamma\vdash(x=a)\cdot t : * \textcolor{gray}{\ \Rightarrow\Delta}}
-      \quad {\small [y=a:u] \in\Gamma}
-    \end{mathpar}
-  \end{block}
-  \pause
-  \begin{block}{Application}
-    \[ \infer
-      {\Gamma\vdash a : (y:u)\cdot t \textcolor{gray}{\ \Rightarrow\Delta}}
-      {\Gamma\vdash a\ x : t \{x/y\} \textcolor{gray}{\ \Rightarrow\Delta}}
-      \quad [x : u] \in \Gamma
-      \]
-  \end{block}
-\end{frame}
-\hrule
+\slide
 
@@
 
-\hrule
-\begin{frame}{Content-aware names (implementation)}
-  Given ``$a$'', how to decide efficiently ``$[y = a : u] \in
-  \Gamma$''?
-  \pause
-  \[ \begin{array}{lllr}
-    | \cdot | &:& \vec\kappa \rightarrow \kappa & \text{Hash function}\\
-    \equiv    &:& \kappa \rightarrow \kappa \rightarrow \mathbb{B} & \text{Efficient comparison}\\
-    \nu       &:& \textsf{unit} \rightarrow \kappa & \text{Fresh key generator}
-  \end{array} \]
-  \pause
-  \begin{mathpar}
-    \infer
-    {\Gamma\vdash a : u \textcolor{gray}{\ \Rightarrow \_} 
-      \and \Gamma[|a|=a:u]\vdash t\{x/|a|\} :
-      *\textcolor{gray}{\ \Rightarrow\Delta} }
-    {\Gamma\vdash (x=a)\cdot t : * \textcolor{gray}{\ \Rightarrow\Delta}}
-    \quad {\small [y=a:u] \notin\Gamma}
-    \and
-    \pause
-    \infer
-    {\Gamma\vdash t : *\uncover<2->{ \Rightarrow \_} \and \Gamma[k:t]\vdash u\{x/k\} :
-      *\uncover<2->{\textcolor{gray}{\ \Rightarrow\Delta}}}
-    {\Gamma\vdash(x:t)\cdot u : * \textcolor{gray}{\textcolor{gray}{\ \Rightarrow\Delta}}}
-    \quad{\small k=\nu()}
-  \end{mathpar}
-  \pause
-  \Large
-  \[ \Gamma\ :\ \kappa \to \vec\kappa * \tau \]
-\end{frame}
-\hrule
+\slide
 
@@
 
-\hrule
-\begin{frame}{Further Work}
-  \Large
-What if we reintroduce {\Huge $[x:t]\cdot t$} ?
-\begin{enumerate} \normalsize
-\pause
-\item Constructive metatheory
-\pause
-\item A language to express patches?
-\end{enumerate}
-\end{frame}
-\hrule
+\slide
 

### 2010-07-09 talks/MIPS2010/slides.tex
Mod images git, ajout mug, dernière mods sur le handout
Cette fois-ci c'est la bonne!

git-svn-id: http://gasp.dev.regis-gianas.com/svn/trunk@151 403e655e-980a-47bd-8c60-1ee5348bb6b7

@@
 
-\section{Motivation}
-
-I am going to present a work-in-progress, realized together with Yann
-Régis-Gianas. It is far from a finished product, so feel free to
-interrupt, comment or criticize. So this talk will be about some
-observations and directions to improve the way we edit and store
-formal proofs in proof assistants.
-
-\hrule
 \begin{frame}{How are \emph{constructed} formal mathematics?}
   \large
+  \vfill
   \begin{tabular}{ll}
@@
     \pause
-    {\Huge A :} & They both spend more time \emph{editing} than \emph{writing}
+    {\Huge A :} & They both spend more time \emph{editing} than
+    \emph{writing}
   \end{tabular}
+  \vfill
+  \begin{center}
+    \includegraphics[width=0.15\textwidth]{images/math-mug.jpg}
+  \end{center}
 \end{frame}
-\hrule
 
-Let me start with a simple observation relating the activity of both
-the working mathematician and the computer scientist. Both activities
-could be percieved, from the outside, as apodictic because the final
-object of their study, the proof or the program, seems to be
-unquestionably valid (or not).
+\section{Motivation}
 
-If we observe the day-to-day work of these scientists though, their
-workflow resembles a lot more that of experimental sciences: it is
-highly non-linear, involving multiple parallel experiments, fixes,
-backtrack on previous modifications\ldots\ eventually validated or
-invalidated by some \emph{a posteriori} criterion, the absence of
-bugs, the validity of a proof.
-
-In other words, the mathematician and the programmer spend more time
-\emph{editing} their developments non-linearly than monotonously
-\emph{writing} new material.
-
-It is the case for programs, for classical, paper mathematics so it
-should be the case for formal mathematics, edited in a proof
-assistant.
-
-\hrule
 \begin{frame}{A paradoxical situation}  
   \begin{block}{Observation}
-    Workflow of formal mathematics was largely inspired by software
-    development:
+    We have powerful tools to mechanize the metatheory of (proof) languages
+  \end{block}
+  \pause
+  \begin{block}{\ldots\ And yet,}
+    Workflow of formal mathematics is still largely inspired by legacy
+    software development:
     \begin{itemize}
@@
     \end{itemize}
-  \end{block}
-  \pause
-  \begin{block}{\ldots\ And yet,}
-    Powerful tools to mechanize the metatheory of (proof) languages
   \end{block}
@@
 \end{frame}
-\hrule
 
-Since the advent of modern proof assistants, large bodies of
-mathematics have been formalized and the way people managed these
-developments is very similar to the one used in software development
-for years: we write proofs in text files, which are linear, and launch
-proof-checking when we are done. To avoid re-checking the whole
-development, we split it into inter-dependent files and re-check only
-those who have changed. To keep track of old versions, we use
-line-based version control systems that have no awareness of the
-actual content of the files. In a sense, the non-linearity of proof
-development is managed outside the scope of the tool, by manual,
-textual transformations that are often very unsatisfactory.
-
-The paradox about this situation is that these very tools (Coq,
-Matita, Agda) can be used to reason about formal languages and
-especially proof languages, some of these tools are even dedicated to
-the mechanization of the metatheory of proof languages. But we still
-use this legacy generic tools to manage our developments.
-
-Couldn't we adapt the methods developed to reason about formal
-languages to the language of the proof assistant itself? Could we
-replace this legacy toolchain and make it language-aware, even
-semantic-aware?
-
-\hrule
 \begin{frame}{Motivations}
@@
 \end{frame}
-\hrule
 
-More precisely, here are some problems that arise with the traditional
-tools when editing proofs. Whereas the usual edit, compile, commit
-loop may be a satisfactory approximation when developing software, the
-extra computational cost of proof-checking makes it unsatisfactory as
-the development grows: often the time of recompilation is prohibitive.
-
-But it is not only a matter of time: once a concept is proof-checked,
-it is frozen and any future modification of it will require to
-re-check a large part of the development, often much more than what is
-actually required. We believe that this fact inhibits the use of proof
-assistants as tools to help the mathematician in the discovery
-process, as an experimentation tool, and not only to formally
-transcribe paper proofs.
-
-The same way, the linearity of proof scripts does not allow any
-flexibility in the alternate definition of concepts.
-
-On the other side, the way we store mathematical developments, that is
-as textual files or textual diffs does not reflect neither their
-syntax nor their semantics, and as I will try to show, there is a lot
-to gain in changing this base representation.
-
-In fact, some will say that this model of interaction wasn't even
-adapted to programming in the first place.
-
-\hrule
 \begin{frame}{The impact of changes}
@@
 \end{frame}
-\hrule
 
-We can then see a number of extra-logical features of modern proof
-assistants as ways to cope with these problems.
-
-Instead of writing our complete development in one unique file and let
-the program check it when done, we can split it in several files or
-modules and establish dependencies among them: only those who changed
-(and their dependencies) have to be rechecked.
-
-But this separation serves two different purpose, and the fact that
-they coincide is pure contingency: the purpose of splitting a
-development into logical units, and the purpose of providing a coarse
-notion of dependency to ease the non-linear process of proof editing
-by accelerating recompilation upon changes.
-
-Secondly, the traditional interaction loop makes unnecessary the
-recompilation upon local changes by providing a global undo feature of
-the whole state of the proof assistant. But if I want to change this
-proof, I have to backtrack all the way here (because the green region
-is locked), do my changes and then replay the whole proofs in between,
-even if my change had no impact on this one, even if I just added a
-comment.
-
-So both these facilities provide a coarse approximation of the
-dependency between objects. But this notion of dependency \emph{is} a
-metatheoretical property of a logic, and it could be used to provide
-an \emph{exact} way of dealing with the impact of changes, in
-particular to guide proof-checking to recheck only minimal parts of
-the developments.
-
-This is what we will try to do in the following.
-
-\hrule
 \begin{frame}{Methodology}
@@
 \end{frame}
-\hrule
 
-Here is a schematic view of the traditional way to manage formal
-developments. The user writes scripts which are version controlled,
-and these scripts are parsed and checked on demand. By pushing version
-management inwards, we abstract from syntax, and more informations
-become available. For instance we can compute the acyclic graph of
-dependencies. At this level already, the situation is greatly
-enhanced: syntax, and file-based development, get the place they
-deserve: they are relegated to a tool for user interaction, the actual
-development being stored and manipulated in an abstract form. We could
-even imagine syntax being dynamically switchable, part of user
-preferences!
-
-But let's go one step further and embed version management into the
-type-checker. We gain access to type informations at the level of
-changes, that is, we can track changes, dependencies and types at the
-same time, and decide to type-check only local changes of the
-development and their global effect, in a minimal way. We can perform
-\emph{incremental type-checking}.
-
-\hrule
 \begin{frame}{A core meta-language for incremental type-checking}
@@
 \end{frame}
-\hrule
 
-In the following, I will present a preliminary implementation of these
-ideas. It is a meta-language for incremental proof-checking, that is a
-language in which one can define abstract syntaxes, logical rules and
-proofs, and allows to check validity, express the incrementality of
-the constructions, and their dependency. It can be viewed as a kernel
-for a generic, typed version control system, and for the first
-iteration of this project, we were inspired by the way \textsf{git}
-stores the history of directories.
-
-\hrule
-\begin{frame}{Inspiration: the \textsf{git} storage model}
+\begin{frame}{A repository of directories}
   \begin{center}
-    \begin{overlayarea}{\textwidth}{12em}
+    \begin{overlayarea}{\textwidth}{14em}
+      \begin{center}
         \only<2>{\includegraphics[width=0.85\textwidth]{images/git1.pdf}}%
@@
           \begin{center}
-          \vspace{5em}
-          {\Large Let's design a \emph{typed} \textsf{git}}
+            \vspace{5em} {\Large Let's do it with \emph{proofs}}
           \end{center}
         }%
+      \end{center}
     \end{overlayarea}
@@
 \end{frame}
-\hrule
 
-\textsf{Git} is one of the most successful version control system, and
-part of its success is due to the very simple, yet very powerful
-storage model. It doesn't keep track of changes \emph{per se}, but of
-similarities between version. Given a base directory structure,
-already committed in its database, and a new version of it with, let's
-say, one file changed -- \texttt{f3.txt}, it stores the new version by
-linking all common subdirectory structure to the previous
-version. This way, similarities are physically relalized by this kind
-of ``compression''. This is possible because every node in the tree,
-directories and files, have a unique name. So actually, the database
-of \textsf{git} looks like this: it associates to every name, a
-content (for files) or a list of names (for directories). The
-invariant is that this database forms a DAG. This naming policy makes
-all content adressable.
-
-Moreover, similarities are easily tracked because every name is
-carefully chosen to reflect the \emph{content} of a file or directory.
-In practice, \textsf{git} uses a hash function to produce all names in
-the database: files are given the hash of their content, and
-directories are given the hash of the hash of all the files it
-contains, recursively.
-
-This way, we ensure that no two subdirectories with the same content
-appear in the database, that is, we enforce maximal sharing among
-subdirectories. In functional programming, this is called
-hash-consing.
-
-Let's apply the same discipline to proofs, and validate them by
-typing. What would we get?
-
-\hrule
-\begin{frame}{A typed repository}
+\begin{frame}{A typed repository of proofs}
   \only<1>{\includegraphics[width=\textwidth]{images/gasp1.pdf}}%
@@
 \end{frame}
-\hrule
 
-On the same model, let's start with a proof of $C$ already committed,
-and starting with this structure. $\pi_1$, $\pi_2$ and $\pi_3$ are the
-rest of the proof, upwards. Each node is labelled with the logical
-rule (or constructor) used, but also annotated by the judgement it
-proves. Now let's imagine that we come up with a new, better proof for
-$B$. Integrating it in the repository means, like before, to share all
-common sub-derivations. But now, we don't need to retype the whole new
-proof, just the new parts (in red). Like before, this is possible if
-we give a name to all sub-derivations, and the database looks like
-this: it associates with every name, how it is formed.
-
-\hrule
 \begin{frame}{Incremental type-checking, incrementally}%
@@
 \end{frame}
-\hrule
 
-Let's construct a calculus for this. We start from the usual Pure Type
-Systems, with abstraction, products, variables, applications and just
-one sort for the sake of simplicity. Typing is done in the usual
-name-type environment, by the usual judgement: ``In $\Gamma$, $t$ has
-type $u$''. We modify it gradually, keeping its expressivity but
-tailoring the syntax to name all subterms.
-
-First we don't want to allow compound, deep applications. We remove
-general application and make it a list of variables. Then, we need a
-new binder to remember these flat applications and give a name to
-them. It will behave like a \textsf{let} binding, and we add a
-corresponding entry to the environment. Note that in terms, the type
-is not present whereas it is in the environment.
-
-To understand how incrementality is handled, we focus for the moment
-on a minimal system. It happens that it is already interesting enough
-without abstraction. For the moment, we remove it, and with it all
-notion of computation.
-
-These simple modifications allow us to see this system as a typed
-repository of proofs with a last modification: the usual judgement
-becomes ``From repository $\Gamma$, term $t$ of type $u$ leads to
-repository $\Delta$''. The analogy here is: $\Gamma$ is the current
-repository, $t$ is a piece of proof that has to be checked and
-committed, it can have redundancies (it's not maximally shared), and
-integration of $t$ into $\Gamma$ leads to $\Delta$, which is maximally
-shared.
-
-We thread the environment along the typing process and return it as
-the result.
-
-\newcommand\g[1]{\textcolor{gray}{#1}}
-
-\hrule
 \begin{frame}{Typing rules}
@@
 \end{frame}
-\hrule
 
-I will present briefly some of the typing rules of this calculus. Here
-is for example the usual product: we thread the return environment in
-its main branch. Here is the usual init rule, we return the enlarged
-environment.
-
-\hrule
 \begin{frame}{Typing rules}
@@
 \end{frame}
-\hrule
 
-The rules for the equality binder are two. The first resembles a lot
-that of the \textsf{let} binding: we check the body, then the rest by
-putting a new definition in the environment. But it applies only when
-there is \emph{no} other equal definitions with body $a$ in the
-environment. If there is, then the second rule apply.
-
-In this case, we just ignore the first premise, as $a$ was already
-type-checked. This rule is the key to incrementality.
-
-Last rule, the application, is not very different from the usual one,
-except that since there can be only a variable in the argument
-position, we inline the Init rule and get rid of the second premise.
-
-\hrule
 \begin{frame}{Content-aware names (implementation)}
@@
 \end{frame}
-\hrule
 
-Let's delve a moment into a particular implementation detail: how to
-choose names. We saw a quite unusual query to the environment before:
-look for all variables with body exactly $a$. How to decide this query
-efficiently?
-
-We inspire once more from \textsf{git}, and give \emph{internal}
-names, or keys, to all objects, that reflect their contents. We
-introduce a morally injective hash function, from lists of keys to
-keys, an efficient comparison function between two hashes, and a fresh
-key generator.
-
-Here are the introduction rules of the system: When we introduce a
-definition $a$, we choose the hash of $a$ as a name for it, and do the
-substitution in the rest of the term. When we introduce a declaration
-(a product), we choose a new key for its name, and do the
-substitution.
-
-Therefore, the environment is then a function from a key to the
-content of its definition (if it's a definition) and its type.
-
-But let me finish with a remark: all of this is very static, content
-is determined exactly by its syntactic shape. There seems to lack a
-notion of computation into the calculus.
-
-\hrule
 \begin{frame}{Further Work}
@@
 \end{frame}
-\hrule
-
-What would happen if we reintroduce computation, and in particular the
-abstraction? We could not only declare logics but also develop its
-metatheory, as we do in Pure Type Systems. For example, the weakening
-property would have an existence, and we could even compute it since
-it's constructively defined. In a sense, there is hope to relax the
-very strict notion of equality implicit in the calculus, from
-syntactic equality to ``semantic'' equality.
-
-Moreover, we would be able to express functions that transform the
-syntax, that is \emph{patches}. This opens a set of research
-directions : are they valid with respect to a logic? what can we say
-about dependencies between patches? Their commutation etc.
 

### 2011-01-14 papers/ICFP2011/paper.tex
ICFP: C'est parti!
git-svn-id: http://gasp.dev.regis-gianas.com/svn/trunk@234 403e655e-980a-47bd-8c60-1ee5348bb6b7

\documentclass[preprint]{sigplanconf}

\usepackage{amsmath}
\usepackage{ucs}
\usepackage[utf8x]{inputenc}
\usepackage{color}
\usepackage[pdftex,backref=page,colorlinks=true]{hyperref}

\definecolor{bwgreen}{rgb}{0.183,1,0.5}
\definecolor{bwmagenta}{rgb}{1,0.169,0.909}
\definecolor{bwblue}{rgb}{0.317,0.161,1}
\hypersetup{
  linkcolor=blue,
  citecolor=blue
}

\newcommand{\remplan}[1]{\noindent\textcolor{bwblue}{$\triangleright$ \textbf{#1}}}
\newcommand{\remtext}[1]{\textcolor{bwgreen}{$\triangleright$ \textsl{#1}}}

\begin{document}

\conferenceinfo{ICFP '11}{June 4--8, Tokyo, Japan} 
\copyrightyear{2010}
\copyrightdata{[to be supplied]}

%\titlebanner{banner above paper title}        % These are ignored unless
%\preprintfooter{short description of paper}   % 'preprint' option specified.

\title{TBD}
%\subtitle{Subtitle Text, if any}

\authorinfo{Matthias Puech}
           {University of Bologna, University Paris Diderot}
           {puech@cs.unibo.it}
\authorinfo{Yann Régis-Gianas}
           {University Paris Diderot, CNRS, and INRIA}
           {yrg@pps.jussieu.fr}

\maketitle

\begin{abstract}
This is the text of the abstract.
\end{abstract}

\category{CR-number}{subcategory}{third-level}

\terms
term1, term2

\keywords
keyword1, keyword2

\section{Introduction}

\remtext{Une intro que j'avais écrit il y a longtemps et que je sais
  pas où mettre. Sûrement à jeter\ldots}

Programs are nowadays faster to write and safer to execute, thanks to
high-level programming concepts and expressive type systems. Still,
one aspect of programming didn't change very much---if not at
all---since the early days of high-level programming: we still write
programs in text files and when we are done, we let the compiler check
and compile them.  Though, every programmer knows that writing a
program longer than a few dozen of lines is a highly non-linear task
involving constant experiments, fixes, rollback on previous
modifications etc., all of those eventually validated by some a
posteriori criterion: execution and ``empirical'' test, and
increasingly type-checking. In a nutshell, programmers spend more time
\emph{editing} than \emph{writing}. Yet, this aspect of their
day-to-day workflow is not, or poorly taken into account in a
developer's toolchain:
\begin{itemize}
\item To avoid recompiling a whole program upon a change, we split
  them into files, and (re-)compile them separatly. Very often yet,
  this separation serve different purpose: abstraction, splitting into
  logical units, namespace management\ldots\ Why should these purposes
  coincide?
\item To keep track of past versions of a program, we use text-based
  version control systems that have no awareness of the actual content
  of the file. In a sense, the non-linearity of the task is managed
  outside the scope of the language, by manual, textual
  transformations. Isn't there a lot to gain in making these tools
  syntax-aware, even semantics-aware?
\item As a result, all the semantic information that could be
  available to the programmer and act as a \emph{guidance tool} to
  write programs (e.g. type informations) are only used by the
  type-checker as a \emph{repressive} measure against programming
  errors (type errors).
\end{itemize}

These observations apply to the usual, wide-spread programming
languages, but even if the current toolchain in use up to now was a
reasonable approximation to the incremental interaction with the
compiler intended by the user, they become \emph{a fortiori} necessary
when it come to the new, demanding languages embedding formal
verification aspects, like proof assistants or rich specification
languages. In these emerging languages, interaction with a type- or
proof-checker is unavoidable due to the difficulty of writing correct
proofs without guidance from the system, and the constant modification
of the source makes necessary to observe incrementally the effect of a
small change on the whole edifice. Moreover, as the complexity of the
object to check increases, we cannot afford anymore to rely on the
usual unit of compilation, the file, a need a finer-grained tool.

We propose to devise here a data structure for repositories of
programs and proofs allowing to take advantage of the incremental
nature of a programmer's workflow.

% The problem (1 page)
% My idea (2 pages)
% The details (5 pages)
% related works (1-2 pages)
%conclusion and further work (0.5 page)

\bibliographystyle{abbrvnat}

\end{document}

### 2011-03-01 papers/ICFP2011/paper.tex
* Quelques notes sur le papier
git-svn-id: http://gasp.dev.regis-gianas.com/svn/trunk@291 403e655e-980a-47bd-8c60-1ee5348bb6b7

@@
 posteriori criterion: execution and ``empirical'' test, and
-increasingly type-checking. In a nutshell, programmers spend more time
-\emph{editing} than \emph{writing}. Yet, this aspect of their
-day-to-day workflow is not, or poorly taken into account in a
-developer's toolchain:
+increasingly complex type-checking and type inference. In a nutshell,
+programmers spend more time \emph{editing} than \emph{writing}. Yet,
+this aspect of their day-to-day workflow is not, or poorly taken into
+account in a developer's toolchain:
 \begin{itemize}
@@
   coincide?
+
+% On propose quelque chose d'encore plus extreme: faire descendre
+% les VCS dans le type-checker. Voir deux paras plus loin.
 \item To keep track of past versions of a program, we use text-based
@@
   syntax-aware, even semantics-aware?
+
+% Je comprends mais je ne sais pas trop ou tu veux en venir.
 \item As a result, all the semantic information that could be
@@
 when it come to the new, demanding languages embedding formal
-verification aspects, like proof assistants or rich specification
-languages. In these emerging languages, interaction with a type- or
-proof-checker is unavoidable due to the difficulty of writing correct
-proofs without guidance from the system, and the constant modification
-of the source makes necessary to observe incrementally the effect of a
-small change on the whole edifice. Moreover, as the complexity of the
-object to check increases, we cannot afford anymore to rely on the
-usual unit of compilation, the file, a need a finer-grained tool.
+verification aspects, like proof assistants or programming languages
+with rich type system. In these emerging languages, interaction with a
+type- or proof-checker is unavoidable due to the difficulty of writing
+correct proofs without guidance from the system, and the constant
+modification of the source makes necessary to observe incrementally
+the effect of a small change on the whole edifice. Moreover, as the
+complexity of the object to check increases, we cannot afford anymore
+to rely on the usual unit of compilation, the file, a need a
+finer-grained tool. 
+% On pourrait parler de l'utilisation de l'incrementalite pour le type-checking
+% parallele et aussi pour le developpement concurrent.
+
+Current text-based version control technology is designed to help
+concurrent and collaborative development but are not used inside
+compilers or proof checkers. There is a good reason for that: a
+representation of changes at the textual level is not structured
+enough to be used easily and efficiently by such symbolic tools. They
+need instead a more structured representation of changes with a clear
+meaning with respect to the language semantics. For instance, if a
+formal argument of a function has been renamed in a source file
+edition, a textual representation would refer to a change in every
+line that contains an occurrence of that formal argument. A proof
+kernel would prefer a higher level description of that edition from
+which it would be simple to deduce that no proof has to be rechecked
+because the old and the new version of the development are
+$\alpha$-equivalent.
+
+% Prochain para: LF. Finir sur les petits problemes de LF.
+
+% Expliciter clairement les contributions du papier:
+% Une extension de LF avec des variables meta + lambda-selecteur-de-boite. (SLF)
+% Un langage pour representer efficacement un ensemble de termes validés. (NLF)
+% Une compilation de SLF vers NLF qui vérifie le bon typage.
+% Un typage incrémental par proof-search dans NLF.
+% La méta-théorie de NLF qui pose des bases pour la formalisation des VCS sémantiques.
+% Un prototype.
 
@@
 
-% The problem (1 page)
-% My idea (2 pages)
-% The details (5 pages)
-% related works (1-2 pages)
-%conclusion and further work (0.5 page)
+% The problem (1 page) et illustration sur un exemple utilisant SLF.
 
+% My idea (2 pages) - un peu vague.
+
+% Presentation de NLF.
+
+% SLF vers NLF
+
+% Metathéorie de NLF
+
+% Applications (typage incrémental, quelques premières pistes pour commit/merge)
+
+% Related works (1-2 pages) (à ne surtout pas négliger!)
+
+% Conclusion and further work (0.5 page)
+% Dans le future work, en vrac:
+%    - un meta-langage au dessus de NLF = un langage de tactique typé pour Coq?
+%    - des algorithmes de parsing incrémentaux
+%    - modélisation des conflits et de leur résolution dirigée par des types.
+%    - intégration dans Coq pour la parallélisation du noyau. 
+%   -  certification du checker pour être intégré dans le noyau de Coq 
 \bibliographystyle{abbrvnat}

### 2011-03-13 doc/papers/MIPS2010/ui.tex
Moved coq/ papers/ talks/ to doc/. Removed old implementation in experiments/
git-svn-id: http://gasp.dev.regis-gianas.com/svn/trunk@341 403e655e-980a-47bd-8c60-1ee5348bb6b7

\section{Development through semantic repositories}
\label{devel}

\remplan{Fri Jun 18, 2010 10:50 AM. L'objectif ici, c'est de décrire
  une (possible) architecture pour rendre utilisable le langage 
  décrit dans la section précédente. En gros, on énumère les 
  opérations qui interviennent dans l'utilisation et on les décrit
  informellement à l'aide d'un exemple.}

The kernel language described in the previous section is clearly not
adapted to human interaction, for several reasons~: 
\begin{itemize}

\item Every single object is flattened, which makes unobvious the
  structure of a program or a proof.

\item A term of this language denotes the entire history of a
  development. Most of the time, the user is more interested in a
  instantaneous snapshot of his object of study.

\item Even though we were able to provide a snapshot of a particular
  state of the development, the repository represents this state 
  using proof terms, which contain too many syntactic details from the
  user point of view and are not suitable for edition.

\end{itemize}

\remtext{Fri Jun 18, 2010 10:52 AM. Par contre, je ne sais pas quel
  exemple utilisé. Le STLC? Est-ce que ça fait ``assez'' système
  logique pour ce workshop?}

\subsection{A layered architecture}

\remtext{Fri Jun 18, 2010  1:32 PM. Ici, on décrit la figure en suivant
  le modèle de flot du système. On décrit le rôle des fonctions, 
les problèmes posés.}

% À mettre dans une figure. 
\begin{figure}
\begin{center}
% FIXME: il y a un problème avec la flèche double phi/psi.
\xymatrix{
\textrm{History of metatheory objects}
&R 
\ar@{->}[d]_{\phi}
& \Delta \ar@{->}[r]^{\theta(R)} & R' \\
\textrm{Metatheory objects} 
&t \ar@{->}[d]^{\pi}
\ar@{->}[u]_{\psi}
& t' \ar@{->}[u]_{\nabla(t)}\\
\textrm{Concrete views on metatheory objects} 
&w \ar@{.>}[r] & w' \ar@{->}[u]_{\sigma(t, w)}
}
\end{center}
\label{fig:architecture}
\caption{A layered architecture for a semantic repository framework.}
\end{figure}

With the previous remarks in mind, we designed a generic framework of semantic
repositories that can be described using three layers which are summarized in 
Figure~\ref{fig:architecture}. We now describe the role of each layer.

The most internal layer focused on the history of metatheory
objects. It is composed of a type-checker for the kernel language. The
purpose of this layer is to maintain a repository~$R$ and the
invariant that it is well-typed. Through the layer interface, a client
can ask for the \textit{integration} $\theta(R)$ of a patch $\Delta$
to obtain an extended version $R'$ of the repository $R$.

The middle layer is a mechanized formalization of a metatheory. This
module provides a syntax for the objects of the metatheory (formulae,
judgments, \ldots) as well as the admissible rules (that we also call
metatheorems).  An \textit{internalization} function $\phi$ translates
the constructors of these objects and the specification of the
admissible rules as terms of the kernel language. As a consequence,
every object~$t$ of the theory can be injected into the
repository~$R$. An \textit{externalization} function $\psi$ is
implementing the inverse operation. Given a modified version $t'$ of
an object~$t$, we are able to compute the patch $\Delta$ using
the \textit{differentiation} function $\nabla(t)$.

The external layer offers views on the objects of the theory. A
view~$w$ is a human readable, and editable, partial representation of
an object~$t$. A view is obtained from an object through a
projection~$\pi$.  The user transforms the view $w$ into a view $w'$
which must be $\textit{put back}$ into the metatheory as a modified
object $t'$ using a $\textit{putback}$ function $\sigma(t, w)$.  In
concrete syntax, an object~$t$ may be denoted simply using its
syntactic definition but also, more interestingly, by a reference to this 
object into the repository.

The implementation of this architecture raises some interesting
technical difficulties that can be handled, but only partially, 
by existing technology. Indeed, even if internalization and 
externalization are easily implemented thanks to the expressivity
of the kernel language, the projection, putback, differentiation
and integration operations are more intricate to realize.

\paragraph{From an abstract proof tree to a concrete view, and back}
\-
\remtext{Fri Jun 18, 2010  1:36 PM. Un arbre de preuves, ce n'est
  pas très pratique. Pourtant, c'est moralement sur cet objet 
  que s'opère le développement et non sur le code source, qui
  n'en est qu'une représentation. On justifie la nécessité de
  projection et on fait référence aux lentilles de Pierce pour
  expliquer $\sigma$ et $\pi$.}

The problem of updating some data according to a modification that was
done on a partial view of it is a well-known issue in database
literature called the \textit{view-update} problem. Pierce \textit{et
  al} have recently attacked it from a language design approach. Their
idea is to develop projection and putback functions simultaneously by
writing a so-called lens in a bidirectional programming
language~\cite{pierce-bidirectional}. Roughly speaking, a lens is a
software component that can be used to produce output from input but
also input from output (hence the bidirectionality). A lens is able
to keep track of the information that was erased by a projection to
produce a projected output from the initial input. When we supply a
modified output to the lens, it can elaborate, from the initial input
and under certain conditions, the missing part of the modified input.

We plan to design a bidirectional language to define the
projection~function~$\pi$ and the putback~function~$\sigma$ as a
lens. Yet, we cannot just import existing lenses combinators from the
Harmony system that already works on trees. Indeed, most of the time,
our trees have extra structural constraints induced by the presence of
binders.

\remtext{Fri Jun 18, 2010 1:46 PM. Autre point important sur les vues
  concrètes~: elles peuvent servir à parler de l'historique.  Voir le
  module MetaAST de l'implémentation. Ces vues sont utiles par exemple
  lorsque l'utilisateur doit construire un patch d'adaptation. Il a
  alors toute la puissance des métathéorèmes ainsi que tous les objets
  déjà construits pour décrire son patch.}

\paragraph{Capturing the evolution of abstract proof trees}
\-

Computing the minimal edition script between two abstract syntax trees
is a NP-hard problem that has been heavily studied, in particular in
bio-informatics.  We guess that, in practice, we can restrict the
search to local transformations that preserve substructures (that
roughly consist in inserting, removing or moving around entire
sub-trees). In that case, an existing polynomial algorithm may be
used~\cite{Touzet05alinear}. Besides, abstract trees are relatively small in
practice.

We are more concerned about the fact that existing algorithms are not
aware of binders that occur in proof-terms. Indeed, some standard
edition scripts must be forbidden because they could break
well-scopedness. As a consequence, we will have to adapt standard
algorithms to only look for scope-preserving edition scripts.

%% We are more concerned about our ability to elaborate high-level patches
%% from these low-level edition scripts. Actually, it is easy to deduce
%% from a low-level edition script, a low-level patch, based on object
%% syntax constructor elimination and introduction. However, such a
%% low-level patch may forbid a more efficient reuse of existing
%% objects. For instance, let us assume a transformation of an
%% object-level term~$x + x$ into~ he term~$2 + 2$. A low-level patch
%% would represents the transformation as the replacement of the first
%% sub-term $x$ by the sub-term~$2$ followed by the replacement of the
%% second sub-term $x$ by the sub-term~$2$. If we had an object-level type
%% derivation for $x + x$, the integration function would have to build
%% two sub-derivations, one for each replaced sub-terms.  Instead, a
%% high-level patch would represent that transformation as a substitution
%% of the free occurrences of~$x$ by the term $2$ in the term~$x + x$. In
%% that case, the integration function has the opportunity to apply a
%% metatheorem about the stability of typing derivations under
%% substitution and finally reuse the initial typing derivation
%% directly. Inferring automatically such a high-level patch seems an
%% unreachable objective~: there is clearly no most general high-level
%% patch. Nonetheless, an interactive process can be designed to help the
%% user building his high-level patch from a set of pre-computed
%% interpretations of the low-level patch.

\remtext{Fri Jun 18, 2010 1:38 PM. Il faut différencier deux arbres de
  syntaxe. Une méthode : la distance d'édition syntaxique permet de
  déduire des applications des règles d'élimination et d'introduction
  que l'on peut utiliser pour construire $\Delta$. Référence sur la
  littérature sur les distances d'édition entre arbres. Discussion sur
  l'élaboration de patchs de haut niveau (comme le renommage etc). }

\paragraph{Propagating changes through dependencies}
\-

There is several ways of integrating a patch $\Delta$ inside a
repository~$R$.  The naive attempt would be a simple concatenation
of~$\Delta$ at the end of~$R$.  This kind of integration, though
possible and correct, is of little interest because it does not take
into account the fact that $\Delta$ probably defines new versions of
objects already present in~$R$.

A more useful integration propagates changes to the context that uses
the modified object in order to trigger the updating of this context
(which may trigger the updating of other tree, and so forth). Again,
in full generality, it is not possible to update the context
automatically. Instead, we choose to generate \textit{challenges},
which are adaptation patches with holes. These challenges can be
addressed by the author that wrote the patch to be integrated, by the
author of the context or automatically when there is no ambiguity
(think of a context that uses a function whose body has changed but
not its type).

\remtext{Fri Jun 18, 2010 1:42 PM. Si on se contente de concaténer
  $\Delta$ à $R$, on n'a intégré qu'une nouvelle version de l'objet
  initial~$t$ dans $R$ mais on n'a pas propagé ces modifications
  auprès des clients de $t$. Cette propagation n'est \textit{a priori}
  pas triviale~: rien ne dit que $t'$ à un sens équivalent à $t$. Il
  faut donc adapter les clients en demandant un patch d'adaptation à
  l'utilisateur. On présente ces demandes sous la forme de patch à
  compléter (on fournit un peu de contexte et on a un trou pour y
  injecter une nouvelle sous-dérivation mais on peut aussi modifier le
  contexte pour l'adapter). La construction d'un patch d'adaptation
  est un processus interactif~: on ne sait pas a priori borner le
  contexte nécessaire à la construction du patchs d'adaptation donc on
  peut imaginer un sorte de jeu où le système fait augmenter le
  contexte (en suivant les dépendances) si l'utilisateur ne sait pas
  construire le patch d'adaptation. En passant, quand on parle
  d'utilisateur ici, ce peut être aussi bien un programme, comme un
  type-checker pour le langage objet, ou bien un programmeur.}

\subsection{Application on simply typed $\lambda$-calculus}

### 2011-03-13 doc/talks/MIPS2010/handout.tex
Moved coq/ papers/ talks/ to doc/. Removed old implementation in experiments/
git-svn-id: http://gasp.dev.regis-gianas.com/svn/trunk@341 403e655e-980a-47bd-8c60-1ee5348bb6b7

--- 2010-07-09 talks/MIPS2010/handout.tex
@@
 
-I am going to present a work-in-progress, realized together with Yann
-Régis-Gianas. It is far from a finished product, so feel free to
-interrupt, comment or criticize. So this talk will be about some
-observations and directions to improve the way we edit and store
-formal proofs in proof assistants.
-
-\slide
-
-Let me start with a simple observation relating the activity of both
-the working mathematician and the computer scientist. Both activities
-could be percieved, from the outside, as apodictic because the final
-object of their study, the proof or the program, seems to be
-unquestionably valid (or not).
-
-If we observe the day-to-day work of these scientists though, their
-workflow resembles a lot more that of experimental sciences: it is
-highly non-linear, involving multiple parallel experiments, fixes,
-backtrack on previous modifications\ldots\ eventually validated or
-invalidated by some \emph{a posteriori} criterion, the absence of
-bugs, the validity of a proof.
+Let me start with a simple observation on the activity of both the
+working mathematician and the computer scientist. If we observe the
+day-to-day work of these scientists, their workflow resembles a lot
+that of researchers in experimental sciences: it is highly non-linear,
+involving multiple parallel experiments, fixes, backtrack on previous
+modifications\ldots\ eventually validated or invalidated by some
+\emph{a posteriori} criterion, the absence of bugs, the validity of a
+proof.
 
@@
 
-It is the case for programs, for classical, paper mathematics so it
-should be the case for formal mathematics, edited in a proof
-assistant.
+I am going to present a work-in-progress, realized together with Yann
+Régis-Gianas. It is far from a finished product, so feel free to
+interrupt, comment or criticize. It will be about some observations
+and directions to improve the way we edit and store formal proofs in
+proof assistants.
 
@@
 
-Since the advent of modern proof assistants, large bodies of
-mathematics have been formalized and the way people managed these
-developments is very similar to the one used in software development
-for years: we write proofs in text files, which are linear, and launch
+We now have very capable tools available to reason about formal
+languages and especially proof languages. Some of these tools are even
+dedicated to the mechanization of the metatheory of proof languages.
+
+But still, the way people managed large development of formal proofs
+is very similar to the one used in software development for years: we
+write proofs in text files, which are linear, and launch
 proof-checking when we are done. To avoid re-checking the whole
-development, we split it into inter-dependent files and re-check only
-those who have changed. To keep track of old versions, we use
-line-based version control systems that have no awareness of the
-actual content of the files. In a sense, the non-linearity of proof
-development is managed outside the scope of the tool, by manual,
-textual transformations that are often very unsatisfactory.
+development, we split it into files. To keep track of old versions, we
+use line-based version control systems that have no awareness of the
+actual content of the files.
 
-The paradox about this situation is that these very tools (Coq,
-Matita, Agda) can be used to reason about formal languages and
-especially proof languages, some of these tools are even dedicated to
-the mechanization of the metatheory of proof languages. But we still
-use this legacy generic tools to manage our developments.
+In a sense, the non-linearity of proof development is managed outside
+the scope of the tool, by manual, textual transformations that are
+often very unsatisfactory.
 
@@
 
-Instead of writing our complete development in one unique file and let
-the program check it when done, we can split it in several files or
-modules and establish dependencies among them: only those who changed
-(and their dependencies) have to be rechecked.
+Instead of writing our complete development in one unique file, we can
+split it in several files or modules and establish dependencies among
+them: only those who changed (and their dependencies) have to be
+rechecked upon a change.
 
@@
 they coincide is pure contingency: the purpose of splitting a
-development into logical units, and the purpose of providing a coarse
-notion of dependency to ease the non-linear process of proof editing
-by accelerating recompilation upon changes.
+development into logical units, and the purpose of accelerating
+recompilation upon changes.
 
-Secondly, the traditional interaction loop makes unnecessary the
-recompilation upon local changes by providing a global undo feature of
-the whole state of the proof assistant. But if I want to change this
-proof, I have to backtrack all the way here (because the green region
-is locked), do my changes and then replay the whole proofs in between,
-even if my change had no impact on this one, even if I just added a
-comment.
+Secondly, the traditional interaction loop facilitates local changes
+by providing a global undo feature of the whole state of the tool. But
+if I want to change this proof, I have to backtrack all the way here
+(because the green region is locked), do my changes and then replay
+the whole proofs in between, even if my change had no impact on this
+one, even if I just added a comment.
 
@@
 the constructions, and their dependency. It can be viewed as a kernel
-for a generic, typed version control system, and for the first
-iteration of this project, we were inspired by the way \textsf{git}
-stores the history of directories.
+for a generic, typed version control system.
 
@@
 
-\textsf{Git} is one of the most successful version control system, and
-part of its success is due to the very simple, yet very powerful
-storage model. It doesn't keep track of changes \emph{per se}, but of
-similarities between version. Given a base directory structure,
-already committed in its database, and a new version of it with, let's
-say, one file changed -- \texttt{f3.txt}, it stores the new version by
-linking all common subdirectory structure to the previous
-version. This way, similarities are physically relalized by this kind
-of ``compression''. This is possible because every node in the tree,
-directories and files, have a unique name. So actually, the database
-of \textsf{git} looks like this: it associates to every name, a
-content (for files) or a list of names (for directories). The
-invariant is that this database forms a DAG. This naming policy makes
-all content adressable.
+As an inspiration, let's see a very simple, common technique for
+storing history of directories. We won't keep track of changes
+\emph{per se}, but of similarities between version. Given a base
+directory structure, already committed in its database, and a new
+version of it with, let's say, one file changed -- \texttt{f3.txt}, we
+stores the new version by linking all common subdirectory structure to
+the previous version. This way, similarities are physically relalized
+by this kind of ``compression''. This is possible if we give to every
+node in the tree, directories and files, a unique name. So actually,
+our database looks like this: it associates to every name, a content
+(for files) or a list of names (for directories). The invariant is
+that this database forms a DAG. This naming policy makes all content
+adressable.
 
-Moreover, similarities are easily tracked because every name is
-carefully chosen to reflect the \emph{content} of a file or directory.
-In practice, \textsf{git} uses a hash function to produce all names in
-the database: files are given the hash of their content, and
-directories are given the hash of the hash of all the files it
-contains, recursively.
+Moreover, similarities are easily tracked if every name is carefully
+chosen to reflect the \emph{content} of a file or directory.  In
+practice, a hash function is used to produce all names in the
+database: files are given the hash of their content, and directories
+are given the hash of the hash of all the files it contains,
+recursively.
 
@@
 hash-consing.
+
+This is exactly the storage model of \textsf{git}, a popular version
+control system.
 

### 2011-03-13 experiments/alagit/tests/gasp-interface/config.sh

@@ -1,3 +1,0 @@
-set -e # If a test fails in a sequence of tests, 
-       # we abort the entire sequence.
-GASP=../../gasp.native

### 2011-03-13 experiments/alagit/tests/gasp-interface/test1.sh

@@ -1,11 +1,0 @@
-#!/usr/bin/env bash
-
-source config.sh
-echo '* Create a repository.'
-$GASP --do init
-echo '* Require a "just-save" integration of "foo.module" in the repository.'
-$GASP --do commit foo foo.module save 
-echo '* Again, just save the "foo2.module" in the repository.'
-$GASP --do commit foo2 foo2.module save
-echo '* Now, we check out this saved module, again, just as raw syntax.'
-$GASP --do checkout foo2 foo3.raw-fragment

### 2011-03-13 tools/latexpp/whizzy.sh

@@ -1,1 +1,0 @@
-export MKSLICE="./latexpp.opt -w -e ocaml ocaml-sf"

### 2011-10-07 doc/papers/TLDI2012/abstract.tex
started TLDI abstract

\documentclass[9pt,authoryear]{sigplanconf}

\usepackage{ntheorem}
\usepackage{amsmath}
\usepackage{amssymb}
\newtheorem{theorem}{Theorem}
\newtheorem{definition}{Definition}
\usepackage{mathpartir}
\usepackage{macros}
\usepackage{breqn}

\begin{document}

\conferenceinfo{TLDI '12}{January 28 2012, Philadelphia, USA.}
\copyrightyear{2012} 
\copyrightdata{[to be supplied]}

\titlebanner{banner above paper title}        % These are ignored unless
\preprintfooter{short description of paper}   % 'preprint' option specified.

\title{Incremental Type Checking}
%\subtitle{Subtitle Text, if any}

\authorinfo{Name1}
           {Affiliation1}
           {Email1}
\authorinfo{Name2\and Name3}
           {Affiliation2/3}
           {Email2/3}

\maketitle

\begin{abstract}
  We study the problem of verifying the well-typing of terms, not in a
  batch fashion, as it is usually the case for typed languages, but
  incrementally, that is by sequentially modifying a term, and
  re-verifying each time only a smaller amount of information.
\end{abstract}

\category{CR-number}{subcategory}{third-level}

\terms
term1, term2

\keywords
keyword1, keyword2

\section{Introduction}

As developments grow and type systems become more involved, writing a
correct program in one shot becomes quite difficult. On the other
side, writing a program in many correct steps can be tedious because
generally, the verification tools rechecks the whole development at
each step. This second option is the usual practice when the time for
verification is neglectible wrt. the time for making the change, but
this is less and less the case, especially when the language in
question embed proof aspects, and verification involves proof
search. Some mechanisms exist already to cope with the incrementality
of proofs or program development: separate compilation, interactive
toplevel with undo, tactic languages; they all provide in different
ways a rough approximation of the process of modifying and checking
incrementally a large term.

We propose here an architecture for a generic incremental
type-checker, a data structure for repositories of typed proofs and a
language for describing proof deltas. It is based on the simple idea
of sharing common subterms to avoid recheckings and exploits the idea
of encoding a derivation in a metalanguage. This way, given a
signature declaring the typing rules and an (untrusted) typing
algorithm for my language of choice, I get an incremental type-checker
for that language. The metalanguage approach gives us the ability to
encode all the aforementioned usual incrementality mechanisms, and
more.

\section{Building up the metalanguage}

As a first example, let us consider a purposedly simplistic sorted
language of boolean and arithmetic expressions:
$$ e, e' \gequal n \gor e + e' \gor e \land e' \gor e \leq e' $$
The algorithm to determine in a batch fashion wether the term
$$ e_1 = (1 + 3 \leq 2 + 4) \land (8 \leq 3) $$
is well-sorted is trivial (we don't care about its evaluation here,
just well-sortedness). But what if I change subterm $2+4$ in $e_1$
into $7 \leq 2+4$, to obtain $e_2$? Clearly, it should be verified
that context $7\leq []$ is well-sorted (it is), that $2+4$ ``fits''
into its hole (it does), that the whole thing ``fits'' into its new
context $(1+3\leq [])\land(8\leq 3)$ (it is not); but the other,
unchanged subterms need not be. For that, we would have to
``remember'' the state of the verifier in some way.

If only we had \emph{names} (memory addresses, hashes) for
sufficiently enough subterms of our initial term, $$ e_1 =
(\overbrace{1 + 3}^\mmeta \leq \overbrace{2+4}^\mmmeta) \land
\overbrace{(8 \leq 3)}^\mmmmeta\text{\ ,} $$ we could express
concisely the change as a \emph{delta} $$\delta_1 = (\mmeta\leq(7\leq
\mmmeta))\land \mmmmeta \text{\ ,}$$ sharing unchanged subterms. If
only we had \emph{annotated} our initial term with the states of the
verifier,
$$e_1 = (\overbrace{1 + 3}^{\mmeta:\nat} \leq \overbrace{2 + 4}^{\mmmeta:\nat}) \land
\overbrace{(8 \leq 3)}^{\mmmmeta:\bool}\text{\ ,} $$ we would have a
simple process to verify $e_2$ taking advantage of $e_1$'s derivation,
in $O(|\delta_2|)$: verify $\delta_1$ as a term, retrieving the sort
of names from a stored map. This suggests a data structure for a
\emph{repository} of named, annotated, verified subexpressions: a map
from names, or \emph{metavariables}, to terms and types, together with
a \emph{head} metavariable identifying the top of the term: $$\mr
\gequal X, \Delta \quad\text{where}\quad \Delta : (\mmeta\mapsto M :
A) \text{\ .}$$

Terms $M$ should be written in a language allowing to encode our expressions with metavariables,
and types $A$ should contain the whole state of the batch verifier
(here the sort). One modular choice for this metalanguage is \emph{LF}
\cite{harper1993framework}: it allows to specify syntax and rules of
an object language as a \emph{signature} $\Sigma$, and check terms
against this signature. We'll use an increasing fragment of it. The
so-called \emph{intrinsic} style of LF signature for our language
could be:
\begin{align*}
  tp &: \type,\quad nat : tp,\quad bool : tp,\quad exp : tp\to nat,\\
  atom &: \nat\to exp\ nat,\\
  plus &: exp\ nat \to exp\ nat \to exp\ nat,\\
  and &: exp\ bool\to exp\ bool\to exp\ bool,\\
  leq &: exp\ nat\to exp\ nat\to exp\ bool
\end{align*}
In this style, the encoding of an expression and its (object) type is
a term in the metalanguage, but appears in the type of this
expression. The repository associated with expression $e_1$ is
\begin{align*}
  T,
  \begin{array}{rl}
  X &\mapsto plus\ 1\ 3 : exp\ nat \\
  Y &\mapsto plus\ 2\ 4 : exp\ nat \\
  Z &\mapsto leq\ 8\ 3 : exp\ bool \\
  T &\mapsto and\ (leq\ X\ Y)\ Z : exp\ bool \\
\end{array}
\end{align*}

\section{Expressivity}

\section{Architecture}

The \emph{kernel} is the component in charge of verifying terms
against a signature and a repository, and updating this repository. It
supports two basic operations:
\begin{itemize}
\item $\pfunction{push}{\Sigma}{\mr,M}$ checks $M$ against $\Sigma$
  in $\mr$, synthetizes its type $A$, chooses a fresh metavariable
  $\mmeta$ for $M$ and returns $\mr[\mmeta\mapsto M:A]$ and $\mmeta$.
\item $\pfunction{pull}\Sigma{\mr, \mmeta}$ returns the term $M$
  associated with $\mmeta$ in $\mr$ recursively: all metavariables are
  unfolded to their definitions.
\end{itemize}

The \emph{slicer} is the component in charge of slicing a term $M$
into many terms for the kernel to push in the repository.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\bibliographystyle{abbrvnat}
\bibliography{../../../../../Documents/These/biblio.bib}

% \begin{thebibliography}{}
% \softraggedright
% \bibitem[Smith et~al.(2009)Smith, Jones]{smith02}
% P. Q. Smith, and X. Y. Jones. ...reference text...
% \end{thebibliography}

\end{document}

### 2011-10-09 doc/papers/TLDI2012/abstract.tex
TLDI abstract, suite

@@
 \usepackage{macros}
-\usepackage{breqn}
 
@@
 
-\authorinfo{Name1}
-           {Affiliation1}
-           {Email1}
-\authorinfo{Name2\and Name3}
-           {Affiliation2/3}
-           {Email2/3}
+\authorinfo{Matthias Puech}
+           {University of Bologna, University Paris Diderot}
+           {puech@cs.unibo.it}
+\authorinfo{Yann R\'egis-Gianas}
+           {University Paris Diderot, CNRS, and INRIA}
+           {yrg@pps.jussieu.fr}
 
@@
 
-As developments grow and type systems become more involved, writing a
+As programs grow and type systems become more involved, writing a
 correct program in one shot becomes quite difficult. On the other
-side, writing a program in many correct steps can be tedious because
-generally, the verification tools rechecks the whole development at
+hand, writing a program in many correct steps can be tedious because
+generally, the verification tool rechecks the entire development at
 each step. This second option is the usual practice when the time for
-verification is neglectible wrt. the time for making the change, but
-this is less and less the case, especially when the language in
-question embed proof aspects, and verification involves proof
-search. Some mechanisms exist already to cope with the incrementality
-of proofs or program development: separate compilation, interactive
-toplevel with undo, tactic languages; they all provide in different
-ways a rough approximation of the process of modifying and checking
-incrementally a large term.
+verification is negligible with respect to the time to make the
+change, but this is less and less the case, especially when the
+language in question contains proof aspects, and verification involves
+proof search. Some mechanisms already exist to cope with the
+incrementality of proofs or program development: separate compilation,
+interactive toplevel with undo, tactic languages; they all provide in
+different ways a rough approximation of the process of modifying and
+checking incrementally a large term.
 
@@
 language for describing proof deltas. It is based on the simple idea
-of sharing common subterms to avoid recheckings and exploits the idea
-of encoding a derivation in a metalanguage. This way, given a
-signature declaring the typing rules and an (untrusted) typing
-algorithm for my language of choice, I get an incremental type-checker
-for that language. The metalanguage approach gives us the ability to
-encode all the aforementioned usual incrementality mechanisms, and
-more.
+of sharing common subterms to avoid rechecking and exploits encoding a
+derivation in a metalanguage. This way, given a signature declaring
+the typing rules and an (untrusted) typing algorithm for my language
+of choice, I get an incremental type-checker for that language. The
+metalanguage approach gives us the ability to encode all the
+aforementioned usual incrementality mechanisms, and more, making our
+system akin to a typed version control system.
 
-\section{Building up the metalanguage}
+\section{Sharing-based incrementality}
 
@@
 The algorithm to determine in a batch fashion wether the term
-$$ e_1 = (1 + 3 \leq 2 + 4) \land (8 \leq 3) $$
-is well-sorted is trivial (we don't care about its evaluation here,
-just well-sortedness). But what if I change subterm $2+4$ in $e_1$
+$$ e_1 = (1 + 3 \leq 2 + 4) \land (8 \leq 3) $$ is well-sorted is
+trivial (we don't care about its evaluation here, just its
+well-sortedness). But what if I then change subterm $2+4$ in $e_1$
 into $7 \leq 2+4$, to obtain $e_2$? Clearly, it should be verified
 that context $7\leq []$ is well-sorted (it is), that $2+4$ ``fits''
-into its hole (it does), that the whole thing ``fits'' into its new
-context $(1+3\leq [])\land(8\leq 3)$ (it is not); but the other,
-unchanged subterms need not be. For that, we would have to
-``remember'' the state of the verifier in some way.
+into its hole (it does), that the whole expression ``fits'' into its
+new context $(1+3\leq [])\land(8\leq 3)$ (it does not); but the other,
+unchanged subterms need not be verified again. To achieve this
+incremental verification, the system would have to ``remember'' the
+states of the verifier in some way.
 
-If only we had \emph{names} (memory addresses, hashes) for
-sufficiently enough subterms of our initial term, $$ e_1 =
-(\overbrace{1 + 3}^\mmeta \leq \overbrace{2+4}^\mmmeta) \land
-\overbrace{(8 \leq 3)}^\mmmmeta\text{\ ,} $$ we could express
-concisely the change as a \emph{delta} $$\delta_1 = (\mmeta\leq(7\leq
-\mmmeta))\land \mmmmeta \text{\ ,}$$ sharing unchanged subterms. If
-only we had \emph{annotated} our initial term with the states of the
+If only we had \emph{names} (memory addresses, hashes) for enough
+subterms of our initial term, $$ e_1 = (\overbrace{1 + 3}^\mmeta \leq
+\overbrace{2+4}^\mmmeta) \land \overbrace{(8 \leq 3)}^\mmmmeta\text{\
+  ,} $$ we could express concisely the change as a
+\emph{delta} $$\delta_1 = (\mmeta\leq(7\leq \mmmeta))\land \mmmmeta
+\text{\ ,}$$ using these names to refer to unchanged subterms. If only
+we had \emph{annotated} our initial term with the states of the
 verifier,
-$$e_1 = (\overbrace{1 + 3}^{\mmeta:\nat} \leq \overbrace{2 + 4}^{\mmmeta:\nat}) \land
-\overbrace{(8 \leq 3)}^{\mmmmeta:\bool}\text{\ ,} $$ we would have a
-simple process to verify $e_2$ taking advantage of $e_1$'s derivation,
-in $O(|\delta_2|)$: verify $\delta_1$ as a term, retrieving the sort
-of names from a stored map. This suggests a data structure for a
-\emph{repository} of named, annotated, verified subexpressions: a map
-from names, or \emph{metavariables}, to terms and types, together with
-a \emph{head} metavariable identifying the top of the term: $$\mr
-\gequal X, \Delta \quad\text{where}\quad \Delta : (\mmeta\mapsto M :
-A) \text{\ .}$$
+$$e_1 = (\overbrace{1 + 3}^{\mmeta:\nat} \leq \overbrace{2 +
+  4}^{\mmmeta:\nat}) \land \overbrace{(8 \leq
+  3)}^{\mmmmeta:\bool}\text{\ ,} $$ we would have a simple process to
+verify $e_2$ taking advantage of $e_1$'s derivation, in
+$O(|\delta_2|)$: verify $\delta_1$ as a term, retrieving the sort of
+names from a stored map. This suggests a data structure for a
+\emph{repository} of named, annotated, verified subexpressions: a
+monotonously growing map, from names, or \emph{metavariables}, to terms
+and types, together with a \emph{head} metavariable identifying the
+top of the expression: $$\mr \gequal X, \Delta \quad\text{where}\quad
+\Delta : (\mmeta\mapsto M : A) \text{\ .}$$
 
-Terms $M$ should be written in a language allowing to encode our expressions with metavariables,
-and types $A$ should contain the whole state of the batch verifier
-(here the sort). One modular choice for this metalanguage is \emph{LF}
-\cite{harper1993framework}: it allows to specify syntax and rules of
-an object language as a \emph{signature} $\Sigma$, and check terms
-against this signature. We'll use an increasing fragment of it. The
-so-called \emph{intrinsic} style of LF signature for our language
-could be:
+\section{The metalanguage}
+
+What language are terms $M$ and types $A$ written into? Terms should
+encode our expressions with metavariables, and types $A$ should encode
+the whole state of the batch verifier (here the sort). An obvious
+choice is to take $M\gequal(e \text{ with metavariables})$ and $A\gequal
+\bool\gor\nat$, but switching to another language, we'd have to
+redefine another repository language. Another more modular choice for
+this is the \emph{metalanguage} \emph{LF} \cite{harper1993framework}: it allows
+to specify syntax and rules of an object language as a
+\emph{signature} $\Sigma$, and check terms against this signature with
+a generic algorithm. We'll use an increasing fragment of it. The
+so-called \emph{intrinsic} style of LF signature for our expression
+language is:
 \begin{align*}
@@
 \end{align*}
-In this style, the encoding of an expression and its (object) type is
-a term in the metalanguage, but appears in the type of this
-expression. The repository associated with expression $e_1$ is
+In this style, both the encoding of an expression and its sort are
+\emph{terms} in the metalanguage, but the sort appears in the
+\emph{type} of the encoded expression. As an example, the repository
+associated with expression $e_1$ is
 \begin{align*}
-  T,
-  \begin{array}{rl}
-  X &\mapsto plus\ 1\ 3 : exp\ nat \\
-  Y &\mapsto plus\ 2\ 4 : exp\ nat \\
-  Z &\mapsto leq\ 8\ 3 : exp\ bool \\
-  T &\mapsto and\ (leq\ X\ Y)\ Z : exp\ bool \\
-\end{array}
+  T, \left(
+  \begin{array}{l}
+  X \mapsto plus\ 1\ 3 : exp\ nat \\
+  Y \mapsto plus\ 2\ 4 : exp\ nat \\
+  Z \mapsto leq\ 8\ 3 : exp\ bool \\
+  T \mapsto and\ (leq\ X\ Y)\ Z : exp\ bool \\
+\end{array}\right)
 \end{align*}
+
+The dependent nature of types in LF allows to express more complex
+languages. We can for example add functions, applications and
+variables to our expressions in a purely first-order style (using de
+Bruijn indices for variables) if we annotate them not only with sorts
+but with an environment of free variables:
+\begin{align*}
+  exp &: env\to tp\to \type,\\
+  atom &: \prd E {env} \nat\to exp\ E\ nat,\\
+  var &: \prd E {env} \prd A {tp} var\ E\ A \to exp\ E\ A,\\
+  leq &: \prd E {env} exp\ e\ nat\to exp\ E\ nat\to exp\ E\ bool, \\
+  lam &: \prd E {env} \prd {A, B} {tp} exp\ (cons\ A\ E)\ B \to
+  exp\ E\ (arr\ A\ B)\\
+  \ldots
+\end{align*}
+
+The encoded expressions are however very verbose: each term
+constructor takes as argument all these annotations. We can
+nonetheless make these information \emph{implicit} in terms (but
+explicit in types) and let a reconstruction algorithm infer them, as
+in \cite{necula1997efficient}. This reconstruction is
+language-dependent, user-provided but does not impair the safety of
+the system for the whole term is still checked afterwards.
+
+There is plenty of room for improvement here: one could for instance
+wish to make use of admissible rules to increase sharing (e.g. adding
+a $lam$ constructor at the top of a term requires to rewrite the whole
+term, or better, apply weakening on it); this supposes to support
+computations on derivation. Besides, LF promotes the use of
+lambda-tree syntax to represent binders; it would require a mechanism
+toimplicitly manage and share the environment in our metalanguage.
 
@@
 
+Aside from enabling to encode a large class of languages generically,
+the metalanguage approach allows to express incrementality features
+usually implemented in an ad-hoc manner, simply by adding new
+constants to the signature.
+
+Suppose we want to implement an \emph{undo system}, storing successive
+versions of a closed expression of sort $bool$ and able to rollback to
+a previous version. We add constants
+\begin{align*}
+  &version : \type,\quad vnil : version,\\
+  &vcons : exp\ nil\ bool \to version \to version
+\end{align*}
+to the signature. The empty repository is now represented as
+$vnil$. Each time we have pushed a full expression $M$, and if $S$ was
+the previous head (a $version$ called its \emph{ancestor}), we push
+$vcons\ M\ S$. This gives us a data structure for an undo stack, and a
+\emph{commit} algorithm. But the sharing inherent to our repositories
+lets us actually represent \emph{trees} of versions, by sharing common
+stack tails, each list head being a \emph{branch}. Reconciling two
+branches' changes into a unique head is called \emph{merging} in
+version control system's terminology: a merge is a version with
+several ancestors. We can represent merges by revising our previous
+addition to the signature into
+\begin{align*}
+  &version : \type,\quad ancestors : \type,\quad anil : ancestors,\\
+  &acons : version \to ancestors \to ancestors,\\
+  &vcons : exp\ nil\ bool \to ancestors \to version
+\end{align*}
+This defines a data structure to represent (acyclic) \emph{graphs} of
+versions; it is the exact data structure of repository used by version
+control systems \textsf{Git}, \textsf{Monotone} and \textsf{Mercurial}
+(see \cite{baudis2009current}).
+
+While our system is based on \emph{bottom-up} term construction, we
+can encode \emph{top-down} construction common to some programming
+environments (e.g. \textsf{Agda}) and tactic-based proof assistants
+(\textsf{Coq}). The user constructs terms by
+successively filling \emph{holes} with terms containing other
+holes. To add (linear) holes to our expressions, add constant
+$$ hole : \prd E {env} \prd A {tp} exp\ E\ A $$
+to the signature. To instantiate a hole with an expression, commit the
+substituted term preserving sharing of subterms.
+
 \section{Architecture}
+
+The system follows a layered architecture.
 
@@
 The \emph{slicer} is the component in charge of slicing a term $M$
-into many terms for the kernel to push in the repository.
+into many terms, pushing them to the repository to enable future
+sharing, and adding version markers. It supports operations:
+\begin{itemize}
+\item $\pfunction{commit}\Sigma{\mr, M}$ pushes $vcons\ M\ (acons\ \mmeta\
+  anil)$ to $\mr$ in several $\function{push}{}$ operations, where
+  $\mmeta$ is the current head.
+\item $\pfunction{merge}\Sigma{\mr, M, \mmmeta}$ pushes $vcons\ M\
+  (acons\ \mmeta\ (acons\ \mmmeta\ anil))$ to $\mr$. Note that it
+  doesn't actually perform the merge, it simply commits a previously
+  computed merge node with value $M$.
+\end{itemize}
+
+The \emph{reconstructor} performs the reconstruction of the derivation
+(an $M$) from the initial expression (an $e$), given the
+derivations for its metavariables (an $\mr$), and the expected type
+($exp\ nil\ bool$), and commits $M$.
+
+Finally, the \emph{compressor} computes a delta $e'$ from a
+metavariable-free expression $e$ by recognizing equal subterms. This
+can be achieved by \emph{hash-consing}.
 

### 2011-10-10 doc/papers/TLDI2012/abstract.tex
TLDI: bibtex
TLDI: version finale

@@
-\documentclass[9pt,authoryear]{sigplanconf}
+\documentclass[9pt]{sigplanconf}
 
@@
 
-\title{Incremental Type Checking}
+\title{Safe Incremental Type Checking}
 %\subtitle{Subtitle Text, if any}
@@
 \authorinfo{Matthias Puech}
-           {University of Bologna, University Paris Diderot}
+           {Department of Comp. Sci., Univ. of Bologna,\\
+             PPS, Team $\pi r^2$ (Univ. Paris Diderot, CNRS, INRIA)}
            {puech@cs.unibo.it}
+
 \authorinfo{Yann R\'egis-Gianas}
-           {University Paris Diderot, CNRS, and INRIA}
+           {PPS, Team $\pi r^2$ (Univ. Paris Diderot, CNRS, INRIA)}
            {yrg@pps.jussieu.fr}
@@
   incrementally, that is by sequentially modifying a term, and
-  re-verifying each time only a smaller amount of information.
+  re-verifying each time only a smaller amount of information than the
+  whole term, still ensuring that it is well-typed.
 \end{abstract}
 
-\category{CR-number}{subcategory}{third-level}
+\category{D.3.3}{Language Constructs and Features}{Data types and
+  structures} \category{F.3.1}{Logics and Meanings of
+  Programs}{Specifying and Verifying and Reasoning about
+  Programs --- Logics of programs}
 
 \terms
-term1, term2
+Theory, Languages
 
 \keywords
-keyword1, keyword2
+incrementality, type checking, logical framework, version control
 
@@
 correct program in one shot becomes quite difficult. On the other
-hand, writing a program in many correct steps can be tedious because
-generally, the verification tool rechecks the entire development at
-each step. This second option is the usual practice when the time for
-verification is negligible with respect to the time to make the
-change, but this is less and less the case, especially when the
-language in question contains proof aspects, and verification involves
-proof search. Some mechanisms already exist to cope with the
-incrementality of proofs or program development: separate compilation,
-interactive toplevel with undo, tactic languages; they all provide in
-different ways a rough approximation of the process of modifying and
-checking incrementally a large term.
+hand, writing a program in many correct steps is the usual practice
+when the time for verification is negligible; the verification tool
+then rechecks the entire development at each step. But this gets more
+tedious especially when the language in question contains proof
+aspects, and verification involves proof search. Some mechanisms
+already exist to cope with the incrementality of proofs or program
+development: separate compilation, interactive toplevel with undo,
+tactic languages; they all provide in different ways a rough
+approximation of the process of modifying and checking incrementally a
+large term.
 
-We propose here an architecture for a generic incremental
-type-checker, a data structure for repositories of typed proofs and a
+We propose here an architecture for a generic and safe incremental
+type checker, a data structure for repositories of typed proofs and a
 language for describing proof deltas. It is based on the simple idea
-of sharing common subterms to avoid rechecking and exploits encoding a
-derivation in a metalanguage. This way, given a signature declaring
-the typing rules and an (untrusted) typing algorithm for my language
-of choice, I get an incremental type-checker for that language. The
-metalanguage approach gives us the ability to encode all the
-aforementioned usual incrementality mechanisms, and more, making our
+of sharing common subterms to avoid rechecking, and exploits encoding
+a derivation in a metalanguage to guarantee the well-typing of the
+result. This way, given a signature declaring the typing rules and an
+(untrusted) typing algorithm for my language of choice, I get an
+incremental type checker for that language. The metalanguage approach
+gives us the ability to encode all the aforementioned usual
+incrementality mechanisms in a type-safe way, and more, making our
 system akin to a typed version control system.
@@
 $$ e, e' \gequal n \gor e + e' \gor e \land e' \gor e \leq e' $$
-The algorithm to determine in a batch fashion wether the term
+The algorithm to determine in a batch fashion whether the term
 $$ e_1 = (1 + 3 \leq 2 + 4) \land (8 \leq 3) $$ is well-sorted is
@@
 
-\section{The metalanguage}
+\section{A metalanguage to encode derivations}
 
@@
 the whole state of the batch verifier (here the sort). An obvious
-choice is to take $M\gequal(e \text{ with metavariables})$ and $A\gequal
-\bool\gor\nat$, but switching to another language, we'd have to
-redefine another repository language. Another more modular choice for
-this is the \emph{metalanguage} \emph{LF} \cite{harper1993framework}: it allows
-to specify syntax and rules of an object language as a
-\emph{signature} $\Sigma$, and check terms against this signature with
-a generic algorithm. We'll use an increasing fragment of it. The
-so-called \emph{intrinsic} style of LF signature for our expression
-language is:
-\begin{align*}
-  tp &: \type,\quad nat : tp,\quad bool : tp,\quad exp : tp\to nat,\\
-  atom &: \nat\to exp\ nat,\\
-  plus &: exp\ nat \to exp\ nat \to exp\ nat,\\
-  and &: exp\ bool\to exp\ bool\to exp\ bool,\\
-  leq &: exp\ nat\to exp\ nat\to exp\ bool
-\end{align*}
+choice is to take $M\gequal(e \text{ with metavariables})$ and
+$A\gequal \bool\gor\nat$, but switching to another language, we'd have
+to redefine another repository language. Moreover, this choice is no
+longer expressive enough when introducing binders. Another more
+modular choice for this is the \emph{metalanguage} \emph{LF}
+\cite{harper1993framework}: it allows to specify syntax and rules of
+an object language as a \emph{signature} $\Sigma$, and check terms
+against this signature with a generic algorithm. We'll use an
+increasing fragment of it. The so-called \emph{intrinsic} style of LF
+signature for our expression language is:
+$$
+\begin{array}{rl}
+  \mathsf{tp} &: \type,\quad
+  \mathsf{nat} : \mathsf{tp},\quad
+  \mathsf{bool} : \mathsf{tp},\quad
+  \mathsf{exp} : \mathsf{tp}\to \mathsf{nat},\\
+  \mathsf{atom} &: \nat\to \mathsf{exp}\ \mathsf{nat},\\
+  \mathsf{plus} &: \mathsf{exp}\ \mathsf{nat} \to \mathsf{exp}\ \mathsf{nat} \to \mathsf{exp}\ \mathsf{nat},\\
+  \mathsf{and} &: \mathsf{exp}\ \mathsf{bool}\to \mathsf{exp}\ \mathsf{bool}\to \mathsf{exp}\ \mathsf{bool},\\
+  \mathsf{leq} &: \mathsf{exp}\ \mathsf{nat}\to \mathsf{exp}\ \mathsf{nat}\to \mathsf{exp}\ \mathsf{bool}
+\end{array}
+$$
 In this style, both the encoding of an expression and its sort are
@@
   \begin{array}{l}
-  X \mapsto plus\ 1\ 3 : exp\ nat \\
-  Y \mapsto plus\ 2\ 4 : exp\ nat \\
-  Z \mapsto leq\ 8\ 3 : exp\ bool \\
-  T \mapsto and\ (leq\ X\ Y)\ Z : exp\ bool \\
+  X \mapsto \mathsf{plus}\ 1\ 3 : \mathsf{exp}\ \mathsf{nat} \\
+  Y \mapsto \mathsf{plus}\ 2\ 4 : \mathsf{exp}\ \mathsf{nat} \\
+  Z \mapsto \mathsf{leq}\ 8\ 3 : \mathsf{exp}\ \mathsf{bool} \\
+  T \mapsto \mathsf{and}\ (\mathsf{leq}\ X\ Y)\ Z : \mathsf{exp}\ \mathsf{bool} \\
 \end{array}\right)
@@
 but with an environment of free variables:
-\begin{align*}
-  exp &: env\to tp\to \type,\\
-  atom &: \prd E {env} \nat\to exp\ E\ nat,\\
-  var &: \prd E {env} \prd A {tp} var\ E\ A \to exp\ E\ A,\\
-  leq &: \prd E {env} exp\ e\ nat\to exp\ E\ nat\to exp\ E\ bool, \\
-  lam &: \prd E {env} \prd {A, B} {tp} exp\ (cons\ A\ E)\ B \to
-  exp\ E\ (arr\ A\ B)\\
+$$
+\begin{array}{rl}
+  \mathsf{exp} &: \mathsf{env}\to \mathsf{tp}\to \type,\\
+  \mathsf{atom} &: \prd E {\mathsf{env}} \nat\to \mathsf{exp}\ E\ \mathsf{nat},\\
+  \mathsf{var} &: \prd E {\mathsf{env}} \prd A {\mathsf{tp}} \mathsf{var}\ E\ A \to \mathsf{exp}\ E\ A,\\
+  \mathsf{leq} &: \prd E {\mathsf{env}} \mathsf{exp}\ E\ \mathsf{nat}\to \mathsf{exp}\ E\ \mathsf{nat}\to \mathsf{exp}\ E\ \mathsf{bool}, \\
+  \mathsf{lam} &: \prd E {\mathsf{env}} \prd {A, B} {\mathsf{tp}}
+  \mathsf{exp}\ (\mathsf{cons}\ A\ E)\ B \\ &
+  \qquad\to\mathsf{exp}\ E\ (\mathsf{arr}\ A\ B)\\
   \ldots
-\end{align*}
-
+\end{array}
+$$
 The encoded expressions are however very verbose: each term
@@
 
-There is plenty of room for improvement here: one could for instance
-wish to make use of admissible rules to increase sharing (e.g. adding
-a $lam$ constructor at the top of a term requires to rewrite the whole
-term, or better, apply weakening on it); this supposes to support
-computations on derivation. Besides, LF promotes the use of
-lambda-tree syntax to represent binders; it would require a mechanism
-toimplicitly manage and share the environment in our metalanguage.
+LF promotes the use of lambda-tree syntax to represent binders:
+instead of encoding the syntax first-order, it uses the $\lambda$
+binder built in LF to encode binders in the object language. This
+style of encoding has the advantage of making the manipulation of the
+environment (weakening, exchange\ldots) implicit in the deltas, but
+raises new challenges for the delta language and the verification
+process: how to share a subterm underneith a lambda? How to
+efficiently verify that such a delta is well-typed?
 
@@
 
-Aside from enabling to encode a large class of languages generically,
-the metalanguage approach allows to express incrementality features
-usually implemented in an ad-hoc manner, simply by adding new
-constants to the signature.
+Aside from enabling to encode a large class of deductive systems
+safely and generically, the metalanguage approach allows to express
+incrementality features usually implemented in an ad-hoc manner,
+simply by adding new constants to the signature.
+
+\paragraph{Version control}
 
@@
 a previous version. We add constants
-\begin{align*}
-  &version : \type,\quad vnil : version,\\
-  &vcons : exp\ nil\ bool \to version \to version
-\end{align*}
+$$
+\begin{array}{rl}
+  &\mathsf{version} : \type,\quad \mathsf{vnil} : \mathsf{version},\\
+  &\mathsf{vcons} : \mathsf{exp}\ \mathsf{nil}\ \mathsf{bool} \to \mathsf{version} \to \mathsf{version}
+\end{array}
+$$
 to the signature. The empty repository is now represented as
 $vnil$. Each time we have pushed a full expression $M$, and if $S$ was
-the previous head (a $version$ called its \emph{ancestor}), we push
-$vcons\ M\ S$. This gives us a data structure for an undo stack, and a
+the previous head (a $\mathsf{version}$ called its \emph{ancestor}), we push
+$\mathsf{vcons}\ M\ S$. This gives us a data structure for an undo stack, and a
 \emph{commit} algorithm. But the sharing inherent to our repositories
@@
 addition to the signature into
-\begin{align*}
-  &version : \type,\quad ancestors : \type,\quad anil : ancestors,\\
-  &acons : version \to ancestors \to ancestors,\\
-  &vcons : exp\ nil\ bool \to ancestors \to version
-\end{align*}
+$$\begin{array}{rl}
+  &\mathsf{version} : \type,\quad
+  \mathsf{ancestors} : \type,\quad
+  \mathsf{anil} : \mathsf{ancestors},\\
+  &\mathsf{acons} : \mathsf{version} \to \mathsf{ancestors} \to \mathsf{ancestors},\\
+  &\mathsf{vcons} : \mathsf{exp}\ \mathsf{nil}\ \mathsf{bool} \to \mathsf{ancestors} \to \mathsf{version}
+\end{array}$$
 This defines a data structure to represent (acyclic) \emph{graphs} of
@@
 control systems \textsf{Git}, \textsf{Monotone} and \textsf{Mercurial}
-(see \cite{baudis2009current}).
+(see e.g. \cite{chacon2009git}) except that where they have directories
+and text files we have arbitrary typed terms.
+
+\paragraph{Top-down construction}
 
@@
 environments (e.g. \textsf{Agda}) and tactic-based proof assistants
-(\textsf{Coq}). The user constructs terms by
+(e.g. \textsf{Coq}). The user constructs terms by
 successively filling \emph{holes} with terms containing other
 holes. To add (linear) holes to our expressions, add constant
-$$ hole : \prd E {env} \prd A {tp} exp\ E\ A $$
+$$ \mathsf{hole} : \prd E {\mathsf{env}} \prd A {\mathsf{tp}} \mathsf{exp}\ E\ A $$
 to the signature. To instantiate a hole with an expression, commit the
@@
 
-The system follows a layered architecture.
+We can implement this system following a layered architecture.
 
@@
 \begin{itemize}
-\item $\pfunction{push}{\Sigma}{\mr,M}$ checks $M$ against $\Sigma$
-  in $\mr$, synthetizes its type $A$, chooses a fresh metavariable
-  $\mmeta$ for $M$ and returns $\mr[\mmeta\mapsto M:A]$ and $\mmeta$.
+\item $\pfunction{push}{\Sigma}{\mr,M}$ checks a small part $M$ of a
+  larger term against $\Sigma$ in $\mr$, synthetizes its type $A$,
+  chooses a fresh metavariable $\mmeta$ for $M$ and returns
+  $\mr[\mmeta\mapsto M:A]$ and $\mmeta$.
 \item $\pfunction{pull}\Sigma{\mr, \mmeta}$ returns the term $M$
@@
 \begin{itemize}
-\item $\pfunction{commit}\Sigma{\mr, M}$ pushes $vcons\ M\ (acons\ \mmeta\
-  anil)$ to $\mr$ in several $\function{push}{}$ operations, where
-  $\mmeta$ is the current head.
-\item $\pfunction{merge}\Sigma{\mr, M, \mmmeta}$ pushes $vcons\ M\
-  (acons\ \mmeta\ (acons\ \mmmeta\ anil))$ to $\mr$. Note that it
-  doesn't actually perform the merge, it simply commits a previously
-  computed merge node with value $M$.
+\item $\pfunction{commit}\Sigma{\mr, M}$ pushes $\mathsf{vcons}\ M\
+  (\mathsf{acons}\ \mmeta\ \mathsf{anil})$ to $\mr$ in several
+  $\function{push}{}$ operations, where $\mmeta$ is the current head.
+\item $\pfunction{merge}\Sigma{\mr, M, \mmmeta}$ pushes
+  $\mathsf{vcons}\ M\ (\mathsf{acons}\ \mmeta\ (\mathsf{acons}\
+  \mmmeta\ \mathsf{anil}))$ to $\mr$. Note that it doesn't actually
+  perform the merge, it simply commits a previously computed merge
+  node with value $M$.
 \end{itemize}
@@
 derivations for its metavariables (an $\mr$), and the expected type
-($exp\ nil\ bool$), and commits $M$.
+($\mathsf{exp}\ \mathsf{nil}\ \mathsf{bool}$), and commits $M$.
 
 Finally, the \emph{compressor} computes a delta $e'$ from a
-metavariable-free expression $e$ by recognizing equal subterms. This
-can be achieved by \emph{hash-consing}.
+metavariable-free expression $e$ by recognizing equal subterms in
+$\mr$. This can be achieved by \emph{hash-consing}.
 
@@
 \bibliographystyle{abbrvnat}
-\bibliography{../../../../../Documents/These/biblio.bib}
+\bibliography{../../english.bib}
 

### 2012-04-27 doc/papers/LFMTP2012/paper.tex
LFMTP'12: new article tentative

\documentclass[9pt]{sigplanconf}

\usepackage{ntheorem}
\usepackage{amsmath}
\usepackage{amssymb}
\newtheorem{theorem}{Theorem}
\newtheorem{definition}{Definition}
\usepackage{macros}

\begin{document}

\conferenceinfo{LFMTP '12}{September 9, 2012 -- Copenhagen, Denmark}
\copyrightyear{2012}
\copyrightdata{[to be supplied]}

\titlebanner{banner above paper title}        % These are ignored unless
\preprintfooter{short description of paper}   % 'preprint' option specified.

\title{Certified incremental program transformations}
%\subtitle{Subtitle Text, if any}

\authorinfo{Matthias Puech}
           {Department of Comp. Sci., Univ. of Bologna,\\
             PPS, Team $\pi r^2$ (Univ. Paris Diderot, CNRS, INRIA)}
           {puech@cs.unibo.it}

\authorinfo{Yann R\'egis-Gianas}
           {PPS, Team $\pi r^2$ (Univ. Paris Diderot, CNRS, INRIA)}
           {yrg@pps.jussieu.fr}

\maketitle

\begin{abstract}
\end{abstract}

\category{D.3.3}{Language Constructs and Features}{Data types and
  structures} \category{F.3.1}{Logics and Meanings of
  Programs}{Specifying and Verifying and Reasoning about
  Programs --- Logics of programs}

\terms
Theory, Languages

\keywords
incrementality, type checking, logical framework, version control

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{Introduction}

\section{The calculus}

\subsection{Core calculus}

\paragraph{Syntax}

We defined the syntax of canonical LF (TODO cite) enriched with
metavariables and substitutions. Canonicity against $\beta$-reduction
is enforced in the syntax, but we think of them also as $\eta$-long;
it will be enforced by typing.

\begin{align*}
  K &\gequal
  \prd x A K \gor
  \type &
  \text{Kind}\\
  A &\gequal
  \prd x A A \gor
  \app {\cst a} S &
  \text{Type family}\\
  M &\gequal
  \lam x M \gor
  \app H S \gor
  \smeta X \sigma &
  \text{Object}\\
  H &\gequal
  \var x \gor
  \cst c &
  \text{Head}\\
  S &\gequal
  \spinenil \gor
  \spinecons S M &
  \text{Spine}\\
  \sigma &\gequal
  \msubstnil \gor
  \msubstcons \sigma x M &
  \text{Substitution} \\
  \Gamma &\gequal
  \gnil \gor
  \gcons \Gamma {\tp {\var x} A} &
  \text{Environment} \\
  \Sigma &\gequal
  \gnil \gor
  \gcons \Sigma {\cst c : A} \gor
  \gcons \Sigma {\cst a : K} &
  \text{Signature}
\end{align*}

As usual, we write $\meta X$ for $\smeta X \msubstnil$, $A\to B$ for
$\prd x A B$ when $x\notin \FV{B}$ and adopt list notations for $S,
\sigma, \Gamma, \Sigma$ when convenient.

A \emph{repository} is a finite map from metavariable to judgements of
object well-typing, along with a distinguished metavariable, the
\emph{head}:
$$ \repo R\;:\;(\meta X \mapsto (\Gamma\vdash\tp M A)), \smeta X \sigma $$

\paragraph{Substitution} We overload \emph{substitution} as a partial
operation on terms, given a parallel substitution $\sigma$. Since the
syntax of object is not stable by substitution, it is defined
\emph{hereditarily}: it is mutually recursive with a
\emph{normalization} function $\gcut M S$ which can in turn trigger a
chain of substitutions, leading to a canonical term, a stuck
substitution or divergence. % TODO vraiment?

\begin{align*}
  \gsubst {(\lam x M)} \sigma &= \lam x {\gsubst M \sigma} &
  \text{if $x\notin dom{(\sigma)}$}
  \\
  \gsubst {(\app {\cst c} S)} \sigma &= \app {\cst c} {(\gsubst S
    \sigma)}
  \\
  \gsubst {(\app {\var x} S)} \sigma &=
  \app {\var x} {\gsubst S \sigma} &
  \text{if $x\notin dom{(\sigma)}$}
  \\
  \gsubst {(\app {\var x} S)} \sigma &=
  \gcut M {\gsubst S \sigma} &
  \text{if $\sigma(x) = M$}
  \\
  \gsubst {\smeta X {\sigma}} {\sigma'} &=
  \smeta X {\gcomp\sigma{\sigma'}}
  \\
  \gsubst\spinenil\sigma &=
  \spinenil \\
  \gsubst{\spinecons S M}\sigma &=
  \spinecons{\gsubst S\sigma}{\gsubst M\sigma}
  \\
  \gcut {\lam x M} {\spinecons N S} &=
  \gcut {\gsubst M N} S
  \\
  % TODO cas meta/S ????
  % \gcut {\smeta X\sigma} S &=
  % \smeta X{\gcomp\sigma\sigma'}
  % \\
  \gcut {\app H S} \spinenil &=
  \app H S
  \\
  \gcomp \sigma \msubstnil &=
  \sigma
  \\
  \gcomp \sigma {(\msubstcons{\sigma'} x M)} &=
  \gcomp {\gsubst\sigma {\msubst x M}} \sigma'
  \\
  \gsubst {(\msubstcons{\sigma'} x M)} \sigma &=
  \msubstcons{\gsubst{\sigma'}\sigma} x {\gsubst M \sigma}
  \\
  \gsubst {\msubstnil}{\sigma} &= \msubstnil
\end{align*}

Note that this notion of substitution presupposes $\eta$-long normal
forms: a spine applied to another application $\gcut {\app H S} {S'}$
blocks the substitution.

% TODO terminaison??

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\bibliographystyle{abbrvnat}
\bibliography{../../english.bib}

% \begin{thebibliography}{}
% \softraggedright
% \bibitem[Smith et~al.(2009)Smith, Jones]{smith02}
% P. Q. Smith, and X. Y. Jones. ...reference text...
% \end{thebibliography}

\end{document}

### 2012-04-27 doc/papers/LFMTP2012/paper.tex
Paper

@@
 
+The core calculus defines the \emph{values} we want to
+manipulate. These contain constants, and --- contrarily to \eg\ ML
+values --- binders and variables enabling $\lambda$-tree syntax (TODO
+cite). Thus, they embed a superficial notion of computation:
+hereditary substitution. This calculus is an extension of \emph{spine
+  canonical LF} (TODO cite) enriched with metavariables. We present
+its typing algorithm relatively to a signature. It features an
+explicit caching mechanism using metavariables.
+
 \paragraph{Syntax}
 
-We defined the syntax of canonical LF (TODO cite) enriched with
-metavariables and substitutions. Canonicity against $\beta$-reduction
-is enforced in the syntax, but we think of them also as $\eta$-long;
-it will be enforced by typing.
+In this variant, only $\eta$-long $\beta$-normal forms have an
+existence: canonicity for $\beta$-reduction is enforced in the syntax;
+canonicity for $\eta$-expansion will be enforced by typing. Also,
+application of a \emph{head} to a \emph{spine} (\ie\ a constant or
+variable) is $n$-ary as opposed to the more standard binary
+application: this eases reasoning about substitution and
+$\eta$-expansion (TODO cite).
 
@@
   \app {\cst a} S &
-  \text{Type family}\\
+  \text{Type family} \\
   M &\gequal
   \lam x M \gor
-  \app H S \gor
+  F \gor
   \smeta X \sigma &
-  \text{Object}\\
+  \text{Object} \\
+  F &\gequal \app H S &
+  \text{Application} \\
   H &\gequal
@@
   \gcons \Gamma {\tp {\var x} A} &
-  \text{Environment} \\
-  \Sigma &\gequal
-  \gnil \gor
-  \gcons \Sigma {\cst c : A} \gor
-  \gcons \Sigma {\cst a : K} &
-  \text{Signature}
+  \text{Environment}
 \end{align*}
@@
 
-A \emph{repository} is a finite map from metavariable to judgements of
-object well-typing, along with a distinguished metavariable, the
-\emph{head}:
-$$ \repo R\;:\;(\meta X \mapsto (\Gamma\vdash\tp M A)), \smeta X \sigma $$
+Metavariables $\smeta X \sigma$ stand for open objects with free
+variables in $\dom{\sigma}$; they are defined in a repository (see below). We
+write $\FMV{M}$ for the set of metavariables in $M$.
 
-\paragraph{Substitution} We overload \emph{substitution} as a partial
-operation on terms, given a parallel substitution $\sigma$. Since the
-syntax of object is not stable by substitution, it is defined
-\emph{hereditarily}: it is mutually recursive with a
-\emph{normalization} function $\gcut M S$ which can in turn trigger a
-chain of substitutions, leading to a canonical term, a stuck
-substitution or divergence. % TODO vraiment?
+\paragraph{Substitution}
 
+We overload \emph{substitution} as a partial operation on terms, given
+a parallel substitution $\sigma$. Since the syntax of object is not
+stable by substitution, it is defined \emph{hereditarily}: it is
+mutually recursive with a \emph{normalization} function $\gcut M S$
+which can in turn trigger a chain of substitutions, leading to a
+canonical term, a stuck substitution or divergence. % TODO vraiment?
 \begin{align*}
   \gsubst {(\lam x M)} \sigma &= \lam x {\gsubst M \sigma} &
-  \text{if $x\notin dom{(\sigma)}$}
+  \text{if $x\notin \dom{\sigma}$}
   \\
@@
   \app {\var x} {\gsubst S \sigma} &
-  \text{if $x\notin dom{(\sigma)}$}
+  \text{if $x\notin \dom{\sigma}$}
   \\
@@
 
+We extend this operation trivially on families $\gsubst A \sigma$
+and kinds $\gsubst K \sigma$
+
 % TODO terminaison??
+% TODO stabilité de la eta
+
+\paragraph{Typing algorithm}
+
+A \emph{repository}
+is a finite map from metavariable to judgements of object well-typing,
+along with a distinguished metavariable, the \emph{head}:
+$$ \repo R\;:\;(\meta X \mapsto (\Gamma\vdash\tp M A)), \smeta X \sigma $$
+We maintain the invariant that a repository $\repo R$ is
+\emph{acyclic}, that is for all $\meta X$ in $\repo R$, the fixpoint
+of $\FMV{\meta X}$ does not contain $\meta X$.
+
+A \emph{signature} declares object and type constants to be used in
+objects:
+$$  \Sigma \gequal
+  \gnil \gor
+  \gcons \Sigma {\cst c : A} \gor
+  \gcons \Sigma {\cst a : K} \qquad
+  \text{Signature}
+$$
+
+\subsection{The computation calculus}
+
+We now turn to the definition of our computation calculus on LF
+objects. This language is untyped
+\begin{align*}
+  T &\gequal
+   M \gor
+  \match T \casenobar P T\ |\ \ldots\ \case P T \\
+  P &\gequal \app {\cst c} {\vec x}
+\end{align*}
+
+We extend our notion of signature to allow object constants to embed a
+computational behaviour.
+
+$$  \Sigma \gequal
+  \ldots \gor
+  \gcons \Sigma {\cst c : A = T}
+  \qquad
+  \text{Signature}
+$$
 

### 2012-04-30 doc/papers/LFMTP2012/paper.tex
paper

@@
 \usepackage{macros}
+\usepackage{mathpartir}
 
@@
   \spinenil \gor
-  \spinecons S M &
+  \spinecons M S &
   \text{Spine}\\
@@
   \msubstcons \sigma x M &
-  \text{Substitution} \\
+  \text{Parallel substitution} \\
   \Gamma &\gequal
-  \gnil \gor
-  \gcons \Gamma {\tp {\var x} A} &
+  \envnil \gor
+  \envcons \Gamma x A &
   \text{Environment}
@@
 
-As usual, we write $\meta X$ for $\smeta X \msubstnil$, $A\to B$ for
+As is customary, we adopt the Barendregt convention for naming bound
+variables, write $\meta X$ for $\smeta X \msubstnil$, $A\to B$ for
 $\prd x A B$ when $x\notin \FV{B}$ and adopt list notations for $S,
@@
 Metavariables $\smeta X \sigma$ stand for open objects with free
-variables in $\dom{\sigma}$; they are defined in a repository (see below). We
-write $\FMV{M}$ for the set of metavariables in $M$.
+variables in $\dom{\sigma}$; they are defined in a repository (see
+below). We write $\FMV{M}$ for the set of metavariables in $M$ and
+speak of a \emph{metaclosed} object $M$ if $\FMV{M} = \emptyset$.
 
@@
 a parallel substitution $\sigma$. Since the syntax of object is not
-stable by substitution, it is defined \emph{hereditarily}: it is
-mutually recursive with a \emph{normalization} function $\gcut M S$
-which can in turn trigger a chain of substitutions, leading to a
-canonical term, a stuck substitution or divergence. % TODO vraiment?
+stable by (textual) substitution, it is defined \emph{hereditarily}: it
+is mutually recursive with a \emph{cut} function $\gcut M S$ which can
+in turn trigger a chain of substitutions, leading to a canonical term,
+a stuck substitution or divergence. % TODO vraiment?
 \begin{align*}
@@
   \spinenil \\
-  \gsubst{\spinecons S M}\sigma &=
-  \spinecons{\gsubst S\sigma}{\gsubst M\sigma}
+  \gsubst{\spinecons M S}\sigma &=
+  \spinecons{\gsubst M\sigma}{\gsubst S\sigma}
   \\
@@
   \app H S
+  \\
+  \gcut {\smeta X\sigma} \spinenil &= \smeta X\sigma
   \\
@@
 
+We extend this operation trivially on families $\gsubst A \sigma$ and
+kinds $\gsubst K \sigma$. We also allow to abusively write the
+non-canonical application $\app M S$ as a shorthand for $\gcut M S$ if
+it is defined.
+
 Note that this notion of substitution presupposes $\eta$-long normal
-forms: a spine applied to another application $\gcut {\app H S} {S'}$
-blocks the substitution.
+forms: a non-empty spine cut against to another non-empty spine $\gcut
+{\app H S} {S'}$ blocks the substitution.
 
-We extend this operation trivially on families $\gsubst A \sigma$
-and kinds $\gsubst K \sigma$
-
+% TODO pourquoi cette def?
 % TODO terminaison??
-% TODO stabilité de la eta
+% TODO stabilité de la eta?
 
@@
 
-A \emph{repository}
-is a finite map from metavariable to judgements of object well-typing,
-along with a distinguished metavariable, the \emph{head}:
-$$ \repo R\;:\;(\meta X \mapsto (\Gamma\vdash\tp M A)), \smeta X \sigma $$
-We maintain the invariant that a repository $\repo R$ is
-\emph{acyclic}, that is for all $\meta X$ in $\repo R$, the fixpoint
-of $\FMV{\meta X}$ does not contain $\meta X$.
+Typing an object while storing intermediate derivations requires
+threading and maintaining a data structure of named partial
+derivations (a repository) throughout the typing process. We present a
+bidirectional typing algorithm updating an input repository.
+
+A \emph{repository} $\mr$ is a finite map from metavariable to
+judgements of object well-typing, along with a distinguished
+metavariable, the \emph{head}:
+$$ \mr\;:\;(\meta X \mapsto (\Gamma\vdash\tp M A)), \smeta X \sigma $$
+We maintain the invariant that a repository $\mr$ is
+\emph{acyclic}, that is for all $\meta X$ in $\mr$, the fixpoint
+of $\FMV{\meta X}$ does not contain $\meta X$. With $\mr =
+(\Delta, X[\sigma])$, we write $\hat {\mr}$ for $X[\sigma]$,
+abusively $\mr$ for $\Delta$ and $\mr (\smeta X \sigma)$ for
+$\gsubst M\sigma$ if $\Delta(\meta X) = (\Gamma\vdash M : A)$.
 
@@
 objects:
-$$  \Sigma \gequal
+$$
+\Sigma \gequal
 \gnil \gor
@@
 
+The algorithm is presented in Fig. \ref{fig:typing-algo} as a
+syntax-directed inference system. All judgements are of the form $\gj
+U V W$ where $U$ and $V$ are inputs, $V$ is the induction hypothesis
+and $W$ the output. They are parameterized by a constant and implicit
+signature $\Sigma$.
+
+\begin{figure*}
+
+  \fbox{$\jm\mr\Gamma M A \mmr {M'}$}
+  \qquad
+  Object typing
+
+  \begin{mathpar}
+    % Lambda
+    \infer{
+      \jm\mr{\envcons\Gamma x A} M {A'} \mmr {M'}
+    }{
+      \jm\mr\Gamma {\lam x M} {\prd x A {A'}} \mmr {\lam x M'}
+    }
+
+    % App
+    \infer{
+      \jf\mr\Gamma F \mmr {M'} {A'}
+       \and
+      \jea\mmr\Gamma A {A'}
+    }{
+      \jm\mr\Gamma F A \mmr {M'}
+    }
+
+    % Meta
+    \infer{
+      \mr(\meta X) = (\Gamma'\vdash M:A')
+      \and
+      \js\mr\Gamma \sigma {\Gamma'} \mmr {\sigma'}
+      \and
+      \jea\mmr\Gamma A {\gsubst {A'} {\sigma'}}
+    }{
+      \jm\mr\Gamma {\smeta X \sigma} A \mmr {M'}
+    }
+  \end{mathpar}
+
+  \fbox{$\js\mr\Gamma\sigma{\Gamma'}\mmr{\sigma'}$}
+  \qquad
+  {Substitution typing}
+
+  \begin{mathpar}
+
+    % Cons
+    \infer{
+      \js\mr\Gamma\sigma{\Gamma'}\mmr{\sigma'}
+      \and
+      \jm\mmr\Gamma M {\gsubst A {\sigma'}} \mmmr {M'}
+    }{
+      \js\mr\Gamma{(\msubstcons \sigma x M)} {(\envcons {\Gamma'} x {A'})}
+      \mmmr {(\msubstcons {\sigma'} x {M'})}
+    }
+
+    % Nil
+    \infer{ }{
+      \js\mr\Gamma\msubstnil\envnil\mr\envnil
+    }
+
+  \end{mathpar}
+
+  \caption{Typing algorithm}
+  \label{fig:typing-algo}
+\end{figure*}
+
+\paragraph{Checkout}
+
+We define the partial \emph{checkout} operation $\checkout{\mr}$
+(resp. $\checkoutr{M}$, $\checkoutr{S}$) from a repository
+(resp. object, spine) to a metaclosed object as unfolding the
+definitions of all metavariables recursively:
+\begin{align*}
+  \checkout{\mr} &= \checkoutr{\hat{\mr}} \\
+  \checkoutr{\smeta X \sigma} &= \checkoutr{\mr(\smeta X\sigma)}\\
+  \checkoutr{\lam x M} &= \lam x \checkoutr{M} \\
+  \checkoutr{\app H S} &= \app H {(\checkoutr{S})} \\
+  \checkoutr{\spinecons M S} &= \spinecons {\checkoutr M} {\checkoutr S} \\
+  \checkoutr{\spinenil} &= \spinenil
+\end{align*}
+
+A repository is said to be \emph{well-formed} if $\checkout {\mr}$
+is defined.
+
 \subsection{The computation calculus}
@@
 We now turn to the definition of our computation calculus on LF
-objects. This language is untyped
+objects. This language is untyped. % TODO??
 \begin{align*}
   T &\gequal
-   M \gor
-  \match T \casenobar P T\ |\ \ldots\ \case P T \\
+  F \gor
+  \matchin T \Gamma \casenobar P T\ |\ \ldots\ \case P T \\
   P &\gequal \app {\cst c} {\vec x}
@@
 
+\section{Implementation}
+
+\section{Examples}
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

### 2012-05-01 doc/papers/LFMTP2012/paper.tex
paper

@@
 \section{Introduction}
+
+\subsection{The LF representation language}
+\subsection{Storing computation results}
+\subsection{Naming intermediate results}
 
@@
 As is customary, we adopt the Barendregt convention for naming bound
-variables, write $\meta X$ for $\smeta X \msubstnil$, $A\to B$ for
-$\prd x A B$ when $x\notin \FV{B}$ and adopt list notations for $S,
-\sigma, \Gamma, \Sigma$ when convenient.
+variables, write $A\to B$ for $\prd x A B$ when $x\notin \FV{B}$,
+$\meta X$ for $\smeta X \msubstnil$, and adopt indifferently list or
+map notations for $\sigma$ and $\Gamma$, \ie\ write $\sigma(x) = M$ or
+$\Gamma(x) = A$.
 
@@
 A \emph{signature} declares object and type constants to be used in
-objects:
-$$
-\Sigma \gequal
+objects. To each object constant we assign an annotation $L$ setting
+the slicing behaviour of the constant during type-checking: either
+sliceable $\annsliceable$ (each application of this constant will have
+its own name in the repository), non-sliceable $\annnonsliceable$ (we
+don't create a name for applications of this constant), or defined
+$\anndefined{F}$ (see \ref{sec:computational-calculus}).
+\begin{align*}
+  L &\gequal
+  \annsliceable \gor
+  \annnonsliceable \gor
+  \anndefined{F}
+  \\
+  \Sigma &\gequal
   \gnil \gor
-\gcons \Sigma {\cst c : A} \gor
-\gcons \Sigma {\cst a : K} \qquad
-\text{Signature}
-$$
+  \gcons \Sigma {\cst c :^L A} \gor
+  \gcons \Sigma {\cst a :^L K}
+  &\text{Signature}
+\end{align*}
+We write $\Sigma(\cst c) = A; L$ for $(\cst c :^L A) \in \Sigma$
+suggesting the implementation of signatures as maps.
 
 The algorithm is presented in Fig. \ref{fig:typing-algo} as a
-syntax-directed inference system. All judgements are of the form $\gj
+syntax-directed inference system. All judgements are of the form $\gjo
 U V W$ where $U$ and $V$ are inputs, $V$ is the induction hypothesis
@@
 
+  \fbox{$\jf\mr\Gamma F \mmr M A$}
+  \qquad
+  {Application typing}
+
+  \begin{mathpar}
+
+    % Non-sliceable
+    \infer{
+      \jh\mr\Gamma H A \annnonsliceable
+      \and
+      \jl\mr\Gamma A S \mmr {S'} {A'}
+    }{
+      \jf\mr\Gamma {\app H S} \mmr {\app H {S'}} {A'}
+    }
+
+    % Sliceable
+    \infer{
+      \jh\mr\Gamma H A \annsliceable
+      \and
+      \jl\mr\Gamma A S \mmr {S'} {A'}
+      \and
+      \jp\mmr\Gamma {\app H S} A \mmmr
+    }{
+      \jf\mr\Gamma {\app H S} \mmmr {\hat\mmmr} {A'}
+    }
+
+  \end{mathpar}
+
+  \fbox{$\jh\mr\Gamma H A L$}
+  \qquad
+  {Head typing}
+
+  \begin{mathpar}
+    % constant
+    \infer{
+      \Sigma(\cst c) = A; L
+    }{
+      \jh\mr\Gamma {\cst c} A L
+    }
+
+    % variable
+    \infer{
+      \Gamma(\var x) = A
+    }{
+      \jh\mr\Gamma {\var x} A \annnonsliceable
+    }
+  \end{mathpar}
+
+  \fbox{$\jp\mr\Gamma M A \mmr$}
+  \qquad
+  {Push}
+
+  \begin{mathpar}
+    \infer{
+      \jstr{\Gamma\vdash\tp M A}{\Gamma'\vdash\tp {M'} {A'}}\sigma
+      \and
+      \text{$\meta X$ fresh in $\Delta$}
+    }{
+      \jp{(\Delta; \smeta Z \sigma)}\Gamma M A {(\mapadd\Delta{\meta
+          X}{(\Gamma'\vdash M':A')}; \smeta X \sigma)}
+    }
+  \end{mathpar}
+
   \caption{Typing algorithm}
@@
 \subsection{The computation calculus}
+\label{sec:computational-calculus}
 
@@
   F \gor
-  \matchin T \Gamma \casenobar P T\ |\ \ldots\ \case P T \\
-  P &\gequal \app {\cst c} {\vec x}
+  \matchin T \Gamma \casenobar {\app {\cst c} {\vec x}} T\ |\ \ldots\
+  \case {\app {\cst c} {\vec x}} T
 \end{align*}
-
-We extend our notion of signature to allow object constants to embed a
-computational behaviour.
-
-$$  \Sigma \gequal
-  \ldots \gor
-  \gcons \Sigma {\cst c : A = T}
-  \qquad
-  \text{Signature}
-$$
 

### 2012-05-04 doc/papers/LFMTP2012/paper.tex
paper

@@
 
+%% Here is a problem
 \begin{abstract}
+  %% 4 sentences:
+  %% State the problem
+  %% Say why it’s an interesting problem
+  %% Say what your solution achieves
+  %% Say what follows from your solution
 \end{abstract}
@@
 
-\section{Introduction}
+\section*{Introduction}
+%% It’s an interesting problem (1 page)
 
-\subsection{The LF representation language}
+As programming languages develop more static analysis like rich type
+systems, programs get safer as bugs are caught earlier.
+
+%% Describe the problem (Use an example)
+%% State your contributions (Bulleted list of contributions)
+
+\section{Background}
+%% It’s an unsolved problem (1 page)
+
+\section{Here is my idea}
+%% Here is my idea (2 pages)
+
+\subsection{The \LF\ representation language}
+
+The emission of certificates by untrusted applications first raises
+the question of the \emph{data structure} to use for these
+certificates. The requirements are:
+\begin{itemize}
+\item conciseness of the certificates, since they are usually stored
+  or transfered remotely; in particular,
+\item a well-understood, short and independent verification algorithm;
+  this is known as the \emph{de Bruijn principle}.
+\item universality with respect to the particular problem: having a
+  certificate language for each domain of application would just shift
+  the problem of trust; having one universally accepted certificate
+  language that can be tailored to many problems puts a much stronger
+  trust on the unique verifier.
+\end{itemize}
+
+The Edimburgh Logical Framework \cite{harper1993framework}, or \LF\
+for short, is a well-accepted solution to this question. It is a typed
+lambda-calculus and can be viewed as a language for
+\emph{representing} terms, derivations and statements of $\Pi_0^1$
+meta-theorems (but not their proofs) in a given inference system
+supporting hypothetical judgements \cite{pfenning2001logical}. For
+these, it provides built-in support for encoding languages with
+binders or hypothesis (a feature called $\lambda$-tree syntax or HOAS
+\cite{pfenning1988higher}). The user gives a \emph{signature} defining
+abstract syntax and inference rules of his \emph{object language}, and
+gets an algorithm to verify terms and derivations expressed in the
+\LF\ syntax.
+
+This framework has been used to issue certificates for a number of
+practical applications like proof-carrying code
+\cite{necula1997proof}, authentication framework \cite{appel1999proof}
+\ldots %TODO ATP
+
+In all these applications, the user feeds the generation program with
+a query (the program to compile to machine-code, the access query, the
+statement of a theorem \ldots) and gets in return a result (machine
+code, yes/no\ldots) together with a certificate that the result
+verifies the wanted property (semantics preservation, justification
+for granting access, proof of the theorem\ldots). A potential client
+can then verify independently that this result is adequate given the
+query: it suffices to batch-verify the certificate.
+
+\subsection{Proof-carrying static analysis}
+
 \subsection{Storing computation results}
@@
 
+\section{Examples}
+
 \section{The calculus}
+%% My idea works (details, data) (5 pages)
 
@@
 The core calculus defines the \emph{values} we want to
-manipulate. These contain constants, and --- contrarily to \eg\ ML
-values --- binders and variables enabling $\lambda$-tree syntax (TODO
-cite). Thus, they embed a superficial notion of computation:
+manipulate. These contain constants, and --- contrarily to \eg\
+\sysname{ML} values --- binders and variables enabling $\lambda$-tree
+syntax. Thus, they embed a superficial notion of computation:
 hereditary substitution. This calculus is an extension of \emph{spine
-  canonical LF} (TODO cite) enriched with metavariables. We present
-its typing algorithm relatively to a signature. It features an
-explicit caching mechanism using metavariables.
+  canonical \LF} \cite{pfenning2007term} enriched with metavariables
+which are references to open terms as in
+\cite{nanevski2008contextual}. We present its typing algorithm
+relatively to a signature. It features an explicit caching mechanism
+using metavariables.
 
@@
 In this variant, only $\eta$-long $\beta$-normal forms have an
-existence: canonicity for $\beta$-reduction is enforced in the syntax;
-canonicity for $\eta$-expansion will be enforced by typing. Also,
-application of a \emph{head} to a \emph{spine} (\ie\ a constant or
-variable) is $n$-ary as opposed to the more standard binary
-application: this eases reasoning about substitution and
-$\eta$-expansion (TODO cite).
+existence: canonicity for $\beta$-reduction is enforced in the syntax
+and canonicity for $\eta$-expansion will be enforced by typing,
+following \cite{hl07mechanizing}. Besides, application of a
+\emph{head} (\ie\ a constant or variable) is $n$-ary (\ie\ to a
+\emph{spine} of objects) as opposed to the more standard binary
+application: this eases the definition substitution and
+$\eta$-expansion, and corresponds to a focused sequent calculus known
+as \sysname{LJT} \cite{herbelin1995λ}.
 
@@
 As is customary, we adopt the Barendregt convention for naming bound
-variables, write $A\to B$ for $\prd x A B$ when $x\notin \FV{B}$,
-$\meta X$ for $\smeta X \msubstnil$, and adopt indifferently list or
-map notations for $\sigma$ and $\Gamma$, \ie\ write $\sigma(x) = M$ or
-$\Gamma(x) = A$.
+variables, write $A\to B$ for $\prd x A B$ when $\var x\notin \FV{B}$,
+$\meta X$ for $\smeta X \msubstnil$, and adopt indifferently list, map
+or set notations for $\sigma$ and $\Gamma$. The identity substitution
+$\msubstid{\Gamma}$ is a notation for $\{\msubst x {\var x}\ |\ \var x
+\in\dom\Gamma\}$.
 
@@
 
-We overload \emph{substitution} as a partial operation on terms, given
+We overload \emph{substitution} as a partial operation on objects, given
 a parallel substitution $\sigma$. Since the syntax of object is not
@@
 is mutually recursive with a \emph{cut} function $\gcut M S$ which can
-in turn trigger a chain of substitutions, leading to a canonical term,
-a stuck substitution or divergence. % TODO vraiment?
+in turn trigger a chain of substitutions, leading to a canonical objects
+or an error.
 \begin{align*}
   \gsubst {(\lam x M)} \sigma &= \lam x {\gsubst M \sigma} &
-  \text{if $x\notin \dom{\sigma}$}
+  \text{if $\var x\notin \dom{\sigma}$}
   \\
@@
   \app {\var x} {\gsubst S \sigma} &
-  \text{if $x\notin \dom{\sigma}$}
+  \text{if $\var x\notin \dom{\sigma}$}
   \\
@@
   \gcut M {\gsubst S \sigma} &
-  \text{if $\sigma(x) = M$}
+  \text{if $\sigma(\var x) = M$}
   \\
@@
   \gcut {\lam x M} {\spinecons N S} &=
-  \gcut {\gsubst M N} S
+  \gcut {\gsubst M {\msubst x N}} S
   \\
@@
 We extend this operation trivially on families $\gsubst A \sigma$ and
-kinds $\gsubst K \sigma$. We also allow to abusively write the
-non-canonical application $\app M S$ as a shorthand for $\gcut M S$ if
-it is defined.
+kinds $\gsubst K \sigma$. We also allow to abusively write
+non-canonical applications $\app M S$ as a shorthand for $\gcut M S$
+if they are defined.
 
-Note that this notion of substitution presupposes $\eta$-long normal
-forms: a non-empty spine cut against to another non-empty spine $\gcut
-{\app H S} {S'}$ blocks the substitution.
+Note that this notion of substitution differs from hereditary
+substitution found in \eg\ \cite{hl07mechanizing} as it presupposes
+$\eta$-long normal forms: a non-empty spine cut against to another
+non-empty spine $\gcut {\app H S} {S'}$ blocks the substitution
+instead of reducing to $\app H {(\gconcat S S')}$ (an implicit
+$\eta$-expansion). For example, the reduction $\gcut {\lam x \app
+  {\var x}{\var x}} {\lam x \app {\var x}{\var x}}$ gets stuck because
+the second $\var x$ in the left part is substituted by a function yet
+is not itself a function.
 
@@
 % TODO stabilité de la eta?
+% TODO comparer avec la def usuelle
 
@@
 threading and maintaining a data structure of named partial
-derivations (a repository) throughout the typing process. We present a
-bidirectional typing algorithm updating an input repository.
+derivations (\emph{slices}) throughout the typing process. We present
+a bidirectional typing algorithm updating an input repository.
+
+A \emph{signature} declares object and type constants to be used in
+objects. To each object constant we assign an annotation $L$ setting
+the slicing behaviour of the constant during type-checking: either
+sliceable $\annsliceable$ (each application of this constant will be
+cached and given a name), non-sliceable $\annnonsliceable$ (we don't
+create a name for applications of this constant), or defined
+$\anndefined{T}$ (see \ref{sec:computational-calculus}).
+\begin{align*}
+  L &\gequal
+  \annsliceable \gor
+  \annnonsliceable \gor
+  \anndefined{T}
+  \\
+  \Sigma &\gequal
+  \gnil \gor
+  \gcons \Sigma {\cst c :^L A} \gor
+  \gcons \Sigma {\cst a :^L K}
+  &\text{Signature}
+\end{align*}
+We write $\Sigma(\cst c) = A; L$ for $(\cst c :^L A) \in \Sigma$
+suggesting the implementation of signatures as maps.
+
+Our algorithm types exactly the well-typed objects of the usual
+\LF, except that we thread a repository $\mr$ and store in it
+all well-typed applicative subterms $F$ along with their classifier
+$A$ and the environment in which they are well-typed.
+
+% TODO parler de la redondance en LF normal
 
@@
 
-A \emph{signature} declares object and type constants to be used in
-objects. To each object constant we assign an annotation $L$ setting
-the slicing behaviour of the constant during type-checking: either
-sliceable $\annsliceable$ (each application of this constant will have
-its own name in the repository), non-sliceable $\annnonsliceable$ (we
-don't create a name for applications of this constant), or defined
-$\anndefined{F}$ (see \ref{sec:computational-calculus}).
-\begin{align*}
-  L &\gequal
-  \annsliceable \gor
-  \annnonsliceable \gor
-  \anndefined{F}
-  \\
-  \Sigma &\gequal
-  \gnil \gor
-  \gcons \Sigma {\cst c :^L A} \gor
-  \gcons \Sigma {\cst a :^L K}
-  &\text{Signature}
-\end{align*}
-We write $\Sigma(\cst c) = A; L$ for $(\cst c :^L A) \in \Sigma$
-suggesting the implementation of signatures as maps.
+We refer to cached subterms by metavariables. We force instantiation
+of all variables in their environment by attaching a substitution to
+them ($\smeta X \sigma$).
+
+When checking large objects, the environment can grow fast. As an
+optimization to reduce the size of the stored environments and
+substitutions in slices $M$, we strengthen environments to their
+minimum size (the free variables of $M$): We define the operation of
+\emph{strengthening} from an environment $\Gamma$ and an object $M$
+well-typed in $\Gamma$ to the minimal environment in which $M$ can
+actually be typed, \ie\ containing only its free variables:
+$$
+\stren{\Gamma,M} = \{x : A\ |\ x\in \FV{M} \wedge \Gamma(x)=A\}
+$$
+% TODO lemme qui dit que c'est correct wrt variables libres des A et
+% du type
 
@@
 and $W$ the output. They are parameterized by a constant and implicit
-signature $\Sigma$.
+signature $\Sigma$, and all premisses are ordered in the algorithmical
+order of computation.
 
@@
     \infer{
-      \jstr{\Gamma\vdash\tp M A}{\Gamma'\vdash\tp {M'} {A'}}\sigma
+      \stren{\Gamma, M}=\Gamma'
       \and
@@
     }{
-      \jp{(\Delta; \smeta Z \sigma)}\Gamma M A {(\mapadd\Delta{\meta
-          X}{(\Gamma'\vdash M':A')}; \smeta X \sigma)}
+      \jp{(\Delta; \smeta Z {\sigma})}\Gamma M A {(\mapadd\Delta{\meta
+          X}{(\Gamma'\vdash M:A)}; \smeta X {\msubstid\Gamma})}
     }
@@
 \end{figure*}
+
+\begin{theorem}
+  If \/ $\jm\mr\Gamma M A \mmr {M'}$ then \/ $\jem\mmr\Gamma M {M'} A$
+\end{theorem}
+
+\begin{theorem}
+  If \/ $\jm\mr\Gamma M A \mmr {M'}$ then \/ $\jlfm\Gamma {(\checkoutr
+    M)} (\checkoutr A)$
+\end{theorem}
+
+% TODO example
 
@@
 
-We now turn to the definition of our computation calculus on LF
-objects. This language is untyped. % TODO??
+We now turn to the definition of our computation calculus. As we saw,
+it is attached to special, \emph{defined} constants in the signature
+and manipulates \LF\ objects. This language is un(i)typed, in that all
+\LF\ objects have the same implicit type $\cst{obj}$. We have the
+ability to construct objects out of other, and deconstruct them by
+case analysis.
 \begin{align*}
   T &\gequal
+  \lam x T \gor
+  U \\
+  U &\gequal
   F \gor
-  \matchin T \Gamma \casenobar {\app {\cst c} {\vec x}} T\ |\ \ldots\
-  \case {\app {\cst c} {\vec x}} T
+  \matchin U \Gamma \casenobar {\app {\cst c} {\vec x}} U\ |\ \ldots\
+  \case {\app {\cst c} {\vec x}} U
 \end{align*}
@@
 
-\section{Examples}
+\section{Related work}
+% 1-2 pages
+
+\section*{Conclusion and future work}
+% 0.5 pages
 

### 2012-05-05 doc/papers/LFMTP2012/paper.tex
paper

@@
 \usepackage{mathpartir}
+\usepackage{url}
 
@@
 
-\title{Certified incremental program transformations}
+\title{Incremental tactics}
 %\subtitle{Subtitle Text, if any}
@@
 
-  \caption{Typing algorithm}
+  \caption{Typing algorithm for the core calculus}
   \label{fig:typing-algo}
@@
 
-\begin{theorem}
-  If \/ $\jm\mr\Gamma M A \mmr {M'}$ then \/ $\jem\mmr\Gamma M {M'} A$
-\end{theorem}
+The main judgement is application typing $\jf\mr\Gamma F \mmr M A$,
+which means that in repository $\mr$ and environment $\Gamma$, the
+applicative term $F$ is well-typed of type $A$, and the slicing of $F$
+produces a new repository $\mmr$. The object $M$ is the
+\emph{residual} of this operation: only certain kind of applications
+are sliced, and it is possible that the ``tip'' of application $F$ is
+not sliced (because it is declared as non-sliceable). In that case,
+$M$ is this tip; if $F$ is directly sliceable, then $M$ is the
+metavariable referring to it.
 
-\begin{theorem}
-  If \/ $\jm\mr\Gamma M A \mmr {M'}$ then \/ $\jlfm\Gamma {(\checkoutr
-    M)} (\checkoutr A)$
-\end{theorem}
+Definitional equality judgement between two families $\jea\mr\Gamma A
+{A'}$ is used by these typing rules. This equivalence is usually
+defined to be just $\alpha$-equivalence since our objects are in
+$\beta\eta$-canonical form. Here, objects can contain metavariables
+defined in $\mr$, so we shall thread it. Moreover, we will see in next
+section how defined constants compute and thus influence definitional
+equality. Definitional equality is defined later in
 
+\paragraph{Example}
+
+Let us work out an example.
 % TODO example
@@
 
+\begin{theorem}
+  If \/ $\jm\mr\Gamma M A \mmr {M'}$ then \/ $\jem\mmr\Gamma M {M'} A$
+\end{theorem}
+
+\begin{theorem}
+  If \/ $\jm\mr\Gamma M A \mmr {M'}$ then \/ $\jlfm\Gamma {(\checkoutr
+    M)} (\checkoutr A)$
+\end{theorem}
+
 \subsection{The computation calculus}
@@
 ability to construct objects out of other, and deconstruct them by
-case analysis.
+case analysis:
 \begin{align*}
@@
   F \gor
-  \matchin U \Gamma \casenobar {\app {\cst c} {\vec x}} U\ |\ \ldots\
-  \case {\app {\cst c} {\vec x}} U
+  \matchin U \Gamma \casenobar {\app {H} {\vec x}} U\ |\ \ldots\
+  \case {\app {H} {\vec x}} U
 \end{align*}
+
+\begin{figure*}
+  \centering
+  TODO
+  \caption{Weak head-reduction}
+  \label{fig:whr}
+\end{figure*}
+
+\begin{figure*}
+  \centering
+  TODO
+  \caption{Definitional equality}
+  \label{fig:def-eq}
+\end{figure*}
 

### 2012-05-05 doc/papers/LFMTP2012/paper.tex
paper

@@
 As is customary, we adopt the Barendregt convention for naming bound
-variables, write $A\to B$ for $\prd x A B$ when $\var x\notin \FV{B}$,
+variables, write $A\to A'$ for $\prd x A A'$ when $\var x\notin \FV{A'}$,
 $\meta X$ for $\smeta X \msubstnil$, and adopt indifferently list, map
@@
   \gcons \Sigma {\cst c :^L A} \gor
-  \gcons \Sigma {\cst a :^L K}
+  \gcons \Sigma {\cst a : K}
   &\text{Signature}
@@
 
-The algorithm is presented in Fig. \ref{fig:typing-algo} as a
+The algorithm is presented in Fig. \ref{fig:obj-typing} as a
 syntax-directed inference system. All judgements are of the form $\gjo
@@
 
+  \fbox{$\jl\mr\Gamma A S \mmr{S'}{A'}$}
+  \qquad
+  {Spine typing}
+
   \fbox{$\jp\mr\Gamma M A \mmr$}
@@
 
-  \caption{Typing algorithm for the core calculus}
-  \label{fig:typing-algo}
+  \caption{Typing algorithm for objects}
+  \label{fig:obj-typing}
 \end{figure*}
@@
 {A'}$ is used by these typing rules. This equivalence is usually
-defined to be just $\alpha$-equivalence since our objects are in
+defined to be just $\alpha$-equivalence since objects are in
 $\beta\eta$-canonical form. Here, objects can contain metavariables
-defined in $\mr$, so we shall thread it. Moreover, we will see in next
-section how defined constants compute and thus influence definitional
-equality. Definitional equality is defined later in
+defined in $\mr$, so we shall thread it. Moreover, we will see in the
+next section how defined constants compute and thus influence
+definitional equality. Definitional equality is defined later in
+\ref{fig:def-eq}.
+
+Signatures themselves need to be checked. We present the checking
+algorithm in Fig. \ref{fig:sign-typing}.
+
+\begin{figure*}
+  \fbox{$\jsig\mr\Sigma\mmr{\Sigma'}$}
+  \qquad
+  {Signature}
+
+  \begin{mathpar}
+    \infer{ }{
+      \jsig\mr\gnil\mr\gnil
+    }
+
+    \infer{
+      \jsig\mr\Sigma\mmr{\Sigma'}
+      \and
+      \jfam\mmr\gnil {\!_{\Sigma'} A}\mmmr {A'}
+    }{
+      \jsig\mr{\gcons\Sigma{{\cst c :^L A}}}\mmmr{\gcons{\Sigma'}{{\cst c :^L A'}}}
+    }
+
+    \infer{
+      \jsig\mmr\Sigma\mmmr{\Sigma'}
+      \and
+      \jkind\mr\gnil {\!_{\Sigma'} K}\mmr {K'}
+    }{
+      \jsig\mr{\gcons\Sigma{{\cst a : K}}}\mmmr{\gcons{\Sigma'}{{\cst a : K'}}}
+    }
+  \end{mathpar}
+
+  \fbox{$\jfam\mr\Gamma A \mmr {A'}$}
+  \qquad
+  {Family}
+
+  \begin{mathpar}
+
+    \infer{
+      \jfam\mr\Gamma {A_1} \mmr {A_1'}
+      \and
+      \jfam\mmr{\gcons\Gamma{\tp {\var x} {A_1'}}} {A_2} \mmmr {A_2'}
+    }{
+      \jfam\mr\Gamma {\prd x {A_1} {A_2}} \mmmr {\prd x {A_1'} {A_2'}}
+    }
+
+    \infer{
+      \Sigma(a) = K
+      % TODO
+    }{
+      \jfam\mr\Gamma {\app{\cst a} S} \mmr {\app{\cst a} {S'}}
+    }
+
+  \end{mathpar}
+
+  \fbox{$\jkind\mr\Gamma K \mmr {K'}$}
+  \qquad
+  {Kind}
+
+  \begin{mathpar}
+
+  \end{mathpar}
+
+  \caption{Typing algorithm for signatures}
+  \label{fig:sign-typing}
+\end{figure*}
 

### 2012-05-07 doc/papers/LFMTP2012/paper.tex
paper

@@
 substitution found in \eg\ \cite{hl07mechanizing} as it presupposes
-$\eta$-long normal forms: a non-empty spine cut against to another
-non-empty spine $\gcut {\app H S} {S'}$ blocks the substitution
-instead of reducing to $\app H {(\gconcat S S')}$ (an implicit
-$\eta$-expansion). For example, the reduction $\gcut {\lam x \app
-  {\var x}{\var x}} {\lam x \app {\var x}{\var x}}$ gets stuck because
-the second $\var x$ in the left part is substituted by a function yet
-is not itself a function.
+$\eta$-long normal forms: once a cut starts, there must be exactly as
+many formal and actual arguments. A non-empty spine cut against
+another non-empty spine $\gcut {\app H S} {S'}$ blocks the
+substitution instead of reducing to $\app H {(\gconcat S S')}$ (an
+implicit $\eta$-expansion); conversely, exhausting actual but not
+formal arguments as in $\gcut{\lam x M}{\gnil}$ blocks the
+substitution too. For example, the reduction $\gcut {\lam x \app {\var
+    x}{\var x}} {\lam x \app {\var x}{\var x}}$ gets stuck because the
+second $\var x$ in the left part is substituted by a function yet is
+not itself a function.
 
@@
   \qquad
-  Object typing
+  Object
 
@@
        \and
-      \jea\mmr\Gamma A {A'}
+      \jea\mmr\Gamma {\app{\cst a} S} {A'}
     }{
-      \jm\mr\Gamma F A \mmr {M'}
+      \jm\mr\Gamma F {\app{\cst a} S} \mmr {M'}
     }
+    % TODO commenter type ground
 
@@
   \qquad
-  {Substitution typing}
+  {Substitution}
 
@@
   \qquad
-  {Application typing}
+  {Application}
 
@@
   \qquad
-  {Head typing}
+  {Head}
 
@@
   \qquad
-  {Spine typing}
+  {Spine}
+
+  \begin{mathpar}
+    \infer{
+      \jm\mr\Gamma M A \mmr {M'}
+      \and
+      \jl\mmr\Gamma {\gsubst{A'}{M'}} S \mmmr {S'} {A''}
+    }{
+      \jl\mr\Gamma {\prd x {A} {A'}} {\spinecons M S} \mmmr {\spinecons
+      {M'}{S'}} {A''}
+    }
+
+    \infer{ }{
+      \jl\mr\Gamma {\app{\cst a} S} \spinenil \mr \spinenil {{\app{\cst a} S}}
+    }
+  \end{mathpar}
 
@@
 next section how defined constants compute and thus influence
-definitional equality. Definitional equality is defined later in
+definitional equality. Definitional equality is defined later in Fig.
 \ref{fig:def-eq}.
@@
       \Sigma(a) = K
+      \and
+      \jfl\mr\Gamma K S \mmr {S'}
       % TODO
@@
     }
+  \end{mathpar}
 
+  \fbox{$\jfl\mr\Gamma K S \mmr {S'}$}
+  \qquad
+  {Family spine}
+
+  \begin{mathpar}
+    \infer{
+      \jm\mr\Gamma M A \mmr {M'}
+      \and
+      \jfl\mmr\Gamma {\gsubst{K}{M'}} S \mmmr {S'}
+    }{
+      \jfl\mr\Gamma {\prd x {A} {K}} {\spinecons M S} \mmmr {\spinecons
+      {M'}{S'}}
+    }
+
+    \infer{ }{
+      \jfl\mr\Gamma\type\spinenil\mr\spinenil
+    }
   \end{mathpar}
@@
   \begin{mathpar}
+    \infer{
+      \jfam\mr\Gamma A\mmr{A'}
+      \and
+      \jkind\mmr{\envcons\Gamma x A} K \mmmr {K'}
+    }{
+      \jkind\mr\Gamma{\prd x A K}\mmmr{\prd x {A'}{K'}}
+    }
 
+    \infer{ }{\jkind\mr\Gamma\type\mr\type}
   \end{mathpar}

### 2012-05-07 doc/papers/LFMTP2012/paper.tex
paper

@@
   \\
-  % TODO cas meta/S ????
+  % TODO cas meta/S ? rem: on en a pas besoin ds le proto.
   % \gcut {\smeta X\sigma} S &=
@@
 kinds $\gsubst K \sigma$. We also allow to abusively write
-non-canonical applications $\app M S$ as a shorthand for $\gcut M S$
-if they are defined.
+non-canonical application $\app M S$ as a shorthand for $\gcut M S$
+if it is defined.
 
@@
 % TODO pourquoi cette def?
-% TODO terminaison??
+% TODO terminaison?
 % TODO stabilité de la eta?
-% TODO comparer avec la def usuelle
 
@@
   \begin{mathpar}
-    \infer{ }{
-      \jsig\mr\gnil\mr\gnil
-    }
 
@@
       \jsig\mr{\gcons\Sigma{{\cst a : K}}}\mmmr{\gcons{\Sigma'}{{\cst a : K'}}}
+    }
+
+    \infer{ }{
+      \jsig\mr\gnil\mr\gnil
     }
@@
       \jfl\mr\Gamma K S \mmr {S'}
-      % TODO
     }{
@@
 
+\paragraph{Commit}
+% TODO commit
+
 \paragraph{Example}
@@
 \begin{theorem}
-  If \/ $\jm\mr\Gamma M A \mmr {M'}$ then \/ $\jlfm\Gamma {(\checkoutr
-    M)} (\checkoutr A)$
+  If \/ $\jm\mr\Gamma M A \mmr {M'}$ then \/ $\jlfm\Gamma {(\checkoutrr
+    M)} (\checkoutrr A)$
 \end{theorem}
@@
 We now turn to the definition of our computation calculus. As we saw,
-it is attached to special, \emph{defined} constants in the signature
-and manipulates \LF\ objects. This language is un(i)typed, in that all
-\LF\ objects have the same implicit type $\cst{obj}$. We have the
-ability to construct objects out of other, and deconstruct them by
-case analysis:
+code is attached to special, \emph{defined} constants in the signature
+that manipulates \LF\ objects. We have the ability to construct
+objects out of other, and deconstruct them by case analysis:
 \begin{align*}
@@
   F \gor
-  \matchin U \Gamma \casenobar {\app {H} {\vec x}} U\ |\ \ldots\
-  \case {\app {H} {\vec x}} U
+  \matchin U \Gamma \casenobar {\app {H} {\vec {\var x}}} U\ |\ \ldots\
+  \case {\app {H} {\vec {\var x}}} U
 \end{align*}
+
+This language is un(i)typed, in that all \LF\ objects have the same
+implicit type $\cst{obj}$: there is no static guarantee that terms
+produced by a constant $\cst c : A_1\to A_2$ will have type $A_2$; it
+is also non-terminating: we could declare a constant $\cst c : \cst
+a\to \cst a = \lam x \app {\cst c} {\var x}$. We expose here only the
+minimal building blocks, but other features could be added at will to
+this language. You can think of defined constants as black boxes
+taking and producing terms, and being able to call other black
+boxes. We detail in section \ref{sec:implem} how to embed
+\sysname{OCaml} as computation calculus.
+
+Nevertheless, in order to maintain the well-typing of the generated
+term, we want to maintain the invariant that arguments fed to these
+defined constant as well as the computation result have the announced
+type. We thus maintain \emph{dynamic typing} of computations.
 
@@
 \section{Implementation}
-
+\label{sec:implem}
 \section{Related work}

### 2012-05-08 doc/papers/LFMTP2012/paper.tex
paper

@@
 
-\subsection{Core calculus}
+\subsection{Value calculus}
 
-The core calculus defines the \emph{values} we want to
+The value calculus defines the \emph{values} we want to
 manipulate. These contain constants, and --- contrarily to \eg\
@@
   \lam x M \gor
-  F \gor
+  F &
+  \text{Object} \\
+  F &\gequal \app H S
+  \gor
   \smeta X \sigma &
-  \text{Object} \\
-  F &\gequal \app H S &
   \text{Application} \\
@@
 
+% TODO com M check/F infer
+
 Metavariables $\smeta X \sigma$ stand for open objects with free
-variables in $\dom{\sigma}$; they are defined in a repository (see
-below). We write $\FMV{M}$ for the set of metavariables in $M$ and
-speak of a \emph{metaclosed} object $M$ if $\FMV{M} = \emptyset$.
+variables in $\dom{\sigma}$; To each use of them is attached a
+substitution $\sigma$ defining these free variables. They are defined
+in a repository (see below). We write $\FMV{M}$ for the set of
+metavariables in $M$ and speak of a \emph{metaclosed} object $M$ if
+$\FMV{M} = \emptyset$.
 
@@
 \emph{acyclic}, that is for all $\meta X$ in $\mr$, the fixpoint
-of $\FMV{\meta X}$ does not contain $\meta X$. With $\mr =
+of $\FMV{\meta X}$ does not contain $\meta X$. If $\mr =
 (\Delta, X[\sigma])$, we write $\hat {\mr}$ for $X[\sigma]$,
@@
     \infer{
-      \jf\mr\Gamma F \mmr {M'} {A'}
+      \jf\mr\Gamma F \mmr {F'} {A'}
        \and
@@
     }{
-      \jm\mr\Gamma F {\app{\cst a} S} \mmr {M'}
+      \jm\mr\Gamma F {\app{\cst a} S} \mmr {F'}
     }
     % TODO commenter type ground
-
-    % Meta
-    \infer{
-      \mr(\meta X) = (\Gamma'\vdash M:A')
-      \and
-      \js\mr\Gamma \sigma {\Gamma'} \mmr {\sigma'}
-      \and
-      \jea\mmr\Gamma A {\gsubst {A'} {\sigma'}}
-    }{
-      \jm\mr\Gamma {\smeta X \sigma} A \mmr {M'}
-    }
   \end{mathpar}
@@
 
-  \fbox{$\jf\mr\Gamma F \mmr M A$}
+  \fbox{$\jf\mr\Gamma F \mmr F A$}
   \qquad
@@
 
+    % Meta
+    \infer{
+      \mr(\meta X) = (\Gamma'\vdash M:A)
+      \and
+      \js\mr\Gamma \sigma {\Gamma'} \mmr {\sigma'}
+    }{
+      \jf\mr\Gamma {\smeta X \sigma} \mmr {\smeta X {\sigma'}} {\gsubst{A}\sigma'}
+    }
+
     % Sliceable
@@
       \and
-      \jp\mmr\Gamma {\app H S} A \mmmr
+      \stren{\Gamma, \app H {S'}}=\Gamma'
+      \and
+      \text{$\meta X$ fresh in $\mr$}
     }{
-      \jf\mr\Gamma {\app H S} \mmmr {\hat\mmmr} {A'}
+      \jf\mr\Gamma {\app H S} {\mapadd\mmr{\meta X}{(\Gamma'\vdash
+          {\app H {S'}}:A)}} {\smeta X{\msubstid{\Gamma'}}} {A'}
     }
@@
 
-  \fbox{$\jp\mr\Gamma M A \mmr$}
-  \qquad
-  {Push}
-
-  \begin{mathpar}
-    \infer{
-      \stren{\Gamma, M}=\Gamma'
-      \and
-      \text{$\meta X$ fresh in $\Delta$}
-    }{
-      \jp{(\Delta; \smeta Z {\sigma})}\Gamma M A {(\mapadd\Delta{\meta
-          X}{(\Gamma'\vdash M:A)}; \smeta X {\msubstid\Gamma})}
-    }
-  \end{mathpar}
-
   \caption{Typing algorithm for objects}
@@
 
-The main judgement is application typing $\jf\mr\Gamma F \mmr M A$,
+The main judgement is application typing $\jf\mr\Gamma F \mmr {F'} A$,
 which means that in repository $\mr$ and environment $\Gamma$, the
 applicative term $F$ is well-typed of type $A$, and the slicing of $F$
-produces a new repository $\mmr$. The object $M$ is the
+produces a new repository $\mmr$. The object $F'$ is the
 \emph{residual} of this operation: only certain kind of applications
@@
 
-Definitional equality judgement between two families $\jea\mr\Gamma A
-{A'}$ is used by these typing rules. This equivalence is usually
-defined to be just $\alpha$-equivalence since objects are in
+These typing rules use definitional equality judgement between two
+families $\jea\mr\Gamma A {A'}$. This equivalence is usually defined
+to be just $\alpha$-equivalence since objects are in
 $\beta\eta$-canonical form. Here, objects can contain metavariables

### 2012-05-08 doc/papers/LFMTP2012/paper.tex
paper

@@
 %% It’s an interesting problem (1 page)
-
-As programming languages develop more static analysis like rich type
-systems, programs get safer as bugs are caught earlier.
 
@@
   F \gor
-  \matchin U \Gamma \casenobar {\app {H} {\vec {\var x}}} U\ |\ \ldots\
-  \case {\app {H} {\vec {\var x}}} U
+  \matchin U \Gamma C \\
+  C &\gequal \casenobar P U \gor
+  C\ \case P U \\
+  P &\gequal
+  \app H {\var x} \ldots\ {\var x}
 \end{align*}
 
-This language is un(i)typed, in that all \LF\ objects have the same
-implicit type $\cst{obj}$: there is no static guarantee that terms
-produced by a constant $\cst c : A_1\to A_2$ will have type $A_2$; it
-is also non-terminating: we could declare a constant $\cst c : \cst
-a\to \cst a = \lam x \app {\cst c} {\var x}$. We expose here only the
-minimal building blocks, but other features could be added at will to
-this language. You can think of defined constants as black boxes
-taking and producing terms, and being able to call other black
-boxes. We detail in section \ref{sec:implem} how to embed
-\sysname{OCaml} as computation calculus.
+This language is un(i)typed, in that all \LF\ objects are viewed as
+having the same implicit type $\cst{obj}$: there is no static
+guarantee that terms produced by a constant $\cst c : A_1\to A_2$ will
+have type $A_2$. It is also non-terminating: we could declare a
+recursive constant $\cst c : \cst a\to \cst a = \lam x \app {\cst c}
+{\var x}$. We expose here only the minimal building blocks of this
+language, but really any language with algebraic datatypes would do
+since we will be manipulating directly the first-order representation
+of \LF\ objects. In fact, whatever the language is, you can think of
+defined constants as mere black boxes taking and producing terms, and
+being able to call other black boxes. We detail for example in section
+\ref{sec:implem} how to embed \sysname{OCaml} as computation calculus.
 
 Nevertheless, in order to maintain the well-typing of the generated
-term, we want to maintain the invariant that arguments fed to these
-defined constant as well as the computation result have the announced
-type. We thus maintain \emph{dynamic typing} of computations.
+term, we want to maintain the invariant that each time a piece of term
+goes in or out of a black box, it is checked against its specification
+(\ie\ its type). We thus maintain \emph{dynamic typing} of
+computations: arguments and results are type-checked at run-time.
+
+\paragraph{Substitution}
+
+\begin{align*}
+  \gsubst {(\lam x T)} \sigma &= \lam x {\gsubst T\sigma} \qquad
+  \text{if $\var x\notin\dom\sigma$}\\
+  \gsubst {(\matchin U \Gamma C)} \sigma &=
+  \matchin {\gsubst
+    U\sigma} {\gsubst\Gamma\sigma} {\gsubst C\sigma} \\
+  \gsubst {(\casenobar{\app H {\spinecons {\var x_1} {\spinecons \ldots {\var
+          x_n}}}} U)} \sigma &= {\casenobar{\app H {\spinecons {\var x_1} {\spinecons \ldots {\var
+          x_n}}}} \gsubst U \sigma} \quad \text{if $\var x_i\notin\dom\sigma$} \\
+\end{align*}
 
 \begin{figure*}
-  \centering
-  TODO
+  \fbox{$\jcc\mr\Gamma T S A \mmr F$}
+  \qquad
+  {Computational cut}
+
+  \begin{mathpar}
+    \infer{
+      \jm\mr\Gamma M A \mmr {M'}
+      \and
+      \jcc\mmr\Gamma{\gsubst T {\msubst x M}} S {A_2}\mmmr F
+    }{
+      \jcc\mr\Gamma{\lam x T}{\spinecons M S}{(\prd x {A_1} {A_2})}
+      \mmmr F
+    }
+
+    \infer{
+      \jcr\mr\Gamma U \mmr F
+      \and
+      \jm\mmr\Gamma F {\app {\cst a} S} \mmmr {F'}
+    }{
+      \jcc\mr\Gamma U \spinenil {\app {\cst a} S} \mmmr {F'}
+    }
+  \end{mathpar}
+
+  \fbox{$\jcr\mr\Gamma U \mmr F$}
+  \qquad
+  {Computational reduction}
+
+  \begin{mathpar}
+    \infer{ }{
+      \jcr\mr\Gamma{\app {\var x} S} \mr {\app {\var x} S}
+    }
+
+    \infer{
+      \Sigma(c) = A; (\annsliceable\text{ or }\annnonsliceable)
+    }{
+      \jcr\mr\Gamma{\app {\cst c} S} \mr {\app {\cst c} S}
+    }
+
+    \infer{
+      \Sigma(c) = A; \anndefined T
+      \and
+      \jcc\mr\Gamma T S A \mmr F
+    }{
+      \jcr\mr\Gamma{\app {\cst c} S} \mr {\app {\cst c} S}
+    }
+
+    \infer{
+      \jcr\mr{\gconcat\Gamma{\Gamma'}} U \mmr {\app {H_i} {\spinecons
+          {M_1}{\spinecons\ldots{M_{m_i}}}}}
+      \and
+      \jcr\mmr\Gamma {\gsubst{U_i}{\msubst {x$_1$} {M_1} \ldots\
+          \msubst {x$_{m_i}$} {M_{m_i}}}} \mmmr F a
+    }{
+      \jcr\mr\Gamma{\matchin{U}{\Gamma'}{
+          \casenobar{\app {H_1}{\spinecons {\var x_{11}} {\spinecons
+                \ldots\ {\var x_{1m_1}}}}}{U_1}
+          |\ \ldots\
+          \case{\app {H_n}{\spinecons {\var x_{n1}} {\spinecons
+              \ldots\ {\var x_{nm_n}}}}}{U_n}
+        }}\mmr F
+    }
+
+  \end{mathpar}
+
   \caption{Weak head-reduction}

### 2012-05-09 doc/papers/LFMTP2012/paper.tex
paper

@@
   F &
-  \text{Object} \\
+  \text{Canonical object} \\
   F &\gequal \app H S
@@
   \smeta X \sigma &
-  \text{Application} \\
+  \text{Atomic object} \\
   H &\gequal
@@
 
-% TODO com M check/F infer
-
 Metavariables $\smeta X \sigma$ stand for open objects with free
@@
 $\FMV{M} = \emptyset$.
+
+We distinguish syntactically between \emph{checkable} and
+\emph{inferrable} objects (resp. \emph{canonical} and \emph{atomic}),
+an idea dating back to \cite{pierce2000local}: canonical objects are
+checked against a given type, whereas we can synthesize the type of
+atomic objects; the coercion from the first to the second will be the
+place where to verify equality between the synthesized type and the
+given one.
 
@@
 
+% TODO parler d'effacer les R et retomber sur LF
+% TODO expl. c'est tout comme LF sauf qqes règles...
 % TODO parler de la redondance en LF normal
@@
   \qquad
-  Object
+  Canonical object
 
@@
 
-  \fbox{$\js\mr\Gamma\sigma{\Gamma'}\mmr{\sigma'}$}
-  \qquad
-  {Substitution}
-
-  \begin{mathpar}
-
-    % Cons
-    \infer{
-      \js\mr\Gamma\sigma{\Gamma'}\mmr{\sigma'}
-      \and
-      \jm\mmr\Gamma M {\gsubst A {\sigma'}} \mmmr {M'}
-    }{
-      \js\mr\Gamma{(\msubstcons \sigma x M)} {(\envcons {\Gamma'} x {A'})}
-      \mmmr {(\msubstcons {\sigma'} x {M'})}
-    }
-
-    % Nil
-    \infer{ }{
-      \js\mr\Gamma\msubstnil\envnil\mr\envnil
-    }
-
-  \end{mathpar}
-
   \fbox{$\jf\mr\Gamma F \mmr F A$}
   \qquad
-  {Application}
+  {Atomic object}
 
@@
       \stren{\Gamma, \app H {S'}}=\Gamma'
-      \and
-      \text{$\meta X$ fresh in $\mr$}
     }{
@@
     }
+    \quad\text{($\meta X$ fresh in $\mr$)}
 
@@
 
+  \fbox{$\js\mr\Gamma\sigma{\Gamma'}\mmr{\sigma'}$}
+  \qquad
+  {Substitution}
+
+  \begin{mathpar}
+
+    % Cons
+    \infer{
+      \js\mr\Gamma\sigma{\Gamma'}\mmr{\sigma'}
+      \and
+      \jm\mmr\Gamma M {\gsubst A {\sigma'}} \mmmr {M'}
+    }{
+      \js\mr\Gamma{(\msubstcons \sigma x M)} {(\envcons {\Gamma'} x {A'})}
+      \mmmr {(\msubstcons {\sigma'} x {M'})}
+    }
+
+    % Nil
+    \infer{ }{
+      \js\mr\Gamma\msubstnil\envnil\mr\envnil
+    }
+
+  \end{mathpar}
+
   \caption{Typing algorithm for objects}
@@
 
-The main judgement is application typing $\jf\mr\Gamma F \mmr {F'} A$,
-which means that in repository $\mr$ and environment $\Gamma$, the
+The main judgement is atomic object typing $\jf\mr\Gamma F \mmr {F'}
+A$, which means that in repository $\mr$ and environment $\Gamma$, the
 applicative term $F$ is well-typed of type $A$, and the slicing of $F$

### 2012-05-10 doc/papers/LFMTP2012/paper.tex
paper

@@
 %% It’s an interesting problem (1 page)
+
+When writing critical software, high assurance against bugs is an
+absolute necessity. To certify the absence of bugs in a large body of
+code given its specification, human reviewing is one option, but
+certainly a failible one. Formal, mathematical proof that our program
+respects a certain \emph{dependability claim} gives us the assurance
+that ``its trustworthyness [\ldots] would presumably last for eternity
+\cite{shao2010certified}''.
+
+Certifying
+
+\cite{miller2011proposal}
 
@@
 
+    % Defined
+    \infer{
+      \jh\mr\Gamma H A {\anndefined T}
+      \and
+      \jl\mr\Gamma A S {\mmr} {S'} {A'}
+      \and
+      \jcc\mr\Gamma A T S \mmmr F
+      \and
+      \jm\mmmr\Gamma F {A'} {\repo {R'''} } {F'}
+    }{
+      \jf\mr\Gamma {\app H S} {\repo{R'''}} {F'} {A'}
+    }
+
   \end{mathpar}
@@
     \infer{
-      \jm\mr\Gamma M A \mmr {M'}
+      \jm\mr\Gamma M {A_1} \mmr {M'}
       \and
-      \jcc\mmr\Gamma{\gsubst T {\msubst x M}} S {A_2}\mmmr F
+      \jcc\mmr\Gamma{\gsubst {A_2} {\msubst x {M'}}}{\gsubst T {\msubst x {M'}}} S \mmmr F
     }{
-      \jcc\mr\Gamma{\lam x T}{\spinecons M S}{(\prd x {A_1} {A_2})}
+      \jcc\mr\Gamma{\prd x {A_1} {A_2}}{\lam x T}{\spinecons M S}
       \mmmr F
@@
     }{
-      \jcc\mr\Gamma U \spinenil {\app {\cst a} S} \mmmr {F'}
+      \jcc\mr\Gamma {\app {\cst a} S} U \spinenil \mmmr {F'}
     }

### 2012-05-11 doc/papers/LFMTP2012/paper.tex
paper + new Test_check3

@@
 %% It’s an unsolved problem (1 page)
+
+A student is learning about System \sysname T, and wants to write a
+program to compute $4!$. He (or she) knows that System \sysname T is a
+typed language, and wants to take advantage of this to build his
+program incrementally using the teacher's provided incremental
+verifier. As a first attempt, he writes the following program $P_1$:
+$$
+\recb{\s\z}{\s\z}{\lam x\lam y \s {\var x}}
+$$
+The system type-checks it correctly by constructing a typing
+derivation $\md_1$ of \/ $\vdash P_1 : \cst{nat}$, valid in an
+empty context, and names all subderivations constructed for future
+reuse. The student then decides to factorize a definition for
+$\var{add}$ out of his code. He writes:
+\begin{align*}
+  &\letb{add}{\tlam x {\cst{nat}} \tlam y {\cst{nat}} \recb {\var x}
+    {\var y} {\lam x \lam y \s{\var x}}}
+  \\
+  &\quad\app{\var{add}}\app{(\s\z)}{(\s\z)} \ \text{.}
+\end{align*}
+But he (or his text editor) realizes that some parts of this program
+have not changed: both $\s\z$, and the function $\lam x\lam y\s{\var
+  x}$. The system must then have a name for the already-built
+associated derivations $\md_2$, $\md_3$ and $\md_4$. Provided there is
+an operator $|\md|$ from the derivation to the program it types, what
+is really to be checked is:
+\begin{align*}
+  &\letb{add}{\tlam x {\cst{nat}} \tlam y {\cst{nat}} \recb {\var x}
+    {\var y} {|\md_4|}}
+  \app{\var{add}}\app{|\md_2|}{|\md_2|} \ \text{,}
+\end{align*}
+that is, only the subterms of the program that changed. Derivations
+$\md_2$ and $\md_4$ used to live in the empty context, we are then
+free to use it in the new context where
+$\var{add}:\cst{nat}\to\cst{nat}$, because System \sysname T enjoys
+the \emph{weakening} property. So the system answers with a new
+derivation made out of pieces of the old one, without having to
+recheck the already built ones. It also gives the name $\md_5$ to the
+derivation for $\tlam x {\cst{nat}} \tlam y {\cst{nat}} \recb {\var x}
+{\var y} {|\md_4|}$, $\md_6$ to the arguments $\app{|\md_2|}{|\md_2|}$
+and $\md_7$ to the call $\app{\var{add}} {|\md_6|}$. Even if it
+occured in the scope of $\var{add}$, $\md_6$ doesn't depend on the
+fact that $\vdash \var{add}:\cst{nat}\to\cst{nat}$, but
+$\md_7$ does: %TODO
+. The
+student then types:
+\begin{align*}
+  &\letb{add}{ |\md_5| }
+  \\
+  &\letb{mul}{ \tlam x {\cst{nat}} \tlam y
+    {\cst{nat}} \recb \z {\var y} {\lam z \lam t
+      \app{\var{add}}\app{\var x} {\var z}} }
+  \\
+  &\app{\var{mul}}\app{|\md_2|}{|\md_2|}
+\end{align*}
+to see if his definition of multiplication is well-typed.
 

### 2012-05-11 doc/papers/LFMTP2012/paper.tex
paper (background)

--- 2012-05-10 doc/papers/LFMTP2012/paper.tex
@@
 %% It’s an unsolved problem (1 page)
+
+\paragraph{Motivational example}
+
+A student is learning about System \sysname T, and wants to write a
+program computing $(2+2)\times 3$. He (or she) knows that sometimes those
+$\cst{rec}$ arguments are quite hard to get correct, but System
+\sysname T is a typed language; he wants to take advantage of this to
+build his program incrementally using the teacher's provided
+incremental verifier. As a first attempt, he wants to verify $P_1$:
+$$
+\recb{\s\z}{\s\z}{\lam x\lam y \s {\var x}}\ \text{.}
+$$
+For this, he sends the command $\infer{P_1}$ to the system. The
+$\infer{}$ operator takes a well-formed program and tries to construct
+a valid typing derivation for its argument in the current, empty
+context. The system succeeds and returns a typing derivation $\md_1$
+of \/ $\vdash P_1 : \cst{nat}$, valid in an empty context.
+
+The student then decides to factorize a definition for $\var{add}$ out
+of his code:
+\begin{align*}
+  &\letb{add}{\tlam x {\cst{nat}} \tlam y {\cst{nat}} \recb {\var x}
+    {\var y} {\lam x \lam y \s{\var x}}}
+  \\
+  &\quad\app{\var{add}}\app{(\s\z)}{(\s\z)} \ \text{.}
+\end{align*}
+But he (or his text editor) realizes that some parts of this program
+have not changed: both $\s\z$, and the function $\lam x\lam y\s{\var
+  x}$, and he could reuse the pieces of derivation just built. Let us
+call them $\md_2$, $\md_3$ and $\md_4$. Provided there is an operator
+$\get\md$ from the derivation to the program it types, what needs really
+to be sent to the system is:
+\begin{align*}
+  &\infer{(
+    \letb{add}{\tlam x {\cst{nat}} \tlam y {\cst{nat}} \recb {\var x}
+      {\var y} {\get{\md_4}}}
+    \app{\var{add}}\app{\get{\md_2}}{\get{\md_2}}
+  )}
+\ \text{,}
+\end{align*}
+that is, only the subterms of the program that changed. Derivations
+$\md_2$ and $\md_4$ used to live in the empty context, we are then
+free to use it in the new context where
+$\var{add}:\cst{nat}\to\cst{nat}$, because System \sysname T enjoys
+the \emph{weakening} property. Taking advantage of $\md_2$ and $\md_4$
+the system succeeds with a new derivation made out of these, without
+having to regenerate them because it knows that $\infer{\get\md} =
+\md$. Let us call $\md_5$ the subderivation generated for $\tlam x
+{\cst{nat}} \tlam y {\cst{nat}} \recb {\var x} {\var y}
+{\get{\md_4}}$, $\md_6$ for the arguments
+$\app{\get{\md_2}}{\get{\md_2}}$ and $\md_7$ for the call
+$\app{\var{add}} {\get{\md_6}}$. Even if it occurs originally in the
+scope of $\var{add}$, $\md_6$ doesn't depend on the hypothesis $\var
+H$ that $\vdash \var{add}:\cst{nat}\to\cst{nat}$ because of the
+\emph{strengthening} property. Contrarily, $\md_7$ does: each time we
+want to reuse it, we will have to provide a proof of this fact.
+
+The student then provides a definition for multiplication, and writes
+the operation $(2 + 2)\times3$:
+\begin{align*}
+  \infer{}(
+  &\letb{add}{\get{\md_5}}
+  \\
+  &\quad\letb{mul} { \tlam x {\cst{nat}} \tlam y
+    {\cst{nat}} \recb \z {\var y} {\lam z \lam t
+      \app{\var{add}}\app{\var x} {\var z}} }
+  \\
+  &\quad\quad\app{\var{mul}}\app
+  {(\get{\gsubst{\md_7}{\msubst H {\infer{\var{add}}}}})}
+  {(\s{\get{\md_2}})}
+  )
+  \ \text{.}
+\end{align*}
+Here again, he took care of reusing all derivations for equal
+subterms. In particular, $\md_7$ expects a proof that the variable
+$\var{add}$ is of type $\cst{nat}\to\cst{nat}$; he provides
+$\infer{\var{add}}$ meaning: at this point we will have generated (or
+reused in this case) the derivation of $\var{add}$, so just use it.
+
+Finally, he decides (in an odd move) to inline the definition of
+$\var{mul}$ into its call site. Call $\md_8$ the derivation of the
+definition of $\var{mul}$ (it depends on an $\var H$ of
+$\vdash\var{add}:\cst{nat}\to\cst{nat}$) and $\md_9$ the derivation of
+$(2+2)\times 3$ depending on derivations $\var{H}$ and $\var{H'}$ for $+$ and
+$\times$. He orders:
+\begin{align*}
+\infer{}(
+\letb{add}{\get{\md_5}}
+\get{({\gsubst{\md_9} {\msubst {H} {(\get{\var{add}})}; \msubst {H'}
+      {\gsubst{\md_8}{\msubst H {(\get{\var{add}})}}}}})}
+)
+  \ \text{.}
+\end{align*}
+The definition for $\var{add}$ is unchanged, but the instantiation of
+open derivations enables to directly substitute one in the other and
+get the appropriate program: call $\md_{10}$ the whole derivation just
+returned, $\get{\md_{10}}$ returns:
+\begin{align*}
+  &\letb{add}{\tlam x {\cst{nat}} \tlam y {\cst{nat}} \recb {\var x}
+    {\var y} {\lam x \lam y \s{\var x}}}
+  \\&\quad
+  {\tlam x {\cst{nat}} \tlam y
+    {\cst{nat}} \recb \z {\var y} {\lam z \lam t
+      \app{\var{add}}\app{\var x} {\var z}}}
+  \\&\quad\quad
+  \app{(\app {\var{add}} \app {(\s{\s\z})}
+    {(\s{\s\z})})}{(\s{\s{\s\z}})}
+  \ \text{.}
+\end{align*}
+
+\paragraph{Discussion}
+
+% scale to big programs
+% memoization ds le cas HO
+%   nécessité de fcts inverses?
+% séparer generation des diffs et vérification
 

### 2012-05-12 doc/papers/LFMTP2012/paper.tex
paper

@@
 
-When writing critical software, high assurance against bugs is an
-absolute necessity. To certify the absence of bugs in a large body of
-code given its specification, human reviewing is one option, but
-certainly a failible one. Formal, mathematical proof that our program
-respects a certain \emph{dependability claim} gives us the assurance
-that ``its trustworthyness [\ldots] would presumably last for eternity
-\cite{shao2010certified}''.
-
-Certifying
-
-\cite{miller2011proposal}
-
 %% Describe the problem (Use an example)
@@
   x}$, and he could reuse the pieces of derivation just built. Let us
-call them $\md_2$, $\md_3$ and $\md_4$. Provided there is an operator
-$\get\md$ from the derivation to the program it types, what needs really
-to be sent to the system is:
+call these $\md_2$, $\md_3$ and $\md_4$. Provided there is an operator
+$\get\md$ from the derivation to the program it types, what needs
+really to be sent to the system is:
 \begin{align*}
@@
 $\var{add}:\cst{nat}\to\cst{nat}$, because System \sysname T enjoys
-the \emph{weakening} property. Taking advantage of $\md_2$ and $\md_4$
-the system succeeds with a new derivation made out of these, without
-having to regenerate them because it knows that $\infer{\get\md} =
-\md$. Let us call $\md_5$ the subderivation generated for $\tlam x
-{\cst{nat}} \tlam y {\cst{nat}} \recb {\var x} {\var y}
-{\get{\md_4}}$, $\md_6$ for the arguments
-$\app{\get{\md_2}}{\get{\md_2}}$ and $\md_7$ for the call
-$\app{\var{add}} {\get{\md_6}}$. Even if it occurs originally in the
-scope of $\var{add}$, $\md_6$ doesn't depend on the hypothesis $\var
-H$ that $\vdash \var{add}:\cst{nat}\to\cst{nat}$ because of the
+the \emph{weakening} property. Taking advantage of the fact that the
+judgement $\md_2$ and $\md_4$ proves is statically known, the system
+succeeds with a new derivation made out of these, without having to
+regenerate them because it knows that $\infer{\get\md} = \md$. Let us
+call $\md_5$ the subderivation generated for $\tlam x {\cst{nat}}
+\tlam y {\cst{nat}} \recb {\var x} {\var y} {\get{\md_4}}$, $\md_6$
+for the arguments $\app{\get{\md_2}}{\get{\md_2}}$ and $\md_7$ for the
+call $\app{\var{add}} {\get{\md_6}}$. Even if it occurs originally in
+the scope of $\var{add}$, $\md_6$ doesn't depend on the hypothesis
+$\var H$ that $\vdash \var{add}:\cst{nat}\to\cst{nat}$ because of the
 \emph{strengthening} property. Contrarily, $\md_7$ does: each time we
@@
 \end{align*}
-Here again, he took care of reusing all derivations for equal
-subterms. In particular, $\md_7$ expects a proof that the variable
-$\var{add}$ is of type $\cst{nat}\to\cst{nat}$; he provides
-$\infer{\var{add}}$ meaning: at this point we will have generated (or
-reused in this case) the derivation of $\var{add}$, so just use it.
+Here again, he took care of reusing all possible subderivations. In
+particular, $\md_7$ expects a proof that the variable $\var{add}$ is
+of type $\cst{nat}\to\cst{nat}$; he provides $\infer{\var{add}}$
+meaning: at this point we will have generated (or reused in this case)
+the derivation of $\var{add}$, so just use it.
 
@@
 
-% scale to big programs
-% memoization ds le cas HO
-%   nécessité de fcts inverses?
-% séparer generation des diffs et vérification
+Although this small example may seem anecdotal, the interest of a
+mechanism for reusing results of derivation generation becomes clear
+on very large programs: then the cost of rechecking well-typing of a
+small modification amounts to recheking the path in the term from the
+root to the changed part, and the changed part itself.
+
+A type-checker is usually viewed as an algorithm returning a binary
+answer (ok/error), specified by a set of inference rules. Seeing
+type-checking as issueing a completely explicit typing derivation
+enables to think of the relationship between the input and the
+output. Feeding part of the output $\md_i$ back to the input $P_i$ is
+a well-known technique for handling changes in the input: combined
+with a way to automatically recognize already-computed input, it is
+\emph{memoization}, a general-purpose technique for incremental
+computation that has been extensively studied and combined with other
+techniques \cite{pugh1989incremental, acar2003selective}. However, it
+is best understood in the case of purely first-order data, and to our
+best knowledge remains unexplored for higher-order structures
+containing binders, like programs and proofs. In this case,
+
+To simplify the problem,
+we separate the task in:
+\begin{itemize}
+\item from a modified program, generate the \emph{diff} against the
+  previous version, that is, observing and linking to unchanged
+  derivations.
+\item given a \emph{diff}, construct a new derivation and verify it.
+\end{itemize}
+The latter is the subject of this paper.
 
@@
 %% Here is my idea (2 pages)
+
+% lambda-lifting
 

### 2012-05-25 doc/papers/LFMTP2012/paper.tex
paper: intro wip

--- 2012-05-11 doc/papers/LFMTP2012/paper.tex
@@
 
-When writing critical software, high assurance against bugs is an
-absolute necessity. To certify the absence of bugs in a large body of
-code given its specification, human reviewing is one option, but
-certainly a failible one. Formal, mathematical proof that our program
-respects a certain \emph{dependability claim} gives us the assurance
-that ``its trustworthyness [\ldots] would presumably last for eternity
-\cite{shao2010certified}''.
+%% Describe the problem (Use an example)
 
-Certifying
+The positive impact of rich type systems on programming language
+design is not to be demonstrated anymore, both in terms of the safety
+of programs --- they ``do not go wrong'', and programming practices
+\eg\ modularity, early bug catching. Verifying well-typing of
+real-world programs is thus a critical matter, but it is a complex
+operation, for the compiler designer and for the human operator.
 
-\cite{miller2011proposal}
+\paragraph{For the compiler designer}
 
-%% Describe the problem (Use an example)
+Type systems are usually presented in a declarative manner as a set of
+inference rules, together with proofs that all programs accepting a
+derivation in this system respect a certain dependability claim, the
+famous ``progress and preservation'' pair. Yet, actual algorithms of
+verification often differ very much from their declarative
+description. This is of course true when programs are provided with
+little type annotations (type inference), but is also true in many
+cases for what is usually called ``type checking'', or
+``church-style'' calculi. Take for example dependent type systems,
+relying on a certain notion of definitional equality on types, or type
+systems with subtype polymorphism: both rely on a rule of the form
+$$
+\infer{\Gamma\vdash M : A \\ A\leq B}{\Gamma\vdash M : B}\ \text{,}
+$$
+for a given notion of $\leq$, which is not syntax-directed. The
+implementer then has to convert the declarative system to an
+algorithm, and convince himself that both are equivalent. This task,
+often neglected, is nonetheless highly non-trivial and non-modular.
+
+The most common option is to formally prove the equivalence, \ie\ that
+the (typing) relation is actually a function. It is of course the most
+satisfactory because it allows to ``forget'' about the declarative
+system and take only the result of the algorithm into consideration,
+although these proofs can be difficult to obtain. But for the sake of
+the discussion, let us choose another path: derivations in the
+declarative systems, if they cannot always be algorithmically
+inferred, can be \emph{verified} if provided by an oracle. This way,
+even if the oracle is wrong, we have a way to decide it. Let us then
+imagine a type-checker to be the pair of 1/ an untrusted procedure
+$\finfer M$ for inferring a typing derivations out of programs $M$; it
+doesn't have to be complete, or even sound, and 2/ a trusted kernel
+$\fcheck{\md}$ for verifying these derivations in a given set of
+inference rules $\Sigma$. This scheme has the following advantages
+over the first approach:
+\begin{itemize}
+\item it is lightweight: the untrusted algorithm is usually shorter
+  than the corresponding equivalence proof
+\item it is certifying: if $\fcheck{\finfer M} = \cst{true}$ then
+  there really \emph{is} a derivation $\md$ for $M$. Of course the
+  converse is not necessarily true: function $\finfer{}\!\!$ could
+  fail, or have bugs (return an incorrect derivation).
+\item it inherits benefits from certifying approaches like
+  proof-carrying code \cite{necula1997proof}: %TODO third-party
+                                              %verification
+\item for optimization purpose, compilers often need to propagate type
+  information in the chain of transformation; the typing derivation
+  produced by $\finfer{}$ being explicit, it could constitute the data
+  structure being exploited.
+\item \ldots\ but there is more:
+\end{itemize}
+
+\paragraph{For the programmer}
+
+Compilers are generally monolithic programs, taking source code and
+producing machine code in a series of transformations, one of the
+first being type checking. But as type systems become richer, it
+becomes increasingly difficult for the programmer to write a correct
+program in one try, compile it and run it: writing a program longer
+than a few dozen of lines becomes a tight and non-linear interaction
+between the human and the type-checker involving constant experiments,
+fixes, etc. If this interaction is possible for short programs and
+fast checkers, it becomes increasingly difficult %TODO
+
 %% State your contributions (Bulleted list of contributions)
@@
 
-A student is learning about System \sysname T, and wants to write a
-program computing $(2+2)\times 3$. He (or she) knows that sometimes those
-$\cst{rec}$ arguments are quite hard to get correct, but System
-\sysname T is a typed language; he wants to take advantage of this to
-build his program incrementally using the teacher's provided
-incremental verifier. As a first attempt, he wants to verify $P_1$:
+A student (the first author) is learning about System \sysname T, and
+wants to write a program computing $(2+2)\times 3$. He knows that
+sometimes those $\cst{rec}$ arguments are quite hard to get correct,
+but System \sysname T is a typed language; he wants to take advantage
+of this to build his program incrementally using the teacher's (the
+first author too) provided incremental verifier. As a first attempt,
+he wants to verify $P_1$:
 $$
@@
 $$
-For this, he sends the command $\infer{P_1}$ to the system. The
-$\infer{}$ operator takes a well-formed program and tries to construct
+For this, he sends the command $\finfer{P_1}$ to the system. The
+$\finfer{}$ operator takes a well-formed program and tries to construct
 a valid typing derivation for its argument in the current, empty
@@
 \begin{align*}
-  &\infer{(
+  &\finfer{(
     \letb{add}{\tlam x {\cst{nat}} \tlam y {\cst{nat}} \recb {\var x}
@@
 the system succeeds with a new derivation made out of these, without
-having to regenerate them because it knows that $\infer{\get\md} =
+having to regenerate them because it knows that $\finfer{\get\md} =
 \md$. Let us call $\md_5$ the subderivation generated for $\tlam x
@@
 \begin{align*}
-  \infer{}(
+  \finfer{}(
   &\letb{add}{\get{\md_5}}
@@
   &\quad\quad\app{\var{mul}}\app
-  {(\get{\gsubst{\md_7}{\msubst H {\infer{\var{add}}}}})}
+  {(\get{\gsubst{\md_7}{\msubst H {\finfer{\var{add}}}}})}
   {(\s{\get{\md_2}})}
@@
 $\var{add}$ is of type $\cst{nat}\to\cst{nat}$; he provides
-$\infer{\var{add}}$ meaning: at this point we will have generated (or
+$\finfer{\var{add}}$ meaning: at this point we will have generated (or
 reused in this case) the derivation of $\var{add}$, so just use it.
@@
 \begin{align*}
-\infer{}(
+\finfer{}(
 \letb{add}{\get{\md_5}}
@@
 objects. To each object constant we assign an annotation $L$ setting
-the slicing behaviour of the constant during type-checking: either
+the slicing behaviour of the constant during type checking: either
 sliceable $\annsliceable$ (each application of this constant will be
@@
 \end{document}
+
+%  LocalWords:  subtype

### 2012-05-28 doc/papers/LFMTP2012/paper.tex
paper

@@
 of programs --- they ``do not go wrong'', and programming practices
-\eg\ modularity, early bug catching. Verifying well-typing of
+--- modularity, early bug catching. Verifying well-typing of
 real-world programs is thus a critical matter, but it is a complex
@@
 description. This is of course true when programs are provided with
-little type annotations (type inference), but is also true in many
-cases for what is usually called ``type checking'', or
-``church-style'' calculi. Take for example dependent type systems,
-relying on a certain notion of definitional equality on types, or type
-systems with subtype polymorphism: both rely on a rule of the form
+little type annotations (``type inference''), in which case the
+missing information has to be algorithmically synthesized, but is also
+true in many cases for what is commonly refered as ``type checking'',
+or explicit ``church-style'' calculi. Take for example dependent type
+systems, relying on a certain notion of definitional equality on
+types, or type systems with subtype polymorphism: both rely on a rule
+of the form
 $$
@@
 The most common option is to formally prove the equivalence, \ie\ that
-the (typing) relation is actually a function. It is of course the most
-satisfactory because it allows to ``forget'' about the declarative
-system and take only the result of the algorithm into consideration,
-although these proofs can be difficult to obtain. But for the sake of
-the discussion, let us choose another path: derivations in the
-declarative systems, if they cannot always be algorithmically
-inferred, can be \emph{verified} if provided by an oracle. This way,
-even if the oracle is wrong, we have a way to decide it. Let us then
-imagine a type-checker to be the pair of 1/ an untrusted procedure
-$\finfer M$ for inferring a typing derivations out of programs $M$; it
-doesn't have to be complete, or even sound, and 2/ a trusted kernel
-$\fcheck{\md}$ for verifying these derivations in a given set of
-inference rules $\Sigma$. This scheme has the following advantages
-over the first approach:
+the (typing) relation is actually a function. Although these proofs
+can be difficult to obtain, it is the most satisfactory because it
+allows to ``forget'' about the declarative system and take only the
+result of the algorithm into consideration. But for the sake of the
+discussion, let us choose another path: derivations in the declarative
+systems, if they cannot always be easily inferred, can be
+\emph{verified} if provided by an oracle. Even if the oracle is wrong,
+we can tell cheaply by verifying the witness typing derivation it
+provides. Let us then imagine a type-checker to be the pair of 1/ an
+untrusted procedure $\finfer M$ for inferring a typing derivations out
+of programs $M$; it doesn't have to be complete, or even sound, and 2/
+a trusted and fast kernel $\fcheck{\md}$ for verifying these
+derivations in a given set of inference rules $\Sigma$. This is a
+certifying scheme: if $\fcheck{\finfer M} = \cst{true}$ then there
+really \emph{is} a derivation $\md$ for $M$. Of course the converse is
+not necessarily true: function $\finfer{}\!\!$ could fail, or have
+bugs (return an incorrect derivation). Explicitly generating the
+derivation in the declarative system has the following advantages over
+the first approach:
 \begin{itemize}
@@
   than the corresponding equivalence proof
-\item it is certifying: if $\fcheck{\finfer M} = \cst{true}$ then
-  there really \emph{is} a derivation $\md$ for $M$. Of course the
-  converse is not necessarily true: function $\finfer{}\!\!$ could
-  fail, or have bugs (return an incorrect derivation).
 \item it inherits benefits from certifying approaches like
-  proof-carrying code \cite{necula1997proof}: %TODO third-party
-                                              %verification
+  proof-carrying code \cite{necula1997proof}: certificates can be
+  transmitted with the program, saving regeneration time and avoiding
+  to disclose the generation program.
 \item for optimization purpose, compilers often need to propagate type
@@
   structure being exploited.
-\item \ldots\ but there is more:
 \end{itemize}
+\ldots\ but there is more:
 
@@
 fixes, etc. If this interaction is possible for short programs and
-fast checkers, it becomes increasingly difficult %TODO
+fast checkers, it becomes increasingly hindered by the latency of
+complete rechecks: even if the modifications made to the program
+between two interactions are small, the whole program will be
+rechecked. This is especially true for languages embedding formal
+verification aspects, like proof assistants: there, tight interaction
+with a type or proof checker is unavoidable due to the difficulty of
+writing correct proofs without guidance from the system, and the time
+taken by the tool to infer trivial parts of the proofs (proof
+search). The constant modification of the source makes it necessary to
+observe \emph{incrementally} the effect of a small change on the whole
+edifice: we need to take advantage of the results of previous checks
+in order to build the result for a modified version of the program.
 

### 2012-05-28 doc/papers/LFMTP2012/paper.tex
paper

@@
 missing information has to be algorithmically synthesized, but is also
-true in many cases for what is commonly refered as ``type checking'',
+true in many cases for what is commonly referred as ``type checking'',
 or explicit ``church-style'' calculi. Take for example dependent type
@@
 provides. Let us then imagine a type-checker to be the pair of 1/ an
-untrusted procedure $\finfer M$ for inferring a typing derivations out
-of programs $M$; it doesn't have to be complete, or even sound, and 2/
-a trusted and fast kernel $\fcheck{\md}$ for verifying these
-derivations in a given set of inference rules $\Sigma$. This is a
-certifying scheme: if $\fcheck{\finfer M} = \cst{true}$ then there
-really \emph{is} a derivation $\md$ for $M$. Of course the converse is
-not necessarily true: function $\finfer{}\!\!$ could fail, or have
-bugs (return an incorrect derivation). Explicitly generating the
-derivation in the declarative system has the following advantages over
-the first approach:
+untrusted procedure $\finfer M$ to infer a typing derivations out of
+programs $M$; it doesn't have to be complete, or even sound, and 2/ a
+trusted and fast kernel $\fcheck{\md}$ to verify these derivations in
+a given set of inference rules $\Sigma$. This is a \emph{certifying}
+scheme: if $\fcheck{\finfer M} = \cst{true}$ then there really
+\emph{is} a derivation $\md$ for $M$. Of course the converse is not
+necessarily true: function $\finfer{}\!\!$ could fail, or have bugs
+(return an incorrect derivation). Explicitly generating the derivation
+in the declarative system has the following advantages over the first
+approach:
 \begin{itemize}
@@
 
-Compilers are generally monolithic programs, taking source code and
-producing machine code in a series of transformations, one of the
-first being type checking. But as type systems become richer, it
-becomes increasingly difficult for the programmer to write a correct
-program in one try, compile it and run it: writing a program longer
-than a few dozen of lines becomes a tight and non-linear interaction
-between the human and the type-checker involving constant experiments,
-fixes, etc. If this interaction is possible for short programs and
-fast checkers, it becomes increasingly hindered by the latency of
-complete rechecks: even if the modifications made to the program
-between two interactions are small, the whole program will be
-rechecked. This is especially true for languages embedding formal
-verification aspects, like proof assistants: there, tight interaction
-with a type or proof checker is unavoidable due to the difficulty of
-writing correct proofs without guidance from the system, and the time
-taken by the tool to infer trivial parts of the proofs (proof
-search). The constant modification of the source makes it necessary to
-observe \emph{incrementally} the effect of a small change on the whole
-edifice: we need to take advantage of the results of previous checks
-in order to build the result for a modified version of the program.
-
-%% State your contributions (Bulleted list of contributions)
+As type systems become richer, it becomes increasingly difficult for
+the programmer to write a correct program in one try, compile it and
+run it: writing a program longer than a few dozen of lines becomes a
+tight and non-linear interaction between the human and the
+type-checker involving constant experiments, fixes, etc. If this
+interaction is possible for short programs and fast checkers, it
+becomes increasingly hindered by the latency of complete rechecks:
+even if the modifications made to the program between two interactions
+are small, the whole program is usually rechecked. This is especially
+true for languages embedding formal verification aspects, like proof
+assistants: there, tight interaction with a type or proof checker is
+unavoidable due to the difficulty of writing correct proofs without
+guidance from the system, and the time taken by many tools to infer
+easy parts of the proofs (proof search). The constant modification of
+the source makes it necessary to observe \emph{incrementally} the
+effect of a small change on the whole edifice: we need to take
+advantage of the results of previous checks in order to build the
+result for a modified version of the program.
+
+Several methods have been devised for turning a batch procedure into
+an incremental one \cite{pugh1989incremental,
+  acar2003selective}. Unfortunately, they all suffer from their
+genericity on the particular problem of type checking:
+\begin{itemize}
+\item first, they ignore the higher-order nature of program syntax,
+  and the kind of structural quotients we make on syntax trees with
+  binders and environments ($\alpha$-equivalence, weakening\ldots): a
+  ground derivation of $\vdash 2+2 : \cst{nat}$ is still valid when
+  put under a binder $\vdash\lam x 2 + 2 - x:\cst{nat}\to\cst{nat}$
+  and should not be regenerated;
+\item for very large programs, we might want to feed the type checker
+  only with a small \emph{delta} on the previous version and not the
+  whole program, thus saving the time to recognize already checked
+  parts;
+\item finally, there is no way to \emph{a posteriori} validate the
+  result of the incremental process: the user has to trust not only
+  the type-checking algorithm but also its transformation into an
+  incremental one.
+\end{itemize}
+It just so happens that, if we consider as above a type-checker as a
+derivation-generating procedure, and storing these derivations between
+calls to the procedure, we could recognize already-built
+subderivations for subterms common to a checked and a new version to
+check of the program, and graft these pieces of derivation to form a
+new, large one for the whole modified program, saving the regeneration
+of all equal subterms' derivations in compatible contexts. This way we
+maintain the verifiable witness of well-typing for the program, and
+take advantage of the quotients made on higher-order syntax.
+
+We can thus add to the previous list that explicitly generating typing
+derivations makes possible a form of safe incrementality by sharing
+common subderivations.
+
+In this paper, we present a framework to incrementally generate typing
+certificates represented in the \LF\ logical framework, and implemented
+as an \sysname{ML} library. We specifically propose:
+\begin{itemize}
+\item a simple language for representing programs and proof
+  deltas, derived from contextual \LF\ \cite{nanevski2008contextual};
+\item an untyped, unsafe calculus to write derivation-generating type
+  checkers, based on a notion of function inverse %TODO le nom?
+  to refer an already-computed value;
+\item a generic algorithm to evaluate these type-checkers on program
+  deltas and verify on-the-fly the generated certificates
+  incrementally against a \emph{repository} of already constructed and
+  checked derivations.
+\end{itemize}
 
@@
 
-A student (the first author) is learning about System \sysname T, and
-wants to write a program computing $(2+2)\times 3$. He knows that
-sometimes those $\cst{rec}$ arguments are quite hard to get correct,
-but System \sysname T is a typed language; he wants to take advantage
-of this to build his program incrementally using the teacher's (the
-first author too) provided incremental verifier. As a first attempt,
-he wants to verify $P_1$:
+A student is learning about System \sysname T, and wants to write a
+program computing $(2+2)\times 3$. He knows that sometimes those
+$\cst{rec}$ arguments are quite hard to get correct, but System
+\sysname T is a typed language; he wants to take advantage of this to
+build his program incrementally using the teacher's provided
+incremental verifier. As a first attempt, he wants to verify $P_1$:
 $$
@@
 call these $\md_2$, $\md_3$ and $\md_4$. Provided there is an operator
-$\get\md$ from the derivation to the program it types, what needs
+$\fget\md$ from the derivation to the program it types, what needs
 really to be sent to the system is:
@@
     \letb{add}{\tlam x {\cst{nat}} \tlam y {\cst{nat}} \recb {\var x}
-      {\var y} {\get{\md_4}}}
-    \app{\var{add}}\app{\get{\md_2}}{\get{\md_2}}
+      {\var y} {\fget{\md_4}}}
+    \app{\var{add}}\app{\fget{\md_2}}{\fget{\md_2}}
   )}
@@
 the system succeeds with a new derivation made out of these, without
-having to regenerate them because it knows that $\finfer{\get\md} =
+having to regenerate them because it knows that $\finfer{\fget\md} =
 \md$. Let us call $\md_5$ the subderivation generated for $\tlam x
 {\cst{nat}} \tlam y {\cst{nat}} \recb {\var x} {\var y}
-{\get{\md_4}}$, $\md_6$ for the arguments
-$\app{\get{\md_2}}{\get{\md_2}}$ and $\md_7$ for the call
-$\app{\var{add}} {\get{\md_6}}$. Even if it occurs originally in the
+{\fget{\md_4}}$, $\md_6$ for the arguments
+$\app{\fget{\md_2}}{\fget{\md_2}}$ and $\md_7$ for the call
+$\app{\var{add}} {\fget{\md_6}}$. Even if it occurs originally in the
 scope of $\var{add}$, $\md_6$ doesn't depend on the hypothesis $\var
@@
   \finfer{}(
-  &\letb{add}{\get{\md_5}}
+  &\letb{add}{\fget{\md_5}}
   \\
@@
   &\quad\quad\app{\var{mul}}\app
-  {(\get{\gsubst{\md_7}{\msubst H {\finfer{\var{add}}}}})}
-  {(\s{\get{\md_2}})}
+  {(\fget{\gsubst{\md_7}{\msubst H {\finfer{\var{add}}}}})}
+  {(\s{\fget{\md_2}})}
   )
@@
 \finfer{}(
-\letb{add}{\get{\md_5}}
-\get{({\gsubst{\md_9} {\msubst {H} {(\get{\var{add}})}; \msubst {H'}
-      {\gsubst{\md_8}{\msubst H {(\get{\var{add}})}}}}})}
+\letb{add}{\fget{\md_5}}
+\fget{({\gsubst{\md_9} {\msubst {H} {(\fget{\var{add}})}; \msubst {H'}
+      {\gsubst{\md_8}{\msubst H {(\fget{\var{add}})}}}}})}
 )
@@
 get the appropriate program: call $\md_{10}$ the whole derivation just
-returned, $\get{\md_{10}}$ returns:
+returned, $\fget{\md_{10}}$ returns:
 \begin{align*}
@@
 
-Although this small example may seem anecdotal, the interest of a
-mechanism for reusing results of derivation generation becomes clear
-on very large programs: then the cost of rechecking well-typing of a
-small modification amounts to recheking the path in the term from the
-root to the changed part, and the changed part itself.
-
-A type-checker is usually viewed as an algorithm returning a binary
-answer (ok/error), specified by a set of inference rules. Seeing
-type-checking as issueing a completely explicit typing derivation
-enables to think of the relationship between the input and the
-output. Feeding part of the output $\md_i$ back to the input $P_i$ is
-a well-known technique for handling changes in the input: combined
-with a way to automatically recognize already-computed input, it is
-\emph{memoization}, a general-purpose technique for incremental
-computation that has been extensively studied and combined with other
-techniques \cite{pugh1989incremental, acar2003selective}. However, it
-is best understood in the case of purely first-order data, and to our
-best knowledge remains unexplored for higher-order structures
-containing binders, like programs and proofs. In this case,
+We choose here to represent deltas as usual terms with variables $\var
+x, \var y$ and special, \emph{meta}-variables referring to
+already-computed results $\md_i$. Commands sent to the system consists
+of terms, possibly containing special typed operators having a
+computational content: $\finfer{\cdot}$ takes a term $M$ and produces
+the (dependent) pair of a type $A$ and a derivation of the judgment
+$\vdash M:A$ in the current context; $\fget{\cdot}$ does the exact
+converse: it takes a derivation $\vdash M:A$ and projects it back as
+the term $M$. The successful result of a command execution is a
+derivation where each subderivation has been given a unique
+metavariable name for later reuse (a \emph{repository}). As
+subderivations can have free variables, we close them by a local
+environment that must be instantiated by a \emph{substitution}.
+
+Although this small example may seem anecdotal, the interest of such a
+mechanism becomes clear on very large programs: then the cost of
+rechecking well-typing of a small modification amounts to rechecking
+the path in the term from the root to the changed part, and the
+changed part itself.
+
+%TODO figure?
+
+Seeing type-checking as issuing a completely explicit typing
+derivation enables thus to think of the relationship between the input
+and the output. Feeding subterms of the output $\md_i$ back into
+subterms of the input $P_i$ is a well-known technique for handling
+changes in the input: combined with a way to automatically recognize
+already-computed input, it is \emph{memoization}, a general-purpose
+technique for incremental computation that has been extensively
+studied and combined with other techniques \cite{pugh1989incremental,
+  acar2003selective}. However, it is best understood in the case of
+purely first-order data, and to our best knowledge remains unexplored
+for higher-order structures containing binders, like programs and
+proofs.
+
+A memoized type-checker stores in a table bindings from its input (the
+program) to its output (the derivation), and returns automatically the
+stored output if the current input matches a binding in the table. It
+thus performs two operations at once: output reuse and recognition of
+already-seen input. Since       %TODO
 
@@
 \begin{itemize}
-\item from a modified program, generate the \emph{diff} against the
+\item from a modified program, generate the \emph{delta} against the
   previous version, that is, observing and linking to unchanged
   derivations.
-\item given a \emph{diff}, construct a new derivation and verify it.
+\item given a \emph{delta}, construct a new derivation and verify it.
 \end{itemize}
@@
 \item conciseness of the certificates, since they are usually stored
-  or transfered remotely; in particular,
+  or transferred remotely; in particular,
 \item a well-understood, short and independent verification algorithm;
@@
 
-The Edimburgh Logical Framework \cite{harper1993framework}, or \LF\
+The Edinburgh Logical Framework \cite{harper1993framework}, or \LF\
 for short, is a well-accepted solution to this question. It is a typed
@@
 meta-theorems (but not their proofs) in a given inference system
-supporting hypothetical judgements \cite{pfenning2001logical}. For
+supporting hypothetical judgments \cite{pfenning2001logical}. For
 these, it provides built-in support for encoding languages with
@@
 
-We overload \emph{substitution} as a partial operation on objects, given
-a parallel substitution $\sigma$. Since the syntax of object is not
-stable by (textual) substitution, it is defined \emph{hereditarily}: it
-is mutually recursive with a \emph{cut} function $\gcut M S$ which can
-in turn trigger a chain of substitutions, leading to a canonical objects
-or an error.
+We overload \emph{substitution} as a partial operation on objects,
+given a parallel substitution $\sigma$. Since the syntax of object is
+not stable by (textual) substitution, it is defined as
+\emph{hereditary}: it is mutually recursive with a \emph{cut} function
+$\gcut M S$ which can in turn trigger a chain of substitutions,
+leading to a canonical objects or an error.
 \begin{align*}
@@
 substitution found in \eg\ \cite{hl07mechanizing} as it presupposes
-$\eta$-long normal forms: once a cut starts, there must be exactly as
-many formal and actual arguments. A non-empty spine cut against
-another non-empty spine $\gcut {\app H S} {S'}$ blocks the
+$\eta$-long normal forms: once a \emph{cut} starts, there must be
+exactly as many formal and actual arguments. A non-empty spine cut
+against another non-empty spine $\gcut {\app H S} {S'}$ blocks the
 substitution instead of reducing to $\app H {(\gconcat S S')}$ (an
@@
 The algorithm is presented in Fig. \ref{fig:obj-typing} as a
-syntax-directed inference system. All judgements are of the form $\gjo
+syntax-directed inference system. All judgments are of the form $\gjo
 U V W$ where $U$ and $V$ are inputs, $V$ is the induction hypothesis
 and $W$ the output. They are parameterized by a constant and implicit
-signature $\Sigma$, and all premisses are ordered in the algorithmical
+signature $\Sigma$, and all premises are ordered in the algorithmic
 order of computation.

### 2012-05-29 doc/papers/LFMTP2012/paper.tex
paper

@@
 real-world programs is thus a critical matter, but it is a complex
-operation, for the compiler designer and for the human operator.
+operation, for the compiler designer and for the human operator:
 
@@
 $$
-\infer{\Gamma\vdash M : A \\ A\leq B}{\Gamma\vdash M : B}\ \text{,}
+\infer{\Gamma\vdash M : A \\ A\leq B}{\Gamma\vdash M : B}
 $$
-for a given notion of $\leq$, which is not syntax-directed. The
+(for a given notion of $\leq$) which is not syntax-directed. The
 implementer then has to convert the declarative system to an
-algorithm, and convince himself that both are equivalent. This task,
-often neglected, is nonetheless highly non-trivial and non-modular.
+algorithm, and convince himself that both are equivalent. This task is
+highly non-trivial and non-modular: for instance, how to infer the
+type of a System \sysname{T} recursor
+$$
+\infer{
+  \Gamma\vdash M : \cst{nat}
+  \and
+  \Gamma\vdash N : A
+  \and
+  \Gamma, \var x:\cst{nat}, \var y:A\vdash P : A
+}{
+  \Gamma\vdash \recb M N x y P
+}
+$$
+in presence of subtyping? The problem is that we have to guess a type
+$A$ more general than that of $N$ and $P$, while giving that type to
+variable $\var y$. A well-known algorithm if the subtyping relation
+forms a complete lattice is to iterate the typing of $P$ with the join
+of the previously computed and the new type until a fixpoint is
+reached.
 
@@
 can be difficult to obtain, it is the most satisfactory because it
-allows to ``forget'' about the declarative system and take only the
-result of the algorithm into consideration. But for the sake of the
-discussion, let us choose another path: derivations in the declarative
-systems, if they cannot always be easily inferred, can be
-\emph{verified} if provided by an oracle. Even if the oracle is wrong,
-we can tell cheaply by verifying the witness typing derivation it
-provides. Let us then imagine a type-checker to be the pair of 1/ an
-untrusted procedure $\finfer M$ to infer a typing derivations out of
-programs $M$; it doesn't have to be complete, or even sound, and 2/ a
-trusted and fast kernel $\fcheck{\md}$ to verify these derivations in
-a given set of inference rules $\Sigma$. This is a \emph{certifying}
-scheme: if $\fcheck{\finfer M} = \cst{true}$ then there really
-\emph{is} a derivation $\md$ for $M$. Of course the converse is not
-necessarily true: function $\finfer{}\!\!$ could fail, or have bugs
-(return an incorrect derivation). Explicitly generating the derivation
-in the declarative system has the following advantages over the first
-approach:
+allows to take only the result of the algorithm into
+consideration. But let us choose another path for a minute:
+derivations in the declarative systems, if they cannot always be
+easily inferred, can be \emph{verified} if provided by an oracle. If
+the oracle is wrong, we can tell cheaply by verifying the witness
+typing derivation it provides. Let us then imagine a type-checker to
+be the pair of 1/ an untrusted procedure $\finfer M$ to infer a typing
+derivation out of program $M$; it doesn't have to be complete, or even
+sound, and 2/ a trusted and fast kernel $\fcheck{\md}$ to verify these
+derivations in a given set of inference rules $\Sigma$. This is a
+\emph{certifying} scheme: %TODO cite Leroy
+if $\fcheck{\finfer M} = \cst{true}$ then there really \emph{is} a
+derivation $\md$ for $M$, but the converse is not necessarily true:
+function $\finfer{}\!\!$ could fail, or have bugs (return an incorrect
+derivation). Explicitly generating the derivation in the declarative
+system has the following advantages over the first approach:
 \begin{itemize}
@@
   than the corresponding equivalence proof
+\item compilers often need to propagate type information further in
+  the chain of transformation; the typing derivation produced by
+  $\finfer{}$ being explicit, it could constitute the data structure
+  being exploited.
 \item it inherits benefits from certifying approaches like
@@
   to disclose the generation program.
-\item for optimization purpose, compilers often need to propagate type
-  information in the chain of transformation; the typing derivation
-  produced by $\finfer{}$ being explicit, it could constitute the data
-  structure being exploited.
 \end{itemize}
@@
 the programmer to write a correct program in one try, compile it and
-run it: writing a program longer than a few dozen of lines becomes a
+run it: writing a program longer than a few dozens of lines becomes a
 tight and non-linear interaction between the human and the
@@
   binders and environments ($\alpha$-equivalence, weakening\ldots): a
-  ground derivation of $\vdash 2+2 : \cst{nat}$ is still valid when
-  put under a binder $\vdash\lam x 2 + 2 - x:\cst{nat}\to\cst{nat}$
-  and should not be regenerated;
+  hypothetical derivation of\ \ $\vdash x+2 : \cst{nat}$ is still valid
+  when put under a binder\ \ $\vdash\lam y (x + 2) -
+  y:\cst{nat}\to\cst{nat}$ and should not be regenerated;
 \item for very large programs, we might want to feed the type checker
@@
   the type-checking algorithm but also its transformation into an
-  incremental one.
+  incremental one, which increases the size of the trusted base.
 \end{itemize}
-It just so happens that, if we consider as above a type-checker as a
-derivation-generating procedure, and storing these derivations between
-calls to the procedure, we could recognize already-built
-subderivations for subterms common to a checked and a new version to
-check of the program, and graft these pieces of derivation to form a
-new, large one for the whole modified program, saving the regeneration
-of all equal subterms' derivations in compatible contexts. This way we
-maintain the verifiable witness of well-typing for the program, and
-take advantage of the quotients made on higher-order syntax.
+
+Now, if we consider as above a type-checker as a derivation-generating
+procedure, and storing these derivations between calls to the
+procedure, we could recognize already-built subderivations for
+subterms common to a checked and a new version of the program, and
+graft these pieces of derivation to form a new, large one for the
+whole modified program, saving the regeneration of all equal subterms'
+derivations in compatible contexts. This way we constantly maintain
+the verifiable witness of well-typing for the program, and take
+advantage of the quotients made on higher-order syntax.
 
@@
   deltas, derived from contextual \LF\ \cite{nanevski2008contextual};
-\item an untyped, unsafe calculus to write derivation-generating type
-  checkers, based on a notion of function inverse %TODO le nom?
-  to refer an already-computed value;
+\item a computational language to write untrusted
+  derivation-generating type checkers, based on a notion of function
+  inverse %TODO le nom?
+  to refer to an already-computed value;
 \item a generic algorithm to evaluate these type-checkers on program
@@
 
-A student is learning about System \sysname T, and wants to write a
+A student is learning about System \sysname T --- a simply typed
+$\lambda$-calculus with natural numbers introduction $\z$ and $\s{M}$,
+and elimination ($\recb M N x y P$).  He (or she) wants to write a
 program computing $(2+2)\times 3$. He knows that sometimes those
@@
 $$
-\recb{\s\z}{\s\z}{\lam x\lam y \s {\var x}}\ \text{.}
+\recb{\s\z}{\s\z} x y {\s {\var x}}\ \text{.}
 $$
@@
   &\letb{add}{\tlam x {\cst{nat}} \tlam y {\cst{nat}} \recb {\var x}
-    {\var y} {\lam x \lam y \s{\var x}}}
+    {\var y} u v {\s{\var u}}}
   \\
@@
 But he (or his text editor) realizes that some parts of this program
-have not changed: both $\s\z$, and the function $\lam x\lam y\s{\var
-  x}$, and he could reuse the pieces of derivation just built. Let us
-call these $\md_2$, $\md_3$ and $\md_4$. Provided there is an operator
-$\fget\md$ from the derivation to the program it types, what needs
-really to be sent to the system is:
+have not changed: both $\s\z$, and the function body $\s{\var x}$, and
+he could reuse the pieces of derivation just built. Let us call these
+respectively $\md_2$, $\md_3$ and $\md_4$. $\md_4$ is hypothetical: it
+depends on the hypothesis $H$ that\ \ $\vdash \var x:\cst{nat}$ .
+Provided there is an operator $\fget\md$ from the derivation to the
+program it types, what needs really to be sent to the system is:
 \begin{align*}
-  &\finfer{(
+  &\finfer{}(
     \letb{add}{\tlam x {\cst{nat}} \tlam y {\cst{nat}} \recb {\var x}
-      {\var y} {\fget{\md_4}}}
+      {\var y} u v {\fget{\gsubst{\md_4}{\msubst{H}{\finfer{\var u}}}}}}
+    \\&\qquad
     \app{\var{add}}\app{\fget{\md_2}}{\fget{\md_2}}
-  )}
+  )
 \ \text{,}
 \end{align*}
-that is, only the subterms of the program that changed. Derivations
-$\md_2$ and $\md_4$ used to live in the empty context, we are then
-free to use it in the new context where
-$\var{add}:\cst{nat}\to\cst{nat}$, because System \sysname T enjoys
-the \emph{weakening} property. Taking advantage of $\md_2$ and $\md_4$
-the system succeeds with a new derivation made out of these, without
-having to regenerate them because it knows that $\finfer{\fget\md} =
-\md$. Let us call $\md_5$ the subderivation generated for $\tlam x
-{\cst{nat}} \tlam y {\cst{nat}} \recb {\var x} {\var y}
-{\fget{\md_4}}$, $\md_6$ for the arguments
-$\app{\fget{\md_2}}{\fget{\md_2}}$ and $\md_7$ for the call
+that is, only the changed subterms of the program. Derivations $\md_2$
+used to live in the empty context, we are then free to use it in the
+new context where $\var{add}:\cst{nat}\to\cst{nat}$, because System
+\sysname T enjoys the \emph{weakening} property; $\md_4$ depending on
+a derivation\ \ $\vdash\var x:\cst{nat}$, we instantiate this hole
+with the derivation $\finfer{\var u}$ inferred recursively for then
+new recursive argument $\var u$. Taking advantage of $\md_2$ and
+$\md_4$ the system succeeds with a new derivation made out of these,
+without having to regenerate them because it knows that
+$\finfer{\fget\md} = \md$.
+
+Let us call $\md_5$ the subderivation generated for $\tlam x
+{\cst{nat}} \tlam y {\cst{nat}} \recb {\var x} {\var y} u v
+{\fget{\gsubst{\md_4}{\msubst H {\finfer{\var u}}}}}$, $\md_6$ for the
+arguments $\app{\fget{\md_2}}{\fget{\md_2}}$ and $\md_7$ for the call
 $\app{\var{add}} {\fget{\md_6}}$. Even if it occurs originally in the
@@
   &\quad\letb{mul} { \tlam x {\cst{nat}} \tlam y
-    {\cst{nat}} \recb \z {\var y} {\lam z \lam t
+    {\cst{nat}} \recb \z {\var y} z t {
       \app{\var{add}}\app{\var x} {\var z}} }
@@
 $\finfer{\var{add}}$ meaning: at this point we will have generated (or
-reused in this case) the derivation of $\var{add}$, so just use it.
+reused in this case) the derivation of $\var{add}$, so we just have to
+use it here.
 
@@
   &\letb{add}{\tlam x {\cst{nat}} \tlam y {\cst{nat}} \recb {\var x}
-    {\var y} {\lam x \lam y \s{\var x}}}
+    {\var y} x y {\s{\var x}}}
   \\&\quad
   {\tlam x {\cst{nat}} \tlam y
-    {\cst{nat}} \recb \z {\var y} {\lam z \lam t
+    {\cst{nat}} \recb \z {\var y} z t {
       \app{\var{add}}\app{\var x} {\var z}}}
@@
 thus performs two operations at once: output reuse and recognition of
-already-seen input. Since       %TODO
-
-To simplify the problem,
-we separate the task in:
-\begin{itemize}
-\item from a modified program, generate the \emph{delta} against the
-  previous version, that is, observing and linking to unchanged
-  derivations.
-\item given a \emph{delta}, construct a new derivation and verify it.
-\end{itemize}
-The latter is the subject of this paper.
+already-seen input. Recognition --- that is from a modified program,
+generate a delta from previous versions --- is in the higher-order
+case a complex task that we hope to address in the future; we tackle
+here the verification of output reuse: given a delta, construct a new
+derivation and verify it.
 
@@
 
-\subsection{The \LF\ representation language}
+\subsection{The representation language}
 
@@
 \begin{itemize}
-\item conciseness of the certificates, since they are usually stored
-  or transferred remotely; in particular,
-\item a well-understood, short and independent verification algorithm;
-  this is known as the \emph{de Bruijn principle}.
+\item conciseness of the certificates, since they are stored or even
+  transferred remotely;
+\item a well-understood, short and independent verification algorithm.
+  This is known as the \emph{de Bruijn principle};
 \item universality with respect to the particular problem: having a
@@
 these, it provides built-in support for encoding languages with
-binders or hypothesis (a feature called $\lambda$-tree syntax or HOAS
+binders and hypothesis (a feature called $\lambda$-tree syntax or HOAS
 \cite{pfenning1988higher}). The user gives a \emph{signature} defining
@@
 gets an algorithm to verify terms and derivations expressed in the
-\LF\ syntax.
-
-This framework has been used to issue certificates for a number of
-practical applications like proof-carrying code
+\LF\ syntax. This framework has been used to issue certificates for a
+number of practical applications like proof-carrying code
 \cite{necula1997proof}, authentication framework \cite{appel1999proof}
@@
 
-In all these applications, the user feeds the generation program with
-a query (the program to compile to machine-code, the access query, the
-statement of a theorem \ldots) and gets in return a result (machine
-code, yes/no\ldots) together with a certificate that the result
-verifies the wanted property (semantics preservation, justification
-for granting access, proof of the theorem\ldots). A potential client
-can then verify independently that this result is adequate given the
-query: it suffices to batch-verify the certificate.
-
-\subsection{Proof-carrying static analysis}
-
-\subsection{Storing computation results}
-\subsection{Naming intermediate results}
-
-\section{Examples}
+To write deltas by sharing subterms, we need to be able to address any
+subterm and instantiate its free variables. For this, we define the
+the \emph{slicing} of a well-typed \LF\ term: we extend \LF\ with
+metavariables $\meta X$, $\meta Y$\ldots\ as in
+\cite{nanevski2008contextual}. Metavariables stand for well-typed
+terms defined in a repository, and are instantiated wrt. their free
+variables. A repository also contains a distinguished metavariable
+corresponding to the head of the term stored in it. For instance, the
+repository
+$$
+\left(\begin{array}{llll}
+    \meta X &\mapsto
+    \envnil&\vdash
+    \app{\cst{lam}} \lam x \smeta Y {\msubst u {\smeta T {\msubst w x}}} &:
+    \cst{tm}
+    \\
+    \meta Y &\mapsto
+    \env u {\cst{tm}} &\vdash
+    \app{\cst{lam}} \lam y \smeta Z {\msubstcons {\msubst u
+      {\var u}} v {\var y} } &:
+    \cst{tm}
+    \\
+    \meta Z &\mapsto
+    \envcons {\env u {\cst{tm}}} v {\cst{tm}} &\vdash
+    \app{\app{\cst{app}}{\var v}}{\var u} &:
+    \cst{tm}
+    \\
+    \meta T &\mapsto
+    \env x {\cst{tm}} &\vdash
+    \app{\cst{app}}\app{\var x}{\cst{0}} &:
+    \cst{tm}
+
+\end{array}\right), \smeta X {}
+$$
+is one of the possible way of slicing of the term
+$$
+\fbox{$\app {\cst{lam}} \lam x
+\fbox{$\app {\cst{lam}} \lam y
+\fbox{\app {\cst{app}} \app {\var y}
+}$}
+{(\fbox{\app
+  {\cst{app}}\app{\var x}{\cst 0}})
+}
+$}$$
+where all boxed subterms are referable slices.
+
+\subsection{Computational calculus}
+\subsection{Slicing results of computation}
 
@@
 
-In this variant, only $\eta$-long $\beta$-normal forms have an
+In this variant, only $\eta$-long and $\beta$-normal values have an
 existence: canonicity for $\beta$-reduction is enforced in the syntax

### 2012-05-29 doc/papers/LFMTP2012/paper.tex
paper

@@
 
-\section{Background}
+\section{Use case}
 %% It’s an unsolved problem (1 page)
 
-\paragraph{Motivational example}
+\paragraph{A backend for incremental type checking}
 
@@
 
+\paragraph{A language to write incremental type-checkers}
+
+A few days earlier, the teacher is preparing the tool for the students
+to train composing System \sysname{T} programs. He (or she) provides
+the abstract syntax, the typing rules and the typing algorithm in
+order to build the incremental type checker. How should he write the
+algorithm? Usually, it takes a program and an environment and returns
+nothing but we know that we want to return a derivation, and take
+advantage of a higher-order representation of the syntax, so it
+shouldn't be necessary to thread any explicit environment. %TODO
+$$
+
+$$
 \section{Here is my idea}

### 2012-05-30 doc/papers/LFMTP2012/paper.tex
paper

--- 2012-05-29 doc/papers/LFMTP2012/paper.tex
@@
   inverse %TODO le nom?
-  to refer to an already-computed value;
+  to refer to already-computed values;
 \item a generic algorithm to evaluate these type-checkers on program
@@
 
-\section{Background}
+\section{Use case}
 %% It’s an unsolved problem (1 page)
 
-\paragraph{Motivational example}
+\subsection{A backend for incremental type checking}
+\label{sec:use-incremental}
 
@@
 
+\subsection{A language to write incremental type-checkers}
+
+A few days earlier, the teacher is preparing the tool for the students
+to train composing System \sysname{T} programs. He (or she) provides
+the abstract syntax, the typing rules and the typing untrusted
+algorithm in order to build the incremental type checker. How should
+he write the algorithm? We know that we want to return a derivation,
+and take advantage of a higher-order representation of the syntax, so
+it shouldn't be necessary to thread any explicit environment. The
+definition should then start with:
+
+\begin{mathleft}
+  \finfer{}\ :\ \prd M {\cst{tm}} \sig A {\cst{tp}} (\vdash M : A) =
+\end{mathleft}
+
+\noindent
+It takes a term $M$ and producing the pair of a type $A$ and a
+derivation that $M$ has type $A$. Its code will be untyped, but
+the teacher provides this type so that the checker can verify that
+each call is fed and returns values of the right type: it is its
+\emph{contract} so to say. He first decomposes term $M$ and treats
+the easy cases of $\cst{o}$, $\cst{s}$ and $\cst{app}$:
+
+\begin{mathleft}
+  \lamd M \match{M} \\
+  \quad\case{\z} \pair {\cst{nat}}{\infer{ }{\vdash \cst o : \cst{nat}}} \\
+  \quad\case{\s M}
+  \letd {\pair {\cst{nat}} \md} {\finfer{M}}
+  \pair {\cst{nat}}{\infer\md{\vdash\app{\cst{s}} M : \cst{nat}}} \\
+  \quad\case{\app{M}N}
+  \letd {\pair {A_1\to B} {\md_1}} {\finfer{M}} \\
+  \quad\quad\letd {\pair {A_2} {\md_2}} {\finfer{N}} \\
+  \quad\quad\letd {\cst{true}} {\feq {A_1} {A_2}}
+  \pair {B} {\infer{\md_1 \and \md_2}{\vdash \app M N : B}}
+\end{mathleft}
+
+\noindent
+We write $\pair{\cdot}{\cdot}$ the constructor of $\Sigma$-types; the
+$\syntax{let}$ construct is just syntactic sugar for a one-branch
+$\syntax{case}$; patterns don't have to be exhaustive since we are in
+an untrusted setting; for conciseness, we use case failure for
+signaling typing error; we use an auxiliary function $\!\!\feq{}{}\!\!
+: \cst{tp}\to\cst{tp}\to\cst{bool}$ for syntactic comparison of types,
+defined the same way.
+
+Then comes the abstraction case $\tlam x A M$. The difficulty is that
+the recursive call $\finfer{M}$ must be done in an enlarged
+environment where $\vdash x : A$, otherwise the check that the
+returned open derivation $\md$ is correct would fail ($x$ is
+unknown). We thus introduce a new construct $\matchin M \Gamma \ldots$
+and its $\syntax{let}$ counterpart ($\match M \ldots$ being syntactic
+sugar for $\matchin M \envnil \ldots$). Even then, in the recursive
+call, arriving to variable $\var x$ would fail since we would have no
+way of telling its type anymore. Thus we substitute all occurrences of
+$\var x$ with the \emph{output} of function $\finfer{}$ on it (since
+it is known at this point), \ie\ a pair of $A$ and $\md_1$. For this
+substitution to be well-typed, we now need to coerce this output to a
+valid input (a term). We introduce an operator which is in some way
+the \emph{inverse} of $\finfer{}$: operator $\fget{}$ maps a pair of a
+type and a derivation to the term it types.
+
+\begin{mathleft}
+  \quad\case{\tlam x A M} \\
+  \quad\quad
+  \letdin{\pair B \md}{\env {$\md_{\var x}$} {(\vdash {\var x} : A)}}
+  {\finfer {\gsubst M {\msubst x {\fget{\pair A {\md_{\var x}}}}}}}
+\end{mathleft}
+
+At this point, we get back $B$ and $\md$ from the recursive call. We
+know that $B$ can't depend on $\md_x$ (because a type is always
+closed), but $\md$ can: it is not \emph{per se} a hypothetical
+derivation but an \emph{open} derivation where a special judgement
+named $\md_x$ can occur. To return the complete derivation of
+$\vdash\tlam A x M : A \to B$, we must then close, or ``hypothesize''
+$\md$ by $\md_x$, and then apply rule \rulename{Lam}:
+
+\begin{mathleft}
+  \quad\quad
+  \pair {A\to B}
+  {\infer{
+\mbox{$      \begin{array}{c}
+        {[\md_{\var x}]} \\
+        {\md}
+      \end{array}
+    $}    }{
+    \vdash\lam x M : A\to B
+  }}
+\end{mathleft}
+
+The case of $\cst{rec}$ is similar: we first compute recursively the
+derivations $\md_M$ and $\md_N$ which should be well-typed in the
+current environment, and then $\md_P$ in the enlarged environment,
+with variables $\var x$ and $\var y$ in $P$ substituted with the
+(future) result of $\finfer{}$.
+
+\begin{mathleft}
+  \quad\case{\recb M N x y P} \\
+  \quad\quad
+  \letd {\pair {\cst{nat}} {\md_M}} {\finfer{M}} \\
+  \quad\quad
+  \letd {\pair {A_1} {\md_N}} {\finfer{N}} \\
+  \quad\quad
+  \syntax{let} {\pair {A_2} {\md_P}} \syntax{in}
+  {(\envcons {\env {$\md_x$} {(\vdash
+        \var x : \cst{nat})}} {$\md_y$} {(\vdash\var y : A_1)})} = \\
+  \quad\quad\quad
+  {\finfer{\gsubst P {\msubstcons {\msubst x
+          {\fget{\pair{\cst{nat}}{\md_x}}}} y
+        {\fget{\pair{A_1}{\md_y}}}}}} \\
+  \quad\quad\letd {\cst{true}} {\feq {A_1} {A_2}} \\
+  \quad\quad
+  \pair {A} {
+    \infer{
+      \md_M \and
+      \md_N \and
+      \mbox{$
+        \begin{array}{c}
+          [\md_x] [\md_y] \\
+          \md_P
+      \end{array}
+        $}
+    }{
+      \vdash \recb M N x y P
+    }
+  }
+\end{mathleft}
+
+The type-checker is now done. There is no case for variables, as we
+won't ever meet this case: we already substituted all variables with
+their result derivation. We only need to make sure that in this case,
+the derivation will be returned:
+$$
+\finfer{\fget{\pair A \md}} = \md
+$$
+Now $\fget{}$ can be derived automatically from the type of
+$\finfer{}$: it is its inverse. It has definition:
+
+\begin{mathleft}
+  \fget{}\ :\ \prd{\{M\}}{\cst{tm}} \prd {A}{\cst{tp}} \prd{\md}{(\vdash
+    M:A)} \cst{tm} =
+  \\\quad \lamd M \lamd A \lamd \md M
+\end{mathleft}
+
+\noindent
+where argument $M$ is implicit.
+
+\paragraph{Discussion}
+
+We exploit here an idea similar to run-time
+\emph{contracts}: %TODO cite
+there is no static guarantee that computations won't return ill-typed
+values, but arguments and results are checked at run-time at the
+boundaries of defined functions. This allows to omit formal
+justifications, like the fact that $B$ can't depend on $\md_x$ in the
+$\lambda$ case, but of course incurs on the performance. This is
+compensated by the slicing mechanism described below in
+\ref{sec:repr}.
+
+Introducing the ``inverse'' of the derivation inference function has
+two purpose: as we saw in \ref{sec:use-incremental}, it allows to
+refer to already-constructed derivations when writing deltas. It also
+permits to abstract from the concrete environment when writing the
+type-checker itself, that one must thread throughout the process in a
+traditional, first-order type-checker. It is a generalization of the
+so-called \emph{context-free} typing presented in
+\cite[chap. 4]{boespflug2011conception} where a special variable
+construct $\{x:A\}$ denoting free variables is added to the syntax of
+terms, and the rules
+$$
+\infer{
+  \vdash \gsubst M {\msubst x {\{\var x:A\}}} : B
+}{
+  \vdash \tlam x A M : A\to B
+}
+\qquad\text{and}\qquad
+\infer{ }{
+  \vdash \{\var x : A\} : A
+}
+$$
+replace the usual ones.
+
+The difficulty with this technique, particularly in a dependent
+setting, is to ensure that term $\{\var x:A\}$ reduces to $\var x$ so
+that they are considered equal, \ie\ we can strip off the annotations,
+while not forgetting them too soon: otherwise we would arrive to the
+unhandled case $\finfer{\var x}$. This requires evaluating the
+type-checker with a carefully designed strategy.
+
 \section{Here is my idea}
@@
 \subsection{The representation language}
+\label{sec:repr}
 
@@
 gets an algorithm to verify terms and derivations expressed in the
-\LF\ syntax. This framework has been used to issue certificates for a
+\LF\ syntax.
+
+For instance, the rule for System \sysname{T}'s recursor can be
+encoded by the following constant:
+
+\begin{mathleft}
+  \cst{is\_rec}\ :\
+  \prd{M}{\cst{tm}}\prd{N}{\cst{tm}}\prd{P}{\cst{tm}\to\cst{tm}\to\cst{tm}}
+  \\\quad
+  \prd A {\cst{tp}}
+  \app {\cst{is}}\app M {\cst{nat}} \to \app{\cst{is}}\app N A \to \\\quad
+  \left(
+  \prd x {\cst{tm}} \prd y {\cst{tm}}
+  \app{\cst{is}}\app x \cst{nat} \to \app{\cst{is}}\app y A \to
+  \app{\cst{is}}\app {(\app P \app x y)} A
+  \right)\to\\\quad
+  \app{\cst{is}}\app{(\app{\cst{rec}}\app M \app N (\lam x\lam y \app
+    P\app x y))} A
+\end{mathleft}
+
+This framework has been used to issue certificates for a
 number of practical applications like proof-carrying code
@@
     \env x {\cst{tm}} &\vdash
-    \app{\cst{app}}\app{\var x}{\cst{0}} &:
+    \app{\cst{app}}\app{\var x}{\cst{o}} &:
     \cst{tm}
@@
 {(\fbox{\app
-  {\cst{app}}\app{\var x}{\cst 0}})
+  {\cst{app}}\app{\var x}{\cst o}})
 }
@@
 
-\subsection{Computational calculus}
-\subsection{Slicing results of computation}
+\subsection{The computational calculus}
 

### 2012-05-31 doc/papers/LFMTP2012/paper.tex
paper
Revert some debug message

@@
 
-A student is learning about System \sysname T --- a simply typed
-$\lambda$-calculus with natural numbers introduction $\z$ and $\s{M}$,
-and elimination ($\recb M N x y P$).  He (or she) wants to write a
-program computing $(2+2)\times 3$. He knows that sometimes those
-$\cst{rec}$ arguments are quite hard to get correct, but System
-\sysname T is a typed language; he wants to take advantage of this to
-build his program incrementally using the teacher's provided
-incremental verifier. As a first attempt, he wants to verify $P_1$:
+A student is learning about System \sysname {T_{<:}} --- a simply
+typed $\lambda$-calculus with natural numbers $\cst{nat}$ coming in
+two sets $\cst{even}$ and $\cst{odd}$, introduction $\z$ and $\s{M}$,
+elimination ($\recb M N x y P$), and subtyping those types. He (or
+she) wants to write a program computing $(2+2)\times 3$. He knows that
+sometimes those $\cst{rec}$ arguments are quite hard to get correct,
+but \sysname{T_{<:}} is a typed language; he wants to take advantage
+of this to build his program incrementally using the teacher's
+provided incremental verifier. As a first attempt, he wants to verify
+$P_1$:
 $$
@@
 But he (or his text editor) realizes that some parts of this program
-have not changed: both $\s\z$, and the function body $\s{\var x}$, and
+have not changed: both $\s\z$, and the function body $\s{\var u}$, and
 he could reuse the pieces of derivation just built. Let us call these
 respectively $\md_2$, $\md_3$ and $\md_4$. $\md_4$ is hypothetical: it
-depends on the hypothesis $H$ that\ \ $\vdash \var x:\cst{nat}$ .
+depends on the hypothesis $H$ that\ \ $\vdash \var u:\cst{nat}$ .
 Provided there is an operator $\fget\md$ from the derivation to the
-program it types, what needs really to be sent to the system is:
+program it types, what really needs to be sent to the system is:
 \begin{align*}
@@
 \end{align*}
-that is, only the changed subterms of the program. Derivations $\md_2$
+that is, only the changed subterms of the program. Derivation $\md_2$
 used to live in the empty context, we are then free to use it in the
@@
 with the derivation $\finfer{\var u}$ inferred recursively for then
-new recursive argument $\var u$. Taking advantage of $\md_2$ and
-$\md_4$ the system succeeds with a new derivation made out of these,
-without having to regenerate them because it knows that
-$\finfer{\fget\md} = \md$.
+new recursive argument $\var u$. $\md_2$ derives $\vdash \s\z :
+\cst{even}$, but $\var{add}$ awaits $\cst{nat}$ arguments. When the
+system will generate this derivation, it will insert a coercion at
+these two points.
+% TODO j'en ss la ds la relecture F<
+Taking advantage of $\md_2$ and $\md_4$ the system
+succeeds with a new derivation made out of these, without having to
+regenerate them because it knows that $\finfer{\fget\md} = \md$.
 
@@
 $$
-replace the usual ones.
-
-The difficulty with this technique, particularly in a dependent
-setting, is to ensure that term $\{\var x:A\}$ reduces to $\var x$ so
-that they are considered equal, \ie\ we can strip off the annotations,
-while not forgetting them too soon: otherwise we would arrive to the
-unhandled case $\finfer{\var x}$. This requires evaluating the
-type-checker with a carefully designed strategy.
+replace the usual ones. The difficulty with this technique,
+particularly in a dependent setting, is to ensure that term $\{\var
+x:A\}$ reduces to $\var x$ so that they are considered equal, \ie\ we
+can strip off the annotations, while not forgetting them too soon:
+otherwise we would arrive to the unhandled case $\finfer{\var
+  x}$. This requires evaluating the type-checker with a carefully
+designed strategy.
+
+For the sake of conciseness, we took a simple, syntax-directed system
+as example; this scales nonetheless to declaratively presented
+features like subtyping or conversion.
 
@@
 \subsection{The computational calculus}
+
+% inverse
 

### 2012-05-31 doc/papers/LFMTP2012/paper.tex
Revert some debug message

@@
 derivations in a given set of inference rules $\Sigma$. This is a
-\emph{certifying} scheme: %TODO cite Leroy
+\emph{certifying} scheme (see \cite{leroy2006compcert} for a
+discussion on approaches to reliable compilers):
 if $\fcheck{\finfer M} = \cst{true}$ then there really \emph{is} a
@@
 two sets $\cst{even}$ and $\cst{odd}$, introduction $\z$ and $\s{M}$,
-elimination ($\recb M N x y P$), and subtyping those types. He (or
+elimination ($\recb M N x y P$), and subtyping on those types. He (or
 she) wants to write a program computing $(2+2)\times 3$. He knows that
@@
 new recursive argument $\var u$. $\md_2$ derives $\vdash \s\z :
-\cst{even}$, but $\var{add}$ awaits $\cst{nat}$ arguments. When the
-system will generate this derivation, it will insert a coercion at
-these two points.
-% TODO j'en ss la ds la relecture F<
-Taking advantage of $\md_2$ and $\md_4$ the system
-succeeds with a new derivation made out of these, without having to
-regenerate them because it knows that $\finfer{\fget\md} = \md$.
+\cst{odd}$, but $\var{add}$ awaits $\cst{nat}$ arguments. When the
+system will generate this derivation, it will insert a coercion
+$\cst{odd} \leq \cst{nat}$ at these two points. Taking advantage of
+$\md_2$ and $\md_4$ the system succeeds with a new derivation made out
+of these, without having to regenerate them because it knows that
+$\finfer{\fget\md} = \md$.
 
@@
 reused in this case) the derivation of $\var{add}$, so we just have to
-use it here.
+use it here. $\md_2$ was deriving $\vdash\s\z:\cst{nat}$ (with a
+coercion since it was used as an argument to $\var{add}$); the system
+thus infer the more general type $\cst{nat}$ for $\s{\fget{\md_2}}$.
 
@@
 The definition for $\var{add}$ is unchanged, but the instantiation of
-open derivations enables to directly substitute one in the other and
+open derivations enables to directly substitute one into the other and
 get the appropriate program: call $\md_{10}$ the whole derivation just
@@
 A few days earlier, the teacher is preparing the tool for the students
-to train composing System \sysname{T} programs. He (or she) provides
-the abstract syntax, the typing rules and the typing untrusted
-algorithm in order to build the incremental type checker. How should
-he write the algorithm? We know that we want to return a derivation,
-and take advantage of a higher-order representation of the syntax, so
-it shouldn't be necessary to thread any explicit environment. The
-definition should then start with:
+to train composing System \sysname{T_{<:}} programs. He (or she)
+provides the abstract syntax, the typing rules and the untrusted
+typing algorithm in order to build the incremental type checker. How
+should he write the algorithm? We know that we want to return a
+derivation, and take advantage of a higher-order representation of the
+syntax, so it shouldn't be necessary to thread any explicit
+environment. The definition should then start with:
 
@@
 \noindent
-It takes a term $M$ and producing the pair of a type $A$ and a
-derivation that $M$ has type $A$. Its code will be untyped, but
-the teacher provides this type so that the checker can verify that
-each call is fed and returns values of the right type: it is its
-\emph{contract} so to say. He first decomposes term $M$ and treats
-the easy cases of $\cst{o}$, $\cst{s}$ and $\cst{app}$:
+Function $\finfer{}$ takes a term $M$ and produces the pair of an
+inferred type $A$ and a derivation that $M$ has type $A$. Its code
+will be untyped, but the teacher provides this type so that the
+checker can verify that each call is fed and returns values of the
+right type: it is its \emph{contract} so to say. He first decomposes
+term $M$ and treats the easy cases of $\cst{o}$ and $\cst{s}$:
 
@@
   \lamd M \match{M} \\
-  \quad\case{\z} \pair {\cst{nat}}{\infer{ }{\vdash \cst o : \cst{nat}}} \\
+  \quad\case{\z} \pair {\cst{even}}{\infer{ }{\vdash \cst o : \cst{even}}} \\
   \quad\case{\s M}
-  \letd {\pair {\cst{nat}} \md} {\finfer{M}}
+  \match{\finfer M} \\
+  \quad\quad\case{\pair{\cst{even}}\md}
+  \pair {\cst{odd}}{\infer\md{\vdash\app{\cst{s}} M : \cst{odd}}} \\
+  \quad\quad\case{\pair{\cst{odd}}\md}
+  \pair {\cst{even}}{\infer\md{\vdash\app{\cst{s}} M : \cst{even}}} \\
+  \quad\quad\case{\pair{\cst{nat}}\md}
   \pair {\cst{nat}}{\infer\md{\vdash\app{\cst{s}} M : \cst{nat}}} \\
-  \quad\case{\app{M}N}
-  \letd {\pair {A_1\to B} {\md_1}} {\finfer{M}} \\
+\end{mathleft}
+
+\noindent
+We write $\pair{\cdot}{\cdot}$ the constructor of $\Sigma$-types;
+patterns don't have to be exhaustive since we are in an untrusted
+setting; for conciseness, we use case failure for signaling typing
+error. There are three cases for $\s{M}$ in the system, corresponding
+to the three rules in the declarative system. For the application
+case, we rely on the well-known property that we may need to insert a
+coercion from the actual type of the argument to its expected type:
+
+\begin{mathleft}
+  \quad\case{\app{M}N} \\
+  \quad\quad\letd {\pair {A_1\to B} {\md_1}} {\finfer{M}} \\
   \quad\quad\letd {\pair {A_2} {\md_2}} {\finfer{N}} \\
-  \quad\quad\letd {\cst{true}} {\feq {A_1} {A_2}}
-  \pair {B} {\infer{\md_1 \and \md_2}{\vdash \app M N : B}}
+  \quad\quad\letd {\md_{\leq}} {\fleq {A_1} {A_2}} \\
+  \quad\quad
+  \match{\md_{\leq}} \\
+  \quad\quad\quad\case{\infer{ }{A\leq A}}
+  \pair {B} {\infer{\md_1 \and \md_2}{\vdash \app M N : B}} \\
+  \quad\quad\quad\case{\_}
+  \pair {B} {\infer{\md_1 \and \infer*{\md_2 \and \md_{\leq}}{\vdash N
+      : A_1}}{\vdash \app M N : B}}
 \end{mathleft}
@@
 \noindent
-We write $\pair{\cdot}{\cdot}$ the constructor of $\Sigma$-types; the
-$\syntax{let}$ construct is just syntactic sugar for a one-branch
-$\syntax{case}$; patterns don't have to be exhaustive since we are in
-an untrusted setting; for conciseness, we use case failure for
-signaling typing error; we use an auxiliary function $\!\!\feq{}{}\!\!
-: \cst{tp}\to\cst{tp}\to\cst{bool}$ for syntactic comparison of types,
-defined the same way.
+The $\syntax{let}$ construct is just syntactic sugar for a one-branch
+$\syntax{case}$; we use an auxiliary function $$\fleq{}{}\ :\
+\prd A {\cst{tp}} \prd B {\cst{tp}} \vdash A\leq B = \ldots$$ trying to
+construct a subtyping derivation for its argument, or failing with an
+error (definition not included here). Notice the slight optimization
+to the size of the derivation: if the types $A_1$ and $A_2$ are the
+same syntactically, we don't generate a subtyping rule. This is done by
+matching on the returned derivation $\md_\leq$.
 
@@
 
-The case of $\cst{rec}$ is similar: we first compute recursively the
-derivations $\md_M$ and $\md_N$ which should be well-typed in the
-current environment, and then $\md_P$ in the enlarged environment,
-with variables $\var x$ and $\var y$ in $P$ substituted with the
-(future) result of $\finfer{}$.
+For what concerns the treatment of environments, the case of $\recb M
+N x y P$ is similar: we first compute recursively the derivations
+$\md_M$ and $\md_N$ which should be well-typed in the current
+environment, and then $\md_P$ in the enlarged environment, with
+variables $\var x$ and $\var y$ in $P$ substituted with the (future)
+result of $\finfer{}$. The problem is with subtyping: terms $N : A_N$
+and $P : A_P$ should be coercible to the same type $A$, but $P$ should
+be typed under $\vdash y:A$, which is not known before typing $P$. The
+trick to break the loop is to iterate the typing of $P$ with more and
+more general types, until we reach a fixpoint (this is finite if there
+is a unique more general type); in our case, that means iterating
+twice: we will then attain \cst{nat}.
 
@@
   \quad\quad
-  \letd {\pair {\cst{nat}} {\md_M}} {\finfer{M}} \\
+  \letd {\pair {A_M} {\md_M}} {\finfer{M}} \\
+  \quad\quad\letd {\md_{A_M}} {{A_M} \leq \cst{nat}} \\
   \quad\quad
-  \letd {\pair {A_1} {\md_N}} {\finfer{N}} \\
+  \letd {\pair {A_N} {\md_N}} {\finfer{N}} \\
   \quad\quad
-  \syntax{let} {\pair {A_2} {\md_P}} \syntax{in}
+  \syntax{let} {\pair {A_P} {\md_P}} \syntax{in}
   {(\envcons {\env {$\md_x$} {(\vdash
-        \var x : \cst{nat})}} {$\md_y$} {(\vdash\var y : A_1)})} = \\
+        \var x : \cst{nat})}} {$\md_y$} {(\vdash\var y : A_N)})} = \\
   \quad\quad\quad
@@
           {\fget{\pair{\cst{nat}}{\md_x}}}} y
-        {\fget{\pair{A_1}{\md_y}}}}}} \\
-  \quad\quad\letd {\cst{true}} {\feq {A_1} {A_2}} \\
+        {\fget{\pair{A_N}{\md_y}}}}}} \syntax{in}\\
+  \quad\quad\letd {\pair A {\pair{\md_{A_N}}{\md_{A_P}}}} {{A_N} \sqcap {A_P}} \\
   \quad\quad
-  \pair {A} {
-    \infer{
-      \md_M \and
-      \md_N \and
-      \mbox{$
+  \syntax{let} {\pair {\_} {\md_P}} \syntax{in}
+  {(\envcons {\env {$\md_x$} {(\vdash
+        \var x : \cst{nat})}} {$\md_y$} {(\vdash\var y : A)})} = \\
+  \quad\quad\quad
+  {\finfer{\gsubst P {\msubstcons {\msubst x
+          {\fget{\pair{\cst{nat}}{\md_x}}}} y
+        {\fget{\pair{A}{\md_y}}}}}} \syntax{in}\\
+  \hspace{-2em}\pair {A} {
+    \infer{
+      \infer*{\md_M \and \md_{A_M}}
+      {\vdash M : A} \and
+      \infer*{\md_N \and \md_{A_N}}
+      {\vdash N : A}\and
+      \infer*{\infer*{}{\mbox{$
         \begin{array}{c}
@@
       \end{array}
-        $}
-    }{
-      \vdash \recb M N x y P
+        $}} \and \md_{A_P}}{\vdash P : A}
+    }{
+      \vdash \recb M N x y P : A
     }
@@
 \end{mathleft}
+
+\noindent
+Again, we use a new auxiliary function
+$$\sqcap\ :\ \prd{A}{\cst{tp}} \prd{B}{\cst{tp}} \sig C {\cst{tp}}
+(\vdash A\leq C )\times(\vdash B\leq C) = \ldots$$ computing the sup
+of two types $A$ and $B$, and returning proofs of of subtyping. Note
+that we didn't perform here the optimization seen in the application
+case.
 
@@
 boundaries of defined functions. This allows to omit formal
-justifications, like the fact that $B$ can't depend on $\md_x$ in the
-$\lambda$ case, but of course incurs on the performance. This is
-compensated by the slicing mechanism described below in
+justifications, like the fact that we need to iterate twice the typing
+of $P$ in $\recb M N x y P$, or that $B$ can't depend on $\md_x$ in
+the $\lambda$ case. Of course this incurs on the performance; we will
+see a compensation for it by the slicing mechanism described below in
 \ref{sec:repr}.
+
+A similar idea is very wide-spread in \sysname{LCF}-style proof
+assistants: during manual proof search, the interaction is driven by
+\emph{tactics}: %TODO cite
+they are untrusted programs generating pieces of proofs, assembled by
+other, higher-order tactics; verification of the generated proofs is
+done when the evaluation of all tactics is complete. The differences
+with our mechanism are two-fold: first, tactics usually take one (or
+several) \emph{goal}(s) as input, \ie\ holes in a proof, whereas our
+functions can take any term; secondly, proof-check is usually done at
+the end of the evaluation, which can lead to less tractable bugs; by
+making evaluation and checking mutually recursive, we choose to detect
+errors earlier in these witnesses: exactly at the point where they are
+generated.
 
@@
   x}$. This requires evaluating the type-checker with a carefully
-designed strategy.
-
-For the sake of conciseness, we took a simple, syntax-directed system
-as example; this scales nonetheless to declaratively presented
-features like subtyping or conversion.
+designed strategy (see \ref{sec:comput}).
+
+% TODO cite lee2007mechanizing
 
@@
 \subsection{The computational calculus}
+\label{sec:comput}
 

### 2012-05-31 doc/papers/LFMTP2012/paper.tex
paper

@@
 design is not to be demonstrated anymore, both in terms of the safety
-of programs --- they ``do not go wrong'', and programming practices
---- modularity, early bug catching. Verifying well-typing of
+of program execution --- they ``do not go wrong'', and programming
+practices --- modularity, early bug catching. Verifying well-typing of
 real-world programs is thus a critical matter, but it is a complex
@@
 highly non-trivial and non-modular: for instance, how to infer the
-type of a System \sysname{T} recursor
-$$
-\infer{
-  \Gamma\vdash M : \cst{nat}
-  \and
-  \Gamma\vdash N : A
-  \and
-  \Gamma, \var x:\cst{nat}, \var y:A\vdash P : A
-}{
-  \Gamma\vdash \recb M N x y P
-}
-$$
-in presence of subtyping? The problem is that we have to guess a type
-$A$ more general than that of $N$ and $P$, while giving that type to
-variable $\var y$. A well-known algorithm if the subtyping relation
-forms a complete lattice is to iterate the typing of $P$ with the join
-of the previously computed and the new type until a fixpoint is
-reached.
+type of a System \sysname{T} recursor $\recb M N x y P$, in presence
+of subtyping? The problem is that we have to guess a type $A$ more
+general than that of $N$ and $P$, while giving that type to variable
+$\var y$. A well-known algorithm if the subtyping relation forms a
+complete lattice is to iterate the typing of $P$ with the join of the
+previously computed and the new type until a fixpoint is reached.
 
@@
 typing derivation it provides. Let us then imagine a type-checker to
-be the pair of 1/ an untrusted procedure $\finfer M$ to infer a typing
-derivation out of program $M$; it doesn't have to be complete, or even
-sound, and 2/ a trusted and fast kernel $\fcheck{\md}$ to verify these
-derivations in a given set of inference rules $\Sigma$. This is a
-\emph{certifying} scheme (see \cite{leroy2006compcert} for a
-discussion on approaches to reliable compilers):
-if $\fcheck{\finfer M} = \cst{true}$ then there really \emph{is} a
-derivation $\md$ for $M$, but the converse is not necessarily true:
-function $\finfer{}\!\!$ could fail, or have bugs (return an incorrect
-derivation). Explicitly generating the derivation in the declarative
-system has the following advantages over the first approach:
+be the pair of \emph{(i)} an untrusted procedure $\finfer M$ to infer
+a typing derivation out of program $M$; it doesn't have to be
+complete, or even sound, and \emph{(ii)} a trusted and fast kernel
+$\fcheck{\md}$ to verify these derivations \wrt\ a given set of inference
+rules $\Sigma$. This is a \emph{certifying} scheme (see
+\cite{leroy2006compcert} for a discussion on approaches to reliable
+compilers): if $\fcheck{\finfer M} = \cst{true}$ then there really
+\emph{is} a derivation $\md$ for $M$, but the converse is not
+necessarily true: function $\finfer{}\!\!$ could fail, or have bugs
+(return an incorrect derivation). Explicitly generating the derivation
+in the declarative system has the following advantages over the first
+approach:
 \begin{itemize}
@@
 As type systems become richer, it becomes increasingly difficult for
-the programmer to write a correct program in one try, compile it and
-run it: writing a program longer than a few dozens of lines becomes a
-tight and non-linear interaction between the human and the
+the programmer to write a well-typed program in one try, compile it
+and run it: writing a program longer than a few dozens of lines
+becomes a tight and non-linear interaction between the human and the
 type-checker involving constant experiments, fixes, etc. If this
@@
 procedure, we could recognize already-built subderivations for
-subterms common to a checked and a new version of the program, and
-graft these pieces of derivation to form a new, large one for the
-whole modified program, saving the regeneration of all equal subterms'
-derivations in compatible contexts. This way we constantly maintain
-the verifiable witness of well-typing for the program, and take
-advantage of the quotients made on higher-order syntax.
-
-We can thus add to the previous list that explicitly generating typing
-derivations makes possible a form of safe incrementality by sharing
-common subderivations.
+subterms common to an old, checked version and a new, to-be-checked
+version, and graft together these pieces of derivation to form a new,
+large one for the new version of the program, saving the regeneration
+of all equal subterms' derivations in compatible contexts. This way we
+constantly maintain the verifiable witness of well-typing for the
+program, and take advantage of the quotients made on higher-order
+syntax.
+
+We can thus add to the list of advantages begun earlier that
+explicitly generating typing derivations makes possible a form of safe
+incrementality by sharing common subderivations.
 
@@
 
+\begin{figure}
+  \centering
+%  \small
+  \begin{mathpar}
+    \infer{ }{\jts{\cst o} {\even}}
+
+    \infer{\jts M \odd} {\jts {\s M} \even}
+
+    \infer{\jts M \even} {\jts {\s M} \odd}
+
+    \infer{\jts M \nat} {\jts {\s M} \nat}
+
+    \infer{\jts M {A\to B} \and \jts N A} {\jts {\app M N} B}
+
+    \infer{\infer*{}{\mbox{$
+          \begin{array}{c}
+            [\jts {\var x} A] \\[-0.3em]
+            \vdots\\
+            \jts M B
+          \end{array}$
+        }}} {\jts {\lam x M} {A\to B}}
+
+    \infer{\jts M A \and \jsub A B}{\jts M B}
+
+    \infer{\jts M \nat
+      \and
+      \jts N A
+      \and
+      \infer*{}{\mbox{$
+        \begin{array}{c}
+          [\jts {\var x} \nat] \quad
+          [\jts {\var y} A] \\[-0.3em]
+          \vdots \\
+          \jts P A
+        \end{array}
+        $}}
+    }{\jts {\recb M N x y P} A}
+
+    \infer{ }{\jsub A A}
+
+    \infer{ }{\jsub \even\nat}
+
+    \infer{ }{\jsub \odd\nat}
+
+    \infer{\jsub {A'} A \and \jsub B {B'} }{\jsub {A\to B}{A'\to B'}}
+  \end{mathpar}
+  \caption{System \sysname{T_{<:}}, declaratively}
+  \label{fig:tsub}
+\end{figure}
+
 A student is learning about System \sysname {T_{<:}} --- a simply
@@
 two sets $\cst{even}$ and $\cst{odd}$, introduction $\z$ and $\s{M}$,
-elimination ($\recb M N x y P$), and subtyping on those types. He (or
-she) wants to write a program computing $(2+2)\times 3$. He knows that
-sometimes those $\cst{rec}$ arguments are quite hard to get correct,
-but \sysname{T_{<:}} is a typed language; he wants to take advantage
-of this to build his program incrementally using the teacher's
-provided incremental verifier. As a first attempt, he wants to verify
-$P_1$:
+elimination ($\recb M N x y P$), and subtyping on those types; see
+Fig. \ref{fig:tsub}. He (or she) wants to write a program computing
+$(2+2)\times 3$. He knows that sometimes those $\cst{rec}$ arguments
+are quite hard to get correct, but \sysname{T_{<:}} is a typed
+language; he wants to take advantage of this to build his program
+incrementally using the teacher's provided incremental verifier. As a
+first attempt, he wants to verify $P_1$:
 $$
@@
 a derivation\ \ $\vdash\var x:\cst{nat}$, we instantiate this hole
-with the derivation $\finfer{\var u}$ inferred recursively for then
-new recursive argument $\var u$. $\md_2$ derives $\vdash \s\z :
+with the derivation $\finfer{\var u}$ inferred recursively for the new
+recursive argument $\var u$. $\md_2$ derives $\vdash \s\z :
 \cst{odd}$, but $\var{add}$ awaits $\cst{nat}$ arguments. When the
@@
   \match{\md_{\leq}} \\
-  \quad\quad\quad\case{\infer{ }{A\leq A}}
+  \quad\quad\quad\case{\infer{ }{\vdash A\leq A}}
   \pair {B} {\infer{\md_1 \and \md_2}{\vdash \app M N : B}} \\
@@
 The $\syntax{let}$ construct is just syntactic sugar for a one-branch
-$\syntax{case}$; we use an auxiliary function $$\fleq{}{}\ :\
-\prd A {\cst{tp}} \prd B {\cst{tp}} \vdash A\leq B = \ldots$$ trying to
+$\syntax{case}$; we use an auxiliary function $$\fleq{}{}\ :\ \prd A
+{\cst{tp}} \prd B {\cst{tp}} \vdash A\leq B = \ldots$$ trying to
 construct a subtyping derivation for its argument, or failing with an
-error (definition not included here). Notice the slight optimization
-to the size of the derivation: if the types $A_1$ and $A_2$ are the
-same syntactically, we don't generate a subtyping rule. This is done by
-matching on the returned derivation $\md_\leq$.
+error. Notice the slight optimization to the size of the derivation:
+if the types $A_1$ and $A_2$ are the same syntactically, we don't
+generate a subtyping rule. This is done by matching on the returned
+derivation $\md_\leq$.
 
@@
 sugar for $\matchin M \envnil \ldots$). Even then, in the recursive
-call, arriving to variable $\var x$ would fail since we would have no
-way of telling its type anymore. Thus we substitute all occurrences of
+call, the variable case $\var x$ would fail since we would have no way
+of telling its type anymore. Thus we substitute all occurrences of
 $\var x$ with the \emph{output} of function $\finfer{}$ on it (since
@@
 substitution to be well-typed, we now need to coerce this output to a
-valid input (a term). We introduce an operator which is in some way
+valid input (a term). We introduce an operator which is in some sense
 the \emph{inverse} of $\finfer{}$: operator $\fget{}$ maps a pair of a
@@
 At this point, we get back $B$ and $\md$ from the recursive call. We
-know that $B$ can't depend on $\md_x$ (because a type is always
-closed), but $\md$ can: it is not \emph{per se} a hypothetical
-derivation but an \emph{open} derivation where a special judgement
-named $\md_x$ can occur. To return the complete derivation of
-$\vdash\tlam A x M : A \to B$, we must then close, or ``hypothesize''
-$\md$ by $\md_x$, and then apply rule \rulename{Lam}:
+know that $B$ can't depend on $\md_x$ (because a System
+\sysname{T_{<:}} type is always closed), but $\md$ can: it is not
+\emph{per se} a hypothetical derivation but an \emph{open} derivation
+where a special judgement named $\md_x$ can occur. To return the
+complete derivation of $\vdash\tlam A x M : A \to B$, we must then
+close, or ``hypothesize'' $\md$ by $\md_x$, and then apply rule
+\rulename{Lam}:
 
@@
 
-For what concerns the treatment of environments, the case of $\recb M
-N x y P$ is similar: we first compute recursively the derivations
-$\md_M$ and $\md_N$ which should be well-typed in the current
-environment, and then $\md_P$ in the enlarged environment, with
-variables $\var x$ and $\var y$ in $P$ substituted with the (future)
-result of $\finfer{}$. The problem is with subtyping: terms $N : A_N$
-and $P : A_P$ should be coercible to the same type $A$, but $P$ should
-be typed under $\vdash y:A$, which is not known before typing $P$. The
-trick to break the loop is to iterate the typing of $P$ with more and
-more general types, until we reach a fixpoint (this is finite if there
-is a unique more general type); in our case, that means iterating
-twice: we will then attain \cst{nat}.
+Concerning the treatment of environments, the case of $\recb M N x y
+P$ is similar: we first compute recursively the derivations $\md_M$
+and $\md_N$ which should be well-typed in the current environment, and
+then $\md_P$ in the enlarged environment, with variables $\var x$ and
+$\var y$ in $P$ substituted with the (future) result of $\finfer{}$.
+
+Remains one problem with subtyping: terms $N : A_N$ and $P : A_P$
+should be coercible to the same type $A$, but $P$ should be typed
+under $\vdash y:A$, which is not known before typing $P$. A trick to
+break the loop is to iterate the typing of $P$ with more and more
+general types, until we reach a fixpoint (we reach a fixpoint if there
+is a unique more general type); in our particular case of subtyping
+relation, this means iterating only twice: we will then attain
+\cst{nat}.
 
@@
 won't ever meet this case: we already substituted all variables with
-their result derivation. We only need to make sure that in this case,
-the derivation will be returned:
+their result derivation. Evaluation only need to make sure that in
+this case, the derivation will be returned, \ie\ that it reduces the
+following way:
 $$
-\finfer{\fget{\pair A \md}} = \md
+\finfer{\fget{\pair A \md}} = \pair A \md
 $$
@@
 
+% TODO confirmer qu'on met bien la def. des f^0
 \begin{mathleft}
@@
     M:A)} \cst{tm} =
-  \\\quad \lamd M \lamd A \lamd \md M
+  \\\qquad\quad \lamd M \lamd A \lamd \md M
 \end{mathleft}
@@
 
-A similar idea is very wide-spread in \sysname{LCF}-style proof
+The idea of untrustedly generating certificates verified \emph{a
+  posteriori} is very wide-spread in \sysname{LCF}-style proof
 assistants: during manual proof search, the interaction is driven by
@@
 Introducing the ``inverse'' of the derivation inference function has
-two purpose: as we saw in \ref{sec:use-incremental}, it allows to
+two purposes: as we saw in \ref{sec:use-incremental}, it allows to
 refer to already-constructed derivations when writing deltas. It also
@@
 $$
+% TODO: vérif ce paragraphe après avoir parlé de fcts inv
 replace the usual ones. The difficulty with this technique,
@@
 subterm and instantiate its free variables. For this, we define the
-the \emph{slicing} of a well-typed \LF\ term: we extend \LF\ with
-metavariables $\meta X$, $\meta Y$\ldots\ as in
+\emph{slicing} of a well-typed \LF\ term: we extend \LF\ with
+contextual metavariables $\meta X$, $\meta Y$\ldots\ as in
 \cite{nanevski2008contextual}. Metavariables stand for well-typed
@@
 $$
-\fbox{$\app {\cst{lam}} \lam x
-\fbox{$\app {\cst{lam}} \lam y
-\fbox{\app {\cst{app}} \app {\var y}
-}$}
-{(\fbox{\app
-  {\cst{app}}\app{\var x}{\cst o}})
-}
-$}$$
+\fbox{$
+  \app {\cst{lam}} \lam x
+  \fbox{$
+    \app {\cst{lam}} \lam y
+    \fbox{$
+      \app {\cst{app}} \app {\var y}
+      (\fbox{$
+        \app{\cst{app}}\app{\var x}{\cst o}
+      $})
+    $}
+  $}
+$}
+$$
 where all boxed subterms are referable slices.
@@
 
-We distinguish syntactically between \emph{checkable} and
-\emph{inferrable} objects (resp. \emph{canonical} and \emph{atomic}),
-an idea dating back to \cite{pierce2000local}: canonical objects are
-checked against a given type, whereas we can synthesize the type of
-atomic objects; the coercion from the first to the second will be the
-place where to verify equality between the synthesized type and the
-given one.
+We distinguish syntactically between \emph{checkable} $M$ and
+\emph{inferrable} $F$ objects (resp. \emph{canonical} and
+\emph{atomic}), an idea dating back to \cite{pierce2000local}:
+canonical objects are checked against a given type, whereas we can
+synthesize the type of atomic objects; the coercion from the first
+grammatical category to the second is where we verify equality between
+the synthesized and the checked types.
 
@@
 not stable by (textual) substitution, it is defined as
-\emph{hereditary}: it is mutually recursive with a \emph{cut} function
-$\gcut M S$ which can in turn trigger a chain of substitutions,
-leading to a canonical objects or an error.
+\emph{hereditary}\footnote{ Note that this notion of substitution,
+  present in \cite{pfenning2007term} differs from
+  \cite{hl07mechanizing} as it presupposes $\eta$-long normal forms:
+  once a \emph{cut} starts, there must be exactly as many formal and
+  actual arguments. A non-empty spine cut against another non-empty
+  spine $\gcut {\app H S} {S'}$ blocks the substitution instead of
+  reducing to $\app H {(\gconcat S S')}$ (an implicit
+  $\eta$-expansion); conversely, exhausting actual but not formal
+  arguments as in $\gcut{\lam x M}{\gnil}$ blocks the substitution
+  too. For example, the reduction $\gcut {\lam x \app {\var x}{\var
+      x}} {\lam x \app {\var x}{\var x}}$ gets stuck because the
+  second $\var x$ in the left part is substituted by a function yet is
+  not itself a function. }: it is mutually recursive with a \emph{cut}
+function $\gcut M S$ which can in turn trigger a chain of
+substitutions, leading to a canonical objects or an error.
 \begin{align*}
@@
 if it is defined.
-
-Note that this notion of substitution differs from hereditary
-substitution found in \eg\ \cite{hl07mechanizing} as it presupposes
-$\eta$-long normal forms: once a \emph{cut} starts, there must be
-exactly as many formal and actual arguments. A non-empty spine cut
-against another non-empty spine $\gcut {\app H S} {S'}$ blocks the
-substitution instead of reducing to $\app H {(\gconcat S S')}$ (an
-implicit $\eta$-expansion); conversely, exhausting actual but not
-formal arguments as in $\gcut{\lam x M}{\gnil}$ blocks the
-substitution too. For example, the reduction $\gcut {\lam x \app {\var
-    x}{\var x}} {\lam x \app {\var x}{\var x}}$ gets stuck because the
-second $\var x$ in the left part is substituted by a function yet is
-not itself a function.
 
@@
 
-Our algorithm types exactly the well-typed objects of the usual
-\LF, except that we thread a repository $\mr$ and store in it
-all well-typed applicative subterms $F$ along with their classifier
-$A$ and the environment in which they are well-typed.
+Our algorithm accepts exactly the well-typed objects of the usual \LF,
+except that we thread a repository $\mr$ and store in it all
+well-typed applicative subterms $F$ along with their classifier $A$
+and the environment in which they are well-typed.
 
@@
 judgements of object well-typing, along with a distinguished
-metavariable, the \emph{head}:
-$$ \mr\;:\;(\meta X \mapsto (\Gamma\vdash\tp M A)), \smeta X \sigma $$
-We maintain the invariant that a repository $\mr$ is
-\emph{acyclic}, that is for all $\meta X$ in $\mr$, the fixpoint
-of $\FMV{\meta X}$ does not contain $\meta X$. If $\mr =
-(\Delta, X[\sigma])$, we write $\hat {\mr}$ for $X[\sigma]$,
-abusively $\mr$ for $\Delta$ and $\mr (\smeta X \sigma)$ for
-$\gsubst M\sigma$ if $\Delta(\meta X) = (\Gamma\vdash M : A)$.
+metavariable, corresponding to the \emph{tip} of the term contained in
+it:
+$$ \mr\;:\;(\meta X \mapsto (\Gamma\vdash\tp M A)), \smeta X
+\sigma $$
+The algorithm will maintain the invariant that a repository
+$\mr$ is \emph{acyclic}, that is for all $\meta X$ in $\mr$, the
+fixpoint of $\FMV{\meta X}$ does not contain $\meta X$. If $\mr =
+(\Delta, X[\sigma])$, we write $\hat {\mr}$ for $X[\sigma]$, abusively
+$\mr$ for $\Delta$ and $\mr (\smeta X \sigma)$ for $\gsubst M\sigma$
+if $\Delta(\meta X) = (\Gamma\vdash M : A)$.
 
@@
 
-\begin{theorem}
-  If \/ $\jm\mr\Gamma M A \mmr {M'}$ then \/ $\jem\mmr\Gamma M {M'} A$
-\end{theorem}
-
-\begin{theorem}
-  If \/ $\jm\mr\Gamma M A \mmr {M'}$ then \/ $\jlfm\Gamma {(\checkoutrr
-    M)} (\checkoutrr A)$
-\end{theorem}
+% \begin{theorem}
+%   If \/ $\jm\mr\Gamma M A \mmr {M'}$ then \/ $\jem\mmr\Gamma M {M'} A$
+% \end{theorem}
+
+% \begin{theorem}
+%   If \/ $\jm\mr\Gamma M A \mmr {M'}$ then \/ $\jlfm\Gamma {(\checkoutrr
+%     M)} (\checkoutrr A)$
+% \end{theorem}
 
@@
 defined constants as mere black boxes taking and producing terms, and
-being able to call other black boxes. We detail for example in section
-\ref{sec:implem} how to embed \sysname{OCaml} as computation calculus.
+being able to call other black boxes.
 
@@
 
-\section{Implementation}
-\label{sec:implem}
 \section{Related work}

### 2012-06-07 doc/papers/LFMTP2012/paper.tex
paper (minor corrections)

@@
 $$
-\infer{\Gamma\vdash M : A \\ A\leq B}{\Gamma\vdash M : B}
+\infer{\vdash M : A \\ A\leq B}{\vdash M : B}
 $$
@@
 
-Several methods have been devised for turning a batch procedure into
-an incremental one \cite{pugh1989incremental,
+Several generic methods have been devised for turning a batch program
+into an incremental one \cite{pugh1989incremental,
   acar2003selective}. Unfortunately, they all suffer from their

### 2012-06-08 doc/papers/LFMTP2012/paper.tex
paper: rename case to caseb (clashes with llncs)
paper: switch style from ACM to LNCS

@@
-\documentclass[9pt]{sigplanconf}
+\documentclass{llncs}
 
@@
 \usepackage{amssymb}
-\newtheorem{theorem}{Theorem}
-\newtheorem{definition}{Definition}
 \usepackage{macros}
@@
 
+\allowdisplaybreaks[1]
+
 \begin{document}
 
-\conferenceinfo{LFMTP '12}{September 9, 2012 -- Copenhagen, Denmark}
-\copyrightyear{2012}
-\copyrightdata{[to be supplied]}
-
-\titlebanner{banner above paper title}        % These are ignored unless
-\preprintfooter{short description of paper}   % 'preprint' option specified.
-
 \title{Incremental tactics}
-%\subtitle{Subtitle Text, if any}
-
-\authorinfo{Matthias Puech}
-           {Department of Comp. Sci., Univ. of Bologna,\\
-             PPS, Team $\pi r^2$ (Univ. Paris Diderot, CNRS, INRIA)}
-           {puech@cs.unibo.it}
-
-\authorinfo{Yann R\'egis-Gianas}
-           {PPS, Team $\pi r^2$ (Univ. Paris Diderot, CNRS, INRIA)}
-           {yrg@pps.jussieu.fr}
+
+\author{Matthias Puech\inst{1,2} \and Yann R\'egis-Gianas\inst{1}}
+\institute{
+  PPS, Team $\pi r^2$ (Universit\'e Paris Diderot, CNRS, INRIA)
+  \and
+  Department of Computer Science, University of Bologna
+}
 
@@
 
-%% Here is a problem
 \begin{abstract}
@@
 \end{abstract}
-
-\category{D.3.3}{Language Constructs and Features}{Data types and
-  structures} \category{F.3.1}{Logics and Meanings of
-  Programs}{Specifying and Verifying and Reasoning about
-  Programs --- Logics of programs}
-
-\terms
-Theory, Languages
-
-\keywords
-incrementality, type checking, logical framework, version control
 
@@
 Several generic methods have been devised for turning a batch program
-into an incremental one \cite{pugh1989incremental,
-  acar2003selective}. Unfortunately, they all suffer from their
-genericity on the particular problem of type checking:
+into an incremental one
+\cite{pugh1989incremental,acar2003selective}. Unfortunately, they all
+suffer from their genericity on the particular problem of type
+checking:
 \begin{itemize}
@@
   parts;
-\item finally, there is no way to \emph{a posteriori} validate the
-  result of the incremental process: the user has to trust not only
-  the type-checking algorithm but also its transformation into an
+\item finally, there is no way to a posteriori validate the result of
+  the incremental process: the user has to trust not only the
+  type-checking algorithm but also its transformation into an
   incremental one, which increases the size of the trusted base.
@@
 technique for incremental computation that has been extensively
-studied and combined with other techniques \cite{pugh1989incremental,
-  acar2003selective}. However, it is best understood in the case of
-purely first-order data, and to our best knowledge remains unexplored
-for higher-order structures containing binders, like programs and
-proofs.
+studied and combined with other techniques
+\cite{pugh1989incremental,acar2003selective}. However, it is best
+understood in the case of purely first-order data, and to our best
+knowledge remains unexplored for higher-order structures containing
+binders, like programs and proofs.
 
@@
   \lamd M \match{M} \\
-  \quad\case{\z} \pair {\cst{even}}{\infer{ }{\vdash \cst o : \cst{even}}} \\
-  \quad\case{\s M}
+  \quad\caseb{\z} \pair {\cst{even}}{\infer{ }{\vdash \cst o : \cst{even}}} \\
+  \quad\caseb{\s M}
   \match{\finfer M} \\
-  \quad\quad\case{\pair{\cst{even}}\md}
+  \quad\quad\caseb{\pair{\cst{even}}\md}
   \pair {\cst{odd}}{\infer\md{\vdash\app{\cst{s}} M : \cst{odd}}} \\
-  \quad\quad\case{\pair{\cst{odd}}\md}
+  \quad\quad\caseb{\pair{\cst{odd}}\md}
   \pair {\cst{even}}{\infer\md{\vdash\app{\cst{s}} M : \cst{even}}} \\
-  \quad\quad\case{\pair{\cst{nat}}\md}
+  \quad\quad\caseb{\pair{\cst{nat}}\md}
   \pair {\cst{nat}}{\infer\md{\vdash\app{\cst{s}} M : \cst{nat}}} \\
@@
 \begin{mathleft}
-  \quad\case{\app{M}N} \\
+  \quad\caseb{\app{M}N} \\
   \quad\quad\letd {\pair {A_1\to B} {\md_1}} {\finfer{M}} \\
@@
   \match{\md_{\leq}} \\
-  \quad\quad\quad\case{\infer{ }{\vdash A\leq A}}
+  \quad\quad\quad\caseb{\infer{ }{\vdash A\leq A}}
   \pair {B} {\infer{\md_1 \and \md_2}{\vdash \app M N : B}} \\
-  \quad\quad\quad\case{\_}
+  \quad\quad\quad\caseb{\_}
   \pair {B} {\infer{\md_1 \and \infer*{\md_2 \and \md_{\leq}}{\vdash N
@@
 \begin{mathleft}
-  \quad\case{\tlam x A M} \\
+  \quad\caseb{\tlam x A M} \\
   \quad\quad
@@
 \begin{mathleft}
-  \quad\case{\recb M N x y P} \\
+  \quad\caseb{\recb M N x y P} \\
   \quad\quad
@@
   \matchin U \Gamma C \\
-  C &\gequal \casenobar P U \gor
-  C\ \case P U \\
+  C &\gequal \casebnobar P U \gor
+  C\ \caseb P U \\
   P &\gequal
@@
     U\sigma} {\gsubst\Gamma\sigma} {\gsubst C\sigma} \\
-  \gsubst {(\casenobar{\app H {\spinecons {\var x_1} {\spinecons \ldots {\var
-          x_n}}}} U)} \sigma &= {\casenobar{\app H {\spinecons {\var x_1} {\spinecons \ldots {\var
+  \gsubst {(\casebnobar{\app H {\spinecons {\var x_1} {\spinecons \ldots {\var
+          x_n}}}} U)} \sigma &= {\casebnobar{\app H {\spinecons {\var x_1} {\spinecons \ldots {\var
           x_n}}}} \gsubst U \sigma} \quad \text{if $\var x_i\notin\dom\sigma$} \\
@@
       \jcr\mr\Gamma{\matchin{U}{\Gamma'}{
-          \casenobar{\app {H_1}{\spinecons {\var x_{11}} {\spinecons
+          \casebnobar{\app {H_1}{\spinecons {\var x_{11}} {\spinecons
                 \ldots\ {\var x_{1m_1}}}}}{U_1}
           |\ \ldots\
-          \case{\app {H_n}{\spinecons {\var x_{n1}} {\spinecons
+          \caseb{\app {H_n}{\spinecons {\var x_{n1}} {\spinecons
               \ldots\ {\var x_{nm_n}}}}}{U_n}
@@
 
-\bibliographystyle{abbrvnat}
-\bibliography{../../english.bib}
+\bibliographystyle{splncs}
+\bibliography{../../english}
 

### 2012-06-08 doc/papers/LFMTP2012/paper.tex
paper: head -> tip + misc. little improvements

@@
 variables. A repository also contains a distinguished metavariable
-corresponding to the head of the term stored in it. For instance, the
+corresponding to the ``tip'' of the term stored in it. For instance, the
 repository
 $$
+\small
 \left(\begin{array}{llll}
@@
 \emph{hereditary}\footnote{ Note that this notion of substitution,
-  present in \cite{pfenning2007term} differs from
-  \cite{hl07mechanizing} as it presupposes $\eta$-long normal forms:
-  once a \emph{cut} starts, there must be exactly as many formal and
-  actual arguments. A non-empty spine cut against another non-empty
-  spine $\gcut {\app H S} {S'}$ blocks the substitution instead of
-  reducing to $\app H {(\gconcat S S')}$ (an implicit
-  $\eta$-expansion); conversely, exhausting actual but not formal
-  arguments as in $\gcut{\lam x M}{\gnil}$ blocks the substitution
-  too. For example, the reduction $\gcut {\lam x \app {\var x}{\var
-      x}} {\lam x \app {\var x}{\var x}}$ gets stuck because the
-  second $\var x$ in the left part is substituted by a function yet is
-  not itself a function. }: it is mutually recursive with a \emph{cut}
-function $\gcut M S$ which can in turn trigger a chain of
-substitutions, leading to a canonical objects or an error.
+  present in \cite{pfenning2007term}, differs from the more standard
+  definition in \cite{hl07mechanizing} as it presupposes $\eta$-long
+  normal forms: once a \emph{cut} starts, there must be exactly as
+  many formal and actual arguments. A non-empty spine cut against
+  another non-empty spine $\gcut {\app H S} {S'}$ blocks the
+  substitution instead of reducing to $\app H {(\gconcat S S')}$ (an
+  implicit $\eta$-expansion); conversely, exhausting actual but not
+  formal arguments as in $\gcut{\lam x M}{\gnil}$ blocks the
+  substitution too. For example, the reduction $\gcut {\lam x \app
+    {\var x}{\var x}} {\lam x \app {\var x}{\var x}}$ gets stuck
+  because the second $\var x$ in the left part is substituted by a
+  function yet is not itself a function. }: it is mutually recursive
+with a \emph{cut} function $\gcut M S$ which can in turn trigger a
+chain of substitutions, leading to a canonical objects or an error.
 \begin{align*}

### 2012-06-08 doc/papers/LFMTP2012/paper.tex
paper: fix & improve system rules
* only thread Delta and not Repo
* add syntactic cat. of atomic types, strengthen rules
* repo contains atomic terms

@@
   \prd x A A \gor
+  P &
+  \text{Type family} \\
+  P &\gequal
   \app {\cst a} S &
-  \text{Type family} \\
+  \text{Atomic type} \\
   M &\gequal
@@
 A \emph{repository} $\mr$ is a finite map from metavariable to
-judgements of object well-typing, along with a distinguished
+judgments of atomic object well-typing, along with a distinguished
 metavariable, corresponding to the \emph{tip} of the term contained in
 it:
-$$ \mr\;:\;(\meta X \mapsto (\Gamma\vdash\tp M A)), \smeta X
+$$ \mr\;:\;(\meta X \mapsto (\Gamma\vdash\tp F P)), \smeta X
 \sigma $$
@@
 
-  \fbox{$\jm\mr\Gamma M A \mmr {M'}$}
+  \fbox{$\jm\Delta\Gamma M A {\Delta'} {M'}$}
   \qquad
@@
   \begin{mathpar}
-    % Lambda
-    \infer{
-      \jm\mr{\envcons\Gamma x A} M {A'} \mmr {M'}
-    }{
-      \jm\mr\Gamma {\lam x M} {\prd x A {A'}} \mmr {\lam x M'}
-    }
-
-    % App
-    \infer{
-      \jf\mr\Gamma F \mmr {F'} {A'}
-       \and
-      \jea\mmr\Gamma {\app{\cst a} S} {A'}
-    }{
-      \jm\mr\Gamma F {\app{\cst a} S} \mmr {F'}
+    \infer[MLam]{
+      \jm\Delta{\envcons\Gamma x A} M {A'} {\Delta'} {M'}
+    }{
+      \jm\Delta\Gamma {\lam x M} {\prd x A {A'}} {\Delta'} {\lam x M'}
+    }
+
+    \infer[MAtom]{
+      \jf\Delta\Gamma F {\Delta'} {F'} {P}
+       \and
+      \jea{\Delta'}\Gamma P {\app{\cst a} S}
+    }{
+      \jm\Delta\Gamma F {\app{\cst a} S} {\Delta'} {F'}
     }
@@
 
-  \fbox{$\jf\mr\Gamma F \mmr F A$}
+  \fbox{$\jf\Delta\Gamma F {\Delta'} {F'} P$}
   \qquad
@@
     % Non-sliceable
-    \infer{
-      \jh\mr\Gamma H A \annnonsliceable
-      \and
-      \jl\mr\Gamma A S \mmr {S'} {A'}
-    }{
-      \jf\mr\Gamma {\app H S} \mmr {\app H {S'}} {A'}
+    \infer[FAppNoSlice]{
+      \jh\Delta\Gamma H A \annnonsliceable
+      \and
+      \jl\Delta\Gamma A S {\Delta'} {S'} P
+    }{
+      \jf\Delta\Gamma {\app H S} {\Delta'} {\app H {S'}} P
     }
@@
     % Meta
-    \infer{
-      \mr(\meta X) = (\Gamma'\vdash M:A)
-      \and
-      \js\mr\Gamma \sigma {\Gamma'} \mmr {\sigma'}
-    }{
-      \jf\mr\Gamma {\smeta X \sigma} \mmr {\smeta X {\sigma'}} {\gsubst{A}\sigma'}
+    \infer[FMeta]{
+      \Delta(\meta X) = (\Gamma'\vdash F:P)
+      \and
+      \js\Delta\Gamma \sigma {\Gamma'} {\Delta'} {\sigma'}
+    }{
+      \jf\Delta\Gamma {\smeta X \sigma} {\Delta'} {\smeta X {\sigma'}} {\gsubst{P}{\sigma'}}
     }
@@
     % Sliceable
-    \infer{
-      \jh\mr\Gamma H A \annsliceable
-      \and
-      \jl\mr\Gamma A S \mmr {S'} {A'}
+    \infer[FAppSlice]{
+      \jh\Delta\Gamma H A \annsliceable
+      \and
+      \jl\Delta\Gamma A S {\Delta'} {S'} {P}
       \and
@@
     }{
-      \jf\mr\Gamma {\app H S} {\mapadd\mmr{\meta X}{(\Gamma'\vdash
-          {\app H {S'}}:A)}} {\smeta X{\msubstid{\Gamma'}}} {A'}
-    }
-    \quad\text{($\meta X$ fresh in $\mr$)}
+      \jf\Delta\Gamma {\app H S} {\mapadd{\Delta'}{\meta X}{(\Gamma'\vdash
+          {\app H {S'}}:P)}} {\smeta X{\msubstid{\Gamma'}}} P
+    }
+    \quad\text{($\meta X$ fresh in $\Delta'$)}
 
     % Defined
-    \infer{
-      \jh\mr\Gamma H A {\anndefined T}
-      \and
-      \jl\mr\Gamma A S {\mmr} {S'} {A'}
-      \and
-      \jcc\mr\Gamma A T S \mmmr F
-      \and
-      \jm\mmmr\Gamma F {A'} {\repo {R'''} } {F'}
-    }{
-      \jf\mr\Gamma {\app H S} {\repo{R'''}} {F'} {A'}
+    \infer[FAppEval]{
+      \jh\Delta\Gamma H A {\anndefined T}
+      \and
+      % check args pas nécessaire (fait ds la red)
+      % \jl\Delta\Gamma A S {{\Delta'}} {S'} P
+      % \and
+      \jcc\Delta\Gamma A T S {\Delta''} F
+      % check result pas nécessaire (fait ds la red)
+      % \and
+      % \jm{\Delta''}\Gamma F P {\Delta'''} {F'}
+    }{
+      \jf\Delta\Gamma {\app H S} {\Delta'''} {F'} P
     }
@@
 
-  \fbox{$\jh\mr\Gamma H A L$}
+  \fbox{$\jh\Delta\Gamma H A L$}
   \qquad
@@
     % constant
-    \infer{
+    \infer[HConst]{
       \Sigma(\cst c) = A; L
     }{
-      \jh\mr\Gamma {\cst c} A L
+      \jh\Delta\Gamma {\cst c} A L
     }
@@
     % variable
-    \infer{
+    \infer[HVar]{
       \Gamma(\var x) = A
     }{
-      \jh\mr\Gamma {\var x} A \annnonsliceable
+      \jh\Delta\Gamma {\var x} A \annnonsliceable
     }
@@
 
-  \fbox{$\jl\mr\Gamma A S \mmr{S'}{A'}$}
+  \fbox{$\jl\Delta\Gamma A S {\Delta'}{S'} P$}
   \qquad
@@
   \begin{mathpar}
-    \infer{
-      \jm\mr\Gamma M A \mmr {M'}
-      \and
-      \jl\mmr\Gamma {\gsubst{A'}{M'}} S \mmmr {S'} {A''}
-    }{
-      \jl\mr\Gamma {\prd x {A} {A'}} {\spinecons M S} \mmmr {\spinecons
-      {M'}{S'}} {A''}
-    }
-
-    \infer{ }{
-      \jl\mr\Gamma {\app{\cst a} S} \spinenil \mr \spinenil {{\app{\cst a} S}}
+    \infer[SCons]{
+      \jm\Delta\Gamma M A {\Delta'} {M'}
+      \and
+      \jl{\Delta'}\Gamma {\gsubst{A'}{\msubst x {M'}}} S {\Delta''} {S'} P
+    }{
+      \jl\Delta\Gamma {\prd x {A} {A'}} {\spinecons M S} {\Delta''} {\spinecons
+      {M'}{S'}} P
+    }
+
+    \infer[SNil]{ }{
+      \jl\Delta\Gamma {\app{\cst a} S} \spinenil \Delta \spinenil {{\app{\cst a} S}}
     }
@@
 
-  \fbox{$\js\mr\Gamma\sigma{\Gamma'}\mmr{\sigma'}$}
+  \fbox{$\js\Delta\Gamma\sigma{\Gamma'}{\Delta'}{\sigma'}$}
   \qquad
@@
     % Cons
-    \infer{
-      \js\mr\Gamma\sigma{\Gamma'}\mmr{\sigma'}
-      \and
-      \jm\mmr\Gamma M {\gsubst A {\sigma'}} \mmmr {M'}
-    }{
-      \js\mr\Gamma{(\msubstcons \sigma x M)} {(\envcons {\Gamma'} x {A'})}
-      \mmmr {(\msubstcons {\sigma'} x {M'})}
+    \infer[$\sigma$Cons]{
+      \js\Delta\Gamma\sigma{\Gamma'}{\Delta'}{\sigma'}
+      \and
+      \jm{\Delta'}\Gamma M {\gsubst A {\sigma'}} {\Delta''} {M'}
+    }{
+      \js\Delta\Gamma{(\msubstcons \sigma x M)} {(\envcons {\Gamma'} x {A'})}
+      {\Delta''} {(\msubstcons {\sigma'} x {M'})}
     }
@@
     % Nil
-    \infer{ }{
-      \js\mr\Gamma\msubstnil\envnil\mr\envnil
+    \infer[$\sigma$Nil]{ }{
+      \js\Delta\Gamma\msubstnil\envnil\Delta\envnil
     }
@@
 
-The main judgement is atomic object typing $\jf\mr\Gamma F \mmr {F'}
-A$, which means that in repository $\mr$ and environment $\Gamma$, the
+The main judgement is atomic object typing $\jf\Delta\Gamma F
+{\Delta'} {F'} P$, which means that in repository $\mr$ and environment $\Gamma$, the
 applicative term $F$ is well-typed of type $A$, and the slicing of $F$
@@
 \begin{figure*}
-  \fbox{$\jsig\mr\Sigma\mmr{\Sigma'}$}
+  \fbox{$\jsig\Delta\Sigma{\Delta'}{\Sigma'}$}
   \qquad
@@
     \infer{
-      \jsig\mr\Sigma\mmr{\Sigma'}
-      \and
-      \jfam\mmr\gnil {\!_{\Sigma'} A}\mmmr {A'}
-    }{
-      \jsig\mr{\gcons\Sigma{{\cst c :^L A}}}\mmmr{\gcons{\Sigma'}{{\cst c :^L A'}}}
+      \jsig\Delta\Sigma{\Delta'}{\Sigma'}
+      \and
+      \jfam{\Delta'}\gnil {\!_{\Sigma'} A}{\Delta''} {A'}
+    }{
+      \jsig\Delta{\gcons\Sigma{{\cst c :^L A}}}{\Delta''}{\gcons{\Sigma'}{{\cst c :^L A'}}}
     }
@@
     \infer{
-      \jsig\mmr\Sigma\mmmr{\Sigma'}
-      \and
-      \jkind\mr\gnil {\!_{\Sigma'} K}\mmr {K'}
-    }{
-      \jsig\mr{\gcons\Sigma{{\cst a : K}}}\mmmr{\gcons{\Sigma'}{{\cst a : K'}}}
+      \jsig{\Delta'}\Sigma{\Delta''}{\Sigma'}
+      \and
+      \jkind\Delta\gnil {\!_{\Sigma'} K}{\Delta'} {K'}
+    }{
+      \jsig\Delta{\gcons\Sigma{{\cst a : K}}}{\Delta''}{\gcons{\Sigma'}{{\cst a : K'}}}
     }
@@
     \infer{ }{
-      \jsig\mr\gnil\mr\gnil
+      \jsig\Delta\gnil\Delta\gnil
     }
@@
 
-  \fbox{$\jfam\mr\Gamma A \mmr {A'}$}
+  \fbox{$\jfam\Delta\Gamma A {\Delta'} {A'}$}
   \qquad
@@
     \infer{
-      \jfam\mr\Gamma {A_1} \mmr {A_1'}
-      \and
-      \jfam\mmr{\gcons\Gamma{\tp {\var x} {A_1'}}} {A_2} \mmmr {A_2'}
-    }{
-      \jfam\mr\Gamma {\prd x {A_1} {A_2}} \mmmr {\prd x {A_1'} {A_2'}}
+      \jfam\Delta\Gamma {A_1} {\Delta'} {A_1'}
+      \and
+      \jfam{\Delta'}{\gcons\Gamma{\tp {\var x} {A_1'}}} {A_2} {\Delta''} {A_2'}
+    }{
+      \jfam\Delta\Gamma {\prd x {A_1} {A_2}} {\Delta''} {\prd x {A_1'} {A_2'}}
     }
@@
       \and
-      \jfl\mr\Gamma K S \mmr {S'}
-    }{
-      \jfam\mr\Gamma {\app{\cst a} S} \mmr {\app{\cst a} {S'}}
+      \jfl\Delta\Gamma K S {\Delta'} {S'}
+    }{
+      \jfam\Delta\Gamma {\app{\cst a} S} {\Delta'} {\app{\cst a} {S'}}
     }
@@
 
-  \fbox{$\jfl\mr\Gamma K S \mmr {S'}$}
+  \fbox{$\jfl\Delta\Gamma K S {\Delta'} {S'}$}
   \qquad
@@
     \infer{
-      \jm\mr\Gamma M A \mmr {M'}
-      \and
-      \jfl\mmr\Gamma {\gsubst{K}{M'}} S \mmmr {S'}
-    }{
-      \jfl\mr\Gamma {\prd x {A} {K}} {\spinecons M S} \mmmr {\spinecons
+      \jm\Delta\Gamma M A {\Delta'} {M'}
+      \and
+      \jfl{\Delta'}\Gamma {\gsubst{K}{\msubst x {M'}}} S {\Delta''} {S'}
+    }{
+      \jfl\Delta\Gamma {\prd x {A} {K}} {\spinecons M S} {\Delta''} {\spinecons
       {M'}{S'}}
@@
     \infer{ }{
-      \jfl\mr\Gamma\type\spinenil\mr\spinenil
+      \jfl\Delta\Gamma\type\spinenil\Delta\spinenil
     }
@@
 
-  \fbox{$\jkind\mr\Gamma K \mmr {K'}$}
+  \fbox{$\jkind\Delta\Gamma K {\Delta'} {K'}$}
   \qquad
@@
     \infer{
-      \jfam\mr\Gamma A\mmr{A'}
-      \and
-      \jkind\mmr{\envcons\Gamma x A} K \mmmr {K'}
-    }{
-      \jkind\mr\Gamma{\prd x A K}\mmmr{\prd x {A'}{K'}}
-    }
-
-    \infer{ }{\jkind\mr\Gamma\type\mr\type}
+      \jfam\Delta\Gamma A{\Delta'}{A'}
+      \and
+      \jkind{\Delta'}{\envcons\Gamma x A} K {\Delta''} {K'}
+    }{
+      \jkind\Delta\Gamma{\prd x A K}{\Delta''}{\prd x {A'}{K'}}
+    }
+
+    \infer{ }{\jkind\Delta\Gamma\type\Delta\type}
   \end{mathpar}

### 2012-06-08 doc/papers/LFMTP2012/paper.tex
paper: abstract

@@
 
-\title{Incremental tactics}
+\title{Certifying, incremental type checking}
 
 \author{Matthias Puech\inst{1,2} \and Yann R\'egis-Gianas\inst{1}}
-\institute{
-  PPS, Team $\pi r^2$ (Universit\'e Paris Diderot, CNRS, INRIA)
-  \and
-  Department of Computer Science, University of Bologna
-}
+\institute{Univ Paris Diderot, Sorbonne Paris Cit\'e, PPS, UMR 7126
+  CNRS, $\pi r^2$, INRIA Paris-Rocquencourt, F-75205 Paris, France
+  \and Department of Computer Science, University of Bologna, 40126
+  Bologna, Italy}
 
@@
  %% State the problem
+  Type checkers are traditionally implemented as batch processes,
+  taking a program and returning a binary information: well-typed or
+  not.
  %% Say why it’s an interesting problem
+  This laconicism presents a double problem: first, how can I trust
+  that the checking algorithm (a complex program) is sound \wrt\ the
+  declarative type system? Secondly, how can I avoid costly rechecks
+  of the whole program when I only make a small, localized change?
  %% Say what your solution achieves
+  A lightweight solution to both problems is letting the checker
+  return a typing derivation: trust is then ensured by checking that
+  this derivation is well-formed, and incrementality by the
+  possibility of reusing already-computed pieces of derivations.
  %% Say what follows from your solution
+  We present a system turning a type system and its checking algorithm
+  into a safe and incremental one, consisting of a generic data
+  structure to represent reusable derivations based on \LF\ and a
+  general-purpose programming language to generate them.
 \end{abstract}
@@
     {\var y} u v {\s{\var u}}}
-  \\
-  &\quad\app{\var{add}}\app{(\s\z)}{(\s\z)} \ \text{.}
+  \app{\var{add}}\app{(\s\z)}{(\s\z)} \ \text{.}
 \end{align*}
@@
       {\var y} u v {\fget{\gsubst{\md_4}{\msubst{H}{\finfer{\var u}}}}}}
-    \\&\qquad
     \app{\var{add}}\app{\fget{\md_2}}{\fget{\md_2}}
@@
 recursive argument $\var u$. $\md_2$ derives $\vdash \s\z :
-\cst{odd}$, but $\var{add}$ awaits $\cst{nat}$ arguments. When the
-system will generate this derivation, it will insert a coercion
-$\cst{odd} \leq \cst{nat}$ at these two points. Taking advantage of
-$\md_2$ and $\md_4$ the system succeeds with a new derivation made out
-of these, without having to regenerate them because it knows that
+\cst{odd}$, but $\var{add}$ awaits two $\cst{nat}$ arguments. When the
+system generates this derivation, it inserts a coercion $\cst{odd}
+\leq \cst{nat}$ at these two points. Taking advantage of $\md_2$ and
+$\md_4$ the system succeeds with a new derivation made out of these,
+without having to regenerate them because it knows that
 $\finfer{\fget\md} = \md$.
@@
   \\&\quad
-  {\tlam x {\cst{nat}} \tlam y
+  {(\tlam x {\cst{nat}} \tlam y
     {\cst{nat}} \recb \z {\var y} z t {
-      \app{\var{add}}\app{\var x} {\var z}}}
+      \app{\var{add}}\app{\var x} {\var z}})}
   \\&\quad\quad
@@
 
-We exploit here an idea similar to run-time
-\emph{contracts}: %TODO cite
-there is no static guarantee that computations won't return ill-typed
-values, but arguments and results are checked at run-time at the
-boundaries of defined functions. This allows to omit formal
-justifications, like the fact that we need to iterate twice the typing
-of $P$ in $\recb M N x y P$, or that $B$ can't depend on $\md_x$ in
-the $\lambda$ case. Of course this incurs on the performance; we will
-see a compensation for it by the slicing mechanism described below in
-\ref{sec:repr}.
+We exploit here an idea similar to run-time \emph{contracts}
+\cite{wadler2009well}: there is no static guarantee that computations
+won't return ill-typed values, but arguments and results are checked
+at run-time at the boundaries of defined functions. This allows to
+omit formal justifications, like the fact that we need to iterate
+twice the typing of $P$ in $\recb M N x y P$, or that $B$ can't depend
+on $\md_x$ in the $\lambda$ case. Of course this incurs on the
+performance; we will see a compensation for it by the slicing
+mechanism described below in \ref{sec:repr}.
 
@@
 These typing rules use definitional equality judgement between two
-families $\jea\mr\Gamma A {A'}$. This equivalence is usually defined
+families $\jea\Delta\Gamma A {A'}$. This equivalence is usually defined
 to be just $\alpha$-equivalence since objects are in
@@
   \lam x T \gor
-  U \\
+  U & \text{Term} \\
   U &\gequal
   F \gor
-  \matchin U \Gamma C \\
+  \matchin U \Gamma C & \text{Atomic term}\\
   C &\gequal \casebnobar P U \gor
-  C\ \caseb P U \\
+  C\ \caseb P U & \text{Branches}\\
   P &\gequal
-  \app H {\var x} \ldots\ {\var x}
+  \app H {\var x} \ldots\ {\var x} &\text{Pattern}
 \end{align*}
@@
 \begin{figure*}
-  \fbox{$\jcc\mr\Gamma T S A \mmr F$}
+  \fbox{$\jcc\Delta\Gamma T S A {\Delta'} F$}
   \qquad
-  {Computational cut}
+  {Term evaluation}
 
   \begin{mathpar}
-    \infer{
-      \jm\mr\Gamma M {A_1} \mmr {M'}
-      \and
-      \jcc\mmr\Gamma{\gsubst {A_2} {\msubst x {M'}}}{\gsubst T {\msubst x {M'}}} S \mmmr F
-    }{
-      \jcc\mr\Gamma{\prd x {A_1} {A_2}}{\lam x T}{\spinecons M S}
-      \mmmr F
-    }
-
-    \infer{
-      \jcr\mr\Gamma U \mmr F
-      \and
-      \jm\mmr\Gamma F {\app {\cst a} S} \mmmr {F'}
-    }{
-      \jcc\mr\Gamma {\app {\cst a} S} U \spinenil \mmmr {F'}
+    \infer[TLam]{
+      \jm\Delta\Gamma M {A_1} {\Delta'} {M'}
+      \and
+      \jcc{\Delta'}\Gamma{\gsubst {A_2} {\msubst x {M'}}}{\gsubst T {\msubst x {M'}}} S {\Delta''} F
+    }{
+      \jcc\Delta\Gamma{\prd x {A_1} {A_2}}{\lam x T}{\spinecons M S}
+      {\Delta''} F
+    }
+
+    \infer[TAtom]{
+      \jcr\Delta\Gamma U {\Delta'} F
+      \and
+      \jm{\Delta'}\Gamma F {\app {\cst a} S} {\Delta''} {F'}
+    }{
+      \jcc\Delta\Gamma {\app {\cst a} S} U \spinenil {\Delta''} {F'}
     }
@@
 
-  \fbox{$\jcr\mr\Gamma U \mmr F$}
+  \fbox{$\jcr\Delta\Gamma U {\Delta'} F$}
   \qquad
-  {Computational reduction}
+  {Atomic term evaluation}
 
   \begin{mathpar}
-    \infer{ }{
-      \jcr\mr\Gamma{\app {\var x} S} \mr {\app {\var x} S}
-    }
-
-    \infer{
+    \infer[UVar]{ }{
+      \jcr\Delta\Gamma{\app {\var x} S} \Delta {\app {\var x} S}
+    }
+
+    \infer[UConst]{
       \Sigma(c) = A; (\annsliceable\text{ or }\annnonsliceable)
     }{
-      \jcr\mr\Gamma{\app {\cst c} S} \mr {\app {\cst c} S}
-    }
-
-    \infer{
+      \jcr\Delta\Gamma{\app {\cst c} S} \Delta {\app {\cst c} S}
+    }
+
+    \infer[UEval]{
       \Sigma(c) = A; \anndefined T
       \and
-      \jcc\mr\Gamma T S A \mmr F
-    }{
-      \jcr\mr\Gamma{\app {\cst c} S} \mr {\app {\cst c} S}
-    }
-
-    \infer{
-      \jcr\mr{\gconcat\Gamma{\Gamma'}} U \mmr {\app {H_i} {\spinecons
+      \jcc\Delta\Gamma T S A {\Delta'} F
+    }{
+      \jcr\Delta\Gamma{\app {\cst c} S} \Delta {\app {\cst c} S}
+    }
+
+    \infer[UCase]{
+      \jcr\Delta{\gconcat\Gamma{\Gamma'}} U {\Delta'} {\app {H_i} {\spinecons
           {M_1}{\spinecons\ldots{M_{m_i}}}}}
       \and
-      \jcr\mmr\Gamma {\gsubst{U_i}{\msubst {x$_1$} {M_1} \ldots\
-          \msubst {x$_{m_i}$} {M_{m_i}}}} \mmmr F a
-    }{
-      \jcr\mr\Gamma{\matchin{U}{\Gamma'}{
-          \casebnobar{\app {H_1}{\spinecons {\var x_{11}} {\spinecons
-                \ldots\ {\var x_{1m_1}}}}}{U_1}
+      \jcr{\Delta'}\Gamma {\gsubst{U_i}{\msubst {x$_1$} {M_1} \ldots\
+          \msubst {x$_{m_i}$} {M_{m_i}}}} {\Delta''} F
+    }{
+      \jcr\Delta\Gamma{\matchin{U}{\Gamma'}{
+          (\casebnobar{\app {H_1}{\spinecons {\var x_{11}} {\spinecons
+                \ldots\ {\var x_{1m_1}}}}}{U_1})\
           |\ \ldots\
-          \caseb{\app {H_n}{\spinecons {\var x_{n1}} {\spinecons
-              \ldots\ {\var x_{nm_n}}}}}{U_n}
-        }}\mmr F
+          \caseb{(\app {H_n}{\spinecons {\var x_{n1}} {\spinecons
+              \ldots\ {\var x_{nm_n}}}}}{U_n})
+        }}{\Delta''} F
     }
@@
 
-  \caption{Weak head-reduction}
+  \caption{Evaluation of terms}
   \label{fig:whr}

### 2012-06-11 doc/papers/LFMTP2012/paper.tex
paper

@@
  %% State the problem
-  Type checkers are traditionally implemented as batch processes,
-  taking a program and returning a binary information: well-typed or
-  not.
- %% Say why it’s an interesting problem
-  This laconicism presents a double problem: first, how can I trust
-  that the checking algorithm (a complex program) is sound \wrt\ the
-  declarative type system? Secondly, how can I avoid costly rechecks
-  of the whole program when I only make a small, localized change?
- %% Say what your solution achieves
-  A lightweight solution to both problems is letting the checker
-  return a typing derivation: trust is then ensured by checking that
-  this derivation is well-formed, and incrementality by the
-  possibility of reusing already-computed pieces of derivations.
- %% Say what follows from your solution
-  We present a system turning a type system and its checking algorithm
-  into a safe and incremental one, consisting of a generic data
-  structure to represent reusable derivations based on \LF\ and a
-  general-purpose programming language to generate them.
+
+  A lightweight way to design a trusted type-checker is to let it
+  return a \emph{certificate} of well-typing (for instance
+  \sysname{Agda} and \sysname{GHC} adopt this architecture
+  internally).  Most of the time, these type-checkers are confronted
+  sequentially with very similar pieces of input: the same program,
+  each time slightly modified. Yet they rebuild the entire
+  certificates, even for unchanged parts of the code. Instead of
+  discarding them between runs, we can take advantage of these
+  certificates to improve the reactivity of type-checking, by feeding
+  parts of the certificates for unchanged parts of the original
+  program back into the new input, thus building an incremental
+  type-checking.
+
+  We present a language-independent framework making possible to turn
+  a type system and its certifying checking algorithm into a safe,
+  incremental one. It consists in a model of incrementality by
+  sharing, a generic data structure to represent reusable derivations
+  based on contextual \LF, and a general-purpose language to write
+  certifying programs.
 \end{abstract}
@@
 
-\section{Use case}
+\section{Using the framework}
 %% It’s an unsolved problem (1 page)
 
-\subsection{A backend for incremental type checking}
+\subsection{As a programmer}
 \label{sec:use-incremental}
 
-\begin{figure}
-  \centering
-%  \small
-  \begin{mathpar}
-    \infer{ }{\jts{\cst o} {\even}}
-
-    \infer{\jts M \odd} {\jts {\s M} \even}
-
-    \infer{\jts M \even} {\jts {\s M} \odd}
-
-    \infer{\jts M \nat} {\jts {\s M} \nat}
-
-    \infer{\jts M {A\to B} \and \jts N A} {\jts {\app M N} B}
-
-    \infer{\infer*{}{\mbox{$
-          \begin{array}{c}
-            [\jts {\var x} A] \\[-0.3em]
-            \vdots\\
-            \jts M B
-          \end{array}$
-        }}} {\jts {\lam x M} {A\to B}}
-
-    \infer{\jts M A \and \jsub A B}{\jts M B}
-
-    \infer{\jts M \nat
-      \and
-      \jts N A
-      \and
-      \infer*{}{\mbox{$
-        \begin{array}{c}
-          [\jts {\var x} \nat] \quad
-          [\jts {\var y} A] \\[-0.3em]
-          \vdots \\
-          \jts P A
-        \end{array}
-        $}}
-    }{\jts {\recb M N x y P} A}
-
-    \infer{ }{\jsub A A}
-
-    \infer{ }{\jsub \even\nat}
-
-    \infer{ }{\jsub \odd\nat}
-
-    \infer{\jsub {A'} A \and \jsub B {B'} }{\jsub {A\to B}{A'\to B'}}
-  \end{mathpar}
-  \caption{System \sysname{T_{<:}}, declaratively}
-  \label{fig:tsub}
-\end{figure}
+% \begin{figure}
+%   \centering
+% %  \small
+%   \begin{mathpar}
+%     \infer{ }{\jts{\cst o} {\even}}
+
+%     \infer{\jts M \odd} {\jts {\s M} \even}
+
+%     \infer{\jts M \even} {\jts {\s M} \odd}
+
+%     \infer{\jts M \nat} {\jts {\s M} \nat}
+
+%     \infer{\jts M {A\to B} \and \jts N A} {\jts {\app M N} B}
+
+%     \infer{\infer*{}{\mbox{$
+%           \begin{array}{c}
+%             [\jts {\var x} A] \\[-0.3em]
+%             \vdots\\
+%             \jts M B
+%           \end{array}$
+%         }}} {\jts {\lam x M} {A\to B}}
+
+%     \infer{\jts M A \and \jsub A B}{\jts M B}
+
+%     \infer{\jts M \nat
+%       \and
+%       \jts N A
+%       \and
+%       \infer*{}{\mbox{$
+%         \begin{array}{c}
+%           [\jts {\var x} \nat] \quad
+%           [\jts {\var y} A] \\[-0.3em]
+%           \vdots \\
+%           \jts P A
+%         \end{array}
+%         $}}
+%     }{\jts {\recb M N x y P} A}
+
+%     \infer{ }{\jsub A A}
+
+%     \infer{ }{\jsub \even\nat}
+
+%     \infer{ }{\jsub \odd\nat}
+
+%     \infer{\jsub {A'} A \and \jsub B {B'} }{\jsub {A\to B}{A'\to B'}}
+%   \end{mathpar}
+%   \caption{System \sysname{T_{<:}}, declaratively}
+%   \label{fig:tsub}
+% \end{figure}
 
@@
 two sets $\cst{even}$ and $\cst{odd}$, introduction $\z$ and $\s{M}$,
-elimination ($\recb M N x y P$), and subtyping on those types; see
-Fig. \ref{fig:tsub}. He (or she) wants to write a program computing
-$(2+2)\times 3$. He knows that sometimes those $\cst{rec}$ arguments
-are quite hard to get correct, but \sysname{T_{<:}} is a typed
-language; he wants to take advantage of this to build his program
-incrementally using the teacher's provided incremental verifier. As a
-first attempt, he wants to verify $P_1$:
+elimination ($\recb M N x y P$), and subtyping on those types. He (or
+she) wants to write a program computing $(2+2)\times 3$. He knows that
+sometimes those $\cst{rec}$ arguments are quite hard to get correct,
+but \sysname{T_{<:}} is a typed language; he wants to take advantage
+of this to build his program incrementally using the teacher's
+provided incremental verifier. As a first attempt, he wants to verify
+$P_1$:
 $$
@@
 
-\subsection{A language to write incremental type-checkers}
+\subsection{As a type system designer}
+\label{sec:lang-incr}
 
@@
 
-\section{Here is my idea}
+\section{Designing the framework}
 %% Here is my idea (2 pages)
@@
 the question of the \emph{data structure} to use for these
-certificates. The requirements are:
-\begin{itemize}
-\item conciseness of the certificates, since they are stored or even
-  transferred remotely;
-\item a well-understood, short and independent verification algorithm.
-  This is known as the \emph{de Bruijn principle};
-\item universality with respect to the particular problem: having a
-  certificate language for each domain of application would just shift
-  the problem of trust; having one universally accepted certificate
-  language that can be tailored to many problems puts a much stronger
-  trust on the unique verifier.
-\end{itemize}
-
+certificates. % The requirements are:
+% \begin{itemize}
+% \item conciseness of the certificates, since they are stored or even
+%   transferred remotely;
+% \item a well-understood, short and independent verification algorithm.
+%   This is known as the \emph{de Bruijn principle};
+% \item universality with respect to the particular problem: having a
+%   certificate language for each domain of application would just shift
+%   the problem of trust; having one universally accepted certificate
+%   language that can be tailored to many problems puts a much stronger
+%   trust on the unique verifier.
+% \end{itemize}
 The Edinburgh Logical Framework \cite{harper1993framework}, or \LF\
-for short, is a well-accepted solution to this question. It is a typed
-lambda-calculus and can be viewed as a language for
+for short, is one of the well-accepted solution to this question. It
+is a typed lambda-calculus and can be viewed as a language for
 \emph{representing} terms, derivations and statements of $\Pi_0^1$
@@
 \LF\ syntax.
-
-For instance, the rule for System \sysname{T}'s recursor can be
-encoded by the following constant:
-
-\begin{mathleft}
-  \cst{is\_rec}\ :\
-  \prd{M}{\cst{tm}}\prd{N}{\cst{tm}}\prd{P}{\cst{tm}\to\cst{tm}\to\cst{tm}}
-  \\\quad
-  \prd A {\cst{tp}}
-  \app {\cst{is}}\app M {\cst{nat}} \to \app{\cst{is}}\app N A \to \\\quad
-  \left(
-  \prd x {\cst{tm}} \prd y {\cst{tm}}
-  \app{\cst{is}}\app x \cst{nat} \to \app{\cst{is}}\app y A \to
-  \app{\cst{is}}\app {(\app P \app x y)} A
-  \right)\to\\\quad
-  \app{\cst{is}}\app{(\app{\cst{rec}}\app M \app N (\lam x\lam y \app
-    P\app x y))} A
-\end{mathleft}
-
-This framework has been used to issue certificates for a
-number of practical applications like proof-carrying code
-\cite{necula1997proof}, authentication framework \cite{appel1999proof}
-\ldots %TODO ATP
+% For instance, the rule for System \sysname{T}'s recursor can be
+% encoded by the following constant:
+% \begin{mathleft}
+%   \cst{is\_rec}\ :\
+%   \prd{M}{\cst{tm}}\prd{N}{\cst{tm}}\prd{P}{\cst{tm}\to\cst{tm}\to\cst{tm}}
+%   \\\quad
+%   \prd A {\cst{tp}}
+%   \app {\cst{is}}\app M {\cst{nat}} \to \app{\cst{is}}\app N A \to \\\quad
+%   \left(
+%   \prd x {\cst{tm}} \prd y {\cst{tm}}
+%   \app{\cst{is}}\app x \cst{nat} \to \app{\cst{is}}\app y A \to
+%   \app{\cst{is}}\app {(\app P \app x y)} A
+%   \right)\to\\\quad
+%   \app{\cst{is}}\app{(\app{\cst{rec}}\app M \app N (\lam x\lam y \app
+%     P\app x y))} A
+% \end{mathleft}
+This framework has been used to issue certificates for a number of
+practical applications like proof-carrying
+code~\cite{necula1997proof}, authentication
+framework~\cite{appel1999proof} \ldots %TODO ATP
 
 To write deltas by sharing subterms, we need to be able to address any
-subterm and instantiate its free variables. For this, we define the
-\emph{slicing} of a well-typed \LF\ term: we extend \LF\ with
-contextual metavariables $\meta X$, $\meta Y$\ldots\ as in
-\cite{nanevski2008contextual}. Metavariables stand for well-typed
-terms defined in a repository, and are instantiated wrt. their free
-variables. A repository also contains a distinguished metavariable
-corresponding to the ``tip'' of the term stored in it. For instance, the
-repository
+subterm. Since a subterm can be open, we must instantiate its free
+variables. For this, we define the \emph{slicing} of a well-typed \LF\
+term: we extend \LF\ with contextual metavariables $\meta X$, $\meta
+Y$, \ldots\ as in \cite{nanevski2008contextual}. Metavariables stand
+for well-typed terms defined in a repository, and are instantiated
+wrt. their free variables. A repository also contains a distinguished
+metavariable corresponding to the ``tip'' of the term stored in
+it. For instance, the repository
 $$
@@
 $$
-is one of the possible way of slicing of the term
+is one of the possible way of slicing the term
 $$
@@
 
+Secondly, we need a mean to transform our \LF\ objects as in
+\ref{sec:lang-incr}. For this we introduce \emph{interpreted
+  functions} $\interp f$, $\interp g$, \ldots in the signatures,
+besides constants. Each $\interp f$ comes with its type and its code:
+
 % inverse
@@
 \cite{nanevski2008contextual}. We present its typing algorithm
-relatively to a signature. It features an explicit caching mechanism
-using metavariables.
+relatively to a signature. % It features an explicit caching mechanism
+% using metavariables.
 
@@
     }{
-      \jcr\Delta\Gamma{\app {\cst c} S} \Delta {\app {\cst c} S}
+      \jcr\Delta\Gamma{\app {\cst c} S} {\Delta'} F
     }

### 2012-06-12 doc/papers/LFMTP2012/paper.tex
paper

@@
 \item a generic algorithm to evaluate these type-checkers on program
-  deltas and verify on-the-fly the generated certificates
-  incrementally against a \emph{repository} of already constructed and
-  checked derivations.
+  deltas and verify on-the-fly the generated incremental certificates
+  against a \emph{repository} of already constructed and checked
+  derivations.
 \end{itemize}
@@
 can strip off the annotations, while not forgetting them too soon:
-otherwise we would arrive to the unhandled case $\finfer{\var
-  x}$. This requires evaluating the type-checker with a carefully
-designed strategy (see \ref{sec:comput}).
+otherwise we would hit the unhandled case $\finfer{\var x}$. This
+requires evaluating the program deltas with a carefully designed
+strategy (see \ref{sec:comput}).
 
@@
 Y$, \ldots\ as in \cite{nanevski2008contextual}. Metavariables stand
-for well-typed terms defined in a repository, and are instantiated
-wrt. their free variables. A repository also contains a distinguished
-metavariable corresponding to the ``tip'' of the term stored in
-it. For instance, the repository
+for well-typed terms defined in a \emph{repository}, and are
+instantiated wrt. their free variables. A repository also contains a
+distinguished metavariable corresponding to the ``tip'' of the term
+stored in it.
+
+We thus introduce Sliced \LF\ (\SLF, see
+Fig. \ref{fig:syntax-LF}). Specifically, it is an extension of
+\emph{spine canonical \LF} \cite{pfenning2007term} enriched with
+metavariables which are references to open terms as in
+\cite{nanevski2008contextual}.
+
+\begin{figure}\centering
+  \begin{align*}
+    K &\gequal \prd x A K \gor \type &
+    \text{Kind}\\
+    A &\gequal \prd x A A \gor P &
+    \text{Type family} \\
+    P &\gequal \app {\cst a} S &
+    \text{Atomic type} \\
+    M &\gequal \lam x M \gor F &
+    \text{Canonical object} \\
+    F &\gequal \app H S \gor \smeta X \sigma &
+    \text{Atomic object} \\
+    H &\gequal \var x \gor \cst c &
+    \text{Head}\\
+    S &\gequal \spinenil \gor \spinecons M S &
+    \text{Spine}\\
+    \sigma &\gequal \msubstnil \gor \msubstcons \sigma x M &
+    \text{Parallel substitution} \\
+    \Gamma &\gequal \envnil \gor \envcons \Gamma x A &
+    \text{Environment}
+  \end{align*}
+  \caption{Syntax of \SLF}
+  \label{fig:syntax-LF}
+\end{figure}
+
+In this variant, only $\eta$-long and $\beta$-normal values have an
+existence: canonicity for $\beta$-reduction is enforced in the syntax
+and canonicity for $\eta$-expansion will be enforced by typing,
+following \cite{hl07mechanizing}. Besides, application of a
+\emph{head} (\ie\ a constant or variable) is $n$-ary (\ie\ to a
+\emph{spine} of objects) as opposed to the more standard binary
+application: this eases the definition substitution and
+$\eta$-expansion, and corresponds to a focused sequent calculus known
+as \sysname{LJT} \cite{herbelin1995λ}. The definition of hereditary
+substitution $\gsubst M \sigma$ is skipped and can be found in
+\cite{pfenning2007term} (it is extended to metavariables as in
+\cite{nanevski2008contextual}).
+
+We distinguish syntactically between \emph{checkable} $M$ and
+\emph{inferrable} $F$ objects (resp. \emph{canonical} and
+\emph{atomic}), as in \eg\ \cite{pierce2000local}:
+canonical objects are checked against a given type, whereas we can
+synthesize the type of atomic objects; the coercion from the first
+grammatical category to the second is where we verify equality between
+the synthesized and the checked types.
+
+As is customary, we adopt the Barendregt convention for naming bound
+variables, write $A\to A'$ for $\prd x A A'$ when $\var x\notin
+\FV{A'}$, $\meta X$ for $\smeta X \msubstnil$, and adopt indifferently
+list, map or set notations for $\sigma$ and $\Gamma$. The identity
+substitution $\msubstid{\Gamma}$ is a notation for $\{\msubst x {\var
+  x}\ |\ \var x \in\dom\Gamma\}$. Metavariables $\smeta X \sigma$
+stand for open objects with free variables in $\dom{\sigma}$; To each
+use of them is attached a substitution $\sigma$ defining these free
+variables. We write $\FMV{M}$ for the set of metavariables in $M$ and
+speak of a \emph{metaclosed} object $M$ if $\FMV{M} = \emptyset$.
+
+% \paragraph{Substitution}
+
+% We overload \emph{substitution} as a partial operation on objects,
+% given a parallel substitution $\sigma$. Since the syntax of object is
+% not stable by (textual) substitution, it is defined as
+% \emph{hereditary}\footnote{ Note that this notion of substitution,
+%   present in \cite{pfenning2007term}, differs from the more standard
+%   definition in \cite{hl07mechanizing} as it presupposes $\eta$-long
+%   normal forms: once a \emph{cut} starts, there must be exactly as
+%   many formal and actual arguments. A non-empty spine cut against
+%   another non-empty spine $\gcut {\app H S} {S'}$ blocks the
+%   substitution instead of reducing to $\app H {(\gconcat S S')}$ (an
+%   implicit $\eta$-expansion); conversely, exhausting actual but not
+%   formal arguments as in $\gcut{\lam x M}{\gnil}$ blocks the
+%   substitution too. For example, the reduction $\gcut {\lam x \app
+%     {\var x}{\var x}} {\lam x \app {\var x}{\var x}}$ gets stuck
+%   because the second $\var x$ in the left part is substituted by a
+%   function yet is not itself a function. }: it is mutually recursive
+% with a \emph{cut} function $\gcut M S$ which can in turn trigger a
+% chain of substitutions, leading to a canonical objects or an error.
+% \begin{align*}
+%   \gsubst {(\lam x M)} \sigma &= \lam x {\gsubst M \sigma} &
+%   \text{if $\var x\notin \dom{\sigma}$}
+%   \\
+%   \gsubst {(\app {\cst c} S)} \sigma &= \app {\cst c} {(\gsubst S
+%     \sigma)}
+%   \\
+%   \gsubst {(\app {\var x} S)} \sigma &=
+%   \app {\var x} {\gsubst S \sigma} &
+%   \text{if $\var x\notin \dom{\sigma}$}
+%   \\
+%   \gsubst {(\app {\var x} S)} \sigma &=
+%   \gcut M {\gsubst S \sigma} &
+%   \text{if $\sigma(\var x) = M$}
+%   \\
+%   \gsubst {\smeta X {\sigma}} {\sigma'} &=
+%   \smeta X {\gcomp\sigma{\sigma'}}
+%   \\
+%   \gsubst\spinenil\sigma &=
+%   \spinenil \\
+%   \gsubst{\spinecons M S}\sigma &=
+%   \spinecons{\gsubst M\sigma}{\gsubst S\sigma}
+%   \\
+%   \gcut {\lam x M} {\spinecons N S} &=
+%   \gcut {\gsubst M {\msubst x N}} S
+%   \\
+%   % TODO cas meta/S ? rem: on en a pas besoin ds le proto.
+%   % \gcut {\smeta X\sigma} S &=
+%   % \smeta X{\gcomp\sigma\sigma'}
+%   % \\
+%   \gcut {\app H S} \spinenil &=
+%   \app H S
+%   \\
+%   \gcut {\smeta X\sigma} \spinenil &= \smeta X\sigma
+%   \\
+%   \gcomp \sigma \msubstnil &=
+%   \sigma
+%   \\
+%   \gcomp \sigma {(\msubstcons{\sigma'} x M)} &=
+%   \gcomp {\gsubst\sigma {\msubst x M}} \sigma'
+%   \\
+%   \gsubst {(\msubstcons{\sigma'} x M)} \sigma &=
+%   \msubstcons{\gsubst{\sigma'}\sigma} x {\gsubst M \sigma}
+%   \\
+%   \gsubst {\msubstnil}{\sigma} &= \msubstnil
+% \end{align*}
+
+% We extend this operation trivially on families $\gsubst A \sigma$ and
+% kinds $\gsubst K \sigma$. We also allow to abusively write
+% non-canonical application $\app M S$ as a shorthand for $\gcut M S$
+% if it is defined.
+
+% TODO pourquoi cette def?
+% TODO terminaison?
+% TODO stabilité de la eta?
+
+\begin{definition}
+  A \emph{repository} $\mr$ is a finite map from metavariable to
+  judgments of atomic object well-typing, along with a distinguished
+  metavariable, corresponding to the \emph{tip} of the term contained
+  in it:
+$$ \mr\;:\;(\meta X \mapsto (\Gamma\vdash\tp F P)), \smeta X
+\sigma $$
+\end{definition}
+
+The algorithm will maintain the invariant that a repository $\mr$ is
+\emph{acyclic}, that is for all $\meta X$ in $\mr$, the fixpoint of
+$\FMV{\meta X}$ does not contain $\meta X$. If $\mr = (\Delta,
+X[\sigma])$, we write $\hat {\mr}$ for the tip $X[\sigma]$, abusively
+$\mr$ for $\Delta$ and $\mr (\smeta X \sigma)$ for $\gsubst M\sigma$
+if $\Delta(\meta X) = (\Gamma\vdash M : A)$.
+
+\begin{definition}
+  We define the partial \emph{checkout} operation $\checkout{\mr}$
+  (resp. $\checkoutr{M}$, $\checkoutr{S}$) from a repository
+  (resp. object, spine) to a metaclosed object as unfolding the
+  definitions of all metavariables recursively:
+  \begin{align*}
+    \checkout{\mr} &= \checkoutr{\hat{\mr}} \\
+    \checkoutr{\smeta X \sigma} &= \checkoutr{\mr(\smeta X\sigma)}\\
+    \checkoutr{\lam x M} &= \lam x \checkoutr{M} \\
+    \checkoutr{\app H S} &= \app H {(\checkoutr{S})} \\
+    \checkoutr{\spinecons M S} &= \spinecons {\checkoutr M} {\checkoutr S} \\
+    \checkoutr{\spinenil} &= \spinenil
+  \end{align*}
+\end{definition}
+
+\begin{example}
+  The repository
 $$
@@
 $$
-is one of the possible way of slicing the term
+checks out as the metaclosed atomic object
 $$
@@
 $$
-where all boxed subterms are referable slices.
-
-\subsection{The computational calculus}
+where all referable slices have been boxed.
+\end{example}
+
+\subsection{The computation language}
 \label{sec:comput}
 
-Secondly, we need a mean to transform our \LF\ objects as in
-\ref{sec:lang-incr}. For this we introduce \emph{interpreted
-  functions} $\interp f$, $\interp g$, \ldots in the signatures,
-besides constants. Each $\interp f$ comes with its type and its code:
-
-% inverse
+Secondly, we introduce the \CL\ language to transform our \SLF\
+objects (Fig.~\ref{fig:syntax-CL}). We can construct objects out of
+other, and deconstruct them by case analysis on an applicative
+term. These objects are \SLF\ objects, except that they can contain
+special function symbols $\fct f$ and $\fcti f n$. Hence, we name them
+with underlined metavariables $\cl F$ instead of $F$.
+
+\begin{figure}
+  \centering
+  \begin{align*}
+    T &\gequal \lam x T \gor
+    U & \text{Term} \\
+    U &\gequal {\cl F} \gor
+    \matchin U \Gamma C & \text{Atomic term}\\
+    C &\gequal \casebnobar P U \gor
+    C\ \caseb P U & \text{Branches}\\
+    P &\gequal \app H {\var x}\ \ldots\ {\var x} &\text{Pattern} \\
+    \cl M &\gequal \lam x {\cl M} \gor \cl F &
+    \text{Canonical object} \\
+    \cl F &\gequal \app {\cl H} {\cl S} \gor \smeta X \sigma &
+    \text{Atomic object} \\
+    \cl H &\gequal \var x \gor \cst c \gor \fct f \gor \fcti f n&
+    \text{Head}\\
+    \cl S &\gequal \spinenil \gor \spinecons {\cl M} {\cl S} &
+    \text{Spine}\\
+    \cl\sigma &\gequal \msubstnil \gor \msubstcons {\cl\sigma} x {\cl M} &
+    \text{Parallel substitution} \\
+  \end{align*}
+
+  \caption{Syntax of \CL}
+\label{fig:syntax-CL}
+\end{figure}
+
+This language is dynamically typed: there is no static guarantee that
+object $\app {\fct f} x$ produced by a function $\fct f : A_1\to A_2 =
+T$ will have type $A_2$, nor that pattern matching is exhaustive. It
+is also non-terminating: we can define a recursive function $\fct f :
+\cst a\to \cst a = \lam x \app {\fct f} {\var x}$. Nevertheless, we
+want to maintain the invariant that each time an object goes in or out
+of a function $\fct f : A_1\to A_2$, it is checked against the type of
+this function (resp. $A_1$ or $A_2$). You can think of interpreted
+functions as mere black boxes taking and producing objects, and being
+able to feed objects to other black boxes; yet the wires linking them
+are labeled with types and objects circulating in them are checked
+\wrt\ that type.
+
+Besides constants, the signature defines \emph{interpreted functions}
+$\fct f$, $\fct g$. Each function comes with its type and its
+code.
+
+\begin{definition}
+  A \emph{signature} declares object and type constants and functions
+  to be used in objects.
+  \begin{align*}
+    \Sigma &\gequal
+    \gnil
+    \gor
+    \gcons \Sigma {\cst a : K}
+    \gor
+    \gcons \Sigma {\cst c : A}
+    \gor
+    \gcons \Sigma {\fct f : A = T}
+    &\text{Signature}
+  \end{align*}
+\end{definition}
+
+Again, we use indifferently set, map or list notations for signatures.
+
+\begin{definition}
+  To each function $\fct f : A = T$ in $\Sigma$ we associate a family
+  of \emph{inverses} $\fcti f n : (A)^n = \pi_n(A)$ for $n\in\mathbb
+  N$, where type $(A)^n$ and term $\pi_n(A)$ are defined recursively:
+  \begin{align*}
+    (\prd x {A_1} {A_2})^{n+1} &= \prd x {A_1} (A_2)^n &
+    (\prd x {A_1} {A_2})^0 &= \prd x {A_1} (A_2)^{A_1} \\
+    (\prd x {A_1} {A_2})^A &= \prd x {A_1} (A_2)^A &
+    (\app {\cst a} S)^A &= \prd x {\app{\cst a} S} A \\
+    \pi_{n+1}(\prd x {A_1}{A_2}) &= \lam x \pi_n(A_2) &
+    \pi_0(\prd x {A_1} {A_2}) &= \lam {y} \pi^y(A_n) \\
+    \pi^y(\prd x {A_1}{A_2}) &= \lam x \pi^y (A_2) &
+    \pi^y(\app{\cst a} S) &= \var y
+  \end{align*}
+\end{definition}
+
+\begin{example}
+  Function $\fct{infer} : (\prd M {\cst{tm}} \app{\cst{sig}} M) =
+  T$ (called $\finfer{}$ above, with $\cst{sig}$ the \LF\ encoding of
+  the previous $\Sigma$-type) has one inverse $\fcti{infer}{0} :
+  (\prd M {\cst{tm}} {\app{\cst{sig}} M} \to \cst{tm}) = \lam x \lam y
+  \var x$ (called $\fget{}$ above).
+\end{example}
+
+\begin{example}
+  Function $\fct{equal} : \prd M {\cst{tm}} \prd N {\cst{tm}}
+  \app{\cst{eq}}\app M N = T'$ has two inverses: $\fcti{equal}{0}$ and
+  $\fcti{equal}{1}$ of the same type $\prd M {\cst{tm}} \prd N
+  {\cst{tm}} \app{\cst{eq}}\app M N \to \cst{tm}$ with code
+  respectively $\lam m \lam n \lam h \var m$ and $\lam m \lam n \lam h
+  \var n$.
+\end{example}
 
@@
 
-The value calculus defines the \emph{values} we want to
-manipulate. These contain constants, and --- contrarily to \eg\
-\sysname{ML} values --- binders and variables enabling $\lambda$-tree
-syntax. Thus, they embed a superficial notion of computation:
-hereditary substitution. This calculus is an extension of \emph{spine
-  canonical \LF} \cite{pfenning2007term} enriched with metavariables
-which are references to open terms as in
-\cite{nanevski2008contextual}. We present its typing algorithm
-relatively to a signature. % It features an explicit caching mechanism
-% using metavariables.
-
-\paragraph{Syntax}
-
-In this variant, only $\eta$-long and $\beta$-normal values have an
-existence: canonicity for $\beta$-reduction is enforced in the syntax
-and canonicity for $\eta$-expansion will be enforced by typing,
-following \cite{hl07mechanizing}. Besides, application of a
-\emph{head} (\ie\ a constant or variable) is $n$-ary (\ie\ to a
-\emph{spine} of objects) as opposed to the more standard binary
-application: this eases the definition substitution and
-$\eta$-expansion, and corresponds to a focused sequent calculus known
-as \sysname{LJT} \cite{herbelin1995λ}.
-
-\begin{align*}
-  K &\gequal
-  \prd x A K \gor
-  \type &
-  \text{Kind}\\
-  A &\gequal
-  \prd x A A \gor
-  P &
-  \text{Type family} \\
-  P &\gequal
-  \app {\cst a} S &
-  \text{Atomic type} \\
-  M &\gequal
-  \lam x M \gor
-  F &
-  \text{Canonical object} \\
-  F &\gequal \app H S
-  \gor
-  \smeta X \sigma &
-  \text{Atomic object} \\
-  H &\gequal
-  \var x \gor
-  \cst c &
-  \text{Head}\\
-  S &\gequal
-  \spinenil \gor
-  \spinecons M S &
-  \text{Spine}\\
-  \sigma &\gequal
-  \msubstnil \gor
-  \msubstcons \sigma x M &
-  \text{Parallel substitution} \\
-  \Gamma &\gequal
-  \envnil \gor
-  \envcons \Gamma x A &
-  \text{Environment}
-\end{align*}
-
-As is customary, we adopt the Barendregt convention for naming bound
-variables, write $A\to A'$ for $\prd x A A'$ when $\var x\notin \FV{A'}$,
-$\meta X$ for $\smeta X \msubstnil$, and adopt indifferently list, map
-or set notations for $\sigma$ and $\Gamma$. The identity substitution
-$\msubstid{\Gamma}$ is a notation for $\{\msubst x {\var x}\ |\ \var x
-\in\dom\Gamma\}$.
-
-Metavariables $\smeta X \sigma$ stand for open objects with free
-variables in $\dom{\sigma}$; To each use of them is attached a
-substitution $\sigma$ defining these free variables. They are defined
-in a repository (see below). We write $\FMV{M}$ for the set of
-metavariables in $M$ and speak of a \emph{metaclosed} object $M$ if
-$\FMV{M} = \emptyset$.
-
-We distinguish syntactically between \emph{checkable} $M$ and
-\emph{inferrable} $F$ objects (resp. \emph{canonical} and
-\emph{atomic}), an idea dating back to \cite{pierce2000local}:
-canonical objects are checked against a given type, whereas we can
-synthesize the type of atomic objects; the coercion from the first
-grammatical category to the second is where we verify equality between
-the synthesized and the checked types.
-
-\paragraph{Substitution}
-
-We overload \emph{substitution} as a partial operation on objects,
-given a parallel substitution $\sigma$. Since the syntax of object is
-not stable by (textual) substitution, it is defined as
-\emph{hereditary}\footnote{ Note that this notion of substitution,
-  present in \cite{pfenning2007term}, differs from the more standard
-  definition in \cite{hl07mechanizing} as it presupposes $\eta$-long
-  normal forms: once a \emph{cut} starts, there must be exactly as
-  many formal and actual arguments. A non-empty spine cut against
-  another non-empty spine $\gcut {\app H S} {S'}$ blocks the
-  substitution instead of reducing to $\app H {(\gconcat S S')}$ (an
-  implicit $\eta$-expansion); conversely, exhausting actual but not
-  formal arguments as in $\gcut{\lam x M}{\gnil}$ blocks the
-  substitution too. For example, the reduction $\gcut {\lam x \app
-    {\var x}{\var x}} {\lam x \app {\var x}{\var x}}$ gets stuck
-  because the second $\var x$ in the left part is substituted by a
-  function yet is not itself a function. }: it is mutually recursive
-with a \emph{cut} function $\gcut M S$ which can in turn trigger a
-chain of substitutions, leading to a canonical objects or an error.
-\begin{align*}
-  \gsubst {(\lam x M)} \sigma &= \lam x {\gsubst M \sigma} &
-  \text{if $\var x\notin \dom{\sigma}$}
-  \\
-  \gsubst {(\app {\cst c} S)} \sigma &= \app {\cst c} {(\gsubst S
-    \sigma)}
-  \\
-  \gsubst {(\app {\var x} S)} \sigma &=
-  \app {\var x} {\gsubst S \sigma} &
-  \text{if $\var x\notin \dom{\sigma}$}
-  \\
-  \gsubst {(\app {\var x} S)} \sigma &=
-  \gcut M {\gsubst S \sigma} &
-  \text{if $\sigma(\var x) = M$}
-  \\
-  \gsubst {\smeta X {\sigma}} {\sigma'} &=
-  \smeta X {\gcomp\sigma{\sigma'}}
-  \\
-  \gsubst\spinenil\sigma &=
-  \spinenil \\
-  \gsubst{\spinecons M S}\sigma &=
-  \spinecons{\gsubst M\sigma}{\gsubst S\sigma}
-  \\
-  \gcut {\lam x M} {\spinecons N S} &=
-  \gcut {\gsubst M {\msubst x N}} S
-  \\
-  % TODO cas meta/S ? rem: on en a pas besoin ds le proto.
-  % \gcut {\smeta X\sigma} S &=
-  % \smeta X{\gcomp\sigma\sigma'}
-  % \\
-  \gcut {\app H S} \spinenil &=
-  \app H S
-  \\
-  \gcut {\smeta X\sigma} \spinenil &= \smeta X\sigma
-  \\
-  \gcomp \sigma \msubstnil &=
-  \sigma
-  \\
-  \gcomp \sigma {(\msubstcons{\sigma'} x M)} &=
-  \gcomp {\gsubst\sigma {\msubst x M}} \sigma'
-  \\
-  \gsubst {(\msubstcons{\sigma'} x M)} \sigma &=
-  \msubstcons{\gsubst{\sigma'}\sigma} x {\gsubst M \sigma}
-  \\
-  \gsubst {\msubstnil}{\sigma} &= \msubstnil
-\end{align*}
-
-We extend this operation trivially on families $\gsubst A \sigma$ and
-kinds $\gsubst K \sigma$. We also allow to abusively write
-non-canonical application $\app M S$ as a shorthand for $\gcut M S$
-if it is defined.
-
-% TODO pourquoi cette def?
-% TODO terminaison?
-% TODO stabilité de la eta?
+% The value calculus defines the \emph{values} we want to
+% manipulate. % These contain constants, and --- contrarily to \eg\
+% \sysname{ML} values --- binders and variables enabling $\lambda$-tree
+% syntax. Thus, they embed a superficial notion of computation:
+% hereditary substitution.
 
@@
 
-A \emph{signature} declares object and type constants to be used in
-objects. To each object constant we assign an annotation $L$ setting
-the slicing behaviour of the constant during type checking: either
-sliceable $\annsliceable$ (each application of this constant will be
-cached and given a name), non-sliceable $\annnonsliceable$ (we don't
-create a name for applications of this constant), or defined
-$\anndefined{T}$ (see \ref{sec:computational-calculus}).
-\begin{align*}
-  L &\gequal
-  \annsliceable \gor
-  \annnonsliceable \gor
-  \anndefined{T}
-  \\
-  \Sigma &\gequal
-  \gnil \gor
-  \gcons \Sigma {\cst c :^L A} \gor
-  \gcons \Sigma {\cst a : K}
-  &\text{Signature}
-\end{align*}
-We write $\Sigma(\cst c) = A; L$ for $(\cst c :^L A) \in \Sigma$
-suggesting the implementation of signatures as maps.
-
 Our algorithm accepts exactly the well-typed objects of the usual \LF,
@@
 
-A \emph{repository} $\mr$ is a finite map from metavariable to
-judgments of atomic object well-typing, along with a distinguished
-metavariable, corresponding to the \emph{tip} of the term contained in
-it:
-$$ \mr\;:\;(\meta X \mapsto (\Gamma\vdash\tp F P)), \smeta X
-\sigma $$
-The algorithm will maintain the invariant that a repository
-$\mr$ is \emph{acyclic}, that is for all $\meta X$ in $\mr$, the
-fixpoint of $\FMV{\meta X}$ does not contain $\meta X$. If $\mr =
-(\Delta, X[\sigma])$, we write $\hat {\mr}$ for $X[\sigma]$, abusively
-$\mr$ for $\Delta$ and $\mr (\smeta X \sigma)$ for $\gsubst M\sigma$
-if $\Delta(\meta X) = (\Gamma\vdash M : A)$.
-
-We refer to cached subterms by metavariables. We force instantiation
-of all variables in their environment by attaching a substitution to
-them ($\smeta X \sigma$).
+% We refer to cached subterms by metavariables. We force instantiation
+% of all variables in their environment by attaching a substitution to
+% them ($\smeta X \sigma$).
 
@@
 
-\paragraph{Checkout}
-
-We define the partial \emph{checkout} operation $\checkout{\mr}$
-(resp. $\checkoutr{M}$, $\checkoutr{S}$) from a repository
-(resp. object, spine) to a metaclosed object as unfolding the
-definitions of all metavariables recursively:
-\begin{align*}
-  \checkout{\mr} &= \checkoutr{\hat{\mr}} \\
-  \checkoutr{\smeta X \sigma} &= \checkoutr{\mr(\smeta X\sigma)}\\
-  \checkoutr{\lam x M} &= \lam x \checkoutr{M} \\
-  \checkoutr{\app H S} &= \app H {(\checkoutr{S})} \\
-  \checkoutr{\spinecons M S} &= \spinecons {\checkoutr M} {\checkoutr S} \\
-  \checkoutr{\spinenil} &= \spinenil
-\end{align*}
-
-A repository is said to be \emph{well-formed} if $\checkout {\mr}$
-is defined.
+% A repository is said to be \emph{well-formed} if $\checkout {\mr}$
+% is defined.
 
@@
 code is attached to special, \emph{defined} constants in the signature
-that manipulates \LF\ objects. We have the ability to construct
-objects out of other, and deconstruct them by case analysis:
-\begin{align*}
-  T &\gequal
-  \lam x T \gor
-  U & \text{Term} \\
-  U &\gequal
-  F \gor
-  \matchin U \Gamma C & \text{Atomic term}\\
-  C &\gequal \casebnobar P U \gor
-  C\ \caseb P U & \text{Branches}\\
-  P &\gequal
-  \app H {\var x} \ldots\ {\var x} &\text{Pattern}
-\end{align*}
-
-This language is un(i)typed, in that all \LF\ objects are viewed as
-having the same implicit type $\cst{obj}$: there is no static
-guarantee that terms produced by a constant $\cst c : A_1\to A_2$ will
-have type $A_2$. It is also non-terminating: we could declare a
-recursive constant $\cst c : \cst a\to \cst a = \lam x \app {\cst c}
-{\var x}$. We expose here only the minimal building blocks of this
-language, but really any language with algebraic datatypes would do
-since we will be manipulating directly the first-order representation
-of \LF\ objects. In fact, whatever the language is, you can think of
-defined constants as mere black boxes taking and producing terms, and
-being able to call other black boxes.
-
-Nevertheless, in order to maintain the well-typing of the generated
-term, we want to maintain the invariant that each time a piece of term
-goes in or out of a black box, it is checked against its specification
-(\ie\ its type). We thus maintain \emph{dynamic typing} of
-computations: arguments and results are type-checked at run-time.
-
+that manipulates \LF\ objects.
 \paragraph{Substitution}

### 2012-06-12 doc/papers/LFMTP2012/paper.tex
paper

@@
 
-The positive impact of rich type systems on programming language
-design is not to be demonstrated anymore, both in terms of the safety
-of program execution --- they ``do not go wrong'', and programming
-practices --- modularity, early bug catching. Verifying well-typing of
-real-world programs is thus a critical matter, but it is a complex
+Verifying the well-typing of real-world programs is a critical matter:
+on it relies their safety, and it encourages good programming
+practices: modularity, early bug catching\ldots. But it is a complex
 operation, for the compiler designer and for the human operator:
@@
 missing information has to be algorithmically synthesized, but is also
-true in many cases for what is commonly referred as ``type checking'',
-or explicit ``church-style'' calculi. Take for example dependent type
-systems, relying on a certain notion of definitional equality on
-types, or type systems with subtype polymorphism: both rely on a rule
-of the form
+true in many cases for mere ``type checking'', or explicit
+``church-style'' calculi. Take for example dependent type systems,
+relying on a certain notion of definitional equality on types, or type
+systems with subtype polymorphism: both rely on a rule of the form
 $$
@@
 (for a given notion of $\leq$) which is not syntax-directed. The
-implementer then has to convert the declarative system to an
-algorithm, and convince himself that both are equivalent. This task is
-highly non-trivial and non-modular: for instance, how to infer the
-type of a System \sysname{T} recursor $\recb M N x y P$, in presence
-of subtyping? The problem is that we have to guess a type $A$ more
+implementer has to convert the declarative system to an algorithm, and
+convince himself that both are equivalent. This task is highly
+non-trivial and non-modular: for instance, how to infer the type of a
+System \sysname{T} recursor $\recb M N x y P$, in presence of
+subtyping? The problem is that we have to guess a type $A$ more
 general than that of $N$ and $P$, while giving that type to variable
@@
 
-Let us call $\md_5$ the subderivation generated for $\tlam x
-{\cst{nat}} \tlam y {\cst{nat}} \recb {\var x} {\var y} u v
-{\fget{\gsubst{\md_4}{\msubst H {\finfer{\var u}}}}}$, $\md_6$ for the
-arguments $\app{\fget{\md_2}}{\fget{\md_2}}$ and $\md_7$ for the call
+Let us call $\md_5$ the subderivation generated for the definition of
+$\var{add}$, $\md_6$ for the arguments
+$\app{\fget{\md_2}}{\fget{\md_2}}$ and $\md_7$ for the call
 $\app{\var{add}} {\fget{\md_6}}$. Even if it occurs originally in the
@@
 
-A few days earlier, the teacher is preparing the tool for the students
-to train composing System \sysname{T_{<:}} programs. He (or she)
-provides the abstract syntax, the typing rules and the untrusted
+Earlier, the teacher is preparing the tool for the students. He (or
+she) provides the abstract syntax, the typing rules and the untrusted
 typing algorithm in order to build the incremental type checker. How
-should he write the algorithm? We know that we want to return a
-derivation, and take advantage of a higher-order representation of the
-syntax, so it shouldn't be necessary to thread any explicit
-environment. The definition should then start with:
+should he write the algorithm? We want it to return a derivation, and
+take advantage of a higher-order representation of the syntax, so it
+shouldn't be necessary to thread any explicit environment. The
+definition should then start with:
 
@@
   \quad\caseb{\app{M}N} \\
-  \quad\quad\letd {\pair {A_1\to B} {\md_1}} {\finfer{M}} \\
-  \quad\quad\letd {\pair {A_2} {\md_2}} {\finfer{N}} \\
-  \quad\quad\letd {\md_{\leq}} {\fleq {A_1} {A_2}} \\
+  \quad\quad\letd {\pair {A_1\to B} {\md_1}} {\finfer{M}}
+  \letd {\pair {A_2} {\md_2}} {\finfer{N}}
+  \letd {\md_{\leq}} {\fleq {A_1} {A_2}} \\
   \quad\quad
@@
 \begin{mathleft}
-  \quad\caseb{\tlam x A M} \\
-  \quad\quad
+  \quad\caseb{\tlam x A M}
   \letdin{\pair B \md}{\env {$\md_{\var x}$} {(\vdash {\var x} : A)}}
@@
   \quad\quad
-  \letd {\pair {A_M} {\md_M}} {\finfer{M}} \\
-  \quad\quad\letd {\md_{A_M}} {{A_M} \leq \cst{nat}} \\
-  \quad\quad
+  \letd {\pair {A_M} {\md_M}} {\finfer{M}}
+  \letd {\md_{A_M}} {{A_M} \leq \cst{nat}}
   \letd {\pair {A_N} {\md_N}} {\finfer{N}} \\
@@
         {\fget{\pair{A}{\md_y}}}}}} \syntax{in}\\
-  \hspace{-2em}\pair {A} {
+  \quad\quad\pair {A} {
     \infer{
@@
 their result derivation. Evaluation only need to make sure that in
-this case, the derivation will be returned, \ie\ that it reduces the
-following way:
-$$
-\finfer{\fget{\pair A \md}} = \pair A \md
-$$
-Now $\fget{}$ can be derived automatically from the type of
+this case, the derivation will be returned, \ie\ that equation
+$\finfer{\fget{\pair A \md}} = \pair A \md$ holds.
+
+Note that $\fget{}$ can be derived automatically from the type of
 $\finfer{}$: it is its inverse. It has definition:
 
-% TODO confirmer qu'on met bien la def. des f^0
 \begin{mathleft}
@@
     M:A)} \cst{tm} =
-  \\\qquad\quad \lamd M \lamd A \lamd \md M
+  \lamd M \lamd A \lamd \md M
 \end{mathleft}
@@
 twice the typing of $P$ in $\recb M N x y P$, or that $B$ can't depend
-on $\md_x$ in the $\lambda$ case. Of course this incurs on the
-performance; we will see a compensation for it by the slicing
-mechanism described below in \ref{sec:repr}.
+on $\md_x$ in the $\lambda$ case, while detecting errors early. Of
+course this incurs on the performance; we will see a compensation for
+it by the slicing mechanism described below in \ref{sec:repr}.
 
@@
 so-called \emph{context-free} typing presented in
-\cite[chap. 4]{boespflug2011conception} where a special variable
-construct $\{x:A\}$ denoting free variables is added to the syntax of
-terms, and the rules
+\cite[chap. 4]{boespflug2011conception} where a special term construct
+$\{x:A\}$ denoting free variables is added to the syntax of terms, and
+rules
 $$
@@
 for short, is one of the well-accepted solution to this question. It
-is a typed lambda-calculus and can be viewed as a language for
-\emph{representing} terms, derivations and statements of $\Pi_0^1$
-meta-theorems (but not their proofs) in a given inference system
-supporting hypothetical judgments \cite{pfenning2001logical}. For
-these, it provides built-in support for encoding languages with
-binders and hypothesis (a feature called $\lambda$-tree syntax or HOAS
-\cite{pfenning1988higher}). The user gives a \emph{signature} defining
-abstract syntax and inference rules of his \emph{object language}, and
-gets an algorithm to verify terms and derivations expressed in the
-\LF\ syntax.
+is a dependently typed lambda-calculus for \emph{representing} terms,
+derivations and statements of $\Pi_0^1$ meta-theorems (but not their
+proofs) in a given inference system supporting hypothetical judgments
+\cite{pfenning2001logical}. For these, it provides built-in support
+for encoding languages with binders and hypothesis (a feature called
+$\lambda$-tree syntax or HOAS \cite{pfenning1988higher}). The user
+gives a \emph{signature} defining abstract syntax and inference rules
+of his \emph{object language}, and gets an algorithm to verify terms
+and derivations expressed in the \LF\ syntax.
 % For instance, the rule for System \sysname{T}'s recursor can be
@@
 term: we extend \LF\ with contextual metavariables $\meta X$, $\meta
-Y$, \ldots\ as in \cite{nanevski2008contextual}. Metavariables stand
-for well-typed terms defined in a \emph{repository}, and are
-instantiated wrt. their free variables. A repository also contains a
-distinguished metavariable corresponding to the ``tip'' of the term
-stored in it.
-
-We thus introduce Sliced \LF\ (\SLF, see
-Fig. \ref{fig:syntax-LF}). Specifically, it is an extension of
-\emph{spine canonical \LF} \cite{pfenning2007term} enriched with
-metavariables which are references to open terms as in
+Y$, \ldots\ as in \cite{nanevski2008contextual}. We thus introduce
+Sliced \LF\ (\SLF, see Fig. \ref{fig:syntax-LF}). Specifically, it is
+an extension of \emph{spine canonical \LF} \cite{pfenning2007term}
+enriched with metavariables which are references to open terms as in
 \cite{nanevski2008contextual}.
@@
     \text{Atomic object} \\
-    H &\gequal \var x \gor \cst c &
+    H &\gequal \var x \gor \cst c \gor \fct{f} &
     \text{Head}\\
@@
 
-In this variant, only $\eta$-long and $\beta$-normal values have an
+In this variant, only $\eta$-long and $\beta$-normal objects have an
 existence: canonicity for $\beta$-reduction is enforced in the syntax
@@
 following \cite{hl07mechanizing}. Besides, application of a
-\emph{head} (\ie\ a constant or variable) is $n$-ary (\ie\ to a
-\emph{spine} of objects) as opposed to the more standard binary
-application: this eases the definition substitution and
+\emph{head} (\ie\ a variable, constant, or function symbol) is $n$-ary
+(\ie\ to a \emph{spine} of objects) as opposed to the more standard
+binary application: this eases the definition substitution and
 $\eta$-expansion, and corresponds to a focused sequent calculus known
-as \sysname{LJT} \cite{herbelin1995λ}. The definition of hereditary
+as \sysname{LJT} \cite{herbelin1995λ}. We differ the discussion of
+function symbols $\fct f$ to \ref{sec:comput}. A \emph{value} is an
+object containing no function symbols. The definition of hereditary
 substitution $\gsubst M \sigma$ is skipped and can be found in
-\cite{pfenning2007term} (it is extended to metavariables as in
+\cite{pfenning2007term} (and extended to metavariables as
 \cite{nanevski2008contextual}).
@@
 \emph{inferrable} $F$ objects (resp. \emph{canonical} and
-\emph{atomic}), as in \eg\ \cite{pierce2000local}:
-canonical objects are checked against a given type, whereas we can
-synthesize the type of atomic objects; the coercion from the first
-grammatical category to the second is where we verify equality between
-the synthesized and the checked types.
+\emph{atomic}), as in \eg\ \cite{pierce2000local}: canonical objects
+are checked against a given type, whereas we can synthesize the type
+of atomic objects; the coercion from the first grammatical category to
+the second is where we verify equality between synthesized and checked
+types.
 
 As is customary, we adopt the Barendregt convention for naming bound
-variables, write $A\to A'$ for $\prd x A A'$ when $\var x\notin
-\FV{A'}$, $\meta X$ for $\smeta X \msubstnil$, and adopt indifferently
+variables, write $A_1\to A_2$ for $\prd x {A_1} {A_2}$ when $\var x\notin
+\FV{A_2}$, $\meta X$ for $\smeta X \msubstnil$, and adopt indifferently
 list, map or set notations for $\sigma$ and $\Gamma$. The identity
@@
   A \emph{repository} $\mr$ is a finite map from metavariable to
-  judgments of atomic object well-typing, along with a distinguished
-  metavariable, corresponding to the \emph{tip} of the term contained
-  in it:
-$$ \mr\;:\;(\meta X \mapsto (\Gamma\vdash\tp F P)), \smeta X
+  triplets of environments, atomic objects and atomic types (\ie\
+  judgments of well-typing), along with a distinguished metavariable,
+  corresponding to the \emph{tip} of the term contained in it:
+$$ \mr\;:\;(\meta X \mapsto (\Gamma\vdash\tp F P))\times\smeta X
 \sigma $$
@@
 
-The algorithm will maintain the invariant that a repository $\mr$ is
+The algorithm in \ref{sec:eval} will maintain the invariant that a repository $\mr$ is
 \emph{acyclic}, that is for all $\meta X$ in $\mr$, the fixpoint of
@@
   \begin{align*}
-    \checkout{\mr} &= \checkoutr{\hat{\mr}} \\
+    \checkout{\mr} &= \checkoutr{\hat{\mr}} &
     \checkoutr{\smeta X \sigma} &= \checkoutr{\mr(\smeta X\sigma)}\\
-    \checkoutr{\lam x M} &= \lam x \checkoutr{M} \\
+    \checkoutr{\lam x M} &= \lam x \checkoutr{M} &
     \checkoutr{\app H S} &= \app H {(\checkoutr{S})} \\
-    \checkoutr{\spinecons M S} &= \spinecons {\checkoutr M} {\checkoutr S} \\
+    \checkoutr{\spinecons M S} &= \spinecons {\checkoutr M} {\checkoutr S} &
     \checkoutr{\spinenil} &= \spinenil
@@
   The repository
-$$
-\small
-\left(\begin{array}{llll}
-    \meta X &\mapsto
-    \envnil&\vdash
-    \app{\cst{lam}} \lam x \smeta Y {\msubst u {\smeta T {\msubst w x}}} &:
+$
+(
+\meta X \mapsto
+\envnil\vdash
+\app{\cst{lam}} \lam x \smeta Y {\msubst u {\smeta T {\msubst w x}}} :
+\cst{tm},
+\meta Y \mapsto
+\env u {\cst{tm}} \vdash
+\app{\cst{lam}} \lam y \smeta Z {\msubstcons {\msubst u
+    {\var u}} v {\var y} } :
 \cst{tm}
-    \\
-    \meta Y &\mapsto
-    \env u {\cst{tm}} &\vdash
-    \app{\cst{lam}} \lam y \smeta Z {\msubstcons {\msubst u
-      {\var u}} v {\var y} } &:
+,
+\meta Z \mapsto
+\envcons {\env u {\cst{tm}}} v {\cst{tm}} \vdash
+\app{\app{\cst{app}}{\var v}}{\var u} :
 \cst{tm}
-    \\
-    \meta Z &\mapsto
-    \envcons {\env u {\cst{tm}}} v {\cst{tm}} &\vdash
-    \app{\app{\cst{app}}{\var v}}{\var u} &:
+,
+\meta T \mapsto
+\env x {\cst{tm}} \vdash
+\app{\cst{app}}\app{\var x}{\cst{o}} :
 \cst{tm}
-    \\
-    \meta T &\mapsto
-    \env x {\cst{tm}} &\vdash
-    \app{\cst{app}}\app{\var x}{\cst{o}} &:
-    \cst{tm}
-
-\end{array}\right), \smeta X {}
-$$
+), \smeta X {}
+$
 checks out as the metaclosed atomic object
@@
 objects (Fig.~\ref{fig:syntax-CL}). We can construct objects out of
-other, and deconstruct them by case analysis on an applicative
-term. These objects are \SLF\ objects, except that they can contain
-special function symbols $\fct f$ and $\fcti f n$. Hence, we name them
-with underlined metavariables $\cl F$ instead of $F$.
+others, and deconstruct them by case analysis on an applicative
+term. The definition $T$ of a function always begins with a series of
+$\lambda$.
 
@@
     U & \text{Term} \\
-    U &\gequal {\cl F} \gor
+    U &\gequal F \gor
     \matchin U \Gamma C & \text{Atomic term}\\
@@
     P &\gequal \app H {\var x}\ \ldots\ {\var x} &\text{Pattern} \\
-    \cl M &\gequal \lam x {\cl M} \gor \cl F &
-    \text{Canonical object} \\
-    \cl F &\gequal \app {\cl H} {\cl S} \gor \smeta X \sigma &
-    \text{Atomic object} \\
-    \cl H &\gequal \var x \gor \cst c \gor \fct f \gor \fcti f n&
-    \text{Head}\\
-    \cl S &\gequal \spinenil \gor \spinecons {\cl M} {\cl S} &
-    \text{Spine}\\
-    \cl\sigma &\gequal \msubstnil \gor \msubstcons {\cl\sigma} x {\cl M} &
-    \text{Parallel substitution} \\
   \end{align*}
@@
 are labeled with types and objects circulating in them are checked
-\wrt\ that type.
+against that type.
 
@@
 \begin{example}
-  Function $\fct{infer} : (\prd M {\cst{tm}} \app{\cst{sig}} M) =
-  T$ (called $\finfer{}$ above, with $\cst{sig}$ the \LF\ encoding of
-  the previous $\Sigma$-type) has one inverse $\fcti{infer}{0} :
-  (\prd M {\cst{tm}} {\app{\cst{sig}} M} \to \cst{tm}) = \lam x \lam y
-  \var x$ (called $\fget{}$ above).
+  \begin{itemize}
+  \item
+    Function $\fct{infer} : (\prd M {\cst{tm}} \app{\cst{sig}} M)
+    = T$ (called $\finfer{}$ above, with $\cst{sig}$ the \LF\ encoding
+    of the previous $\Sigma$-type) has one inverse $\fcti{infer}{0} :
+    (\prd M {\cst{tm}} {\app{\cst{sig}} M} \to \cst{tm}) = \lam x \lam
+    y \var x$ (called $\fget{}$ above).
+  \item
+    Function $\fct{equal} : \prd M {\cst{tm}} \prd N {\cst{tm}}
+    \app{\cst{eq}}\app M N = T'$ has two inverses: $\fcti{equal}{0}$
+    and $\fcti{equal}{1}$ of the same type $\prd M {\cst{tm}} \prd N
+    {\cst{tm}} \app{\cst{eq}}\app M N \to \cst{tm}$ with code
+    respectively $\lam m \lam n \lam h \var m$ and $\lam m \lam n \lam
+    h \var n$.
+  \end{itemize}
 \end{example}
 
-\begin{example}
-  Function $\fct{equal} : \prd M {\cst{tm}} \prd N {\cst{tm}}
-  \app{\cst{eq}}\app M N = T'$ has two inverses: $\fcti{equal}{0}$ and
-  $\fcti{equal}{1}$ of the same type $\prd M {\cst{tm}} \prd N
-  {\cst{tm}} \app{\cst{eq}}\app M N \to \cst{tm}$ with code
-  respectively $\lam m \lam n \lam h \var m$ and $\lam m \lam n \lam h
-  \var n$.
-\end{example}
-
-\section{The calculus}
-%% My idea works (details, data) (5 pages)
-
-\subsection{Value calculus}
-
-% The value calculus defines the \emph{values} we want to
-% manipulate. % These contain constants, and --- contrarily to \eg\
-% \sysname{ML} values --- binders and variables enabling $\lambda$-tree
-% syntax. Thus, they embed a superficial notion of computation:
-% hereditary substitution.
-
-\paragraph{Typing algorithm}
-
-Typing an object while storing intermediate derivations requires
-threading and maintaining a data structure of named partial
-derivations (\emph{slices}) throughout the typing process. We present
-a bidirectional typing algorithm updating an input repository.
-
-Our algorithm accepts exactly the well-typed objects of the usual \LF,
-except that we thread a repository $\mr$ and store in it all
-well-typed applicative subterms $F$ along with their classifier $A$
-and the environment in which they are well-typed.
-
-% TODO parler d'effacer les R et retomber sur LF
-% TODO expl. c'est tout comme LF sauf qqes règles...
-% TODO parler de la redondance en LF normal
-
-% We refer to cached subterms by metavariables. We force instantiation
-% of all variables in their environment by attaching a substitution to
-% them ($\smeta X \sigma$).
-
-When checking large objects, the environment can grow fast. As an
-optimization to reduce the size of the stored environments and
-substitutions in slices $M$, we strengthen environments to their
-minimum size (the free variables of $M$): We define the operation of
-\emph{strengthening} from an environment $\Gamma$ and an object $M$
-well-typed in $\Gamma$ to the minimal environment in which $M$ can
-actually be typed, \ie\ containing only its free variables:
-$$
-\stren{\Gamma,M} = \{x : A\ |\ x\in \FV{M} \wedge \Gamma(x)=A\}
-$$
-% TODO lemme qui dit que c'est correct wrt variables libres des A et
-% du type
-
-The algorithm is presented in Fig. \ref{fig:obj-typing} as a
-syntax-directed inference system. All judgments are of the form $\gjo
-U V W$ where $U$ and $V$ are inputs, $V$ is the induction hypothesis
-and $W$ the output. They are parameterized by a constant and implicit
-signature $\Sigma$, and all premises are ordered in the algorithmic
-order of computation.
+\subsection{The evaluation algorithm}
+\label{sec:eval}
+
+The evaluation algorithm, called $\function{commit$_\mr$}{F}$ takes an
+atomic object $F$ (possibly containing function symbols to evaluate)
+and a repository $\mr$, and produces a new repository $\mr'$ which is
+a strict extension of $\mr$ with a different tip, and which checks out
+to the value of $F$. It has three roles, which are necessarily
+interleaved:
+
+\begin{itemize}
+\item it evaluates $M$ by reducing function symbols,
+\item it type-checks all values fed and produced by functions \wrt\ $\mr$,
+\item it slices those values
+\end{itemize}
+
+This requires threading the map of $\mr$ throughout evaluation and
+typing. We first present the bidirectional typing and slicing part of
+the algorithm (Fig. \ref{fig:obj-typing}) as a syntax-directed
+inference system. All judgments are of the form $\gjo U V W$ where $U$
+and $V$ are inputs, $V$ is the induction hypothesis and $W$ the
+output. They are parameterized by a constant and implicit signature
+$\Sigma$, and all premises are ordered in the algorithmic order of
+computation.
 
@@
     % Non-sliceable
-    \infer[FAppNoSlice]{
-      \jh\Delta\Gamma H A \annnonsliceable
+    \infer[FAppVar]{
+      (x:A)\in\Gamma
       \and
@@
     }{
-      \jf\Delta\Gamma {\app H S} {\Delta'} {\app H {S'}} P
+      \jf\Delta\Gamma {\app {\var x} S} {\Delta'} {\app {\var x} {S'}} P
+    }
+
+    % Sliceable
+    \infer[FAppConst]{
+      (c:A)\in\Sigma
+      \and
+      \jl\Delta\Gamma A S {\Delta'} {S'} {P}
+      \and
+      \stren{\Gamma, \app {\cst c} {S'}}=\Gamma'
+    }{
+      \jf\Delta\Gamma {\app {\cst c} S} {\mapadd{\Delta'}{\meta X}{(\Gamma'\vdash
+          {\app {\cst c} {S'}}:P)}} {\smeta X{\msubstid{\Gamma'}}} P
+    }
+    \quad\text{($\meta X$ fresh in $\Delta'$)}
+
+    % Defined
+    \infer[FAppEval]{
+      (\fct f : A = T) \in \Sigma
+      \and
+      % check args pas nécessaire (fait ds la red)
+      % \jl\Delta\Gamma A S {{\Delta'}} {S'} P
+      % \and
+      \jcc\Delta\Gamma A T S {\Delta'} F P
+      % check result pas nécessaire (fait ds la red)
+      % \and
+      % \jm{\Delta''}\Gamma F P {\Delta'''} {F'}
+    }{
+      \jf\Delta\Gamma {\app {\fct f} S} {\Delta'} {F'} P
     }
@@
 
-    % Sliceable
-    \infer[FAppSlice]{
-      \jh\Delta\Gamma H A \annsliceable
-      \and
-      \jl\Delta\Gamma A S {\Delta'} {S'} {P}
-      \and
-      \stren{\Gamma, \app H {S'}}=\Gamma'
-    }{
-      \jf\Delta\Gamma {\app H S} {\mapadd{\Delta'}{\meta X}{(\Gamma'\vdash
-          {\app H {S'}}:P)}} {\smeta X{\msubstid{\Gamma'}}} P
-    }
-    \quad\text{($\meta X$ fresh in $\Delta'$)}
-
-    % Defined
-    \infer[FAppEval]{
-      \jh\Delta\Gamma H A {\anndefined T}
-      \and
-      % check args pas nécessaire (fait ds la red)
-      % \jl\Delta\Gamma A S {{\Delta'}} {S'} P
-      % \and
-      \jcc\Delta\Gamma A T S {\Delta''} F
-      % check result pas nécessaire (fait ds la red)
-      % \and
-      % \jm{\Delta''}\Gamma F P {\Delta'''} {F'}
-    }{
-      \jf\Delta\Gamma {\app H S} {\Delta'''} {F'} P
-    }
-
-  \end{mathpar}
-
-  \fbox{$\jh\Delta\Gamma H A L$}
-  \qquad
-  {Head}
-
-  \begin{mathpar}
-    % constant
-    \infer[HConst]{
-      \Sigma(\cst c) = A; L
-    }{
-      \jh\Delta\Gamma {\cst c} A L
-    }
-
-    % variable
-    \infer[HVar]{
-      \Gamma(\var x) = A
-    }{
-      \jh\Delta\Gamma {\var x} A \annnonsliceable
-    }
   \end{mathpar}
@@
 
+To reduce the size of the stored environments and substitutions to
+their minimum in slices $\Gamma\vdash F : P$, we use an auxiliary
+operation:
+
+\begin{definition}
+  \emph{Strengthening} takes an environment $\Gamma$ and an object $F$
+  well-typed in $\Gamma$ to an environment $\Gamma'$ containing only
+  the free variables of $F$:
+$$
+\stren{\Gamma,F} = \{x : A\ |\ x\in \FV{F} \wedge \Gamma(x)=A\}
+$$
+\end{definition}
+
+\begin{lemma}
+  If $\jlfm\Gamma F A$, then $\jlfm{\stren{\Gamma, F}} F A$
+\end{lemma}
+
+Our algorithm accepts exactly the well-typed objects of the usual \LF,
+except that we thread a repository $\mr$ and store in it all
+well-typed applicative subterms $F$ along with their classifier $A$
+and the environment in which they are well-typed.
+
+% TODO parler d'effacer les R et retomber sur LF
+% TODO expl. c'est tout comme LF sauf qqes règles...
+% TODO parler de la redondance en LF normal
+
 The main judgement is atomic object typing $\jf\Delta\Gamma F
-{\Delta'} {F'} P$, which means that in repository $\mr$ and environment $\Gamma$, the
-applicative term $F$ is well-typed of type $A$, and the slicing of $F$
-produces a new repository $\mmr$. The object $F'$ is the
-\emph{residual} of this operation: only certain kind of applications
-are sliced, and it is possible that the ``tip'' of application $F$ is
-not sliced (because it is declared as non-sliceable). In that case,
-$M$ is this tip; if $F$ is directly sliceable, then $M$ is the
-metavariable referring to it.
-
-These typing rules use definitional equality judgement between two
-families $\jea\Delta\Gamma A {A'}$. This equivalence is usually defined
-to be just $\alpha$-equivalence since objects are in
-$\beta\eta$-canonical form. Here, objects can contain metavariables
-defined in $\mr$, so we shall thread it. Moreover, we will see in the
-next section how defined constants compute and thus influence
-definitional equality. Definitional equality is defined later in Fig.
-\ref{fig:def-eq}.
-
-Signatures themselves need to be checked. We present the checking
-algorithm in Fig. \ref{fig:sign-typing}.
-
-\begin{figure*}
-  \fbox{$\jsig\Delta\Sigma{\Delta'}{\Sigma'}$}
-  \qquad
-  {Signature}
-
-  \begin{mathpar}
-
-    \infer{
-      \jsig\Delta\Sigma{\Delta'}{\Sigma'}
-      \and
-      \jfam{\Delta'}\gnil {\!_{\Sigma'} A}{\Delta''} {A'}
-    }{
-      \jsig\Delta{\gcons\Sigma{{\cst c :^L A}}}{\Delta''}{\gcons{\Sigma'}{{\cst c :^L A'}}}
-    }
-
-    \infer{
-      \jsig{\Delta'}\Sigma{\Delta''}{\Sigma'}
-      \and
-      \jkind\Delta\gnil {\!_{\Sigma'} K}{\Delta'} {K'}
-    }{
-      \jsig\Delta{\gcons\Sigma{{\cst a : K}}}{\Delta''}{\gcons{\Sigma'}{{\cst a : K'}}}
-    }
-
-    \infer{ }{
-      \jsig\Delta\gnil\Delta\gnil
-    }
-  \end{mathpar}
-
-  \fbox{$\jfam\Delta\Gamma A {\Delta'} {A'}$}
-  \qquad
-  {Family}
-
-  \begin{mathpar}
-
-    \infer{
-      \jfam\Delta\Gamma {A_1} {\Delta'} {A_1'}
-      \and
-      \jfam{\Delta'}{\gcons\Gamma{\tp {\var x} {A_1'}}} {A_2} {\Delta''} {A_2'}
-    }{
-      \jfam\Delta\Gamma {\prd x {A_1} {A_2}} {\Delta''} {\prd x {A_1'} {A_2'}}
-    }
-
-    \infer{
-      \Sigma(a) = K
-      \and
-      \jfl\Delta\Gamma K S {\Delta'} {S'}
-    }{
-      \jfam\Delta\Gamma {\app{\cst a} S} {\Delta'} {\app{\cst a} {S'}}
-    }
-  \end{mathpar}
-
-  \fbox{$\jfl\Delta\Gamma K S {\Delta'} {S'}$}
-  \qquad
-  {Family spine}
-
-  \begin{mathpar}
-    \infer{
-      \jm\Delta\Gamma M A {\Delta'} {M'}
-      \and
-      \jfl{\Delta'}\Gamma {\gsubst{K}{\msubst x {M'}}} S {\Delta''} {S'}
-    }{
-      \jfl\Delta\Gamma {\prd x {A} {K}} {\spinecons M S} {\Delta''} {\spinecons
-      {M'}{S'}}
-    }
-
-    \infer{ }{
-      \jfl\Delta\Gamma\type\spinenil\Delta\spinenil
-    }
-  \end{mathpar}
-
-  \fbox{$\jkind\Delta\Gamma K {\Delta'} {K'}$}
-  \qquad
-  {Kind}
-
-  \begin{mathpar}
-    \infer{
-      \jfam\Delta\Gamma A{\Delta'}{A'}
-      \and
-      \jkind{\Delta'}{\envcons\Gamma x A} K {\Delta''} {K'}
-    }{
-      \jkind\Delta\Gamma{\prd x A K}{\Delta''}{\prd x {A'}{K'}}
-    }
-
-    \infer{ }{\jkind\Delta\Gamma\type\Delta\type}
-  \end{mathpar}
-
-  \caption{Typing algorithm for signatures}
-  \label{fig:sign-typing}
-\end{figure*}
-
-\paragraph{Commit}
-% TODO commit
-
-\paragraph{Example}
-
-Let us work out an example.
-% TODO example
-
-% A repository is said to be \emph{well-formed} if $\checkout {\mr}$
-% is defined.
+{\Delta'} {F'} P$, which reads: in repository $\mr$ and environment
+$\Gamma$, applicative term $F$ is well-typed of type $A$, and the
+slicing of its value produces a new repository $\mmr$. The object $F'$
+is the \emph{residual} of this operation: only constant applications
+$\app{\cst c} S$ are sliced, $F'$ is the ``tip'' of $F$ that was not
+sliced.
+
+These typing rules use definitional equality between two families
+$\jea\Delta\Gamma A {A'}$.
+% TODO expliquer, et commenter reduction des fonctions ou pas,
+% conv. typee ou pas
+
+% Signatures themselves need to be checked. We present the checking
+% algorithm in Fig. \ref{fig:sign-typing}.
+
+% \begin{figure*}
+%   \fbox{$\jsig\Delta\Sigma{\Delta'}{\Sigma'}$}
+%   \qquad
+%   {Signature}
+
+%   \begin{mathpar}
+
+%     \infer{
+%       \jsig\Delta\Sigma{\Delta'}{\Sigma'}
+%       \and
+%       \jfam{\Delta'}\gnil {\!_{\Sigma'} A}{\Delta''} {A'}
+%     }{
+%       \jsig\Delta{\gcons\Sigma{{\cst c :^L A}}}{\Delta''}{\gcons{\Sigma'}{{\cst c :^L A'}}}
+%     }
+
+%     \infer{
+%       \jsig{\Delta'}\Sigma{\Delta''}{\Sigma'}
+%       \and
+%       \jkind\Delta\gnil {\!_{\Sigma'} K}{\Delta'} {K'}
+%     }{
+%       \jsig\Delta{\gcons\Sigma{{\cst a : K}}}{\Delta''}{\gcons{\Sigma'}{{\cst a : K'}}}
+%     }
+
+%     \infer{ }{
+%       \jsig\Delta\gnil\Delta\gnil
+%     }
+%   \end{mathpar}
+
+%   \fbox{$\jfam\Delta\Gamma A {\Delta'} {A'}$}
+%   \qquad
+%   {Family}
+
+%   \begin{mathpar}
+
+%     \infer{
+%       \jfam\Delta\Gamma {A_1} {\Delta'} {A_1'}
+%       \and
+%       \jfam{\Delta'}{\gcons\Gamma{\tp {\var x} {A_1'}}} {A_2} {\Delta''} {A_2'}
+%     }{
+%       \jfam\Delta\Gamma {\prd x {A_1} {A_2}} {\Delta''} {\prd x {A_1'} {A_2'}}
+%     }
+
+%     \infer{
+%       \Sigma(a) = K
+%       \and
+%       \jfl\Delta\Gamma K S {\Delta'} {S'}
+%     }{
+%       \jfam\Delta\Gamma {\app{\cst a} S} {\Delta'} {\app{\cst a} {S'}}
+%     }
+%   \end{mathpar}
+
+%   \fbox{$\jfl\Delta\Gamma K S {\Delta'} {S'}$}
+%   \qquad
+%   {Family spine}
+
+%   \begin{mathpar}
+%     \infer{
+%       \jm\Delta\Gamma M A {\Delta'} {M'}
+%       \and
+%       \jfl{\Delta'}\Gamma {\gsubst{K}{\msubst x {M'}}} S {\Delta''} {S'}
+%     }{
+%       \jfl\Delta\Gamma {\prd x {A} {K}} {\spinecons M S} {\Delta''} {\spinecons
+%       {M'}{S'}}
+%     }
+
+%     \infer{ }{
+%       \jfl\Delta\Gamma\type\spinenil\Delta\spinenil
+%     }
+%   \end{mathpar}
+
+%   \fbox{$\jkind\Delta\Gamma K {\Delta'} {K'}$}
+%   \qquad
+%   {Kind}
+
+%   \begin{mathpar}
+%     \infer{
+%       \jfam\Delta\Gamma A{\Delta'}{A'}
+%       \and
+%       \jkind{\Delta'}{\envcons\Gamma x A} K {\Delta''} {K'}
+%     }{
+%       \jkind\Delta\Gamma{\prd x A K}{\Delta''}{\prd x {A'}{K'}}
+%     }
+
+%     \infer{ }{\jkind\Delta\Gamma\type\Delta\type}
+%   \end{mathpar}
+
+%   \caption{Typing of signatures}
+%   \label{fig:sign-typing}
+% \end{figure*}
+
+\begin{definition}
+  Let $\mr = (\Delta, \smeta X \sigma)$.  If $\jf\Delta\Gamma F
+  {\Delta'} {\smeta Y {\sigma'}} P$, let $\function{commit$_\mr$}{F}$ be
+  $\mmr = (\Delta', \smeta X\sigma)$.
+\end{definition}
 
@@
 
-\subsection{The computation calculus}
-\label{sec:computational-calculus}
-
-We now turn to the definition of our computation calculus. As we saw,
-code is attached to special, \emph{defined} constants in the signature
-that manipulates \LF\ objects.
-\paragraph{Substitution}
-
-\begin{align*}
-  \gsubst {(\lam x T)} \sigma &= \lam x {\gsubst T\sigma} \qquad
-  \text{if $\var x\notin\dom\sigma$}\\
-  \gsubst {(\matchin U \Gamma C)} \sigma &=
-  \matchin {\gsubst
-    U\sigma} {\gsubst\Gamma\sigma} {\gsubst C\sigma} \\
-  \gsubst {(\casebnobar{\app H {\spinecons {\var x_1} {\spinecons \ldots {\var
-          x_n}}}} U)} \sigma &= {\casebnobar{\app H {\spinecons {\var x_1} {\spinecons \ldots {\var
-          x_n}}}} \gsubst U \sigma} \quad \text{if $\var x_i\notin\dom\sigma$} \\
-\end{align*}
-
 \begin{figure*}
-  \fbox{$\jcc\Delta\Gamma T S A {\Delta'} F$}
+  \fbox{$\jcc\Delta\Gamma A T S {\Delta'} F P$}
   \qquad
@@
       \and
-      \jcc{\Delta'}\Gamma{\gsubst {A_2} {\msubst x {M'}}}{\gsubst T {\msubst x {M'}}} S {\Delta''} F
+      \jcc{\Delta'}\Gamma{\gsubst {A_2} {\msubst x {M'}}}{\gsubst T
+        {\msubst x {M'}}} S {\Delta''} F P
     }{
       \jcc\Delta\Gamma{\prd x {A_1} {A_2}}{\lam x T}{\spinecons M S}
-      {\Delta''} F
+      {\Delta''} F P
     }
@@
       \and
-      \jm{\Delta'}\Gamma F {\app {\cst a} S} {\Delta''} {F'}
+      \jm{\Delta'}\Gamma F P {\Delta''} {F'}
     }{
-      \jcc\Delta\Gamma {\app {\cst a} S} U \spinenil {\Delta''} {F'}
+      \jcc\Delta\Gamma P U \spinenil {\Delta''} {F'} P
     }
@@
 
-    \infer[UConst]{
-      \Sigma(c) = A; (\annsliceable\text{ or }\annnonsliceable)
+    \infer[UConst]{ }{
+      \jcr\Delta\Gamma{\app {\cst c} S} \Delta {\app {\cst c} S}
+    }
+
+    % TODO UMeta?
+
+    \infer[UEval]{
+      (\fct f : A = T)\in\Sigma
+      \and
+      \jcc\Delta\Gamma A T S {\Delta'} F P
     }{
-      \jcr\Delta\Gamma{\app {\cst c} S} \Delta {\app {\cst c} S}
-    }
-
-    \infer[UEval]{
-      \Sigma(c) = A; \anndefined T
-      \and
-      \jcc\Delta\Gamma T S A {\Delta'} F
-    }{
-      \jcr\Delta\Gamma{\app {\cst c} S} {\Delta'} F
+      \jcr\Delta\Gamma{\app {\fct f} S} {\Delta'} F
     }
@@
 
-\begin{figure*}
-  \centering
-  TODO
-  \caption{Definitional equality}
-  \label{fig:def-eq}
-\end{figure*}
-
 \section{Related work}

### 2012-06-13 doc/papers/LFMTP2012/paper.tex
paper

@@
 \documentclass{llncs}
-
+\usepackage{ucs}
+\usepackage[utf8x]{inputenc}
+\usepackage[protrusion=true,expansion=true]{microtype}
 \usepackage{ntheorem}
@@
 \allowdisplaybreaks[1]
+\renewcommand{\floatpagefraction}{0.8}
 
@@
   A lightweight way to design a trusted type-checker is to let it
-  return a \emph{certificate} of well-typing (for instance
-  \sysname{Agda} and \sysname{GHC} adopt this architecture
-  internally).  Most of the time, these type-checkers are confronted
-  sequentially with very similar pieces of input: the same program,
-  each time slightly modified. Yet they rebuild the entire
+  return a \emph{certificate} of well-typing, and check it a
+  posteriori (for instance \sysname{Agda} and \sysname{GHC} adopt this
+  architecture internally).  Most of the time, these type-checkers are
+  confronted sequentially with very similar pieces of input: the same
+  program, each time slightly modified. Yet they rebuild the entire
   certificates, even for unchanged parts of the code. Instead of
@@
 ``church-style'' calculi. Take for example dependent type systems,
-relying on a certain notion of definitional equality on types, or type
-systems with subtype polymorphism: both rely on a rule of the form
+relying definitional equality on types, or type systems with subtype
+polymorphism: both rely on a rule of the form
 $$
@@
 (for a given notion of $\leq$) which is not syntax-directed. The
-implementer has to convert the declarative system to an algorithm, and
+implementer has to turn the declarative system into an algorithm, and
 convince himself that both are equivalent. This task is highly
@@
 complete, or even sound, and \emph{(ii)} a trusted and fast kernel
-$\fcheck{\md}$ to verify these derivations \wrt\ a given set of inference
-rules $\Sigma$. This is a \emph{certifying} scheme (see
+$\fcheck{\md}$ to verify these derivations \wrt\ a given set of
+inference rules $\Sigma$. This is a \emph{certifying} scheme (see
 \cite{leroy2006compcert} for a discussion on approaches to reliable
@@
 \emph{is} a derivation $\md$ for $M$, but the converse is not
-necessarily true: function $\finfer{}\!\!$ could fail, or have bugs
-(return an incorrect derivation). Explicitly generating the derivation
-in the declarative system has the following advantages over the first
+necessarily true: function $\finfer{}\!\!$ could fail, or return an
+incorrect derivation. Explicitly generating the derivation in the
+declarative system has the following advantages over the first
 approach:
@@
 into an incremental one
-\cite{pugh1989incremental,acar2003selective}. Unfortunately, they all
+\cite{pugh1989incremental,acar2003selective}. Unfortunately, they
 suffer from their genericity on the particular problem of type
@@
 provided incremental verifier. As a first attempt, he wants to verify
-$P_1$:
-$$
-\recb{\s\z}{\s\z} x y {\s {\var x}}\ \text{.}
-$$
-For this, he sends the command $\finfer{P_1}$ to the system. The
-$\finfer{}$ operator takes a well-formed program and tries to construct
-a valid typing derivation for its argument in the current, empty
-context. The system succeeds and returns a typing derivation $\md_1$
-of \/ $\vdash P_1 : \cst{nat}$, valid in an empty context.
+$P_1 = \recb{\s\z}{\s\z} x y {\s {\var x}}$: He sends the command
+$\finfer{P_1}$ to the system. The $\finfer{}$ operator takes a
+well-formed program and tries to construct a valid typing derivation
+for its argument in the current, empty context. The system succeeds
+and returns a typing derivation $\md_1$ of \/ $\vdash P_1 :
+\cst{nat}$, valid in an empty context.
 
@@
 used to live in the empty context, we are then free to use it in the
-new context where $\var{add}:\cst{nat}\to\cst{nat}$, because System
+new context where $\vdash\var{add}:\cst{nat}\to\cst{nat}$, because System
 \sysname T enjoys the \emph{weakening} property; $\md_4$ depending on
@@
   &\letb{add}{\fget{\md_5}}
-  \\
-  &\quad\letb{mul} { \tlam x {\cst{nat}} \tlam y
+  \letb{mul} { \tlam x {\cst{nat}} \tlam y
     {\cst{nat}} \recb \z {\var y} z t {
@@
 
-Finally, he decides (in an odd move) to inline the definition of
-$\var{mul}$ into its call site. Call $\md_8$ the derivation of the
-definition of $\var{mul}$ (it depends on an $\var H$ of
-$\vdash\var{add}:\cst{nat}\to\cst{nat}$) and $\md_9$ the derivation of
-$(2+2)\times 3$ depending on derivations $\var{H}$ and $\var{H'}$ for $+$ and
-$\times$. He orders:
-\begin{align*}
-\finfer{}(
-\letb{add}{\fget{\md_5}}
-\fget{({\gsubst{\md_9} {\msubst {H} {(\fget{\var{add}})}; \msubst {H'}
-      {\gsubst{\md_8}{\msubst H {(\fget{\var{add}})}}}}})}
-)
-  \ \text{.}
-\end{align*}
-The definition for $\var{add}$ is unchanged, but the instantiation of
-open derivations enables to directly substitute one into the other and
-get the appropriate program: call $\md_{10}$ the whole derivation just
-returned, $\fget{\md_{10}}$ returns:
-\begin{align*}
-  &\letb{add}{\tlam x {\cst{nat}} \tlam y {\cst{nat}} \recb {\var x}
-    {\var y} x y {\s{\var x}}}
-  \\&\quad
-  {(\tlam x {\cst{nat}} \tlam y
-    {\cst{nat}} \recb \z {\var y} z t {
-      \app{\var{add}}\app{\var x} {\var z}})}
-  \\&\quad\quad
-  \app{(\app {\var{add}} \app {(\s{\s\z})}
-    {(\s{\s\z})})}{(\s{\s{\s\z}})}
-  \ \text{.}
-\end{align*}
+% Finally, he decides (in an odd move) to inline the definition of
+% $\var{mul}$ into its call site. Call $\md_8$ the derivation of the
+% definition of $\var{mul}$ (it depends on an $\var H$ of
+% $\vdash\var{add}:\cst{nat}\to\cst{nat}$) and $\md_9$ the derivation of
+% $(2+2)\times 3$ depending on derivations $\var{H}$ and $\var{H'}$ for $+$ and
+% $\times$. He orders:
+% \begin{align*}
+% \finfer{}(
+% \letb{add}{\fget{\md_5}}
+% \fget{({\gsubst{\md_9} {\msubst {H} {(\fget{\var{add}})}; \msubst {H'}
+%       {\gsubst{\md_8}{\msubst H {(\fget{\var{add}})}}}}})}
+% )
+%   \ \text{.}
+% \end{align*}
+% The definition for $\var{add}$ is unchanged, but the instantiation of
+% open derivations enables to directly substitute one into the other and
+% get the appropriate program: call $\md_{10}$ the whole derivation just
+% returned, $\fget{\md_{10}}$ returns:
+% \begin{align*}
+%   &\letb{add}{\tlam x {\cst{nat}} \tlam y {\cst{nat}} \recb {\var x}
+%     {\var y} x y {\s{\var x}}}
+%   \\&\quad
+%   {(\tlam x {\cst{nat}} \tlam y
+%     {\cst{nat}} \recb \z {\var y} z t {
+%       \app{\var{add}}\app{\var x} {\var z}})}
+%   \app{(\app {\var{add}} \app {(\s{\s\z})}
+%     {(\s{\s\z})})}{(\s{\s{\s\z}})}
+%   \ \text{.}
+% \end{align*}
 
@@
 
+% explicit LF
 % lambda-lifting
@@
 is a dependently typed lambda-calculus for \emph{representing} terms,
-derivations and statements of $\Pi_0^1$ meta-theorems (but not their
-proofs) in a given inference system supporting hypothetical judgments
+derivations and statements of $\Pi_0^1$ meta-theorems in a given
+inference system supporting hypothetical judgments
 \cite{pfenning2001logical}. For these, it provides built-in support
@@
 
-\begin{figure}\centering
+\begin{figure}[t]
   \begin{align*}
@@
 $\eta$-expansion, and corresponds to a focused sequent calculus known
-as \sysname{LJT} \cite{herbelin1995λ}. We differ the discussion of
+as \sysname{LJT} \cite{herbelin1995lambda}. We differ the discussion of
 function symbols $\fct f$ to \ref{sec:comput}. A \emph{value} is an
@@
 speak of a \emph{metaclosed} object $M$ if $\FMV{M} = \emptyset$.
+
+% TODO relate to LF
 
@@
 
-The algorithm in \ref{sec:eval} will maintain the invariant that a repository $\mr$ is
-\emph{acyclic}, that is for all $\meta X$ in $\mr$, the fixpoint of
-$\FMV{\meta X}$ does not contain $\meta X$. If $\mr = (\Delta,
-X[\sigma])$, we write $\hat {\mr}$ for the tip $X[\sigma]$, abusively
-$\mr$ for $\Delta$ and $\mr (\smeta X \sigma)$ for $\gsubst M\sigma$
-if $\Delta(\meta X) = (\Gamma\vdash M : A)$.
+If $\mr = (\Delta, X[\sigma])$, we write $\hat {\mr}$ for the tip
+$X[\sigma]$, abusively $\mr$ for $\Delta$ and $\mr (\smeta X \sigma)$
+for $\gsubst M\sigma$ if $\Delta(\meta X) = (\Gamma\vdash M : A)$.
+The algorithm in \ref{sec:eval} will maintain the invariant that a
+repository $\mr$ is \emph{acyclic}, that is for all $\meta X$ in
+$\mr$, the fixpoint of $\FMV{\mr(\meta X)}$ does not contain $\meta X$.
 
@@
 objects (Fig.~\ref{fig:syntax-CL}). We can construct objects out of
-others, and deconstruct them by case analysis on an applicative
-term. The definition $T$ of a function always begins with a series of
+others, and deconstruct them by case analysis. The definition $T$ of a function always begins with a series of
 $\lambda$.
 
-\begin{figure}
-  \centering
+\begin{figure}[t]
   \begin{align*}
@@
     C\ \caseb P U & \text{Branches}\\
-    P &\gequal \app H {\var x}\ \ldots\ {\var x} &\text{Pattern} \\
+    P &\gequal \app H {\var x}\ \ldots\ {\var x} &\text{Pattern}
   \end{align*}
-
   \caption{Syntax of \CL}
@@
 
-Again, we use indifferently set, map or list notations for signatures.
-
 \begin{definition}
   To each function $\fct f : A = T$ in $\Sigma$ we associate a family
-  of \emph{inverses} $\fcti f n : (A)^n = \pi_n(A)$ for $n\in\mathbb
+  of \emph{inverses} $\fcti f {\;n} : (A)^n = \pi_n(A)$ for $n\in\mathbb
   N$, where type $(A)^n$ and term $\pi_n(A)$ are defined recursively:
@@
 \item it evaluates $M$ by reducing function symbols,
-\item it type-checks all values fed and produced by functions \wrt\ $\mr$,
-\item it slices those values
+\item it type-checks all values fed to and produced by functions \wrt\
+  $\mr$,
+\item it slices those values. This requires threading the map of $\mr$
+  throughout evaluation and typing.
 \end{itemize}
 
-This requires threading the map of $\mr$ throughout evaluation and
-typing. We first present the bidirectional typing and slicing part of
-the algorithm (Fig. \ref{fig:obj-typing}) as a syntax-directed
-inference system. All judgments are of the form $\gjo U V W$ where $U$
-and $V$ are inputs, $V$ is the induction hypothesis and $W$ the
-output. They are parameterized by a constant and implicit signature
-$\Sigma$, and all premises are ordered in the algorithmic order of
-computation.
-
-\begin{figure*}
-
+\paragraph{Typing}
+
+We first present the bidirectional typing and slicing part of the
+algorithm (Fig. \ref{fig:obj-typing}) as a syntax-directed inference
+system. All judgments are of the form $\gjo U V W$ where $U$ and $V$
+are inputs, $V$ is the induction hypothesis and $W$ the output. They
+are parameterized by a constant and implicit signature $\Sigma$, and
+all premises are ordered in the algorithmic order of computation. The
+main judgement is atomic object typing $\jf\Delta\Gamma F {\Delta'}
+{F'} P$, which reads: in repository $\mr$ and environment $\Gamma$,
+applicative term $F$ is well-typed of type $A$, and the slicing of its
+value produces a new repository $\mmr$. $F'$ is the \emph{residual} of
+this operation: only constant applications $\app{\cst c} S$ are
+sliced, $F'$ is the ``tip'' of $F$ that was not sliced.
+
+These typing rules use definitional equality between two families
+$\jea\Delta\Gamma A {A'}$.
+% TODO expliquer, et commenter reduction des fonctions ou pas,
+% conv. typee ou pas
+
+\begin{figure}
   \fbox{$\jm\Delta\Gamma M A {\Delta'} {M'}$}
@@
     \infer[SNil]{ }{
-      \jl\Delta\Gamma {\app{\cst a} S} \spinenil \Delta \spinenil {{\app{\cst a} S}}
+      \jl\Delta\Gamma P \spinenil \Delta \spinenil P
     }
@@
   \label{fig:obj-typing}
-\end{figure*}
+\end{figure}
 
@@
 
+It is correct to strengthen $\Gamma$ only \wrt\ $F$ before storing $F$
+as a slice because of the following:
+
+% TODO verif qu'on parle bien de LF normal
 \begin{lemma}
-  If $\jlfm\Gamma F A$, then $\jlfm{\stren{\Gamma, F}} F A$
+  If $\jm\Delta\Gamma F A {\Delta'} {F'}$, then
+  $\jm\Delta{\stren{\Gamma, F}} F A {\Delta'}{F'}$
 \end{lemma}
-
-Our algorithm accepts exactly the well-typed objects of the usual \LF,
-except that we thread a repository $\mr$ and store in it all
-well-typed applicative subterms $F$ along with their classifier $A$
-and the environment in which they are well-typed.
-
-% TODO parler d'effacer les R et retomber sur LF
-% TODO expl. c'est tout comme LF sauf qqes règles...
-% TODO parler de la redondance en LF normal
-
-The main judgement is atomic object typing $\jf\Delta\Gamma F
-{\Delta'} {F'} P$, which reads: in repository $\mr$ and environment
-$\Gamma$, applicative term $F$ is well-typed of type $A$, and the
-slicing of its value produces a new repository $\mmr$. The object $F'$
-is the \emph{residual} of this operation: only constant applications
-$\app{\cst c} S$ are sliced, $F'$ is the ``tip'' of $F$ that was not
-sliced.
-
-These typing rules use definitional equality between two families
-$\jea\Delta\Gamma A {A'}$.
-% TODO expliquer, et commenter reduction des fonctions ou pas,
-% conv. typee ou pas
 
@@
 
-% \begin{figure*}
+% \begin{figure}
 %   \fbox{$\jsig\Delta\Sigma{\Delta'}{\Sigma'}$}
@@
 %   \label{fig:sign-typing}
-% \end{figure*}
+% \end{figure}
 
@@
   Let $\mr = (\Delta, \smeta X \sigma)$.  If $\jf\Delta\Gamma F
-  {\Delta'} {\smeta Y {\sigma'}} P$, let $\function{commit$_\mr$}{F}$ be
-  $\mmr = (\Delta', \smeta X\sigma)$.
+  {\Delta'} {\smeta {X'} {\sigma'}} P$, then we define $\commit\mr F =
+  (\Delta', \smeta X\sigma)$. We say that $\mr$ is
+  \emph{well-constructed} if $\mr = \commit\mmr F$ with either
+  $\mmr=(\mapempty, \smeta X\sigma)$ or $\mmr$ well-constructed.
 \end{definition}
+
+\begin{lemma}
+  If $\mr$ well-constructed, then $\checkout\mr$ is defined and is a
+  value.
+\end{lemma}
+
+We can already state a conservativity result stating that our
+algorithm accepts exactly the well-typed values of \LF:
+
+\begin{theorem}
+  If $F$ is a value, and $\mr$ is well-constructed then $\commit\mr F
+  = \mmr$ iff $\jlfm\Gamma{\checkoutr F}A$
+\end{theorem}
+
+\paragraph{Evaluation}
+
+Rule \textsc{FAppEval} calls the evaluation of a function. We now turn
+to its description (Fig. \ref{fig:evaluation}).
+
+% TODO
 
@@
 
-\begin{figure*}
+\begin{figure}
   \fbox{$\jcc\Delta\Gamma A T S {\Delta'} F P$}
@@
   \caption{Evaluation of terms}
-  \label{fig:whr}
-\end{figure*}
+  \label{fig:evaluation}
+\end{figure}
 

### 2012-06-14 doc/papers/LFMTP2012/paper.tex
paper: various

@@
 on $\md_x$ in the $\lambda$ case, while detecting errors early. Of
-course this incurs on the performance; we will see a compensation for
-it by the slicing mechanism described below in \ref{sec:repr}.
+course this incurs on the performance.
 
@@
 
-\section{Designing the framework}
+\section{Design of the framework}
 %% Here is my idea (2 pages)
 
-% explicit LF
-% lambda-lifting
+% TODO explicit LF
+% TODO lambda-lifting
 
@@
 Secondly, we introduce the \CL\ language to transform our \SLF\
-objects (Fig.~\ref{fig:syntax-CL}). We can construct objects out of
-others, and deconstruct them by case analysis. The definition $T$ of a function always begins with a series of
-$\lambda$.
+objects (Fig.~\ref{fig:syntax-CL}). We can construct atomic objects
+out of others, and deconstruct them by case analysis. The definition
+$T$ of a function always begins with a series of $\lambda$.
 
@@
 
-This language is dynamically typed: there is no static guarantee that
-object $\app {\fct f} x$ produced by a function $\fct f : A_1\to A_2 =
-T$ will have type $A_2$, nor that pattern matching is exhaustive. It
-is also non-terminating: we can define a recursive function $\fct f :
-\cst a\to \cst a = \lam x \app {\fct f} {\var x}$. Nevertheless, we
-want to maintain the invariant that each time an object goes in or out
-of a function $\fct f : A_1\to A_2$, it is checked against the type of
-this function (resp. $A_1$ or $A_2$). You can think of interpreted
-functions as mere black boxes taking and producing objects, and being
-able to feed objects to other black boxes; yet the wires linking them
-are labeled with types and objects circulating in them are checked
-against that type.
+This language is untyped, but types are dynamically checked: there is
+no static guarantee that object $\app {\fct f} x$ produced by a
+function $\fct f : A_1\to A_2 = T$ will have type $A_2$, nor that
+pattern matching is exhaustive. It is also non-terminating: we can
+define a recursive function $\fct f : \cst a\to \cst a = \lam x \app
+{\fct f} {\var x}$. Nevertheless, we want to maintain the invariant
+that each time an object goes in or out of a function $\fct f : A_1\to
+A_2$, it is checked against the type of this function (resp. $A_1$ or
+$A_2$). You can think of interpreted functions as mere black boxes
+taking and producing objects, and being able to feed objects to other
+black boxes; yet the wires linking them are labeled with types and
+objects circulating in them are checked against that type.
 
@@
     (\prd x {A_1} {A_2})^A &= \prd x {A_1} (A_2)^A &
-    (\app {\cst a} S)^A &= \prd x {\app{\cst a} S} A \\
+    (\app P)^A &= \prd x P A \\
     \pi_{n+1}(\prd x {A_1}{A_2}) &= \lam x \pi_n(A_2) &
@@
     \pi^y(\prd x {A_1}{A_2}) &= \lam x \pi^y (A_2) &
-    \pi^y(\app{\cst a} S) &= \var y
+    \pi^y(P) &= \var y
   \end{align*}
@@
     \infer[MAtom]{
-      \jf\Delta\Gamma F {\Delta'} {F'} {P}
+      \jf\Delta\Gamma F {\Delta'} {F'} {P'}
        \and
-      \jea{\Delta'}\Gamma P {\app{\cst a} S}
+      \jea{\Delta'}\Gamma P {P'}
     }{
-      \jm\Delta\Gamma F {\app{\cst a} S} {\Delta'} {F'}
+      \jm\Delta\Gamma F P {\Delta'} {F'}
     }
@@
 
+% TODO
+
+\begin{theorem}
+  If $\mr$ well-constructed, and $\commit\mr F = \mmr$, then
+  $\checkout{\mmr}$ is a value.
+\end{theorem}
+
 \section{Related work}

### 2012-06-15 doc/papers/LFMTP2012/paper.tex
paper: finished!

@@
 be the pair of \emph{(i)} an untrusted procedure $\finfer M$ to infer
-a typing derivation out of program $M$; it doesn't have to be
+a typing derivation out of program $M$; it does not have to be
 complete, or even sound, and \emph{(ii)} a trusted and fast kernel
@@
 Several generic methods have been devised for turning a batch program
-into an incremental one
-\cite{pugh1989incremental,acar2003selective}. Unfortunately, they
-suffer from their genericity on the particular problem of type
+into an incremental one \cite{acar2006,carlsson2002}. Unfortunately,
+they suffer from their genericity on the particular problem of type
 checking:
@@
   derivation-generating type checkers, based on a notion of function
-  inverse %TODO le nom?
+  inverse
   to refer to already-computed values;
@@
 
-% \begin{figure}
-%   \centering
-% %  \small
-%   \begin{mathpar}
-%     \infer{ }{\jts{\cst o} {\even}}
-
-%     \infer{\jts M \odd} {\jts {\s M} \even}
-
-%     \infer{\jts M \even} {\jts {\s M} \odd}
-
-%     \infer{\jts M \nat} {\jts {\s M} \nat}
-
-%     \infer{\jts M {A\to B} \and \jts N A} {\jts {\app M N} B}
-
-%     \infer{\infer*{}{\mbox{$
-%           \begin{array}{c}
-%             [\jts {\var x} A] \\[-0.3em]
-%             \vdots\\
-%             \jts M B
-%           \end{array}$
-%         }}} {\jts {\lam x M} {A\to B}}
-
-%     \infer{\jts M A \and \jsub A B}{\jts M B}
-
-%     \infer{\jts M \nat
-%       \and
-%       \jts N A
-%       \and
-%       \infer*{}{\mbox{$
-%         \begin{array}{c}
-%           [\jts {\var x} \nat] \quad
-%           [\jts {\var y} A] \\[-0.3em]
-%           \vdots \\
-%           \jts P A
-%         \end{array}
-%         $}}
-%     }{\jts {\recb M N x y P} A}
-
-%     \infer{ }{\jsub A A}
-
-%     \infer{ }{\jsub \even\nat}
-
-%     \infer{ }{\jsub \odd\nat}
-
-%     \infer{\jsub {A'} A \and \jsub B {B'} }{\jsub {A\to B}{A'\to B'}}
-%   \end{mathpar}
-%   \caption{System \sysname{T_{<:}}, declaratively}
-%   \label{fig:tsub}
-% \end{figure}
-
 A student is learning about System \sysname {T_{<:}} --- a simply
@@
 two sets $\cst{even}$ and $\cst{odd}$, introduction $\z$ and $\s{M}$,
-elimination ($\recb M N x y P$), and subtyping on those types. He (or
-she) wants to write a program computing $(2+2)\times 3$. He knows that
-sometimes those $\cst{rec}$ arguments are quite hard to get correct,
-but \sysname{T_{<:}} is a typed language; he wants to take advantage
-of this to build his program incrementally using the teacher's
-provided incremental verifier. As a first attempt, he wants to verify
-$P_1 = \recb{\s\z}{\s\z} x y {\s {\var x}}$: He sends the command
+elimination ($\recb M N x y P$), and subtyping on these types (for
+formal details, see appendix). He (or she) wants to write a program
+computing $(2+2)\times 3$ in that system. He knows that sometimes the
+arguments of $\cst{rec}$ are quite hard to get correct, but
+\sysname{T_{<:}} is a typed language; he wants to take advantage of
+this to build his program incrementally using the teacher's provided
+incremental verifier. As a first attempt, he wants to verify $P_1 =
+\recb{\s\z}{\s\z} x y {\s {\var x}}$: he sends the command
 $\finfer{P_1}$ to the system. The $\finfer{}$ operator takes a
@@
 But he (or his text editor) realizes that some parts of this program
-have not changed: both $\s\z$, and the function body $\s{\var u}$, and
-he could reuse the pieces of derivation just built. Let us call these
+have not changed: both $\s\z$, and the function body $\s{\var u}$. He
+could reuse the pieces of derivation just built. Let us call these
 respectively $\md_2$, $\md_3$ and $\md_4$. $\md_4$ is hypothetical: it
@@
 \end{align*}
-that is, only the changed subterms of the program. Derivation $\md_2$
-used to live in the empty context, we are then free to use it in the
-new context where $\vdash\var{add}:\cst{nat}\to\cst{nat}$, because System
-\sysname T enjoys the \emph{weakening} property; $\md_4$ depending on
-a derivation\ \ $\vdash\var x:\cst{nat}$, we instantiate this hole
-with the derivation $\finfer{\var u}$ inferred recursively for the new
+that is, only the changed subterms of the program: this is a
+\emph{delta}. Derivation $\md_2$ used to live in the empty context, we
+are then free to use it in the new context where
+$\vdash\var{add}:\cst{nat}\to\cst{nat}$, because System \sysname T
+enjoys the \emph{weakening} property; $\md_4$ depending on a
+derivation\ \ $\vdash\var x:\cst{nat}$, we instantiate this hole with
+the derivation $\finfer{\var u}$ inferred recursively for the new
 recursive argument $\var u$. $\md_2$ derives $\vdash \s\z :
@@
 $\md_4$ the system succeeds with a new derivation made out of these,
-without having to regenerate them because it knows that
-$\finfer{\fget\md} = \md$.
+without having to regenerate them.
 
@@
 $\var{add}$, $\md_6$ for the arguments
-$\app{\fget{\md_2}}{\fget{\md_2}}$ and $\md_7$ for the call
+$\app{\fget{\md_2}}{\fget{\md_2}}$ and $\md_7$ for the application
 $\app{\var{add}} {\fget{\md_6}}$. Even if it occurs originally in the
-scope of $\var{add}$, $\md_6$ doesn't depend on the hypothesis $\var
+scope of $\var{add}$, $\md_6$ does not depend on the hypothesis $\var
 H$ that $\vdash \var{add}:\cst{nat}\to\cst{nat}$ because of the
@@
 coercion since it was used as an argument to $\var{add}$); the system
-thus infer the more general type $\cst{nat}$ for $\s{\fget{\md_2}}$.
+thus infers the more general type $\cst{nat}$ for $\s{\fget{\md_2}}$.
 
@@
 
-%TODO figure?
-
 Seeing type-checking as issuing a completely explicit typing
@@
 technique for incremental computation that has been extensively
-studied and combined with other techniques
-\cite{pugh1989incremental,acar2003selective}. However, it is best
-understood in the case of purely first-order data, and to our best
-knowledge remains unexplored for higher-order structures containing
-binders, like programs and proofs.
+studied and combined with other techniques \cite{acar2006}. However,
+it is best understood in the case of purely first-order data, and to
+our best knowledge remains unexplored for higher-order structures
+containing binders, like programs and proofs.
 
@@
 take advantage of a higher-order representation of the syntax, so it
-shouldn't be necessary to thread any explicit environment. The
+should not be necessary to thread any explicit environment. The
 definition should then start with:
@@
 right type: it is its \emph{contract} so to say. He first decomposes
-term $M$ and treats the easy cases of $\cst{o}$ and $\cst{s}$:
+term $M$ and treats the easy cases of $\cst{o}$ and $\cst{s}$. The
+code continues with:
 
@@
 \noindent
-We write $\pair{\cdot}{\cdot}$ the constructor of $\Sigma$-types;
-patterns don't have to be exhaustive since we are in an untrusted
-setting; for conciseness, we use case failure for signaling typing
+We write $\pair{\cdot}{\cdot}$ for the constructor of $\Sigma$-types.
+Patterns do not have to be exhaustive since we are in an untrusted
+setting. Fr conciseness, we use case failure for signaling typing
 error. There are three cases for $\s{M}$ in the system, corresponding
@@
 error. Notice the slight optimization to the size of the derivation:
-if the types $A_1$ and $A_2$ are the same syntactically, we don't
+if the types $A_1$ and $A_2$ are the same syntactically, we do not
 generate a subtyping rule. This is done by matching on the returned
-derivation $\md_\leq$.
+derivation $\md_\leq$ in the case of the \textsc{Refl} rule.
 
@@
 the recursive call $\finfer{M}$ must be done in an enlarged
-environment where $\vdash x : A$, otherwise the check that the
+environment where $\vdash x : A$. Otherwise the check that the
 returned open derivation $\md$ is correct would fail ($x$ is
 unknown). We thus introduce a new construct $\matchin M \Gamma \ldots$
-and its $\syntax{let}$ counterpart ($\match M \ldots$ being syntactic
-sugar for $\matchin M \envnil \ldots$). Even then, in the recursive
-call, the variable case $\var x$ would fail since we would have no way
-of telling its type anymore. Thus we substitute all occurrences of
-$\var x$ with the \emph{output} of function $\finfer{}$ on it (since
-it is known at this point), \ie\ a pair of $A$ and $\md_1$. For this
-substitution to be well-typed, we now need to coerce this output to a
-valid input (a term). We introduce an operator which is in some sense
-the \emph{inverse} of $\finfer{}$: operator $\fget{}$ maps a pair of a
-type and a derivation to the term it types.
+($\match M \ldots$ being syntactic sugar for $\matchin M \envnil
+\ldots$). Even then, in the recursive call, we would have no way of
+synthesizing a variable $\var x$'s type anymore. Thus we substitute
+all occurrences of $\var x$ with the \emph{output} of function
+$\finfer{}$ on it (since it is known at this point), \ie\ a pair of
+$A$ and $\md_1$. For this substitution to be well-typed, we now need
+to coerce this output to a valid input (a term). We introduce an
+operator which is in some sense the \emph{inverse} of $\finfer{}$:
+operator $\fget{}$ maps a pair of a type and a derivation to the term
+it types.
 
@@
 At this point, we get back $B$ and $\md$ from the recursive call. We
-know that $B$ can't depend on $\md_x$ (because a System
+know that $B$ cannot depend on $\md_x$ (because a System
 \sysname{T_{<:}} type is always closed), but $\md$ can: it is not
@@
 where a special judgement named $\md_x$ can occur. To return the
-complete derivation of $\vdash\tlam A x M : A \to B$, we must then
+complete derivation of $\vdash\tlam x A M : A \to B$, we must then
 close, or ``hypothesize'' $\md$ by $\md_x$, and then apply rule
@@
 
-Remains one problem with subtyping: terms $N : A_N$ and $P : A_P$
+One problem remains with subtyping: terms $N : A_N$ and $P : A_P$
 should be coercible to the same type $A$, but $P$ should be typed
 under $\vdash y:A$, which is not known before typing $P$. A trick to
-break the loop is to iterate the typing of $P$ with more and more
-general types, until we reach a fixpoint (we reach a fixpoint if there
-is a unique more general type); in our particular case of subtyping
+break the loop is to iterate the typing of $P$ with a more and more
+general type, until we reach a fixpoint. We reach a fixpoint if there
+is a unique more general type, in our particular case of subtyping
 relation, this means iterating only twice: we will then attain
@@
 (\vdash A\leq C )\times(\vdash B\leq C) = \ldots$$ computing the sup
-of two types $A$ and $B$, and returning proofs of of subtyping. Note
-that we didn't perform here the optimization seen in the application
-case.
+of two types $A$ and $B$, and returning proofs of subtyping. Note that
+we could have performed the optimization seen in the application case
+as well.
 
 The type-checker is now done. There is no case for variables, as we
-won't ever meet this case: we already substituted all variables with
-their result derivation. Evaluation only need to make sure that in
-this case, the derivation will be returned, \ie\ that equation
+will not ever meet this case: we already substituted all variables
+with their result derivation. Evaluation only needs to make sure that
+in this case, the derivation will be returned, \ie\ that equation
 $\finfer{\fget{\pair A \md}} = \pair A \md$ holds.
@@
 Note that $\fget{}$ can be derived automatically from the type of
-$\finfer{}$: it is its inverse. It has definition:
+$\finfer{}$: we will call it its inverse, since it maps the return
+type to the arguments. It has definition:
 
 \begin{mathleft}
-  \fget{}\ :\ \prd{\{M\}}{\cst{tm}} \prd {A}{\cst{tp}} \prd{\md}{(\vdash
+  \fget{}\ :\ \prd{M}{\cst{tm}} \prd {A}{\cst{tp}} \prd{\md}{(\vdash
     M:A)} \cst{tm} =
@@
 \noindent
-where argument $M$ is implicit.
+where argument $M$ is left implicit (it can be inferred).
 
@@
 
-We exploit here an idea similar to run-time \emph{contracts}
+We exploit an idea similar to run-time \emph{contracts}
 \cite{wadler2009well}: there is no static guarantee that computations
-won't return ill-typed values, but arguments and results are checked
-at run-time at the boundaries of defined functions. This allows to
-omit formal justifications, like the fact that we need to iterate
-twice the typing of $P$ in $\recb M N x y P$, or that $B$ can't depend
-on $\md_x$ in the $\lambda$ case, while detecting errors early. Of
-course this incurs on the performance.
+will not return ill-typed values, but arguments and results are
+checked at run-time at the boundaries of defined functions. This
+allows to omit formal justifications, like the fact that we need to
+iterate twice the typing of $P$ in $\recb M N x y P$, or that $B$
+cannot depend on $\md_x$ in the $\lambda$ case, while detecting errors
+early.
 
@@
 assistants: during manual proof search, the interaction is driven by
-\emph{tactics}: %TODO cite
-they are untrusted programs generating pieces of proofs, assembled by
-other, higher-order tactics; verification of the generated proofs is
-done when the evaluation of all tactics is complete. The differences
-with our mechanism are two-fold: first, tactics usually take one (or
-several) \emph{goal}(s) as input, \ie\ holes in a proof, whereas our
-functions can take any term; secondly, proof-check is usually done at
-the end of the evaluation, which can lead to less tractable bugs; by
-making evaluation and checking mutually recursive, we choose to detect
-errors earlier in these witnesses: exactly at the point where they are
-generated.
+\emph{tactics}: they are untrusted programs generating pieces of
+proofs, assembled by other, higher-order tactics; verification of the
+generated proofs is done when the evaluation of all tactics is
+complete. The differences with our mechanism are two-fold: first,
+tactics usually take one (or several) \emph{goal}(s) as input, \ie\
+holes in a proof, whereas our functions can take any term; secondly,
+proof-check is usually done at the end of the evaluation, which can
+lead to less tractable bugs; by making evaluation and checking
+mutually recursive, we choose to detect errors earlier in these
+witnesses: exactly at the point where they are generated.
 
@@
 $$
-% TODO: vérif ce paragraphe après avoir parlé de fcts inv
 replace the usual ones. The difficulty with this technique,
@@
 
-% TODO cite lee2007mechanizing
-
 \section{Design of the framework}
 %% Here is my idea (2 pages)
-
-% TODO explicit LF
-% TODO lambda-lifting
 
@@
 for short, is one of the well-accepted solution to this question. It
-is a dependently typed lambda-calculus for \emph{representing} terms,
-derivations and statements of $\Pi_0^1$ meta-theorems in a given
-inference system supporting hypothetical judgments
-\cite{pfenning2001logical}. For these, it provides built-in support
-for encoding languages with binders and hypothesis (a feature called
-$\lambda$-tree syntax or HOAS \cite{pfenning1988higher}). The user
-gives a \emph{signature} defining abstract syntax and inference rules
-of his \emph{object language}, and gets an algorithm to verify terms
-and derivations expressed in the \LF\ syntax.
+is a dependently typed lambda-calculus for \emph{representing}
+higher-order terms and derivations
+\cite{pfenning1988higher}. The user gives a
+\emph{signature} defining abstract syntax and inference rules of his
+\emph{object language}, and gets an algorithm to verify terms and
+derivations expressed in the \LF\ syntax.
 % For instance, the rule for System \sysname{T}'s recursor can be
@@
 % \end{mathleft}
-This framework has been used to issue certificates for a number of
-practical applications like proof-carrying
-code~\cite{necula1997proof}, authentication
-framework~\cite{appel1999proof} \ldots %TODO ATP
-
-To write deltas by sharing subterms, we need to be able to address any
-subterm. Since a subterm can be open, we must instantiate its free
-variables. For this, we define the \emph{slicing} of a well-typed \LF\
-term: we extend \LF\ with contextual metavariables $\meta X$, $\meta
-Y$, \ldots\ as in \cite{nanevski2008contextual}. We thus introduce
-Sliced \LF\ (\SLF, see Fig. \ref{fig:syntax-LF}). Specifically, it is
-an extension of \emph{spine canonical \LF} \cite{pfenning2007term}
-enriched with metavariables which are references to open terms as in
-\cite{nanevski2008contextual}.
+
+To write deltas by sharing subterms, we need to be able to refer to
+any subterm. For this, we first extend \LF\ with contextual
+metavariables (written $\meta X$, $\meta Y$) standing for open terms,
+following \cite{nanevski2008contextual}, secondly, we make sure that
+every subterm is named by a metavariable (an operation called
+\emph{slicing}). We thus introduce Sliced \LF\ (\SLF, see
+Fig. \ref{fig:syntax-LF}): it is an extension of \emph{spine canonical
+  \LF} \cite{pfenning2007term} enriched with metavariables which are
+references to open terms as in \cite{nanevski2008contextual}.
 
@@
 In this variant, only $\eta$-long and $\beta$-normal objects have an
-existence: canonicity for $\beta$-reduction is enforced in the syntax
+existence. Canonicity for $\beta$-reduction is enforced in the syntax
 and canonicity for $\eta$-expansion will be enforced by typing,
@@
 (\ie\ to a \emph{spine} of objects) as opposed to the more standard
-binary application: this eases the definition substitution and
-$\eta$-expansion, and corresponds to a focused sequent calculus known
-as \sysname{LJT} \cite{herbelin1995lambda}. We differ the discussion of
-function symbols $\fct f$ to \ref{sec:comput}. A \emph{value} is an
-object containing no function symbols. The definition of hereditary
-substitution $\gsubst M \sigma$ is skipped and can be found in
-\cite{pfenning2007term} (and extended to metavariables as
-\cite{nanevski2008contextual}).
+binary application: this eases the definition of substitution and
+$\eta$-long forms, and corresponds to a focused sequent calculus known
+as \sysname{LJT}~\cite{herbelin1995lambda}. We differ the discussion
+of function symbols $\fct f$ to section \ref{sec:comput}. A
+\emph{value} is an object containing no function symbols. The
+definition of hereditary substitution $\gsubst M \sigma$ is skipped
+and can be found in the appendix.
 
@@
 \emph{inferrable} $F$ objects (resp. \emph{canonical} and
-\emph{atomic}), as in \eg\ \cite{pierce2000local}: canonical objects
+\emph{atomic}), as in \eg\ \cite{pierce2000local}. Canonical objects
 are checked against a given type, whereas we can synthesize the type
-of atomic objects; the coercion from the first grammatical category to
-the second is where we verify equality between synthesized and checked
-types.
+of atomic objects; the coercion from canonical to atomic objects
+triggers the equality check between synthesized and checked types.
 
 As is customary, we adopt the Barendregt convention for naming bound
-variables, write $A_1\to A_2$ for $\prd x {A_1} {A_2}$ when $\var x\notin
-\FV{A_2}$, $\meta X$ for $\smeta X \msubstnil$, and adopt indifferently
-list, map or set notations for $\sigma$ and $\Gamma$. The identity
-substitution $\msubstid{\Gamma}$ is a notation for $\{\msubst x {\var
-  x}\ |\ \var x \in\dom\Gamma\}$. Metavariables $\smeta X \sigma$
-stand for open objects with free variables in $\dom{\sigma}$; To each
-use of them is attached a substitution $\sigma$ defining these free
-variables. We write $\FMV{M}$ for the set of metavariables in $M$ and
-speak of a \emph{metaclosed} object $M$ if $\FMV{M} = \emptyset$.
-
-% TODO relate to LF
-
-% \paragraph{Substitution}
-
-% We overload \emph{substitution} as a partial operation on objects,
-% given a parallel substitution $\sigma$. Since the syntax of object is
-% not stable by (textual) substitution, it is defined as
-% \emph{hereditary}\footnote{ Note that this notion of substitution,
-%   present in \cite{pfenning2007term}, differs from the more standard
-%   definition in \cite{hl07mechanizing} as it presupposes $\eta$-long
-%   normal forms: once a \emph{cut} starts, there must be exactly as
-%   many formal and actual arguments. A non-empty spine cut against
-%   another non-empty spine $\gcut {\app H S} {S'}$ blocks the
-%   substitution instead of reducing to $\app H {(\gconcat S S')}$ (an
-%   implicit $\eta$-expansion); conversely, exhausting actual but not
-%   formal arguments as in $\gcut{\lam x M}{\gnil}$ blocks the
-%   substitution too. For example, the reduction $\gcut {\lam x \app
-%     {\var x}{\var x}} {\lam x \app {\var x}{\var x}}$ gets stuck
-%   because the second $\var x$ in the left part is substituted by a
-%   function yet is not itself a function. }: it is mutually recursive
-% with a \emph{cut} function $\gcut M S$ which can in turn trigger a
-% chain of substitutions, leading to a canonical objects or an error.
-% \begin{align*}
-%   \gsubst {(\lam x M)} \sigma &= \lam x {\gsubst M \sigma} &
-%   \text{if $\var x\notin \dom{\sigma}$}
-%   \\
-%   \gsubst {(\app {\cst c} S)} \sigma &= \app {\cst c} {(\gsubst S
-%     \sigma)}
-%   \\
-%   \gsubst {(\app {\var x} S)} \sigma &=
-%   \app {\var x} {\gsubst S \sigma} &
-%   \text{if $\var x\notin \dom{\sigma}$}
-%   \\
-%   \gsubst {(\app {\var x} S)} \sigma &=
-%   \gcut M {\gsubst S \sigma} &
-%   \text{if $\sigma(\var x) = M$}
-%   \\
-%   \gsubst {\smeta X {\sigma}} {\sigma'} &=
-%   \smeta X {\gcomp\sigma{\sigma'}}
-%   \\
-%   \gsubst\spinenil\sigma &=
-%   \spinenil \\
-%   \gsubst{\spinecons M S}\sigma &=
-%   \spinecons{\gsubst M\sigma}{\gsubst S\sigma}
-%   \\
-%   \gcut {\lam x M} {\spinecons N S} &=
-%   \gcut {\gsubst M {\msubst x N}} S
-%   \\
-%   % TODO cas meta/S ? rem: on en a pas besoin ds le proto.
-%   % \gcut {\smeta X\sigma} S &=
-%   % \smeta X{\gcomp\sigma\sigma'}
-%   % \\
-%   \gcut {\app H S} \spinenil &=
-%   \app H S
-%   \\
-%   \gcut {\smeta X\sigma} \spinenil &= \smeta X\sigma
-%   \\
-%   \gcomp \sigma \msubstnil &=
-%   \sigma
-%   \\
-%   \gcomp \sigma {(\msubstcons{\sigma'} x M)} &=
-%   \gcomp {\gsubst\sigma {\msubst x M}} \sigma'
-%   \\
-%   \gsubst {(\msubstcons{\sigma'} x M)} \sigma &=
-%   \msubstcons{\gsubst{\sigma'}\sigma} x {\gsubst M \sigma}
-%   \\
-%   \gsubst {\msubstnil}{\sigma} &= \msubstnil
-% \end{align*}
-
-% We extend this operation trivially on families $\gsubst A \sigma$ and
-% kinds $\gsubst K \sigma$. We also allow to abusively write
-% non-canonical application $\app M S$ as a shorthand for $\gcut M S$
-% if it is defined.
-
-% TODO pourquoi cette def?
-% TODO terminaison?
-% TODO stabilité de la eta?
+variables, write $A_1\to A_2$ for $\prd x {A_1} {A_2}$ when $\var
+x\notin \FV{A_2}$, $\meta X$ for $\smeta X \msubstnil$, and adopt
+indifferently list, map or set notations for $\sigma$ and
+$\Gamma$. The identity substitution $\msubstid{\Gamma}$ is a notation
+for $\{\msubst x {\var x}\ |\ \var x \in\dom\Gamma\}$. Metavariables
+$\smeta X \sigma$ stand for open objects with free variables in
+$\dom{\sigma}$. To each use of them is attached a substitution
+$\sigma$ defining these free variables. We write $\FMV{M}$ for the set
+of metavariables in $M$ and call an object $M$ \emph{metaclosed} if
+$\FMV{M} = \emptyset$.
 
@@
   judgments of well-typing), along with a distinguished metavariable,
-  corresponding to the \emph{tip} of the term contained in it:
+  corresponding to the \emph{tip} of the term it contains:
 $$ \mr\;:\;(\meta X \mapsto (\Gamma\vdash\tp F P))\times\smeta X
@@
 for $\gsubst M\sigma$ if $\Delta(\meta X) = (\Gamma\vdash M : A)$.
-The algorithm in \ref{sec:eval} will maintain the invariant that a
-repository $\mr$ is \emph{acyclic}, that is for all $\meta X$ in
-$\mr$, the fixpoint of $\FMV{\mr(\meta X)}$ does not contain $\meta X$.
+The algorithm in \ref{sec:eval} will maintain two invariants: a
+repository $\mr$ is \emph{acyclic}, (for all $\meta X$ in
+$\mr$, the fixpoint of $\FMV{\mr(\meta X)}$ does not contain $\meta
+X$) and all $F$ and $P$ in it are values.
 
@@
 checks out as the metaclosed atomic object
-$$
-\fbox{$
+$
   \app {\cst{lam}} \lam x
-  \fbox{$
     \app {\cst{lam}} \lam y
-    \fbox{$
       \app {\cst{app}} \app {\var y}
-      (\fbox{$
+      (
         \app{\cst{app}}\app{\var x}{\cst o}
-      $})
-    $}
-  $}
-$}
-$$
-where all referable slices have been boxed.
+      )
+$.
 \end{example}
@@
 
-Secondly, we introduce the \CL\ language to transform our \SLF\
+Secondly, we introduce the \CL\ language compute over our \SLF\
 objects (Fig.~\ref{fig:syntax-CL}). We can construct atomic objects
@@
     (\prd x {A_1} {A_2})^{n+1} &= \prd x {A_1} (A_2)^n &
-    (\prd x {A_1} {A_2})^0 &= \prd x {A_1} (A_2)^{A_1} \\
-    (\prd x {A_1} {A_2})^A &= \prd x {A_1} (A_2)^A &
-    (\app P)^A &= \prd x P A \\
+    (\prd x {A_1} {A_2})^0 &= \prd x {A_1} (A_2)_{A_1} \\
+    (\prd x {A_1} {A_2})_A &= \prd x {A_1} (A_2)_A &
+    (P)_A &= \prd x P A \\
     \pi_{n+1}(\prd x {A_1}{A_2}) &= \lam x \pi_n(A_2) &
-    \pi_0(\prd x {A_1} {A_2}) &= \lam {y} \pi^y(A_n) \\
+    \pi_0(\prd x {A_1} {A_2}) &= \lam {y} \pi^y(A_2) \\
     \pi^y(\prd x {A_1}{A_2}) &= \lam x \pi^y (A_2) &
@@
 \end{definition}
+Functions $(A)^n$ and $\pi_n(A)$ use auxiliary functions $(A)_B$ and
+$\pi^y(A)$ to put the $n$-th argument at the end of the type and term.
 
@@
 
-The evaluation algorithm, called $\function{commit$_\mr$}{F}$ takes an
-atomic object $F$ (possibly containing function symbols to evaluate)
-and a repository $\mr$, and produces a new repository $\mr'$ which is
-a strict extension of $\mr$ with a different tip, and which checks out
-to the value of $F$. It has three roles, which are necessarily
-interleaved:
-
+The evaluation algorithm, called \emph{commit}
+$\function{commit$_\mr$}{F}$ takes an atomic object $F$ (possibly
+containing function symbols to evaluate) and a repository $\mr$, and
+produces a new repository $\mr'$ which is a strict extension of $\mr$
+with a different tip, and which checks out to the value of $F$. It has
+three roles, which are necessarily interleaved:
 \begin{itemize}
@@
 
-\paragraph{Typing}
-
-We first present the bidirectional typing and slicing part of the
-algorithm (Fig. \ref{fig:obj-typing}) as a syntax-directed inference
-system. All judgments are of the form $\gjo U V W$ where $U$ and $V$
-are inputs, $V$ is the induction hypothesis and $W$ the output. They
-are parameterized by a constant and implicit signature $\Sigma$, and
-all premises are ordered in the algorithmic order of computation. The
-main judgement is atomic object typing $\jf\Delta\Gamma F {\Delta'}
-{F'} P$, which reads: in repository $\mr$ and environment $\Gamma$,
-applicative term $F$ is well-typed of type $A$, and the slicing of its
-value produces a new repository $\mmr$. $F'$ is the \emph{residual} of
-this operation: only constant applications $\app{\cst c} S$ are
-sliced, $F'$ is the ``tip'' of $F$ that was not sliced.
-
-These typing rules use definitional equality between two families
-$\jea\Delta\Gamma A {A'}$.
-% TODO expliquer, et commenter reduction des fonctions ou pas,
-% conv. typee ou pas
+We now turn to the bidirectional typed reduction and slicing of the
+algorithm (Fig. \ref{fig:obj-typing}, \ref{fig:obj-typing2},
+\ref{fig:evaluation}), presented as an inference system. All judgments
+are of the form $\gjao L U V W$ where $U$, $V$ and $L$ are inputs, are
+directed by the syntax of $V$ and $L$, and $W$ is the output. They are
+parameterized by a constant and implicit signature $\Sigma$.
+
+Label $L$ is a mode influencing the strategy of evaluation of defined
+function applications. They range over $\top$, $\bot$ and $\fct f$,
+meaning respectively strong reduction, pure checking with no reduction
+and call-by-name, head reduction. The purpose of the $\fct f$
+annotation is shown in rule \textsc{FAppEvalInv}: it detects the
+special redex form $\app{\fct f}\app {(\app{\fcti f
+    0}S_0)}\app\ldots{(\app{\fcti f n} S_n)}$, reducing it to the last
+argument of the spines $\function{last}{S_0}$, provided all $S_i$ are
+equal. When $n=0$, we recover the equality $\finfer{\fget{X}} = X$.
+
+The main judgment is atomic object typed reduction
+$\jf\aeval\Delta\Gamma F {\Delta'} {F'} P$, which reads: in repository
+$\Delta$ and environment $\Gamma$, applicative term $F$ is well-typed of
+type $P$, and the slicing of its value produces a new repository
+$\Delta'$. $F'$ is the \emph{residual} of this operation: only constant
+applications $\app{\cst c} S$ are sliced, $F'$ is the ``tip'' of $F$
+that was not sliced.
+
+These typing rules use typed definitional equality between families
+$\jea\Delta\Gamma {A_1} {A_2}$ and between spines $\jel\Delta\Gamma A
+{S_1} {S_2} P$. The formal definition of these judgments is eluded
+here. It compares syntactically the values of its arguments. Since it
+potentially triggers function evaluation (and thus typing
+verification), we carry the repository, the environment and the type
+of the compared objects.
 
 \begin{figure}
-  \fbox{$\jm\Delta\Gamma M A {\Delta'} {M'}$}
+  \fbox{$\jm L \Delta\Gamma M A {\Delta'} {M'}$}
   \qquad
@@
     \infer[MLam]{
-      \jm\Delta{\envcons\Gamma x A} M {A'} {\Delta'} {M'}
-    }{
-      \jm\Delta\Gamma {\lam x M} {\prd x A {A'}} {\Delta'} {\lam x M'}
+      \jm {\fann L} \Delta{\envcons\Gamma x A} M {A'} {\Delta'} {M'}
+    }{
+      \jm {L} \Delta\Gamma {\lam x M} {\prd x A {A'}} {\Delta'} {\lam x M'}
     }
 
+    \begin{array}{rl}
+      \fann \top &= \top \\
+      \fann \bot &= \bot \\
+      \fann {\fct f} &= \bot \\
+    \end{array}
+
     \infer[MAtom]{
-      \jf\Delta\Gamma F {\Delta'} {F'} {P'}
+      \jf L \Delta\Gamma F {\Delta'} {F'} {P'}
        \and
@@
     }{
-      \jm\Delta\Gamma F P {\Delta'} {F'}
+      \jm L \Delta\Gamma F P {\Delta'} {F'}
     }
-    % TODO commenter type ground
   \end{mathpar}
 
-  \fbox{$\jf\Delta\Gamma F {\Delta'} {F'} P$}
+  \fbox{$\jf L \Delta\Gamma F {\Delta'} {F'} P$}
   \qquad
@@
 
-    % Non-sliceable
+    % Variable
     \infer[FAppVar]{
-      (x:A)\in\Gamma
-      \and
-      \jl\Delta\Gamma A S {\Delta'} {S'} P
-    }{
-      \jf\Delta\Gamma {\app {\var x} S} {\Delta'} {\app {\var x} {S'}} P
+      \jl {\fann L} \Delta\Gamma A S {\Delta'} {S'} P
+    }{
+      \jf L \Delta\Gamma {\app {\var x} S} {\Delta'} {\app {\var x} {S'}} P
+    }\quad{\small (x:A)\in\Gamma}
+
+    % Variable
+    \infer[FAppConst]{
+      \jl {\fann L} \Delta\Gamma A S {\Delta'} {S'} P
+    }{
+      \jf L \Delta\Gamma {\app {\cst c} S} {\Delta'} {\app {\cst c} {S'}} P
+    }\quad{\small
+      \begin{array}{l}
+        (c:A)\in\Sigma\\
+        L\neq\top
+      \end{array}
     }
 
-    % Sliceable
-    \infer[FAppConst]{
-      (c:A)\in\Sigma
-      \and
-      \jl\Delta\Gamma A S {\Delta'} {S'} {P}
+    % Constant
+    \infer[FAppConstSlice]{
+      \jl {\top} \Delta\Gamma A S {\Delta'} {S'} {P}
       \and
@@
     }{
-      \jf\Delta\Gamma {\app {\cst c} S} {\mapadd{\Delta'}{\meta X}{(\Gamma'\vdash
+      \jf \top \Delta\Gamma {\app {\cst c} S} {\mapadd{\Delta'}{\meta X}{(\Gamma'\vdash
           {\app {\cst c} {S'}}:P)}} {\smeta X{\msubstid{\Gamma'}}} P
     }
-    \quad\text{($\meta X$ fresh in $\Delta'$)}
-
-    % Defined
+    \quad
+    \begin{array}{l}
+      \small
+      (c:A)\in\Sigma\\
+      \text{($\meta X$ fresh in $\Delta'$)}
+    \end{array}
+
+    % Evaluation
     \infer[FAppEval]{
-      (\fct f : A = T) \in \Sigma
+      \jl{L'}\Delta\Gamma A S {{\Delta'}} {S'} P
+      \\\\
+      \jcc{\Delta'}\Gamma T {S'} {\Delta''} F
       \and
-      % check args pas nécessaire (fait ds la red)
-      % \jl\Delta\Gamma A S {{\Delta'}} {S'} P
-      % \and
-      \jcc\Delta\Gamma A T S {\Delta'} F P
-      % check result pas nécessaire (fait ds la red)
-      % \and
-      % \jm{\Delta''}\Gamma F P {\Delta'''} {F'}
-    }{
-      \jf\Delta\Gamma {\app {\fct f} S} {\Delta'} {F'} P
-    }
+      \jm L {\Delta''}\Gamma F P {\Delta'''} {F'}
+    }{
+      \jf L \Delta\Gamma {\app {\fct f} S} {\Delta'''} {F'} P
+    } \quad
+    \begin{array}{l}\small
+      (\fct f : A = T) \in \Sigma \\
+      {S' \neq \app {(\app {\fcti f 0} S_0)} \app\ldots {(\app
+          {\fcti f n} {S_n})}} \\
+      (L = \fct g \wedge L'=\bot) \vee (L=\top\wedge L'=\fct f)
+    \end{array}
+
+    \infer[FAppEvalInv]{
+      \jl{\fct f} \Delta\Gamma A S {\Delta'}{\app {(\app {\fcti f 0} S_0)} \app\ldots {(\app
+          {\fcti f n} {S_n})}} P
+      \\\\
+      \text{for all $i$,$j$, }\ (\jel\Delta\Gamma A {S_i}{S_j} P)
+    }{
+      \jf L \Delta\Gamma {\app {\fct f} S}
+      {\Delta'}{\function{last}{S_0}} P
+    }\quad{\small (L = \top \vee L = \fct g)}
+
+    \infer[FAppNoEval]{
+      \jl\bot\Delta\Gamma A S {\Delta'} {S'} P
+    } {
+      \jf \bot \Delta\Gamma {\app{\fct f} S} {\Delta'} {\app{\fct f}
+        {S'}} P
+    }\quad{(\fct f : A = T\in\Sigma)}
 
@@
       \and
-      \js\Delta\Gamma \sigma {\Gamma'} {\Delta'} {\sigma'}
-    }{
-      \jf\Delta\Gamma {\smeta X \sigma} {\Delta'} {\smeta X {\sigma'}} {\gsubst{P}{\sigma'}}
+      \js{\fann L}\Delta\Gamma \sigma {\Gamma'} {\Delta'} {\sigma'}
+    }{
+      \jf L \Delta\Gamma {\smeta X \sigma} {\Delta'} {\smeta X {\sigma'}} {\gsubst{P}{\sigma'}}
     }
@@
 
-  \fbox{$\jl\Delta\Gamma A S {\Delta'}{S'} P$}
+  \caption{Typed evaluation for objects (1)}
+  \label{fig:obj-typing}
+\end{figure}
+
+\begin{figure}
+  \fbox{$\jl L\Delta\Gamma A S {\Delta'}{S'} P$}
   \qquad
@@
   \begin{mathpar}
+    \hspace{-1.5em}
     \infer[SCons]{
-      \jm\Delta\Gamma M A {\Delta'} {M'}
+      \jm L \Delta\Gamma M A {\Delta'} {M'}
       \and
-      \jl{\Delta'}\Gamma {\gsubst{A'}{\msubst x {M'}}} S {\Delta''} {S'} P
-    }{
-      \jl\Delta\Gamma {\prd x {A} {A'}} {\spinecons M S} {\Delta''} {\spinecons
+      \jl L {\Delta'}\Gamma {\gsubst{A'}{\msubst x {M'}}} S {\Delta''} {S'} P
+    }{
+      \jl L \Delta\Gamma {\prd x {A} {A'}} {\spinecons M S} {\Delta''} {\spinecons
       {M'}{S'}} P
     }
-
+    \quad
     \infer[SNil]{ }{
-      \jl\Delta\Gamma P \spinenil \Delta \spinenil P
+      \jl L\Delta\Gamma P \spinenil \Delta \spinenil P
     }
   \end{mathpar}
-
-  \fbox{$\js\Delta\Gamma\sigma{\Gamma'}{\Delta'}{\sigma'}$}
+  \fbox{$\js L \Delta\Gamma\sigma{\Gamma'}{\Delta'}{\sigma'}$}
   \qquad
   {Substitution}
-
   \begin{mathpar}
-
     % Cons
     \infer[$\sigma$Cons]{
-      \js\Delta\Gamma\sigma{\Gamma'}{\Delta'}{\sigma'}
+      \js L \Delta\Gamma\sigma{\Gamma'}{\Delta'}{\sigma'}
       \and
-      \jm{\Delta'}\Gamma M {\gsubst A {\sigma'}} {\Delta''} {M'}
-    }{
-      \js\Delta\Gamma{(\msubstcons \sigma x M)} {(\envcons {\Gamma'} x {A'})}
+      \jm {\fann L} {\Delta'}\Gamma M {\gsubst A {\sigma'}} {\Delta''} {M'}
+    }{
+      \js L \Delta\Gamma{(\msubstcons \sigma x M)} {(\envcons {\Gamma'} x {A'})}
       {\Delta''} {(\msubstcons {\sigma'} x {M'})}
     }
-
+    \quad
     % Nil
     \infer[$\sigma$Nil]{ }{
-      \js\Delta\Gamma\msubstnil\envnil\Delta\envnil
+      \js L \Delta\Gamma\msubstnil\envnil\Delta\envnil
     }
@@
 
-  \caption{Typing algorithm for objects}
-  \label{fig:obj-typing}
+  \caption{Typed evaluation for objects (2)}
+  \label{fig:obj-typing2}
 \end{figure}
@@
   the free variables of $F$:
-$$
-\stren{\Gamma,F} = \{x : A\ |\ x\in \FV{F} \wedge \Gamma(x)=A\}
-$$
+$
+\stren{\Gamma,F} = \{x : P\ |\ x\in \FV{F} \wedge \Gamma(x)=P\}
+$.
 \end{definition}
@@
 
-% TODO verif qu'on parle bien de LF normal
 \begin{lemma}
-  If $\jm\Delta\Gamma F A {\Delta'} {F'}$, then
-  $\jm\Delta{\stren{\Gamma, F}} F A {\Delta'}{F'}$
+  If $\jm L \Delta\Gamma F A {\Delta'} {F'}$, then
+  $\jm L \Delta{\stren{\Gamma, F}} F A {\Delta'}{F'}$
 \end{lemma}
@@
 \begin{definition}
-  Let $\mr = (\Delta, \smeta X \sigma)$.  If $\jf\Delta\Gamma F
-  {\Delta'} {\smeta {X'} {\sigma'}} P$, then we define $\commit\mr F =
-  (\Delta', \smeta X\sigma)$. We say that $\mr$ is
-  \emph{well-constructed} if $\mr = \commit\mmr F$ with either
-  $\mmr=(\mapempty, \smeta X\sigma)$ or $\mmr$ well-constructed.
+  Let $\mr = (\Delta, \smeta X \sigma)$.  If $\jf \top\Delta\Gamma F
+  {\Delta'} {\smeta {X'} {\sigma'}} P$, then we define the
+  \emph{commit} operation $\commit\mr F = (\Delta', \smeta
+  X\sigma)$. We say that $\mr$ is \emph{well-constructed} if $\mr =
+  \commit\mmr F$ with either $\mmr=(\mapempty, \smeta X\sigma)$ or
+  $\mmr$ well-constructed.
 \end{definition}
@@
   If $F$ is a value, and $\mr$ is well-constructed then $\commit\mr F
-  = \mmr$ iff $\jlfm\Gamma{\checkoutr F}A$
+  = \mmr$ iff $\jlfm\Gamma{\checkoutrr F}A$
 \end{theorem}
 
-\paragraph{Evaluation}
-
-Rule \textsc{FAppEval} calls the evaluation of a function. We now turn
-to its description (Fig. \ref{fig:evaluation}).
-
-% TODO
-
-% \begin{theorem}
-%   If \/ $\jm\mr\Gamma M A \mmr {M'}$ then \/ $\jem\mmr\Gamma M {M'} A$
-% \end{theorem}
-
-% \begin{theorem}
-%   If \/ $\jm\mr\Gamma M A \mmr {M'}$ then \/ $\jlfm\Gamma {(\checkoutrr
-%     M)} (\checkoutrr A)$
-% \end{theorem}
-
 \begin{figure}
-  \fbox{$\jcc\Delta\Gamma A T S {\Delta'} F P$}
+  \fbox{$\jcc\Delta\Gamma T S {\Delta'} F$}
   \qquad
@@
     \infer[TLam]{
-      \jm\Delta\Gamma M {A_1} {\Delta'} {M'}
-      \and
-      \jcc{\Delta'}\Gamma{\gsubst {A_2} {\msubst x {M'}}}{\gsubst T
-        {\msubst x {M'}}} S {\Delta''} F P
-    }{
-      \jcc\Delta\Gamma{\prd x {A_1} {A_2}}{\lam x T}{\spinecons M S}
-      {\Delta''} F P
+      \jcc{\Delta}\Gamma{\gsubst T
+        {\msubst x {M}}} S {\Delta'} F
+    }{
+      \jcc\Delta\Gamma{\lam x T}{\spinecons M S}
+      {\Delta'} F
     }
@@
       \jcr\Delta\Gamma U {\Delta'} F
-      \and
-      \jm{\Delta'}\Gamma F P {\Delta''} {F'}
-    }{
-      \jcc\Delta\Gamma P U \spinenil {\Delta''} {F'} P
+    }{
+      \jcc\Delta\Gamma U \spinenil {\Delta'} {F}
     }
@@
   \begin{mathpar}
-    \infer[UVar]{ }{
-      \jcr\Delta\Gamma{\app {\var x} S} \Delta {\app {\var x} S}
-    }
-
-    \infer[UConst]{ }{
-      \jcr\Delta\Gamma{\app {\cst c} S} \Delta {\app {\cst c} S}
-    }
-
-    % TODO UMeta?
-
-    \infer[UEval]{
-      (\fct f : A = T)\in\Sigma
-      \and
-      \jcc\Delta\Gamma A T S {\Delta'} F P
-    }{
-      \jcr\Delta\Gamma{\app {\fct f} S} {\Delta'} F
-    }
+    % \infer[UVar]{ }{
+    %   \jcr\Delta\Gamma{\app {\var x} S} \Delta {\app {\var x} S}
+    % }
+
+    % \infer[UConst]{ }{
+    %   \jcr\Delta\Gamma{\app {\cst c} S} \Delta {\app {\cst c} S}
+    % }
+    % \infer[UEval]{
+    %   (\fct f : A = T)\in\Sigma
+    %   \and
+    %   \jcc\Delta\Gamma A T S {\Delta'} F P
+    % }{
+    %   \jcr\Delta\Gamma{\app {\fct f} S} {\Delta'} F
+    % }
+
+    \infer[UAtom]{
+      \jf{\fct{f}}\Delta\Gamma F {\Delta'} {F'} P
+    }{
+      \jcr\Delta\Gamma F {\Delta'}{F'}
+    }\quad
+    \begin{array}{l}
+      \text{$\fct f$ fresh} \\
+      F'\neq\smeta X\sigma
+    \end{array}
+
+    \infer[UMeta]{
+      \jf{\fct{f}}\Delta\Gamma F {\Delta'} {\smeta X \sigma} P
+    }{
+      \jcr\Delta\Gamma F {\Delta'}{\Delta'(\smeta X \sigma)}
+    }\quad
+    {\text{($\fct f$ fresh)}}
 
@@
     }
-
   \end{mathpar}
-
-  \caption{Evaluation of terms}
+  \caption{Evaluation of \CL\ terms}
   \label{fig:evaluation}
@@
 
-% TODO
+Rule \textsc{FAppEval} calls the evaluation of a function, written as
+a \CL\ term, on a spine. The rules of Fig. \ref{fig:evaluation} follow
+the call-by-name strategy, so that an atom is only head-reduced when
+destructed by a case analysis. We finish by a soundness result:
 
 \begin{theorem}
-  If $\mr$ well-constructed, and $\commit\mr F = \mmr$, then
-  $\checkout{\mmr}$ is a value.
+  If $\mr$ well-constructed, and $\commit\mr F$ succeeds with $\mmr$,
+  then $\checkout{\mmr}$ is a closed value, well-typed \wrt\ \LF.
 \end{theorem}
@@
 \section{Related work}
-% 1-2 pages
+
+\paragraph{Symbolic manipulation of proofs}
+
+Several attempts to design a functional programming language are
+proposed to compute over valid proof
+derivations~\cite{cave2012,stampoulis2012,poswolsky2008}. Two
+essential design aspects of these tools are the choice of the
+representation for proof trees and the choice of the meta-language to
+manipulate them. Like them, we commit ourselves to Higher-Order
+Abstract Syntax (HOAS), and more specifically to a language of the
+\LF\ family.
+
+Unlike them, we follow a contract-based approach~\cite{wadler2009well}
+to guarantee the validity of the proofs generated by the programs
+written using our framework. Thus, our technique is similar to
+\LF-based proof-carrying code~\cite{necula1997proof,appel1999proof},
+and to the \sysname{LCF} architecture: an untrusted proof generator is
+composed with a (small) trusted proof checker. A noticeable difference
+is the fact that, in our setting, the checking of proofs is
+interleaved with their generation.
+
+This recent work~\cite{cave2012,stampoulis2012,poswolsky2008} tackles
+the ambitious challenge of deciding statically if a proof-generating
+function will always produce valid proof terms. On one hand, these
+approaches are more satisfying than contract-based ones because of the
+static guarantees. Besides, they are also more efficient since they
+totally remove the need for dynamic verification of proof
+well-formedness. On the other hand, they are based on more sophisticated
+type systems, which augments the trusted base, and put more
+constraints on the way programs are written.
+
+% We write our proof-generating functions in a functional language
+% equipped with a pattern matching capable of dealing with binders. Our
+% syntax for patterns is close to Delphin's~\cite{poswolsky2008} or
+% FreshML's~\cite{shinwell2003} in the sense that it uses a meta-level
+% operation reminiscent of the $\nabla$ quantifier~\cite{gacek2011}
+% instead of the meta-level binding operation~\cite{jay2009}in order to
+% introduce bound names in the evaluation context. Unlike FreshML, our
+% operational semantics does not allow bound names to escape their
+% scope: a variable escaping a local scope introduced by a pattern
+% matching would be rejected by dynamic contract-based verification just
+% before the function exit. Yet, in contrast with Pure
+% FreshML~\cite{pottier2007} or NomPa~\cite{pouillard2010}, we do not
+% enforce this non-escaping property statically but dynamically.
+
+\paragraph{Incremental computation}
+
+Memoization is a widely used technique to reuse previously computed
+outputs. The slicing of our values implements the same idea: naming
+intermediate results of computation (derivations), and reuse of
+computation results by metavariables. Whereas memoization embeds the
+recognition of a previously seen input by searching in the associative
+table, we delegate this to the user (or a front-end) who is
+responsible for giving the delta: these deltas are more involved to
+infer in the higher-order case, and may invoke reasons that are
+beyond the scope of a basic memoized type-checker to justify the reuse
+of a specific typing derivation. For instance, if term~$\tlam x A M$
+has been assigned the type $A \rightarrow B$ leading to a typing
+derivation $\md$, one can specialize its type to $A'\rightarrow B$ by
+applying the subtyping rule directly on $\md$ and providing a proof
+that $A' \rightarrow B \leq A \rightarrow B$.
+
+There is more to incrementality than just memoization. Adaptive
+functional programming techniques~\cite{acar2006,carlsson2002}
+includes memoization but also, thanks to dependency tracking, allows
+to recompute the output only for changed parts without even
+considering unchanged parts. We do have an explicit representation of
+changes (the deltas), but they still must describe the path down to
+the changed part.
 
@@
 % 0.5 pages
+
+We have presented a lightweight framework to program incremental
+certifying type-checker. A prototype implementation is available
+online\footnote{\url{http://www.cs.unibo.it/~puech/}} as well as
+examples of type-checkers for several variants of
+$\lambda$-calculus. This prototype comes as an \sysname{OCaml} library
+with its \sysname{CamlP4} syntax extension, offering the commit and
+checkout operations on repositories given an \LF\ signature and
+\sysname{OCaml} implementations of the defined functions.
+
+This system is meant to serve as a kernel for checking deltas inferred
+by a front-end whose development is still to be done. This piece of
+software will be responsible for comparing several versions of the same
+source code to guess which modifications have been applied to it.
 
@@
 
+\pagebreak
+\appendix
+\section{Appendix}
+
+\begin{figure}
+  \centering
+%  \small
+  \begin{mathpar}
+    \infer{ }{\jts{\cst o} {\even}}
+
+    \infer{\jts M \odd} {\jts {\s M} \even}
+
+    \infer{\jts M \even} {\jts {\s M} \odd}
+
+    \infer{\jts M \nat} {\jts {\s M} \nat}
+
+    \infer{\jts M {A\to B} \and \jts N A} {\jts {\app M N} B}
+
+    \infer{\infer*{}{\mbox{$
+          \begin{array}{c}
+            [\jts {\var x} A] \\[-0.3em]
+            \vdots\\
+            \jts M B
+          \end{array}$
+        }}} {\jts {\lam x M} {A\to B}}
+
+    \infer{\jts M A \and \jsub A B}{\jts M B}
+
+    \infer{\jts M \nat
+      \and
+      \jts N A
+      \and
+      \infer*{}{\mbox{$
+        \begin{array}{c}
+          [\jts {\var x} \nat] \quad
+          [\jts {\var y} A] \\[-0.3em]
+          \vdots \\
+          \jts P A
+        \end{array}
+        $}}
+    }{\jts {\recb M N x y P} A}
+
+    \infer{ }{\jsub A A}
+
+    \infer{ }{\jsub \even\nat}
+
+    \infer{ }{\jsub \odd\nat}
+
+    \infer{\jsub {A'} A \and \jsub B {B'} }{\jsub {A\to B}{A'\to B'}}
+  \end{mathpar}
+  \caption{System \sysname{T_{<:}}, declaratively}
+  \label{fig:tsub}
+\end{figure}
+
+\paragraph{Substitution}
+
+We overload \emph{substitution} as a partial operation on objects,
+given a parallel substitution $\sigma$. Since the syntax of object is
+not stable by (textual) substitution, it is defined as
+\emph{hereditary}\footnote{ Note that this notion of substitution,
+  present in \cite{pfenning2007term}, differs from the more standard
+  definition in \cite{hl07mechanizing} as it presupposes $\eta$-long
+  normal forms: once a \emph{cut} starts, there must be exactly as
+  many formal and actual arguments. A non-empty spine cut against
+  another non-empty spine $\gcut {\app H S} {S'}$ blocks the
+  substitution instead of reducing to $\app H {(\gconcat S S')}$ (an
+  implicit $\eta$-expansion); conversely, exhausting actual but not
+  formal arguments as in $\gcut{\lam x M}{\gnil}$ blocks the
+  substitution too. For example, the reduction $\gcut {\lam x \app
+    {\var x}{\var x}} {\lam x \app {\var x}{\var x}}$ gets stuck
+  because the second $\var x$ in the left part is substituted by a
+  function yet is not itself a function. }: it is mutually recursive
+with a \emph{cut} function $\gcut M S$ which can in turn trigger a
+chain of substitutions, leading to a canonical objects or an error.
+\begin{align*}
+  \gsubst {(\lam x M)} \sigma &= \lam x {\gsubst M \sigma} &
+  \text{if $\var x\notin \dom{\sigma}$}
+  \\
+  \gsubst {(\app {\cst c} S)} \sigma &= \app {\cst c} {(\gsubst S
+    \sigma)}
+  \\
+  \gsubst {(\app {\var x} S)} \sigma &=
+  \app {\var x} {\gsubst S \sigma} &
+  \text{if $\var x\notin \dom{\sigma}$}
+  \\
+  \gsubst {(\app {\var x} S)} \sigma &=
+  \gcut M {\gsubst S \sigma} &
+  \text{if $\sigma(\var x) = M$}
+  \\
+  \gsubst {\smeta X {\sigma}} {\sigma'} &=
+  \smeta X {\gcomp\sigma{\sigma'}}
+  \\
+  \gsubst\spinenil\sigma &=
+  \spinenil \\
+  \gsubst{\spinecons M S}\sigma &=
+  \spinecons{\gsubst M\sigma}{\gsubst S\sigma}
+  \\
+  \gcut {\lam x M} {\spinecons N S} &=
+  \gcut {\gsubst M {\msubst x N}} S
+  \\
+  % TODO cas meta/S ? rem: on en a pas besoin ds le proto.
+  % \gcut {\smeta X\sigma} S &=
+  % \smeta X{\gcomp\sigma\sigma'}
+  % \\
+  \gcut {\app H S} \spinenil &=
+  \app H S
+  \\
+  \gcut {\smeta X\sigma} \spinenil &= \smeta X\sigma
+  \\
+  \gcomp \sigma \msubstnil &=
+  \sigma
+  \\
+  \gcomp \sigma {(\msubstcons{\sigma'} x M)} &=
+  \gcomp {\gsubst\sigma {\msubst x M}} \sigma'
+  \\
+  \gsubst {(\msubstcons{\sigma'} x M)} \sigma &=
+  \msubstcons{\gsubst{\sigma'}\sigma} x {\gsubst M \sigma}
+  \\
+  \gsubst {\msubstnil}{\sigma} &= \msubstnil
+\end{align*}
+
+We extend this operation trivially on families $\gsubst A \sigma$ and
+kinds $\gsubst K \sigma$. We also allow to abusively write
+non-canonical application $\app M S$ as a shorthand for $\gcut M S$
+if it is defined.
+
+% TODO pourquoi cette def?
+% TODO terminaison?
+% TODO stabilité de la eta?
+
 \end{document}

### 2012-06-16 doc/papers/LFMTP2012/paper.tex
paper: dernieres corrections

@@
 ``church-style'' calculi. Take for example dependent type systems,
-relying definitional equality on types, or type systems with subtype
-polymorphism: both rely on a rule of the form
+relying on definitional equality on types, or type systems with subtype
+polymorphism: both feature a rule of the form
 $$
@@
 allows to take only the result of the algorithm into
-consideration. But let us choose another path for a minute:
-derivations in the declarative systems, if they cannot always be
-easily inferred, can be \emph{verified} if provided by an oracle. If
-the oracle is wrong, we can tell cheaply by verifying the witness
-typing derivation it provides. Let us then imagine a type-checker to
-be the pair of \emph{(i)} an untrusted procedure $\finfer M$ to infer
-a typing derivation out of program $M$; it does not have to be
-complete, or even sound, and \emph{(ii)} a trusted and fast kernel
-$\fcheck{\md}$ to verify these derivations \wrt\ a given set of
-inference rules $\Sigma$. This is a \emph{certifying} scheme (see
-\cite{leroy2006compcert} for a discussion on approaches to reliable
-compilers): if $\fcheck{\finfer M} = \cst{true}$ then there really
-\emph{is} a derivation $\md$ for $M$, but the converse is not
-necessarily true: function $\finfer{}\!\!$ could fail, or return an
-incorrect derivation. Explicitly generating the derivation in the
-declarative system has the following advantages over the first
-approach:
+consideration. But let us choose another path: derivations in the
+declarative systems, if they cannot always be easily inferred, can be
+\emph{verified} if provided by an oracle. We can tell cheaply if it is
+wrong by verifying the witness typing derivation it provided. Let us
+then imagine a type-checker to be the pair of \emph{(i)} an untrusted
+procedure $\finfer M$ to infer a typing derivation out of program $M$;
+it does not have to be complete, or even sound, and \emph{(ii)} a
+trusted and fast kernel $\fcheck{\md}$ to verify these derivations
+\wrt\ a given set of inference rules $\Sigma$. This is a
+\emph{certifying} scheme (see \cite{leroy2006compcert}). If $\fcheck{\finfer
+  M} = \cst{true}$ then there really \emph{is} a derivation $\md$ for
+$M$, but the converse is not necessarily true: function
+$\finfer{}\!\!$ could fail, or return an incorrect
+derivation. Explicitly generating the derivation in the declarative
+system has the following advantages over the first approach:
 \begin{itemize}
-\item it is lightweight: the untrusted algorithm is usually shorter
-  than the corresponding equivalence proof
-\item compilers often need to propagate type information further in
+\item It is lightweight: the untrusted algorithm is usually shorter
+  than the corresponding equivalence proof.
+\item Compilers often need to propagate type information further in
   the chain of transformation; the typing derivation produced by
@@
   being exploited.
-\item it inherits benefits from certifying approaches like
+\item It inherits benefits from certifying approaches like
   proof-carrying code \cite{necula1997proof}: certificates can be
@@
 the programmer to write a well-typed program in one try, compile it
-and run it: writing a program longer than a few dozens of lines
-becomes a tight and non-linear interaction between the human and the
-type-checker involving constant experiments, fixes, etc. If this
-interaction is possible for short programs and fast checkers, it
-becomes increasingly hindered by the latency of complete rechecks:
-even if the modifications made to the program between two interactions
-are small, the whole program is usually rechecked. This is especially
-true for languages embedding formal verification aspects, like proof
-assistants: there, tight interaction with a type or proof checker is
-unavoidable due to the difficulty of writing correct proofs without
-guidance from the system, and the time taken by many tools to infer
-easy parts of the proofs (proof search). The constant modification of
-the source makes it necessary to observe \emph{incrementally} the
-effect of a small change on the whole edifice: we need to take
-advantage of the results of previous checks in order to build the
-result for a modified version of the program.
+and run it: writing a program becomes a tight and non-linear
+interaction between the human and the type-checker involving constant
+experiments, fixes, etc. If this interaction is possible for short
+programs and fast checkers, it becomes increasingly hindered by the
+latency of complete rechecks: even if the modifications made to the
+program between two interactions are small, the whole program is
+usually rechecked. This is especially true for languages embedding
+formal verification aspects, like proof assistants: there, tight
+interaction with a type or proof checker is unavoidable due to the
+difficulty of writing correct proofs without guidance from the system,
+and the time taken by many tools to infer easy parts of the proofs
+(proof search). The constant modification of the source makes it
+necessary to observe \emph{incrementally} the effect of a small change
+on the whole edifice: we need to take advantage of the results of
+previous checks in order to build the result for a modified version of
+the program.
 
 Several generic methods have been devised for turning a batch program
-into an incremental one \cite{acar2006,carlsson2002}. Unfortunately,
-they suffer from their genericity on the particular problem of type
+into an incremental one
+\cite{pugh1989incremental,acar2006,carlsson2002}. Unfortunately, they
+suffer from their genericity on the particular problem of type
 checking:
 \begin{itemize}
-\item first, they ignore the higher-order nature of program syntax,
+\item First, they ignore the higher-order nature of program syntax,
   and the kind of structural quotients we make on syntax trees with
   binders and environments ($\alpha$-equivalence, weakening\ldots): a
-  hypothetical derivation of\ \ $\vdash x+2 : \cst{nat}$ is still valid
-  when put under a binder\ \ $\vdash\lam y (x + 2) -
-  y:\cst{nat}\to\cst{nat}$ and should not be regenerated;
-\item for very large programs, we might want to feed the type checker
-  only with a small \emph{delta} on the previous version and not the
-  whole program, thus saving the time to recognize already checked
-  parts;
-\item finally, there is no way to a posteriori validate the result of
+  hypothetical derivation of\ \ $\vdash x+2 : \cst{nat}$ is still
+  valid when put under a binder $\lam y (x + 2) - y$ and should not be
+  regenerated.
+\item Since it is easy to track edits on a very large programs (trace
+  of the editor), we may want to feed the type checker only with a
+  small \emph{delta} on the previous version and not the whole
+  program, thus saving the time to recognize already checked parts;
+\item Finally, there is no way to a posteriori validate the result of
   the incremental process: the user has to trust not only the
@@
 procedure, and storing these derivations between calls to the
-procedure, we could recognize already-built subderivations for
+procedure, we can reuse already-built subderivations for
 subterms common to an old, checked version and a new, to-be-checked
@@
 constantly maintain the verifiable witness of well-typing for the
-program, and take advantage of the quotients made on higher-order
-syntax.
+program, and take advantage of the quotients made on syntax with
+binders.
 
@@
 two sets $\cst{even}$ and $\cst{odd}$, introduction $\z$ and $\s{M}$,
-elimination ($\recb M N x y P$), and subtyping on these types (for
-formal details, see appendix). He (or she) wants to write a program
-computing $(2+2)\times 3$ in that system. He knows that sometimes the
-arguments of $\cst{rec}$ are quite hard to get correct, but
-\sysname{T_{<:}} is a typed language; he wants to take advantage of
-this to build his program incrementally using the teacher's provided
-incremental verifier. As a first attempt, he wants to verify $P_1 =
-\recb{\s\z}{\s\z} x y {\s {\var x}}$: he sends the command
-$\finfer{P_1}$ to the system. The $\finfer{}$ operator takes a
-well-formed program and tries to construct a valid typing derivation
-for its argument in the current, empty context. The system succeeds
-and returns a typing derivation $\md_1$ of \/ $\vdash P_1 :
-\cst{nat}$, valid in an empty context.
+elimination ($\recb M N x y P$), and subtyping on these types. He (or
+she) wants to write a program computing $(2+2)\times 3$ in that
+system. He knows that sometimes the arguments of $\cst{rec}$ are quite
+hard to get correct, but \sysname{T_{<:}} is a typed language; he
+wants to take advantage of this to build his program incrementally
+using the teacher's provided incremental verifier. As a first attempt,
+he wants to verify $P_1 = \recb{\s\z}{\s\z} x y {\s {\var x}}$: he
+sends the command $\finfer{P_1}$ to the system. The $\finfer{}$
+operator takes a well-formed program and tries to construct a valid
+typing derivation for its argument in the current, empty context. The
+system succeeds and returns a typing derivation $\md_1$ of \/ $\vdash
+P_1 : \cst{nat}$, valid in an empty context.
 
@@
 \cst{odd}$, but $\var{add}$ awaits two $\cst{nat}$ arguments. When the
-system generates this derivation, it inserts a coercion $\cst{odd}
-\leq \cst{nat}$ at these two points. Taking advantage of $\md_2$ and
-$\md_4$ the system succeeds with a new derivation made out of these,
-without having to regenerate them.
+system generates the derivation for this application, it inserts a
+coercion $\cst{odd} \leq \cst{nat}$ at these two points. Taking
+advantage of $\md_2$ and $\md_4$ the system succeeds with a new
+derivation made out of these, without having to regenerate them.
 
@@
 of terms, possibly containing special typed operators having a
-computational content: $\finfer{\cdot}$ takes a term $M$ and produces
-the (dependent) pair of a type $A$ and a derivation of the judgment
-$\vdash M:A$ in the current context; $\fget{\cdot}$ does the exact
+computational content: $\finfer{}$ takes a term $M$ and produces the
+(dependent) pair of a type $A$ and a derivation of the judgment
+$\vdash M:A$ in the current context; $\fget{}$ does the exact
 converse: it takes a derivation $\vdash M:A$ and projects it back as
@@
 mechanism becomes clear on very large programs: then the cost of
-rechecking well-typing of a small modification amounts to rechecking
-the path in the term from the root to the changed part, and the
-changed part itself.
+rechecking well-typing of a small modification amounts to recheck the
+path in the term from the root to the changed part, and the changed
+part itself.
 
@@
 derivation enables thus to think of the relationship between the input
-and the output. Feeding subterms of the output $\md_i$ back into
-subterms of the input $P_i$ is a well-known technique for handling
-changes in the input: combined with a way to automatically recognize
-already-computed input, it is \emph{memoization}, a general-purpose
-technique for incremental computation that has been extensively
-studied and combined with other techniques \cite{acar2006}. However,
-it is best understood in the case of purely first-order data, and to
-our best knowledge remains unexplored for higher-order structures
+and the output. Feeding subresults of the output $\md_i$ back into the
+input $P_i$ is a well-known technique for handling changes in the
+input: combined with a way to automatically recognize already-computed
+input, it is \emph{memoization}, a general-purpose technique for
+incremental computation that has been extensively studied and combined
+with other techniques \cite{acar2006,pugh1989incremental}. However, it
+is best understood in the case of purely first-order data, and to our
+best knowledge remains mainly unexplored for higher-order structures
 containing binders, like programs and proofs.
@@
 Patterns do not have to be exhaustive since we are in an untrusted
-setting. Fr conciseness, we use case failure for signaling typing
+setting. For conciseness, we use case failure for signaling typing
 error. There are three cases for $\s{M}$ in the system, corresponding
@@
 generate a subtyping rule. This is done by matching on the returned
-derivation $\md_\leq$ in the case of the \textsc{Refl} rule.
+derivation $\md_\leq$, in the case of the \textsc{Refl} rule.
 
@@
 the recursive call $\finfer{M}$ must be done in an enlarged
-environment where $\vdash x : A$. Otherwise the check that the
-returned open derivation $\md$ is correct would fail ($x$ is
-unknown). We thus introduce a new construct $\matchin M \Gamma \ldots$
-($\match M \ldots$ being syntactic sugar for $\matchin M \envnil
-\ldots$). Even then, in the recursive call, we would have no way of
-synthesizing a variable $\var x$'s type anymore. Thus we substitute
-all occurrences of $\var x$ with the \emph{output} of function
-$\finfer{}$ on it (since it is known at this point), \ie\ a pair of
-$A$ and $\md_1$. For this substitution to be well-typed, we now need
-to coerce this output to a valid input (a term). We introduce an
-operator which is in some sense the \emph{inverse} of $\finfer{}$:
-operator $\fget{}$ maps a pair of a type and a derivation to the term
-it types.
+environment where $\vdash x : A$. Otherwise the check that the open
+derivation $\md$ returned by the recursive call is correct would fail
+($x$ is unknown). We thus introduce a new construct $\matchin M \Gamma
+\ldots$ ($\match M \ldots$ being syntactic sugar for $\matchin M
+\envnil \ldots$). Even then, we would have no way of synthesizing a
+variable $\var x$'s type anymore ($\caseb{\var x} \text{???}$). We
+therefore substitute all occurrences of $\var x$ with the
+\emph{output} of function $\finfer{}$ on it (since it is known at this
+point), \ie\ a pair of $A$ and $\md_1$. For this substitution to be
+well-typed, we now need to coerce this output to a valid input (a
+term). We introduce an operator which is in some sense the
+\emph{inverse} of $\finfer{}$: operator $\fget{}$ maps a pair of a
+type and a derivation to the term it types.
 
@@
 is a unique more general type, in our particular case of subtyping
-relation, this means iterating only twice: we will then attain
+relation, this means iterating only twice: we will then reach
 \cst{nat}.
@@
 $\finfer{}$: we will call it its inverse, since it maps the return
-type to the arguments. It has definition:
+type to the argument. It has definition:
 
@@
 The idea of untrustedly generating certificates verified \emph{a
-  posteriori} is very wide-spread in \sysname{LCF}-style proof
-assistants: during manual proof search, the interaction is driven by
+  posteriori} is well-known to \sysname{LCF}-style proof assistants:
+during manual proof search, the interaction is driven by
 \emph{tactics}: they are untrusted programs generating pieces of
@@
 requires evaluating the program deltas with a carefully designed
-strategy (see \ref{sec:comput}).
+strategy (see \ref{sec:eval}).
 
@@
 The Edinburgh Logical Framework \cite{harper1993framework}, or \LF\
-for short, is one of the well-accepted solution to this question. It
+for short, is one of the well-accepted solutions to this question. It
 is a dependently typed lambda-calculus for \emph{representing}
-higher-order terms and derivations
-\cite{pfenning1988higher}. The user gives a
-\emph{signature} defining abstract syntax and inference rules of his
-\emph{object language}, and gets an algorithm to verify terms and
+higher-order terms and derivations \cite{pfenning1988higher}. The user
+gives a \emph{signature} defining abstract syntax and inference rules
+of his object language, and gets an algorithm to verify terms and
 derivations expressed in the \LF\ syntax.
@@
 metavariables (written $\meta X$, $\meta Y$) standing for open terms,
-following \cite{nanevski2008contextual}, secondly, we make sure that
+following \cite{nanevski2008contextual}. Secondly, we make sure that
 every subterm is named by a metavariable (an operation called
@@
 \emph{value} is an object containing no function symbols. The
-definition of hereditary substitution $\gsubst M \sigma$ is skipped
-and can be found in the appendix.
-
-We distinguish syntactically between \emph{checkable} $M$ and
-\emph{inferrable} $F$ objects (resp. \emph{canonical} and
-\emph{atomic}), as in \eg\ \cite{pierce2000local}. Canonical objects
+definition of hereditary substitution $\gsubst M \sigma$ is standard,
+and it is eluded here.
+
+We distinguish syntactically between \emph{checkable} and
+\emph{inferrable} objects (resp. \emph{canonical} $M$ and
+\emph{atomic} $F$), as in \eg\ \cite{pierce2000local}. Canonical objects
 are checked against a given type, whereas we can synthesize the type
@@
 If $\mr = (\Delta, X[\sigma])$, we write $\hat {\mr}$ for the tip
-$X[\sigma]$, abusively $\mr$ for $\Delta$ and $\mr (\smeta X \sigma)$
-for $\gsubst M\sigma$ if $\Delta(\meta X) = (\Gamma\vdash M : A)$.
-The algorithm in \ref{sec:eval} will maintain two invariants: a
-repository $\mr$ is \emph{acyclic}, (for all $\meta X$ in
-$\mr$, the fixpoint of $\FMV{\mr(\meta X)}$ does not contain $\meta
-X$) and all $F$ and $P$ in it are values.
+$X[\sigma]$, and $\mr (\smeta X \sigma)$ for $\gsubst M\sigma$ if
+$\Delta(\meta X) = (\Gamma\vdash M : A)$.  The algorithm in
+\ref{sec:eval} will maintain two invariants: a repository $\mr$ is
+\emph{acyclic}, (for all $\meta X$ in $\mr$, the fixpoint of
+$\FMV{\mr(\meta X)}$ does not contain $\meta X$) and all $F$ and $P$
+in it are values.
 
@@
 $
-(
+((
 \meta X \mapsto
@@
 \cst{tm}
-), \smeta X {}
+), \smeta X {})
 $
@@
 
-Secondly, we introduce the \CL\ language compute over our \SLF\
+Secondly, we introduce the \CL\ language to compute over our \SLF\
 objects (Fig.~\ref{fig:syntax-CL}). We can construct atomic objects
 out of others, and deconstruct them by case analysis. The definition
-$T$ of a function always begins with a series of $\lambda$.
+$T$ of a function always begins with a series of $\lambda$ (as many as
+there are arguments in its type).
 
@@
 {\fct f} {\var x}$. Nevertheless, we want to maintain the invariant
-that each time an object goes in or out of a function $\fct f : A_1\to
-A_2$, it is checked against the type of this function (resp. $A_1$ or
-$A_2$). You can think of interpreted functions as mere black boxes
-taking and producing objects, and being able to feed objects to other
-black boxes; yet the wires linking them are labeled with types and
-objects circulating in them are checked against that type.
+that each time an object goes in (resp. out) of a function $\fct f :
+A_1\to A_2$, it is checked against the type of this function ($A_1$,
+resp. $A_2$). You can think of interpreted functions as mere black
+boxes taking and producing objects, and being able to feed objects to
+other black boxes; yet the wires linking them are labeled with types,
+and objects circulating in them are checked against these types.
 
@@
 
-We now turn to the bidirectional typed reduction and slicing of the
-algorithm (Fig. \ref{fig:obj-typing}, \ref{fig:obj-typing2},
+We now turn to the bidirectional typed reduction and slicing algorithm
+(Fig. \ref{fig:obj-typing}, \ref{fig:obj-typing2},
 \ref{fig:evaluation}), presented as an inference system. All judgments
-are of the form $\gjao L U V W$ where $U$, $V$ and $L$ are inputs, are
-directed by the syntax of $V$ and $L$, and $W$ is the output. They are
-parameterized by a constant and implicit signature $\Sigma$.
+are of the form $\gjao {L} {\mathcal U} {\mathcal V} {\mathcal W}$
+where $\mathcal U$, $\mathcal V$ and $L$ are inputs, are directed by
+the syntax of $\mathcal V$ and $L$, and $\mathcal W$ is the
+output. They are parameterized by a constant and implicit signature
+$\Sigma$.
 
@@
 These typing rules use typed definitional equality between families
-$\jea\Delta\Gamma {A_1} {A_2}$ and between spines $\jel\Delta\Gamma A
-{S_1} {S_2} P$. The formal definition of these judgments is eluded
-here. It compares syntactically the values of its arguments. Since it
+$\jea\Delta\Gamma {A_1} {A_2}$ and spines $\jel\Delta\Gamma A {S_1}
+{S_2} P$. The formal definition of these judgments is eluded here. It
+compares syntactically the respective values of its objects. Since it
 potentially triggers function evaluation (and thus typing
@@
 a \CL\ term, on a spine. The rules of Fig. \ref{fig:evaluation} follow
-the call-by-name strategy, so that an atom is only head-reduced when
-destructed by a case analysis. We finish by a soundness result:
+the call-by-name strategy, so that when destructed by a case analysis
+an atom is only head-reduced. We finish by a soundness result:
 
@@
 
-Several attempts to design a functional programming language are
-proposed to compute over valid proof
+Several attempts to design a functional programming language are being
+developed to compute over valid proof
 derivations~\cite{cave2012,stampoulis2012,poswolsky2008}. Two
@@
 Abstract Syntax (HOAS), and more specifically to a language of the
-\LF\ family.
-
-Unlike them, we follow a contract-based approach~\cite{wadler2009well}
-to guarantee the validity of the proofs generated by the programs
-written using our framework. Thus, our technique is similar to
-\LF-based proof-carrying code~\cite{necula1997proof,appel1999proof},
-and to the \sysname{LCF} architecture: an untrusted proof generator is
-composed with a (small) trusted proof checker. A noticeable difference
-is the fact that, in our setting, the checking of proofs is
-interleaved with their generation.
-
-This recent work~\cite{cave2012,stampoulis2012,poswolsky2008} tackles
-the ambitious challenge of deciding statically if a proof-generating
-function will always produce valid proof terms. On one hand, these
-approaches are more satisfying than contract-based ones because of the
-static guarantees. Besides, they are also more efficient since they
-totally remove the need for dynamic verification of proof
-well-formedness. On the other hand, they are based on more sophisticated
-type systems, which augments the trusted base, and put more
-constraints on the way programs are written.
+\LF\ family. This recent work tackles the ambitious challenge of
+deciding statically if a proof-generating function will always produce
+valid proof terms. Unlike them, we follow a contract-based
+approach~\cite{wadler2009well} to guarantee the validity of the
+generated proofs. Thus, our technique is similar to \LF-based
+proof-carrying code~\cite{necula1997proof,appel1999proof}, and to the
+\sysname{LCF} architecture: an untrusted proof generator is composed
+with a small, trusted proof checker. A noticeable difference is the
+fact that, in our setting, the checking of proofs is interleaved with
+their generation, allowing earlier adn easier bug catching during
+their development. On one hand, these approaches are more satisfying
+and effective because of the static guarantees. On the other hand,
+they are based on more sophisticated type systems, which augments the
+trusted base, and put more constraints on the way programs are
+written. When programming type checkers as well as in other examples,
+we encounter the problem of linking a variable $\var x$ to a typing
+hypothesis $\vdash \var x : A$ to retrieve its type. In
+\cite{cave2012}, a machinery of environment-side $\Sigma$-types is
+proposed. We replace it by the use of inverse function which seems
+more general, but puts more burden on the reduction strategy.
 
@@
 Memoization is a widely used technique to reuse previously computed
-outputs. The slicing of our values implements the same idea: naming
-intermediate results of computation (derivations), and reuse of
-computation results by metavariables. Whereas memoization embeds the
-recognition of a previously seen input by searching in the associative
-table, we delegate this to the user (or a front-end) who is
-responsible for giving the delta: these deltas are more involved to
-infer in the higher-order case, and may invoke reasons that are
-beyond the scope of a basic memoized type-checker to justify the reuse
-of a specific typing derivation. For instance, if term~$\tlam x A M$
-has been assigned the type $A \rightarrow B$ leading to a typing
-derivation $\md$, one can specialize its type to $A'\rightarrow B$ by
-applying the subtyping rule directly on $\md$ and providing a proof
-that $A' \rightarrow B \leq A \rightarrow B$.
+output \cite{pugh1989incremental}. The slicing of our values
+implements the same idea: naming intermediate results of computation
+(derivations), and reuse of these computation results by
+metavariables. Whereas memoization embeds the recognition of a
+previously seen input, we delegate this to the user (or a front-end)
+who is responsible for providing the delta: these deltas are more
+involved to infer in the higher-order case, and may invoke reasons
+that are beyond the scope of a basic memoized type-checker to justify
+the reuse of a specific typing derivation. For instance, if
+term~$\tlam x A M$ has been assigned the type $A \rightarrow B$
+leading to a typing derivation $\md$, one can specialize its type to
+$A'\rightarrow B$ by applying the subtyping rule directly on $\md$ and
+providing a proof that $A' \rightarrow B \leq A \rightarrow B$.
 
@@
 online\footnote{\url{http://www.cs.unibo.it/~puech/}} as well as
-examples of type-checkers for several variants of
+examples of type-checkers for several variants of the
 $\lambda$-calculus. This prototype comes as an \sysname{OCaml} library
-with its \sysname{CamlP4} syntax extension, offering the commit and
-checkout operations on repositories given an \LF\ signature and
-\sysname{OCaml} implementations of the defined functions.
+with a \sysname{CamlP4} syntax extension, offering the \emph{commit}
+and \emph{checkout} operations on repositories, given an \LF\
+signature and \sysname{OCaml} implementations for the defined
+functions.
 
@@
 
-\pagebreak
-\appendix
-\section{Appendix}
-
-\begin{figure}
-  \centering
-%  \small
-  \begin{mathpar}
-    \infer{ }{\jts{\cst o} {\even}}
-
-    \infer{\jts M \odd} {\jts {\s M} \even}
-
-    \infer{\jts M \even} {\jts {\s M} \odd}
-
-    \infer{\jts M \nat} {\jts {\s M} \nat}
-
-    \infer{\jts M {A\to B} \and \jts N A} {\jts {\app M N} B}
-
-    \infer{\infer*{}{\mbox{$
-          \begin{array}{c}
-            [\jts {\var x} A] \\[-0.3em]
-            \vdots\\
-            \jts M B
-          \end{array}$
-        }}} {\jts {\lam x M} {A\to B}}
-
-    \infer{\jts M A \and \jsub A B}{\jts M B}
-
-    \infer{\jts M \nat
-      \and
-      \jts N A
-      \and
-      \infer*{}{\mbox{$
-        \begin{array}{c}
-          [\jts {\var x} \nat] \quad
-          [\jts {\var y} A] \\[-0.3em]
-          \vdots \\
-          \jts P A
-        \end{array}
-        $}}
-    }{\jts {\recb M N x y P} A}
-
-    \infer{ }{\jsub A A}
-
-    \infer{ }{\jsub \even\nat}
-
-    \infer{ }{\jsub \odd\nat}
-
-    \infer{\jsub {A'} A \and \jsub B {B'} }{\jsub {A\to B}{A'\to B'}}
-  \end{mathpar}
-  \caption{System \sysname{T_{<:}}, declaratively}
-  \label{fig:tsub}
-\end{figure}
-
-\paragraph{Substitution}
-
-We overload \emph{substitution} as a partial operation on objects,
-given a parallel substitution $\sigma$. Since the syntax of object is
-not stable by (textual) substitution, it is defined as
-\emph{hereditary}\footnote{ Note that this notion of substitution,
-  present in \cite{pfenning2007term}, differs from the more standard
-  definition in \cite{hl07mechanizing} as it presupposes $\eta$-long
-  normal forms: once a \emph{cut} starts, there must be exactly as
-  many formal and actual arguments. A non-empty spine cut against
-  another non-empty spine $\gcut {\app H S} {S'}$ blocks the
-  substitution instead of reducing to $\app H {(\gconcat S S')}$ (an
-  implicit $\eta$-expansion); conversely, exhausting actual but not
-  formal arguments as in $\gcut{\lam x M}{\gnil}$ blocks the
-  substitution too. For example, the reduction $\gcut {\lam x \app
-    {\var x}{\var x}} {\lam x \app {\var x}{\var x}}$ gets stuck
-  because the second $\var x$ in the left part is substituted by a
-  function yet is not itself a function. }: it is mutually recursive
-with a \emph{cut} function $\gcut M S$ which can in turn trigger a
-chain of substitutions, leading to a canonical objects or an error.
-\begin{align*}
-  \gsubst {(\lam x M)} \sigma &= \lam x {\gsubst M \sigma} &
-  \text{if $\var x\notin \dom{\sigma}$}
-  \\
-  \gsubst {(\app {\cst c} S)} \sigma &= \app {\cst c} {(\gsubst S
-    \sigma)}
-  \\
-  \gsubst {(\app {\var x} S)} \sigma &=
-  \app {\var x} {\gsubst S \sigma} &
-  \text{if $\var x\notin \dom{\sigma}$}
-  \\
-  \gsubst {(\app {\var x} S)} \sigma &=
-  \gcut M {\gsubst S \sigma} &
-  \text{if $\sigma(\var x) = M$}
-  \\
-  \gsubst {\smeta X {\sigma}} {\sigma'} &=
-  \smeta X {\gcomp\sigma{\sigma'}}
-  \\
-  \gsubst\spinenil\sigma &=
-  \spinenil \\
-  \gsubst{\spinecons M S}\sigma &=
-  \spinecons{\gsubst M\sigma}{\gsubst S\sigma}
-  \\
-  \gcut {\lam x M} {\spinecons N S} &=
-  \gcut {\gsubst M {\msubst x N}} S
-  \\
-  % TODO cas meta/S ? rem: on en a pas besoin ds le proto.
-  % \gcut {\smeta X\sigma} S &=
-  % \smeta X{\gcomp\sigma\sigma'}
+% \pagebreak
+% \appendix
+% \section{Appendix}
+
+% \begin{figure}
+%   \centering
+% %  \small
+%   \begin{mathpar}
+%     \infer{ }{\jts{\cst o} {\even}}
+
+%     \infer{\jts M \odd} {\jts {\s M} \even}
+
+%     \infer{\jts M \even} {\jts {\s M} \odd}
+
+%     \infer{\jts M \nat} {\jts {\s M} \nat}
+
+%     \infer{\jts M {A\to B} \and \jts N A} {\jts {\app M N} B}
+
+%     \infer{\infer*{}{\mbox{$
+%           \begin{array}{c}
+%             [\jts {\var x} A] \\[-0.3em]
+%             \vdots\\
+%             \jts M B
+%           \end{array}$
+%         }}} {\jts {\lam x M} {A\to B}}
+
+%     \infer{\jts M A \and \jsub A B}{\jts M B}
+
+%     \infer{\jts M \nat
+%       \and
+%       \jts N A
+%       \and
+%       \infer*{}{\mbox{$
+%         \begin{array}{c}
+%           [\jts {\var x} \nat] \quad
+%           [\jts {\var y} A] \\[-0.3em]
+%           \vdots \\
+%           \jts P A
+%         \end{array}
+%         $}}
+%     }{\jts {\recb M N x y P} A}
+
+%     \infer{ }{\jsub A A}
+
+%     \infer{ }{\jsub \even\nat}
+
+%     \infer{ }{\jsub \odd\nat}
+
+%     \infer{\jsub {A'} A \and \jsub B {B'} }{\jsub {A\to B}{A'\to B'}}
+%   \end{mathpar}
+%   \caption{System \sysname{T_{<:}}, declaratively}
+%   \label{fig:tsub}
+% \end{figure}
+
+% \paragraph{Substitution}
+
+% We overload \emph{substitution} as a partial operation on objects,
+% given a parallel substitution $\sigma$. Since the syntax of object is
+% not stable by (textual) substitution, it is defined as
+% \emph{hereditary}\footnote{ Note that this notion of substitution,
+%   present in \cite{pfenning2007term}, differs from the more standard
+%   definition in \cite{hl07mechanizing} as it presupposes $\eta$-long
+%   normal forms: once a \emph{cut} starts, there must be exactly as
+%   many formal and actual arguments. A non-empty spine cut against
+%   another non-empty spine $\gcut {\app H S} {S'}$ blocks the
+%   substitution instead of reducing to $\app H {(\gconcat S S')}$ (an
+%   implicit $\eta$-expansion); conversely, exhausting actual but not
+%   formal arguments as in $\gcut{\lam x M}{\gnil}$ blocks the
+%   substitution too. For example, the reduction $\gcut {\lam x \app
+%     {\var x}{\var x}} {\lam x \app {\var x}{\var x}}$ gets stuck
+%   because the second $\var x$ in the left part is substituted by a
+%   function yet is not itself a function. }: it is mutually recursive
+% with a \emph{cut} function $\gcut M S$ which can in turn trigger a
+% chain of substitutions, leading to a canonical objects or an error.
+% \begin{align*}
+%   \gsubst {(\lam x M)} \sigma &= \lam x {\gsubst M \sigma} &
+%   \text{if $\var x\notin \dom{\sigma}$}
 %   \\
-  \gcut {\app H S} \spinenil &=
-  \app H S
-  \\
-  \gcut {\smeta X\sigma} \spinenil &= \smeta X\sigma
-  \\
-  \gcomp \sigma \msubstnil &=
-  \sigma
-  \\
-  \gcomp \sigma {(\msubstcons{\sigma'} x M)} &=
-  \gcomp {\gsubst\sigma {\msubst x M}} \sigma'
-  \\
-  \gsubst {(\msubstcons{\sigma'} x M)} \sigma &=
-  \msubstcons{\gsubst{\sigma'}\sigma} x {\gsubst M \sigma}
-  \\
-  \gsubst {\msubstnil}{\sigma} &= \msubstnil
-\end{align*}
-
-We extend this operation trivially on families $\gsubst A \sigma$ and
-kinds $\gsubst K \sigma$. We also allow to abusively write
-non-canonical application $\app M S$ as a shorthand for $\gcut M S$
-if it is defined.
-
-% TODO pourquoi cette def?
-% TODO terminaison?
-% TODO stabilité de la eta?
+%   \gsubst {(\app {\cst c} S)} \sigma &= \app {\cst c} {(\gsubst S
+%     \sigma)}
+%   \\
+%   \gsubst {(\app {\var x} S)} \sigma &=
+%   \app {\var x} {\gsubst S \sigma} &
+%   \text{if $\var x\notin \dom{\sigma}$}
+%   \\
+%   \gsubst {(\app {\var x} S)} \sigma &=
+%   \gcut M {\gsubst S \sigma} &
+%   \text{if $\sigma(\var x) = M$}
+%   \\
+%   \gsubst {\smeta X {\sigma}} {\sigma'} &=
+%   \smeta X {\gcomp\sigma{\sigma'}}
+%   \\
+%   \gsubst\spinenil\sigma &=
+%   \spinenil \\
+%   \gsubst{\spinecons M S}\sigma &=
+%   \spinecons{\gsubst M\sigma}{\gsubst S\sigma}
+%   \\
+%   \gcut {\lam x M} {\spinecons N S} &=
+%   \gcut {\gsubst M {\msubst x N}} S
+%   \\
+%   % TODO cas meta/S ? rem: on en a pas besoin ds le proto.
+%   % \gcut {\smeta X\sigma} S &=
+%   % \smeta X{\gcomp\sigma\sigma'}
+%   % \\
+%   \gcut {\app H S} \spinenil &=
+%   \app H S
+%   \\
+%   \gcut {\smeta X\sigma} \spinenil &= \smeta X\sigma
+%   \\
+%   \gcomp \sigma \msubstnil &=
+%   \sigma
+%   \\
+%   \gcomp \sigma {(\msubstcons{\sigma'} x M)} &=
+%   \gcomp {\gsubst\sigma {\msubst x M}} \sigma'
+%   \\
+%   \gsubst {(\msubstcons{\sigma'} x M)} \sigma &=
+%   \msubstcons{\gsubst{\sigma'}\sigma} x {\gsubst M \sigma}
+%   \\
+%   \gsubst {\msubstnil}{\sigma} &= \msubstnil
+% \end{align*}
+
+% We extend this operation trivially on families $\gsubst A \sigma$ and
+% kinds $\gsubst K \sigma$. We also allow to abusively write
+% non-canonical application $\app M S$ as a shorthand for $\gcut M S$
+% if it is defined.
+
+% % TODO pourquoi cette def?
+% % TODO terminaison?
+% % TODO stabilité de la eta?
 
 \end{document}
-
-%  LocalWords:  subtype
